### 2021-11-30 R Code/Randomise_Papers.Rmd.Rmd
initial commit

---
title: Randomly select papers for review
output:
  html_document: default
  pdf_document: default
---

***

## Summary
This document contains code for randomising paper ranks for the review on Dynamic Occupancy Models. For each of four search parameters, papers are stratified into 5 year time periods and randomly ranked. The top 25% of papers in each ranking category will be reviewed; papers which do not qualify for review will be replaced with the next highest rank.

***

```{r setup, echo = FALSE, message = FALSE}
library("knitr")
library("tidyverse")
library("readxl")

opts_knit$set(root.dir = "/Users/kelleherj/Documents/Research/QAEco/ODM_Review/")
```

```{r Read-in excel sheets, echo = FALSE, message = FALSE, warning = FALSE}
# From the excel review spreadsheet, read in only the identifying characteristics 
identifierSheet <- readxl::read_xlsx("Review Spreadsheet.xlsx", 
                                     sheet = "Identifying", skip = 1) %>%
  # Drop rows prior to start of data
  slice(4:nrow(.)) %>%
  # Only need google search query and the ID reference
  select(Review_ID, Query = `Google Scholar search terms`) 
# Repeat with the study metadata
studyDataSheet <- readxl::read_xlsx("Review Spreadsheet.xlsx", 
                                    sheet = "Study data", skip = 1) %>%
  slice(4:nrow(.)) %>%
  # Only need year data and ID reference
  select(Review_ID, Year) 
```

```{r Combine and merge, echo = FALSE, message = FALSE}
# Merge paper queries and metadata
paperList <- full_join(identifierSheet, studyDataSheet, 
                       by = "Review_ID") %>% 
  # Drop empty rows, excluded, and incidental papers
  filter(!is.na(Query) & Query != "INCIDENTAL" & Year != "EXCLUDE") %>%
  # Break into time periods 
  mutate(YearPeriod = case_when(Year %in% as.character(2000:2005) ~ "2000:2005",
                                Year %in% as.character(2005:2010) ~ "2005:2010",
                                Year %in% as.character(2010:2015) ~ "2010:2015",
                                Year %in% as.character(2015:2021) ~ "2015:2021"))
```

```{r Randomly rank papers within strata}
rankedPapers <- paperList %>%
  # Add random ranks for all papers
  mutate(randRank = sample(x = 1:nrow(.), size = nrow(.))) %>%
  # Re-rank within strata
  group_by(Query, YearPeriod) %>%
  arrange(Query, YearPeriod, randRank) %>%
  mutate(groupRank = row_number())

```

### 2021-11-30 R Code/Randomise_Papers.Rmd.Rmd
Added rankings

@@
 
-opts_knit$set(root.dir = "/Users/kelleherj/Documents/Research/QAEco/ODM_Review/")
+opts_knit$set(root.dir = "/Users/kelleherj/Documents/Research/QAEco/ODM_Review/dynamic-occupancy-model-review/")
 ```
@@
   arrange(Query, YearPeriod, randRank) %>%
-  mutate(groupRank = row_number())
+  mutate(GroupRank = row_number()) 
+```
 
+```{r Highlight papers to read}
+# Identify which papers qualify to be read
+highlightedPapers <- rankedPapers %>%
+  group_by(Query, YearPeriod) %>%
+  mutate(PercentRank =  GroupRank/max(GroupRank)) %>%
+  mutate(ToRead = case_when(
+    # If there are fewer than 5, read them all
+    max(GroupRank) <= 5 ~ "YES",
+    # If there are 20 or fewer, read the top 5
+    max(GroupRank) <= 20 & GroupRank <= 5 ~ "YES",
+    max(GroupRank) <= 20 & GroupRank >= 5 ~ "NO",
+    # If there are 21 or more, read the top 25%
+    max(GroupRank) >= 21 & PercentRank <= 0.25  ~ "YES",
+    max(GroupRank) >= 21 & PercentRank > 0.25 ~ "NO")) %>%
+  ungroup() %>%
+  # Get Review_ID, GroupRank, and ToRead
+  select(Review_ID, GroupRank, ToRead)
+
+# Write to disk
+write_csv(highlightedPapers, file = "RankedPapers.csv")
 ```

### 2022-07-26 Code/Randomisation/Randomise_Papers.Rmd
Update review july 2022

---
title: Randomly select papers for review
output:
  html_document: default
  pdf_document: default
---

***

## Summary
This document contains code for randomising paper ranks for the review on Dynamic Occupancy Models. For each of four search parameters, papers are stratified into 5 year time periods and randomly ranked. The top 25% or top 5 papers in each ranking category, whichever is larger, will be evaluated for inclusion; papers which do not qualify for review will be replaced with the next highest rank.

***

```{r setup, echo = FALSE, message = FALSE}
library("knitr")
library("tidyverse") 
library("readxl")

# Need seed set for reproducibility
set.seed(5378)

opts_knit$set(root.dir = "/Users/kelleherj/Documents/Research/QAEco/ODM_Review/dynamic-occupancy-model-review/")
```

Read in the excel sheets and process out the header
```{r Read-in excel sheets, echo = FALSE, message = FALSE, warning = FALSE}
# From the excel review spreadsheet, read in only the identifying characteristics 
identifierSheet <- readxl::read_xlsx("Review Spreadsheet.xlsx", 
                                     sheet = "Metadata", skip = 1) %>%
  # Drop rows prior to start of data
  slice(4:nrow(.)) %>%
  # Only need google search query and the ID reference
  select(Review_ID, Query = `Search terms`) 
# Repeat with the study metadata
studyDataSheet <- readxl::read_xlsx("Review Spreadsheet.xlsx", 
                                    sheet = "Study data", skip = 1) %>%
  # Cut out non-data rows
  slice(4:nrow(.)) %>%
  # Only need year data and ID reference
  select(Review_ID, Year) 
```

Combine the two sheets, the former containing metadata and the latter containing information on the papers
```{r Combine and merge, echo = FALSE, message = FALSE}
# Merge paper queries and metadata
paperList <- full_join(identifierSheet, studyDataSheet, 
                       by = "Review_ID")

# Exclude some non-query papers and add year periods to the rest
includedPapers <- paperList %>%
  # Drop empty rows, excluded, and incidental papers
  filter(!is.na(Query) & Query != "INCIDENTAL" & Year != "EXCLUDE") %>%
  # Break into time periods 
  mutate(YearPeriod = case_when(Year %in% as.character(2000:2005) ~ "2000:2005",
                                Year %in% as.character(2005:2010) ~ "2005:2010",
                                Year %in% as.character(2010:2015) ~ "2010:2015",
                                Year %in% as.character(2015:2021) ~ "2015:2021"))
```

Assign each paper a random overll rank, then order the papers within their strata by the overall rank
```{r Randomly rank papers within strata}
# Randomly rank papers, then rank within the strata
rankedPapers <- includedPapers %>%
  # Add random ranks for all papers
  mutate(randRank = sample(x = 1:nrow(.), size = nrow(.))) %>%
  # Re-rank within strata
  group_by(Query, YearPeriod) %>%
  arrange(Query, YearPeriod, randRank) %>%
  mutate(GroupRank = row_number()) 
```

Highlight which papers are ranked highly enough to qualify for evaluation 
```{r Highlight papers to read}
# Identify which papers qualify to be read
highlightedPapers <- rankedPapers %>%
  group_by(Query, YearPeriod) %>%
  mutate(PercentRank =  GroupRank/max(GroupRank)) %>%
  mutate(ToRead = case_when(
    # If there are fewer than 5, read them all
    max(GroupRank) <= 5 ~ "YES",
    # If there are 20 or fewer, read the top 5
    max(GroupRank) <= 20 & GroupRank <= 5 ~ "YES",
    max(GroupRank) <= 20 & GroupRank >= 5 ~ "NO",
    # If there are 21 or more, read the top 25%
    max(GroupRank) >= 21 & PercentRank <= 0.25  ~ "YES",
    max(GroupRank) >= 21 & PercentRank > 0.25 ~ "NO")) %>%
  ungroup() %>%
  unite(col = Group, c(Query, YearPeriod)) %>%
  # Get Review_ID, GroupRank, and ToRead
  select(Review_ID, Group, OverallRank = randRank, GroupRank, ToRead)
```

Add previously unranked papers, remove extraneous columns, and save both this workspace and the rankings to disk.
```{r Reformat and save}
# Get the unranked papers to add to the list
unrankedPapers <- paperList %>% 
  # Drop empty rows, excluded, and incidental papers
  filter(is.na(Query) | Query == "INCIDENTAL" | Year == "EXCLUDE") %>%
  mutate(ToRead = "NO", Group = "Unranked", GroupRank = "NA", OverallRank = "NA") %>%
  select(Review_ID, Group, OverallRank, GroupRank, ToRead)

# Combine ranked and unranked papers
outputList <- rbind(highlightedPapers, unrankedPapers) %>%
  mutate(Review_ID = as.numeric(Review_ID)) %>%
  arrange(Review_ID)

# Write to disk
write_csv(outputList, file = "RankedPapers.csv")
# Save environment as well for reproducibility
save.image("Randomise_Papers.RData")
```

### 2022-07-29 Analysis/Outputs/confirmationStats.Rmd
restructure files

---
title: key stats for confirmation
output:
  html_document: default
  pdf_document: default
---

***

```{r setup, echo = FALSE, message = FALSE}
library("knitr")
library("tidyverse") 
library("readxl")
library("sf")
```

```{r readin}
reviewTable <- read_xlsx("Data/ConfirmationTable.xlsx")

nonSPOM <- reviewTable %>%
  filter(SearchTerm %in% c("Dynamic occupancy model", "Multi-season occupancy model"))

```

```{r locations}
reviewTable %>%
  group_by(Biorealm) %>%
  summarize(TotalPapers = n()) %>%
  arrange(-TotalPapers)

reviewTable %>%
  group_by(Taxa) %>%
  summarize(TotalPapers = n()) %>%
  arrange(-TotalPapers)

reviewTable %>%
  group_by(IUCN) %>%
  summarize(TotalPapers = n()) %>%
  arrange(-TotalPapers)

reviewTable %>%
  group_by(AuthorCountry) %>%
  summarize(TotalPapers = n()) %>%
  arrange(-TotalPapers)

reviewTable %>%
  separate_rows(ObjectiveCategories, sep = ", ") %>%
  group_by(ObjectiveCategories) %>%
  summarise(count = n()) %>% 
  mutate(proportion = count/58) %>%
  arrange(-proportion) %>%
  filter(ObjectiveCategories != 'invasives') %>%

  ggplot() +
  geom_col(aes(x = ObjectiveCategories, y = proportion), fill = 'dodgerblue4') +
  ylim(c(0,1)) +
  scale_x_discrete(limits = c('conservation', 'management', 'methods', 'research'),
                    labels = c('Conservation', 'Management', 'Methods', 'Research')) +
  labs(x = "", y = "Proportion of papers") +
  theme_minimal() +
  theme(axis.text = element_text(size = 15),axis.title = element_text(size = 15))

```

```{r split covariate column}
nonSPOM %>%
  separate_rows(CovTypes, sep = ", ") %>%
  group_by(CovTypes) %>%
  summarise(count = n()) %>% 
  mutate(proportion = count/38) %>%
  filter(proportion > 0.125) %>%
  arrange(proportion) %>%

  ggplot(aes(x = reorder(CovTypes, -proportion),
             y = proportion)) +
  geom_col(fill = "dodgerblue4") +
  ylim(c(0,1)) + 
  labs(title = "Covariates usage in non-SPOM models",
       x = "Covariate class",
       y = "Proportion of papers") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 15),
        axis.title = element_text(size = 15)) +
  scale_x_discrete(limits = c('land_cover', 'habitat', 'biotic_interaction',
                              'hydro', 'land_use', 'anthropogenic'),
                   labels = c('Land Cover', 'Habitat', 'Biotic Interactions',
                              'Hydrologic', 'Land Use', 'Anthropogenic'))
```

```{r prop infer}
reviewTable %>% 
  count(InfSpatial) %>%
  mutate(Prop = n/58)

reviewTable %>% 
  count(InfFuture)%>%
  mutate(Prop = n/58)

reviewTable %>% 
  count(InfPast)%>%
  mutate(Prop = n/58)

reviewTable %>% 
  count(Maps)%>%
  mutate(Prop = n/58)

reviewTable %>%
  count(Taxa) %>%
  arrange(-n)%>%
  mutate(Prop = n/58)

reviewTable %>%
  count(ModelSel) %>%
  mutate(Prop = n/58)

reviewTable %>%
  count(ModelEval) %>%
  mutate(Prop = n/58)

t<-reviewTable %>%
  separate_rows(CovTypes, sep = ", ") %>%
  filter(!CovTypes %in% c("connectivity", "habitat", "survey", "none"))

reviewTable %>%
  separate_rows(CovTypes, sep = ", ") %>%
  group_by(CovTypes) %>%
  summarise(count = n()) %>% 
  mutate(proportion = count/58) %>%
  arrange(-proportion)  %>%
  ggplot(aes(x = reorder(CovTypes, -proportion),
             y = proportion)) +
  geom_col(fill = "dodgerblue4") +
  ylim(c(0,0.5)) + 
  labs(title = "Covariates usage in non-SPOM models",
       x = "Covariate class",
       y = "Proportion of papers") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 15),
        axis.title = element_text(size = 15)) +
  scale_x_discrete(limits = c('land_cover', 'biotic_interaction',
                              'hydro', 'land_use', 'anthropogenic'),
                   labels = c('Land Cover', 'Biotic Interactions',
                              'Hydrologic', 'Land Use', 'Anthropogenic'))

```

```{r maps}
ecoregions <- st_read("Data/Ecoregions2017/Ecoregions2017.shp") %>%
  st_make_valid() %>%
  group_by(REALM) %>%
  summarise() %>%
  st_simplify(dTolerance = 10000)

ecoregions <- ecoregions %>%
  st_simplify(dTolerance = 20000)

reviewTable %>%
  group_by(Biorealm) %>%
  summarize(TotalPapers = n()) %>%
  arrange(-TotalPapers)

ggplot(data = ecoregions) +
  geom_sf(aes(fill = REALM))
```

```{r}
data.frame(selMethod = c("A priori", "Stepwise", "Bayesian", "Predefined alternatives"),
           Proportion = c(.44, .24, .22, .11)) %>%
  ggplot(aes(x = reorder(selMethod, -Proportion),
             y = Proportion)) +
  geom_col(fill = "dodgerblue4") +
  ylim(c(0,1)) + 
  labs(x = "Model Selection Method",
       y = "Proportion of papers") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 15),
        axis.title = element_text(size = 15))

data.frame(infer = c("Limited to study\nduration and area  ", "Spatial", "Future"),
           Proportion = c(.62, .12, .29)) %>%
  ggplot(aes(x = reorder(infer, -Proportion),
             y = Proportion)) +
  geom_col(fill = "dodgerblue4") +
  ylim(c(0,1)) + 
  labs(x = "Type of inference",
       y = "Proportion of papers") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 15),
        axis.title = element_text(size = 15))
```

### 2022-08-30 Writing/Randomisation/Randomise_Papers.Rmd
need to get some first-round results first to determine new methods

--- 2022-07-26 Code/Randomisation/Randomise_Papers.Rmd
@@
 set.seed(5378)
-
-opts_knit$set(root.dir = "/Users/kelleherj/Documents/Research/QAEco/ODM_Review/dynamic-occupancy-model-review/")
 ```
@@
 # From the excel review spreadsheet, read in only the identifying characteristics 
-identifierSheet <- readxl::read_xlsx("Review Spreadsheet.xlsx", 
+identifierSheet <- readxl::read_xlsx("Data/Review Spreadsheet.xlsx", 
                                      sheet = "Metadata", skip = 1) %>%
@@
 # Repeat with the study metadata
-studyDataSheet <- readxl::read_xlsx("Review Spreadsheet.xlsx", 
+studyDataSheet <- readxl::read_xlsx("Data/Review Spreadsheet.xlsx", 
                                     sheet = "Study data", skip = 1) %>%

### 2023-01-16 Analysis/Code/ScannedPapersSummary.Rmd

---
title: Basic summary of included papers
output:
  html_document: default
  pdf_document: default
---

***

```{r setup, echo = FALSE, message = FALSE}
library("knitr")
library("tidyverse") 
library("readxl")

opts_knit$set(root.dir = "/Users/kelleherj/Documents/Research/QAEco/ODM_Review/dynamic-occupancy-model-review/")
```

```{r Read spreadsheet, echo = FALSE, message = FALSE}
regex <- c("Academic|Government|NGO|Private")

summarySheet <- readxl::read_xlsx("Data/SelectedSummaries.xlsx") %>%
  mutate(Include = case_when(Include == "Y" ~ "Included",
                             Include == "N" ~ "Not Included")) %>%
  mutate(Query = case_when(Query == "Dynamic occupancy model" ~ "DOM",
                           Query == "Occupancy dynamics model" ~ "ODM",
                           Query == "Multi-season occupancy model" ~ "MSOM",
                           Query == "Stochastic patch occupancy model" ~ "SPOM")) %>%
  mutate(Query = factor(Query, levels = c("DOM", "ODM", "MSOM", "SPOM"))) %>%
  mutate(Academic = grepl("Academic", Practitioner)) %>%
  mutate(Government = grepl("Government", Practitioner)) %>%
  mutate(Private = grepl("Private", Practitioner)) %>%
  mutate(NGO = grepl("NGO", Practitioner)) %>%
  mutate(Covariates = case_when(Covariates == "Y" ~ 1,
                                Covariates == "N" ~ 0))
```

## Qualified papers 
```{R overall qualified, echo = FALSE, message = FALSE}
summarySheet %>%
  mutate(Include = as.factor(Include)) %>%
  count(Query, Include, name = "Papers", .drop = FALSE) %>%

  ggplot(aes(x = Query, y = Papers, fill = Include)) +
  geom_col(position = "dodge") + 
  scale_fill_manual(values = c("dodgerblue3", "firebrick3")) +
  ggtitle("Inclusion by query") +
  theme(legend.title = element_blank())
```

## Authorship Organisation
```{r Organisation, echo = FALSE, message = FALSE}
summarySheet %>%
  filter(Include == "Included") %>%
  summarize(Total = n(),
            Academic = sum(Academic)/Total,
            Government = sum(Government)/Total, 
            NGO = sum(NGO)/Total,
            Private = sum(Private)/Total) %>%
  pivot_longer(cols = c(Academic, Government, NGO, Private), names_to = "Practitioner", values_to = "Proportion") %>%

  ggplot(aes(x = Practitioner, y = Proportion, fill = Practitioner)) +
  geom_col(position = "dodge") +
  ggtitle("Proportion of papers by authorship category") +
  theme(legend.title = element_blank()) +
  ylim(0,1)
```

## Covariates
```{r Covariate use, echo = FALSE, message = FALSE}
summarySheet %>%
  filter(Include == "Included") %>%
  summarize(Prop = sum(Covariates)/n()) %>%

  ggplot(aes(x = , y = Prop)) +
  geom_col(fill = "dodgerblue4") + 
  ggtitle("Proportion using covariates") +
  ylim(0,1)
```

```{r spreadsheet for goals}
regex <- c("Management|Methods|Invasives|Conservation|Research")

summarySheet2 <- readxl::read_xlsx("Data/SelectedSummaries.xlsx") %>%
  mutate(Include = case_when(Include == "Y" ~ "Included",
                             Include == "N" ~ "Not Included")) %>%
  mutate(Query = case_when(Query == "Dynamic occupancy model" ~ "DOM",
                           Query == "Occupancy dynamics model" ~ "ODM",
                           Query == "Multi-season occupancy model" ~ "MSOM",
                           Query == "Stochastic patch occupancy model" ~ "SPOM")) %>%
  mutate(Query = factor(Query, levels = c("DOM", "ODM", "MSOM", "SPOM"))) %>%
  mutate(Methods = grepl("Methods", Objective)) %>%
  mutate(Management = grepl("Management", Objective)) %>%
  mutate(Conservation = grepl("Conservation", Objective)) %>%
  mutate(Research = grepl("Research", Objective)) %>%
  mutate(Invasives = grepl("Invasives", Objective)) %>%
  mutate(Covariates = case_when(Covariates == "Y" ~ 1,
                                Covariates == "N" ~ 0))

summarySheet2 %>%
  filter(Include == "Included") %>%
  summarize(Total = n(),
            Methods = sum(Methods)/Total,
            Research = sum(Research)/Total, 
            Management = sum(Management)/Total,
            Conservation = sum(Conservation)/Total,
            Invasives = sum(Invasives)/Total) %>%
  pivot_longer(cols = c(Methods, Research, Management, Conservation, Invasives), 
               names_to = "Objectives", values_to = "Proportion") %>%

  ggplot(aes(x = Objectives, y = Proportion, fill = Objectives)) +
  geom_col(position = "dodge") +
  ggtitle("Proportion of papers by objective category") +
  theme(legend.title = element_blank()) +
  ylim(0,1)
```

### 2023-01-16 Analysis/Code/Formal_Randomisation.R
Resolve "Reprocess stage1"

@@ -1,0 +1,17 @@
+## Rerandomisation of papers from the preliminary review to be re-read in the 
+## formal review
+
+library(tidyverse)
+library(readxl)
+
+set.seed(1234)
+
+# Read review IDs
+review_sheet <- read_xlsx("Spreadsheets/Stage2_Spreadsheet.xlsx")
+
+# Shuffle review ID S
+random_order <- data.frame(order = seq(1:nrow(review_sheet)),
+                           reviewID = sample(review_sheet$`Review ID`, nrow(review_sheet)))
+
+       
+write_csv(random_order, "Analysis/Outputs/formal_randomisation.csv")

### 2023-01-23 Analysis/Code/1_DataPrep.Rmd
Started data prep rmd

---
title: "DOM review: Data preparation"
output: html_document
---

## Summary

This document reads in the review spreadsheet and converts the fields to a usable format. It also checks each field for any possible input errors and indicates the offending paper to the user.

```{r Packages, include = FALSE}
library(tidyverse)
library(readxl)
library(knitr)
```

## Read-in
The review data spreadsheet is comprised of six sheets, unified by the shared 'Review ID' field.
- Study: Authorship and publication information
- Objectives and Inference: Purpose of the study and key outputs
- Species and Location: Taxa and geographic information
- Data collection: Sampling and survey methods
- Covariates: Estimated parameters and predictors used
- Modelling: Model selection and evaluation methods

Each must be read-in individually.
```{r Read spreadsheet, echo = FALSE}
# Load the spreadsheet as a list of dataframes (sheets)
spreadsheet_path <- "Spreadsheets/Stage2_Spreadsheet.xlsx"
spreadsheet <- map(excel_sheets(spreadsheet_path),
                   read_excel,
                   path = spreadsheet_path)
# Reapply sheet names
names(spreadsheet) <- excel_sheets(spreadsheet_path)
```

## Study
This sheet contains information on authorship and publication history.
```{r Process study, echo = FALSE}
study_sheet <- spreadsheet[["Study"]] %>%
  # Select only included papers
  filter(Status == "Complete") %>%
  # Coerce all journal names to uppercase
  mutate(Journal = toupper(Journal)) %>%
  # Convert each possible 'Authorship group` to binary TRUE/FALSE
  mutate(Author_NGO = str_detect(`Authorship groups`, "NGO"),
         Author_Government = str_detect(`Authorship groups`, "Government"),
         Author_Private = str_detect(`Authorship groups`, "Private"),
         Author_Academic = str_detect(`Authorship groups`, "Academic")) %>%
  # Select only necessary columns for analysis
  select(`Review ID`, Title, Journal, Year, `First author country`, 
         Author_NGO, Author_Government, Author_Private, Author_Academic, `Authorship groups`)

# Print head
kable(head(study_sheet))
```

Next, each column of the study sheet must be checked for possible errors
```{r Check study}
# Are there any journals whose titles should match but don't?
study_sheet %>% count(Journal) %>% arrange(-n)

# Are all country names correct?
study_sheet %>% count(`First author country`) %>% arrange(-n)

# Are all years between 2000-2022?
study_sheet %>% count(Year) %>% arrange(Year)

# Do all papers have at least one of the Authorship categories set to 'True'?
study_sheet %>% select(Author_NGO,Author_Academic,Author_Government,Author_Private)
```

## Objectives 
This sheet contains information on the authors' stated research objectives and whether their papers produced certain products (maps/predictions). This sheet is relatively simple and does not require checks.
```{r Process objectives, echo = FALSE}
objectives_sheet <- spreadsheet[["Objectives and Inference"]] %>%
  # Select only completed papers
  filter(Status == "Complete") %>%
  # Convert objective columns to TRUE/FALSE
  mutate(Obj_Methods = case_when(`Methods?` == "Y" ~ TRUE,
                                 `Methods?` == "N" ~ FALSE),
         Obj_Projections = case_when(`Future predictions?` == "Y" ~ TRUE,
                                     `Future predictions?` == "N" ~ FALSE),
         Obj_Spatial = case_when(`Spatial predictions?` == "Y" ~ TRUE,
                                 `Spatial predictions?` == "N" ~ FALSE),
         Obj_Trends = case_when(`Observing trends?` == "Y" ~ TRUE,
                                `Observing trends?` == "N" ~ FALSE),
         Obj_Hypothesis = case_when(`Hypothesis testing?` == "Y" ~ TRUE,
                                    `Hypothesis testing?` == "N" ~ FALSE)) %>%
  # Drop unneeded columns
  select(`Review ID`, Title, Obj_Methods, Obj_Projections, 
         Obj_Spatial, Obj_Trends, Obj_Hypothesis)

# Print first few rows
kable(head(objectives_sheet))
```

## Species and Location
This sheet describes the study species and area. Some papers with multi-species models may have multiple entries.
```{r Process Species/Location, echo = FALSE}
taxa_location_sheet <- spreadsheet[["Species and Location"]] %>%
  # Filter to included papers
  filter(Status == "Complete") %>%
  # Create binary threat columns
  mutate(Threat_HabitatLoss = str_detect(`Threat categories`, "HabitatLoss"),
         Threat_Exploitation = str_detect(`Threat categories`, "Exploitation"),
         Threat_ClimateChange = str_detect(`Threat categories`, "ClimateChange"),
         Threat_Invasives = str_detect(`Threat categories`, "Invasives"),
         Threat_Disease = str_detect(`Threat categories`, "Disease"),
         Threat_Disaster = str_detect(`Threat categories`, "Disturbance")) %>%
  # Select only necessary files
  select(`Review ID`, Title, Taxa, `Taxa keywords`, Number, 
         Threat_HabitatLoss, Threat_Exploitation, Threat_ClimateChange,
         Threat_Invasives, Threat_Disease, Threat_Disaster,
         Country, Continent, Realm, `Study Size`)
```

This section includes several fields which require checks
```{r Check Species/Location, echo = FALSE}

```

### 2023-01-24 Analysis/Code/1_DataPrep.Rmd
working on survey page of spreadsheet

@@
          Threat_Disaster = str_detect(`Threat categories`, "Disturbance")) %>%
-  # Select only necessary files
+  # Split 'taxa keywords' column by commas
+  separate_rows(`Taxa keywords`, sep = ", ") %>%
+  # Split country column by commas
+  separate_rows(Country, sep = ", ") %>%
+  # Select only necessary fields
   select(`Review ID`, Title, Taxa, `Taxa keywords`, Number, 
@@
 
-This section includes several fields which require checks
+This section includes a few fields which require checks
 ```{r Check Species/Location, echo = FALSE}
+# Are all taxa names appropriate?
+unique(taxa_location_sheet$Taxa)
+
+# Should any taxa categories be condensed or removed?
+unique(taxa_location_sheet$`Taxa keywords`)
+
+# Are all country names correct?
+taxa_location_sheet %>% count(Country) %>% arrange(-n)
+```
+
+## Data collection
+This sheet includes information on survey and sampling techniques
+```{r Process Data Collection, echo = FALSE}
+survey_sheet <- spreadsheet[["Data collection"]] %>%
+  # Filter to included papers
+  filter(Status == "Complete") %>% 
+  # Split capture method and sampling method fields
+  separate_rows(`Capture method`, sep = ", ") %>%
+  separate_rows(`Sampling method`, sep = ", ") %>%
+  #
+
+  mutate(MinPrimary = `Number of primary occasions`,
+         MaxPrimary = `Number of primary occasions`) %>%
+  map_dfr(.f = function(.x){
+    if (str_detect(.x$MinPrimary, "-")){
+      .x$MinPrimary <- str_split(.x$MinPrimary[[1]])
+      .x$MaxPrimary <- str_split(.x$MaxPrimary[[2]])
+    }
+  })
+
+  mutate(MinPrimary = case_when(str_detect(`Number of primary occasions`, "-") ~
+                                  str_split(`Number of primary occasions`, "-")[[1]],
+                                !str_detect(`Number of primary occasions`, "-") ~
+                                  `Number of primary occasions`),
+         MaxPrimary = case_when(str_detect(`Number of primary occasions`, "-") ~
+                                  str_split(`Number of primary occasions`, "-")[[2]],
+                                !str_detect(`Number of primary occasions`, "-") ~
+                                  `Number of primary occasions`))
+
+        function(test){}
+
+  # Select important fields
+  select(`Review ID`, Title, `Study duration`, `Capture method`, 
+         `Num primary` = `Number of primary occasions`,
+         `Num secondary` = `Number of secondary occasions`,
+         `Type secondary` = `Type of secondary occasions`,
+         `Type site` = `Type of site`, `Num sites` = `Number of sites`,
+         `Sampling method`)
 

### 2023-01-26 Analysis/Code/1_DataPrep.Rmd
finished processing review data

--- 2023-01-23 Analysis/Code/1_DataPrep.Rmd
@@
          Country, Continent, Realm, `Study Size`)
+kable(head(taxa_location_sheet))
 ```
 
-This section includes several fields which require checks
+This section includes a few fields which require checks
 ```{r Check Species/Location, echo = FALSE}
+# Are all taxa names appropriate?
+unique(taxa_location_sheet$Taxa)
+
+# Should any taxa categories be condensed or removed?
+unique(taxa_location_sheet$`Taxa keywords`)
+
+# Are all country names correct?
+taxa_location_sheet %>% count(Country) %>% arrange(-n)
+```
+
+## Data collection
+This sheet includes information on survey and sampling techniques
+```{r Process Data Collection, echo = FALSE}
+survey_sheet <- spreadsheet[["Data collection"]] %>%
+  # Filter to included papers
+  filter(Status == "Complete") %>% 
+  # Split capture method and sampling method fields
+  separate_rows(`Capture method`, sep = ", ") %>%
+  separate_rows(`Sampling method`, sep = ", ") %>%
+  # Split min/max number of primary occasions if present
+  mutate(MinPrimary = str_split(`Number of primary occasions`, "-", simplify = TRUE)[,1],
+         MaxPrimary = str_split(`Number of primary occasions`, "-", simplify = TRUE)[,2]) %>%
+  mutate(MaxPrimary = case_when(MaxPrimary == "" ~ MinPrimary,
+                                MaxPrimary != "" ~ MaxPrimary)) %>%
+  # Repeat for secondary occasions
+  mutate(MinSecondary = str_split(`Number of secondary occasions`, "-", simplify = TRUE)[,1],
+         MaxSecondary = str_split(`Number of secondary occasions`, "-", simplify = TRUE)[,2]) %>%
+  mutate(MaxSecondary = case_when(MaxSecondary == "" ~ MinSecondary,
+                                  MaxSecondary != "" ~ MaxSecondary)) %>%
+  # Coerce columns to numeric
+  mutate(MinPrimary = as.numeric(MinPrimary),
+         MaxPrimary = as.numeric(MaxPrimary),
+         MinSecondary = as.numeric(MinSecondary),
+         MaxSecondary = as.numeric(MaxSecondary),
+         NumSites = as.numeric(`Number of sites`)) %>%
+  # Select important fields
+  select(`Review ID`, Title, `Study duration`, `Capture method`, 
+         MinPrimary, MaxPrimary, MinSecondary, MaxSecondary,
+         `Type site` = `Type of site`, NumSites,
+         `Sampling method`)
+
+kable(head(survey_sheet))
+```
+
+This section still needs to be checked for any input errors.
+```{r Check Data Collection, echo = FALSE}
 
 ```
+
+## Covariates
+This section includes estimated parameters and covariates considered for inclusion
+
+```{r Process covariates, echo = FALSE}
+covariates_sheet <- spreadsheet[["Covariates"]] %>%
+  # Filter down to only included papers
+  filter(Status == "Complete") %>%
+  # Note modifiers for parameters, then split them off from parameters
+  mutate(MultiSpecies = case_when(str_detect(Parameter,"MS") ~ TRUE,
+                                  !str_detect(Parameter, "MS") ~ FALSE),
+         FalsePositive = case_when(str_detect(Parameter, "FP") & str_detect(Parameter, "Detection") ~ TRUE)) %>%
+  mutate(Parameter = str_split(Parameter, "_", simplify = TRUE)[,1]) %>%
+  # Split out each covariate
+  separate_rows(`Categories`, sep =  ", ") %>%
+  # Convert 'none' to NA
+  separate(`Categories`, c("Category", "Scale", "Dynanimity"), sep = "_") %>%
+  select(`Review ID`, Title, `Total covariates`, Parameter, Category, Scale, 
+         Dynanimity, `Higher order`, `Interactive terms`,
+         MultiSpecies, FalsePositive)
+
+covariates_sheet$Category[covariates_sheet$Category %in% c("None", "NA")] <- NA
+covariates_sheet$Category[covariates_sheet$Category == c("Topography")] <- "Topographic"
+covariates_sheet$Category[covariates_sheet$Category == c("Hydrology")] <- "Hydrologic"
+
+kable(head(covariates_sheet))
+```
+
+```{r Check covariates, echo = FALSE}
+```
+
+## Modelling
+This contains information about the model selection and evaluation process
+```{r Process modelling, echo = FALSE}
+modelling_sheet <- spreadsheet[["Modelling"]] %>%
+  filter(Status == "Complete") %>%
+  separate_rows(`Selection category`, sep = ", ") %>%
+  separate(`Selection category`, into = c("Select_Category", "Select_Type"), sep = "_") %>%
+  select(`Review ID`, Title, Bayesian, Select_Category, Select_Type)
+
+modelling_sheet$Select_Category[modelling_sheet$Select_Category == "APriori"] <- "Apriori"
+
+```
+
+```{r Save output, echo = FALSE}
+
+reviewResults <- list(Study = study_sheet,
+                      Objectives = objectives_sheet,
+                      Taxa = taxa_location_sheet,
+                      Survey = survey_sheet,
+                      Covariates = covariates_sheet,
+                      Modelling = modelling_sheet)
+
+write_rds(reviewResults, "Analysis/Outputs/ReviewResults.rds")
+```

### 2023-01-26 Analysis/Code/2_InitialAnalysis.Rmd
first few basic figs

---
title: "DOM Review: Initial analysis"
output: html_document
---

```{r Setup, include = FALSE}
library(tidyverse)
```

```{r Readin, echo = FALSE}
reviewResults <- read_rds("Analysis/Outputs/ReviewResults.rds")
```

```{r Study, echo = FALSE}

```

```{r Objectives, echo = FALSE}

reviewResults$Objectives %>%
  pivot_longer(c("Obj_Methods","Obj_Projections","Obj_Spatial","Obj_Trends","Obj_Hypothesis"),
               "Objective") %>%
  mutate(value = as.numeric(value)) %>%
  group_by(Objective) %>%
  summarise(n=sum(value)) %>%

  ggplot(aes(x = Objective, y = n/54)) +
  geom_col() +
  ylim(0,1)
```

```{r Taxa, echo = FALSE}

```

```{r Survey, echo = FALSE}
hist(as.numeric(reviewResults$Survey$`Study duration`))

reviewResults$Survey %>%
  group_by(`Capture method`) %>%
  summarize(n = n()) %>%

  ggplot(aes(x = reorder(`Capture method`, -n), y = n)) +
  geom_col() +
  theme(axis.text.x = element_text(angle=90, vjust = 0.5,hjust=1))

```

```{r Covariates, echo = FALSE}
nrow(reviewResults$Covariates)

reviewResults$Covariates %>%
  filter(!is.na(Category)) %>%
  group_by(Category, Scale) %>%
  summarize(Count = n()) %>%
  arrange(Count) %>%

  ggplot(aes(x = reorder(Category, -Count), y = Count, fill = Scale)) +
  geom_col() + 
  scale_fill_discrete(limits = c("a","p","d","g","u", NA), 
                    labels = c("area","point","distance","global", "Unclear", "NA")) +
  theme(axis.text.x = element_text(angle=90, vjust = 0.5,hjust=1))

reviewResults$Covariates %>%
  filter(!is.na(Category)) %>%
  group_by(Category, Dynanimity) %>%
  summarize(Count = n()) %>%
  arrange(Count) %>%

  ggplot(aes(x = reorder(Category, -Count), y = Count, fill = Dynanimity)) +
  geom_col() + 
  scale_fill_discrete(limits = c("d","s","u"), 
                    labels = c("Dynamic", "Static", "Unclear")) +
  theme(axis.text.x = element_text(angle=90, vjust = 0.5,hjust=1))

```

```{r Modelling, echo = FALSE}

```

### 2023-01-30 Analysis/Code/2_InitialAnalysis.Rmd
added study summaries

@@
 library(tidyverse)
+library(sf)
+library(knitr)
 ```
 
-```{r Readin, echo = FALSE}
+## Summary
+This document includes some preliminary analyses of the review results, arranged by the six categories included in the review:
+
+* Study
+* Objectives
+* Taxa and location
+* Survey methods
+* Parameters and covariates
+* Modelling methods
+
+```{r Read-in, echo = FALSE}
 reviewResults <- read_rds("Analysis/Outputs/ReviewResults.rds")
@@
 
+## Study 
+Basic information on the locations and affiliations of authors involved with each paper was collected.
+Two thirds of papers (36/54) were by authors from U.S.
+institutions; the next greatest total was 3 papers from Australia.
+
 ```{r Study, echo = FALSE}
-
+studyData <- reviewResults[["Study"]]
 ```
 
+```{r Study authorship countries, echo = FALSE, message = FALSE}
+countriesShapefile <- st_read("Analysis/SpatialData/Countries/World_Countries__Generalized_.shp")
+
+## Map of authorship countries by # of papers
+studyData %>%
+  # Get count of 
+  group_by(`First author country`) %>%
+  summarise(Count = n()) %>%
+  arrange(-Count) %>% 
+  mutate(`First author country` = case_when(`First author country` == "United States of America" ~ "United States",
+                                            `First author country` != "United States of America" ~ `First author country`)) %>%
+  # Merge by country name
+  right_join(countriesShapefile, by = c('First author country' = 'COUNTRY')) %>%
+  # Plot map
+  ggplot() +
+  geom_sf(aes(geometry = geometry, fill = log(Count))) +
+  labs(title = "Log paper count by country") +
+  theme_void() + 
+  scale_fill_continuous(name = "Log counts", low="thistle2", high="darkred",na.value="white")
+
+# Table with the same data
+studyData %>%
+  group_by(`First author country`) %>%
+  summarise(`Total papers` = n()) %>%
+  arrange(-`Total papers`) %>%
+  kable()
+```
+
+Affiliated institutions were split into four categories: Academic, Non-governmental organisations, Governmental agencies, and private sector groups.
+These were fairly varied, with academic groups participating in the vast majority of papers but with substantial representation from governmental and NGO groups.
+
+```{r Affiliation types, echo = FALSE}
+studyData %>%
+  summarise(NGO = sum(Author_NGO)/n()*100,
+            Academic = sum(Author_Academic)/n()*100,
+            Government = sum(Author_Government/n()*100),
+            Private = sum(Author_Private)/n()*100) %>% 
+  pivot_longer(cols = c(NGO, Academic, Government, Private), names_to = 'Affiliation', values_to = "% of papers") %>%
+  arrange(-`% of papers`) %>%
+  kable(digits = 1)
+```
+
+## Objectives
+Five categories of objectives and goals were assessed for each paper:
+* Assessing trends - interested in the trajector of estimated or derived parameters
+* Testing hypotheses - evaluating attributes of a **specific** covariate on parameters
+* Spatial mapping - Extrapolating estimates of papameter values beyond surveyed sites
+* Future projections - Making predictions of parameters into the future
+* Methods - introducing, testing, or demonstrating a new extension to these models
+
 ```{r Objectives, echo = FALSE}
-
-reviewResults$Objectives %>%
-  pivot_longer(c("Obj_Methods","Obj_Projections","Obj_Spatial","Obj_Trends","Obj_Hypothesis"),
-               "Objective") %>%
-  mutate(value = as.numeric(value)) %>%
-  group_by(Objective) %>%
-  summarise(n=sum(value)) %>%
-
-  ggplot(aes(x = Objective, y = n/54)) +
-  geom_col() +
-  ylim(0,1)
+objData <- reviewResults[["Objectives"]]
 ```
@@
 ```{r Survey, echo = FALSE}
-hist(as.numeric(reviewResults$Survey$`Study duration`))
-
-reviewResults$Survey %>%
-  group_by(`Capture method`) %>%
-  summarize(n = n()) %>%
-
-  ggplot(aes(x = reorder(`Capture method`, -n), y = n)) +
-  geom_col() +
-  theme(axis.text.x = element_text(angle=90, vjust = 0.5,hjust=1))
 
@@
 ```{r Covariates, echo = FALSE}
-nrow(reviewResults$Covariates)
-
-reviewResults$Covariates %>%
-  filter(!is.na(Category)) %>%
-  group_by(Category, Scale) %>%
-  summarize(Count = n()) %>%
-  arrange(Count) %>%
-
-  ggplot(aes(x = reorder(Category, -Count), y = Count, fill = Scale)) +
-  geom_col() + 
-  scale_fill_discrete(limits = c("a","p","d","g","u", NA), 
-                    labels = c("area","point","distance","global", "Unclear", "NA")) +
-  theme(axis.text.x = element_text(angle=90, vjust = 0.5,hjust=1))
-
-reviewResults$Covariates %>%
-  filter(!is.na(Category)) %>%
-  group_by(Category, Dynanimity) %>%
-  summarize(Count = n()) %>%
-  arrange(Count) %>%
-
-  ggplot(aes(x = reorder(Category, -Count), y = Count, fill = Dynanimity)) +
-  geom_col() + 
-  scale_fill_discrete(limits = c("d","s","u"), 
-                    labels = c("Dynamic", "Static", "Unclear")) +
-  theme(axis.text.x = element_text(angle=90, vjust = 0.5,hjust=1))
 

### 2023-02-03 Analysis/Code/2_InitialAnalysis.Rmd
added some obj plots

@@
 studyData %>%
-  # Get count of 
+  # Get count of papers by author's first country
   group_by(`First author country`) %>%
@@
 ```{r Affiliation types, echo = FALSE}
+# Calc % of each type
 studyData %>%
@@
             Private = sum(Author_Private)/n()*100) %>% 
+  # Make legible for table
   pivot_longer(cols = c(NGO, Academic, Government, Private), names_to = 'Affiliation', values_to = "% of papers") %>%
@@
 Five categories of objectives and goals were assessed for each paper:
+
 * Assessing trends - interested in the trajector of estimated or derived parameters
@@
 
-```{r Taxa, echo = FALSE}
+Monitoring trends and testing hypotheses were the most popular objectives; making predictions (spatial or temporal) were quite rare.
+```{r Objectives table, echo = FALSE}
+# Make table 
+objData %>% 
+  pivot_longer(cols = c(Obj_Methods, Obj_Projections, Obj_Hypothesis, Obj_Spatial, Obj_Trends),
+               names_to = "Objective", values_to = "TF") %>%
+  # Drop obj_ prefix
+  mutate(Objective = str_replace(Objective, "Obj_","")) %>%
+  # Tally up and make table
+  group_by(Objective) %>%
+  summarise(`Percent of papers` = 100*sum(TF)/54) %>%
+  arrange(-`Percent of papers`) %>%
+  kable(digits = 1)
+```
+
+```{r Obj by Authorship group, echo = FALSe}
+objByAuth <- full_join(objData, studyData, by = c("Review ID", "Title")) %>%
+  # Pivot authorship column and adjust  
+  pivot_longer(cols = c(Author_NGO, Author_Academic, Author_Government, Author_Private), 
+               names_to = 'Authorship', values_to = "Auth_TF") %>%
+  mutate(Authorship = str_replace(Authorship, "Author_","")) %>%
+  # Get count of # of papers by authorship for the plot later
+  filter(Auth_TF == TRUE) %>%
+  group_by(Authorship) %>%
+  mutate(nAuth = n()) %>% 
+  # Pivot objectives column and adjust
+  pivot_longer(cols = c(Obj_Methods, Obj_Projections, Obj_Hypothesis, Obj_Spatial, Obj_Trends),
+               names_to = "Objective", values_to = "Obj_TF") %>%
+  mutate(Objective = str_replace(Objective, "Obj_","")) %>%
+  filter(Obj_TF == TRUE)
+
+objByAuth %>%
+  filter(Obj_TF == TRUE & Auth_TF == TRUE) %>%
+  group_by(Authorship) %>%
+  mutate(TotalAuth = length(unique(Title))) %>%
+  group_by(Authorship, Objective, nAuth) %>%
+  summarise(Count = n()/nAuth) %>%
+  ungroup() %>%
+  complete(Authorship, Objective) %>%
+  mutate(Count = case_when(is.na(Count) ~ 0.01,
+                           !is.na(Count) ~ Count)) %>%
+
+  ggplot(aes(x = Authorship, y = Count, fill = Objective)) +
+  geom_col(position = "dodge") +
+  ylim(0, 1) + 
+  labs(title = "Percent of papers by affiliation pursuing each objective")
+```
+
+```{r Covs by obj, echo = FALSE}
+covData <- reviewResults[["Covariates"]] %>%
+  mutate(`Review ID` = as.numeric(str_replace(`Review ID`, "[:alpha:]", "")))
+
+objData %>%
+   pivot_longer(cols = c(Obj_Methods, Obj_Projections, Obj_Hypothesis, Obj_Spatial, Obj_Trends),
+               names_to = "Objective", values_to = "Obj_TF") %>%
+  mutate(Objective = str_replace(Objective, "Obj_","")) %>%
+  filter(Obj_TF == TRUE) %>%
+  select(`Review ID`, Objective) %>%
+  full_join(covData, by = 'Review ID') %>%
+
+  group_by(`Review ID`, Objective, Parameter) %>%
+  summarise(Count = n()) %>%
+  filter(Parameter %in% c("Colonisation", "Extinction", "InitialOccupancy", "Detection"),
+         Count < 20) %>%
+
+  ggplot(aes(x = Objective, y = Count, colour = Parameter)) +
+  geom_boxplot()
 
 ```
+
+## Taxa and location
+Details on the study organism was collected for each study to assess any trends in the types of species these models are applied to, with a particular focus on those that are vulnerable to conservation threats.
+The locations where surveys were conducted was also noted.
+
+```{r Taxa, echo = FALSE}
+taxaData <- reviewResults[["Taxa"]]
+```
+
+As with authorship, the United States dominated the countries where surveys took place.
+
+### Study site maps {.tabset}
+#### Country
+```{r Site country map, echo = FALSE}
+taxaData %>%
+  # Divide by unique title/country
+  group_by(Country, Title) %>%
+  summarise() %>%
+  # Get count of papers by site country
+  group_by(Country) %>%
+  summarise(Count = n()) %>%
+  arrange(-Count) %>% 
+  mutate(Country = case_when(Country == "United States of America" ~ "United States",
+                             Country != "United States of America" ~ Country)) %>%
+  # Merge by country name
+  right_join(countriesShapefile, by = c('Country' = 'COUNTRY')) %>%
+  # Plot map
+  ggplot() +
+  geom_sf(aes(geometry = geometry, fill = log(Count))) +
+  labs(title = "Log paper count by country") +
+  theme_void() + 
+  scale_fill_continuous(name = "Log counts", 
+                        low = "thistle2", 
+                        high = "darkred",
+                        na.value = "white")
+
+# Table with the same data
+taxaData %>%
+  # Divide by unique title/country
+  group_by(Country, Title) %>%
+  summarise() %>%
+  # Get count of papers by site country
+  group_by(Country) %>%
+  summarise(`Total papers` = n()) %>%
+  arrange(-`Total papers`) %>% 
+  kable()
+```
+
+#### Continent
+```{r Site continent map, echo = FALSE}
+continentShapefile <- st_read("Analysis/SpatialData/Continents/World_Continents.shp")
+
+taxaData %>%
+  # Divide by unique title/country
+  group_by(Country, Continent, Title) %>%
+  summarise() %>%
+  # Get count of papers by site continent
+  group_by(Continent) %>%
+  summarise(Count = n()) %>%
+  arrange(-Count) %>% 
+  mutate(Continent = case_when(Continent == "Oceania" ~ "Australia",
+                               Continent != "Oceania" ~ Continent)) %>%
+
+  # Merge by continent name
+  right_join(continentShapefile, by = c('Continent' = 'CONTINENT')) %>%
+  # Plot map
+  ggplot() +
+  geom_sf(aes(geometry = geometry, fill = Count)) +
+  labs(title = "Paper count by continent") +
+  theme_void() + 
+  scale_fill_continuous(name = "Log counts", 
+                        low = "thistle2", 
+                        high = "darkred",
+                        na.value = "white")
+
+# Table with the same data
+taxaData %>%
+  # Divide by unique title/country
+  group_by(Country, Title, Continent) %>%
+  summarise() %>%
+  # Get count of papers by continent
+  group_by(Continent) %>%
+  summarise(`Total papers` = n()) %>%
+  arrange(-`Total papers`) %>% 
+  kable()
+
+```
+
+#### Realm
+```{r Site realm map, echo = FALSE}
+#ecoregionsShapefile <- st_read("Analysis/SpatialData/Ecoregions/Terrestrial_Ecoregions.shp") %>%
+#  st_make_valid() 
+#
+#t <- ecoregionsShapefile %>%
+#  group_by(WWF_REALM2) %>%
+#  summarise(geometry = st_combine(geometry))
+#
+#taxaData %>%
+#  # Divide by unique title/country
+#  group_by(Country, Realm, Title) %>%
+#  summarise() %>%
+#  # Get count of papers by site continent
+#  group_by(Realm) %>%
+#  summarise(Count = n()) %>%
+#  arrange(-Count) %>% 
+#  
+#  # Merge by realm
+#  right_join(ecoregionsShapefile, by = c('Realm' = 'WWF_REALM2')) %>%
+#  # Plot map
+#  ggplot() +
+#  geom_sf(aes(geometry = geometry, fill = Count)) +
+#  labs(title = "Paper count by Udvardy realm") +
+#  theme_void() + 
+#  scale_fill_continuous(name = "Log counts", 
+#                        low = "thistle2",
+#                        high = "darkred",
+#                        na.value="white")
+#
+# Table with the same data
+taxaData %>%
+  # Divide by unique title/country
+  group_by(Country, Title, Realm) %>%
+  summarise() %>%
+  # Get count of papers by site realm
+  group_by(Realm) %>%
+  summarise(`Total papers` = n()) %>%
+  arrange(-`Total papers`) %>% 
+  kable()
+```
+
+## Survey methods
 
@@
 
+## Parameters and covariates
+
 ```{r Covariates, echo = FALSE}
@@
 
+## Modelling decisions
+
 ```{r Modelling, echo = FALSE}

### 2023-02-08 Analysis/Code/2_InitialAnalysis.Rmd
major additions to summary

--- 2023-01-26 Analysis/Code/2_InitialAnalysis.Rmd
@@
 title: "DOM Review: Initial analysis"
-output: html_document
+output: 
+  html_document:
+    theme: journal
+    toc: true
+    toc_float: true
+    toc_collapsed: true
 ---
@@
 library(tidyverse)
+library(sf)
+library(knitr)
+options(dplyr.summarise.inform = FALSE)
+```
+## Summary
+This document includes some preliminary analyses of the dynamic occupancy model review results, arranged by the six categories included in the review:
+
+* Study
+* Objectives
+* Taxa and location
+* Survey methods
+* Parameters and covariates
+* Modelling methods
+
+```{r Read-in, echo = FALSE, cache = TRUE}
+# Load in processed spreadsheet
+reviewResults <- read_rds("Analysis/Outputs/ReviewResults.rds")
+
+# Separate out each page of the spreadsheet
+studyData <- reviewResults[["Study"]]
+objData   <- reviewResults[["Objectives"]]
+taxaData  <- reviewResults[["Taxa"]] %>%
+  mutate(`Review ID` = as.numeric(str_replace(`Review ID`, "[:alpha:]", "")))
+surveyData <- reviewResults[["Survey"]] %>% 
+  mutate(`Review ID` = as.numeric(str_replace(`Review ID`, "[:alpha:]", "")))
+covData   <- reviewResults[["Covariates"]] %>%
+  mutate(`Review ID` = as.numeric(str_replace(`Review ID`, "[:alpha:]", "")))
+modelData <- reviewResults[["Modelling"]]
 ```
 
-```{r Readin, echo = FALSE}
-reviewResults <- read_rds("Analysis/Outputs/ReviewResults.rds")
+```{r Load spatial data, echo = FALSE, results = FALSE, cache= TRUE}
+# Silently load and cache all spatial data to suppress annoying messages
+
+# Load shapefiles of countries & continents of the world
+countriesShapefile <- st_read("Analysis/SpatialData/Countries/World_Countries__Generalized_.shp")
+continentShapefile <- st_read("Analysis/SpatialData/Continents/World_Continents.shp")
 ```
 
-```{r Study, echo = FALSE}
+## Study 
+Basic information on the locations and affiliations of authors involved with each paper was collected.
+
+### Authorship countries {.tabset}
+Two thirds of papers (36/54) were by authors from U.S. institutions; the next greatest total was 3 papers from Australia.
+
+#### Map
+```{r Authorship countries map, echo = FALSE, results = FALSE, cache = TRUE}
+## Map of authorship countries by # of papers
+studyData %>%
+  # Get count of papers by author's first country
+  group_by(`First author country`) %>%
+  summarise(Count = n()) %>%
+  arrange(-Count) %>% 
+  mutate(`First author country` = case_when(`First author country` == "United States of America" ~ "United States",
+                                            `First author country` != "United States of America" ~ `First author country`)) %>%
+  # Merge by country name
+  right_join(countriesShapefile, by = c('First author country' = 'COUNTRY')) %>%
+  # Plot map
+  ggplot() +
+  geom_sf(aes(geometry = geometry, fill = log(Count))) +
+  labs(title = "Log paper count by country") +
+  theme_void() + 
+  scale_fill_continuous(name = "Log counts", low="thistle2", high="darkred",na.value="white")
+```
+
+#### Table
+```{r Authorship countries table, echo = FALSE}
+# Simple able with the same data as map
+studyData %>%
+  group_by(`First author country`) %>%
+  summarise(`Total papers` = n()) %>%
+  arrange(-`Total papers`) %>%
+  kable()
+```
+
+### Authorship affiliations
+Affiliated institutions were split into four categories: Academic, Non-governmental organisations, Governmental agencies, and private sector groups.
+These were fairly varied, with academic groups participating in the vast majority of papers but with substantial representation from governmental and NGO groups.
+
+```{r Affiliation types, echo = FALSE}
+# Calc % of each type
+studyData %>%
+  summarise(NGO = sum(Author_NGO)/n()*100,
+            Academic = sum(Author_Academic)/n()*100,
+            Government = sum(Author_Government/n()*100),
+            Private = sum(Author_Private)/n()*100) %>% 
+  # Make legible for table
+  pivot_longer(cols = c(NGO, Academic, Government, Private), names_to = 'Affiliation', values_to = "% of papers") %>%
+  arrange(-`% of papers`) %>%
+  kable(digits = 1)
+```
+
+***
+
+## Objectives
+Five categories of objectives and goals were assessed for each paper; these were not mutually exclusive and many papers had multiple objectives.
+
+* Assessing trends: interested in the trajectory of estimated or derived parameters
+* Testing hypotheses: evaluating attributes of a **specific** covariate on parameters
+* Spatial mapping: Extrapolating estimates of parameter values beyond surveyed sites
+* Future projections: Making predictions of parameters into the future
+* Methods: introducing, testing, or demonstrating a new extension to these models
+
+### Papers per objective
+Monitoring trends and testing hypotheses were the most popular objectives; making predictions (spatial or temporal) was quite rare.
+```{r Objectives table, echo = FALSE}
+# Make table 
+objData %>% 
+  pivot_longer(cols = c(Obj_Methods, Obj_Projections, Obj_Hypothesis, Obj_Spatial, Obj_Trends),
+               names_to = "Objective", values_to = "TF") %>%
+  # Drop obj_ prefix
+  mutate(Objective = str_replace(Objective, "Obj_","")) %>%
+  # Tally up and make table
+  group_by(Objective) %>%
+  summarise(`Percent of papers` = 100*sum(TF)/54) %>%
+  arrange(-`Percent of papers`) %>%
+  kable(digits = 1)
+```
+
+### Objectives by authorship group
+Objectives did appear to vary based on whether certain types of authors worked on the paper.
+Where government or NGO authors were present, papers tended to assess trends moreso than testing hypotheses.
+```{r Obj by Authorship group, echo = FALSE}
+objByAuth <- full_join(objData, studyData, by = c("Review ID", "Title")) %>%
+  # Pivot authorship column and adjust  
+  pivot_longer(cols = c(Author_NGO, Author_Academic, Author_Government, Author_Private), 
+               names_to = 'Authorship', values_to = "Auth_TF") %>%
+  mutate(Authorship = str_replace(Authorship, "Author_","")) %>%
+  # Get count of # of papers by authorship for the plot later
+  filter(Auth_TF == TRUE) %>%
+  group_by(Authorship) %>%
+  mutate(nAuth = n()) %>% 
+  # Pivot objectives column and adjust
+  pivot_longer(cols = c(Obj_Methods, Obj_Projections, Obj_Hypothesis, Obj_Spatial, Obj_Trends),
+               names_to = "Objective", values_to = "Obj_TF") %>%
+  mutate(Objective = str_replace(Objective, "Obj_","")) %>%
+  filter(Obj_TF == TRUE)
+
+# Process to format suited for barplot
+objByAuth %>%
+  # Get tallies of papers per objective per affiliation
+  filter(Obj_TF == TRUE & Auth_TF == TRUE) %>%
+  group_by(Authorship) %>%
+  mutate(TotalAuth = length(unique(Title))) %>%
+  group_by(Authorship, Objective, nAuth) %>%
+  summarise(Count = n()/nAuth) %>%
+  ungroup() %>%
+  # Add categories which had 0 representative papers
+  complete(Authorship, Objective) %>%
+  mutate(Count = case_when(is.na(Count) ~ 0.005,
+                           !is.na(Count) ~ Count)) %>%
+  # Reorder factors for barplot
+  mutate(Objective = factor(Objective, levels = c("Trends", "Hypothesis", "Methods", "Projections", "Spatial"))) %>%
+
+  # Generate barplot
+  ggplot(aes(x = Authorship, y = Count, fill = Objective)) +
+  geom_col(position = "dodge") +
+  ylim(0, 1) + 
+  labs(title = "Percent of papers by affiliation pursuing each objective",
+       x = "Affiliation type", y = "% of papers")
+```
+
+### Covariates by objection
+```{r Covs by obj, echo = FALSE}
+# Create cross-table of objective, parameter, and # of covariates by merging relavent sheets
+objData %>%
+  # Process objective column
+  pivot_longer(cols = c(Obj_Methods, Obj_Projections, Obj_Hypothesis, Obj_Spatial, Obj_Trends),
+               names_to = "Objective", values_to = "Obj_TF") %>%
+  mutate(Objective = str_replace(Objective, "Obj_","")) %>%
+  filter(Obj_TF == TRUE) %>%
+  select(`Review ID`, Objective) %>%
+  # Add covariates data
+  full_join(covData, by = 'Review ID') %>%
+  group_by(`Review ID`, Objective, Parameter) %>%
+  summarise(Count = n()) %>%
+  # Filter to key parameters and drop outliers 
+  filter(Parameter %in% c("Colonisation", "Extinction", "InitialOccupancy", "Detection"),
+         Count < 20) %>%
+  # Reorder factors for barplot 
+  mutate(Parameter = factor(Parameter, 
+                            levels = c("InitialOccupancy", "Colonisation", "Extinction", "Detection"))) %>%
+  mutate(Objective = factor(Objective, 
+                            levels = c("Trends", "Hypothesis", "Methods", "Projections", "Spatial"))) %>%
+
+  # Generate boxplot
+  ggplot(aes(x = Objective, y = Count, colour = Parameter)) +
+  geom_boxplot() + 
+  labs(title = "Number of covariates by objective", 
+       x = "Paper objective", y = "Number of covariates")
+```
+
+### Selection by objective
+*A priori* model selection was the most popular method for each of the five objective categories, with more complex options quite rare.
+*Note that the model selection data needs to be cleaned a bit more to fix overlapping categories*
+```{r Obj by selection method, echo = FALSE}
+# Create cross-table of objective, parameter, and # of covariates by merging relevent sheets
+objData %>%
+  # Process objective column
+  pivot_longer(cols = c(Obj_Methods, Obj_Projections, Obj_Hypothesis, Obj_Spatial, Obj_Trends),
+               names_to = "Objective", values_to = "Obj_TF") %>%
+  mutate(Objective = str_replace(Objective, "Obj_","")) %>%
+  filter(Obj_TF == TRUE) %>%
+  select(`Review ID`, Objective) %>% 
+  # Merge with model selection data
+  right_join(modelData, by = "Review ID") %>%
+  # Add total # of papers in each select category
+  group_by(Objective) %>% 
+  mutate(nSelect = n()) %>%
+
+  # Make crosstable for Objective and model selection category
+  group_by(Objective, Select_Category) %>%
+  summarise(Percent = n()/nSelect) %>%
+  ungroup() %>%
+  distinct() %>%
+  # Add missing combos for plot
+  complete(Objective, Select_Category) %>%
+  mutate(Percent = case_when(is.na(Percent) ~ 0.01,
+                             !is.na(Percent) ~ Percent)) %>%
+  # Rearrange factors for plot
+  mutate(Select_Category = factor(Select_Category, levels = c("Apriori", "Exhaustive", "PbyP", 
+                                                              "Step", "Simple", "Other", "None")),
+         Objective = factor(Objective, levels = c("Trends", "Hypothesis", "Methods", "Projections", "Spatial"))) %>%
+
+  #Generate bar plot
+  ggplot(aes(x = Objective, y = Percent, fill = Select_Category)) +
+  geom_col(position = "dodge") +
+  ylim(0, 1) +
+  scale_fill_discrete(name = "Selection category", 
+                    labels = c("A priori", "Exhaustive", "Param by param", "Stepwise",
+                               "Simpler models", "Other", "None")) +
+  labs(title = "Selection methods used by objective type",
+       x = "Objective category", y = "Percentage of papers")
+```
+
+***
+
+## Taxa and location
+Details on the study organism was collected for each study to assess any trends in the types of species these models are applied to, with a particular focus on those that are vulnerable to conservation threats.
+The locations where surveys were conducted was also noted.
+
+As with authorship, the United States dominated the countries where surveys took place.
+
+### Study site maps {.tabset}
+As with authorship, the United States/North America dominated the countries where surveys took place.
+
+#### Country
+```{r Site country map, echo = FALSE}
+taxaData %>%
+  # Divide by unique title/country
+  group_by(Country, Title) %>%
+  summarise() %>%
+  # Get count of papers by site country
+  group_by(Country) %>%
+  summarise(Count = n()) %>%
+  arrange(-Count) %>% 
+  mutate(Country = case_when(Country == "United States of America" ~ "United States",
+                             Country != "United States of America" ~ Country)) %>%
+  # Merge by country name
+  right_join(countriesShapefile, by = c('Country' = 'COUNTRY')) %>%
+  # Plot map
+  ggplot() +
+  geom_sf(aes(geometry = geometry, fill = log(Count))) +
+  labs(title = "Log paper count by country") +
+  theme_void() + 
+  scale_fill_continuous(name = "Log counts", 
+                        low = "thistle2", 
+                        high = "darkred",
+                        na.value = "white")
+
+# Table with the same data
+taxaData %>%
+  # Divide by unique title/country
+  group_by(Country, Title) %>%
+  summarise() %>%
+  # Get count of papers by site country
+  group_by(Country) %>%
+  summarise(`Total papers` = n()) %>%
+  arrange(-`Total papers`) %>% 
+  kable()
+```
+
+#### Continent
+```{r Site continent map, echo = FALSE}
+taxaData %>%
+  # Divide by unique title/country
+  group_by(Country, Continent, Title) %>%
+  summarise() %>%
+  # Get count of papers by site continent
+  group_by(Continent) %>%
+  summarise(Count = n()) %>%
+  arrange(-Count) %>% 
+  mutate(Continent = case_when(Continent == "Oceania" ~ "Australia",
+                               Continent != "Oceania" ~ Continent)) %>%
+
+  # Merge by continent name
+  right_join(continentShapefile, by = c('Continent' = 'CONTINENT')) %>%
+  # Plot map
+  ggplot() +
+  geom_sf(aes(geometry = geometry, fill = Count)) +
+  labs(title = "Paper count by continent") +
+  theme_void() + 
+  scale_fill_continuous(name = "Log counts", 
+                        low = "thistle2", 
+                        high = "darkred",
+                        na.value = "white")
+
+# Table with the same data
+taxaData %>%
+  # Divide by unique title/country
+  group_by(Country, Title, Continent) %>%
+  summarise() %>%
+  # Get count of papers by continent
+  group_by(Continent) %>%
+  summarise(`Total papers` = n()) %>%
+  arrange(-`Total papers`) %>% 
+  kable()
 
@@
 
-```{r Objectives, echo = FALSE}
-
-reviewResults$Objectives %>%
-  pivot_longer(c("Obj_Methods","Obj_Projections","Obj_Spatial","Obj_Trends","Obj_Hypothesis"),
-               "Objective") %>%
-  mutate(value = as.numeric(value)) %>%
-  group_by(Objective) %>%
-  summarise(n=sum(value)) %>%
-
-  ggplot(aes(x = Objective, y = n/54)) +
-  geom_col() +
-  ylim(0,1)
+#### Realm
+```{r Site realm map, echo = FALSE}
+#ecoregionsShapefile <- st_read("Analysis/SpatialData/Ecoregions/Terrestrial_Ecoregions.shp") %>%
+#  st_make_valid() 
+#
+#t <- ecoregionsShapefile %>%
+#  group_by(WWF_REALM2) %>%
+#  summarise(geometry = st_combine(geometry))
+#
+#taxaData %>%
+#  # Divide by unique title/country
+#  group_by(Country, Realm, Title) %>%
+#  summarise() %>%
+#  # Get count of papers by site continent
+#  group_by(Realm) %>%
+#  summarise(Count = n()) %>%
+#  arrange(-Count) %>% 
+#  
+#  # Merge by realm
+#  right_join(ecoregionsShapefile, by = c('Realm' = 'WWF_REALM2')) %>%
+#  # Plot map
+#  ggplot() +
+#  geom_sf(aes(geometry = geometry, fill = Count)) +
+#  labs(title = "Paper count by Udvardy realm") +
+#  theme_void() + 
+#  scale_fill_continuous(name = "Log counts", 
+#                        low = "thistle2",
+#                        high = "darkred",
+#                        na.value="white")
+#
+# Table with the same data
+taxaData %>%
+  # Divide by unique title/country
+  group_by(Country, Title, Realm) %>%
+  summarise() %>%
+  # Get count of papers by site realm
+  group_by(Realm) %>%
+  summarise(`Total papers` = n()) %>%
+  arrange(-`Total papers`) %>% 
+  kable()
 ```
 
-```{r Taxa, echo = FALSE}
+### Study species
+Taxa were roughly split between mammals, birds, and herptiles; fish and invertebrates were scantly represented. 16 were recognised as threatened species by the authors, and 8 were invasives.
 
+```{r Study species, echo = FALSE}
+taxaData %>% 
+  # Simplify out any repeats
+  group_by(`Review ID`, Taxa, `Taxa keywords`) %>%
+  summarise() %>%
+  # Tally up and table
+  group_by(Taxa = `Taxa keywords`) %>%
+  summarise(Count = n()) %>%
+  arrange(-Count) %>%
+  kable
 ```
 
-```{r Survey, echo = FALSE}
-hist(as.numeric(reviewResults$Survey$`Study duration`))
-
-reviewResults$Survey %>%
-  group_by(`Capture method`) %>%
-  summarize(n = n()) %>%
-
-  ggplot(aes(x = reorder(`Capture method`, -n), y = n)) +
-  geom_col() +
-  theme(axis.text.x = element_text(angle=90, vjust = 0.5,hjust=1))
-
+### Threat categories
+Threats to each species which were mentioned by the author were denoted. Habitat loss was the most common threat mentioned, followed by invasive species.
+```{r Threat categories, echo = FALSE}
+taxaData %>% 
+  pivot_longer(c(Threat_HabitatLoss, Threat_Exploitation, Threat_ClimateChange,
+                 Threat_Invasives, Threat_Disaster, Threat_Disease), 
+               names_to = 'Threat', values_to = 'ThreatTF') %>%
+  mutate(Threat = str_replace(Threat, "Threat_", "")) %>%
+  group_by(Threat) %>%
+  summarise(Percent = 100*sum(ThreatTF)/94) %>%
+  arrange(-Percent) %>%
+  kable(digits = 2)
 ```
 
-```{r Covariates, echo = FALSE}
-nrow(reviewResults$Covariates)
+### Study size
+Study area followed an apparent log-normal distribution centered on the 100-10000km^2 range
+```{r Study size, echo = FALSE}
+taxaData %>% 
+  #Simplify out any repeats
+  group_by(`Review ID`, `Study Size`) %>%
+  summarise() %>%
+  # Tally up
+  group_by(`Study Size`) %>%
+  summarise(Count = n()) %>%
+  filter(!`Study Size` %in% c('Not specified', 'Unspecified')) %>%
 
-reviewResults$Covariates %>%
-  filter(!is.na(Category)) %>%
-  group_by(Category, Scale) %>%
-  summarize(Count = n()) %>%
-  arrange(Count) %>%
-
-  ggplot(aes(x = reorder(Category, -Count), y = Count, fill = Scale)) +
-  geom_col() + 
-  scale_fill_discrete(limits = c("a","p","d","g","u", NA), 
-                    labels = c("area","point","distance","global", "Unclear", "NA")) +
-  theme(axis.text.x = element_text(angle=90, vjust = 0.5,hjust=1))
-
-reviewResults$Covariates %>%
-  filter(!is.na(Category)) %>%
-  group_by(Category, Dynanimity) %>%
-  summarize(Count = n()) %>%
-  arrange(Count) %>%
-
-  ggplot(aes(x = reorder(Category, -Count), y = Count, fill = Dynanimity)) +
-  geom_col() + 
-  scale_fill_discrete(limits = c("d","s","u"), 
-                    labels = c("Dynamic", "Static", "Unclear")) +
-  theme(axis.text.x = element_text(angle=90, vjust = 0.5,hjust=1))
-
+  # Make horizontal bar plot
+  ggplot(aes(x = Count, y = `Study Size`)) +
+  geom_col(fill = "darkred") +
+  labs(title = "Papers by study size",
+       x = "Number of papers", y = "Study area in km^2")
 ```
 
-```{r Modelling, echo = FALSE}
+### Covariates by study size
+```{r Covariates by study size, echo = FALSE}
+taxaData %>% 
+  #Simplify out any repeats
+  group_by(`Review ID`, `Study Size`) %>%
+  summarise() %>%
+  # Drop unspecified
+  group_by(`Study Size`) %>%
+  filter(!`Study Size` %in% c('Not specified', 'Unspecified')) %>%
+  # Join to covariate table
+  left_join(covData, by = 'Review ID') %>%
+  # Filter to core parameters
+  filter(Parameter %in% c('InitialOccupancy', 'Colonisation', 'Extinction', 'Detection')) %>%
+  # Tally up
+  group_by(`Review ID`, Parameter) %>%
+  summarise(Count = n(), `Study Size` = `Study Size`) %>%
+  # Reorder factors
+  mutate(Parameter = factor(Parameter,
+                            levels = c('InitialOccupancy', 'Colonisation', 'Extinction', 'Detection'))) %>%
 
+  # Create boxplot
+  ggplot(aes(x = Count, y = `Study Size`, colour = Parameter)) +
+  geom_boxplot() + 
+  scale_colour_discrete(limits = c('Detection', 'Extinction', 'Colonisation', 'InitialOccupancy')) +
+  labs(title = "Number of covariates by study size",
+       x = "Number of covariates", y = "Study size in km^2")
 ```
+
+***
+
+## Survey methods
+Information on the data collection and sampling processes was collected from all papers.
+
+### Study duration
+Average study length from first to last survey was 83 months (~7 years), although they varied from just one month to 40 years in duration.
+
+```{r Survey, echo = FALSE}
+surveyData %>%
+  mutate(`Study duration` = as.numeric(`Study duration`)) %>%
+  # Remove duplicates
+  group_by(`Review ID`, `Study duration`) %>%
+  summarise() %>% 
+  ungroup() %>%
+  mutate(Mean = mean(`Study duration`)) %>%
+
+  # Basic histogram
+  ggplot(aes(x = `Study duration`)) +
+  geom_vline(aes(xintercept = Mean), colour = "thistle4") +
+  geom_histogram(fill = "darkred", bins = 24) +
+  labs(title = "Duration of studies", 
+       x = "Study duration in months", y = "Number of papers") 
+```
+
+### Sampling occasions {.tabset}
+The primary sampling occasion (season) and secondary occasion (generally 'visit', used to calculate detection) were recorded.
+Number of secondary occasions per primary occasion sometimes varied within a study, min and max counts are noted.
+
+#### Primary occasions
+The median number of primary occasions from the reviewed studies was *5*, although two outlying studies had over *200* primary occasions. 
+```{r Primary occasions, echo = FALSE, results = FALSE, message = FALSE}
+surveyData %>% 
+  group_by(`Review ID`, MaxPrimary) %>%
+  summarise() %>%
+  ungroup() %>%
+  mutate(Median = median(MaxPrimary)) %>%
+
+  ggplot(aes(x = MaxPrimary)) +
+  geom_histogram(fill = "darkred") +
+  geom_vline(aes(xintercept = Median), colour = "thistle4") +
+  labs(title = "Number of primary occasions per study",
+       subtitle = "Median Line Indicated", 
+       x = "Number of primary occasions", y = "Number of papers")
+```
+
+#### Secondary occasions
+Studies had a median of *4* secondary occasions, with a maximum of *99*
+```{r Secondary occasions, echo = FALSE, results = FALSE, message = FALSE, warning = FALSE}
+surveyData %>% 
+  group_by(`Review ID`, MaxSecondary) %>%
+  summarise() %>%
+  ungroup() %>%
+  mutate(Median = median(MaxSecondary, na.rm = TRUE)) %>%
+
+  ggplot(aes(x = MaxSecondary)) +
+  geom_histogram(fill = "darkred") +
+  geom_vline(aes(xintercept = Median), colour = "thistle4") +
+  labs(title = "Number of secondary occasions per study",
+       subtitle = "Median line indicated", 
+       x = "Number of secondary occasions", y = "Number of papers")
+```
+
+### Capture method
+The most common method for detections was 'audiovisual,' where human observers visually or acoustically sampled the target taxa in a structured manner (e.g., as a transect or point count).
+Searches (where the site was preferentially searched for the species) and trapping was also common.
+Citizen science was involved in 7 papers.
+```{r Capture method table, echo = FALSE}
+surveyData %>% 
+  group_by(`Review ID`, `Capture method`) %>%
+  summarise() %>%
+  group_by(`Capture method`) %>% 
+  summarise(`Number of papers` = n()) %>%
+  arrange(-`Number of papers`) %>%
+  kable()
+```
+
+### Site types
+The definition of a 'site' (the level of occupied and unoccupied states) was variable.
+Irregular areas (like habitat patches or ponds) were the most common, followed by cartesian points and grid cells.
+```{r Site type, echo = FALSE}
+surveyData %>% 
+  group_by(`Review ID`, `Type site`) %>%
+  summarise() %>%
+  group_by(`Type site`) %>% 
+  summarise(`Number of papers` = n()) %>%
+  arrange(-`Number of papers`) %>%
+  kable()
+```
+
+### Number of sites
+
+The number of sites ranged from 7 to 9394, with a mean of *377* sites.
+```{r Number of sites, echo = FALSE, message = FALSE, warning = FALSE}
+surveyData %>%
+  group_by(`Review ID`, NumSites) %>%
+  summarise() %>%
+
+  ggplot() +
+  geom_histogram(aes(x = NumSites), fill = "darkred") +
+  scale_x_continuous(trans = "log10") +
+  labs(title = "Number of sites per survey", subtitle = "Log transformed",
+       x = "Number of sites", y = "Number of papers")
+```
+
+### Number of sites by study size
+As would be expected, the number of sites increases with the size of the study area.
+```{r Num sites by study size, echo = FALSE, warning = FALSE}
+surveyData %>%
+  left_join(taxaData, by = "Review ID")  %>%
+  group_by(`Review ID`, NumSites, `Study Size`) %>%
+  summarise() %>%
+  filter(!`Study Size` %in% c("Unspecified", "Not specified")) %>%
+
+  ggplot() +
+  geom_boxplot(aes(x = `Study Size`, y = NumSites), colour = "darkred") +
+  scale_y_continuous(trans = "log10") + 
+  labs(title = "Number of sites by study size",
+       x = "Study size (km2)", y = "Number of Sites (log transformed)") +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
+```
+### Covariates by site count
+
+```{r Number of covariates by sites, echo = FALSE, message = FALSE, warning = FALSE}
+surveyData %>%
+  left_join(covData, by = "Review ID")  %>%
+  mutate(`Total covariates` = as.numeric(`Total covariates`)) %>%
+  group_by(`Review ID`, NumSites, Parameter, `Total covariates`) %>%
+  summarise() %>%
+  filter(Parameter %in% c('InitialOccupancy', 'Colonisation', 'Extinction', 'Detection')) %>%
+  group_by(Parameter) %>%
+
+  ggplot(aes(x = NumSites, y = `Total covariates`, colour = Parameter)) +
+  geom_point(alpha = 0.5, size = 4, position = "jitter") +
+  geom_smooth(method = "lm", se = FALSE) +
+  scale_x_continuous(trans = "log10") + scale_y_continuous(trans = "log10") +
+  labs(title = "Covariate count by number of sites", subtitle = "Log transformed",
+       x = "Number of sites", y = "Total number of covariates")
+```
+
+### Sampling method
+Methods used to select sites were documented - this section needs to be revisited, as some categories can likely be combined.
+The most interesting question here may in whether selection falls under a broad 'random' or 'preferential' system, and how that relates to objectives or covariates.
+```{r Sampling method table, echo = FALSE}
+surveyData %>%
+  group_by(`Review ID`, `Sampling method`) %>%
+  summarise() %>%
+  group_by(`Sampling method`) %>%
+  summarise(`Number of papers` = n()) %>%
+  arrange(-`Number of papers`) %>%
+  kable()
+```
+
+***
+
+## Parameters and covariates
+The most complex and in-depth category of data was the covariates, which were organised by parameter and covariate type.
+
+### Parameters estimated
+The vast majority of papers used the four core parameters of the standard DOM: **Initial occupancy**, **colonisation**, **extinction**, and **detection** (although 12 used the reciprocal of **Extinction**, **Persistence**)
+
+```{r Parameter table, echo = FALSE}
+covData %>% 
+  group_by(`Review ID`, Parameter) %>% 
+  summarise() %>%
+  group_by(Parameter) %>%
+  summarise(`Number of papers` = n()) %>%
+  arrange(-`Number of papers`) %>% 
+  mutate(Percentage = `Number of papers` / 54)
+  kable()
+```
+
+### Multi-species
+*9* models included at least one multi-species parameter.
+```{r Multi-species papers, echo = FALSE}
+covData %>%
+  group_by(`Review ID`, MultiSpecies, Parameter) %>%
+  summarise() %>%
+  group_by(`Review ID`) %>%
+  summarise(MultiSpecies = sum(MultiSpecies)) %>%
+  mutate(MultiSpecies = case_when(MultiSpecies >0 ~ 1,
+                                  MultiSpecies == 0 ~ 0)) %>%
+  summarise(sum(MultiSpecies))
+```
+
+### Covariate types by parameter
+```{r Covariates by parameter, echo = FALSE, warning = FALSE}
+covData %>%
+  group_by(`Review ID`, Parameter) %>%
+  summarise() %>%
+  filter(Parameter %in% c('InitialOccupancy', 'Colonisation', 'Extinction', 'Detection')) %>%
+  group_by(Parameter) %>%
+  mutate(ParamTotal = n()) %>%
+  left_join(covData[,c('Review ID', 'Parameter', 'Category')], by = c("Parameter", "Review ID")) %>%
+  group_by(Parameter, Category, ParamTotal) %>%
+  summarise(Percent = n()/ParamTotal) %>%
+  group_by(Parameter, Category, Percent) %>%
+  summarise() %>%
+  ungroup() %>%
+  complete(Parameter,Category) %>%
+  filter(!is.na(Category),
+         !Category %in% c('Other', 'Unspecified')) %>%
+
+  ggplot() +
+  geom_col(aes(x = Category, y = Percent, fill = Category)) +
+  ylim(c(0,1)) +
+  facet_grid(rows = vars(Parameter)) +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
+```
+
+### Number of covariate by parameter
+```{r Covariate count by parameter, echo = FALSE, warning = FALSE, message = FALSE}
+covData %>%
+  group_by(`Review ID`, Parameter, `Total covariates`) %>%
+  summarise() %>%
+  mutate(`Total covariates` = as.numeric(`Total covariates`)) %>%
+  filter(Parameter %in% c('InitialOccupancy', 'Colonisation', 'Extinction', 'Detection')) %>%
+
+  ggplot() +
+  geom_histogram(aes(x = `Total covariates`, fill = Parameter), binwidth = 1) +
+  facet_grid(rows = vars(Parameter)) +
+  labs(title = "Number of covariates by parameter type",
+      x = "Number of covariates", y = "Number of papers")
+```
+
+### Dynamicity in covariate types
+Covariates varied drastically in whether they were **static** or **dynamic**
+```{r Dynamic covariates, echo = FALSE}
+covData %>%
+  group_by(`Review ID`, Category, Parameter, Dynanimity) %>%
+  summarise() %>%
+  filter(!is.na(Category),
+         !Category %in% c("Other", "Unspecified"),
+         Parameter %in% c("InitialOccupancy", "Colonisation", "Extinction", "Detection")) %>%
+  group_by(Category) %>%
+  mutate(nCategory = n()) %>%
+  group_by(Category, nCategory, Dynanimity) %>%
+  summarise(Percent = n()/nCategory) %>%
+  group_by(Category, Dynanimity, Percent) %>%
+  summarise() %>%
+
+  ggplot() +
+  geom_col(aes(y = Category, x = Percent, fill = Dynanimity)) +
+  labs(title = 'Dynanimity by covariate type',
+       x = 'Percent of covariates', y = 'Covariate category') +
+  scale_fill_discrete(labels = c("Dynamic", "Static", "Unclear"))
+```
+
+### Covariate scale
+'Scale' defined as whether the covariate represents a point, area, distance, etc.
+
+```{r Covariate scale, echo = FALSE}
+covData %>%
+  group_by(`Review ID`, Category, Parameter, Scale) %>%
+  summarise() %>%
+  filter(!is.na(Category),
+         !Category %in% c("Other", "Unspecified"),
+         Parameter %in% c("InitialOccupancy", "Colonisation", "Extinction", "Detection")) %>%
+  group_by(Category) %>%
+  mutate(nCategory = n()) %>%
+  group_by(Category, nCategory, Scale) %>%
+  summarise(Percent = n()/nCategory) %>%
+  group_by(Category, Scale, Percent) %>%
+  summarise() %>%
+
+  ggplot() +
+  geom_col(aes(y = Category, x = Percent, fill = Scale)) +
+  labs(title = 'Scale by covariate type',
+       x = 'Percent of covariates', y = 'Covariate category') +
+  scale_fill_discrete(labels = c("Area", "Distance", "Global", "Point", "Unspecified/Unclear"))
+```
+
+***
+
+## Modelling decisions
+
+### Selection method
+The majority of papers selected models **a priori** (either single or multiple)
+```{r Selection category, echo = FALSE}
+modelData %>%
+  group_by(`Review ID`, Select_Category) %>%
+  summarise() %>%
+  group_by('Selection category' = Select_Category) %>%
+  summarise(Percent = n()/54) %>%
+  arrange(-Percent) %>%
+  kable()
+```
+
+### Percent bayesian
+*41%* of papers used a bayesian implementation
+```{r Percent bayesian, echo = FALSE, include = FALSE}
+modelData %>%
+  group_by(`Review ID`, Bayesian) %>%
+  summarise() %>%
+  mutate(Bayesian = case_when(Bayesian == "Y" ~ 1,
+                              Bayesian == "N" ~ 0)) %>%
+  group_by(Bayesian) %>%
+  summarise(Bayesian = sum(Bayesian))
+22/54
+```
+
+### Bayesian selection methods
+
+```{r Bayesian selection methods, echo = FALSE}
+modelData %>%
+  group_by(`Review ID`, Select_Category, Bayesian) %>%
+  summarise() %>% 
+  mutate(n = case_when(Bayesian == "Y" ~ 22,
+                       Bayesian == "N" ~ 32)) %>%
+  mutate(Bayesian = case_when(Bayesian == "Y" ~ "Bayesian",
+                              Bayesian == "N" ~ "Frequentist")) %>%
+  group_by('Selection category' = Select_Category, Bayesian) %>%
+  summarise(Percent = n()/n) %>%
+  group_by(`Selection category`, Bayesian, Percent) %>%
+  summarise() %>%
+
+  ggplot() +
+  geom_col(aes(x = `Selection category`, y = Percent, fill = `Selection category`)) +
+  ylim(c(0,1)) +
+  facet_grid(rows = vars(Bayesian)) 
+```

### 2023-02-09 Analysis/Code/2_InitialAnalysis.Rmd

@@
 ```
+
 ### Covariates by site count
@@
   arrange(-`Number of papers`) %>% 
-  mutate(Percentage = `Number of papers` / 54)
+  mutate(Percentage = `Number of papers` / 54) %>%
   kable()
@@
 *9* models included at least one multi-species parameter.
-```{r Multi-species papers, echo = FALSE}
+```{r Multi-species papers, echo = FALSE, include = FALSE}
 covData %>%
@@
   arrange(-Percent) %>%
-  kable()
+  kable(digits = 2)
 ```

### 2023-11-27 Manuscript/index.qmd

---
title: Dynamic occupancy models review
date: today
authors:
  - name: Saoirse Kelleher
    affiliation: University of Melbourne
    email: saoirse.kelleher@student.unimelb.edu.au
    orcid: 0000-0002-4614-5753
    corresponding: true
  - name: Natalie Briscoe
    affiliation: University of Melbourne
  - name: Damaris Zurell
    affiliation: University of Potsdam
  - name: Marc Kery
    affiliation: Vogelwarte
  - name: Gurutzeta Guillera-Arroita
    affiliation: Consejo Superior de Investigaciones Científicas (CSIC)
  - name: Jane Elith
    affiliation: University of Melbourne

abstract: >
  Dynamic occupancy models allow researchers to estimate species occupancy across space and time while accounting for imperfect detection.

bibliography: references.bib
---

## Introduction

## Methods

## Results

## Discussion

## Conclusions

### 2023-11-27 Randomisation/Randomise_Papers.qmd
reviewed first sample randomisation

---
title: Randomise papers
subtitle: Describes random selection of papers for inclusion in the review
author: Saoirse Kelleher
date: today
format: 
  html:
    code-fold: true
    theme: minty
bibliography: references.bib
---

## Overview

As this review evolved, two samples of target papers were acquired from two distinct sampling protocols.

-   The **First Sample** was acquired in October 2021 from Google Scholar. The first 100 hits for each of four search terms ('dynamic occupancy model,' 'occupancy dynamics model,' 'multi-season occupancy model,' and 'stochastic patch occupancy model', plus allowances for pluralisation) were downloaded. These were stratified by their query and year-window (2000-2005, 2006-2010, 2010-2015, and 2016-2021). A random sample of 25% of the strata *or* 5 papers, whichever was larger, was flagged for inclusion in the review, to be replaced from the same strata if inclusion criteria were not met.

-   After reviewing this initial sample, the need for a **Second Sample** was identified as many of the papers included in the first did not represent our model of interest. This second, more narrowly targeted sample was drawn from papers citing @mackenzie2003, the original description of the model class which all subsequent implementations would be expected to cite. The same 4 year-window strata were used here, with 10 papers randomly drawn from each strata.

## First sample

The full list of papers acquired from the aforementioned queries is read into R, and each is assigned to a year-window.

```{r}
#| message: false
library("knitr")
library("tidyverse") 
library("readxl")

set.seed(5378)

# From the excel review spreadsheet, read in only the identifying characteristics 
identifierSheet <- readxl::read_xlsx("Spreadsheets/Stage1_Spreadsheet.xlsx", 
                                     sheet = "Metadata", skip = 1) %>%
  # Drop rows prior to start of data
  slice(4:nrow(.)) %>%
  # Only need google search query and the ID reference
  select(Review_ID, Query = `Search terms`)

# Repeat with the study metadata
studyDataSheet <- readxl::read_xlsx("Spreadsheets/Stage1_Spreadsheet.xlsx", 
                                    sheet = "Study data", skip = 1) %>%
  # Cut out non-data rows
  slice(4:nrow(.)) %>%
  # Only need year data and ID reference
  select(Review_ID, Year)

# Merge paper queries and metadata
paperList <- full_join(identifierSheet, studyDataSheet, 
                       by = "Review_ID")

# Exclude some non-query papers and add year periods to the rest
includedPapers <- paperList %>%
  # Drop empty rows, excluded, and incidental papers
  filter(!is.na(Query) & Query != "INCIDENTAL" & Year != "EXCLUDE") %>%
  # Break into time periods 
  mutate(YearPeriod = case_when(Year %in% as.character(2000:2005) ~ "2000:2005",
                                Year %in% as.character(2006:2010) ~ "2006:2010",
                                Year %in% as.character(2011:2015) ~ "2011:2015",
                                Year %in% as.character(2016:2021) ~ "2016:2021"))

head(includedPapers) %>%
  kable()
```

A random number is then assigned to *all* papers. Papers are then ranked by number within their year-window/query strata. Papers in the top 5/25% (whichever was larger) were flagged to be read, and the rest were kept available for replacement of papers which may fail to meet inclusion criteria.

```{r Assign random numbers}
# Randomly rank papers, then rank within the strata
rankedPapers <- includedPapers %>%
  # Add random ranks for all papers
  mutate(randRank = sample(x = 1:nrow(.), size = nrow(.))) %>%
  # Re-rank within strata
  group_by(Query, YearPeriod) %>%
  arrange(Query, YearPeriod, randRank) %>%
  mutate(GroupRank = row_number())

# Identify which papers qualify to be read
highlightedPapers <- rankedPapers %>%
  group_by(Query, YearPeriod) %>%
  mutate(PercentRank =  GroupRank/max(GroupRank)) %>%
  mutate(ToRead = case_when(
    # If there are fewer than 5, read them all
    max(GroupRank) <= 5 ~ "YES",
    # If there are 20 or fewer, read the top 5
    max(GroupRank) <= 20 & GroupRank <= 5 ~ "YES",
    max(GroupRank) <= 20 & GroupRank >= 5 ~ "NO",
    # If there are 21 or more, read the top 25%
    max(GroupRank) >= 21 & PercentRank <= 0.25  ~ "YES",
    max(GroupRank) >= 21 & PercentRank > 0.25 ~ "NO")) %>%
  ungroup() %>%
  unite(col = Group, c(Query, YearPeriod)) %>%
  # Get Review_ID, GroupRank, and ToRead
  select(Review_ID, Group, OverallRank = randRank, GroupRank, ToRead)

head(highlightedPapers) %>%
  kable()
```

Incidental papers which were not ranked are returned to the dataframe, and the output is saved.

```{r re-add unranked papers}
unrankedPapers <- paperList %>% 
  # Drop empty rows, excluded, and incidental papers
  filter(is.na(Query) | Query == "INCIDENTAL" | Year == "EXCLUDE") %>%
  mutate(ToRead = "NO", Group = "Unranked", GroupRank = "NA", OverallRank = "NA") %>%
  select(Review_ID, Group, OverallRank, GroupRank, ToRead)

# Combine ranked and unranked papers
outputList <- rbind(highlightedPapers, unrankedPapers) %>%
  mutate(Review_ID = as.numeric(Review_ID)) %>%
  arrange(Review_ID)

# Write to disk
write_csv(outputList, file = "Randomisation/PrimarySampleRanks.csv")
```

## Second sample

## References

### 2023-11-27 Randomisation/Randomise_Papers.qmd
randomised second sample

@@
     theme: minty
+    toc: true
 bibliography: references.bib
@@
 This second, more narrowly targeted sample was drawn from papers citing @mackenzie2003, the original description of the model class which all subsequent implementations would be expected to cite.
-The same 4 year-window strata were used here, with 10 papers randomly drawn from each strata.
+Papers were acquired via Web of Science in September 2023.
+The same 4 year-window strata were used here, with 15 papers randomly drawn from each strata.
+A fifth strata was added for 2022-2023, to capture the most recent papers on these models.
+
+For transparency and reproducibility, the randomisation code is preserved in this document.
 
@@
   # Only need year data and ID reference
-  select(Review_ID, Year)
+  select(Review_ID, Title, Year)
 
@@
   # Get Review_ID, GroupRank, and ToRead
-  select(Review_ID, Group, OverallRank = randRank, GroupRank, ToRead)
+  select(Review_ID, Title, Group, OverallRank = randRank, GroupRank, ToRead)
 
@@
   mutate(ToRead = "NO", Group = "Unranked", GroupRank = "NA", OverallRank = "NA") %>%
-  select(Review_ID, Group, OverallRank, GroupRank, ToRead)
+  select(Review_ID, Title, Group, OverallRank, GroupRank, ToRead)
 
@@
   mutate(Review_ID = as.numeric(Review_ID)) %>%
-  arrange(Review_ID)
+  arrange(Review_ID) %>%
+  mutate(Sample = "First")
 
 # Write to disk
-write_csv(outputList, file = "Randomisation/PrimarySampleRanks.csv")
+write_csv(outputList, file = "Randomisation/FirstSampleRanks.csv")
 ```
@@
 
+The list of papers pulled from Web of Science is loaded in - 1184 papers are represented here. Year-windows are assigned here as in the **First Sample.**
+
+```{r Load second sample}
+#| warning: false
+
+# Load query and extract relevant columns
+wosQuery <- readxl::read_xls("Queries/MacKenzie WOS/MacKenzie03.xls") %>%
+  select(Review_ID, Title = `Article Title`, Year = `Publication Year`) %>%
+  # Assign year period
+  mutate(YearPeriod = case_when(Year %in% 2000:2005 ~ "2000:2005",
+                                Year %in% 2006:2010 ~ "2006:2010",
+                                Year %in% 2011:2015 ~ "2011:2015",
+                                Year %in% 2016:2021 ~ "2016:2021",
+                                Year %in% 2022:2023 ~ "2022:2023"))
+
+head(wosQuery) %>%
+  kable()
+```
+
+Papers are assigned a random number and ranked within their groups, and top-15 papers are flagged to be read. The output is saved for future reference.
+
+```{r randomise second sample}
+
+randomisedWOS <- wosQuery %>%
+  # Add query flag to match with the first sample
+  mutate(Query = "MacKenzie Cited") %>%
+  # Add random ranks for all papers
+  mutate(randRank = sample(x = 1:nrow(.), size = nrow(.))) %>%
+  # Re-rank within strata
+  group_by(Query, YearPeriod) %>%
+  arrange(Query, YearPeriod, randRank) %>%
+  mutate(GroupRank = row_number()) %>% 
+  mutate(ToRead = case_when(
+    GroupRank <= 15 ~ "YES",
+    GroupRank >15 ~ "NO")) %>%
+  unite(col = Group, c(Query, YearPeriod)) %>%
+  mutate(Sample = "Second") %>%
+  # Get Review_ID, GroupRank, and ToRead
+  select(Sample, Review_ID, Title, Group, OverallRank = randRank, GroupRank, ToRead)
+
+write_csv(randomisedWOS, "Randomisation/SecondSampleRanks.csv")
+
+head(randomisedWOS) %>%
+  kable()
+```
+
+## Sample summary
+
+A basic representation of the two sample's coverage is presented here.
+
 ## References

### 2023-11-27 Randomisation/Randomise_Papers.qmd
completed randomisation document

--- 2023-11-27 Randomisation/Randomise_Papers.qmd
@@
 
-```{r}
+```{r load packages}
 #| message: false
@@
 # From the excel review spreadsheet, read in only the identifying characteristics 
-identifierSheet <- readxl::read_xlsx("Spreadsheets/Stage1_Spreadsheet.xlsx", 
+first_idSheet <- readxl::read_xlsx("Spreadsheets/Stage1_Spreadsheet.xlsx", 
                                      sheet = "Metadata", skip = 1) %>%
@@
 # Repeat with the study metadata
-studyDataSheet <- readxl::read_xlsx("Spreadsheets/Stage1_Spreadsheet.xlsx", 
+first_studyDataSheet <- readxl::read_xlsx("Spreadsheets/Stage1_Spreadsheet.xlsx", 
                                     sheet = "Study data", skip = 1) %>%
@@
   # Only need year data and ID reference
-  select(Review_ID, Year)
+  select(Review_ID, Title, Year)
 
 # Merge paper queries and metadata
-paperList <- full_join(identifierSheet, studyDataSheet, 
-                       by = "Review_ID")
-
+first_paperList <- full_join(first_idSheet, first_studyDataSheet, 
+                             by = "Review_ID") %>%
   # Exclude some non-query papers and add year periods to the rest
-includedPapers <- paperList %>%
   # Drop empty rows, excluded, and incidental papers
@@
 
-head(includedPapers) %>%
+head(first_paperList) %>%
   kable()
@@
 # Randomly rank papers, then rank within the strata
-rankedPapers <- includedPapers %>%
+first_rankedPapers <- first_paperList %>%
   # Add random ranks for all papers
@@
   arrange(Query, YearPeriod, randRank) %>%
-  mutate(GroupRank = row_number())
-
+  mutate(GroupRank = row_number()) %>%
 # Identify which papers qualify to be read
-highlightedPapers <- rankedPapers %>%
   group_by(Query, YearPeriod) %>%
@@
   # Get Review_ID, GroupRank, and ToRead
-  select(Review_ID, Group, OverallRank = randRank, GroupRank, ToRead)
+  select(Review_ID, Title, Group, OverallRank = randRank, GroupRank, ToRead)
 
-head(highlightedPapers) %>%
+head(first_rankedPapers) %>%
   kable()
@@
 ```{r re-add unranked papers}
-unrankedPapers <- paperList %>% 
+first_unrankedPapers <- first_paperList %>% 
   # Drop empty rows, excluded, and incidental papers
@@
   mutate(ToRead = "NO", Group = "Unranked", GroupRank = "NA", OverallRank = "NA") %>%
-  select(Review_ID, Group, OverallRank, GroupRank, ToRead)
+  select(Review_ID, Title, Group, OverallRank, GroupRank, ToRead)
 
 # Combine ranked and unranked papers
-outputList <- rbind(highlightedPapers, unrankedPapers) %>%
+firstSample <- rbind(first_rankedPapers, first_unrankedPapers) %>%
   mutate(Review_ID = as.numeric(Review_ID)) %>%
-  arrange(Review_ID)
+  arrange(Review_ID) %>%
+  mutate(Sample = "First")
 
 # Write to disk
-write_csv(outputList, file = "Randomisation/PrimarySampleRanks.csv")
+write_csv(firstSample, file = "Randomisation/FirstSampleRanks.csv")
 ```
@@
 
+The list of papers pulled from Web of Science is loaded in - 1184 papers are represented here. Year-windows are assigned here as in the **First Sample.**
+
+```{r Load second sample}
+#| warning: false
+# Load query and extract relevant columns
+second_Query <- readxl::read_xls("Queries/MacKenzie WOS/MacKenzie03.xls") %>%
+  select(Review_ID, Title = `Article Title`, Year = `Publication Year`) %>%
+  # Assign year period
+  mutate(YearPeriod = case_when(Year %in% 2000:2005 ~ "2000:2005",
+                                Year %in% 2006:2010 ~ "2006:2010",
+                                Year %in% 2011:2015 ~ "2011:2015",
+                                Year %in% 2016:2021 ~ "2016:2021",
+                                Year %in% 2022:2023 ~ "2022:2023"))
+
+head(second_Query) %>%
+  kable()
+```
+
+Papers are assigned a random number and ranked within their groups, and top-15 papers are flagged to be read. The output is saved for future reference.
+
+```{r randomise second sample}
+secondSample <- second_Query %>%
+  # Add query flag to match with the first sample
+  mutate(Query = "MacKenzie Cited") %>%
+  # Add random ranks for all papers
+  mutate(randRank = sample(x = 1:nrow(.), size = nrow(.))) %>%
+  # Re-rank within strata
+  group_by(Query, YearPeriod) %>%
+  arrange(Query, YearPeriod, randRank) %>%
+  mutate(GroupRank = row_number()) %>% 
+  mutate(ToRead = case_when(
+    GroupRank <= 15 ~ "YES",
+    GroupRank >15 ~ "NO")) %>%
+  unite(col = Group, c(Query, YearPeriod)) %>%
+  mutate(Sample = "Second") %>%
+  # Get Review_ID, GroupRank, and ToRead
+  select(Sample, Review_ID, Title, Group, OverallRank = randRank, GroupRank, ToRead)
+
+write_csv(secondSample, "Randomisation/SecondSampleRanks.csv")
+
+head(secondSample) %>%
+  kable()
+```
+
+## Sample summary
+
+A basic representation of the two sample's coverage is presented here.
+The data presented is from after randomisation, **not** after reviewing, and thus does not represent the final coverage of the review as some papers may be excluded.
+
+### First sample categories
+
+```{r plot first sample categories}
+firstSample %>%
+  separate(Group, c("Query", "YearWindow"), sep = "_") %>%
+
+  ggplot() +
+  geom_bar(aes(x = YearWindow, fill = ToRead)) +
+  scale_fill_manual(values = c("gray70", "dodgerblue2")) +
+  facet_wrap(vars(Query), nrow = 2) +
+  labs(title = "Coverage of the Primary Sample",
+       y = "Number of Papers",
+       x = "Year Window",
+       fill = "Included")
+```
+
+### Second sample categories
+
+```{r plot second sample categories}
+secondSample %>%
+  separate(Group, c("Query", "YearWindow"), sep = "_") %>%
+
+  ggplot() +
+  geom_bar(aes(x = YearWindow, fill = ToRead)) +
+  scale_fill_manual(values = c("gray70", "firebrick3")) +
+  labs(title = "Coverage of the Secondary Sample",
+       y = "Number of Papers",
+       x = "Year Window",
+       fill = "Included")
+```
+
+### Combined sample
+
+Adding the second sample improves coverage across the review period, particularly in the first two year-windows where fewer papers were included in the first sample.
+
+```{r combined sample}
+rbind(firstSample, secondSample) %>%
+  separate(Group, c("Query", "YearWindow"), sep = "_") %>%
+  filter(ToRead == "YES") %>%
+
+  ggplot() +
+  geom_bar(aes(x = YearWindow, fill = Sample)) +
+  scale_fill_manual(values = c("dodgerblue2", "firebrick3")) +
+  labs(title = "Coverage of combined sample",
+       y = "Number of papers",
+       x = "Year-window")
+```
+
 ## References

### 2023-11-27 Analysis/Archive/Code/1_DataPrep.Rmd
minor restructure

--- 2023-01-26 Analysis/Code/1_DataPrep.Rmd
@@
                                  `Spatial predictions?` == "N" ~ FALSE),
-         Obj_Trends = case_when(`Observing trends?` == "Y" ~ TRUE,
-                                `Observing trends?` == "N" ~ FALSE),
          Obj_Hypothesis = case_when(`Hypothesis testing?` == "Y" ~ TRUE,
@@
   select(`Review ID`, Title, Obj_Methods, Obj_Projections, 
-         Obj_Spatial, Obj_Trends, Obj_Hypothesis)
+         Obj_Spatial, Obj_Hypothesis)
 
@@
   # Select only necessary fields
-  select(`Review ID`, Title, Taxa, `Taxa keywords`, Number, 
+  select(`Review ID`, Title, Taxa, `Taxa keywords`, `IUCN Status`, Number, 
          Threat_HabitatLoss, Threat_Exploitation, Threat_ClimateChange,
          Threat_Invasives, Threat_Disease, Threat_Disaster,
-         Country, Continent, Realm, `Study Size`)
+         Country, Continent, Realm, `Study Size`, `Center point`)
 kable(head(taxa_location_sheet))
@@
   filter(Status == "Complete") %>%
-  # Note modifiers for parameters, then split them off from parameters
-  mutate(MultiSpecies = case_when(str_detect(Parameter,"MS") ~ TRUE,
-                                  !str_detect(Parameter, "MS") ~ FALSE),
-         FalsePositive = case_when(str_detect(Parameter, "FP") & str_detect(Parameter, "Detection") ~ TRUE)) %>%
+  # Select parameter prefix
   mutate(Parameter = str_split(Parameter, "_", simplify = TRUE)[,1]) %>%
   # Split out each covariate
-  separate_rows(`Categories`, sep =  ", ") %>%
+  separate_rows(`Covariate List`, sep =  ", ") %>%
+  # Convert to uppercase
+  mutate(`Covariate List` = toupper(`Covariate List`)) %>%
   # Convert 'none' to NA
-  separate(`Categories`, c("Category", "Scale", "Dynanimity"), sep = "_") %>%
-  select(`Review ID`, Title, `Total covariates`, Parameter, Category, Scale, 
-         Dynanimity, `Higher order`, `Interactive terms`,
-         MultiSpecies, FalsePositive)
-
-covariates_sheet$Category[covariates_sheet$Category %in% c("None", "NA")] <- NA
-covariates_sheet$Category[covariates_sheet$Category == c("Topography")] <- "Topographic"
-covariates_sheet$Category[covariates_sheet$Category == c("Hydrology")] <- "Hydrologic"
+  separate(`Covariate List`, c("Category", "Scale", "CovObsType", "Complexity"), sep = "_") %>%
+  separate(`Category`, c("CovCategory", "EnvCategory"), sep = "-") %>%
+  select(`Review ID`, Title, `Total covariates`, Parameter, CovCategory, EnvCategory, Scale, 
+         CovObsType, Complexity)
 
@@
   filter(Status == "Complete") %>%
-  separate_rows(`Selection category`, sep = ", ") %>%
-  separate(`Selection category`, into = c("Select_Category", "Select_Type"), sep = "_") %>%
-  select(`Review ID`, Title, Bayesian, Select_Category, Select_Type)
-
-modelling_sheet$Select_Category[modelling_sheet$Select_Category == "APriori"] <- "Apriori"
+  mutate(`Prefiltering category` = tolower(`Prefiltering category`),
+         `Selection category` = tolower(`Selection category`)) %>%
+  select(`Review ID`, Title, Bayesian, `Prefiltering category`, `Selection category`, `Evaluation method`,
+         `Multi-species`, Connectivity)
 
@@
 ```{r Save output, echo = FALSE}
-
 reviewResults <- list(Study = study_sheet,

### 2023-11-27 Analysis/Archive/Code/Archive/2_PaperFigures.Rmd
minor restructure

---
title: "DOM review: Paper figures"
output: html_document
---

```{r Packages, include = FALSE}
library(tidyverse)
library(readxl)
library(knitr)
library(sf)
library(ggalluvial)
```

# Read-in data
```{r read in data, echo = FALSE}
reviewResults <- read_rds("Analysis/Outputs/ReviewResults.rds")
```

# Map of study sites
```{r map of study sites}
# load world map
continents <- st_read("Analysis/SpatialData/Continents/World_Continents.shp")

sitePoints <- reviewResults$Taxa %>%
  select(Title, `Center point`) %>%
  separate(`Center point`, into = c('Latitude', 'Longitude'), sep = ", ") %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = st_crs(continents))

ggplot() +
  geom_sf(data = continents, aes(fill = CONTINENT), show.legend = FALSE) +
  geom_sf(data = sitePoints, colour = "black", shape = 4, size = 4) +
  theme_void()

```

```{r alluvial plot}
reviewResults$Modelling %>% 
  filter(Bayesian %in% c("Y", "N")) %>%
  mutate(Bayesian = case_when(Bayesian == "Y" ~ "Bayesian",
                              .default = "Frequentist")) %>%
  group_by(`Prefiltering category`, `Selection category`, Bayesian) %>%
  summarise(n = n()) %>%

  ggplot(aes(y = n, axis1 = `Prefiltering category`, axis2 = `Selection category`)) +
  geom_alluvium(aes(fill = `Selection category`)) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 7) +
  facet_grid(rows = vars(Bayesian), scales = "free_y") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 20),
        strip.text = element_text(size = 20))

```

```{r}

vec <-  c("test_a20", "test2_n15", "testing3_p3")

str_replace(string = vec, pattern = "(?<=\\_)[:alpha:]{1}", replacement = "s")

```

```{r stacked bar covs}
simpleCovs <- reviewResults$Covariates %>%
  select(Title, Parameter, CovCategory, EnvCategory, Scale, CovObsType) %>%
  left_join(select(reviewResults$Study, Title, Year), by = "Title")

simpleCovs %>%
  mutate(Parameter = case_when(Parameter == "Persistence" ~ "Extinction",
                               .default = Parameter)) %>%
  filter(Parameter %in% c("InitialOccupancy", "Colonisation", "Extinction", "Detection")) %>%
  filter(CovCategory != "NONE") %>%
  filter(!CovCategory %in% c('UNCLEAR', 'UNSPECIFIED')) %>%
  group_by(Parameter, CovCategory) %>% 
  summarise(avgcount = n()/47) %>%

  ggplot() +
  geom_col(aes(x = Parameter, y = avgcount, fill = CovCategory)) +
  scale_fill_discrete("Covariate Category",
                      limits = c("E", "CONN", "INT", "SEAS", "SPEC", "SURV", "OTHER"),
                     labels = c("Environmental", "Connectivity", "Interactions", "Season effect",
                               "Species effect", "Survey details", "Other")) +
  scale_x_discrete(limits = c("InitialOccupancy", "Colonisation", "Extinction", "Detection")) +
  labs(title = "Average number of covariates per parameter", y = "Average number of covariates", x = "Parameter")

# Total number of covariates by year
simpleCovs %>%
  filter(Parameter %in% c("InitialOccupancy", "Colonisation", "Extinction", "Detection")) %>%
  filter(CovCategory != "NONE") %>%
  filter(!is.na(CovCategory)) %>%
  filter(!CovCategory %in% c('UNCLEAR', 'UNSPECIFIED')) %>%
  filter(!is.na(Year)) %>% 
  group_by(Year) %>%
  mutate(npapers = length(unique(Title))) %>%
  ungroup() %>%
  group_by(CovCategory, Year, npapers) %>%
  summarise(count = n()) %>%
  mutate(avgCount = count/npapers) %>%

  ggplot() + 
  geom_col(aes(x = Year, y = avgCount, fill = CovCategory)) +
  scale_fill_discrete("Covariate Category",
                      limits = c("E", "CONN", "INT", "SEAS", "SPEC", "SURV", "OTHER"),
                    labels = c("Environmental", "Connectivity", "Interactions", "Season effect",
                               "Species effect", "Survey details", "Other")) +
  labs(title = "Average total number of covariates by year", 
       y = "Average number of covariates")
```

```{r covs by objective}
reviewResults$Objectives %>%
  select(Title, Obj_Methods, Obj_Projections, Obj_Spatial, Obj_Hypothesis) %>%
  pivot_longer(cols = c(Obj_Methods, Obj_Spatial, Obj_Hypothesis, Obj_Projections),
               names_to = "Objective", values_to = "ObjIncluded") %>%
  filter(ObjIncluded == TRUE) %>%
  mutate(Objective = case_when(Objective == "Obj_Methods" ~ "Methods",
                               Objective == "Obj_Spatial" ~ "Spatial Prediction",
                               Objective == "Obj_Projections" ~ "Future Prediction",
                               Objective == "Obj_Hypothesis" ~ "Hypothesis Testing")) %>%

  right_join(simpleCovs, by = "Title") %>%
  filter(Parameter %in% c("InitialOccupancy", "Colonisation", "Extinction", "Detection")) %>%
  filter(CovCategory != "NONE") %>%
  filter(!is.na(CovCategory)) %>%
  filter(!CovCategory %in% c('UNCLEAR', 'UNSPECIFIED')) %>%
  group_by(Objective) %>%
  mutate(npapers = length(unique(Title))) %>%
  ungroup() %>%
  group_by(CovCategory, Objective, npapers) %>%
  summarise(count = n()) %>%
  mutate(avgCount = count/npapers) %>%  
  filter(!is.na(Objective)) %>%

  ggplot() + 
  geom_col(aes(x = Objective, y = avgCount, fill = CovCategory)) +
  scale_fill_discrete("Covariate Category",
                      limits = c("E", "CONN", "INT", "SEAS", "SPEC", "SURV", "OTHER"),
                    labels = c("Environmental", "Connectivity", "Interactions", "Season effect",
                               "Species effect", "Survey details", "Other")) +
  labs(title = "Average total number of covariates by objective", 
       y = "Average number of covariates")

```

```{r stat tables}
reviewResults$Study %>%
  group_by(`First author country`) %>%
  summarise(n = n())

reviewResults$Survey %>%
  group_by(`Capture method`) %>%
  summarise(n = n())

reviewResults$Covariates %>%
  filter(Parameter %in% c("InitialOccupancy", "Colonisation", "Extinction", "Detection")) %>%
  filter(CovCategory != "NONE") %>%
  filter(!is.na(CovCategory)) %>%
  filter(!CovCategory %in% c('UNCLEAR', 'UNSPECIFIED')) %>%
  group_by(CovCategory, EnvCategory, Title) %>%
  summarise() %>%
  group_by(CovCategory, EnvCategory) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  mutate(percent = n/47)

```

### 2023-11-27 Queries/QueryComparison.Rmd

---
title: "Web of Science and Google Scholar Comparison"
date: "September 14 2023"
output: html_document
---

```{r Setup, include = FALSE}
library(tidyverse)
library(readxl)
papers <- read_excel("Queries/Query Comparison.xlsx")
```

Find overlap
```{r, Find overlaps}
overlaps <- papers %>%
  mutate(Overlap = Query) %>%
  group_by(Title, Query, Year) %>%
  mutate(Overlap = if_else(n()>1, TRUE, FALSE)) %>%
  mutate(Platform = case_when(Overlap == TRUE ~ "Both",
                              .default = Platform)) %>%
  group_by(Title, Query, Year, Platform) %>%
  summarise()
```

```{r overlaps}
ggplot(overlaps) +
  geom_bar(aes(x = Query, fill = Platform))

ggplot(overlaps) +
  geom_bar(aes(x = Year, fill = Platform))

ggplot(overlaps) +
  geom_bar(aes(x = Year, fill = Platform)) +
  facet_wrap(vars(Query))

```

### 2024-03-08 Randomisation/Archive/old_Randomise_Papers.Rmd
turkey paper

--- 2022-08-30 Writing/Randomisation/Randomise_Papers.Rmd
@@
 
-***
+------------------------------------------------------------------------
 
 ## Summary
+
 This document contains code for randomising paper ranks for the review on Dynamic Occupancy Models.
@@
 
-***
+------------------------------------------------------------------------
 
@@
 Read in the excel sheets and process out the header
+
 ```{r Read-in excel sheets, echo = FALSE, message = FALSE, warning = FALSE}
@@
 Combine the two sheets, the former containing metadata and the latter containing information on the papers
+
 ```{r Combine and merge, echo = FALSE, message = FALSE}
@@
 Assign each paper a random overll rank, then order the papers within their strata by the overall rank
+
 ```{r Randomly rank papers within strata}
@@
 Highlight which papers are ranked highly enough to qualify for evaluation
+
 ```{r Highlight papers to read}
@@
 Add previously unranked papers, remove extraneous columns, and save both this workspace and the rankings to disk.
+
 ```{r Reformat and save}

### 2024-03-14 Analysis/ReviewSummary/ReviewSummary.qmd
started outlining sections and data availability

---
title: "DOM Review Summary"
author: "Saoirse Kelleher"
date: "March 13 2024"
date-modifed: today
abstract: "A summary of results from the DOM Review, for use in interpreting outputs and writing the review manuscript. "
abstract-title: "Summary"
format: 
  html:
    theme: minty
    toc: true
---

```{r prepare inputs}
#| include: false

library(tidyverse)
library(readxl)

# Study Details
review_StudyDetails <- read_xlsx("Spreadsheets/Stage3_Spreadsheet.xlsx",
                                 sheet = "Study") |>
  select(ID = "Review ID", Sample, Status, Title, Journal, Year, 
         FirstAuthor = "First author country", 
         AuthorCategories = "Authorship groups")

# Objectives and Inference
review_Objectives <- read_xlsx("Spreadsheets/Stage3_Spreadsheet.xlsx",
                               sheet = "Objectives and Inference") |>
  select(ID = "Review ID", Status, Title, 
         Obj_Methods = "Methods?",
         Obj_Future = "Future predictions?",
         Obj_Spatial = "Spatial predictions?", 
         Obj_Hypothesis = "Hypothesis testing?")

# Species and Location
review_SpeciesLocation <- read_xlsx("Spreadsheets/Stage3_Spreadsheet.xlsx",
                                    sheet = "Species and Location") |>
  select(ID = "Review ID", Status, Title, 
         Taxa_Category = "Taxa keywords", Taxa_Number = "Number",
         IUCN_Status = "IUCN Status", Threat_Category = "Threat categories",
         Country, Continent, Realm, Size_km = "Study Size")

# Data Collection
review_DataCollection <- read_xlsx("Spreadsheets/Stage3_Spreadsheet.xlsx",
                                   sheet = "Data collection") |>
  select(ID = "Review ID", Status, Title,
         Duration = "Study duration", Detection_Method = "Capture method",
         Primary_Occasions = "Number of primary occasions",
         Secondary_Occasions = "Number of secondary occasions",
         Secondary_Type = "Type of secondary occasions",
         Site_Number = "Number of sites", Site_Type = "Type of site",
         Sampling_Method = "Sampling method")

# Covariates
review_Covariates <- read_xlsx("Spreadsheets/Stage3_Spreadsheet.xlsx",
                               sheet = "Covariates") |>
  select(ID = "Review ID", Status, Title, 
         Parameter, Covariate_Count = "Total covariates",
         Covariates = "Covariate List")

# Modelling
review_Modelling <- read_xlsx("Spreadsheets/Stage3_Spreadsheet.xlsx",
                              sheet = "Modelling") |>
  select(ID = "Review ID", Status, Title,
         Bayesian, Prefiltering = "Prefiltering category",
         Selection = "Selection category", Evaluation = "Evaluation method",
         Multispecies = "Multi-species", Connectivity)

```

# Study details

Basic metadata was collected for each paper in the review, with available data including:

| Field             | Description                                      | Possible values                      |
|--------------|------------------------------------|----------------------|
| Journal           | Journal which this article was published in      | *Any*                                |
| Year              | Year of publication                              | *Any*                                |
| Author country    | Nation of the first author's primary affiliation | *ISO 3166 country names*             |
| Author categories | Categories of affiliations for *all* authors     | *Academic, Government, NGO, Private* |

## Publication journals

## Authorship countries

## Authorship categories

## Authorship category trends

## Review coverage

# Objectives and inference

The stated objectives for each article were assessed and divided into four categories - each paper could fall into multiple (or zero) objective categories.

| Objective           | Description                                                                       |
|---------------------|---------------------------------------------------|
| Methods             | Did authors seek to demonstrate or test specific methods for occupancy modelling? |
| Future Predictions  | Did authors make predictions beyond the study period?                             |
| Spatial Predictions | Did authors make predictions at locations where surveys were not conducted?       |
| Hypothesis testing  | Did authors seek to test specific hypotheses of species occupancy dynamics?       |

## Frequency of objectives

## Stated objectives by authorship categories

## Stated objective by year

# Species and location

Data on focal species and the location of study areas was collected, available fields include:

| Field           | Description                                                                                | Possible values                                                                   |
|----------------|----------------------------|----------------------------|
| Taxa category   | Category of study taxa; could fit into multiple categories                                 | *Bird, Mammal, Invertebrate, Herptile, Fish, Threatened, Invasive, Other*         |
| Taxa number     | Number of individual species within the taxa for which models were fit                     | *Numeric*                                                                         |
| IUCN status     | IUCN status for the study species; NA if multi-species or if the study species is invasive | *LC, NT, EN, CR, DD, NA*                                                          |
| Threat category | Category of threats to the species; multiple per species permitted                         | *Habitat loss, Climate change, Exploitation, Invasives, Disease, Disaster, Other* |
| Country         | Country/countries where the study took place                                               | *ISO 3166 country names*                                                          |
| Continent       | Continent(s) where the study took place                                                    | *North America, South America, Europe, Asia, Africa, Oceania*                     |
|                 |                                                                                            |                                                                                   |
|                 |                                                                                            |                                                                                   |
|                 |                                                                                            |                                                                                   |
|                 |                                                                                            |                                                                                   |
|                 |                                                                                            |                                                                                   |

# Data collection

##

# Covariates

# Modelling

### 2024-03-14 Analysis/ReviewSummary/ReviewSummary.qmd
Wrote outline
Generated outline for each section of the review summary, with details on available fields and required plots

@@
     toc: true
+embed-resources: true
 ---
@@
 
++-------------------+--------------------------------------------------+--------------------------------------+
 | Field             | Description                                      | Possible values                      |
-|--------------|------------------------------------|----------------------|
++===================+==================================================+======================================+
 | Journal           | Journal which this article was published in      | *Any*                                |
++-------------------+--------------------------------------------------+--------------------------------------+
 | Year              | Year of publication                              | *Any*                                |
++-------------------+--------------------------------------------------+--------------------------------------+
 | Author country    | Nation of the first author's primary affiliation | *ISO 3166 country names*             |
++-------------------+--------------------------------------------------+--------------------------------------+
 | Author categories | Categories of affiliations for *all* authors     | *Academic, Government, NGO, Private* |
++-------------------+--------------------------------------------------+--------------------------------------+
 
@@
 
++---------------------+-----------------------------------------------------------------------------------+
 | Objective           | Description                                                                       |
-|---------------------|---------------------------------------------------|
++=====================+===================================================================================+
 | Methods             | Did authors seek to demonstrate or test specific methods for occupancy modelling? |
++---------------------+-----------------------------------------------------------------------------------+
 | Future Predictions  | Did authors make predictions beyond the study period?                             |
++---------------------+-----------------------------------------------------------------------------------+
 | Spatial Predictions | Did authors make predictions at locations where surveys were not conducted?       |
++---------------------+-----------------------------------------------------------------------------------+
 | Hypothesis testing  | Did authors seek to test specific hypotheses of species occupancy dynamics?       |
++---------------------+-----------------------------------------------------------------------------------+
 
@@
 
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Field           | Description                                                                                | Possible values                                                                   |
-|----------------|----------------------------|----------------------------|
++=================+============================================================================================+===================================================================================+
 | Taxa category   | Category of study taxa; could fit into multiple categories                                 | *Bird, Mammal, Invertebrate, Herptile, Fish, Threatened, Invasive, Other*         |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Taxa number     | Number of individual species within the taxa for which models were fit                     | *Numeric*                                                                         |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | IUCN status     | IUCN status for the study species; NA if multi-species or if the study species is invasive | *LC, NT, EN, CR, DD, NA*                                                          |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Threat category | Category of threats to the species; multiple per species permitted                         | *Habitat loss, Climate change, Exploitation, Invasives, Disease, Disaster, Other* |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Country         | Country/countries where the study took place                                               | *ISO 3166 country names*                                                          |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Continent       | Continent(s) where the study took place                                                    | *North America, South America, Europe, Asia, Africa, Oceania*                     |
-|                 |                                                                                            |                                                                                   |
-|                 |                                                                                            |                                                                                   |
-|                 |                                                                                            |                                                                                   |
-|                 |                                                                                            |                                                                                   |
-|                 |                                                                                            |                                                                                   |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
+| Realm           | Biogeographic realm where the study took place                                             | *Nearctic, Palearctic, Afrotropical, Indomalayan, Australasian, Neotropical*      |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
+| Size            | Size of the study area extent in square kilometres                                         | *\<1km, 1-10, 10-100, 100-1000, 10000-100000, 100000-1000000, 1000000+*           |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
+
+## Taxa categories
+
+## Number of species
+
+## Threatened taxa
+
+## Study locations
 
@@
 
-##
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
+| Field               | Description                                                              | Possible values                                                                              |
++=====================+==========================================================================+==============================================================================================+
+| Duration            | Length of study in months                                                | *Numeric*                                                                                    |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
+| Capture method      | Method of recording detections; multiple categories possible             | *Audiovisual, Search, Trapping, Bioacoustic, Camera, Citizen, Incidental*                    |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
+| Primary occasions   | Number of primary occasions                                              | *Numeric*                                                                                    |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
+| Secondary occasions | Number of secondary occasions, presented as a range where not consistent | *Numeric range*                                                                              |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
+| Secondary type      | Type of secondary occasion                                               | *Revisits, Spatial replicates, Same-visit replicates, multiple observers, time to detection* |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
+| Type of site        | How a 'site' was defined                                                 | *Point location, Grid cell, Irregular area, Transect*                                        |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
+| Number of sites     | Number of individual sites                                               | Numeric                                                                                      |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
+| Sampling method     | How sites were selected                                                  | *Random, Stratified, Grid, Preferential, Available, Exhaustive, Uniform, Transect*           |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
+
+## Study duration
+
+## Number of occasions
+
+## Data collection method
+
+## Sampling method
+
+## Study size over time
 
@@
 
+All covariates considered for modelling were documented for each article. Details available for each covariate include:
+
++-------------+-------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------+
+| Field       | Description                                     | Possible values                                                                                                                                                     |
++=============+=================================================+=====================================================================================================================================================================+
+| Parameter   | Parameter which this covariate was considered   | *Initial occupancy, colonisation, extinction, detection*                                                                                                            |
++-------------+-------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------+
+| Type        | Category of covariate                           | *Environmental: Habitat type terrestrial, Habitat type aquatic, Anthropogenic, Topographic, Climate and weather, Disaster and disturbance, Hydrologic, Seasonality* |
+|             |                                                 |                                                                                                                                                                     |
+|             |                                                 | *Other: Interactions, Site geometry, Connectivity, Site-level effects, Species-level effects, Season-level effects, Survey details*                                 |
++-------------+-------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------+
+| Variation   | Level of variation for the covariate            | *Site, Season, Survey*                                                                                                                                              |
++-------------+-------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------+
+| Measurement | How the covariate was recorded                  | *Remotely, Observed, Intrinsic*                                                                                                                                     |
++-------------+-------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------+
+| Complexity  | How the covariate was mathmatically implemented | *Linear-only, Quadratics, Interactions, Interactions and Quadratics, Other*                                                                                         |
++-------------+-------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------+
+|             |                                                 |                                                                                                                                                                     |
++-------------+-------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------+
+
+## Covariate quantity
+
+## Covariate categories
+
+## Covariate features
+
+## Frequency of covariate features
+
+## Quantity of covariates by year
+
+## Quantity of covariates by study size
+
 # Modelling
+
+Key details of the modelling process were documented.
+
++---------------+-----------------------------------------------------------+-------------------------------------------------------------------------------------------------+
+| Field         | Description                                               | Possible values                                                                                 |
++===============+===========================================================+=================================================================================================+
+| Bayesian      | Whether the models fit were Bayesian                      | *Logical*                                                                                       |
++---------------+-----------------------------------------------------------+-------------------------------------------------------------------------------------------------+
+| Prefiltering  | Which, if any, prefiltering method was used               | *Expert, Simple, Other*                                                                         |
++---------------+-----------------------------------------------------------+-------------------------------------------------------------------------------------------------+
+| Selection     | Which, if any, selection method was used                  | *Single a priori, Multiple a priori, Exhaustive, Exhaustive by parameter, Forwards, Backwards*  |
++---------------+-----------------------------------------------------------+-------------------------------------------------------------------------------------------------+
+| Evaluation    | How models were compared and evaluated                    | *AIC, AUC, DiC, Model-averaging, Bootstrapping*                                                 |
++---------------+-----------------------------------------------------------+-------------------------------------------------------------------------------------------------+
+| Multi-species | Whether this was a multi-species model                    | *Logical*                                                                                       |
++---------------+-----------------------------------------------------------+-------------------------------------------------------------------------------------------------+
+| Connectivity  | Whether site connectivity was incorporated into the model | *Logical*                                                                                       |
++---------------+-----------------------------------------------------------+-------------------------------------------------------------------------------------------------+

### 2024-03-14 Analysis/ReviewSummary/ReviewSummary.qmd
Clean Study Details
Filtered down study details; started journals table

@@
 
-```{r prepare inputs}
+```{r Load_packages}
 #| include: false
@@
 library(readxl)
+library(knitr)
+```
+
+```{r Load_data}
+#| include: false
 
@@
 
+```{r StudyDetails_Clean}
+#| include: false
+
+studyDetails_clean <- review_StudyDetails |>
+  filter(Status == "Complete")
+```
+
 ## Publication journals
+
+```{r StudyDetails_journals}
+#| echo: false
+
+studyDetails_clean |>
+  summarise(Articles = n(), .by = "Journal") |>
+  mutate(Journal = case_when(Articles == 1 ~ "Other",
+                             .default = Journal)) |>
+  summarise(Articles = sum(Articles), .by = "Journal") |> 
+  arrange(-Articles) |>
+  kable()
+```
 
@@
 
-+---------------+-----------------------------------------------------------+-------------------------------------------------------------------------------------------------+
++---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Field         | Description                                               | Possible values                                                                                |
-+===============+===========================================================+=================================================================================================+
++===============+===========================================================+================================================================================================+
 | Bayesian      | Whether the models fit were Bayesian                      | *Logical*                                                                                      |
-+---------------+-----------------------------------------------------------+-------------------------------------------------------------------------------------------------+
++---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Prefiltering  | Which, if any, prefiltering method was used               | *Expert, Simple, Other*                                                                        |
-+---------------+-----------------------------------------------------------+-------------------------------------------------------------------------------------------------+
++---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Selection     | Which, if any, selection method was used                  | *Single a priori, Multiple a priori, Exhaustive, Exhaustive by parameter, Forwards, Backwards* |
-+---------------+-----------------------------------------------------------+-------------------------------------------------------------------------------------------------+
++---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Evaluation    | How models were compared and evaluated                    | *AIC, AUC, DiC, Model-averaging, Bootstrapping*                                                |
-+---------------+-----------------------------------------------------------+-------------------------------------------------------------------------------------------------+
++---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Multi-species | Whether this was a multi-species model                    | *Logical*                                                                                      |
-+---------------+-----------------------------------------------------------+-------------------------------------------------------------------------------------------------+
++---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Connectivity  | Whether site connectivity was incorporated into the model | *Logical*                                                                                      |
-+---------------+-----------------------------------------------------------+-------------------------------------------------------------------------------------------------+
++---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+

### 2024-03-14 Analysis/ReviewSummary/ReviewSummary.qmd

@@
 library(knitr)
+library(reactable)
 ```
@@
 studyDetails_clean <- review_StudyDetails |>
-  filter(Status == "Complete")
+  filter(Status == "Complete")  |>
+  mutate(YearStrata = case_when(Year %in% 2000:2005 ~ "2000-2005",
+                                Year %in% 2006:2010 ~ "2006-2010",
+                                Year %in% 2011:2015 ~ "2011-2015",
+                                Year %in% 2016:2020 ~ "2016-2020",
+                                Year %in% 2021:2023 ~ "2021-2023"))
+```
+
+## Year coverage
+
+```{r StudyDetails_years}
+studyDetails_clean |>
+  filter(!is.na(YearStrata)) |>
+  mutate(YearStrata = fct(YearStrata,
+                          levels = c("2000-2005", "2006-2010",
+                                     "2011-2015", "2016-2020",
+                                     "2021-2023"))) |>
+  summarise(Articles = n(), .by = YearStrata) |>
+
+  ggplot() +
+  geom_col(aes(x = YearStrata, y = Articles), fill = "darkslategrey") +
+  labs(title = "Number of articles by year strata",
+       x = "Year Strata", y = "Number of articles")
 ```
@@
 
-studyDetails_clean |>
+journalsTable <- studyDetails_clean |>
   summarise(Articles = n(), .by = "Journal") |>
-  mutate(Journal = case_when(Articles == 1 ~ "Other",
-                             .default = Journal)) |>
-  summarise(Articles = sum(Articles), .by = "Journal") |> 
-  arrange(-Articles) |>
-  kable()
+  arrange(-Articles, Journal) 
+```
+
+`{r} nrow(journalsTable)` unique journals are represented in the sample - `{r} nrow(filter(journalsTable, Articles >1))` journals have more than one article.
+
+```{r StudyDetails_journalsTable}
+#| echo: false
+reactable(journalsTable, defaultPageSize = 8)
 ```
@@
 
+```{r StudyDetails_countries}
+#| echo: false
+
+countriesTable <- studyDetails_clean |>
+  summarise(Articles = n(), .by = "FirstAuthor") |>
+  arrange(-Articles)
+```
+
+Counting by the location of the first author's institution, `{r} nrow(countriesTable` countries are represented in the sample.
+A strong majority of articles were published by first authors based in the United States of America.
+
+```{r StudyDetails_countriesTable}
+#| echo: false
+
+reactable(countriesTable, defaultPageSize = 8)
+```
+
 ## Authorship categories
 
-## Authorship category trends
+```{r StudyDetails_Categories}
+#| echo: false
 
-## Review coverage
+authorCategories_overall <- studyDetails_clean |>
+  separate_rows(AuthorCategories, sep = ", ") |> 
+  summarise(Articles = n(),
+            Proportion = n()/nrow(studyDetails_clean),
+            .by = AuthorCategories) |>
+  mutate(Proportion = round(Proportion, 2)) |>
+  rename(Category = AuthorCategories) |>
+  arrange(-Proportion)
+```
+
+The most frequent category of authorship was academic, with `{r} max(authorCategories_overall)`% of articles having at least one author from an academic institution.
+
+```{r StudyDetails_CategoriesTable}
+#| echo: false
+
+kable(authorCategories_overall)
+```
+
+There is limited discernible change in authorship categories over time.
+
+```{r StudyDetails_authorshipTrend}
+#| echo: false
+#| warning: false
+
+studyDetails_clean |>
+  filter(!is.na(YearStrata)) |>
+  mutate(StrataArticles = n(), .by = YearStrata) |>
+  separate_rows(AuthorCategories, sep = ", ") |> 
+  summarise(Articles = n(),
+            .by = c(AuthorCategories, YearStrata, StrataArticles)) |>
+  mutate(Proportion = Articles/StrataArticles) |>
+  mutate(YearStrata = fct(YearStrata,
+                          levels = c("2000-2005", "2006-2010",
+                                     "2011-2015", "2016-2020",
+                                     "2021-2023")),
+         AuthorCategories = fct(AuthorCategories,
+                                levels = c("Academic", "Government",
+                                           "NGO", "Private"))) |>
+  complete(AuthorCategories, YearStrata) |>
+
+  ggplot(aes(x = YearStrata, 
+                 y = Proportion, 
+                 fill = AuthorCategories)) +
+  geom_col(position = "dodge") +
+  labs(title = "Authorship trends over time",
+       x = "Year of publication",
+       y = "Proportion of Articles") +
+  scale_fill_manual("Authorship category",
+                    values = c("darkslategrey", "darkorchid3",
+                               "cyan3", "goldenrod1"))
+```
+
+##
 

### 2024-03-15 Analysis/ReviewSummary/ReviewSummary.qmd
clean objectives
Cleaned objective data, added year strata

@@
 ```{r StudyDetails_years}
+#| echo: false
+
 studyDetails_clean |>
@@
 
-Counting by the location of the first author's institution, `{r} nrow(countriesTable` countries are represented in the sample.
+Counting by the location of the first author's institution, `{r} nrow(countriesTable)` countries are represented in the sample.
 A strong majority of articles were published by first authors based in the United States of America.
@@
 
-The most frequent category of authorship was academic, with `{r} max(authorCategories_overall)`% of articles having at least one author from an academic institution.
+The most frequent category of authorship was academic, with `{r} max(authorCategories_overall$Proportion)` % of articles having at least one author from an academic institution.
 
@@
 +---------------------+-----------------------------------------------------------------------------------+
+
+```{r Objectives_clean}
+#| echo: false
+
+review_Objectives |>
+  filter(Status == "Complete") |> 
+  select(ID, Obj_Methods, Obj_Future, Obj_Spatial, Obj_Hypothesis) |>
+  pivot_longer(cols = c(Obj_Methods, Obj_Future, 
+                        Obj_Spatial, Obj_Hypothesis), 
+               names_to = "Objective", 
+               values_to = "Stated") |>
+  mutate(Objective = str_extract(Objective, "(?<=Obj\\_).+"),
+         Stated = case_when(Stated == "Y" ~ TRUE,
+                            Stated == "N" ~ FALSE)) |>
+  left_join(select(studyDetails_clean, ID, Year, YearStrata), by = "ID")
+
+```
 

### 2024-03-15 Analysis/ReviewSummary/ReviewSummary.qmd

@@
 
-review_Objectives |>
+objectives_Clean <- review_Objectives |>
   filter(Status == "Complete") |> 
   select(ID, Obj_Methods, Obj_Future, Obj_Spatial, Obj_Hypothesis) |>
+  rowwise() |>
+  mutate(Obj_None = case_when(all(c(Obj_Methods, Obj_Future,
+                                    Obj_Spatial, Obj_Hypothesis) == "N") ~ "Y",
+                              any(c(Obj_Methods, Obj_Future,
+                                    Obj_Spatial, Obj_Hypothesis) == "Y") ~ "N")) |>
   pivot_longer(cols = c(Obj_Methods, Obj_Future, 
-                        Obj_Spatial, Obj_Hypothesis), 
+                        Obj_Spatial, Obj_Hypothesis, Obj_None), 
                names_to = "Objective", 
@@
                             Stated == "N" ~ FALSE)) |>
-  left_join(select(studyDetails_clean, ID, Year, YearStrata), by = "ID")
-
+  left_join(select(studyDetails_clean, ID, 
+                   Year, YearStrata), 
+            by = "ID")
 ```
@@
 
+```{r Objectives_frequency}
+#| echo: false
+
+objectives_frequency <- objectives_Clean |>
+  mutate(Total = length(unique(ID))) |>
+  filter(Stated == TRUE) |>
+  summarise(Articles = n(),
+            .by = c(Objective, Total)) |>
+  mutate(Proportion = Articles/Total) |> 
+  arrange(-Proportion)
+```
+
+The most common objective was **Testing Hypotheses**, with `{r} max(objectives_frequency$Proportion)`% of articles examining pre-determined relationships.
+Prediction, whether spatial or temporal, was rare.
+
+```{r Objectives_frequencyPlot}
+#| echo: false
+
+objectives_frequency |>
+  mutate(Objective = fct(Objective,
+                         levels = c("Hypothesis", "Methods",
+                                    "Future", "Spatial",
+                                    "None"))) |>
+
+  ggplot() +
+  geom_col(aes(x = Objective, y = Articles, fill = Objective)) +
+  scale_fill_manual(values = c("darkorchid", "deepskyblue3",
+                               "coral2", "goldenrod", "gray60")) +
+  labs(title = "Number of articles by objective",
+       subtitle = "'None' indicates that none of the objectives were stated,\ngenerally this indicates a focus on occupancy dynamics parameters",
+       x = "Stated objective", y = "Number of articles") +
+  theme(legend.position = "none")
+
+```
+
 ## Stated objectives by authorship categories
 
+Little variation was shown between the objectives of papers by different authorship categories.
+
+```{r Objective_frequencyAuthorship}
+#| echo: false
+
+objectives_frequencyAuthorship <- objectives_Clean |>
+  left_join(select(studyDetails_clean, ID, AuthorCategories), by = "ID") |>
+  separate_rows(AuthorCategories, sep = ", ") |>
+  filter(Stated == TRUE) |>
+  mutate(Total = length(unique(ID)), .by = AuthorCategories) |>
+  summarise(Articles = n(),
+            .by = c(Objective, AuthorCategories, Total)) |>
+  mutate(Proportion = Articles/Total) |>
+  complete(Objective, AuthorCategories, 
+           fill = list(Articles = 0,
+                       Proportion = 0))
+
+# Plot frequency faceted by authorship
+objectives_frequencyAuthorship |>
+  mutate(Objective = fct(Objective,
+                         levels = c("Hypothesis", "Methods",
+                                    "Future", "Spatial",
+                                    "None"))) |>
+
+  ggplot() +
+  geom_col(aes(x = Objective, y = Proportion, fill = Objective)) +
+  scale_fill_manual(values = c("darkorchid", "deepskyblue3",
+                               "coral2", "goldenrod", "gray60")) +
+  facet_wrap(facets = vars(AuthorCategories)) +
+  labs(title = "Proportion of articles with each objective",
+       subtitle = "'None' indicates that none of the listed objectives were stated,\ngenerally this indicates a focus on simply estimating occupancy dynamics",
+       x = "Stated objective", y = "Number of articles") +
+  theme(legend.position = "bottom")
+```
+
 ## Stated objective by year
+
+Objective categories have changed in frequency over time, with more variation in objectives in recent years relative to the early days of DOMs
+
+```{r Objectives_byYear}
+#| echo: false
+
+objectives_Clean |>
+  filter(!is.na(YearStrata)) |>
+  filter(Stated == TRUE) |>
+  mutate(Total = length(unique(ID)), .by = YearStrata) |>
+  summarise(Articles = n(), .by = c(Objective, YearStrata, Total)) |>
+  mutate(Proportion = Articles/Total) |>
+  complete(Objective, YearStrata, fill = list(Articles = 0, Proportion = 0)) |>
+  mutate(Objective = fct(Objective,
+                         levels = c("Hypothesis", "Methods",
+                                    "Future", "Spatial",
+                                    "None"))) |>
+
+  ggplot() +
+  geom_col(aes(x = YearStrata, y = Proportion, fill = Objective),
+           position = "dodge") +
+  scale_fill_manual(values = c("darkorchid", "deepskyblue3",
+                               "coral2", "goldenrod", "gray60")) +
+  labs(title = "Objective categories over time",
+       subtitle = "'None' indicates that none of the listed objectives were stated,\ngenerally this indicates a focus on simply estimating occupancy dynamics",
+       y = "Proportion of articles", x = "Year Strata")
+
+```
 

### 2024-03-15 Analysis/ReviewSummary/ReviewSummary.qmd
Taxa categories/counts
Added table for taxa category and plot for taxa counts

@@
 
+```{r Species_clean}
+species_clean <- review_SpeciesLocation |>
+  filter(Status == "Complete") |>
+  mutate(ID_clean = str_extract(ID, "[:digit:]+"))
+```
+
 Data on focal species and the location of study areas was collected, available fields include:
@@
 
+```{r Species_Taxa}
+#| echo: false
+
+species_taxaTable <-species_clean |>
+  separate_rows(Taxa_Category, sep = ", ") |>
+  mutate(Total = length(unique(ID_clean))) |>
+  select(Taxa_Category, ID_clean, Total) |>
+  distinct() |>
+  summarise(Articles = n(), 
+            Proportion = round(Articles/mean(Total), 2),
+            .by = Taxa_Category) |>
+  arrange(-Articles) |>
+  rename(Taxa = Taxa_Category)
+```
+
+DOMs have been widely used for terrestrial taxa, with numerous examples modelling bird, mammal, and herptile occupancy dynamics.
+`{r} species_taxaTable$Articles[species_taxaTable$Taxa == "Threatened"]` articles described their focal species as threatened in some way; `{r} species_taxaTable$Articles[species_taxaTable$Taxa == "Invasive"]` fit models to invasive species.
+
+```{r Species_TaxaTable}
+#| echo: false
+
+kable(species_taxaTable)
+```
+
 ## Number of species
+
+```{r Species_Number}
+#| echo: false
+species_Number <- species_clean |>
+  select(ID_clean, Taxa_Number) |>
+  mutate(Taxa_Number = sum(Taxa_Number), .by = ID_clean)
+```
+
+Only approximately half of articles modelled occupancy dynamics for a single species.
+Others fit multiple occupancy models: either multi-species models where 2+ species were modelled in concert, or many independent models for a suite of species.
+The largest number of species modelled in a single paper was **`{r} max(species_Number$Taxa_Number)`**.
+
+```{r Species_NumberPlot}
+#| echo: false
+
+species_Number |>
+  mutate(NumberGroup = case_when(Taxa_Number == 1 ~ "1",
+                                 Taxa_Number %in% 2:10 ~ "2-10",
+                                 Taxa_Number %in% 11:100 ~ "11-100",
+                                 Taxa_Number > 100 ~ ">100")) |>
+  summarise(Articles = n(), .by = NumberGroup) |>
+  mutate(NumberGroup = fct(NumberGroup,
+                           levels = c("1", "2-10", "11-100", ">100"))) |>
+
+  ggplot() +
+  geom_col(aes(x = NumberGroup, y = Articles), fill = "darkslategrey") +
+  labs(title = "Number of species modelled per article",
+       x = "Number of species", y = "Number of articles") 
+```
 

### 2024-03-15 Analysis/ReviewSummary/ReviewSummary.qmd
Threatened taxa
Details on threats to taxa and IUCN status of taxa

@@
 kable(species_taxaTable)
+species_taxaTable |>
+  mutate(Taxa = fct_reorder(Taxa, Articles)) |>
+
+  ggplot() +
+  geom_col(aes(x = Articles, y = Taxa, fill = Taxa)) +
+  scale_fill_manual(limits = c("Bird", "Threatened", "Mammal", 
+                               "Herptile", "Invasive", "Invertebrate",
+                               "Lichen"),
+                    values = c("darkslategrey", "firebrick4", "darkslategrey", 
+                               "darkslategrey", "cyan4", "darkslategrey", 
+                               "darkslategrey")) +
+  theme(legend.position = "none") +
+  labs(title = "Modelled taxa", y = "", 
+       x = "number of articles")
+
+```
+
+## Threats to modelled taxa
+
+Of papers which describe any threats to their focal taxa, by far the most frequent threat was habitat loss.
+
+```{r Species_Threats}
+#| echo: false
+
+species_clean |>
+  filter(!Threat_Category %in% c("NA", "None"),
+         !is.na(Threat_Category)) |> 
+  separate_rows(Threat_Category, sep = ", ") |>
+  select(ID_clean, Threat_Category) |>
+  distinct() |>
+  mutate(Total = length(unique(ID_clean))) |>
+  summarise(Articles = n(), 
+            Proportion = round(Articles/mean(Total), 2),
+            .by = Threat_Category)|>
+  mutate(Threat_Category = case_when(Threat_Category == "HabitatLoss" ~ "Habitat loss",
+                                     Threat_Category == "ClimateChange" ~ "Climate change",
+                                     Threat_Category == "Other_Pollution" ~ "Pollution",
+                                     .default = Threat_Category)) |>
+  mutate(Threat_Category = fct_reorder(Threat_Category, Articles)) |>
+
+  ggplot() +
+  geom_col(aes(x = Proportion, y = Threat_Category), fill = "darkslategrey") +
+  xlim(c(0,1)) +
+  labs(title = "Threats to study taxa",
+       subtitle = "Of articles which explicitly mention any threats",
+       x = "Proportion of articles", y = "")
+```
+
+## Threatened taxa
+
+```{r Species_IUCN}
+#| echo: false
+
+species_clean |>
+  summarise(Articles = n(), .by = IUCN_Status) |>
+  filter(!IUCN_Status %in% c("Not applicable", "NA")) |>
+  mutate(IUCN_Status = case_when(IUCN_Status == "NL" ~ "Not listed",
+                                 .default = IUCN_Status)) |>
+  mutate(IUCN_Status = fct(IUCN_Status,
+                           levels = c("LC", "NT", "VU",
+                                      "EN", "CR", "Not listed"))) |>
+
+  ggplot() +
+  geom_col(aes(x = IUCN_Status, y = Articles, fill = IUCN_Status)) +
+  scale_fill_manual(values = c("darkslategrey", "darkorange3",
+                               "darkorange3", "darkorange3",
+                               "darkorange3", "darkslategrey")) +
+  labs(title = "IUCN Red list status",
+       subtitle = "Invasives, higher-order taxa, and grouped species excluded",
+       y = "Number of articles", x = "IUCN Red List Status") +
+  theme(legend.position = "none")
 ```
@@
 
-Only approximately half of articles modelled occupancy dynamics for a single species.
-Others fit multiple occupancy models: either multi-species models where 2+ species were modelled in concert, or many independent models for a suite of species.
+Only approximately half of articles modelled occupancy dynamics for a single taxa.
+Others fit multiple occupancy models: either multi-species models where 2+ taxa were modelled in concert, or many independent models for a suite of taxa.
 The largest number of species modelled in a single paper was **`{r} max(species_Number$Taxa_Number)`**.
@@
 ```
-
-## Threatened taxa
 

### 2024-03-16 Analysis/ReviewSummary/ReviewSummary.qmd
Study area size
Plot for study area size in kilometres

@@
 
+Consistent with authorship trends, a majority of articles conducted studies in the United States.
+
+::: panel-tabset
+## Country
+
+```{r Species_Country}
+#| echo: false
+
+species_clean |> 
+  select(ID_clean, Country) |>
+  separate_rows(Country, sep = ", ") |> 
+  distinct() |>
+  summarise(Articles = n(), .by = Country) |>
+  arrange(-Articles) |>
+  reactable(defaultPageSize = 8)
+```
+
+## Continent
+
+```{r Species_Continent}
+#| echo: false
+
+species_clean |> 
+  select(ID_clean, Continent) |>
+  distinct() |>
+  summarise(Articles = n(), .by = Continent) |>
+  arrange(-Articles) |>
+  reactable(defaultPageSize = 8)
+```
+
+## Realm
+
+```{r Species_Realm}
+#| echo: false
+
+species_clean |> 
+  select(ID_clean, Realm) |>
+  distinct() |>
+  summarise(Articles = n(), .by = Realm) |>
+  arrange(-Articles) |>
+  reactable(defaultPageSize = 8)
+
+```
+:::
+
+```{r Species_locations}
+
+```
+
+## Study area size
+
+The size of the study area in square kilometres follows a log-normal distribution, with studies ranging from less than one to over one million square kilometres.
+
+```{r Species_sizeKM}
+#| echo: false
+
+species_clean |> 
+  select(ID_clean, Size_km) |>
+  distinct() |>
+  mutate(Size_km = case_when(Size_km == "<1" ~ "0-1",
+                             Size_km == "1000-9999" ~ "1000-10000",
+                             .default = Size_km)) |>
+  summarise(Articles = n(), .by = Size_km) |>
+  arrange(Size_km) |>
+  mutate(Size_km = fct_inorder(Size_km)) |> 
+  filter(!Size_km %in% c("Not specified", "Unspecified")) |>
+
+  ggplot() +
+  geom_col(aes(x = Size_km, y = Articles), fill = "darkslategrey") +
+  labs(title = "Study area size",
+       x = "Study area in square kilometres", 
+       y = "Number of articles") +
+  theme(axis.text.x = element_text(angle = 60, vjust = 0.5))
+
+```
+
 # Data collection

### 2024-03-17 Analysis/ReviewSummary/ReviewSummary.qmd
tweaks to study species
changed highlighting for study taxa categories

@@
   summarise(Articles = n(), 
-            Proportion = round(Articles/mean(Total), 2),
+            Percentage = round(Articles/mean(Total), 2)*100,
             .by = Taxa_Category) |>
@@
   ggplot() +
-  geom_col(aes(x = Articles, y = Taxa, fill = Taxa)) +
+  geom_col(aes(x = Percentage, y = Taxa, fill = Taxa)) +
   scale_fill_manual(limits = c("Bird", "Threatened", "Mammal", 
@@
                                "Lichen"),
-                    values = c("darkslategrey", "firebrick4", "darkslategrey", 
-                               "darkslategrey", "cyan4", "darkslategrey", 
+                    values = c("darkslategrey", "darkorange4", "darkslategrey", 
+                               "darkslategrey", "darkorange4", "darkslategrey", 
                                "darkslategrey")) +
   theme(legend.position = "none") +
-  labs(title = "Modelled taxa", y = "", 
-       x = "number of articles")
+  xlim(c(0,50)) +
+  labs(title = "Modelled taxa", 
+       subtitle = "Multiple categories per article permitted",
+       y = "", x = "Percentage of articles")
 
@@
 
+```{r DataCollection_clean}
+#| echo: false
+
+```
+
 ## Study duration

### 2024-03-17 Analysis/ReviewSummary/ReviewSummary.qmd
Clean datacollection
create datacollection_clean object

@@
 
-species_taxaTable <-species_clean |>
+species_taxaTable <- species_clean |>
   separate_rows(Taxa_Category, sep = ", ") |>
@@
 Only approximately half of articles modelled occupancy dynamics for a single taxa.
-Others fit multiple occupancy models: either multi-species models where 2+ taxa were modelled in concert, or many independent models for a suite of taxa.
+Others fit multiple occupancy models: either multi-species models where 2+ taxa were modelled in concert, or many independent models fit for a suite of taxa.
 The largest number of species modelled in a single paper was **`{r} max(species_Number$Taxa_Number)`**.
@@
 #| echo: false
+#| warning: false
 
+dataCollection_clean <- review_DataCollection |>
+  filter(Status == "Complete") |>
+  mutate(Duration = as.numeric(Duration)) |>
+  separate_wider_delim(cols = Primary_Occasions, 
+                       delim = "-",
+                       names = c("Primary_Min", "Primary_Max"),
+                       too_few = "align_start") |>
+  separate_wider_delim(cols = Secondary_Occasions, 
+                       delim = "-",
+                       names = c("Secondary_Min", "Secondary_Max"),
+                       too_few = "align_start") |>
+  separate_wider_delim(cols = Site_Number, 
+                       delim = "-",
+                       names = c("SiteNumber_Min", "SiteNumber_Max"),
+                       too_few = "align_start") |>
+  mutate(Primary_Max = if_else(is.na(Primary_Max),
+                               Primary_Min, Primary_Max),
+         Secondary_Max = if_else(is.na(Secondary_Max),
+                                 Secondary_Min, Secondary_Max),
+         SiteNumber_Max = if_else(is.na(SiteNumber_Max),
+                                  SiteNumber_Min, SiteNumber_Max)) |>
+  mutate(across(c(Primary_Min, Primary_Max, 
+                  Secondary_Min, Secondary_Max,
+                  SiteNumber_Min, SiteNumber_Max),
+                as.numeric))
 ```

### 2024-03-18 Analysis/ReviewSummary/ReviewSummary.qmd
Study duration plot
log-scaled histogram of study duration

@@
 author: "Saoirse Kelleher"
-date: "March 13 2024"
-date-modifed: today
+date: today
 abstract: "A summary of results from the DOM Review, for use in interpreting outputs and writing the review manuscript. "
@@
 
+Study duration, defined as the length of time between the first and last survey, ranged from **`{r} min(dataCollection_clean$Duration)`** month to **`{r} max(dataCollection_clean$Duration)/12`** years.
+
+```{r DataCollection_duration}
+#| echo: false
+
+dataCollection_clean |>
+  ggplot() +
+  geom_histogram(aes(x = Duration), 
+                 binwidth = 0.1, fill = "darkslategrey") +
+  scale_x_continuous(transform = "log10", 
+                     breaks = c(1, 12, 60, 
+                                120, 240, 480), 
+                     labels = c("1 month", "1 year", "5 years", 
+                                "10 years", "20 years", "40 years")) +
+  labs(title = "Study duration distribution",
+       subtitle = "Defined as time between first and last survey",
+       y = "Frequency", x = "Study duration (log10 scale)")
+```
+
 ## Number of occasions

### 2024-03-18 Analysis/ReviewSummary/ReviewSummary.qmd
Number of occasions
Number of primary, secondary, and total occasions plot

@@
 
+```{r DataCollection_PrimaryOccasions}
+#| echo: false
+
+dataCollection_clean |>
+  rowwise() |>
+  mutate(Primary_Max = case_when(Primary_Max > 30 ~ 30,
+                                  .default = Primary_Max)) |>
+
+  ggplot() +
+  geom_histogram(aes(x = Primary_Max),
+                 binwidth = 1, fill = "darkslategrey") +
+  labs(title = "Number of primary occasions",
+       x = "Number of primary occasions", y = "Number of articles") +
+  scale_x_continuous(breaks = c(1, 5, 10, 20, 30),
+                     labels = c("1", "5", "10", "20", "30+"))
+```
+
+```{r DataCollection_SecondaryOccasions}
+#| echo: false
+
+dataCollection_clean |>
+  filter(!is.na(Secondary_Max)) |>
+  rowwise() |>
+  mutate(Secondary_Max = case_when(Secondary_Max > 20 ~ 20,
+                                   .default = Secondary_Max)) |>
+
+  ggplot() +
+  geom_histogram(aes(x = Secondary_Max),
+                 binwidth = 1, fill = "darkcyan") +
+  labs(title = "Number of secondary occasions",
+       subtitle = "Where variable, maximum number of occasions is presented",
+       x = "Number of secondary occasions", y = "Number of articles") +
+  scale_x_continuous(breaks = c(1, 5, 10, 20),
+                     labels = c("1", "5", "10", "20+"))
+```
+
+```{r DataCollection_TotalOccasions}
+#| echo: false
+
+dataCollection_clean |>
+  filter(!is.na(Secondary_Max) & !is.na(Primary_Max)) |>
+  mutate(TotalOccasions = Secondary_Max*Primary_Max) |>
+  ggplot() +
+  geom_histogram(aes(x = TotalOccasions),
+                 binwidth = 0.1,
+                 fill = "darkorchid3") +
+  labs(title = "Total number of occasions",
+       subtitle = "Product of primary and secondary occasions",
+       x = "Number of occasions (log10 scale)", y = "Number of articles") +
+  scale_x_continuous(transform = "log10",
+                     breaks = c(4, 10, 25, 50, 100, 500, 1000, 2000)) +
+  scale_y_continuous(breaks = c(1, 3, 6, 9))
+```
+
 ## Data collection method

### 2024-03-18 Analysis/ReviewSummary/ReviewSummary.qmd
Number of sites fig
Number of sites figure, adjusted min/max/median text

@@
   filter(Status == "Complete") |>
+  mutate(ID_clean = str_extract(ID, "[:digit:]+")) |>
   mutate(Duration = as.numeric(Duration)) |>
@@
 
-Study duration, defined as the length of time between the first and last survey, ranged from **`{r} min(dataCollection_clean$Duration)`** month to **`{r} max(dataCollection_clean$Duration)/12`** years.
+Study duration, defined as the length of time between the first and last survey, ranged from **`{r} min(dataCollection_clean$Duration)`** month to **`{r} max(dataCollection_clean$Duration)/12`** years, with a median of **`{r} median(dataCollection_clean$Duration`** months.
 
@@
 ## Number of occasions
+
+The number of sampling occasions varied considerably.
+The number of primary occasions ranged from just **`{r} min(dataCollection_clean$Primary_Max)`** to **`{r} max(dataCollection_clean$Primary_Max)`** with a median of **`{r} median(dataCollection_clean$Primary_Max)`**.
+
+The number of secondary occasions from **`{r} min(dataCollection_clean$Secondary_Max, na.rm = TRUE)`** to **`{r} max(dataCollection_clean$Secondary_Max, na.rm = TRUE)`** with a median of **`{r} median(dataCollection_clean$Secondary_Max, na.rm = TRUE)`**.
 
@@
 
+The most common method for recording detections was human surveys, either structured (e.g.
+time-bound point counts or transects) or unstructured (searches of sites).
+Autonomous detection methods (camera traps & acoustic monitors) were used infrequently, despite their potential suitability.
+
+```{r DataCollection_SamplingMethod}
+#| echo: false
+
+dataCollection_clean |>
+  mutate(Total = length(unique(ID_clean))) |>
+  separate_rows(Detection_Method, sep = ", ") |>
+  summarise(Articles = n(), 
+            #Proportion = round(Articles/Total, 2),
+            .by = c(Detection_Method, Total)) |>
+  mutate(Percentage = round(Articles/Total, 2)*100) |>
+  select(Detection_Method, Articles, Percentage) |>
+  filter(!Detection_Method %in% c("66", NA, "Unspecificed",
+                                  "Survey", "Incidental")) |>
+  arrange(Percentage) |>
+  mutate(Detection_Method = case_when(Detection_Method == "AudioVisual" ~ "Human structured survey",
+                                      Detection_Method == "Search" ~ "Human unstructured search",
+                                      Detection_Method == "Trapping" ~ "Physical trapping",
+                                      Detection_Method == "Camera" ~ "Camera trapping",
+                                      Detection_Method == "Citizen" ~ "Citizen science",
+                                      Detection_Method == "Bioacoustic" ~ "Acoustic monitor")) |>
+  mutate(Detection_Method = fct_inorder(Detection_Method)) |>
+
+  ggplot() +
+  geom_col(aes(x = Percentage, y = Detection_Method), fill = "darkslategrey") +
+  labs(title = "Detection method",
+       subtitle = "Multiple methods per article permitted",
+       x = "Percentage of articles", y = "") +
+  xlim(c(0,50))
+```
+
+## Number of sites
+
+The number of sites where surveys were conducted ranged widely from **`{r} min(dataCollection_clean$SiteNumber_Max, na.rm = TRUE)`** to **`{r} max(dataCollection_clean$SiteNumber_Max, na.rm = TRUE)`**, with a median site count of **`{r} median(dataCollection_clean$SiteNumber_Max, na.rm = TRUE)`**.
+
+```{r DataCollection_numberSites}
+#| echo: false
+
+dataCollection_clean |>
+  filter(!is.na(SiteNumber_Max)) |>
+
+  ggplot() +
+  geom_histogram(aes(x = SiteNumber_Max), 
+                 binwidth = 0.1, 
+                 fill = "darkslategrey") +
+  scale_x_continuous(transform = "log10",
+                     breaks = c(5, 10, 25, 50, 100, 500, 1000, 10000)) +
+  labs(title = "Number of sites",
+       x = "Number of sites", y = "Number of articles")
+```
+
 ## Sampling method
+
+```{r DataCollection_Sampling}
+
+```
 

### 2024-03-18 Analysis/ReviewSummary/ReviewSummary.qmd
study size over time
plotted trends/boxplots for study size over time

@@
 
-Study duration, defined as the length of time between the first and last survey, ranged from **`{r} min(dataCollection_clean$Duration)`** month to **`{r} max(dataCollection_clean$Duration)/12`** years, with a median of **`{r} median(dataCollection_clean$Duration`** months.
+Study duration, defined as the length of time between the first and last survey, ranged from **`{r} min(dataCollection_clean$Duration)`** month to **`{r} max(dataCollection_clean$Duration)/12`** years, with a median of **`{r} median(dataCollection_clean$Duration)`** months.
 
@@
   summarise(Articles = n(), 
-            #Proportion = round(Articles/Total, 2),
             .by = c(Detection_Method, Total)) |>
@@
 
+Site selection was variable; various studies selected sites randomly, as part of a grid design, preferentially, or exhaustively.
+A considerable amount did not clearly specify how their study sites were determined.
+
 ```{r DataCollection_Sampling}
+#| echo: false
 
+dataCollection_clean |>
+  mutate(Total = length(unique(ID_clean))) |>
+  separate_rows(Sampling_Method, sep = ", ") |>
+  mutate(Sampling_Method = case_when(Sampling_Method == "Unspecified" ~ "Not specified",
+         .default = Sampling_Method)) |>
+  summarise(Articles = n(), 
+            .by = c(Sampling_Method, Total)) |>
+  mutate(Percentage = round(Articles/Total, 2)*100) |>
+  select(-Total) |>
+  arrange(-Percentage) |>
+  kable()
 ```
@@
 ## Study size over time
+
+There does not appear to be a discernible trend in study size (whether defined by duration, number of sites, or total number of survey occasions.
+
+::: panel-tabset
+## Year
+
+```{r DataCollection_sizeTrend}
+#| echo: false
+#| warning: false
+#| message: false
+
+dataCollection_clean |>
+  select(ID_clean, Duration, 
+         Primary_Max, Secondary_Max,
+         SiteNumber_Max) |>
+  mutate(ID_clean = as.numeric(ID_clean)) |>
+  rowwise() |>
+  mutate(TotalOccasions = Primary_Max*Secondary_Max) |>
+  left_join(select(studyDetails_clean, ID, Year), by = join_by(ID_clean == ID)) |>
+  pivot_longer(cols = c(Duration, TotalOccasions, SiteNumber_Max),
+               names_to = "Variable", values_to = "Quantity") |>
+  mutate(Variable = case_when(Variable == "Duration" ~ "Study duration (months)",
+                              Variable == "SiteNumber_Max" ~ "Number of sites",
+                              Variable == "TotalOccasions" ~ "Total survey occasions (Primary x Secondary)")) |>
+
+  ggplot(aes(x = Year, y = Quantity, colour = Variable)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "lm", se = FALSE, alpha = 0.7) +
+  scale_y_continuous(transform = "log10") +
+  xlim(c(2005, 2024)) +
+  scale_x_continuous(breaks = c(2005, 2010, 2015, 2020, 2024)) +
+  scale_colour_manual(values = c("darkslategrey", "cyan4", "darkorchid4")) +
+  facet_wrap(vars(Variable), ncol = 1, scales = "free_y") +
+  theme(legend.position = "none") +
+  labs(title = "Trend in study size",
+       x = "Year", y = "")
+```
+
+## Year Strata
+
+```{r DataCollection_Sizetrendboxplots}
+#| echo: false
+#| warning: false
+#| message: false
+
+dataCollection_clean |>
+  select(ID_clean, Duration, 
+         Primary_Max, Secondary_Max,
+         SiteNumber_Max) |>
+  mutate(ID_clean = as.numeric(ID_clean)) |>
+  rowwise() |>
+  mutate(TotalOccasions = Primary_Max*Secondary_Max) |>
+  left_join(select(studyDetails_clean, ID, YearStrata), by = join_by(ID_clean == ID)) |>
+  filter(!is.na(YearStrata)) |>
+  pivot_longer(cols = c(Duration, TotalOccasions, SiteNumber_Max),
+               names_to = "Variable", values_to = "Quantity") |>
+  mutate(Variable = case_when(Variable == "Duration" ~ "Study duration (months)",
+                              Variable == "SiteNumber_Max" ~ "Number of sites",
+                              Variable == "TotalOccasions" ~ "Total survey occasions (Primary x Secondary)")) |>
+
+  ggplot(aes(x = YearStrata, y = Quantity, colour = Variable)) +
+  geom_point(alpha = 0.3) +
+  geom_boxplot() +
+  scale_y_continuous(transform = "log10") +
+  scale_colour_manual(values = c("darkslategrey", "cyan4", "darkorchid4")) +
+  facet_wrap(vars(Variable), ncol = 1, scales = "free_y") +
+  theme(legend.position = "none") +
+  labs(title = "Trend in study size",
+       x = "Year strata", y = "")
+```
+:::
 

### 2024-03-18 Analysis/ReviewSummary/ReviewSummary.qmd
Covariate quantity
Cleaned covariate data; added plots for number of covariates

@@
 
-The most common objective was **Testing Hypotheses**, with `{r} max(objectives_frequency$Proportion)`% of articles examining pre-determined relationships.
+The most common objective was **Testing Hypotheses**, with `{r} round(max(objectives_frequency$Proportion), 2)`% of articles examining pre-determined relationships.
 Prediction, whether spatial or temporal, was rare.
@@
 ```{r Species_clean}
+#| echo: false
 species_clean <- review_SpeciesLocation |>
@@
 
+```{r Covariates_clean}
+#| echo: false
+
+covariates_clean <- review_Covariates |>
+  filter(Status == "Complete") |>
+  mutate(ID_clean = str_extract(ID, "[:digit:]+")) |>
+  mutate(Covariates = str_replace(Covariates, "\\.", "\\,")) |>
+  separate_rows(Covariates, sep = ", ") |>
+  mutate(Covariates = str_to_upper(Covariates)) |>
+  separate_wider_delim(cols = Covariates, delim = "_",
+                       names = c("Type", 
+                                 "Variation",
+                                 "Measurement",
+                                 "Complexity"),
+                       too_few = "align_start", 
+                       too_many = "drop") |>
+  mutate(ParameterCategory = case_when(Parameter %in% c("InitialOccupancy",
+                                                        "InitialOccupancy_MS",
+                                                        "Occupancy",
+                                                        "Occupancy_MS") ~ "Psi",
+                                       Parameter %in% c("Colonisation",
+                                                        "Colonisation_MS") ~ "Gamma",
+                                       Parameter %in% c("Extinction",
+                                                        "Extinction_MS",
+                                                        "Persistence",
+                                                        "Persistence_MS") ~ "Eps",
+                                       Parameter %in% c("Detection",
+                                                        "Detection_FP",
+                                                        "Detection_FP_MS",
+                                                        "Detection_MS", 
+                                                        "Detection_Other") ~ "p",
+                                       .default = "Other"),
+         Type = case_when(Type %in% c("E-HTT", " E-HTT") ~ "E-HTT",
+                          Type %in% c("E-DIS", "DIS") ~ "E-DIS",
+                          Type %in% c("ANTH", "E-ANTH") ~ "E-ANTH",
+                          Type %in% c("SP", "Spec") ~ "SP",
+                          Type %in% c("NONE", "NA") ~ "NONE",
+                          .default = Type),
+         Variation = case_when(Variation %in% c("SY", "YS") ~ "Site-Year",
+                               Variation == "S" ~ "Site",
+                               Variation == "YSS" ~ "Site-Year-Survey",
+                               Variation == "Y" ~ "Year",
+                               Variation == "SP" ~ "Species",
+                               Variation == "U" ~ "Unclear"),
+         Measurement = case_when(Measurement == "I" ~ "Intrinsic",
+                                 Measurement == "O" ~ "Observed",
+                                 Measurement == "R" ~ "Remote",
+                                 Measurement == "U" ~ "Unclear"),
+         Complexity = case_when(Complexity %in% c("N", "NH", "N,") ~ "Linear",
+                                Complexity == "Q" ~ "Quadratic",
+                                Complexity == "I" ~ "Interaction",
+                                Complexity == "B" ~ "Quadratic + Interaction",
+                                Complexity == "O" ~ "Other"))
+```
+
 ## Covariate quantity
+
+::: panel-tabset
+## Overall
+
+```{r covariates_quantityOverall}
+#| echo: false
+
+covariates_clean |>
+  filter(Type != "NONE") |>
+  summarise(Covariates = n(), .by = ID) |>
+
+  ggplot() +
+  geom_histogram(aes(x = Covariates), 
+                 binwidth = 5,
+                 fill = "darkslategrey") +
+  labs(title = "Total number of covariates considered",
+       subtitle = "All parameters",
+       x = "Number of covariates", y = "Number of articles")
+```
+
+## By parameter
+
+```{r covariates_quantityParameter}
+#| echo: false
+
+covariates_clean |>
+  filter(Type != "NONE") |>
+  summarise(Covariates = n(), .by = c(ID, ParameterCategory)) |>
+  filter(ParameterCategory != "Other") |>
+  mutate(ParameterCategory = case_when(ParameterCategory == "Psi" ~ "Initial Occupancy",
+                                       ParameterCategory == "Gamma" ~ "Colonisation",
+                                       ParameterCategory == "Eps" ~ "Extinction",
+                                       ParameterCategory == "p" ~ "Detection")) |>
+  complete(ID, ParameterCategory, fill= list(Covariates = 0)) |>
+  mutate(ParameterCategory = fct(ParameterCategory,
+                                 levels = c("Initial Occupancy",
+                                            "Colonisation", 
+                                            "Extinction",
+                                            "Detection"))) |>
+  mutate(Covariates = case_when(Covariates > 10 ~ 10,
+                                .default = Covariates)) |>
+
+  ggplot() +
+  geom_histogram(aes(x = Covariates, fill = ParameterCategory), 
+                 binwidth = 1) +
+  labs(title = "Total number of covariates considered",
+       x = "Number of covariates", y = "Number of articles") +
+  scale_fill_manual(values = c("darkslategrey", "darkorchid", 
+                               "cyan2", "goldenrod")) +
+  scale_x_continuous(breaks = c(0, 5, 10), 
+                     labels = c("0", "5", "10+")) +
+  facet_grid(rows = vars(ParameterCategory)) +
+  theme(legend.position = "none")
+```
+:::
 

### 2024-03-18 Analysis/ReviewSummary/ReviewSummary.qmd
Covariate changes plots
Plots for types of covs, attributes, changes in time, study size, etc

@@
 
-+-------------------+--------------------------------------------------+--------------------------------------+
 | Field             | Description                                      | Possible values                      |
-+===================+==================================================+======================================+
+|-------------------|--------------------------------------------------|--------------------------------------|
 | Journal           | Journal which this article was published in      | *Any*                                |
-+-------------------+--------------------------------------------------+--------------------------------------+
 | Year              | Year of publication                              | *Any*                                |
-+-------------------+--------------------------------------------------+--------------------------------------+
 | Author country    | Nation of the first author's primary affiliation | *ISO 3166 country names*             |
-+-------------------+--------------------------------------------------+--------------------------------------+
 | Author categories | Categories of affiliations for *all* authors     | *Academic, Government, NGO, Private* |
-+-------------------+--------------------------------------------------+--------------------------------------+
 
@@
 
-+---------------------+-----------------------------------------------------------------------------------+
 | Objective           | Description                                                                       |
-+=====================+===================================================================================+
+|---------------------|-----------------------------------------------------------------------------------|
 | Methods             | Did authors seek to demonstrate or test specific methods for occupancy modelling? |
-+---------------------+-----------------------------------------------------------------------------------+
 | Future Predictions  | Did authors make predictions beyond the study period?                             |
-+---------------------+-----------------------------------------------------------------------------------+
 | Spatial Predictions | Did authors make predictions at locations where surveys were not conducted?       |
-+---------------------+-----------------------------------------------------------------------------------+
 | Hypothesis testing  | Did authors seek to test specific hypotheses of species occupancy dynamics?       |
-+---------------------+-----------------------------------------------------------------------------------+
 
@@
 
-+-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Field           | Description                                                                                | Possible values                                                                   |
-+=================+============================================================================================+===================================================================================+
+|-----------------|--------------------------------------------------------------------------------------------|-----------------------------------------------------------------------------------|
 | Taxa category   | Category of study taxa; could fit into multiple categories                                 | *Bird, Mammal, Invertebrate, Herptile, Fish, Threatened, Invasive, Other*         |
-+-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Taxa number     | Number of individual species within the taxa for which models were fit                     | *Numeric*                                                                         |
-+-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | IUCN status     | IUCN status for the study species; NA if multi-species or if the study species is invasive | *LC, NT, EN, CR, DD, NA*                                                          |
-+-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Threat category | Category of threats to the species; multiple per species permitted                         | *Habitat loss, Climate change, Exploitation, Invasives, Disease, Disaster, Other* |
-+-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Country         | Country/countries where the study took place                                               | *ISO 3166 country names*                                                          |
-+-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Continent       | Continent(s) where the study took place                                                    | *North America, South America, Europe, Asia, Africa, Oceania*                     |
-+-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Realm           | Biogeographic realm where the study took place                                             | *Nearctic, Palearctic, Afrotropical, Indomalayan, Australasian, Neotropical*      |
-+-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Size            | Size of the study area extent in square kilometres                                         | *\<1km, 1-10, 10-100, 100-1000, 10000-100000, 100000-1000000, 1000000+*           |
-+-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 
@@
 
-+---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Field               | Description                                                              | Possible values                                                                              |
-+=====================+==========================================================================+==============================================================================================+
+|---------------------|--------------------------------------------------------------------------|----------------------------------------------------------------------------------------------|
 | Duration            | Length of study in months                                                | *Numeric*                                                                                    |
-+---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Capture method      | Method of recording detections; multiple categories possible             | *Audiovisual, Search, Trapping, Bioacoustic, Camera, Citizen, Incidental*                    |
-+---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Primary occasions   | Number of primary occasions                                              | *Numeric*                                                                                    |
-+---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Secondary occasions | Number of secondary occasions, presented as a range where not consistent | *Numeric range*                                                                              |
-+---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Secondary type      | Type of secondary occasion                                               | *Revisits, Spatial replicates, Same-visit replicates, multiple observers, time to detection* |
-+---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Type of site        | How a 'site' was defined                                                 | *Point location, Grid cell, Irregular area, Transect*                                        |
-+---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Number of sites     | Number of individual sites                                               | Numeric                                                                                      |
-+---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Sampling method     | How sites were selected                                                  | *Random, Stratified, Grid, Preferential, Available, Exhaustive, Uniform, Transect*           |
-+---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 
@@
                           Type %in% c("ANTH", "E-ANTH") ~ "E-ANTH",
-                          Type %in% c("SP", "Spec") ~ "SP",
+                          Type %in% c("SP", "SPEC") ~ "SP",
+                          Type %in% c("INT", " INT") ~ "INT",
                           Type %in% c("NONE", "NA") ~ "NONE",
@@
 
+::: column-page
+::: panel-tabset
+## Overall
+
+```{r covariate_categoriesOverall}
+#| echo: false
+
+covariates_clean |>
+  mutate(Total = length(unique(ID))) |>
+  select(ID, Type, Total) |>
+  distinct() |>
+  summarise(Articles = n(),
+            .by = c(Type, Total)) |>
+  mutate(Percentage = round(Articles/Total, 2)*100) |>
+  arrange(-Percentage) |>
+  filter(!Type %in% c("UNSPECIFIED", "UNCLEAR", "NONE", NA, "OTHER")) |>
+  mutate(Type = case_when(Type == "E-ANTH" ~ "Anthropogenic",
+                          Type == "E-CLIM" ~ "Climate/Weather",
+                          Type == "E-DIS" ~ "Disaster/Disturbance",
+                          Type == "E-HTAQ" ~ "Habitat (Aquatic)",
+                          Type == "E-HTT" ~ "Habitat (Terrestrial)",
+                          Type == "E-HYD" ~ "Hydrologic",
+                          Type == "E-SEAS" ~ "Phenologic",
+                          Type == "E-TOPO" ~ "Topographic",
+                          Type == "CONN" ~ "Connectivity",
+                          Type == "GEOM" ~ "Site geometry",
+                          Type == "INT" ~ "Biotic interaction",
+                          Type == "SEAS" ~ "Primary occasion",
+                          Type == "SITE" ~ "Site",
+                          Type == "SP" ~ "Species",
+                          Type == "SURV" ~ "Survey detail")) |>
+  mutate(Type_cat = case_when(Type %in% c("Habitat (Terrestrial)",
+                                          "Habitat (Aquatic)",
+                                          "Hydrologic", 
+                                          "Climate/Weather",
+                                          "Disaster/Disturbance",
+                                          "Topographic",
+                                          "Phenologic",
+                                          "Anthropogenic",
+                                          "Biotic interaction") ~ "Env",
+                              .default = "Other")) |>
+  mutate(Type = fct(Type,
+                    levels = c("Habitat (Terrestrial)", "Habitat (Aquatic)",
+                               "Hydrologic", "Climate/Weather",
+                               "Disaster/Disturbance", "Topographic",
+                               "Phenologic", "Anthropogenic",
+                               "Biotic interaction", "Primary occasion",
+                               "Survey detail", "Species", 
+                               "Site", "Site geometry", "Connectivity"))) |>
+  select(Type, Type_cat, Percentage) |>
+
+  ggplot() +
+  geom_col(aes(x = Type, y = Percentage, fill = Type_cat)) +
+  labs(title = "Most frequent covariate types (All parameters)",
+       x = "", y = "Percentage of models (any parameter)") +
+  scale_fill_manual(values = c("darkslategrey", "darkorange3")) +
+  theme(axis.text.x = element_text(angle = 90, 
+                                   vjust = 0.5, hjust = 1),
+        legend.position = "none")
+```
+
+## Initial Occupancy
+
+```{r Covariate_categoriesPsi}
+#| echo: false
+
+covariates_clean |>
+  mutate(Total = length(unique(ID))) |>
+  filter(ParameterCategory == "Psi") |>
+  select(ID, Type, Total) |>
+  distinct() |>
+  summarise(Articles = n(),
+            .by = c(Type, Total)) |>
+  mutate(Percentage = round(Articles/Total, 2)*100) |>
+  arrange(-Percentage) |>
+  filter(!Type %in% c("UNSPECIFIED", "UNCLEAR", "NONE", NA, "SURV", "SEAS", "OTHER")) |>
+  mutate(Type = case_when(Type == "E-ANTH" ~ "Anthropogenic",
+                          Type == "E-CLIM" ~ "Climate/Weather",
+                          Type == "E-DIS" ~ "Disaster/Disturbance",
+                          Type == "E-HTAQ" ~ "Habitat (Aquatic)",
+                          Type == "E-HTT" ~ "Habitat (Terrestrial)",
+                          Type == "E-HYD" ~ "Hydrologic",
+                          Type == "E-SEAS" ~ "Phenologic",
+                          Type == "E-TOPO" ~ "Topographic",
+                          Type == "CONN" ~ "Connectivity",
+                          Type == "GEOM" ~ "Site geometry",
+                          Type == "INT" ~ "Biotic interaction",
+                          Type == "SEAS" ~ "Primary occasion",
+                          Type == "SITE" ~ "Site",
+                          Type == "SP" ~ "Species",
+                          Type == "SURV" ~ "Survey detail",
+                          Type == "OTHER" ~ "Other")) |>  
+  mutate(Type_cat = case_when(Type %in% c("Habitat (Terrestrial)",
+                                          "Habitat (Aquatic)",
+                                          "Hydrologic", 
+                                          "Climate/Weather",
+                                          "Disaster/Disturbance",
+                                          "Topographic",
+                                          "Phenologic",
+                                          "Anthropogenic",
+                                          "Biotic interaction") ~ "Env",
+                              .default = "Other")) |>
+  mutate(Type = fct(Type,
+                    levels = c("Habitat (Terrestrial)", "Habitat (Aquatic)",
+                               "Hydrologic", "Climate/Weather",
+                               "Disaster/Disturbance", "Topographic",
+                               "Phenologic", "Anthropogenic",
+                               "Biotic interaction", "Species", 
+                               "Site", "Site geometry", "Connectivity"))) |>
+  select(Type, Type_cat, Percentage) |>
+
+  ggplot() +
+  geom_col(aes(x = Type, y = Percentage, fill = Type_cat)) +
+  labs(title = "Most frequent covariate types (Initial Occupancy)",
+       x = "", y = "Percentage of models") +
+  scale_fill_manual(values = c("darkslategrey", "darkorange3")) +
+  theme(axis.text.x = element_text(angle = 90, 
+                                   vjust = 0.5, hjust = 1),
+        legend.position = "none")
+```
+
+## Colonisation
+
+```{r covariates_catrgoriesGamma}
+#| echo: false
+
+covariates_clean |>
+  mutate(Total = length(unique(ID))) |>
+  filter(ParameterCategory == "Gamma") |>
+  select(ID, Type, Total) |>
+  distinct() |>
+  summarise(Articles = n(),
+            .by = c(Type, Total)) |>
+  mutate(Percentage = round(Articles/Total, 2)*100) |>
+  arrange(-Percentage) |>
+  filter(!Type %in% c("UNSPECIFIED", "UNCLEAR", "NONE", NA, "SURV", "OTHER")) |>
+  mutate(Type = case_when(Type == "E-ANTH" ~ "Anthropogenic",
+                          Type == "E-CLIM" ~ "Climate/Weather",
+                          Type == "E-DIS" ~ "Disaster/Disturbance",
+                          Type == "E-HTAQ" ~ "Habitat (Aquatic)",
+                          Type == "E-HTT" ~ "Habitat (Terrestrial)",
+                          Type == "E-HYD" ~ "Hydrologic",
+                          Type == "E-SEAS" ~ "Phenologic",
+                          Type == "E-TOPO" ~ "Topographic",
+                          Type == "CONN" ~ "Connectivity",
+                          Type == "GEOM" ~ "Site geometry",
+                          Type == "INT" ~ "Biotic interaction",
+                          Type == "SEAS" ~ "Primary occasion",
+                          Type == "SITE" ~ "Site",
+                          Type == "SP" ~ "Species",
+                          Type == "SURV" ~ "Survey detail",
+                          Type == "OTHER" ~ "Other")) |>  
+  mutate(Type_cat = case_when(Type %in% c("Habitat (Terrestrial)",
+                                          "Habitat (Aquatic)",
+                                          "Hydrologic", 
+                                          "Climate/Weather",
+                                          "Disaster/Disturbance",
+                                          "Topographic",
+                                          "Phenologic",
+                                          "Anthropogenic",
+                                          "Biotic interaction") ~ "Env",
+                              .default = "Other")) |>
+  mutate(Type = fct(Type,
+                    levels = c("Habitat (Terrestrial)", "Habitat (Aquatic)",
+                               "Hydrologic", "Climate/Weather",
+                               "Disaster/Disturbance", "Topographic",
+                               "Phenologic", "Anthropogenic",
+                               "Biotic interaction", "Species",
+                               "Primary occasion", "Site",
+                               "Site geometry", "Connectivity"))) |>
+  select(Type, Type_cat, Percentage) |>
+
+  ggplot() +
+  geom_col(aes(x = Type, y = Percentage, fill = Type_cat)) +
+  labs(title = "Most frequent covariate types (Colonisation)",
+       x = "", y = "Percentage of models") +
+  scale_fill_manual(values = c("darkslategrey", "darkorange3")) +
+  theme(axis.text.x = element_text(angle = 90, 
+                                   vjust = 0.5, hjust = 1),
+        legend.position = "none")
+```
+
+## Extinction
+
+```{r covariates_categoryEps}
+#| echo: false
+
+covariates_clean |>
+  mutate(Total = length(unique(ID))) |>
+  filter(ParameterCategory == "Eps") |>
+  select(ID, Type, Total) |>
+  distinct() |>
+  summarise(Articles = n(),
+            .by = c(Type, Total)) |>
+  mutate(Percentage = round(Articles/Total, 2)*100) |>
+  arrange(-Percentage) |>
+  filter(!Type %in% c("UNSPECIFIED", "UNCLEAR", "NONE", NA, "SURV", "OTHER")) |>
+  mutate(Type = case_when(Type == "E-ANTH" ~ "Anthropogenic",
+                          Type == "E-CLIM" ~ "Climate/Weather",
+                          Type == "E-DIS" ~ "Disaster/Disturbance",
+                          Type == "E-HTAQ" ~ "Habitat (Aquatic)",
+                          Type == "E-HTT" ~ "Habitat (Terrestrial)",
+                          Type == "E-HYD" ~ "Hydrologic",
+                          Type == "E-SEAS" ~ "Phenologic",
+                          Type == "E-TOPO" ~ "Topographic",
+                          Type == "CONN" ~ "Connectivity",
+                          Type == "GEOM" ~ "Site geometry",
+                          Type == "INT" ~ "Biotic interaction",
+                          Type == "SEAS" ~ "Primary occasion",
+                          Type == "SITE" ~ "Site",
+                          Type == "SP" ~ "Species",
+                          Type == "SURV" ~ "Survey detail",
+                          Type == "OTHER" ~ "Other")) |>  
+  mutate(Type_cat = case_when(Type %in% c("Habitat (Terrestrial)",
+                                          "Habitat (Aquatic)",
+                                          "Hydrologic", 
+                                          "Climate/Weather",
+                                          "Disaster/Disturbance",
+                                          "Topographic",
+                                          "Phenologic",
+                                          "Anthropogenic",
+                                          "Biotic interaction") ~ "Env",
+                              .default = "Other")) |>
+  mutate(Type = fct(Type,
+                    levels = c("Habitat (Terrestrial)", "Habitat (Aquatic)",
+                               "Hydrologic", "Climate/Weather",
+                               "Disaster/Disturbance", "Topographic",
+                               "Phenologic", "Anthropogenic",
+                               "Biotic interaction", "Species",
+                               "Primary occasion", "Site",
+                               "Site geometry", "Connectivity"))) |>
+  select(Type, Type_cat, Percentage) |>
+
+  ggplot() +
+  geom_col(aes(x = Type, y = Percentage, fill = Type_cat)) +
+  labs(title = "Most frequent covariate types (Extinction)",
+       x = "", y = "Percentage of models") +
+  scale_fill_manual(values = c("darkslategrey", "darkorange3")) +
+  theme(axis.text.x = element_text(angle = 90, 
+                                   vjust = 0.5, hjust = 1),
+        legend.position = "none")
+
+```
+
+## Detection
+
+```{r covariate_categoriesDet}
+#| echo: false
+
+covariates_clean |>
+  mutate(Total = length(unique(ID))) |>
+  filter(ParameterCategory == "p") |>
+  select(ID, Type, Total) |>
+  distinct() |>
+  summarise(Articles = n(),
+            .by = c(Type, Total)) |>
+  mutate(Percentage = round(Articles/Total, 2)*100) |>
+  arrange(-Percentage) |>
+  filter(!Type %in% c("UNSPECIFIED", "UNCLEAR", "NONE", "OTHER", NA)) |>
+  mutate(Type = case_when(Type == "E-ANTH" ~ "Anthropogenic",
+                          Type == "E-CLIM" ~ "Climate/Weather",
+                          Type == "E-DIS" ~ "Disaster/Disturbance",
+                          Type == "E-HTAQ" ~ "Habitat (Aquatic)",
+                          Type == "E-HTT" ~ "Habitat (Terrestrial)",
+                          Type == "E-HYD" ~ "Hydrologic",
+                          Type == "E-SEAS" ~ "Phenologic",
+                          Type == "E-TOPO" ~ "Topographic",
+                          Type == "CONN" ~ "Connectivity",
+                          Type == "GEOM" ~ "Site geometry",
+                          Type == "INT" ~ "Biotic interaction",
+                          Type == "SEAS" ~ "Primary occasion",
+                          Type == "SITE" ~ "Site",
+                          Type == "SP" ~ "Species",
+                          Type == "SURV" ~ "Survey detail",
+                          Type == "OTHER" ~ "Other")) |>  
+  mutate(Type_cat = case_when(Type %in% c("Habitat (Terrestrial)",
+                                          "Habitat (Aquatic)",
+                                          "Hydrologic", 
+                                          "Climate/Weather",
+                                          "Disaster/Disturbance",
+                                          "Topographic",
+                                          "Phenologic",
+                                          "Anthropogenic",
+                                          "Biotic interaction") ~ "Env",
+                              .default = "Other")) |>
+  mutate(Type = fct(Type,
+                    levels = c("Habitat (Terrestrial)", "Habitat (Aquatic)",
+                               "Hydrologic", "Climate/Weather",
+                               "Disaster/Disturbance", "Topographic",
+                               "Phenologic", "Anthropogenic",
+                               "Biotic interaction", "Species",
+                               "Primary occasion", "Site", "Survey detail",
+                               "Site geometry", "Connectivity"))) |>
+  select(Type, Type_cat, Percentage) |>
+
+  ggplot() +
+  geom_col(aes(x = Type, y = Percentage, fill = Type_cat)) +
+  labs(title = "Most frequent covariate types (Detection)",
+       x = "", y = "Percentage of models") +
+  scale_fill_manual(values = c("darkslategrey", "darkorange3")) +
+  theme(axis.text.x = element_text(angle = 90, 
+                                   vjust = 0.5, hjust = 1),
+        legend.position = "none")
+
+```
+:::
+:::
+
 ## Covariate features
 
-## Frequency of covariate features
+Aspects of each covariate are presented here as percentages; e.g., the frequency that a class of covariate had a quadratic term, or the frequency at which it was measured via remote sensing.
+
+::: column-page
+```{r Covariates_Features table}
+#| echo: false
+
+covariates_clean |> 
+  mutate(Overall_Total = length(unique(ID))) |>
+  mutate(Meas_Observed = case_when(Measurement == "Observed" ~ 1,
+                                   .default = 0),
+         Meas_Intrinsic = case_when(Measurement == "Intrinsic" ~ 1,
+                                    .default = 0),
+         Meas_Remote = case_when(Measurement == "Remote" ~ 1,
+                                 .default = 0),
+         Dynamic = case_when(Variation %in% c("Year",
+                                              "Year-Site", 
+                                              "Year-Site-Survey") ~ 1,
+                             .default =0),
+         Linear = case_when(Complexity == "Linear" ~ 1,
+                               .default = 0),
+         Quadratic = case_when(Complexity %in% c("Quadratic + Interaction",
+                                                 "Quadratic") ~ 1,
+                               .default = 0),
+         Interaction = case_when(Complexity %in% c("Quadratic + Interaction",
+                                                   "Interaction") ~ 1,
+                               .default = 0)) |>
+  summarise(Meas_Observed = sum(Meas_Observed)/n(),
+            Meas_Intrinsic = sum(Meas_Intrinsic)/n(),
+            Meas_Remote = sum(Meas_Remote)/n(),
+            Dynamic = sum(Dynamic)/n(),
+            Linear = sum(Linear)/n(),
+            Quadratic = sum(Quadratic)/n(),
+            Interaction = sum(Interaction)/n(),
+            .by = Type) |>
+  mutate(across(where(is.numeric), ~round(.x, 2))) |>
+  mutate(Type = case_when(Type == "E-ANTH" ~ "Anthropogenic",
+                          Type == "E-CLIM" ~ "Climate/Weather",
+                          Type == "E-DIS" ~ "Disaster/Disturbance",
+                          Type == "E-HTAQ" ~ "Habitat (Aquatic)",
+                          Type == "E-HTT" ~ "Habitat (Terrestrial)",
+                          Type == "E-HYD" ~ "Hydrologic",
+                          Type == "E-SEAS" ~ "Phenologic",
+                          Type == "E-TOPO" ~ "Topographic",
+                          Type == "CONN" ~ "Connectivity",
+                          Type == "GEOM" ~ "Site geometry",
+                          Type == "INT" ~ "Biotic interaction",
+                          Type == "SEAS" ~ "Primary occasion",
+                          Type == "SITE" ~ "Site",
+                          Type == "SP" ~ "Species",
+                          Type == "SURV" ~ "Survey detail",
+                          Type == "OTHER" ~ "Other"))|>
+  filter(!is.na(Type) & Type != "Other") |>
+  reactable()
+
+```
+:::
+
+## Complexity
+
+21% of articles considered at least one quadratic term; 26% considered at least one interaction term.
+
+```{r covariates_complexity}
+#| echo: false
+
+covariates_clean |>
+  mutate(Total = length(unique(ID))) |>
+  mutate(Quadratic = case_when(Complexity %in% c("Quadratic + Interaction",
+                                                 "Quadratic") ~ 1,
+                               .default = 0),
+         Interaction = case_when(Complexity %in% c("Quadratic + Interaction",
+                                                   "Interaction") ~ 1,
+                                 .default = 0)) |>
+  select(ID, Total, Quadratic, Interaction) |>
+  distinct() |>
+  summarise(Quadratic = round(sum(Quadratic)/mean(Total), 2),
+            Interaction = round(sum(Interaction)/mean(Total),2)) |>
+  kable()
+```
 
@@
 
+There is no apparent trend in the quantity of covariates over time.
+
+```{r covariates_quantityYearly}
+#| echo: false
+#| message: false
+
+covariates_clean |>
+  filter(Type != "NONE") |>
+  summarise(Covariates = n(), .by = ID_clean) |>
+  mutate(ID_clean = as.numeric(ID_clean)) |>
+  left_join(select(studyDetails_clean, Year, ID), 
+            by = join_by(ID_clean == ID)) |>
+  filter(Covariates < 60) |>
+
+  ggplot(aes(x = Year, y = Covariates)) +
+  geom_point(colour = "darkslategrey", alpha = 0.5) +
+  geom_smooth(method = "lm", colour = "darkslategrey") +
+  labs(title = "Quantities of covariates over time",
+       x = "Year", y = "Total number of covariates")
+```
+
 ## Quantity of covariates by study size
 
+```{r Covariates_studysize}
+#| echo: false
+#| warning: false
+#| message: false
+
+covariates_count <- covariates_clean |>
+  filter(Type != "NONE") |>
+  summarise(Covariates = n(), .by = ID_clean) |>
+  mutate(ID_clean = as.numeric(ID_clean))
+
+dataCollection_clean |>
+  select(ID_clean, Duration, 
+         Primary_Max, Secondary_Max,
+         SiteNumber_Max) |>
+  mutate(ID_clean = as.numeric(ID_clean)) |>
+  rowwise() |>
+  mutate(TotalOccasions = Primary_Max*Secondary_Max) |>
+  left_join(select(studyDetails_clean, ID, YearStrata), by = join_by(ID_clean == ID)) |>
+  filter(!is.na(YearStrata)) |>
+  pivot_longer(cols = c(Duration, TotalOccasions, SiteNumber_Max),
+               names_to = "Variable", values_to = "Quantity") |>
+  left_join(covariates_count, by = "ID_clean") |>
+  mutate(Variable = case_when(Variable == "Duration" ~ "Study duration (months)",
+                              Variable == "SiteNumber_Max" ~ "Number of sites",
+                              Variable == "TotalOccasions" ~ "Total survey occasions (Primary x Secondary)")) |>
+  filter(Covariates < 60) |>
+
+  ggplot(aes(x = Quantity, y = Covariates, colour = Variable)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "lm", se = FALSE, alpha = 0.7) +
+  scale_x_continuous(transform = "log10") +
+  scale_colour_manual(values = c("darkslategrey", "cyan4", "darkorchid4")) +
+  facet_wrap(vars(Variable), ncol = 1, scales = "free_x") +
+  theme(legend.position = "none") +
+  labs(title = "Covariate count by study size",
+       x = "Study size (log scales)", y = "Number of covariates")
+```
+
+## Covariate count by objective
+
+```{r Covariates_objective}
+#| echo: false
+
+objectives_Clean |>
+  select(ID, Objective, Stated) |>
+  filter(Stated == TRUE) |>
+  left_join(covariates_count, by = join_by(ID == ID_clean)) |>
+  filter(Covariates < 60) |>
+
+  ggplot() +
+  geom_boxplot(aes(x = Objective, y = Covariates, colour = Objective)) +
+  theme(legend.position = "none") +
+  labs(title = "Total covariate count by objective",
+       x = "", y = "Number of covariates") + 
+  scale_fill_manual(values = c("darkorchid", "deepskyblue3",
+                               "coral2", "goldenrod", "gray60"))
+```
+
 # Modelling
@@
 
-+---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Field         | Description                                               | Possible values                                                                                |
-+===============+===========================================================+================================================================================================+
+|---------------|-----------------------------------------------------------|------------------------------------------------------------------------------------------------|
 | Bayesian      | Whether the models fit were Bayesian                      | *Logical*                                                                                      |
-+---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Prefiltering  | Which, if any, prefiltering method was used               | *Expert, Simple, Other*                                                                        |
-+---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Selection     | Which, if any, selection method was used                  | *Single a priori, Multiple a priori, Exhaustive, Exhaustive by parameter, Forwards, Backwards* |
-+---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Evaluation    | How models were compared and evaluated                    | *AIC, AUC, DiC, Model-averaging, Bootstrapping*                                                |
-+---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Multi-species | Whether this was a multi-species model                    | *Logical*                                                                                      |
-+---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Connectivity  | Whether site connectivity was incorporated into the model | *Logical*                                                                                      |
-+---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+

### 2024-03-18 Analysis/ReviewSummary/ReviewSummary.qmd

@@
 
++-------------------+--------------------------------------------------+--------------------------------------+
 | Field             | Description                                      | Possible values                      |
-|-------------------|--------------------------------------------------|--------------------------------------|
++===================+==================================================+======================================+
 | Journal           | Journal which this article was published in      | *Any*                                |
++-------------------+--------------------------------------------------+--------------------------------------+
 | Year              | Year of publication                              | *Any*                                |
++-------------------+--------------------------------------------------+--------------------------------------+
 | Author country    | Nation of the first author's primary affiliation | *ISO 3166 country names*             |
++-------------------+--------------------------------------------------+--------------------------------------+
 | Author categories | Categories of affiliations for *all* authors     | *Academic, Government, NGO, Private* |
++-------------------+--------------------------------------------------+--------------------------------------+
 
@@
 
++---------------------+-----------------------------------------------------------------------------------+
 | Objective           | Description                                                                       |
-|---------------------|-----------------------------------------------------------------------------------|
++=====================+===================================================================================+
 | Methods             | Did authors seek to demonstrate or test specific methods for occupancy modelling? |
++---------------------+-----------------------------------------------------------------------------------+
 | Future Predictions  | Did authors make predictions beyond the study period?                             |
++---------------------+-----------------------------------------------------------------------------------+
 | Spatial Predictions | Did authors make predictions at locations where surveys were not conducted?       |
++---------------------+-----------------------------------------------------------------------------------+
 | Hypothesis testing  | Did authors seek to test specific hypotheses of species occupancy dynamics?       |
++---------------------+-----------------------------------------------------------------------------------+
 
@@
 
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Field           | Description                                                                                | Possible values                                                                   |
-|-----------------|--------------------------------------------------------------------------------------------|-----------------------------------------------------------------------------------|
++=================+============================================================================================+===================================================================================+
 | Taxa category   | Category of study taxa; could fit into multiple categories                                 | *Bird, Mammal, Invertebrate, Herptile, Fish, Threatened, Invasive, Other*         |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Taxa number     | Number of individual species within the taxa for which models were fit                     | *Numeric*                                                                         |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | IUCN status     | IUCN status for the study species; NA if multi-species or if the study species is invasive | *LC, NT, EN, CR, DD, NA*                                                          |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Threat category | Category of threats to the species; multiple per species permitted                         | *Habitat loss, Climate change, Exploitation, Invasives, Disease, Disaster, Other* |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Country         | Country/countries where the study took place                                               | *ISO 3166 country names*                                                          |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Continent       | Continent(s) where the study took place                                                    | *North America, South America, Europe, Asia, Africa, Oceania*                     |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Realm           | Biogeographic realm where the study took place                                             | *Nearctic, Palearctic, Afrotropical, Indomalayan, Australasian, Neotropical*      |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 | Size            | Size of the study area extent in square kilometres                                         | *\<1km, 1-10, 10-100, 100-1000, 10000-100000, 100000-1000000, 1000000+*           |
++-----------------+--------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------+
 
@@
 
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Field               | Description                                                              | Possible values                                                                              |
-|---------------------|--------------------------------------------------------------------------|----------------------------------------------------------------------------------------------|
++=====================+==========================================================================+==============================================================================================+
 | Duration            | Length of study in months                                                | *Numeric*                                                                                    |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Capture method      | Method of recording detections; multiple categories possible             | *Audiovisual, Search, Trapping, Bioacoustic, Camera, Citizen, Incidental*                    |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Primary occasions   | Number of primary occasions                                              | *Numeric*                                                                                    |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Secondary occasions | Number of secondary occasions, presented as a range where not consistent | *Numeric range*                                                                              |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Secondary type      | Type of secondary occasion                                               | *Revisits, Spatial replicates, Same-visit replicates, multiple observers, time to detection* |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Type of site        | How a 'site' was defined                                                 | *Point location, Grid cell, Irregular area, Transect*                                        |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Number of sites     | Number of individual sites                                               | Numeric                                                                                      |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 | Sampling method     | How sites were selected                                                  | *Random, Stratified, Grid, Preferential, Available, Exhaustive, Uniform, Transect*           |
++---------------------+--------------------------------------------------------------------------+----------------------------------------------------------------------------------------------+
 
@@
 
++---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Field         | Description                                               | Possible values                                                                                |
-|---------------|-----------------------------------------------------------|------------------------------------------------------------------------------------------------|
++===============+===========================================================+================================================================================================+
 | Bayesian      | Whether the models fit were Bayesian                      | *Logical*                                                                                      |
++---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Prefiltering  | Which, if any, prefiltering method was used               | *Expert, Simple, Other*                                                                        |
++---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Selection     | Which, if any, selection method was used                  | *Single a priori, Multiple a priori, Exhaustive, Exhaustive by parameter, Forwards, Backwards* |
++---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Evaluation    | How models were compared and evaluated                    | *AIC, AUC, DiC, Model-averaging, Bootstrapping*                                                |
++---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Multi-species | Whether this was a multi-species model                    | *Logical*                                                                                      |
++---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
 | Connectivity  | Whether site connectivity was incorporated into the model | *Logical*                                                                                      |
++---------------+-----------------------------------------------------------+------------------------------------------------------------------------------------------------+
+
+```{r modelling_clean}
+#| echo: false
+
+modelling_clean <- review_Modelling |>
+  filter(Status == "Complete")
+```
+
+## Selection methods
+
+```{r modelling_selection}
+#| echo: false
+
+modelling_prefilter <- modelling_clean |>
+  mutate(Total = length(unique(ID))) |>
+  summarise(Articles = n(), 
+            .by = c(Prefiltering, Total)) |>
+  rename(Method = Prefiltering) |>
+  mutate(Class = "Prefilter")
+
+modelling_clean |> 
+  mutate(Total = length(unique(ID))) |>
+  summarise(Articles = n(), 
+            .by = c(Selection, Total)) |>
+  rename(Method = Selection) |>
+  mutate(Class = "Select") |>
+  bind_rows(modelling_prefilter) |>
+  mutate(Percentage = round(Articles/Total, 2)*100) |>
+  filter(!Method %in% c("None", "APriori_m, Stepwise", "Other")) |>
+  mutate(Method = case_when(Method %in% c("Apriori_s", "APriori_s") ~ "A priori (Single)",
+                            Method == "APriori_m" ~ "A priori (Multiple)",
+                            Method == "Backwards" ~ "Stepwise (Backwards)",
+                            Method == "Forwards" ~ "Stepwise (Forwards)",
+                            Method == "Exhaustive_all" ~ "Exhaustive (True)",
+                            Method == "Exhaustive_p" ~ "Exhaustive (By parameter)",
+                            .default = Method)) |>
+  summarise(Percentage = sum(Percentage),
+            .by = c(Method, Class)) |>
+  arrange(Class, Percentage) |>
+  mutate(Method = fct_inorder(Method)) |>
+
+  ggplot() +
+  geom_col(aes(x = Percentage, y = Method, fill = Class)) +
+  scale_fill_manual(limits = c("Prefilter", "Select"),
+                    values = c("darkorange3", "darkslategrey")) +
+  labs(title = "Covariate selection method frequency", 
+       x = "Percentage of articles", y = "") +
+  theme(legend.position = "bottom")
+```
+
+## Model selection trends
+
+```{r Modelling_trends}
+#| echo: false
+#| warning: false
+
+modelling_clean |>
+  select(ID, Selection) |>
+  mutate(ID = as.numeric(ID)) |>
+  left_join(select(studyDetails_clean, ID, YearStrata),
+            by = "ID") |>
+  filter(!is.na(YearStrata)) |>
+  mutate(Selection = case_when(Selection %in% c("APriori_s",
+                                                "Apriori_s") ~ "A priori (Single)",
+                               Selection == "APriori_m" ~ "A priori (Multiple)",
+
+                               .default = "Other")) |>
+  mutate(Total = n(), .by = YearStrata) |>
+  summarise(Articles = n(), .by = c(YearStrata, Selection, Total)) |>
+  mutate(Proportion = Articles/Total) |>
+
+  ggplot() +
+  geom_col(aes(x = YearStrata, y = Articles, fill = Selection)) +
+  scale_fill_manual(values = c("goldenrod", "darkorange3", "darkorchid4")) +
+  theme(legend.position = "bottom") +
+  labs(title = "Covariate selection trends",
+       x = "Year Strata", 
+       y = "Number of articles")
+
+```
+
+## Frequentist vs Bayesian
+
+```{r modelling_frequentist}
+#| echo: false
+#| warning: false
+
+modelling_clean |>
+  mutate(Class = case_when(Bayesian == "Y" ~ "Bayesian",
+                           Bayesian == "N" ~ "Frequentist")) |>
+  mutate(ID = as.numeric(ID)) |>
+  left_join(select(studyDetails_clean, ID, YearStrata),
+            by = "ID") |>
+  filter(!is.na(YearStrata),
+         !is.na(Class)) |>
+  summarise(Articles = n(),
+            .by = c(YearStrata, Class)) |>
+
+  ggplot() +
+  geom_col(aes(x = YearStrata, y = Articles, fill = Class)) +
+  scale_fill_manual(values = c("cyan4", "darkorchid3")) +
+  theme(legend.position = "bottom") +
+  labs(title = "Model type trends",
+       x = "Year Strata", 
+       y = "Number of articles")
+
+```

### 2024-03-19 Manuscript/domReview.qmd

---
title: Dynamic occupancy models review
date: today
authors:
  - name: Saoirse Kelleher
    affiliation: University of Melbourne
    email: saoirse.kelleher@student.unimelb.edu.au
    orcid: 0000-0002-4614-5753
    corresponding: true
  - name: Natalie Briscoe
    affiliation: University of Melbourne
  - name: Gurutzeta Guillera-Arroita
    affiliation: Consejo Superior de Investigaciones Científicas (CSIC)
  - name: Jane Elith
    affiliation: University of Melbourne

abstract: >
  Dynamic occupancy models allow researchers to estimate species occupancy across space and time while accounting for imperfect detection.

bibliography: references.bib
embed-resources: true
---

## Introduction

## Methods

## Results

## Discussion

## Conclusions

### 2024-03-22 Manuscript/domReview.qmd
Abstract/intro content
First stab at basic overview sections

@@
 ---
-title: Dynamic occupancy models review
+title: "Twenty years of dynamic occupancy modelling – what’s been done, and where to next?"
 date: today
@@
   - name: Gurutzeta Guillera-Arroita
-    affiliation: Consejo Superior de Investigaciones Científicas (CSIC)
+    affiliation: Pyrenean Institute of Ecology, Spanish National Research Council
   - name: Jane Elith
@@
 
-abstract: >
-  Dynamic occupancy models allow researchers to estimate species occupancy across space and time while accounting for imperfect detection.
+abstract: |
+  In the twenty years since their introduction, the dynamic occupancy model has become an important tool for describing trends in species occupancy across landscapes and throughout time while accounting for imperfect detection.
+
+  We present a retrospective of the first two decades of dynamic occupancy modelling based on the results of a systematic review of models fit to field data.
+We describe how authors have used these models to describe patterns of occupancy for a diverse range of study systems and focal taxa, dedicating particular emphasis to how authors engaged with modelling decisions such as covariate selection and model evaluation.
+Modelling approaches vary considerably within the literature, and objective metrics of model performance are rarely estimated.
+
+  From these results we consider what the future of dynamic occupancy modelling may hold.
+Key questions remain on how the modelling process influences parameter estimates and model performance.
+Considerably more emphasis must be placed on evaluating models via processes such as out-of-sample validation.
+Further opportunities also abound; newly proliferating data types such as bioacoustics and camera traps present new questions around model design as well as new ways to study occupancy at scales which have not been historically possible.
+The suitability of dynamic occupancy models to generating predictions, as an alternative to static models, is another area which has been historically under-emphasised.
 
@@
 
-## Introduction
+# Introduction
 
-## Methods
+The dynamic implementation of the occupancy model was introduced two decades ago in @mackenzie2003.
+Since then, the model has become a popular tool for ecologists seeking to describe occupancy dynamics while accounting for imperfect detection, generating hundreds of articles describing model uses and inspiring popular software tools for their implementation [@kery2016].
 
-## Results
+The dynamic occupancy model (DOM) fills an important niche in the ecological modelling landscape as an approachable midpoint between purely correlative species distributions models and more complex process-based models.
+Data input requirements are achievable for many research programs, and estimates of occupancy dynamic parameters (i.e., colonisation and extinction) are of use in addressing many key ecological questions.
+Furthermore, the capacity to control for imperfect detection addresses a ubiquitous issue with field ecological data.
+A description of the basic model form is presented in **Box 1.**
 
-## Discussion
+When implementing DOMs, authors must make several key decisions during the modelling process.
+The duration of primary and secondary sampling occasions has implications both in describing the frame of reference for model outputs as well as in determining whether the closure assumption is fulfilled, and numerous authors have described the importance of this decision.
+
+The selection of covariates has been discussed in considerably less detail.
+Each of the four model parameters (Initial occupancy, colonisation, extinction, and detection) can accept covariates to account for heterogeneity.
+The determination of which covariates to include can be challenging.
+
+*INTRO OUTLINE*
+
+drive the past v future angle; how have they been built&used and how can they be in the future?
+
+-   DOMs were developed 20 years ago, they are popular and useful
+
+-   What is a DOM? \[put in figures\]
+
+-   Key assumptions
+
+-   Modelling decisions as they relate to the key assumptions
+
+-   Closure assumption & primary/secondary occasions
+
+-   Heterogeneity and covariates
+
+-   Which covariates should we use?
+
+-   How do we select covariates?
+
+-   & how do we evaluate our models?
+
+-
+
+-   We review articles which fit DOMs
+
+-   Emphasised model building process
+
+-   Synthesis on what further research needs to be done
+
+-   what new potential opportunities exist?
+
+# Methods
+
+# Results
+
+# Discussion
+
+## Use cases
+
+### Observing trends
+
+### Hypothesis testing
+
+### Generating predictions
+
+## Modelling
+
+### Covariates
+
+#### Covariate types
+
+#### Complexity
+
+### Model selection
+
+#### Bayesian implementation
+
+#### Frequentist implementations
+
+### Model evaluation
+
+###
+
+## Future opportunities
+
+### Data types
+
+### Prediction
 

### 2024-03-22 Manuscript/domReview.qmd
progress on intro
intro progress

@@
 
+Over the past few decades the dynamic occupancy model and its extensions have been used to address a broad range of research questions and objectives.
+Authors have assessed patterns and trends of occupancy for threatened species, tested specific hypotheses on the drivers of occupancy, and generated predictions of occupancy under future conditions.
+These models are well-suited to non-equilibrium situations such as recovering, declining, or invasives species, where assumptions of conventional species distribution models cannot be met.
+
+Dynamic occupancy models have three fundamental assumptions; i) that sites are closed to changes in occupancy between primary sampling occasions, ii) that all heterogeneity in occupancy, occupancy dynamics, and detection is accounted for by covariates, and iii) that detections are independent from one another.
+Much has been discussed with respect to the first assumption, the closure assumption.
+Considerably less has been said regarding heterogeneity and the implications of inadequately modelling it.
+
 When implementing DOMs, authors must make several key decisions during the modelling process.
@@
 The determination of which covariates to include can be challenging.
+
+We present a comprehensive overview of how the dynamic occupancy model has been applied in its first two decades, with a look to the key questions which remain on how DOM can best be used to address ecological questions.
 

### 2024-03-25 Manuscript/domReview.qmd
Restructure
Swapped to a non-standard outline format, away from methods-results-discussion

@@
 
-The dynamic implementation of the occupancy model was introduced two decades ago in @mackenzie2003.
-Since then, the model has become a popular tool for ecologists seeking to describe occupancy dynamics while accounting for imperfect detection, generating hundreds of articles describing model uses and inspiring popular software tools for their implementation [@kery2016].
-
-The dynamic occupancy model (DOM) fills an important niche in the ecological modelling landscape as an approachable midpoint between purely correlative species distributions models and more complex process-based models.
-Data input requirements are achievable for many research programs, and estimates of occupancy dynamic parameters (i.e., colonisation and extinction) are of use in addressing many key ecological questions.
-Furthermore, the capacity to control for imperfect detection addresses a ubiquitous issue with field ecological data.
-A description of the basic model form is presented in **Box 1.**
-
 Over the past few decades the dynamic occupancy model and its extensions have been used to address a broad range of research questions and objectives.
@@
 
-*INTRO OUTLINE*
+# Introduction
 
-drive the past v future angle; how have they been built&used and how can they be in the future?
+The dynamic implementation of the occupancy model was introduced two decades ago in @mackenzie2003.
+Since then, the model has become a popular tool for ecologists seeking to describe occupancy dynamics while accounting for imperfect detection, generating hundreds of articles describing model uses and inspiring popular software tools for their implementation [@kery2016].
 
--   DOMs were developed 20 years ago, they are popular and useful
+The dynamic occupancy model (DOM) fills an important niche in the ecological modelling landscape as an approachable midpoint between purely correlative species distributions models and more complex process-based models.
+Data input requirements are achievable for many research programs, and estimates of occupancy dynamic parameters (i.e., colonisation and extinction) are of use in addressing many key ecological questions.
+Furthermore, the capacity to control for imperfect detection addresses a ubiquitous source of bias and uncertainty in field ecological data.
+A description of the basic model form is presented in **Box 1.**
 
--   What is a DOM? \[put in figures\]
+## How are they used
 
--   Key assumptions
+Outputs of dynamic occupancy model may be put to many uses.
+In many cases, simple estimates of the ecological parameters (i.e.
+occupancy, colonisation, or extinction) can aid in describing understanding changes in range dynamics.
+In other cases, estimates of covariates relating to these parameters are used to test hypotheses, for example, the influence of forest fire on the probability of local extinction.
+Less frequently, DOMs are used to make predictions of site occupancy either across unsurveyed sites or into the future under predictions of future conditions or management actions.
 
--   Modelling decisions as they relate to the key assumptions
+## What does this article do
 
--   Closure assumption & primary/secondary occasions
+We present a comprehensive overview of how the dynamic occupancy model has been applied in its first two decades, with a look to the key questions which remain on how DOM can best be used to address ecological questions.
 
--   Heterogeneity and covariates
+# History of the dynamic occupancy model
 
--   Which covariates should we use?
+## Precursors
 
--   How do we select covariates?
+## Variants
 
--   & how do we evaluate our models?
+## Model extensions
 
--
+# Review of published implementations
 
--   We review articles which fit DOMs
+## Systematic review methodology
 
--   Emphasised model building process
+## Study systems
 
--   Synthesis on what further research needs to be done
+## Data collection
 
--   what new potential opportunities exist?
+## Objectives and authorship
 
-# Methods
+## Model development and evaluation
 
-# Results
+# Future priorities for dynamic occupancy models
 
-# Discussion
+## Model selection and evaluation
 
-## Use cases
+## Continuous detection data
 
-### Observing trends
-
-### Hypothesis testing
-
-### Generating predictions
-
-## Modelling
-
-### Covariates
-
-#### Covariate types
-
-#### Complexity
-
-### Model selection
-
-#### Bayesian implementation
-
-#### Frequentist implementations
-
-### Model evaluation
-
-###
-
-## Future opportunities
-
-### Data types
-
-### Prediction
-
-## Conclusions
+## Predictions

### 2024-03-25 Manuscript/domReview.qmd
History section
framework for the history of occupancy models

@@
 
-Over the past few decades the dynamic occupancy model and its extensions have been used to address a broad range of research questions and objectives.
-Authors have assessed patterns and trends of occupancy for threatened species, tested specific hypotheses on the drivers of occupancy, and generated predictions of occupancy under future conditions.
-These models are well-suited to non-equilibrium situations such as recovering, declining, or invasives species, where assumptions of conventional species distribution models cannot be met.
-
-Dynamic occupancy models have three fundamental assumptions; i) that sites are closed to changes in occupancy between primary sampling occasions, ii) that all heterogeneity in occupancy, occupancy dynamics, and detection is accounted for by covariates, and iii) that detections are independent from one another.
-Much has been discussed with respect to the first assumption, the closure assumption.
-Considerably less has been said regarding heterogeneity and the implications of inadequately modelling it.
-
-When implementing DOMs, authors must make several key decisions during the modelling process.
-The duration of primary and secondary sampling occasions has implications both in describing the frame of reference for model outputs as well as in determining whether the closure assumption is fulfilled, and numerous authors have described the importance of this decision.
-
-The selection of covariates has been discussed in considerably less detail.
-Each of the four model parameters (Initial occupancy, colonisation, extinction, and detection) can accept covariates to account for heterogeneity.
-The determination of which covariates to include can be challenging.
-
-We present a comprehensive overview of how the dynamic occupancy model has been applied in its first two decades, with a look to the key questions which remain on how DOM can best be used to address ecological questions.
-
-# Introduction
-
 The dynamic implementation of the occupancy model was introduced two decades ago in @mackenzie2003.
@@
 
+These models are particularly well-suited to non-equilibrium situations such as recovering, declining, or invasives species, where assumptions of conventional species distribution models cannot be met.
+
 ## What does this article do
@@
 
+Dynamic occupancy models originate in the mask-recapture model literature with roots in the robust sampling design, which is also used in the Jolly-Seber and other related models.
+In the case of the dynamic occupancy model the unit of inference is the site, not the individual.
+
+DOMs bear similarities to the Incidence-Function or Hanksi models frequently used in the study of island biogeography, with similar probabilities of colonisation and extinction; DOMs are typically not spatially-explicit, however.
+
 ## Variants
 
+Various parameterisations of the standard dynamic occupancy have been described, the most frequent being those described by MacKenzie and Nichols.
+The former estimates probabilities of initial occupancy, colonisation, extinction, and detection and derives probabilities of yearly occupancy.
+The latter directly estimates probabilities of occupancy, local extinction, and detection; colonisation is then estimated as a derived parameter.
+
+The Bayesian implementation of the dynamic occupancy model was described by Nichols.
+
 ## Model extensions
+
+The dynamic occupancy model has been extended numerous times, to make it suitable for multiple scales, multiple species, and to accommodate for false positives.
+An in-depth discussion of model extensions is presented in Bailey 2014.
 

### 2024-03-25 Manuscript/domReview.qmd
vague outline
added vague outline details for key sections

@@
 
+We conducted a systematic review of articles fitting the dynamic occupancy model to ecological field data.
+Our initial focus was on a set of four related terms: "dynamic occupancy model", "occupancy dynamics model," "multi-season occupancy model," and "incidence function model." For each of these, the first 100 articles listed on Google Scholar were downloaded.
+For each combination of search term and time strata (2000-2005, 2006-2010, 2011-2015, 2016-2021), a random sample of 5 articles or 25% of articles (whichever was greater) was selected for analysis.
+Inclusion criteria were as follows:
+
+-   Models were fit on field-collected ecological data.
+
+-   Models estimated site occupancy and detection probability over multiple time-steps.
+
+-   Models were Markovian, in that occupancy of one year was conditioned on the previous year.
+
+Articles which did not meet the criteria were removed and replaced by another article from the same strata where available.
+Very few papers from the "incidence-function model" or "occupancy dynamics model" fit the search criteria, and authors deemed that consensus in the names of this model is low and that additional papers were required.
+A second sample was drawn from articles citing MacKenzie 2003; 15 papers from each aforementioned time strata as well as 15 additional papers from 2022-2023 were analysed.
+Articles which did not fit the criteria were again replaced from the same strata.
+
+**\[figure 1: review diagram\]**
+
 ## Study systems
+
+Dynamic occupancy models are flexible models suitable across spatial scales.
+Authors have fit them to data from areas as small as a single square kilometre and as large as entire continents.
+Interpretation of occupancy does shift depending on the spatial scale and density of sites.
+
+DOMs are taxonomically diverse. Many studies have focused on threatened species and invasive taxa.
+
+Many DOMs have fit more than a single species. Some of these are independent single species models, some are hierarchical joint estimations, and some are interactive models.
+
+**\[figure 2: diversity in study systems\]**
 
@@
 
+Survey design is considerably varied. The shortest data sets were for under a month, the longest were 40 years.
+
+There were multiple types of secondary occasions. Most common were revisits, but others have used time to detection, multiple observers, or spatial replicates.
+
+Most data was collected in either structured or unstructured data. Some models incorporated citizen science data effectively.
+
+A few papers used autonomous detection data, like cameras and bioacoustics.
+
+## Model development and evaluation
+
+Model complexity varied quite a bit. Many had low numbers of parameters - particularly on ecological parameters.
+
+Selection was similarly rare. Lots of models were defined a-priori, without structured model selection.
+
+Despite lots of data, model evaluation was quite rare. There is no goodness of fit test, but what about out of sample validation.
+
 ## Objectives and authorship
 
-## Model development and evaluation
+Four objective categories were keyed out.
+
+Hypothesis testing and parameter estimation was common.
+
+Future predictions and spatial mapping was fairly rare.
 
@@
 
+Unclear how much model selection matters.
+
+Testing hypotheses is not an excuse for not doing model selection.
+
+Can model evaluation be better reported, especially since many studies have so much data?
+
 ## Continuous detection data
 
+Autonomous data has lots of opportunities. Offers more choice in selecting primary and secondary sampling occasions; also requires more thought to disentangle short and long term occupancy.
+
 ## Predictions
+
+Predictions were rare, but they're a big opportunity - particularly for SDM users.
+DOMs may better account for non-equilibrium states, plus they handle imperfect detection.
+There have been some excellent examples.

### 2024-04-12 Analysis/ReviewFigures/ReviewFigures.qmd

---
title: "DOM Review Stats and Figures"
author: "Saoirse Kelleher"
date: today
abstract: "Compiles key statistics and figures for use in the DOM Review manuscript"
abstract-title: "Summary"
format: 
  html:
    theme: minty
    toc: true
embed-resources: true
---

```{r Load_packages}
#| include: false

library(tidyverse)
library(readxl)
library(knitr)
library(reactable)
library(sf)
```

```{r Load data}
#| include: false

review_Data <- read_xlsx("Spreadsheets/Manuscript_Spreadsheet.xlsx")
covariate_Data <- read_xlsx("Spreadsheets/Manuscript_Spreadsheet.xlsx", sheet = 2)
```

## Survey locations map

```{r locations map}

countries_shp <- st_read("Analysis/SpatialData/Countries/World_Countries__Generalized_.shp") |>
  st_transform(crs = "ESRI:54030")

countryCount <- review_Data |>
  separate_longer_delim(StudyCountry, ", ") |>
  mutate(StudyCountry = case_when(StudyCountry == "United States of America" ~ "United States",
                                  .default = StudyCountry)) |>
  summarise(Articles = n(), .by = StudyCountry) |>
  select(COUNTRY = StudyCountry, Articles) |>
  mutate(Articles = case_when(Articles == 1 ~ "1", 
                              Articles == 2 ~ "2", 
                              Articles == 3 ~ "3", 
                              Articles == 4 ~ "4", 
                              Articles == 46 ~ "46")) |>
  mutate(Articles = fct(Articles, 
                        levels = c("1", "2", "3", "4", "46")))

countries_map <- countries_shp |>
  left_join(countryCount, by = "COUNTRY") |> 
  filter(COUNTRY != "Antarctica") |>

  ggplot() +
  geom_sf(aes(fill = Articles), colour = "white", linewidth = 0.01) +
  scale_fill_brewer(palette = "PuBuGn", na.value = "gray40") +
  guides(fill = guide_legend(title = "Number of Articles", nrow = 1)) +
  theme_dark() +
  theme(legend.position = "bottom",
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text = element_blank(), axis.ticks = element_blank())

ggsave("Manuscript/Figures/StudiesMap.jpeg", countries_map)
```

## Review coverage

```{r coverage}

sampleCoverage <- read_xlsx("Randomisation/SecondSampleRanks.xlsx") |>
  summarise(n = n(),
            .by = c(Group, Status)) |>
  mutate(Status = case_when(is.na(Status) ~ "Unread",
                            .default = Status)) |>
  pivot_wider(names_from = Status, values_from = n) |>
  mutate(HitRate = Accept/(Accept+Reject)) |>
  mutate(EstRemaining = Unread*HitRate)

sampleCoveragePlot <- sampleCoverage |>
  pivot_longer(c(Accept, EstRemaining),
               names_to = "Category", values_to = "Articles") |>
  mutate(Group = case_when(Group == "MacKenzie Cited_2000:2005" ~ "2000-2005",
                           Group == "MacKenzie Cited_2006:2010" ~ "2006-2010",
                           Group == "MacKenzie Cited_2011:2015" ~ "2011-2015",
                           Group == "MacKenzie Cited_2016:2021" ~ "2016-2021",
                           Group == "MacKenzie Cited_2022:2023" ~ "2022-2023")) |>
  mutate(Group = fct(Group, c("2000-2005", 
                              "2006-2010",
                              "2011-2015",
                              "2016-2021",
                              "2022-2023"))) |>

  ggplot() +
  geom_col(aes(x = Group, y = Articles, fill = Category),
           colour = "white", width = 1,
           position = position_stack(reverse = TRUE)) +
  scale_fill_manual(limits = c("Accept", "EstRemaining"),
                    values = c("turquoise4", "gray70"),
                    labels = c("Processed", "Estimated")) +
  scale_y_continuous(limits = c(0, 250), 
                     expand = c(0,0)) +
  labs(title = "Estimated published articles fitting Dynamic Occupancy Models",
       x = "", y = "") +
  theme(legend.position = "top",
        panel.grid = element_blank(),
        title = element_text(size = 14),
        axis.text = element_text(size = 13),
        legend.text = element_text(size = 13))

ggsave("Manuscript/Figures/CoveragePlot.jpeg", sampleCoveragePlot,
       width = 9, height = 4)
```

## Focal species bar plot

```{r Focal species bar plot}

speciesCounts <- review_Data |>
  select(Taxa, Invasive, Threatened) |>
  separate_longer_delim(Taxa, ", ") |>
  mutate(Taxa = case_when(Taxa == "Lichen" ~ "Non-animal",
                          .default = Taxa)) |>
  mutate(Invasive = case_when(Invasive == "Y" ~ 1,
                              Invasive == "N" ~ 0),
         Threatened = case_when(Threatened == "Y" ~ 1,
                                Threatened == "N" ~ 0),
         Neither = case_when(Invasive == 0 & Threatened == 0 ~ 1,
                             .default = 0)) |>
  summarise(Invasive = sum(Invasive, na.rm = TRUE),
            Threatened = sum(Threatened, na.rm = TRUE),
            Neither = sum(Neither, na.rm = TRUE),
            .by = Taxa) |>
  pivot_longer(c(Invasive, Threatened, Neither), 
               names_to = "Category",
               values_to = "Values") |>
  mutate(Taxa = fct_reorder(Taxa, Values, .fun = sum)) |>
  mutate(Category = fct(Category, 
                        levels = c("Neither", "Invasive", "Threatened")))

speciesPlot <- speciesCounts |>
  ggplot() +
  geom_col(aes(y = Taxa, x = Values, fill = Category)) +
  scale_fill_manual(values = c("darkslategray", "goldenrod", "firebrick4")) +
  labs(x = "", y = "") +
  scale_x_continuous(expand = c(0, 1)) +
  theme(legend.position = "top",
        axis.text.y = element_text(size = 15),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(size = 10), 
        panel.grid = element_blank())

ggsave("Manuscript/Figures/SpeciesPlot.jpeg", speciesPlot, width = 7, height = 5)
```

## Taxa quantity statistics

```{r Taxa quantity statistics}

review_Data |>
  summarise(n(), .by = Quantity)
```

## Survey method statistics

```{r Survey method statistics}

review_Data |>
  separate_longer_delim(Survey, ", ") |>
  summarise(n(), .by = Survey)
```

## Survey size figures

```{r Study area size}

studyArea <- review_Data |>
  summarise(Articles = n(), .by = StudyArea) |>
  filter(!StudyArea %in% c("na", "NA")) |>
  mutate(StudyArea = fct(StudyArea,
                         levels = c("0-1", "1-10", "10-100", 
                                    "100-1000", "1000-10000", "10000-100000", 
                                    "100000-1000000", "1000000+")))

  studyAreaPlot <- ggplot(studyArea) +
    geom_col(aes(y = StudyArea, x = Articles), fill = "darkslategray") +
    labs(title = "Study Area (square kilometres)", x = "Articles", y = "") +
    scale_x_reverse(expand = c(0,0)) +
    scale_y_discrete(position = "right") +
    theme(panel.grid = element_blank(),
          axis.text = element_text(size = 15), 
          axis.title = element_text(size = 14),
          title = element_text(size = 15), 
          axis.ticks.y = element_blank())

  ggsave("Manuscript/Figures/StudyAreaPlot.jpeg", studyAreaPlot)

```

```{r Duration fig}

durationPlot <- ggplot(review_Data) +
  geom_histogram(aes(x = Duration), binwidth = 0.2, fill = "turquoise4") +
  geom_vline(aes(xintercept = median(Duration)), 
             colour = "gray30", linetype = 2, linewidth = 3) +
  scale_x_continuous(transform = "log10", 
                     breaks = c(1, 12, 60, 
                                120, 240, 480), 
                     labels = c("1 month", "1 year", "5 years", 
                                "10 years", "20 years", "40 years")) +
  scale_y_continuous(expand = c(0, 0)) + 
  labs(title = "Study duration", x = "", y = "") +
  theme(panel.grid = element_blank(),
        title = element_text(size = 13), 
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(size = 16, angle = 70, 
                                   hjust = 1, vjust = 1))
ggsave("Manuscript/Figures/Duration.jpeg", durationPlot, 
       height = 4, width = 7)
```

```{r Sites fig}
sitesPlot <- review_Data |>
  mutate(Sites = as.numeric(Sites)) |>

  ggplot() +
  geom_histogram(aes(x = Sites), binwidth = 0.1, fill = "turquoise4") +
  geom_vline(aes(xintercept = median(Sites, na.rm = TRUE)), 
             colour = "gray30", linetype = 2, linewidth = 3) +
  scale_x_continuous(transform = "log10",
                     breaks = c(5, 10, 25, 50, 100, 500, 1000, 10000)) +
  scale_y_continuous(expand = c(0, 0)) + 
  labs(title = "Number of sites", x = "", y = "") +
  theme(panel.grid = element_blank(),
        title = element_text(size = 13), 
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(size = 16, angle = 70, 
                                   hjust = 1, vjust = 1))
ggsave("Manuscript/Figures/Sites.jpeg", sitesPlot, 
       height = 4, width = 7)
```

## Covariates

```{r process covs}

covariates_clean <- covariate_Data |>
  mutate(Covariates = str_replace(Covariates, "\\.", "\\,")) |>
  separate_rows(Covariates, sep = ", ") |>
  mutate(Covariates = str_to_upper(Covariates)) |>
  separate_wider_delim(cols = Covariates, delim = "_",
                       names = c("Type", 
                                 "Variation",
                                 "Measurement",
                                 "Complexity"),
                       too_few = "align_start", 
                       too_many = "drop") |>
  mutate(Type = case_when(Type %in% c("E-HTT", " E-HTT") ~ "E-HTT",
                          Type %in% c("E-DIS", "DIS") ~ "E-DIS",
                          Type %in% c("ANTH", "E-ANTH") ~ "E-ANTH",
                          Type %in% c("SP", "SPEC") ~ "SP",
                          Type %in% c("INT", " INT") ~ "INT",
                          Type %in% c("NONE", "NA") ~ "NONE",
                          .default = Type),
         Variation = case_when(Variation %in% c("SY", "YS") ~ "Site-Year",
                               Variation == "S" ~ "Site",
                               Variation == "YSS" ~ "Site-Year-Survey",
                               Variation == "Y" ~ "Year",
                               Variation == "SP" ~ "Species",
                               Variation == "U" ~ "Unclear"),
         Measurement = case_when(Measurement == "I" ~ "Intrinsic",
                                 Measurement == "O" ~ "Observed",
                                 Measurement == "R" ~ "Remote",
                                 Measurement == "U" ~ "Unclear"),
         Complexity = case_when(Complexity %in% c("N", "NH", "N,") ~ "Linear",
                                Complexity == "Q" ~ "Quadratic",
                                Complexity == "I" ~ "Interaction",
                                Complexity == "B" ~ "Quadratic + Interaction",
                                Complexity == "O" ~ "Other"))
```

```{r Covs by param}

covParamPlot <- covariates_clean |>
  mutate(Type = case_when(Type == "E-ANTH" ~ "Anthropogenic",
                          Type == "E-CLIM" ~ "Climate/Weather",
                          Type == "E-DIS" ~ "Disaster/Disturbance",
                          Type == "E-HTAQ" ~ "Habitat (Aquatic)",
                          Type == "E-HTT" ~ "Habitat (Terrestrial)",
                          Type == "E-HYD" ~ "Hydrologic",
                          Type == "E-SEAS" ~ "Phenologic",
                          Type == "E-TOPO" ~ "Topographic",
                          Type == "CONN" ~ "Connectivity",
                          Type == "GEOM" ~ "Site geometry",
                          Type == "INT" ~ "Biotic interaction",
                          Type == "SEAS" ~ "Primary occasion",
                          Type == "SITE" ~ "Site",
                          Type == "SP" ~ "Species",
                          Type == "SURV" ~ "Survey detail",
                          .default = Type)) |>
  mutate(Type_cat = case_when(Type %in% c("Habitat (Terrestrial)",
                                          "Habitat (Aquatic)",
                                          "Hydrologic", 
                                          "Climate/Weather",
                                          "Disaster/Disturbance",
                                          "Topographic",
                                          "Phenologic",
                                          "Anthropogenic",
                                          "Biotic interaction") ~ "Env",
                              .default = "Other")) |>
  summarise(Covariates = case_when("NONE" %in% Type ~ 0,
                                   .default = n()), 
            .by = c(ID, Parameter, Type_cat)) |>
  filter(Parameter != "Occupancy") |>
  complete(nesting(ID, Parameter), 
           Type_cat, fill = list(Covariates = 0)) |>
  mutate(Parameter = fct(Parameter, 
                         levels = c("InitialOccupancy",
                                    "Colonisation", "Extinction",
                                    "Detection"))) |>

  ggplot() +
  geom_boxplot(aes(x = Covariates, y = Type_cat, 
                   colour = Type_cat, fill = Type_cat),
               alpha = 0.5, outliers = FALSE, staplewidth = 0.5) +
  facet_grid(rows = vars(Parameter), switch = "y") +
  scale_colour_manual("Covariate type",
                      values = c("seagreen4", "darkslategray"),
                      labels = c("Environmental", "Other")) +
  scale_fill_manual("Covariate type",
                      values = c("seagreen4", "darkslategray"),
                      labels = c("Environmental", "Other")) +
  labs(title = "Covariate quantity by parameter", 
       x = "Number of covariates considered", y = "") +
  theme(strip.text.y.left = element_text(angle = 0, size = 14, hjust = 1),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        legend.position = "bottom",
        title = element_text(size = 13),
        axis.title.x = element_text(size = 13),
        legend.text = element_text(size = 13))

ggsave("Manuscript/Figures/CovParamPlot.jpeg", covParamPlot,
       width = 7, height = 5)
```

```{r Covs by objective}

covariates_clean |>
  summarise(Covariates = case_when("NONE" %in% Type ~ 0,
                                   .default = n()), 
            .by = c(ID, Parameter)) |>
  summarise(Covariates = sum(Covariates),
            .by = ID) |>
  left_join(review_Data, by = "ID") |>
  mutate(Predict = if_else(Spatial == "Y" | Future == "Y",
                           "Y", "N")) |>
  pivot_longer(c(Hypothesis, Methods, Parameters, Predict),
               names_to = "Objective", values_to = "ObjIncluded") |>
  filter(ObjIncluded == "Y") |>
  summarise(MedianCovs = median(Covariates), n = n(),
            .by = Objective)

```

```{r percent polynomial}

covariates_clean |>
  mutate(Polynomial = case_when(Complexity %in% c("Quadratic", "Quadratic + Interaction") ~ 1,
                                .default = 0),
         Interaction = case_when(Complexity %in% c("Interaction", "Quadratic + Interaction") ~ 1,
                                .default = 0)) |>
  summarise(Polynomial = max(Polynomial),
            Interaction = max(Interaction),
            .by = ID) |>
  left_join(review_Data, by = "ID") |>
  mutate(Predict = if_else(Spatial == "Y" | Future == "Y",
                           "Y", "N")) |>
  pivot_longer(c(Hypothesis, Methods, Parameters, Predict),
               names_to = "Objective", values_to = "ObjIncluded") |>
  filter(ObjIncluded == "Y") |>
  summarise(Polynomial = sum(Polynomial)/n(),
            Interaction = sum(Interaction)/n(),
            n = n(),
            .by = Objective)

```

```{r select}

review_Data |>
  separate_longer_delim(Selection, ", ") |>  
  mutate(Predict = if_else(Spatial == "Y" | Future == "Y",
                           "Y", "N")) |>
  pivot_longer(c(Hypothesis, Methods, Parameters, Predict),
               names_to = "Objective", values_to = "ObjIncluded") |>
  filter(ObjIncluded == "Y") |>
  mutate(total = length(unique(ID)),
         .by = Objective) |>
  summarise(n = n(), total = mean(total),
            .by = c(Objective, Selection)) |>
  mutate(frac = 1-(n/total)) |>
  filter(Selection == "APrioriS")

```

```{r eval}

review_Data |>
  separate_longer_delim(Evaluation, ", ") |>  
  mutate(Predict = if_else(Spatial == "Y" | Future == "Y",
                           "Y", "N")) |>
  pivot_longer(c(Hypothesis, Methods, Parameters, Predict),
               names_to = "Objective", values_to = "ObjIncluded") |>
  filter(ObjIncluded == "Y") |>
  mutate(total = length(unique(ID)),
         .by = Objective) |>
  summarise(n = n(), total = mean(total),
            .by = c(Objective, Evaluation)) |>
  mutate(frac = n/total) |>
  arrange(Objective)

```

```{r Selection fig}
#| fig-width: 10
#| fig-height: 4

selectData <- review_Data |>
  separate_longer_delim(Selection, ", ") |> 
  mutate(total = length(unique(ID)),
         .by = Bayesian) |>
  summarise(n = n(), total = mean(total),
            .by = c(Selection, Bayesian)) |>
  mutate(fraction = n/total) |>
  filter(Bayesian != "NA") |>
  mutate(Bayesian = case_when(Bayesian == "Y" ~ "Bayesian",
                              Bayesian == "N" ~ "Frequentist"))

selectFigure <- selectData |>
  mutate(Bayesian = fct(Bayesian,
                        levels = c("Frequentist", 
                                   "Bayesian"))) |>
  mutate(Selection = fct(Selection, 
                         levels = c("APrioriS", "APrioriM",
                                    "Stepwise", "Exhaustive",
                                    "Simple", "Other"))) |>

  ggplot() +
  geom_col(aes(x = Selection, y = fraction, fill = Bayesian)) +
  facet_grid(cols = vars(Bayesian)) + 
  scale_fill_manual(values = c("darkorange3", "turquoise4")) +
  scale_y_continuous(limits = c(0, 0.8), expand = c(0, 0)) +
  scale_x_discrete(labels = c("Single\nmodel",
                              "Candidate\nsuite",
                              "Stepwise\nselection",
                              "Exhaustive\nselection",
                              "Simple", "Other\nmethod")) +
  labs(title = "Model selection methods",
       x = "", y = "Fraction of articles") +
  theme(panel.grid = element_blank(),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 15),
        strip.text = element_text(size = 14),
        axis.text.y = element_text(size = 13),
        axis.text.x = element_text(size = 11),
        axis.title.y = element_text(size = 13))

ggsave("Manuscript/Figures/SelectionFigure.jpeg",
       selectFigure, width = 10, height = 4)
```

### 2024-04-12 Manuscript/domReview.qmd

--- 2024-03-25 Manuscript/domReview.qmd
@@
 ---
-title: "Twenty years of dynamic occupancy modelling – what’s been done, and where to next?"
+title: "Twenty years of dynamic occupancy modelling --- what’s been done, and where to next?"
 date: today
@@
 
-abstract: |
-  In the twenty years since their introduction, the dynamic occupancy model has become an important tool for describing trends in species occupancy across landscapes and throughout time while accounting for imperfect detection.
-
-  We present a retrospective of the first two decades of dynamic occupancy modelling based on the results of a systematic review of models fit to field data.
-We describe how authors have used these models to describe patterns of occupancy for a diverse range of study systems and focal taxa, dedicating particular emphasis to how authors engaged with modelling decisions such as covariate selection and model evaluation.
-Modelling approaches vary considerably within the literature, and objective metrics of model performance are rarely estimated.
-
-  From these results we consider what the future of dynamic occupancy modelling may hold.
-Key questions remain on how the modelling process influences parameter estimates and model performance.
-Considerably more emphasis must be placed on evaluating models via processes such as out-of-sample validation.
-Further opportunities also abound; newly proliferating data types such as bioacoustics and camera traps present new questions around model design as well as new ways to study occupancy at scales which have not been historically possible.
-The suitability of dynamic occupancy models to generating predictions, as an alternative to static models, is another area which has been historically under-emphasised.
-
 bibliography: references.bib
@@
 
-The dynamic implementation of the occupancy model was introduced two decades ago in @mackenzie2003.
-Since then, the model has become a popular tool for ecologists seeking to describe occupancy dynamics while accounting for imperfect detection, generating hundreds of articles describing model uses and inspiring popular software tools for their implementation [@kery2016].
+Since its introduction two decades ago by @mackenzie2003, the dynamic multi-season occupancy model has become an important tool for analysis of species occupancy throughout space and time.
+Dynamic occupancy models (DOMs) represent a happy medium in the ecological modelling landscape.
+In explicitly estimating the processes which guide occupancy (i.e., colonisation and extinction) this model class offers more advanced insights than simpler static models of species distribution which have noted deficits in their ability to describe non-equilibrium states **\[cite\]**.
+The capacity to account for imperfect detection provides a strong additional benefit given the near-ubiquity of this source of bias in ecological field data.
+Even with these advantages, data input requirements for DOMs remain achievable for many research programs, particularly compared to other more complex process-explicit models.
 
-The dynamic occupancy model (DOM) fills an important niche in the ecological modelling landscape as an approachable midpoint between purely correlative species distributions models and more complex process-based models.
-Data input requirements are achievable for many research programs, and estimates of occupancy dynamic parameters (i.e., colonisation and extinction) are of use in addressing many key ecological questions.
-Furthermore, the capacity to control for imperfect detection addresses a ubiquitous source of bias and uncertainty in field ecological data.
-A description of the basic model form is presented in **Box 1.**
+Potential use cases of this model class are wide-ranging.
+Simple estimates of occupancy through time have value in many monitoring programs, particularly for threatened or invasive species where distributions are likely to exist in a non-equilibrium state **\[cite?\]**.
+These models can also be used in hypothesis testing of specific relationships, with examples including the effect of fire on extinction probability and the influence of species interactions on occupancy **\[cite,cite\]**.
+A less frequently pursued yet promising application of DOMs lies in their ability to make predictions, either to unsurveyed areas or into the future [@briscoe2021; @kéry2013a].
 
-## How are they used
+This review aims to provide an overview of the first two decades of the dynamic occupancy models, including its development and how authors have implemented it in their investigations of ecological questions.
+We present key guidelines for authors seeking to build these models in their work, and outline the key priorities and opportunities for future research incorporating this flexible and important model class.
 
-Outputs of dynamic occupancy model may be put to many uses.
-In many cases, simple estimates of the ecological parameters (i.e.
-occupancy, colonisation, or extinction) can aid in describing understanding changes in range dynamics.
-In other cases, estimates of covariates relating to these parameters are used to test hypotheses, for example, the influence of forest fire on the probability of local extinction.
-Less frequently, DOMs are used to make predictions of site occupancy either across unsurveyed sites or into the future under predictions of future conditions or management actions.
+### The dynamic occupancy model
 
-These models are particularly well-suited to non-equilibrium situations such as recovering, declining, or invasives species, where assumptions of conventional species distribution models cannot be met.
+Dynamic occupancy models jointly describe two processes: the ecological process of site occupancy describing the presence or absence of a species at a given site in a given point in time; and the observational process of detection describing whether a species is observed given that a site is occupied.
+The basic model form as described by @mackenzie2003 is depicted in @fig-modelform.
+Under this parameterisation sites exist in either occupied or unoccupied states which they may transition between through time.
+Occupancy state in the first time-step is given by the initial occupancy parameter, and in ensuing time-steps occupancy is described as a Markovian process governed by probabilities of colonisation and extinction.
+Given that a site is occupied, the probability of observing the species during a survey is given by the detection parameter.
+Under this formulation, it is assumed that false-positive detections do not occur.
 
-## What does this article do
+![The original parameterisation of the dynamic occupancy model.
+The ecological process of occupancy is guided by the initial occupancy, colonisation, and extinction parameters; the observational process is guided only by the detection parameter.](Figures/ModelForm.png){#fig-modelform width="1000"}
 
-We present a comprehensive overview of how the dynamic occupancy model has been applied in its first two decades, with a look to the key questions which remain on how DOM can best be used to address ecological questions.
+To disentangle the ecological and observation processes and estimate the detection parameter, the DOM relies on a hierarchical sampling design shown in @fig-surveys.
+Under this design, observations at a site occur during distinct, time-bound seasons in which sites are considered 'closed' to changes in occupancy.
+Within seasons multiple observations are conducted, which permit estimation of the probability of detection conditional on occupancy.
+These repeat observations may be conducted by a variety of methods.
+Most frequently surveys are simply repeated on separate occasions, although alternative methods exist.
+These include surveys conducted at multiple locations within a site **\[cite\]**, use of multiple observers during a survey **\[cite\]**, or by recording time elapsed before a detection is recorded **\[cite\]**.
 
-# History of the dynamic occupancy model
+![The sampling design of the dynamic occupancy model.](Figures/SurveyDesign.png){#fig-surveys fig-align="center" width="589"}
 
-## Precursors
+Dynamic occupancy models require a few assumptions with important implications for users.
 
-Dynamic occupancy models originate in the mask-recapture model literature with roots in the robust sampling design, which is also used in the Jolly-Seber and other related models.
-In the case of the dynamic occupancy model the unit of inference is the site, not the individual.
+1.
+**False positive detections do not occur.** While this assumption can be safely met in some studies, it is not guaranteed when working with more cryptic species.
+Several authors have commented on the perils of not accounting for false positives where they occur (@miller2015, @mcclintock2010).
+Furthermore, @berigan2019 notes that in some cases genuine identifications of a species may be considered false positives if they do not represent the right kind of occupancy - in this case, transient individuals where the authors were interested in breeding occupany.
+Extensions to the dynamic occupancy model which account for false positive error are available (@royle2006, @miller2011, @miller2013) and should be carefully considered where this assumption cannot be met.
 
-DOMs bear similarities to the Incidence-Function or Hanksi models frequently used in the study of island biogeography, with similar probabilities of colonisation and extinction; DOMs are typically not spatially-explicit, however.
+2.
+**Sites are closed to changes in occupancy between seasons.** This assumption, best known as the 'closure assumption,' has also been subject to considerable discussion on the bias which is introduced when it is violated (@rota2009, @otto2013).
+Closure is dependent not only on the behaviour of the species, but also the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discusses the difficultly in distinguishing temporary emigration and local extinction.
+Model extensions to relax the closure assumption have been developed, including @kendall2013 's approach using staggered arrival and depature periods between sites.
+More pertinent for most studies, however, is a careful consideration of ecologically and observationally relevant seasons which will allow the estimation of an appropriate 'type' of occupancy, whether that be shorter or longer term.
 
-## Variants
+3.
+**No modelled heterogeneity in ecological or observational parameters.** Relative to the former two assumptions, considerably less research has been conducted on the impact of unmodelled heterogeneity on parameter estimates.
+Particularly at larger spatial scales, environmental factors are near-guaranteed to influence species occupancy and observation.
 
-Various parameterisations of the standard dynamic occupancy have been described, the most frequent being those described by MacKenzie and Nichols.
-The former estimates probabilities of initial occupancy, colonisation, extinction, and detection and derives probabilities of yearly occupancy.
-The latter directly estimates probabilities of occupancy, local extinction, and detection; colonisation is then estimated as a derived parameter.
+# Methods
 
-The Bayesian implementation of the dynamic occupancy model was described by Nichols.
+This review is focused on the dynamic, multi-season, implementation of the occupancy model of @mackenzie2003.
+To qualify for inclusion articles must fit a model which fulfils these criteria: i) multiple sites which can exist in at least two states, including an occupied and unoccupied states; ii) multiple survey periods, between which sites may change state conditional on the prior survey period's occupancy state and at least one transition probability; iii) contains at least one detection parameter; iv) uses real (i.e., not simulated) field ecological data.
 
-## Model extensions
+Papers were solicited for inclusion in two stages.
+In a preliminary analysis, four search strings ("dynamic occupancy model," "occupancy dynamics model," "incidence function model," and "multi-season occupancy model" were fed into Google Scholar **\[cite\]**.
+Up to 100 articles were downloaded for each string in order of appearance.
+Articles were divided into four temporal strata: 2000-2005, 2006-2010, 2011-2015, and 2016-2021.
+For each combination of search string and year strata, either a random 25% or 5 articles (whichever was greater) were processed.
+Articles which did not meet inclusion criteria were replaced with another from the same search string and strata where possible.
 
-The dynamic occupancy model has been extended numerous times, to make it suitable for multiple scales, multiple species, and to accommodate for false positives.
-An in-depth discussion of model extensions is presented in Bailey 2014.
+After completion of the preliminary sample authors recognised significant uncertainty in how the model of interest was described, particularly in their first decade of use, and noted that the incidence-function model is a distinct model class **\[cite\].** To increase sample size, a complete list of articles which have cited @mackenzie2003 was downloaded from Web of Science.
+Articles in this list were split into the same year strata, plus an additional 2022-2023 category.
+15 articles were drawn from each strata and again processed with replacement from their strata.
 
-# Review of published implementations
+Details on methods for each article were recorded in detail, including the full suite of covariates considered for inclusion in modelling.
+The full spreadsheet of reviewed articles is available in **Appendix I**.
 
-## Systematic review methodology
+# Results
 
-## Study systems
+A total of **\[number\]** articles were included in the review, **\[number\]** from the preliminary sample and **\[number\]** from the secondary sample.
+Based on the acceptance rate from the secondary sample, an estimated **\[hit number\]** of the **\[total number\]** articles citing @mackenzie2003 would have met inclusion criteria; i.e., fit a dynamic occupancy model.
+
+![](Figures/CoveragePlot.jpeg){#fig-coverage}
+
+## Study details
+
+Articles were published with first-authors from a total of **\[number\]** countries, with authors from the United States producing **\[zz\]**% of articles.
+**\[x\]**% of articles had at least one author based at an academic institution, **\[y\]**% had authors at government institutions, and \[**z\]**% had authors from non-profit organisations.
+
+Data for DOMs were collected from study areas on all continents and in **\[xx\]** countries, with the United States again representing a majority of the total articles published (@fig-StudyDetails a).
+These study areas ranged in size from less than 1 to over 1 million square kilometres
+
+![](Figures/ReviewBox.png){#fig-StudyDetails}
 
@@
 
-## Objectives and authorship
+## Modelling
 
-## Model development and evaluation
+![](Figures/CovParamPlot.jpeg){#fig-covariates}
 
-# Future priorities for dynamic occupancy models
+## Selection
 
-## Model selection and evaluation
+![](Figures/SelectionFigure.jpeg){#fig-selection}
 
-## Continuous detection data
+### Complexity
 
-## Predictions
+![](Figures/ModelComplexity.png){#fig-complexity}
+
+# Discussion
+
+**\<Covariate inclusion figure\>**
+
+We provide guidelines for fitting dynamic occupancy models.
+
+![](Figures/Guidelines.png){#fig-guidelines}
+
+#

### 2024-04-12 Manuscript/domReview.qmd
Complete sections
completed manuscript sections rough text

@@
 
+## Article solicitation
+
 This review is focused on the dynamic, multi-season, implementation of the occupancy model of @mackenzie2003.
@@
 
+## Analysis
+
+For each included article we collected data collected on authorship, stated research objectives, study taxa and field survey locations, data collection methods, covariates, and model selection and evaluation.
+Particular emphasis was placed on the covariates considered by authors, with no distinction between covariates which were or were not included in final models.
+As we were principally interested in the modelling process, our analysis did not directly explore final models or the results generated by those models.
+A spreadsheet of reviewed articles and collected data is provided in Appendix I.
+
 # Results
@@
 A total of **\[number\]** articles were included in the review, **\[number\]** from the preliminary sample and **\[number\]** from the secondary sample.
-Based on the acceptance rate from the secondary sample, an estimated **\[hit number\]** of the **\[total number\]** articles citing @mackenzie2003 would have met inclusion criteria; i.e., fit a dynamic occupancy model.
+Based on the acceptance rate from the secondary sample, an estimated **\[hit number\]** of the **\[total number\]** articles citing @mackenzie2003 would have met inclusion criteria.
+Total published articles have increased considerably over time.
 
-![](Figures/CoveragePlot.jpeg){#fig-coverage}
-
-## Study details
+![Estimated coverage of articles included in this review](Figures/CoveragePlot.jpeg){#fig-coverage}
 
@@
 
-Data for DOMs were collected from study areas on all continents and in **\[xx\]** countries, with the United States again representing a majority of the total articles published (@fig-StudyDetails a).
-These study areas ranged in size from less than 1 to over 1 million square kilometres
+## Study locations, focal species, and data collection
+
+Data for DOMs were collected from study areas on all continents and in **\[xx\]** countries, with the United States again representing a majority of the total articles published.
+These study areas ranged in size from less than 1 to over 1 million square kilometres (@fig-StudyDetails a).
+
+Focal species were largely vertebrate taxa, although 6 articles fit models to invertebrates.
+A single non-animal taxa was represented in the sample, from an article fitting a DOM to a rare species of lichen @belinchón2017.
+[aa%] of articles modelled threatened taxa, and \[bb%\] modelled species invasive to the study area.
+The majority of articles (50, X%) fit single-species DOMs to one taxa.
+Of the remainder, 26 (X%) fit independent single-species models to more than one taxa.
+8 articles fit (X%) multi-species implementations with explicit interactions between species (@fig-StudyDetails b).
+
+Data used to fit dynamic occupancy models included human survey data such as transects or point counts (X% of articles), live-trapping data (X%), camera traps (X%) or bioacoustic monitors (Y%), and citizen science data (Z%); some articles incorporate multiple detection methods to build their detection datasets.
+Study duration, defined as the time elapsed between the first and last surveys conducted, ranged from 1 month to 40 years with the median study running for X years.
+The number of seasons ranged from 2 to Y with a median of Z; the number of observations per season ranged from X to Y with a median of Z.
+The number of sites surveyed per article ranged from X% to Y%, with the median study including Z sites (@fig-StudyDetails c).
 
@@
 
-## Data collection
+## Covariates
 
-## Modelling
+A summary of the covariates classes considered in the review sample is presented in @tbl-covariates. The most frequent
+
+| Covariate type | Proportion of articles considering | Proportion including on Initial Occupancy | Proportion dynamic | Proportion directly observed | Frequency of interaction terms | Frequency of polynomial terms |
+|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
+|                |                                    |                                           |                    |                              |                                |                               |
+|                |                                    |                                           |                    |                              |                                |                               |
+|                |                                    |                                           |                    |                              |                                |                               |
+|                |                                    |                                           |                    |                              |                                |                               |
+|                |                                    |                                           |                    |                              |                                |                               |
+|                |                                    |                                           |                    |                              |                                |                               |
+|                |                                    |                                           |                    |                              |                                |                               |
+|                |                                    |                                           |                    |                              |                                |                               |
+|                |                                    |                                           |                    |                              |                                |                               |
+|                |                                    |                                           |                    |                              |                                |                               |
+
+: Covariates considered in dynamic occupancy models {#tbl-covariates}
 
 ![](Figures/CovParamPlot.jpeg){#fig-covariates}
-
-## Selection
-
-![](Figures/SelectionFigure.jpeg){#fig-selection}
 
@@
 
+### Selection and evaluation
+
+Bayesian models were far less likely to implement a model selection protocol compared to frequentist models: 30% of the former fit only a single model defined *a priori,* compared to 10% of the latter (@fig-selection).
+In frequentist models, stepwise selection protocols were the most popular choice.
+The vast majority of frequentist models (XX%) used AIC or AICc to identify the best model.
+
+![](Figures/SelectionFigure.jpeg){#fig-selection}
+
+Objective evaluation of model fit was rare throughout the sample, regardless of selection protocol.
+XX% of articles calculated non-relative metrics, either via in-sample validation, bootstrapped goodness-of-fit tests, or out-of-sample validation.
+
 # Discussion
 
-**\<Covariate inclusion figure\>**
+### Usage of dynamic occupancy models
 
-We provide guidelines for fitting dynamic occupancy models.
+Our review sample displays considerable diversity in implementations of dynamic occupancy models with widespread representation from mammals, birds, and herptiles.
+This assemblage contrasts with the related stochastic patch occupancy models (SPOMs, @gutiérrez-arellano), where the majority of models are fit to invertebrates, plants, and fungi.
+This may reflect SPOMs lack of consideration for imperfect detection and assumption of discrete habitat patches, limitations which are unlikely to be fulfilled by most vertebrate taxa.
 
-![](Figures/Guidelines.png){#fig-guidelines}
+DOMs also display great variation in scale, from local studies with a handful of sites to continental-scale studies spanning decades.
 
-#
+Several types of detection data have been successfully incorporated DOMs.
+This flexibility may be one of the more notable aspects of dynamic occupancy models; while the heirarchical data inputs may be more complex than those required by other models of species distributions, authors have been able to successfully manipulate datasets from many sources to meet requirements.
+
+### Practices in the modelling process
+
+Model selection approaches diverge between Bayesian and frequentist implementations of the model.
+While model selection will naturally look different between these two categories, it is unclear what effect a lack of model selection may have on the outputs produced by Bayesian models and it is likely that Bayesian model selection approaches should be further explored @hooten2015.
+
+Notably, the popular sequential approach to model selection (where one sub-model is assessed at a time) may not provide optimal results as shown in @morin2020.
+Considerably more research is required on this topic to clarify how authors should best approach the model building process.
+
+### Key considerations for model users
+
+Implementation of dynamic occupancy models requires considerable thought to ensure that model outputs correspond with author expectations and ecological realities.
+Based upon our review of the existing literature we discuss key questions authors should carefully engage with during the modelling process.
+
+1.
+**Are seasons ecologically meaningful, and how do they affect the definition of occupancy?** As previously discussed, the duration of the season has important implications for both the reliability and proper interpretation of model results.
+Seasons which are not ecologically meaningful may run afoul of the closure assumption resulting in associated biases.
+Also relevant is how the definition of 'occupancy' may shift with season length and life-history of a species — for many species, occupancy within a year and within a week may be considerably different concepts, with shorter intervals more closely relating to the idea of 'use.' With the increasing proliferation of continuous detection methods such as camera traps and acoustic monitors where the definition of 'season' is to some degree arbitrary, it is increasingly important to thoughtfully consider how this element of the DOM is defined.
+
+2.
+**Are key drivers of occupancy and detection incorporated?** While it is effectively impossible to *fully* describe heterogeneity across a landscape, authors should make their best efforts to capture the drivers of occupancy and detection most likely to influence their study species.
+This is true even when another covariate is of principal interest to investigators — failure to include other major determinants of occupancy is likely to bias parameter estimates.
+One place where this is of particular importance is in determining drivers of detection, which is not only subject to observational conditions but also ecological determinants which may cause the species to spend more time at a location and thus be more frequently available for detection.
+
+3.
+**Ensure covariates reflect ecological complexity.** Many ecological relationships are non-linear and may require polynomials terms to be accurately represented.
+This is particularly true at large spatial scales where a larger portion of a species ecological niche is shown in surveyed data.
+Authors should question whether a linear term is sufficient to capture hypothetical relationships in their study systems and consider quadratic terms where appropriate.
+
+4.
+**Consider model evaluation.** Model evaluation remains rare in ecological research relative to other fields embracing complex modelling, likely due to a reluctance to 'throw away' data for use in testing.
+This concept should be challenged, particularly for larger scale long-term projects where reserving a fraction of data for testing is unlikely to substantially reduce precision of parameter estimates.
+For studies where this is genuinely not possible, reporting results of goodness-of-fit tests may still provide valuable information to readers.
+
+### Priorities for future research
+
+Work on dynamic occupancy modelling over the past 20 years has developed this model class into an important tool for describing species distributions and ecological relationships.
+Historically, methods-development research for dynamic occupancy models has focused on questions of model structure — namely ideas around the closure assumption and imperfect detection.
+This emphasis likely reflects the DOM's roots in the mark-recapture literature, where these topics have been discussed for several decades.
+Considerably less emphasis has been placed on the relation of model parameters to environmental factors via covariates and the model selection processes used to do so.
+This stands in stark contrast to related correlative species distribution models, where these topics have been the subject of considerable research over the years.
+There are opportunities for the DOM literature to learn from species distribution models, and more targeted research on the influence of the model selection process is necessary.
+
+The use of DOMs in prediction has remained fairly limited despite significant potential for this use-case.
+Papers in our sample which have taken advantage of this potential included @mcgowan2020 who used DOMs to make predictions of black rail distributions under various management scenarios, and @saunders2020, who validated predictions from ecological niche models using continent-scale DOMs.
+Further implementations of these models may benefit from the aforementioned necessary research on model selection to ensure robust outputs.
+
+A final opportunity for future research relates to autonomous detection methods, such as camera traps and bio-acoustic monitors — these technologies have permitted collection of enormous quantities of data particularly well-suited to dynamic occupancy models, but have also raised additional questions.
+The definition of seasons is of increased importance for these data types, where division of data into 'seasons' and 'observations' is arbitrary.
+Further research should explore the implications of this phenomenon and additional modelling opportunities made possible by this novel data type.
+
+The past two decades have shown that dynamic occupancy models are an excellent tool for addressing a wide range of ecological questions relating to species occupancy throughout space and time.
+The flexibility afforded by model extensions and the considerable advantages of the class over other species distribution models is likely to ensure that these models continue to increase in popularity.
+Given their strong use on threatened taxa and excellent potential to support environemental management, it is important that research continues to support model users in creating the most reliable models possible.

### 2024-04-17 Analysis/ReviewFigures/ReviewFigures.qmd
Final version for supervisors

@@
         axis.text = element_blank(), axis.ticks = element_blank()) 
+countries_map
 
@@
 
+prelimCoverage <- review_Data |>
+  filter(Query != "Mackenzie") |>
+  summarise(Prelim = n(), .by = Strata)
+
 sampleCoveragePlot <- sampleCoverage |>
-  pivot_longer(c(Accept, EstRemaining),
-               names_to = "Category", values_to = "Articles") |>
   mutate(Group = case_when(Group == "MacKenzie Cited_2000:2005" ~ "2000-2005",
@@
                               "2022-2023"))) |>
+  left_join(prelimCoverage, by = join_by("Group" == "Strata")) |>
+  pivot_longer(c(Accept, Prelim, EstRemaining),
+               names_to = "Category", values_to = "Articles") |>
+  mutate(Category = fct(Category,
+                        levels = c("Prelim", "Accept", "EstRemaining"))) |>
 
@@
            position = position_stack(reverse = TRUE)) +
-  scale_fill_manual(limits = c("Accept", "EstRemaining"),
-                    values = c("turquoise4", "gray70"),
-                    labels = c("Processed", "Estimated")) +
-  scale_y_continuous(limits = c(0, 250), 
+  scale_fill_manual("",
+                    limits = c("Prelim", "Accept", "EstRemaining"),
+                    values = c("darkslategray", "turquoise4", "gray70"),
+                    labels = c("Preliminary", "Secondary", "Estimated articles")) +
+  scale_y_continuous(limits = c(0, 300), 
                      expand = c(0,0)) +
@@
         legend.text = element_text(size = 13))
+sampleCoveragePlot
 
@@
        width = 9, height = 4)
+sampleCoverage
+```
+
+## Authorship
+
+```{r Authorship}
+
+review_Data |>
+
 ```
@@
 ```{r Focal species bar plot}
+#| fig-height: 4
+#| fig-width: 8
 
@@
   geom_col(aes(y = Taxa, x = Values, fill = Category)) +
-  scale_fill_manual(values = c("darkslategray", "goldenrod", "firebrick4")) +
+  scale_fill_manual(values = c("#2E848E", "gray70", "#896EA3")) +
   labs(x = "", y = "") +
   scale_x_continuous(expand = c(0, 1)) +
-  theme(legend.position = "top",
-        axis.text.y = element_text(size = 15),
+  theme(legend.position = "none",
+        axis.text.y = element_text(size = 24),
         axis.ticks.y = element_blank(),
-        axis.text.x = element_text(size = 10), 
-        panel.grid = element_blank())
+        axis.text.x = element_blank(), 
+        axis.ticks.x = element_blank(),
+        panel.grid = element_blank(),
+        panel.background = element_rect(fill='transparent', color=NA),
+        plot.background = element_rect(fill='transparent', color=NA))
+speciesPlot
 
-ggsave("Manuscript/Figures/SpeciesPlot.jpeg", speciesPlot, width = 7, height = 5)
+ggsave("Manuscript/Figures/SpeciesPlot.png", speciesPlot, 
+       width = 8, height = 4, bg = "transparent")
 ```
@@
 ```{r Study area size}
+#| fig-width: 7
+#| fig-height: 5
 
@@
  studyAreaPlot <- ggplot(studyArea) +
-    geom_col(aes(y = StudyArea, x = Articles), fill = "darkslategray") +
-    labs(title = "Study Area (square kilometres)", x = "Articles", y = "") +
-    scale_x_reverse(expand = c(0,0)) +
-    scale_y_discrete(position = "right") +
+    geom_col(aes(x = StudyArea, y = Articles), fill = "#2E848E") +
+    labs(title = "", x = "", y = "") +
+    scale_y_continuous(expand = c(0,0)) +
+    scale_x_discrete(labels = c("0-1", "1-10", "10-100",
+                                "100-\n1000", "1000-\n10000",
+                                "10000-\n100000", "100000-\n1000000",
+                                "1000000+")) +
     theme(panel.grid = element_blank(),
-          axis.text = element_text(size = 15), 
-          axis.title = element_text(size = 14),
+          panel.background = element_rect(fill='transparent', color=NA),
+          plot.background = element_rect(fill='transparent', color=NA),
+          axis.text.x = element_text(size = 23,
+                                     angle = 90,
+                                     vjust = 0.5, hjust = 1),
+          axis.text.y = element_blank(),
+          axis.ticks.y = element_blank(),
+          axis.title.y = element_text(size = 20),
           title = element_text(size = 15),
-          axis.ticks.y = element_blank())
+          axis.ticks.x = element_blank())
+  studyAreaPlot
 
-  ggsave("Manuscript/Figures/StudyAreaPlot.jpeg", studyAreaPlot)
+  ggsave("Manuscript/Figures/StudyAreaPlot.png",
+         studyAreaPlot,
+         width = 7, height = 5,
+         bg='transparent')
 
@@
 ```{r Duration fig}
+#| fig-height: 7
+#| fig-width: 4
 
  durationPlot <- ggplot(review_Data) +
-  geom_histogram(aes(x = Duration), binwidth = 0.2, fill = "turquoise4") +
-  geom_vline(aes(xintercept = median(Duration)), 
-             colour = "gray30", linetype = 2, linewidth = 3) +
-  scale_x_continuous(transform = "log10", 
+  geom_histogram(aes(y = Duration), binwidth = 0.2, fill = "#2E848E") +
+  geom_hline(aes(yintercept = median(Duration)), 
+             colour = "gray80", linetype = 2, linewidth = 3) +
+  scale_y_continuous(transform = "log10", 
                      breaks = c(1, 12, 60, 
@@
                                 "10 years", "20 years", "40 years")) +
-  scale_y_continuous(expand = c(0, 0)) + 
-  labs(title = "Study duration", x = "", y = "") +
+  scale_x_continuous(expand = c(0, 0)) + 
+  labs(title = "", x = "", y = "") +
   theme(panel.grid = element_blank(),
+        panel.background = element_rect(fill='transparent', color=NA),
+        plot.background = element_rect(fill='transparent', color=NA),
         title = element_text(size = 13), 
-        axis.text.y = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.x = element_text(size = 16, angle = 70, 
-                                   hjust = 1, vjust = 1))
+        axis.text.x = element_blank(),
+        axis.ticks.x = element_blank(),
+        axis.text.y = element_text(size = 20))
+durationPlot
+
 ggsave("Manuscript/Figures/Duration.jpeg", durationPlot, 
@@
 ```{r Sites fig}
+#| fig-width: 7
+#| fig-height: 4
+
 sitesPlot <- review_Data |>
@@
                                    hjust = 1, vjust = 1))
+sitesPlot
+
 ggsave("Manuscript/Figures/Sites.jpeg", sitesPlot, 
@@
 
-```{r Covs by param}
+```{r Covariate type table}
+covType_overall <- covariates_clean |>
+  select(ID, Type) |>
+  mutate(Included = 1, Total = length(unique(ID))) |>
+  distinct() |>
+  complete(ID, Type, fill = list(Included = 0)) |>
+  summarise(Articles = sum(Included),
+            Total = max(Total, na.rm = TRUE),
+            .by = Type) |>
+  mutate(Proportion = round(Articles/Total, 2)) |>
+  select(Type, Overall_Proportion = Proportion)
 
-covParamPlot <- covariates_clean |>
+covType_psi <- covariates_clean |>
+  filter(Parameter == "InitialOccupancy") |>
+  select(ID, Type) |>
+  mutate(Included = 1, Total = length(unique(ID))) |>
+  distinct() |>
+  complete(ID, Type, fill = list(Included = 0)) |>
+  summarise(Articles = sum(Included),
+            Total = max(Total, na.rm = TRUE),
+            .by = Type) |>
+  mutate(Proportion = round(Articles/Total, 2)) |>
+  select(Type, InitOcc_Proportion = Proportion)
+
+covType_gamma <- covariates_clean |>
+  filter(Parameter == "Colonisation") |>
+  select(ID, Type) |>
+  mutate(Included = 1, Total = length(unique(ID))) |>
+  distinct() |>
+  complete(ID, Type, fill = list(Included = 0)) |>
+  summarise(Articles = sum(Included),
+            Total = max(Total, na.rm = TRUE),
+            .by = Type) |>
+  mutate(Proportion = round(Articles/Total, 2)) |>
+  select(Type, Colonisation_Proportion = Proportion)
+
+covType_epsilon <- covariates_clean |>
+  filter(Parameter == "Extinction") |>
+  select(ID, Type) |>
+  mutate(Included = 1, Total = length(unique(ID))) |>
+  distinct() |>
+  complete(ID, Type, fill = list(Included = 0)) |>
+  summarise(Articles = sum(Included),
+            Total = max(Total, na.rm = TRUE),
+            .by = Type) |>
+  mutate(Proportion = round(Articles/Total, 2)) |>
+  select(Type, Extinction_Proportion = Proportion)
+
+covType_detection <- covariates_clean |>
+  filter(Parameter == "Detection") |>
+  select(ID, Type) |>
+  mutate(Included = 1, Total = length(unique(ID))) |>
+  distinct() |>
+  complete(ID, Type, fill = list(Included = 0)) |>
+  summarise(Articles = sum(Included),
+            Total = max(Total, na.rm = TRUE),
+            .by = Type) |>
+  mutate(Proportion = round(Articles/Total, 2)) |>
+  select(Type, Detection_Proportion = Proportion)
+
+covObserved <- covariates_clean |>
+  filter(Type != "NONE") |>
+  mutate(Observed = case_when(Measurement == "Observed" ~ 1,
+                              .default = 0)) |>
+  mutate(Total = n(), .by = Type) |>
+  summarise(Observed = sum(Observed),
+            .by = c(Type, Total)) |>
+  mutate(Proportion_Observed = round(Observed/Total, 2)) |>
+  select(Type, Proportion_Observed)
+
+covDynamic <- covariates_clean |>
+  filter(Type != "NONE") |>
+  filter(Parameter %in% c("Detection", "Colonisation", "Detection")) |>
+  mutate(Dynamic = case_when(Variation %in% c("Year",
+                                              "Site-Year",
+                                              "Site-Year-Survey") ~ 1,
+                              .default = 0)) |>
+  mutate(Total = n(), .by = Type) |>
+  summarise(Dynamic = sum(Dynamic),
+            .by = c(Type, Total)) |>
+  mutate(Proportion_Dynamic = round(Dynamic/Total, 2)) |>
+  select(Type, Proportion_Dynamic)
+
+covPolynomial <- covariates_clean |>
+  filter(Type != "NONE") |>
+  mutate(Polynomial = case_when(Complexity %in% c("Quadratic",
+                                                  "Quadratic + Interaction",
+                                                  "Other") ~ 1,
+                                .default = 0)) |>
+  mutate(Total = n(), .by = Type) |>
+  summarise(Polynomial = sum(Polynomial),
+            .by = c(Type, Total)) |>
+  mutate(Proportion_Polynomial = round(Polynomial/Total, 2)) |>
+  select(Type, Proportion_Polynomial)
+
+covInteract <- covariates_clean |>
+  filter(Type != "NONE") |>
+  mutate(Interact = case_when(Complexity %in% c("Interaction",
+                                                  "Quadratic + Interaction") ~ 1,
+                                .default = 0)) |>
+  mutate(Total = n(), .by = Type) |>
+  summarise(Interact = sum(Interact),
+            .by = c(Type, Total)) |>
+  mutate(Proportion_Interact = round(Interact/Total, 2)) |>
+  select(Type, Proportion_Interact)
+
+covType_table <- covType_overall |>
+  full_join(covType_psi, by = "Type") |>
+  full_join(covType_gamma, by = "Type") |>
+  full_join(covType_epsilon, by = "Type") |>
+  full_join(covType_detection, by = "Type") |>
+  full_join(covDynamic, by = "Type") |>
+  full_join(covObserved, by = "Type") |>
+  full_join(covPolynomial, by = "Type") |>
+  full_join(covInteract, by = "Type") |>
+
+  filter(!Type %in% c("NONE", "OTHER", "SP", "UNSPECIFIED", "SITE")) |>
+  mutate(Category = case_when(Type %in% c("E-ANTH", "E-CLIM", "E-DIS",
+                                          "E-HTAQ", "E-HTT", "E-HYD",
+                                          "E-SEAS", "E-TOPO", "INT") ~ "Environmental",
+                              .default = "Other")) |>
   mutate(Type = case_when(Type == "E-ANTH" ~ "Anthropogenic",
@@
                           Type == "E-HYD" ~ "Hydrologic",
-                          Type == "E-SEAS" ~ "Phenologic",
-                          Type == "E-TOPO" ~ "Topographic",
+                          Type == "E-SEAS" ~ "Phenology",
+                          Type == "E-TOPO" ~ "Topography",
                           Type == "CONN" ~ "Connectivity",
+                          Type == "INT" ~ "Biotic Interaction",
                           Type == "GEOM" ~ "Site geometry",
-                          Type == "INT" ~ "Biotic interaction",
-                          Type == "SEAS" ~ "Primary occasion",
-                          Type == "SITE" ~ "Site",
-                          Type == "SP" ~ "Species",
-                          Type == "SURV" ~ "Survey detail",
-                          .default = Type)) |>
-  mutate(Type_cat = case_when(Type %in% c("Habitat (Terrestrial)",
-                                          "Habitat (Aquatic)",
-                                          "Hydrologic", 
-                                          "Climate/Weather",
-                                          "Disaster/Disturbance",
-                                          "Topographic",
-                                          "Phenologic",
-                                          "Anthropogenic",
-                                          "Biotic interaction") ~ "Env",
-                              .default = "Other")) |>
+                          Type == "SEAS" ~ "Season",
+                          Type == "SURV" ~ "Survey effect"))
+
+kable(covType_table)
+write_csv(covType_table, "Manuscript/Figures/CovariateTable.csv")
+```
+
+```{r Objectives table}
+
+objs_df <- review_Data |> 
+  select(ID, Parameters, Hypothesis, Drivers, Predict, Methods, 
+         Selection, Evaluation) |>
+  mutate(Any = "Y") |>
+  pivot_longer(c(Parameters, Hypothesis, Drivers, Predict, Methods, Any),
+               names_to = "Objective", values_to = "Included") |>
+  filter(Included == "Y")
+
+covs_summary <- covariates_clean |>
+  mutate(Polynomial = case_when(Complexity %in% c("Quadratic",
+                                                  "Quadratic + Interaction",
+                                                  "Other") ~ 1,
+                                .default = 0)) |>
+  mutate(Interact = case_when(Complexity %in% c("Interaction",
+                                                  "Quadratic + Interaction") ~ 1,
+                                .default = 0)) |>
+  select(ID, Parameter, Type, Polynomial, Interact) |>
+  summarise(Covs = case_when("NONE" %in% Type ~ 0,
+                             .default = n()),
+            Polynomial = max(Polynomial),
+            Interact = max(Interact),
+            .by = c(ID, Parameter)) |>
+  summarise(Covariates = sum(Covs),
+            Polynomial = max(Polynomial),
+            Interact = max(Interact),
+            .by = ID)
+
+objTable <- objs_df |>
+  left_join(covs_summary, by = "ID") |>
+  mutate(OverallTotal = length(unique(ID))) |>
+  mutate(ObjTotal = length(unique(ID)),
+         .by = Objective) |>
+  mutate(ObjEval = case_when(str_detect(Evaluation, "GOF") ~ 1,
+                             str_detect(Evaluation, "InSampleVal") ~ 1,
+                             str_detect(Evaluation, "OutSampleVal") ~ 1,
+                             .default = 0)) |>
+  mutate(Select = case_when(Selection != "APrioriS" ~ 1,
+                            .default = 0)) |>
+  summarise(ObjTotal = mean(ObjTotal),
+            OverallTotal = mean(OverallTotal),
+            MedianCovs = median(Covariates, na.rm = TRUE),
+            Poly = sum(Polynomial), Int = sum(Interact), 
+            Eval = sum(ObjEval), Select = sum(Select),
+            .by = Objective)|>
+  mutate(ObjProportion = ObjTotal/OverallTotal,
+         ProbPoly = Poly/ObjTotal,
+         ProbInt = Int/ObjTotal,
+         ProbEval = Eval/ObjTotal,
+         ProbSelect = Select/ObjTotal) |>
+  select(Objective, Articles = ObjTotal, ArticlesProp = ObjProportion,
+         MedianCovs, ProbPoly, ProbInt, ProbEval, ProbSelect) 
+kable(objTable)
+
+write_csv(objTable, "Manuscript/Figures/ObjectiveTable.csv")
+```
+
+```{r Covs by param}
+
+covParamData <- covariates_clean |>
   summarise(Covariates = case_when("NONE" %in% Type ~ 0,
                                    .default = n()), 
-            .by = c(ID, Parameter, Type_cat)) |>
+            .by = c(ID, Parameter)) |>
   filter(Parameter != "Occupancy") |>
-  complete(nesting(ID, Parameter), 
-           Type_cat, fill = list(Covariates = 0)) |>
+  mutate(Parameter = case_when(Parameter == "InitialOccupancy" ~ "Initial Occupancy",
+                               .default = Parameter)) |>
   mutate(Parameter = fct(Parameter, 
-                         levels = c("InitialOccupancy",
+                         levels = c("Initial Occupancy",
                                     "Colonisation", "Extinction",
                                     "Detection"))) |>
+  mutate(Covariates = case_when(Covariates > 15 ~ 15,
+                                .default = Covariates)) |>
+  mutate(MedianCovs = mean(Covariates),
+         .by = Parameter)
 
+covParamPlot <- covParamData |>
   ggplot() +
-  geom_boxplot(aes(x = Covariates, y = Type_cat, 
-                   colour = Type_cat, fill = Type_cat),
-               alpha = 0.5, outliers = FALSE, staplewidth = 0.5) +
+  geom_histogram(aes(x = Covariates, fill = Parameter), binwidth = 1) +
   facet_grid(rows = vars(Parameter), switch = "y") +
-  scale_colour_manual("Covariate type",
-                      values = c("seagreen4", "darkslategray"),
-                      labels = c("Environmental", "Other")) +
-  scale_fill_manual("Covariate type",
-                      values = c("seagreen4", "darkslategray"),
-                      labels = c("Environmental", "Other")) +
   labs(title = "Covariate quantity by parameter", 
        x = "Number of covariates considered", y = "") +
-  theme(strip.text.y.left = element_text(angle = 0, size = 14, hjust = 1),
-        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
-        legend.position = "bottom",
+  scale_x_continuous(breaks = c(0, 5, 10, 15),
+                     labels = c("0", "5", "10", "15+")) +
+  scale_y_continuous(position = "right") +
+  scale_fill_manual(values = c("turquoise4", "turquoise4", "turquoise4", "orchid4")) +
+  theme(strip.text.y.left = element_text(angle = 0, size = 12, hjust = 1),
+        panel.grid = element_blank(),
+        #axis.text.y = element_blank(), axis.ticks.y = element_blank(),
+        legend.position = "none",
         title = element_text(size = 13),
+        axis.text = element_text(size = 13),
         axis.title.x = element_text(size = 13),
         legend.text = element_text(size = 13)) 
+covParamPlot
 
@@
        width = 7, height = 5)
-```
-
-```{r Covs by objective}
-
-covariates_clean |>
-  summarise(Covariates = case_when("NONE" %in% Type ~ 0,
-                                   .default = n()), 
-            .by = c(ID, Parameter)) |>
-  summarise(Covariates = sum(Covariates),
-            .by = ID) |>
-  left_join(review_Data, by = "ID") |>
-  mutate(Predict = if_else(Spatial == "Y" | Future == "Y",
-                           "Y", "N")) |>
-  pivot_longer(c(Hypothesis, Methods, Parameters, Predict),
-               names_to = "Objective", values_to = "ObjIncluded") |>
-  filter(ObjIncluded == "Y") |>
-  summarise(MedianCovs = median(Covariates), n = n(),
-            .by = Objective)
-
-```
-
-```{r percent polynomial}
-
-covariates_clean |>
-  mutate(Polynomial = case_when(Complexity %in% c("Quadratic", "Quadratic + Interaction") ~ 1,
-                                .default = 0),
-         Interaction = case_when(Complexity %in% c("Interaction", "Quadratic + Interaction") ~ 1,
-                                .default = 0)) |>
-  summarise(Polynomial = max(Polynomial),
-            Interaction = max(Interaction),
-            .by = ID) |>
-  left_join(review_Data, by = "ID") |>
-  mutate(Predict = if_else(Spatial == "Y" | Future == "Y",
-                           "Y", "N")) |>
-  pivot_longer(c(Hypothesis, Methods, Parameters, Predict),
-               names_to = "Objective", values_to = "ObjIncluded") |>
-  filter(ObjIncluded == "Y") |>
-  summarise(Polynomial = sum(Polynomial)/n(),
-            Interaction = sum(Interaction)/n(),
-            n = n(),
-            .by = Objective)
-
-```
-
-```{r select}
-
-review_Data |>
-  separate_longer_delim(Selection, ", ") |>  
-  mutate(Predict = if_else(Spatial == "Y" | Future == "Y",
-                           "Y", "N")) |>
-  pivot_longer(c(Hypothesis, Methods, Parameters, Predict),
-               names_to = "Objective", values_to = "ObjIncluded") |>
-  filter(ObjIncluded == "Y") |>
-  mutate(total = length(unique(ID)),
-         .by = Objective) |>
-  summarise(n = n(), total = mean(total),
-            .by = c(Objective, Selection)) |>
-  mutate(frac = 1-(n/total)) |>
-  filter(Selection == "APrioriS")
-
-```
-
-```{r eval}
-
-review_Data |>
-  separate_longer_delim(Evaluation, ", ") |>  
-  mutate(Predict = if_else(Spatial == "Y" | Future == "Y",
-                           "Y", "N")) |>
-  pivot_longer(c(Hypothesis, Methods, Parameters, Predict),
-               names_to = "Objective", values_to = "ObjIncluded") |>
-  filter(ObjIncluded == "Y") |>
-  mutate(total = length(unique(ID)),
-         .by = Objective) |>
-  summarise(n = n(), total = mean(total),
-            .by = c(Objective, Evaluation)) |>
-  mutate(frac = n/total) |>
-  arrange(Objective)
-
 ```
@@
                               "Candidate\nsuite",
-                              "Stepwise\nselection",
+                              "Procedural\nselection",
                               "Exhaustive\nselection",
-                              "Simple", "Other\nmethod")) +
+                              "Simple\nprecursor", "Other\nmethod")) +
   labs(title = "Model selection methods",
-       x = "", y = "Fraction of articles") +
+       x = "", y = "Proportion of articles") +
   theme(panel.grid = element_blank(),
@@
         axis.title.y = element_text(size = 13)) 
-
+selectFigure
 ggsave("Manuscript/Figures/SelectionFigure.jpeg",

### 2024-04-17 Manuscript/domReview.qmd
Final version for supervisors

--- 2024-04-12 Manuscript/domReview.qmd
@@
 ---
-title: "Twenty years of dynamic occupancy modelling --- what’s been done, and where to next?"
+title: "Twenty years of dynamic occupancy models: a review of applications and a look towards the future"
 date: today
+
 authors:
@@
 
+abstract: | 
+  Dynamic occupancy models are an increasingly important tool for the analysis of wildlife occurence data due to their favourable balance of complexity and ease of implementation.
+Uses of this model class are diverse, including explorations of the drivers of occupancy as well as predictions of species distributions across landscapes and into the future.
+The application of these models does require careful navigation of the modelling process including important decisions on structure, covariate inclusion, and assessment of model performance.
+
+  In this article we present a systematic review of implementations of the dynamic occupancy model, with particular emphasis on how authors have approached the modelling process.
+Our findings indicate that authors have found this model class suitable for a considerably diverse set of ecological questions and have successfully fit models to data collected under a wide range of conditions, highlighting the high degree of flexibility available.
+However, the manner of application appears to be quite variable: approaches to covariate selection vary considerably, and model complexity generally speaking is low.
+The implications of these decisions as they relate to model outputs are yet to be explored in detail and should be a priority for future research.
+
+embed-resources: true
 bibliography: references.bib
-embed-resources: true
 ---
@@
 
-Since its introduction two decades ago by @mackenzie2003, the dynamic multi-season occupancy model has become an important tool for analysis of species occupancy throughout space and time.
-Dynamic occupancy models (DOMs) represent a happy medium in the ecological modelling landscape.
-In explicitly estimating the processes which guide occupancy (i.e., colonisation and extinction) this model class offers more advanced insights than simpler static models of species distribution which have noted deficits in their ability to describe non-equilibrium states **\[cite\]**.
-The capacity to account for imperfect detection provides a strong additional benefit given the near-ubiquity of this source of bias in ecological field data.
-Even with these advantages, data input requirements for DOMs remain achievable for many research programs, particularly compared to other more complex process-explicit models.
+Since its introduction two decades ago by @mackenzie2003 the dynamic multi-season occupancy model has become an important tool for the analysis of species occurrence throughout space and time.
+In the ecological modelling landscape dynamic occupancy models (DOMs) occupy a favourable position between model complexity and ease of data collection.
+In explicitly estimating the processes which guide occupancy while accounting for imperfect detection, the model class offers more advanced insights than simpler static species distribution models; despite this, input data requirements remain achievable compared to other more complex process-explicit models.
 
-Potential use cases of this model class are wide-ranging.
-Simple estimates of occupancy through time have value in many monitoring programs, particularly for threatened or invasive species where distributions are likely to exist in a non-equilibrium state **\[cite?\]**.
-These models can also be used in hypothesis testing of specific relationships, with examples including the effect of fire on extinction probability and the influence of species interactions on occupancy **\[cite,cite\]**.
-A less frequently pursued yet promising application of DOMs lies in their ability to make predictions, either to unsurveyed areas or into the future [@briscoe2021; @kéry2013a].
+The use-cases of this model class are wide-ranging, and overlap considerably with other models such as species distribution models [@elith2009] and stochastic patch occupancy models [@gutiérrez-arellano].
+Estimates of occupancy through time are independently valuable for many monitoring programs, and DOMs are a particularly favourable choice when species are not detected perfectly [@guillera-arroita2017].
+DOMs may also be used in testing specific hypotheses of relationships between occupancy and the environment, or in identifying the drivers of occupancy in a system.
+Although infrequently applied thus far, a promising application of DOMs is their use to make predictions to unsurveyed sites or into the future [@briscoe2021; @kéry2013a].
+In each of these use cases, authors must navigate the modelling process and make important decisions on aspects of model structure, the process of determining which covariates to incorporate, and the critical assessment of model outputs such as parameter estimates or predictions.
 
-This review aims to provide an overview of the first two decades of the dynamic occupancy models, including its development and how authors have implemented it in their investigations of ecological questions.
-We present key guidelines for authors seeking to build these models in their work, and outline the key priorities and opportunities for future research incorporating this flexible and important model class.
+This review seeks to provide an overview dynamic occupancy model's first two decades of use.
+We discuss the DOMs key features, limitations, and implementation process, before presenting a systematic review of articles implementing the model.
+Based on these results, we present key considerations for authors seeking to build these models in their work and outline priorities and opportunities for future research involving this flexible and important model class.
 
-### The dynamic occupancy model
+## Model form
 
-Dynamic occupancy models jointly describe two processes: the ecological process of site occupancy describing the presence or absence of a species at a given site in a given point in time; and the observational process of detection describing whether a species is observed given that a site is occupied.
+Dynamic occupancy models unite two processes: the ecological process of site occupancy describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed given that a site is occupied.
 The basic model form as described by @mackenzie2003 is depicted in @fig-modelform.
-Under this parameterisation sites exist in either occupied or unoccupied states which they may transition between through time.
+Under this parameterisation, sites exist in either occupied or unoccupied states and may transition between these states between each time-step.
 Occupancy state in the first time-step is given by the initial occupancy parameter, and in ensuing time-steps occupancy is described as a Markovian process governed by probabilities of colonisation and extinction.
@@
 
-![The original parameterisation of the dynamic occupancy model.
+![The original parameterisation of the dynamic occupancy model as described by @mackenzie2003.
 The ecological process of occupancy is guided by the initial occupancy, colonisation, and extinction parameters; the observational process is guided only by the detection parameter.](Figures/ModelForm.png){#fig-modelform width="1000"}
 
-To disentangle the ecological and observation processes and estimate the detection parameter, the DOM relies on a hierarchical sampling design shown in @fig-surveys.
+To disentangle the ecological and observation processes and estimate the probability of detection, the DOM relies on a hierarchical sampling design based on Pollock's robust design [@pollock1982] and depicted in @fig-surveys.
 Under this design, observations at a site occur during distinct, time-bound seasons in which sites are considered 'closed' to changes in occupancy.
 Within seasons multiple observations are conducted, which permit estimation of the probability of detection conditional on occupancy.
-These repeat observations may be conducted by a variety of methods.
-Most frequently surveys are simply repeated on separate occasions, although alternative methods exist.
-These include surveys conducted at multiple locations within a site **\[cite\]**, use of multiple observers during a survey **\[cite\]**, or by recording time elapsed before a detection is recorded **\[cite\]**.
+Most frequently these observations are collected by revisiting the site on separate occasions, although alternative methods to generate repeat observations exist [@mackenzie2005].
+These alternatives include surveys conducted at multiple locations within a site, use of multiple observers during a survey, or recording of the time elapsed before a detection is recorded.
 
@@
 
-Dynamic occupancy models require a few assumptions with important implications for users.
+## Key assumptions
+
+Dynamic occupancy models require a small number of assumptions with important implications for users.
 
@@
 **False positive detections do not occur.** While this assumption can be safely met in some studies, it is not guaranteed when working with more cryptic species.
-Several authors have commented on the perils of not accounting for false positives where they occur (@miller2015, @mcclintock2010).
-Furthermore, @berigan2019 notes that in some cases genuine identifications of a species may be considered false positives if they do not represent the right kind of occupancy - in this case, transient individuals where the authors were interested in breeding occupany.
-Extensions to the dynamic occupancy model which account for false positive error are available (@royle2006, @miller2011, @miller2013) and should be carefully considered where this assumption cannot be met.
+Several authors have commented on the perils of not accounting for false positives where they occur [@miller2015; @mcclintock2010].
+Furthermore, @berigan2019 notes that in some cases genuine identifications of a species may be considered false positives if they do not represent the intended definition of occupancy - in the authors' case, transient individuals where the authors were interested in breeding occupancy.
+Extensions to the dynamic occupancy model which account for false positive error are available [@royle2006; @miller2011; @miller2013] and should be carefully considered where this assumption cannot be met.
 
 2.
-**Sites are closed to changes in occupancy between seasons.** This assumption, best known as the 'closure assumption,' has also been subject to considerable discussion on the bias which is introduced when it is violated (@rota2009, @otto2013).
+**Sites are closed to changes in occupancy between seasons.** This assumption, best known as the 'closure assumption,' has also been subject to considerable discussion on the bias which is introduced when it is violated [@rota2009; @otto2013].
 Closure is dependent not only on the behaviour of the species, but also the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discusses the difficultly in distinguishing temporary emigration and local extinction.
-Model extensions to relax the closure assumption have been developed, including @kendall2013 's approach using staggered arrival and depature periods between sites.
+Model extensions to relax the closure assumption have been developed, including @kendall2013 's approach using staggered arrival and departure periods between sites.
 More pertinent for most studies, however, is a careful consideration of ecologically and observationally relevant seasons which will allow the estimation of an appropriate 'type' of occupancy, whether that be shorter or longer term.
@@
 3.
-**No modelled heterogeneity in ecological or observational parameters.** Relative to the former two assumptions, considerably less research has been conducted on the impact of unmodelled heterogeneity on parameter estimates.
-Particularly at larger spatial scales, environmental factors are near-guaranteed to influence species occupancy and observation.
+**No modelled heterogeneity in ecological or observational parameters.** Relative to the former two assumptions, considerably less research has been conducted on the impact of modelled heterogeneity on parameter estimates [@mackenzie2017].
+Environmental factors are near-guaranteed to influence species occupancy and observation in any system, particularly at larger spatial scales.
+While studies should attempt to incorporate the key drivers of occupancy via inclusion of covariates, there is limited discussion on the best methods to achieve this and the implications of failure to capture this variation.
+
+## Model extensions
+
+The DOM has been frequently extended over the years, with alternate parameterisations available to better address many needs.
+In addition to the false-positive implementations previously discussed, prominent model extensions include multi-state DOMs which allow sites to exist in states other than occupied and unoccupied [@nichols2007], multi-scale models describing occupancy at multiple spatial scales [@nichols2008], and multi-species models permitting joint estimation of species with interactions [@devarajan2020].
+For an in-depth discussion of additional model extensions and their applications see @bailey2014.
 
@@
 
-This review is focused on the dynamic, multi-season, implementation of the occupancy model of @mackenzie2003.
-To qualify for inclusion articles must fit a model which fulfils these criteria: i) multiple sites which can exist in at least two states, including an occupied and unoccupied states; ii) multiple survey periods, between which sites may change state conditional on the prior survey period's occupancy state and at least one transition probability; iii) contains at least one detection parameter; iv) uses real (i.e., not simulated) field ecological data.
+## Article collection
 
-Papers were solicited for inclusion in two stages.
-In a preliminary analysis, four search strings ("dynamic occupancy model," "occupancy dynamics model," "incidence function model," and "multi-season occupancy model" were fed into Google Scholar **\[cite\]**.
+This systematic review is focused on the dynamic, multi-season, implementation of the occupancy model of @mackenzie2003.
+To qualify for inclusion articles must have fit a model which fulfils these criteria: i) describes multiple sites which can exist in at least two states, including an occupied and unoccupied states; ii) has multiple survey periods, between which sites may change state conditional on the prior survey period's occupancy state and at least one transition probability; iii) contains at least one detection parameter; and iv) uses real (i.e., not simulated) field ecological data.
+
+Papers were gather for inclusion in two stages.
+
+1.
+In a preliminary analysis, four search strings ("dynamic occupancy model," "occupancy dynamics model," "incidence function model," and "multi-season occupancy model" were entered into Google Scholar.
 Up to 100 articles were downloaded for each string in order of appearance.
 Articles were divided into four temporal strata: 2000-2005, 2006-2010, 2011-2015, and 2016-2021.
-For each combination of search string and year strata, either a random 25% or 5 articles (whichever was greater) were processed.
-Articles which did not meet inclusion criteria were replaced with another from the same search string and strata where possible.
+For each combination of search string and year strata, a random 25% or 5 articles (whichever was greater) was processed.
+Articles which did not meet inclusion criteria were replaced with another from the same search string and strata where possible.After completion of the preliminary sample authors recognised significant uncertainty in how the model of interest was described, particularly in their first decade of use, and noted that the incidence-function model is best interpreted as a distinct model class**.**
 
-After completion of the preliminary sample authors recognised significant uncertainty in how the model of interest was described, particularly in their first decade of use, and noted that the incidence-function model is a distinct model class **\[cite\].** To increase sample size, a complete list of articles which have cited @mackenzie2003 was downloaded from Web of Science.
+2.
+To increase sample size a secondary sample of articles was generated, using the complete list of articles which have cited @mackenzie2003 (downloaded via Web of Science).
 Articles in this list were split into the same year strata, plus an additional 2022-2023 category.
@@
 
-Details on methods for each article were recorded in detail, including the full suite of covariates considered for inclusion in modelling.
-The full spreadsheet of reviewed articles is available in **Appendix I**.
+## Article processing
+
+For each included article data was collected on authorship, stated research objectives, study taxa and field survey locations, data collection methods, covariates, and model selection and evaluation.
+A full spreadsheet of data collected is given in **Appendix I.**
+
+### Authorship
+
+To assess patterns of authorship, the location of each articles first author's first-listed affiliation was denoted.
+For *all* authors[,]{.underline} each affiliation was denoted as one of four categories: academic institutions, governmental organisations and institutes, non-governmental organisations, and private sector companies.
+
+### Objectives
+
+For each article five non-exclusive research objectives were indicated based on the authors' stated aims: 'Estimating parameters,' where authors express interest in estimates of site occupancy, colonisation, or extinction; 'Testing hypotheses,' where authors explore *specific* relationships between covariates and model parameters; 'identifying drivers,' where authors attempt to find *which* covariates influence model parameters; 'generating predictions,' where which authors estimate model parameters either for unsurveyed sites or under possible future conditions; and 'methods development,' where authors introduce or test aspects of dynamic occupancy models.
+
+### Taxa and study areas
+
+Focal taxa were divided into broad categories (bird, mammal, etc.).
+Taxa were denoted as threatened either where the authors refer to them as so or where the species is listed on the IUCN red list (IUCN 2024) - this determination was made to account for unlisted sub-species or species of local concern.
+Study areas are defined as the intended area of inference for a given study in which all sites are contained; size of these areas are reported as orders of magnitude in square kilometres to partly account for uncertainty in reporting.
+
+### Covariates
+
+Particular emphasis was placed on the covariates considered by authors, with no distinction between covariates which were or were not included in final models.
+For each parameter all covariates were recorded in broad mutually-excluded categories.
+For each of these covariates, data was collected on whether the covariate was remotely collected or required direct observation, whether polynomial terms were considered, whether interaction terms included this covariate were considered, and whether the covariate was dynamic – i.e, varied between seasons.
+
+### Model selection and evaluation
+
+To describe the complex model selection process, non-exclusive categories were applied to each article including: 'A priori,' where only one model is considered; 'candidate suite,' where a predefined set of models are considered; 'procedural,' where covariates are selected parameter-by-parameter; 'exhaustive,' where all possible model combinations are fit; 'simple precursors,' where selection is predicated on results of another, simpler model implementation; and 'other.'
+
+Model evaluation was similarly categorised - options in this case included 'relative fit,' for relative metrics of fit such as AIC; 'model averaging,' where model weights are applied to candidate models; 'goodness of fit,' where absolute metrics of fit are used; 'out-of-sample validation' and 'in-sample-validation,' where predictive performance is evaluated using in-sample or out-of-sample data respectively.
 
@@
 
-A total of **\[number\]** articles were included in the review, **\[number\]** from the preliminary sample and **\[number\]** from the secondary sample.
-Based on the acceptance rate from the secondary sample, an estimated **\[hit number\]** of the **\[total number\]** articles citing @mackenzie2003 would have met inclusion criteria; i.e., fit a dynamic occupancy model.
+A total of 84 articles were included in this review, 47 from the preliminary sample and 37from the secondary sample.
+Based on the acceptance rate from the secondary sample, an estimated 448 of the 1066 articles citing @mackenzie2003 would have met inclusion criteria.
 
-![](Figures/CoveragePlot.jpeg){#fig-coverage}
+![Estimated coverage of articles included in this review, estimated based upon the secondary sample's rate of inclusion for each strata.](Figures/CoveragePlot.jpeg){#fig-coverage}
 
-## Study details
+Articles were published with first-authors from a total of 15countries, with authors from the United States producing 64% of articles.
+86% of articles had at least one author based at an academic institution, 46% had authors at government institutions, and 25% had authors from non-profit organisations.
 
-Articles were published with first-authors from a total of **\[number\]** countries, with authors from the United States producing **\[zz\]**% of articles.
-**\[x\]**% of articles had at least one author based at an academic institution, **\[y\]**% had authors at government institutions, and \[**z\]**% had authors from non-profit organisations.
+## Study locations, focal species, and data collection
 
-Data for DOMs were collected from study areas on all continents and in **\[xx\]** countries, with the United States again representing a majority of the total articles published (@fig-StudyDetails a).
-These study areas ranged in size from less than 1 to over 1 million square kilometres
+![An overview of study areas, focal species, and data collection for reviewed articles.](Figures/ReviewBox.png){#fig-StudyDetails}
 
-![](Figures/ReviewBox.png){#fig-StudyDetails}
+Data for DOMs were collected from study areas on all continents and in 21 countries, with studies in the United States again representing a majority of the total articles published.
+These study areas ranged in size from less than 1 to over 1 million square kilometres (@fig-StudyDetails a).
 
-## Data collection
+Focal species were largely vertebrate taxa, although 6 articles fit models to invertebrates.
+A single non-animal taxa was represented in the sample, from an article fitting a DOM to a rare species of lichen @belinchón2017.
+35% of articles modelled threatened taxa, and 10% modelled species invasive to the study area.
+The majority of articles (50, 60%) fit single-species DOMs to one taxa.
+Of the remainder, 26 (31%) fit independent single-species models to more than one taxa.
+8 articles fit (10%) multi-species implementations with explicit interactions between species (@fig-StudyDetails b).
 
-## Modelling
+Data used to fit dynamic occupancy models included human survey data such as transects or point counts (69% of articles), live-trapping data (20%), camera traps (12%) or bioacoustic monitors (1%), and citizen science data (12%); some articles incorporate multiple detection methods to build their detection datasets.
+Study duration, defined as the time elapsed between the first and last surveys conducted, ranged from 1 month to 40 years with the median study running for 39 months.
+The number of seasons ranged from 2 to 204 with a median of 5; the number of observations per season ranged from 2 to 153 with a median of 5.
+The number of sites surveyed per article ranged from 7 to 9394, with the median study including 98 sites (@fig-StudyDetails c).
 
-![](Figures/CovParamPlot.jpeg){#fig-covariates}
+## Covariates
 
-## Selection
+A summary of covariates considered in the review sample is presented in @tbl-covariates.
+For each category of covariate relative frequency and characteristics are given.
+For all categories, the rate of consideration of polynomial terms does not exceed 16%, and the rate of consideration of interactions does not exceed 34%.
+
+::: column-page-inset
+```{r covariate table}
+#| echo: false
+#| message: false
+#| label: tbl-covariates
+#| tbl-cap: "Attributes of covariates found in dynamic occupancy models"
+
+library(tidyverse)
+library(gt)
+
+covTable <- read_csv("Figures/CovariateTable.csv")
+
+covTable |>
+  arrange(Category, -Overall_Proportion) |>
+  gt(rowname_col = "Type") |>
+  tab_header("Covariate attributes") |>
+  tab_row_group(
+    label = md("Environmental covariates"),
+    rows = Category == "Environmental",
+    id = "Env"
+  ) |>
+  tab_row_group(
+    label = md("Other covariates"),
+    rows = Category == "Other",
+    id = "Other"
+  ) |>
+  tab_style(
+    style = cell_fill(color = "#82AD8E"),
+    locations = cells_row_groups(groups = "Env")
+  ) |>
+  tab_style(
+    style = list(
+      cell_fill(color = "gray70")
+    ),
+    locations = cells_row_groups(groups = "Other")
+  ) |>
+  row_group_order(groups = c("Env", "Other")) |>
+  tab_spanner(label = "Articles with covariate on:",
+              columns = c("Overall_Proportion", "InitOcc_Proportion",
+                          "Colonisation_Proportion", 
+                          "Extinction_Proportion",
+                          "Detection_Proportion"),
+              id = "articles") |>
+  tab_spanner(label = "Covariates which are:",
+              columns = c(Proportion_Dynamic, Proportion_Observed,
+                          Proportion_Polynomial, Proportion_Interact),
+              id = "covariates") |>
+   cols_label(
+    Overall_Proportion = "Any parameter",
+    InitOcc_Proportion = "Initial Occ.",
+    Colonisation_Proportion = "Colonisation",
+    Extinction_Proportion = "Extinction",
+    Detection_Proportion = "Detection",
+    Proportion_Dynamic = "Dynamic",
+    Proportion_Observed = "Directly observed",
+    Proportion_Polynomial = "Polynomial",
+    Proportion_Interact = "Interactive"
+  ) |>
+  cols_hide(Category) |>
+  fmt_percent(decimals = 0) |>
+  sub_missing(missing_text = "-")|>
+  tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = Overall_Proportion,
+                                   rows = Overall_Proportion == max(Overall_Proportion, na.rm = TRUE))) |>
+    tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = InitOcc_Proportion,
+                                   rows = InitOcc_Proportion == max(InitOcc_Proportion, na.rm = TRUE))) |>
+    tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = Colonisation_Proportion,
+                                   rows = Colonisation_Proportion == max(Colonisation_Proportion, na.rm = TRUE))) |>
+    tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = Extinction_Proportion,
+                                   rows = Extinction_Proportion == max(Extinction_Proportion, na.rm = TRUE))) |>
+    tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = Detection_Proportion,
+                                   rows = Detection_Proportion == max(Detection_Proportion, na.rm = TRUE))) |>
+    tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = Proportion_Dynamic,
+                                   rows = Proportion_Dynamic == max(Proportion_Dynamic, na.rm = TRUE))) |>
+    tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = Proportion_Observed,
+                                   rows = Proportion_Observed == max(Proportion_Observed, na.rm = TRUE))) |>
+    tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = Proportion_Polynomial,
+                                   rows = Proportion_Polynomial == max(Proportion_Polynomial, na.rm = TRUE))) |>
+    tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = Proportion_Interact,
+                                   rows = Proportion_Interact == max(Proportion_Interact, na.rm = TRUE))) |>
+  as_raw_html()
+
+```
+:::
+
+Quantity of covariates considered varies considerably both between articles and between parameters as shown in @fig-covariates.
+Of the four standard model parameters, models were least likely to consider covariates for initial occupancy (35% of articles with zero covariates considered), and most likely to consider them for detection (81% of articles with at least one covariate considered).
+
+![Distribution of covariate quantity considered for model parameters across articles.](Figures/CovParamPlot.jpeg){#fig-covariates}
+
+Model complexity varies with the stated objectives of articles, with those interested in making predictions and identifying drivers of occupancy being more likely to consider a larger number of covariates (@tbl-objectives).
+These articles were also more likely to incorporate polynomial and interactive terms.
+Articles with objectives involving methods development considered fewer and less complex covariates.
+
+<div>
+
+```{r Objectives table}
+#| echo: false
+#| message: false
+#| label: tbl-objectives
+#| tbl-cap: "Complexity of models by article objective"
+
+objTable <- read_csv("Figures/ObjectiveTable.csv")
+
+objTable |>
+  mutate(Objective = fct(Objective,
+                         levels = c("Hypothesis", "Drivers", "Parameters",
+                                    "Methods", "Predict", "Any"))) |>
+  arrange(Objective) |>
+  mutate(Objective = case_when(Objective == "Hypothesis" ~ "Testing hypotheses",
+                               Objective == "Drivers" ~ "Identifying drivers",
+                               Objective == "Predict" ~ "Making predictions",
+                               Objective == "Methods" ~ "Developing methods",
+                               Objective == "Parameters" ~ "Estimating parameters",
+                               Objective == "Any" ~ "Any objective")) |>
+  gt(rowname_col = "Objective") |>
+  tab_header("Model complexity by article objectives") |>
+  tab_spanner(label = "Articles",
+              columns = c("Articles", "ArticlesProp"),
+              id = "articles") |>
+  tab_spanner(label = "Percent with:",
+              columns = c("ProbPoly", "ProbInt",
+                          "ProbEval", "ProbSelect"),
+              id = "percents") |>
+  cols_label(Articles = "n", ArticlesProp = "%",
+             MedianCovs = "Median covariates considered",
+             ProbPoly = "Polynomial terms",
+             ProbInt = "Interaction terms",
+             ProbEval = "Objective evaluation",
+             ProbSelect = "Model selection") |>
+  fmt_percent(columns = c(ArticlesProp, 
+                          ProbPoly, ProbInt, 
+                          ProbEval, ProbSelect),
+              decimals = 0) |>
+  tab_style(style = list(cell_fill(color = "gray80"),
+                         cell_text(style = "italic")),
+            locations = cells_body(rows = Objective == "Any objective")) |>
+  tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = MedianCovs,
+                                   rows = MedianCovs == max(MedianCovs, na.rm = TRUE))) |>
+  tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = ProbPoly,
+                                   rows = ProbPoly == max(ProbPoly, na.rm = TRUE))) |>
+  tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = ProbInt,
+                                   rows = ProbInt == max(ProbInt, na.rm = TRUE))) |>
+  tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = ProbEval,
+                                   rows = ProbEval == max(ProbEval, na.rm = TRUE))) |>
+  tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = ProbSelect,
+                                   rows = ProbSelect == max(ProbSelect, na.rm = TRUE))) |>
+  as_raw_html()
+```
+
+</div>
+
+Bayesian models were far less likely to implement a model selection protocol compared to frequentist models: 70% of the former fit only a single model defined *a priori,* compared to 10% of the latter (@fig-selection).
+In frequentist models, procedural selection protocols in which parameters are fixed one by one were the most popular choice.
+Nearly all frequentist models used AIC or AICc to identify their best model.
 
@@
 
-### Complexity
-
-![](Figures/ModelComplexity.png){#fig-complexity}
+Objective evaluation of model fit was rare throughout the sample regardless of selection protocol.
+Just 8% of articles calculated non-relative metrics, either via in-sample validation, bootstrapped goodness-of-fit tests, or out-of-sample validation.
 
@@
 
-**\<Covariate inclusion figure\>**
+### Applications of dynamic occupancy models
 
-We provide guidelines for fitting dynamic occupancy models.
+DOMs display great variation of scale in all senses.
+Analyses in our samples range from the hyper-local at individual parks [@zúñiga-vega2019] to continental-scale studies \[@zuckerberg2011\].
+Similar diversity in study duration and site quantity indicates further emphasises the breadth of studies which DOMs may be fit to.
+Method of species detection also need not be a limitation for the implementation of DOMs, as articles in our sample show that many approaches may be coerced into data suitable for this model class.
+While studies pre-designed to match hierarchical data inputs may be simplest to align with model assumptions, post-hoc manipulation of other data such as long-term citizen monitoring programs \[@zuckerberg2011; @peach2019\], disparate agency modelling programs [@mcgowan2020], or camera trapping data [@davis2018] is feasible with careful consideration.
+Detections from multiple sources, with associated variation in efficiency, have also been incorporated by authors [@warrier2020; @pitman2017].
 
-![](Figures/Guidelines.png){#fig-guidelines}
+Reviewed implementations of dynamic occupancy models show widespread representation from mammals, birds, and herptiles.
+This assemblage contrasts with the related stochastic patch occupancy models (SPOMs, @gutiérrez-arellano), where the majority of models are fit to invertebrates, plants, and fungi.
+This difference may reflect SPOMs' assumption of perfect detection and discrete habitat patches, limitations which are unlikely to be fulfilled by most vertebrate taxa.
+Aside from differences in assumptions around detection, SPOMs also differ from DOMs in that they are generally spatially explicit, with probabilities of colonisation and extinction dependent on patch size and the occupancy state of nearby patches.
+Where desired, these features may also added to dynamic occupancy models — @sutherland2014 presents an implementation which may be considered a fusion of the DOM and SPOM methods, and several other authors incorporate patch characteristics aspects into their models as covariates [@duggan2011; @broms2016].
 
-#
+While objectives vary in our sample, DOMs are rarely used to extrapolate beyond study areas — use of prediction remains the least popular use-case in our sample.
+Articles which do make predictions provide some of the most directly applicable outputs found in the review sample: @mcgowan2020 provides projections for a threatened species under multiple putative management scenarios, and @pollentier2021 generates maps of distributions more frequently associated with species distribution models (SDMs).
+DOMs offer advantages over other SDMs for this purpose, particularly where populations are assumed not to be at equilibrium as is the case for many threatened and invasive species.
+By explicitly accounting for changes in occupancy (i.e., colonisation and extinction) DOMs may better capture future range contraction or expansion — an aspect of prediction where other SDMs may struggle.
+
+### Practices in the modelling process
+
+A diverse set of covariates were identified in the model sample reflecting the diversity of systems and ecological questions described by DOMs.
+Notably, many models lack covariates on the initial occupancy parameter, raising questions on how the parameterisation of the model in it's first time-step may affect the quality of estimates at later time-steps.
+The distinction in covariates considered for the ecological parameters vs those considered for detection is also of note: ecological parameters more frequently contained covariates describing environmental features, where detection more frequently contained more structural covariates such as season.
+Covariate complexity is relatively low across the board, with only a fraction of models implementing either non-linear responses to covariates or interactions between covariates.
+While this addition complexity does have implications for model selection, it may be necessary to accurately describe many ecological relationships.
+
+Model selection approaches diverge between Bayesian and frequentist implementations of the model.
+While model selection will naturally be practised differently between these two categories, it is unclear what effect a lack of model selection may have on the outputs produced by Bayesian models, and it is likely that Bayesian model selection approaches should be further explored @hooten2015.
+Model selection approaches in the Frequentist applications show limited consensus, although most do consider multiple models.
+Notably, @morin2020 found that the popular procedural approaches to model selection (where one sub-model is assessed at a time) may not provide optimal results.
+Considerably more research is required on this topic to clarify how authors should best approach covariate selection.
+
+### Key considerations for model users
+
+Implementation of dynamic occupancy models requires considerable thought to ensure that model outputs correspond with author expectations and ecological realities.
+Based upon our review of the existing literature we discuss key questions authors should carefully engage with during the modelling process.
+
+1.
+**Are seasons ecologically meaningful, and how do they affect the definition of occupancy?** As previously discussed, the duration of the season has important implications for both the reliability and proper interpretation of model results.
+Seasons which are not ecologically meaningful may run afoul of the closure assumption resulting in associated biases.
+Also relevant is how the definition of 'occupancy' may shift with season length and life-history of a species — for many species, occupancy within a year and within a week may be considerably different concepts, with shorter intervals more closely relating to the idea of 'use.' With the proliferation of continuous detection methods such as camera traps and acoustic monitors where the definition of 'season' is to some degree arbitrary, it is increasingly important to thoughtfully consider how this element of the DOM is defined.
+
+2.
+**Are key drivers of occupancy and detection incorporated?** While it is effectively impossible to *fully* describe heterogeneity across a landscape, authors should make their best efforts to capture the drivers of occupancy and detection most likely to influence their study species.
+This is true even when another covariate is of principal interest to investigators — failure to include other major determinants of occupancy is likely to bias parameter estimates.
+One place where this is of particular importance is in determining drivers of detection, which is not only subject to observational conditions but also ecological determinants which may cause the species to spend more time at a location and thus be more frequently available for detection.
+
+3.
+**Ensure covariates reflect ecological complexity.** Many ecological relationships are non-linear and may require polynomial terms to be accurately represented.
+This is particularly true at large spatial scales where a larger portion of a species ecological niche is represented in surveyed data.
+Authors should question whether a linear term is sufficient to capture hypothetical relationships in their study systems, and consider quadratic terms as well as interactions between covariates where appropriate.
+
+4.
+**Consider model evaluation.** Model evaluation remains rare in ecological research relative to other fields embracing complex modelling, likely due to a reluctance to 'throw away' data for use in testing.
+This concept should be challenged, particularly for larger scale long-term projects where reserving a fraction of data for testing is unlikely to substantially reduce precision of parameter estimates.
+For studies where this is genuinely not possible, reporting results of goodness-of-fit tests may still provide valuable information to readers.
+
+### Priorities for future research
+
+Work on dynamic occupancy modelling over the past 20 years has developed this model class into an important tool for describing species distributions and ecological relationships.
+Historically, methods-development research for dynamic occupancy models has focused on questions of model structure — namely ideas around the closure assumption and imperfect detection.
+This emphasis likely reflects the DOM's roots in the mark-recapture literature, where these topics have been discussed for several decades.
+Considerably less emphasis has been placed on the relation of model parameters to environmental factors via covariates and the model selection processes used to do so, despite this being an objective in a considerable fraction of studies fitting DOMs.
+This stands in stark contrast to related correlative species distribution models, where these topics have been the subject of considerable research over the years.
+There are opportunities for the DOM literature to learn from species distribution models, and more targeted research on the influence of the model selection process is necessary.
+
+An important opportunity for future research relates to autonomous detection methods, such as camera traps and bio-acoustic monitors — these technologies have permitted collection of enormous quantities of data particularly well-suited to dynamic occupancy models, but have also raised additional questions around their implementation.
+The definition of seasons is of increased importance for these data types, where division of data into 'seasons' and 'observations' is arbitrary.
+Further research should explore the implications of this phenomenon and the additional modelling opportunities made possible by this novel data type.
+
+The past two decades have shown that dynamic occupancy models are an excellent tool for addressing a wide range of ecological questions relating to species occupancy throughout space and time.
+The flexibility afforded by model extensions and the considerable advantages of the class over other species distribution models is likely to ensure that these models continue to increase in popularity.
+Given their strong use on threatened taxa and excellent potential to support environmental management, it is important that research continues to support model users in creating the most reliable models possible.

### 2024-06-11 Manuscript/domReview.qmd
Changes to introduction
Added more general introduction to species distributions

@@
 
-embed-resources: true
 bibliography: references.bib
@@
 # Introduction
+
+Ecologists have long sought to understand the processes by which species are distributed across landscapes and throughout time \[@humboldt1849\].
+Research on these patterns of occurrence is valuable in many ways — simply describing where a species occurs may be of use in monitoring programs, and identifying the drivers of distribution can help to determine how best to manage or conserve a species.
+
+In recent decades, interest in distributions has precipitated numerous methods for quantitatively modelling species distributions \[@elith2009\].
+In their typical form species distribution models (SDMs) correlate species occurrence records with environmental covariates to describe where species occur on the landscape.
+They have been implemented in several thousand scientific articles and are frequently relied upon to support management and conservation programs.
+Despite their popularity SDMs do have considerable downsides.
+Typically, they do not explicitly account for imperfect detection, where a species is present at a site but is not detected.
+Imperfect detection is known to lead to biased estimates of occupancy and is a ubiquitous and near-unavoidable feature of wildlife presence-absence data.
+SDMs also tend to be most effective when species are at equilibrium states, and can struggle to predict to novel environments or future conditions.
+
+Another approach to modelling species occurrence is the occupancy model, first introduced in @mackenzie2002.
+Like SDMs, occupancy models estimate probabilities of species occurrence; unlike SDMs, occupancy models account for imperfect detection by explicitly modelling the process of detection.
+@mackenzie2003 extends the original occupancy model to describe change in species occurrence across multiple time-steps.
+Despite the similarity in their output, SDMs and occupancy models have developed largely distinctly.
 
@@
 DOMs display great variation of scale in all senses.
-Analyses in our samples range from the hyper-local at individual parks [@zúñiga-vega2019] to continental-scale studies \[@zuckerberg2011\].
+Analyses in our samples range from the hyper-local at individual parks [@zúñiga-vega2019] to continental-scale studies [@zuckerberg2011].
 Similar diversity in study duration and site quantity indicates further emphasises the breadth of studies which DOMs may be fit to.
 Method of species detection also need not be a limitation for the implementation of DOMs, as articles in our sample show that many approaches may be coerced into data suitable for this model class.
-While studies pre-designed to match hierarchical data inputs may be simplest to align with model assumptions, post-hoc manipulation of other data such as long-term citizen monitoring programs \[@zuckerberg2011; @peach2019\], disparate agency modelling programs [@mcgowan2020], or camera trapping data [@davis2018] is feasible with careful consideration.
+While studies pre-designed to match hierarchical data inputs may be simplest to align with model assumptions, post-hoc manipulation of other data such as long-term citizen monitoring programs [@zuckerberg2011; @peach2019], disparate agency modelling programs [@mcgowan2020], or camera trapping data [@davis2018] is feasible with careful consideration.
 Detections from multiple sources, with associated variation in efficiency, have also been incorporated by authors [@warrier2020; @pitman2017].

### 2024-06-20 Manuscript/_manuscript/domReview.qmd
Major changes to manuscript structure

---
title: "Twenty years of dynamic occupancy models: a review of applications and a look towards the future"
date: today

authors:
  - name: Saoirse Kelleher
    affiliation: University of Melbourne
    email: saoirse.kelleher@student.unimelb.edu.au
    orcid: 0000-0002-4614-5753
    corresponding: true
  - name: Natalie Briscoe
    affiliation: University of Melbourne
  - name: Gurutzeta Guillera-Arroita
    affiliation: Pyrenean Institute of Ecology, Spanish National Research Council
  - name: Jane Elith
    affiliation: University of Melbourne

abstract: | 
  Describing patterns of species occupancy across landscapes and throughout time is a fundamental aspect of much ecological research. The dynamic multi-season occupancy model is an important tool for analysing wildlife occurence data -- by explicitly estimating colonisation and extinction probabilities while accounting for imperfect detection, it strikes a favourable balance between realism and feasibility. These models can be used in similar ways to more conventional species distribution models, which can struggle to deal with imperfect detection and non-equilibrium populations.

  We present a review of applications of the dynamic occupancy model in the twenty years since its initial publication. Our findings indicate that this model demonstrates exceptional flexibility with suitability to address a wide range of ecological questions, and that it may have  broader potential than commonly thought. We also note that practices in fitting dynamic occupancy models are highly variable, and that the existing literature lacks clear answers for best practices. This contrasts strongly with the robust species distribution model literature on modelling practices, which we suggest may be a valuable resource for improving how authors implement dynamic occupancy models.

bibliography: references.bib
---

# Introduction

Ecologists have long sought to describe the processes by which species are distributed across landscapes and throughout time [@humboldt1849]. An understanding of where a species exists, which factors lead to its presence, and how its distribution may change has always been of interest and is ever more important in a dynamic and changing world. Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and are imperative for assessing conservation status, and further untangling the drivers of occurrence can help us to better understand a species life history and inform potential management actions. A robust understanding of the occupancy patterns of a species also helps us to predict where a species is most likely to occur — now, and under hypothetical future scenarios.

The need to reliably describe species occupancy has led to the development of numerous quantitative approaches, of which the best known are species distribution models [SDMs, @elith2009]. The prototypical SDM correlates records of species occurrence with environmental features to determine where species are most probable to occur. While SDMs are an exceedingly popular tool for describing occupancy, they do have considerable downsides. First and foremost, SDMs do not typically account for imperfect detection, a near-ubiquitous characteristic of wildlife field data that can lead to meaningful bias in estimates of occupancy [@lahoz-monfort2014; @guillera-arroita2017; @gu2004]. Secondly, SDMs have been noted to perform poorly when a species' distribution is not at equilibrium [@elith2010]. This may be problematic when working with invasive species or species whose ranges are expanding or contracting, all of which are increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].

The dynamic occupancy model (DOMs, also termed 'multi-season occupancy models') introduced in @mackenzie2003 provide an alternative framework to modelling occurrence and distributions. By explicitly describing the colonisation and extinction processes which guide occupancy while accounting for imperfect detection, the model class offers more advanced insights than simpler static species distribution models; despite this, input data requirements remain achievable compared to other more complex process-explicit models. Dynamic occupancy models unite two processes: the ecological process of site occupancy describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed given that a site is occupied (@fig-modelform). Under this parameterisation, sites exist in either occupied or unoccupied states. The occupancy state for each step in the first time-step is given by the *initial occupancy* parameter, and in ensuing time-steps occupancy is determined based on probabilities of *colonisation* and *extinction.* Given that a site is occupied, the probability of observing the species during a survey is given by the *detection* parameter. In the DOM's standard form, it is assumed that false-positive detections do not occur.

![The original parameterisation of the dynamic occupancy model as described by @mackenzie2003. The occupancy state of any given site is described in the ecological process. In the first season, occupancy is determined by the *initial occupancy* parameter. In subsequent season, unoccupied sites may become occupied according to the colonisation probability and occupied sites may become unoccupied according to the extinction probability. \n The observational process accounts for imperfect detection: during a survey at sites which are truly occupied the species may be either detected or not detection according to detection probability. At sites which are truly unoccupied, the species is never detected – that is, it is assumed that false positive detections do not occur.](Figures/ModelForm.png){#fig-modelform width="1000"}

To distinguish between the ecological and observation processes and estimate the probability of detection, the DOM employs on a hierarchical sampling design depicted in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons in which sites are considered closed to changes in occupancy. Within each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy. Most frequently these observations are collected by revisiting the site on separate occasions, although alternative methods to generate repeat observations exist such as conducting surveys at multiple locations within a site, using of multiple observers during a survey, or recording the time elapsed until a detection is recorded [@mackenzie2005].

![The sampling design of the standard dynamic occupancy model. During seasons, also called 'primary occasions', sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons. Within each season, multiple observations ('secondary occasions') are conducted to record the observed presence or absence of the species at each site. These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded. It is not necessary for the same number of observations to occur in each year or each site, allowing for greater flexibility in data inputs.](Figures/SurveyDesign.png){#fig-surveys fig-align="center" width="589"}

We have presented the standard form of the dynamic occupancy model as described in @mackenzie2003 — numerous model extensions and alternative formulations are also available, including implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models[@dorazio2010]. For a comprehensive discussion of the most common extensions and their applications see @bailey2014, as well as @devarajan2020 for a review of multi-species occupancy models.

Dynamic occupancy models make a small number of assumptions with important implications. Firstly, it is assumed that false positive detections do not occur — while this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods. Authors have commented on the bias induced when false positives occur and are not accounted for [@mcclintock2010; @miller2015]. Additionally, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019]. Where this assumption can not reasonably met, the aforementioned extensions designed to account for false positive error should be considered.

A second assumption states that sites are closed to changes in occupancy between seasons. This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion on the bias which is introduced when it is violated [@rota2009; @otto2013]. Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discuss the difficultly in distinguishing temporary emigration and local extinction. Model extensions to relax the closure assumption have been developed, including @kendall2013's approach using staggered arrival and departure periods between sites. More pertinent for most studies, however, is a careful consideration of ecologically and observationally relevant seasons which will allow the estimation of an appropriate 'type' of occupancy, whether that be shorter or longer term.

Generally speaking, DOMs may be applied in much the same ways as SDMs. Comparable probabilities of occupancy may be estimated, relationships to environmental covariates can be quantified, and predictions of occupancy across space and into the future can be generated. Despite these similarities, the two models have developed largely distinctly with key differences in directions within their respective research communities. DOMs originated from the mark-recapture literature, with similarities to models such as the Cormack-Jolly-Seber model [@pollock1990]. This history contributes to an emphasis on imperfect detection and survey design in the DOM literature. In contrast, other areas of the modelling process have received comparatively little attention. For any model estimating occupancy, key environmental drivers must be adequately represented using covariates — the SDM community has long debated how to best approach model selection and evaluation [@guisan2006; @araújo2006; @jarnevich2015; @araújo2019]. The DOM literature has by comparison focused less on these aspects of modelling [@mackenzie2017], although what research does exists indicates that choice of model selection approach can impact model performance [@morin2020].

This review aims to provide an overview of the dynamic occupancy model's first two decades of use. We present a systematic review exploring how authors have used the DOM to answer ecological questions, with emphasis on how they have collected data, selected covariates, and evaluated their models. Based on these results we highlight the DOMs flexibility as a tool for understanding occupancy and discuss common misconceptions regarding their use. We also examine how authors have approached the modelling process and discuss areas where lessons can be drawn from the SDM literature. Finally, we present considerations for authors seeking to use DOMs in their work and outline key priorities and opportunities for future research involving this important model class.

# Systematic review methods

Our review focuses on the dynamic, multi-season implementation of the occupancy model of @mackenzie2003. To meet inclusion criteria articles must have fit a model which i) describes multiple sites which can exist in at least two states, including an occupied and unoccupied states; ii) has multiple survey periods, between which sites may change state conditional on the prior survey period's occupancy state and at least one transition probability; iii) contains at least one detection parameter; and iv) uses real (i.e., not simulated) field ecological data. Papers were gather for inclusion in two stages:

In a preliminary analysis, four search strings ("dynamic occupancy model," "occupancy dynamics model," "incidence function model," and "multi-season occupancy model" were entered into Google Scholar. Up to 100 articles were downloaded for each string, in order of appearance. These articles were divided into four strata based on data of publication — 2000-2005, 2006-2010, 2011-2015, and 2016-2021. For each combination of search string and year strata, a random selection of the greater of 25% of articles or 5 articles was processed. Where available, articles which did not meet inclusion criteria were replaced with another from the same search string and strata.

After completion of the preliminary sample, we noted that the articles from the 'incidence function model' search term represented a distinct model class with its own history and features [@gutiérrez-arellano]. Articles fitting this model were dropped. We also noted that early DOMs used varied terminology to refer to the model and that very few articles were identified from the first few years after their introduction. To better capture these early models and increase the size of our sample, a secondary sample was collected from the complete list of articles which have cited @mackenzie2003, accessed via Web of Science. Articles from this list were split into the same year strata as the first sample, plus an additional 2022-2023 category. 15 articles were drawn from each strata and processed, with replacement from within their strata where necessary.

For each article we recorded key details on authorship, research objectives, study taxa and system, survey methods, and modelling approach. To assess patterns of authorship, the location of each article's first author's primary affiliation was recorded. *All* affiliations for all authors were classified into one of four categories: academic institutions, governmental organisations and institutes, non-governmental organisations, and private sector companies.

We recorded details on the type of taxa (bird, mammal, etc.) modelled in each article, and how multiple species were modelled where applicable. Taxa were denoted as threatened either when they are listed on the IUCN Red List of Species or when authors indicate that they are threatened. This latter determination was made to account for unlisted sub-species or species of local concern. Study locality and size was documented; the size of the study area being defined as the intended area of inference containing all sites, recorded as an order of magnitude to partially account for uncertainty in reporting.

To examine the reasons why authors used DOMs, we allocated each article to one or more category of objective based on the study's stated aims. The possible categories were i) *Estimating parameters*, where authors express interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; ii) *Testing hypotheses*, where authors explore specific predefined relationships between covariates and model parameters; iii) *identifying drivers*, where authors attempt to find which covariates influence model parameters; iv) *generating predictions*, where authors estimate model parameters either for unsurveyed sites or under possible future conditions; and v) *methods development*, where authors introduce, test, or demonstrate aspects of dynamic occupancy models.

Particular emphasis was placed on describing how authors conducted their modelling from covariate selection through model evaluation. We recorded all covariates considered in each study, regardless of whether they were or were not included in final models. Data was collected for each covariate including a general classification, whether it was directly observed or remotely sensed, whether it was static or varied between seasons, and how it was represented in the model - as a linear term, a polynomial term, or as part of an interaction with another covariate. Model selection procedures were sorted into non-exclusive categories including *A priori*, where only one model was considered; *candidate suite*, where a predefined set of models was considered; *procedural*, where covariates were selected parameter-by-parameter; *exhaustive*, where all possible model combinations were fit; *simple precursors*, where selection was preceded by another simpler model implementation; and *other*. Model evaluation was similarly categorised - options in this case included *goodness of fit*, where absolute metrics of fit were used; and *out-of-sample validation* and *in-sample-validation*, where predictive performance was evaluated using in-sample or out-of-sample data respectively.

For the full spreadsheet of data collected and further details on categorisation, see **Appendix I.**

A total of 84 articles were included in this review, 47 from the preliminary sample and 37 from the secondary sample. Based on the acceptance rate from the secondary sample, an estimated 448 of the 1066 articles citing @mackenzie2003 would have met inclusion criteria (@fig-coverage).

![Estimated coverage of articles included in this review, based on the secondary sample's rate of inclusion for each strata.](Figures/CoveragePlot.jpeg){#fig-coverage}

# Applications of dynamic occupancy models

One of the strengths of the dynamic occupancy model is its considerable flexibility across several axes — they may used to achieve several different objectives, fit to diverse forms of occurrence data, and applied to studies at a wide range of spatial and temporal scales. This flexibility is reflected in the types of authors who conducted the studies captured in our sample: while 86% of articles had at least one author based at an academic institution, 46% had author(s) at government institutions and 25% had author(s) from non-profit organisations. The high proportion of articles with government and NGO participants indicates a more applied focus for many users of DOMs, suggesting that model outputs may directly guide management or conservation actions in many cases. The most frequent objective for studies using DOMs was hypothesis testing, with 50% of authors expressing interest in a specific relationship. 31% were interested in identifying nondescript drivers of occupancy, 30% in estimating models parameters like occupancy themselves, and 29% worked to develop and expand on methods for DOMs. Just 10% of authors expressed interest in making predictions to unsurveyed locations or into the future.

Study sites were located in all continents except Antarctica and in 21 countries, with the majority occurring in the United States of America (@fig-StudyDetails A). This strong geographic trend stands in stark contrast with SDMS — @araújo2019's review found that SDMs are far more international, with no continent yielding more than 20% of studies. The reasons behind this disparity are unclear, but one possible explanation may be the DOMs roots in the mark-recapture literature and its history of publication in American journals. These factors may also contribute to the seemingly limited overlap between the bodies of literature for SDMs and DOMs.

Differences in use between SDMs and DOMs are also evident in the species targeted for modelling. Where SDMs show broader coverage of plants, invertebrates, and aquatic species [@araújo2019], 93% of studies in our sample of DOMs modelled terrestrial vertebrate fauna with @belinchón2017's study on lichen the sole non-animal taxa representative (@fig-StudyDetails A). These are also taxa where the DOM's advantages, namely their handling of imperfect detection, are most applicable. One group for which DOMs have only rarely been used is aquatic species, as in @fisher2014's application to invasive salmon or @falke2012's model on Great Plains stream fishes.

40% of the studies in our sample fit models to more than one taxa ('taxa' defined here as one or more species modelled as one). Of these, 75% fit DOMs to multiple species, and 25% fit explicitly multi-species implementations (@fig-StudyDetails B). These multi-species models varied in structure, some taking hierarchical approaches with species-specific effects [@mordecai2009; @lee2017] and others explicitly modelling interactions between species [@lesmeister2015; @fidino2019]. For a more comprehensive discussion on multi-species occupancy modelling, see @devarajan2020's review.

\< FLAG \> discuss how it can be used for large scale monitoring type things where you need to model tons of species — this data already exists for a lot of places. flag paper that fit 300 species, note that there are questions about how to appropriately model them.

One notable finding of our review is the diversity of scale in studies where DOMs have been used. The smallest study area in our sample was less than one kilometre squared, and the largest over one million. Study duration shows similar variation — data collection ran for for less than a month in the shortest study, and for 40 years in the longest, with a median duration of 3.25 years. This reiterates the flexibility of dynamic occupancy models, with explorations of hyper-local patterns of occupancy and analyses of continental-scale distribution using the same underlying model structure. The detection data collected from these surveys came from a variety of detection methods: 69% of articles used human observations, 20% conducted live trapping, 12% used camera traps, and 1% used bioacoustic monitors (@fig-StudyDetails); some also incorporated data from multiple detection sources.

![A) Study areas: Data for a majority of studies were collected from study locations in the United States. The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres. \n B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles. 50 articles fit models to only a single taxa, 26 fit independent models to multiple taxa, and 8 fit multi-species models with explicit interactions between taxa. \n C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors. 10 articles incorporated data from citizen science projects. Project scale was also variable, with the median study running for 3.25 years and covering 98 sites.](Figures/ReviewBox.png){#fig-StudyDetails}

It is important to note that the data input requirements for dynamic occupancy models are not as arduous as may be commonly perceived. While it *is* necessary to record multiple observations at individual sites, this does not necessarily require that data be collected according to a strict, preconceived sampling scheme. Post-hoc manipulation of other data such as long-term citizen monitoring programs [@zuckerberg2011; @peach2019], disparate agency modelling programs [@mcgowan2020], or camera trapping data [@davis2018] is feasible with careful consideration. In one example, @riddell2021 used data from as long ago as 1908 by treating grid cells as sites and counting historical surveys within each cell as repeat observations. In a similar vein, data does not necessarily need to come from the same source. Detections from multiple methods, with associated variation in detection probabilities, have also been incorporated by authors [@warrier2020; @pitman2017].

The proliferation of modern survey techniques like camera traps and bioacoustic monitors has also led to an increase in datasets suitable for the dynamic occupancy model. The DOM is a valuable tool for analysing these data [@balantic2019]

Modern survey techniques like camera traps and bioacoustic monitors create other opportunities for modelling, and questions remain on the best ways to approach handling such large quantities of data. One interesting possibility is the use of large numbers of extremely short seasons, which is rarely feasible when using human survey data — @kleiven2020 and @mölle2022 divided their camera trap records into seasons of just a few days, modelling patterns of occupancy at a fine temporal scale. It is important to note that this substantially alters how model outputs should be interpreted, and a larger discussion on this topic is necessary.

The DOMs ability to make predictions has been often underexploited in the existing literature, given its broad potential for the use-case [@kéry2013; @briscoe2021]. Articles which do make predictions provide some of the most directly applicable outputs found in the review sample: @mcgowan2020 provides projections for a threatened species under multiple putative management scenarios, and @pollentier2021 generates maps of distributions more frequently associated with species distribution models (SDMs).

# Current modelling practices

Reviewed articles have approached the modelling pro

A summary of covariates considered in the review sample is presented in @tbl-covariates. For each category of covariate relative frequency and characteristics are given. For all categories, the rate of consideration of polynomial terms does not exceed 16%, and the rate of consideration of interactions does not exceed 34%.

::: column-page-inset
```{r covariate table}
#| echo: false
#| message: false
#| label: tbl-covariates
#| tbl-cap: "All covariates considered for inclusion in a study were classified into mutually exclusive categories. For each parameter (initial occupancy, colonisation, extinction, and detection), the percentage of studies including at least one covariate in a given category is presented. We also present the percentage of covariates in each category which are dynamic (change throughout seasons) or require direct observation (e.g. are not remotely sensed). The percentage of covariates in each category for which non-linear representations or interactions with other covariates were considered is given to indicate which were most likely to be included as more complex terms."

library(tidyverse)
library(gt)

covTable <- read_csv("Figures/CovariateTable.csv")

covTable |>
  arrange(Category, -Overall_Proportion) |>
  gt(rowname_col = "Type") |>
  tab_header("Covariate attributes") |>
  tab_row_group(
    label = md("Environmental covariates"),
    rows = Category == "Environmental",
    id = "Env"
  ) |>
  tab_row_group(
    label = md("Other covariates"),
    rows = Category == "Other",
    id = "Other"
  ) |>
  tab_style(
    style = cell_fill(color = "#82AD8E"),
    locations = cells_row_groups(groups = "Env")
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "gray70")
    ),
    locations = cells_row_groups(groups = "Other")
  ) |>
  row_group_order(groups = c("Env", "Other")) |>
  tab_spanner(label = "Articles with covariate on:",
              columns = c("Overall_Proportion", "InitOcc_Proportion",
                          "Colonisation_Proportion", 
                          "Extinction_Proportion",
                          "Detection_Proportion"),
              id = "articles") |>
  tab_spanner(label = "Covariates which are:",
              columns = c(Proportion_Dynamic, Proportion_Observed),
              id = "covariates") |>
  tab_spanner(label = "Represented as:",
                columns = c(Proportion_Polynomial, Proportion_Interact),
                id = "representations") |>
   cols_label(
    Overall_Proportion = "Any parameter",
    InitOcc_Proportion = "Initial Occ.",
    Colonisation_Proportion = "Colonisation",
    Extinction_Proportion = "Extinction",
    Detection_Proportion = "Detection",
    Proportion_Dynamic = "Dynamic",
    Proportion_Observed = "Directly observed",
    Proportion_Polynomial = "Non-linear",
    Proportion_Interact = "Interaction"
  ) |>
  cols_hide(Category) |>
  fmt_percent(decimals = 0) |>
  sub_missing(missing_text = "-")|>
  tab_style(style = cell_fill(color = "lavender"),
            locations = cells_body(columns = Overall_Proportion,
                                   rows = Overall_Proportion == max(Overall_Proportion, na.rm = TRUE))) |>
    tab_style(style = cell_fill(color = "lavender"),
            locations = cells_body(columns = InitOcc_Proportion,
                                   rows = InitOcc_Proportion == max(InitOcc_Proportion, na.rm = TRUE))) |>
    tab_style(style = cell_fill(color = "lavender"),
            locations = cells_body(columns = Colonisation_Proportion,
                                   rows = Colonisation_Proportion == max(Colonisation_Proportion, na.rm = TRUE))) |>
    tab_style(style = cell_fill(color = "lavender"),
            locations = cells_body(columns = Extinction_Proportion,
                                   rows = Extinction_Proportion == max(Extinction_Proportion, na.rm = TRUE))) |>
    tab_style(style = cell_fill(color = "lavender"),
            locations = cells_body(columns = Detection_Proportion,
                                   rows = Detection_Proportion == max(Detection_Proportion, na.rm = TRUE))) |>
    tab_style(style = cell_fill(color = "lavender"),
            locations = cells_body(columns = Proportion_Dynamic,
                                   rows = Proportion_Dynamic == max(Proportion_Dynamic, na.rm = TRUE))) |>
    tab_style(style = cell_fill(color = "lavender"),
            locations = cells_body(columns = Proportion_Observed,
                                   rows = Proportion_Observed == max(Proportion_Observed, na.rm = TRUE))) |>
    tab_style(style = cell_fill(color = "lavender"),
            locations = cells_body(columns = Proportion_Polynomial,
                                   rows = Proportion_Polynomial == max(Proportion_Polynomial, na.rm = TRUE))) |>
    tab_style(style = cell_fill(color = "lavender"),
            locations = cells_body(columns = Proportion_Interact,
                                   rows = Proportion_Interact == max(Proportion_Interact, na.rm = TRUE))) |>
  as_raw_html()

```
:::

The number of covariates considered varied considerably, both between studies and between parameters as shown in @fig-covariates. Of the four standard model parameters, models were least likely to consider covariates for initial occupancy (35% of articles with zero covariates considered), and most likely to consider them for detection (81% of articles with at least one covariate considered).

![Distribution of covariate quantity considered for model parameters across articles.](Figures/CovParamPlot.jpeg){#fig-covariates}

Model complexity varies with the stated objectives of articles, with those focussed on making predictions and identifying drivers of occupancy being more likely to consider a larger number of covariates (@tbl-objectives). These articles were also more likely to represent covariates in more complex ways by considering non-linear responses and interactions between covariates. Articles with objectives involving methods development considered fewer covariates and were less likely to use these more complex formulations.

<div>

```{r Objectives table}
#| echo: false
#| message: false
#| label: tbl-objectives
#| tbl-cap: "Implementation of models by study objective. Models were assessed based on the number of covariates considered, the inclusion of non-linear responses and interactions between covaraites, the manner of model selection, and whether or not they objectively evaluated model performance."

objTable <- read_csv("Figures/ObjectiveTable.csv")

objTable |>
  mutate(Objective = fct(Objective,
                         levels = c("Hypothesis", "Drivers", "Parameters",
                                    "Methods", "Predict", "Any"))) |>
  arrange(Objective) |>
  mutate(Objective = case_when(Objective == "Hypothesis" ~ "Testing hypotheses",
                               Objective == "Drivers" ~ "Identifying drivers",
                               Objective == "Predict" ~ "Making predictions",
                               Objective == "Methods" ~ "Developing methods",
                               Objective == "Parameters" ~ "Estimating parameters",
                               Objective == "Any" ~ "Any objective")) |>
  gt(rowname_col = "Objective") |>
  tab_spanner(label = "Articles",
              columns = c("Articles", "ArticlesProp"),
              id = "articles") |>
  tab_spanner(label = "Percent which have:",
              columns = c("ProbPoly", "ProbInt",
                          "ProbEval", "ProbSelect"),
              id = "percents") |>
  cols_label(Articles = "n", ArticlesProp = "%",
             MedianCovs = "Median covariates considered",
             ProbPoly = "Non-linear responses",
             ProbInt = "Covariate interactions",
             ProbSelect = "Model selection",
             ProbEval = "Model evaluation / goodness of fit") |>
  fmt_percent(columns = c(ArticlesProp, 
                          ProbPoly, ProbInt, 
                          ProbEval, ProbSelect),
              decimals = 0) |>
  tab_style(style = list(cell_fill(color = "gray80"),
                         cell_text(style = "italic")),
            locations = cells_body(rows = Objective == "Any objective")) |>
  tab_style(style = cell_fill(color = "lavender"),
            locations = cells_body(columns = MedianCovs,
                                   rows = MedianCovs == max(MedianCovs, na.rm = TRUE))) |>
  tab_style(style = cell_fill(color = "lavender"),
            locations = cells_body(columns = ProbPoly,
                                   rows = ProbPoly == max(ProbPoly, na.rm = TRUE))) |>
  tab_style(style = cell_fill(color = "lavender"),
            locations = cells_body(columns = ProbInt,
                                   rows = ProbInt == max(ProbInt, na.rm = TRUE))) |>
  tab_style(style = cell_fill(color = "lavender"),
            locations = cells_body(columns = ProbEval,
                                   rows = ProbEval == max(ProbEval, na.rm = TRUE))) |>
  tab_style(style = cell_fill(color = "lavender"),
            locations = cells_body(columns = ProbSelect,
                                   rows = ProbSelect == max(ProbSelect, na.rm = TRUE))) |>
  as_raw_html()
```

</div>

Bayesian models were far less likely to be implemented using a model selection protocol compared to frequentist models: 70% of the former fit only a single model defined *a priori,* compared to 10% of the latter (@fig-selection). In frequentist models, sequential selection protocols in which parameters are fixed one by one were the most popular choice. Nearly all frequentist models used AIC or AICc to identify their best model.

![Frequency of use for model selection methods. 'Single model' indicates no model selection occurred, 'candidate suite' includes studies which compared a predefined set of models, articles with 'procedural selection' fit parameters one-by-one, 'exhaustive selection' fit all possible combinations of models, and 'simple precursor' describes where results of a less-complex model were used to inform covariate choice.](Figures/SelectionFigure.jpeg){#fig-selection}

Model evaluation and absolute assessment of model fit was rare throughout the sample regardless of selection protocol. Only 8% of articles calculated non-relative metrics, either via in-sample validation, bootstrapped goodness-of-fit tests, or out-of-sample validation.

# Conclusions and key priorities

# OLD —

# Discussion

One of the strengths of the dynamic occupancy model is its considerable flexibility — as seen in our review, applications of DOMs over the past two decades have been fit to data for a diverse cohort of species, using a broad range of detection methods and survey designs. This flexibility permits quite a bit of creativity on the part of authors and has allowed for innovative use-cases well beyond simple estimations of occupancy. Despite this promise, questions do remain on the best practices for fitting DOMs. Studies in our sample are discordant on their approaches the modelling process, and there the literature gives little insight on how these approaches may differ in their ability to identify reliable models.

## Uses and broader potential

## Outstanding questions on approaches to modelling

## Key considerations for users

# Conclusion

### Applications of dynamic occupancy models

DOMs display great variation of scale in all senses. Analyses in our samples range from the hyper-local at individual parks [@zúñiga-vega2019] to continental-scale studies [@zuckerberg2011]. Similar diversity in study duration and site quantity indicates further emphasises the breadth of studies which DOMs may be fit to. Method of species detection also need not be a limitation for the implementation of DOMs, as articles in our sample show that many approaches may be coerced into data suitable for this model class. While studies pre-designed to match hierarchical data inputs may be simplest to align with model assumptions, post-hoc manipulation of other data such as long-term citizen monitoring programs [@zuckerberg2011; @peach2019], disparate agency modelling programs [@mcgowan2020], or camera trapping data [@davis2018] is feasible with careful consideration. Detections from multiple sources, with associated variation in efficiency, have also been incorporated by authors [@warrier2020; @pitman2017].

Reviewed implementations of dynamic occupancy models show widespread representation from mammals, birds, and herptiles. This assemblage contrasts with the related stochastic patch occupancy models (SPOMs, @gutiérrez-arellano), where the majority of models are fit to invertebrates, plants, and fungi. This difference may reflect SPOMs' assumption of perfect detection and discrete habitat patches, limitations which are unlikely to be fulfilled by most vertebrate taxa. Aside from differences in assumptions around detection, SPOMs also differ from DOMs in that they are generally spatially explicit, with probabilities of colonisation and extinction dependent on patch size and the occupancy state of nearby patches. Where desired, these features may also added to dynamic occupancy models — @sutherland2014 presents an implementation which may be considered a fusion of the DOM and SPOM methods, and several other authors incorporate patch characteristics aspects into their models as covariates [@duggan2011; @broms2016].

While objectives vary in our sample, DOMs are rarely used to extrapolate beyond study areas — use of prediction remains the least popular use-case in our sample. Articles which do make predictions provide some of the most directly applicable outputs found in the review sample: @mcgowan2020 provides projections for a threatened species under multiple putative management scenarios, and @pollentier2021 generates maps of distributions more frequently associated with species distribution models (SDMs). DOMs offer advantages over other SDMs for this purpose, particularly where populations are assumed not to be at equilibrium as is the case for many threatened and invasive species. By explicitly accounting for changes in occupancy (i.e., colonisation and extinction) DOMs may better capture future range contraction or expansion — an aspect of prediction where other SDMs may struggle.

### Practices in the modelling process

A diverse set of covariates were identified in the model sample reflecting the diversity of systems and ecological questions described by DOMs. Notably, many models lack covariates on the initial occupancy parameter, raising questions on how the parameterisation of the model in it's first time-step may affect the quality of estimates at later time-steps. The distinction in covariates considered for the ecological parameters vs those considered for detection is also of note: ecological parameters more frequently contained covariates describing environmental features, where detection more frequently contained more structural covariates such as season. Covariate complexity is relatively low across the board, with only a fraction of models implementing either non-linear responses to covariates or interactions between covariates. While this addition complexity does have implications for model selection, it may be necessary to accurately describe many ecological relationships.

Model selection approaches diverge between Bayesian and frequentist implementations of the model. While model selection will naturally be practised differently between these two categories, it is unclear what effect a lack of model selection may have on the outputs produced by Bayesian models, and it is likely that Bayesian model selection approaches should be further explored @hooten2015. Model selection approaches in the Frequentist applications show limited consensus, although most do consider multiple models. Notably, @morin2020 found that the popular procedural approaches to model selection (where one sub-model is assessed at a time) may not provide optimal results. Considerably more research is required on this topic to clarify how authors should best approach covariate selection.

### Key considerations for model users

Implementation of dynamic occupancy models requires considerable thought to ensure that model outputs correspond with author expectations and ecological realities. Based upon our review of the existing literature we discuss key questions authors should carefully engage with during the modelling process.

1.  **Are seasons ecologically meaningful, and how do they affect the definition of occupancy?** As previously discussed, the duration of the season has important implications for both the reliability and proper interpretation of model results. Seasons which are not ecologically meaningful may run afoul of the closure assumption resulting in associated biases. Also relevant is how the definition of 'occupancy' may shift with season length and life-history of a species — for many species, occupancy within a year and within a week may be considerably different concepts, with shorter intervals more closely relating to the idea of 'use.' With the proliferation of continuous detection methods such as camera traps and acoustic monitors where the definition of 'season' is to some degree arbitrary, it is increasingly important to thoughtfully consider how this element of the DOM is defined.

2.  **Are key drivers of occupancy and detection incorporated?** While it is effectively impossible to *fully* describe heterogeneity across a landscape, authors should make their best efforts to capture the drivers of occupancy and detection most likely to influence their study species. This is true even when another covariate is of principal interest to investigators — failure to include other major determinants of occupancy is likely to bias parameter estimates. One place where this is of particular importance is in determining drivers of detection, which is not only subject to observational conditions but also ecological determinants which may cause the species to spend more time at a location and thus be more frequently available for detection.

3.  **Ensure covariates reflect ecological complexity.** Many ecological relationships are non-linear and may require polynomial terms to be accurately represented. This is particularly true at large spatial scales where a larger portion of a species ecological niche is represented in surveyed data. Authors should question whether a linear term is sufficient to capture hypothetical relationships in their study systems, and consider quadratic terms as well as interactions between covariates where appropriate.

4.  **Consider model evaluation.** Model evaluation remains rare in ecological research relative to other fields embracing complex modelling, likely due to a reluctance to 'throw away' data for use in testing. This concept should be challenged, particularly for larger scale long-term projects where reserving a fraction of data for testing is unlikely to substantially reduce precision of parameter estimates. For studies where this is genuinely not possible, reporting results of goodness-of-fit tests may still provide valuable information to readers.

### Priorities for future research

Work on dynamic occupancy modelling over the past 20 years has developed this model class into an important tool for describing species distributions and ecological relationships. Historically, methods-development research for dynamic occupancy models has focused on questions of model structure — namely ideas around the closure assumption and imperfect detection. This emphasis likely reflects the DOM's roots in the mark-recapture literature, where these topics have been discussed for several decades. Considerably less emphasis has been placed on the relation of model parameters to environmental factors via covariates and the model selection processes used to do so, despite this being an objective in a considerable fraction of studies fitting DOMs. This stands in stark contrast to related correlative species distribution models, where these topics have been the subject of considerable research over the years. There are opportunities for the DOM literature to learn from species distribution models, and more targeted research on the influence of the model selection process is necessary.

An important opportunity for future research relates to autonomous detection methods, such as camera traps and bio-acoustic monitors — these technologies have permitted collection of enormous quantities of data particularly well-suited to dynamic occupancy models, but have also raised additional questions around their implementation. The definition of seasons is of increased importance for these data types, where division of data into 'seasons' and 'observations' is arbitrary. Further research should explore the implications of this phenomenon and the additional modelling opportunities made possible by this novel data type.

The past two decades have shown that dynamic occupancy models are an excellent tool for addressing a wide range of ecological questions relating to species occupancy throughout space and time. The flexibility afforded by model extensions and the considerable advantages of the class over other species distribution models is likely to ensure that these models continue to increase in popularity. Given their strong use on threatened taxa and excellent potential to support environmental management, it is important that research continues to support model users in creating the most reliable models possible.

## old

First-authors of studies came from a total of 15 countries, with authors in the United States producing 64% of articles. 86% of articles had at least one author based at an academic institution, 46% had author(s) at government institutions, and 25% had author(s) from non-profit organisations.

## Study locations, focal species, and data collection

Data for DOMs were collected from study areas on all continents and in 21 countries, with studies in the United States again representing a majority of the total articles published. These study areas ranged in size from less than 1 to over 1 million square kilometres (@fig-StudyDetails a).

Focal species were largely vertebrate taxa, although 6 articles in the sample fit models to invertebrates. A single non-animal taxa was represented in the sample, from an article fitting a DOM to a rare species of lichen [@belinchón2017]. 35% of articles modelled threatened taxa, and 10% modelled species invasive to the study area. Most articles (50, 60%) fit single-species DOMs to one taxa. Of the remainder, 26 (31%) fit independent single-species models to more than one taxa. 8 articles fit (10%) multi-species implementations with explicit interactions between species (@fig-StudyDetails b).

Data used to fit DOMs included human survey data such as transects or point counts (69% of articles), live-trapping data (20%), camera traps (12%) or bioacoustic monitors (1%), and citizen science data (12%); some articles incorporate multiple detection methods to build their detection datasets. Study duration, defined as the time elapsed between the first and last surveys conducted, ranged from 1 month to 40 years with the median study running for 3.25 years. The number of seasons ranged from 2 to 204 with a median of 5; the number of observations per season ranged from 2 to 153 with a median of 5. The number of sites surveyed per article ranged from 7 to 9394, with the median study including 98 sites (@fig-StudyDetails c).

### 2024-07-28 Randomisation/Final_Randomisation.qmd
Cleaned up for final stage

---
title: Randomise papers for DOM Review
subtitle: Describes random selection of papers for inclusion in the review
author: Saoirse Kelleher
date: today
format: 
  html:
    code-fold: true
    theme: minty
    toc: true
    embed-resources: true
---

```{r packages}
#| message: false

options(tidyverse.quiet = TRUE)
library("knitr")
library("tidyverse")
library("readxl")
```

## Load Web of Science queries

Two search queries were made on Web of Science on July 26 2024.

1.  ALL=(dynamic occupancy model\*) OR ALL=(occupancy dynamic\$ model\*) OR ALL=(multi\$season occupancy model\*) OR (ALL=(occupancy) AND ALL=(Coloni\$ation) AND ((ALL = (Extinction)) OR (ALL = (Persistence))) AND All=(Detection)) and Ecology or Biodiversity Conservation or Environmental Sciences or Zoology or Evolutionary Biology or Biology or Ornithology or Forestry or Plant Sciences or Environmental Studies or Fisheries or Entomology (Web of Science Categories)
2.  All articles citing MacKenzie et al. 2003

WOS datasheets containing all matching records are loaded in.

```{r load WOS queries}
keyTerms_datasheet <- read_xlsx("Queries/Final Queries/KeyTerms_A.xlsx",
                                skip = 10) |>
  rbind(read_xlsx("Queries/Final Queries/KeyTerms_B.xlsx",
                  skip = 10)) |>
  select(Title, Year = "Publication Year", Journal = "Source Title", 
         Authors, Citations = "Total Citations", DOI) |>
  mutate(Query = "KeyTerms")

citesMackenzie_datasheet <- read_xlsx("Queries/Final Queries/MacKenzie_A.xlsx",
                                      skip = 10) |>
  rbind(read_xlsx("Queries/Final Queries/MacKenzie_B.xlsx",
                  skip = 10)) |>
  select(Title, Year = "Publication Year", Journal = "Source Title", 
         Authors, Citations = "Total Citations", DOI) |>
  mutate(Query = "CitesMackenzie")
```

The datasheets are then combined, and duplicate records are removed. All records prior to 2004 are also removed. Each article is given a unique ID.

```{r combine articles}
combined_datasheet <- rbind(keyTerms_datasheet,
                            citesMackenzie_datasheet) |>
  filter(Year %in% 2004:2023) |>
  mutate(QueryValue = TRUE) |>
  pivot_wider(names_from = Query, values_from = QueryValue,
              values_fill = FALSE) |>
  mutate(Review_ID = 1:n())
```

## Add year strata and randomise

```{r randomise articles}
withr::with_seed(14512, {
  randomised_datasheet <- combined_datasheet |>
    mutate(Strata = case_when(Year %in% 2004:2007 ~ "2004-2007",
                              Year %in% 2008:2011 ~ "2008-2011",
                              Year %in% 2012:2015 ~ "2012-2015",
                              Year %in% 2016:2019 ~ "2016-2019",
                              Year %in% 2020:2023 ~ "2020-2023")) |>
    mutate(StrataRank = sample(1:n(), n()),
           .by = Strata) |>
    arrange(Strata, StrataRank) |>
    select(Review_ID, StrataRank, Strata, Title, Year, 
           Journal, Authors, Citations, DOI, 
           KeyTerms, CitesMackenzie)
})
```

## Write randomised datasheets

```{r Save sheet}
write_csv(randomised_datasheet, "Randomisation/Final_Randomisation.csv")
```

### 2024-07-28 Manuscript/_manuscript/domReview.qmd

@@
 We also note that practices in fitting dynamic occupancy models are highly variable, and that the existing literature lacks clear answers for best practices.
-This contrasts strongly with the robust species distribution model literature on modelling practices, which we suggest may be a valuable resource for improving how authors implement dynamic occupancy models.
+This contrasts with the strong focus on modelling practices in the species distribution model literature, which we suggest may be a valuable resource for improving how authors implement dynamic occupancy models.
 
@@
 
-Ecologists have long sought to describe the processes by which species are distributed across landscapes and throughout time [@humboldt1849].
-An understanding of where a species exists, which factors lead to its presence, and how its distribution may change has always been of interest and is ever more important in a dynamic and changing world.
+Ecologists have long sought to describe the processes by which species are distributed across landscapes and throughout time [@humboldt1849], and an understanding of where a species exists, which factors lead to its presence, and how its distribution may change is ever more important in a dynamic and changing world.
 Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and are imperative for assessing conservation status, and further untangling the drivers of occurrence can help us to better understand a species life history and inform potential management actions.
-A robust understanding of the occupancy patterns of a species also helps us to predict where a species is most likely to occur — now, and under hypothetical future scenarios.
+Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur — now, and under hypothetical future scenarios.
 
-The need to reliably describe species occupancy has led to the development of numerous quantitative approaches, of which the best known are species distribution models [SDMs, @elith2009].
+The need to reliably describe species occupancy has led to the development of many quantitative approaches, of which the best known are species distribution models [SDMs, @elith2009].
 The prototypical SDM correlates records of species occurrence with environmental features to determine where species are most probable to occur.
-While SDMs are an exceedingly popular tool for describing occupancy, they do have considerable downsides.
+While SDMs are an exceedingly popular tool for describing occupancy, the most frequently used implementations are known to struggle under certain common scenarios.
 First and foremost, SDMs do not typically account for imperfect detection, a near-ubiquitous characteristic of wildlife field data that can lead to meaningful bias in estimates of occupancy [@lahoz-monfort2014; @guillera-arroita2017; @gu2004].
 Secondly, SDMs have been noted to perform poorly when a species' distribution is not at equilibrium [@elith2010].
-This may be problematic when working with invasive species or species whose ranges are expanding or contracting, all of which are increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].
+These shortcomings may be problematic when working with invasive species or species whose ranges are expanding or contracting, all increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].
 
-The dynamic occupancy model (DOMs, also termed 'multi-season occupancy models') introduced in @mackenzie2003 provide an alternative framework to modelling occurrence and distributions.
-By explicitly describing the colonisation and extinction processes which guide occupancy while accounting for imperfect detection, the model class offers more advanced insights than simpler static species distribution models; despite this, input data requirements remain achievable compared to other more complex process-explicit models.
-Dynamic occupancy models unite two processes: the ecological process of site occupancy describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed given that a site is occupied (@fig-modelform).
+The dynamic occupancy model (DOM, also termed 'multi-season occupancy model') introduced in @mackenzie2003 provides an alternative framework to modelling occurrence and distributions.
+The DOM, an extension of @mackenzie2002's static occupancy model, explicitly describes the colonisation and extinction processes which guide occupancy while accounting for imperfect detection.
+By incorporating these process-explicit elements, the DOM offers more advanced insights than simpler static species distribution models and accounts for their major weaknesses.
+Despite their added complexity, input data requirements remain more achievable compared to other more complex process-explicit models, enhancing their suitability to applied questions in ecology.
+
+### Dynamic occupancy model form and assumptions
+
+Dynamic occupancy models unite two processes: the ecological process of site occupancy describes the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed given that a site is occupied (@fig-modelform).
 Under this parameterisation, sites exist in either occupied or unoccupied states.
-The occupancy state for each step in the first time-step is given by the *initial occupancy* parameter, and in ensuing time-steps occupancy is determined based on probabilities of *colonisation* and *extinction.* Given that a site is occupied, the probability of observing the species during a survey is given by the *detection* parameter.
-In the DOM's standard form, it is assumed that false-positive detections do not occur.
+The occupancy state for each site in the first time-step is given by an initial occupancy parameter which itself resembles a static SDM.
+In ensuing time-steps, occupancy state is predicated on the site's previous state and probabilities of colonisation and extinction.
+When surveys are conducted at an occupied site, the probability that the species is observed is given by the detection probability.
+In the DOM's standard form, it is assumed that detections never occur at unoccupied sites; that is, that false-positive detections do not occur.
 
@@
 
-To distinguish between the ecological and observation processes and estimate the probability of detection, the DOM employs on a hierarchical sampling design depicted in @fig-surveys.
+To disentangle the ecological and observational processes, the DOM requires a hierarchical sampling design as depicted in @fig-surveys.
 Under this design, observations at a site occur during distinct, time-bound seasons in which sites are considered closed to changes in occupancy.
 Within each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy.
-Most frequently these observations are collected by revisiting the site on separate occasions, although alternative methods to generate repeat observations exist such as conducting surveys at multiple locations within a site, using of multiple observers during a survey, or recording the time elapsed until a detection is recorded [@mackenzie2005].
+Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@mackenzie2005].
 
@@
 
-We have presented the standard form of the dynamic occupancy model as described in @mackenzie2003 — numerous model extensions and alternative formulations are also available, including implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models[@dorazio2010].
+These descriptions represent only the original form of the dynamic occupancy model as prescribed in @mackenzie2003.
+Numerous model extensions and alternative formulations are also available, including implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models[@dorazio2010].
 For a comprehensive discussion of the most common extensions and their applications see @bailey2014, as well as @devarajan2020 for a review of multi-species occupancy models.
@@
 A second assumption states that sites are closed to changes in occupancy between seasons.
-This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion on the bias which is introduced when it is violated [@rota2009; @otto2013].
-Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discuss the difficultly in distinguishing temporary emigration and local extinction.
+This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion around the bias which is introduced when it is violated [@rota2009; @otto2013].
+Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discuss the difficultly in distinguishing between temporary emigration and local extinction.
 Model extensions to relax the closure assumption have been developed, including @kendall2013's approach using staggered arrival and departure periods between sites.
-More pertinent for most studies, however, is a careful consideration of ecologically and observationally relevant seasons which will allow the estimation of an appropriate 'type' of occupancy, whether that be shorter or longer term.
+A more pertinent approach for most studies, however, is careful consideration of ecologically relevant seasons corresponding to an appropriate definition of occupancy.
 
-Generally speaking, DOMs may be applied in much the same ways as SDMs.
-Comparable probabilities of occupancy may be estimated, relationships to environmental covariates can be quantified, and predictions of occupancy across space and into the future can be generated.
-Despite these similarities, the two models have developed largely distinctly with key differences in directions within their respective research communities.
-DOMs originated from the mark-recapture literature, with similarities to models such as the Cormack-Jolly-Seber model [@pollock1990].
-This history contributes to an emphasis on imperfect detection and survey design in the DOM literature.
-In contrast, other areas of the modelling process have received comparatively little attention.
-For any model estimating occupancy, key environmental drivers must be adequately represented using covariates — the SDM community has long debated how to best approach model selection and evaluation [@guisan2006; @araújo2006; @jarnevich2015; @araújo2019].
-The DOM literature has by comparison focused less on these aspects of modelling [@mackenzie2017], although what research does exists indicates that choice of model selection approach can impact model performance [@morin2020].
+Each of the DOM's main parameters (Initial occupancy, Colonisation, Extinction, and Detection) can be estimated with covariates to capture spatial or temporal variation.
+As with any model, it is assumed that heterogeneity in the study system is reasonably captured by these covariates.
+While the importance of relating the ecological parameters to the environment is generally understood, representation of heterogeneity in the Detection probability is particularly nuanced and may be easily misinterpreted.
+The likelihood of detecting a species depends not only on the factors influencing observability (e.g.
+survey efficiency, weather, etc.), but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017]; failing to account for these factors when modelling detections may risk biasing model outputs.
 
-This review aims to provide an overview of the dynamic occupancy model's first two decades of use.
+### Review aims and objectives
+
+DOMs and SDMs have similarities in their application and outputs.
+Using either model, comparable probabilities of occupancy may be estimated, relationships to environmental covariates can be quantified, and predictions of occupancy across space and into the future can be generated.
+Despite these parallels, the two models have developed largely distinctly with differences in priorities within their respective research communities.
+From their beginnings, SDMs have sought to relate species occurrence to environmental factors — this is reflected in their literature, which has extensively discussed how best to incorporate covariates in models [@araújo2006; @guisan2006; @elith2006].
+On the other hand, the DOM has its origins in the mark-recapture literature with pointed similarities to methods like the Cormack-Jolly-Seber model which also relies on a similar sampling scheme [@pollock1990].
+This history has contributed to an emphasis on imperfect detection and survey design in the DOM literature, while other areas of the modelling process have received comparatively little attention [@mackenzie2017].
+
+This review aims to provide an overview of how dynamic occupancy models have been applied and implemented in their first two decades of use.
 We present a systematic review exploring how authors have used the DOM to answer ecological questions, with emphasis on how they have collected data, selected covariates, and evaluated their models.
-Based on these results we highlight the DOMs flexibility as a tool for understanding occupancy and discuss common misconceptions regarding their use.
-We also examine how authors have approached the modelling process and discuss areas where lessons can be drawn from the SDM literature.
-Finally, we present considerations for authors seeking to use DOMs in their work and outline key priorities and opportunities for future research involving this important model class.
+Based on these results we highlight the DOMs flexibility as a tool for understanding occupancy and discuss further opportunities where they may be of use.
+We also examine how authors have approached the modelling process and discuss key areas where lessons can be drawn from the SDM literature.
+Finally, we present considerations for authors seeking to use DOMs in their work and outline key priorities for future research involving this important model class.
 
@@
 
-Our review focuses on the dynamic, multi-season implementation of the occupancy model of @mackenzie2003.
-To meet inclusion criteria articles must have fit a model which i) describes multiple sites which can exist in at least two states, including an occupied and unoccupied states; ii) has multiple survey periods, between which sites may change state conditional on the prior survey period's occupancy state and at least one transition probability; iii) contains at least one detection parameter; and iv) uses real (i.e., not simulated) field ecological data.
-Papers were gather for inclusion in two stages:
+Our review is focused on applications of the dynamic, multi-season occupancy model of @mackenzie2003 and its extensions.
+To reflect this, we included articles which fit a model meeting the following criteria:
 
-In a preliminary analysis, four search strings ("dynamic occupancy model," "occupancy dynamics model," "incidence function model," and "multi-season occupancy model" were entered into Google Scholar.
+I.  Uses non-simulated, field-collected presence-absence data.
+
+II. Describes multiple sites which can exist in at least two states including *occupied* and *unoccupied.*
+
+III. Has multiple seasons, between which sites may change states conditional on the prior season's occupancy state and transition probabilities such as colonisation and extinction.
+
+IV. Contains at least one parameter describing the detection process.
+
+We gathered a representative sample of studies meeting these criteria in two stages.
+In a preliminary analysis, four search strings ("dynamic occupancy model," "occupancy dynamics model," "incidence function model," and "multi-season occupancy model") were entered into Google Scholar.
 Up to 100 articles were downloaded for each string, in order of appearance.
 These articles were divided into four strata based on data of publication — 2000-2005, 2006-2010, 2011-2015, and 2016-2021.
-For each combination of search string and year strata, a random selection of the greater of 25% of articles or 5 articles was processed.
-Where available, articles which did not meet inclusion criteria were replaced with another from the same search string and strata.
+For each combination of search string and year strata, a random 5 articles or 25% of the strata (whichever was greater) was processed.
+Where available, articles which did not meet inclusion criteria were replaced with another from the same search string and strata combination.
 
-After completion of the preliminary sample, we noted that the articles from the 'incidence function model' search term represented a distinct model class with its own history and features [@gutiérrez-arellano].
-Articles fitting this model were dropped.
+After completion of the preliminary sample, we noted that the articles from the 'incidence function model' search term represented a distinct model class with its own history and features: also called 'stochastic patch occupancy models' or 'Hanski models', these generally focus on small-scale patch dynamics and do not typically account for imperfect detection [@gutiérrez-arellano], hence these articles were removed from the sample.
 We also noted that early DOMs used varied terminology to refer to the model and that very few articles were identified from the first few years after their introduction.
 To better capture these early models and increase the size of our sample, a secondary sample was collected from the complete list of articles which have cited @mackenzie2003, accessed via Web of Science.
-Articles from this list were split into the same year strata as the first sample, plus an additional 2022-2023 category.
+Articles from this list were split into the same year strata as the first sample plus an additional 2022-2023 category.
 15 articles were drawn from each strata and processed, with replacement from within their strata where necessary.
@@
 We recorded details on the type of taxa (bird, mammal, etc.) modelled in each article, and how multiple species were modelled where applicable.
-Taxa were denoted as threatened either when they are listed on the IUCN Red List of Species or when authors indicate that they are threatened.
-This latter determination was made to account for unlisted sub-species or species of local concern.
+Taxa were denoted as threatened either when they are listed on the IUCN Red List of Species or when authors otherwise indicate that they are threatened.
+This deference to authors' representation of conservation status was made to account for sub-species which lack listings or species which are of more local concern.
 Study locality and size was documented; the size of the study area being defined as the intended area of inference containing all sites, recorded as an order of magnitude to partially account for uncertainty in reporting.
@@
 To examine the reasons why authors used DOMs, we allocated each article to one or more category of objective based on the study's stated aims.
-The possible categories were i) *Estimating parameters*, where authors express interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; ii) *Testing hypotheses*, where authors explore specific predefined relationships between covariates and model parameters; iii) *identifying drivers*, where authors attempt to find which covariates influence model parameters; iv) *generating predictions*, where authors estimate model parameters either for unsurveyed sites or under possible future conditions; and v) *methods development*, where authors introduce, test, or demonstrate aspects of dynamic occupancy models.
+The possible categories were *Estimating parameters*, where authors express interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; *Testing hypotheses*, where authors explore specific predefined relationships between covariates and model parameters; *Identifying drivers*, where authors attempt to find which covariates influence model parameters; *Generating predictions*, where authors estimate model parameters either for unsurveyed sites or under possible future conditions; and *Methods development*, where authors introduce, test, or demonstrate aspects of dynamic occupancy models.
 
@@
 We recorded all covariates considered in each study, regardless of whether they were or were not included in final models.
-Data was collected for each covariate including a general classification, whether it was directly observed or remotely sensed, whether it was static or varied between seasons, and how it was represented in the model - as a linear term, a polynomial term, or as part of an interaction with another covariate.
+Key traits of each covariate were recorded including their general category, whether they were directly observed or remotely sensed, whether they were static or varied between seasons, and how they were represented in the model (as a linear term, a polynomial term, or as part of an interaction with another covariate).
 Model selection procedures were sorted into non-exclusive categories including *A priori*, where only one model was considered; *candidate suite*, where a predefined set of models was considered; *procedural*, where covariates were selected parameter-by-parameter; *exhaustive*, where all possible model combinations were fit; *simple precursors*, where selection was preceded by another simpler model implementation; and *other*.
 Model evaluation was similarly categorised - options in this case included *goodness of fit*, where absolute metrics of fit were used; and *out-of-sample validation* and *in-sample-validation*, where predictive performance was evaluated using in-sample or out-of-sample data respectively.
-
 For the full spreadsheet of data collected and further details on categorisation, see **Appendix I.**
@@
 
-One of the strengths of the dynamic occupancy model is its considerable flexibility across several axes — they may used to achieve several different objectives, fit to diverse forms of occurrence data, and applied to studies at a wide range of spatial and temporal scales.
-This flexibility is reflected in the types of authors who conducted the studies captured in our sample: while 86% of articles had at least one author based at an academic institution, 46% had author(s) at government institutions and 25% had author(s) from non-profit organisations.
-The high proportion of articles with government and NGO participants indicates a more applied focus for many users of DOMs, suggesting that model outputs may directly guide management or conservation actions in many cases.
-The most frequent objective for studies using DOMs was hypothesis testing, with 50% of authors expressing interest in a specific relationship.
-31% were interested in identifying nondescript drivers of occupancy, 30% in estimating models parameters like occupancy themselves, and 29% worked to develop and expand on methods for DOMs.
-Just 10% of authors expressed interest in making predictions to unsurveyed locations or into the future.
-
-Study sites were located in all continents except Antarctica and in 21 countries, with the majority occurring in the United States of America (@fig-StudyDetails A).
-This strong geographic trend stands in stark contrast with SDMS — @araújo2019's review found that SDMs are far more international, with no continent yielding more than 20% of studies.
-The reasons behind this disparity are unclear, but one possible explanation may be the DOMs roots in the mark-recapture literature and its history of publication in American journals.
-These factors may also contribute to the seemingly limited overlap between the bodies of literature for SDMs and DOMs.
-
-Differences in use between SDMs and DOMs are also evident in the species targeted for modelling.
-Where SDMs show broader coverage of plants, invertebrates, and aquatic species [@araújo2019], 93% of studies in our sample of DOMs modelled terrestrial vertebrate fauna with @belinchón2017's study on lichen the sole non-animal taxa representative (@fig-StudyDetails A).
-These are also taxa where the DOM's advantages, namely their handling of imperfect detection, are most applicable.
-One group for which DOMs have only rarely been used is aquatic species, as in @fisher2014's application to invasive salmon or @falke2012's model on Great Plains stream fishes.
-
-40% of the studies in our sample fit models to more than one taxa ('taxa' defined here as one or more species modelled as one).
-Of these, 75% fit DOMs to multiple species, and 25% fit explicitly multi-species implementations (@fig-StudyDetails B).
-These multi-species models varied in structure, some taking hierarchical approaches with species-specific effects [@mordecai2009; @lee2017] and others explicitly modelling interactions between species [@lesmeister2015; @fidino2019].
-For a more comprehensive discussion on multi-species occupancy modelling, see @devarajan2020's review.
-
-\< FLAG \> discuss how it can be used for large scale monitoring type things where you need to model tons of species — this data already exists for a lot of places.
-flag paper that fit 300 species, note that there are questions about how to appropriately model them.
-
-One notable finding of our review is the diversity of scale in studies where DOMs have been used.
-The smallest study area in our sample was less than one kilometre squared, and the largest over one million.
-Study duration shows similar variation — data collection ran for for less than a month in the shortest study, and for 40 years in the longest, with a median duration of 3.25 years.
-This reiterates the flexibility of dynamic occupancy models, with explorations of hyper-local patterns of occupancy and analyses of continental-scale distribution using the same underlying model structure.
-The detection data collected from these surveys came from a variety of detection methods: 69% of articles used human observations, 20% conducted live trapping, 12% used camera traps, and 1% used bioacoustic monitors (@fig-StudyDetails); some also incorporated data from multiple detection sources.
+Dynamic occupancy models are considerably flexible in many ways: they may used to achieve several different objectives, fit to occurrence data collected by various methods, and applied to studies at a wide range of spatial and temporal scales.
+Some of this diversity in study locations, focal taxa, and survey methods amongst the reviewed studies is presented in @fig-StudyDetails.
 
@@
 
+### Focal systems
+
+Although studies were conducted in 21 countries and on all continents except for Antarctica, the majority of reviewed articles used data collected in the United States of America.
+This strong geographic trend starkly contrasts with SDMs far more international reach; @araújo2019's review of SDM applications found that no more than 20% of studies came from any single continent.
+The reasons behind this disparity are unclear, but a possible explanation could relate to the DOM's roots in the mark-recapture literature and a history of publication in American journals.
+These differences in where users conduct their research may also contribute to the seemingly limited overlap between these models' respective bodies of literature.
+
+Differences in use between SDMs and DOMs are also evident in the species targeted for modelling.
+Where SDMs show broader coverage of plants, invertebrates, and aquatic species [@araújo2019], 93% of studies in our sample of DOMs modelled terrestrial vertebrate fauna with @belinchón2017's study on lichen the sole non-animal taxa representative.
+These differences are most easily explained by the DOM's emphasis on imperfect detection, a concern which is less prioritised (though not non-existent [@chen2013]) in studies on non-mobile species like plants.
+We find DOMs have been frequently used to model threatened taxa (32% of studies) and invasive species (11% of studies), taxa for which they may be expected to excel due to their more explicit handling of changes in occupancy.
+One group for which DOMs have only rarely been used is aquatic species, as in @fisher2014's application to invasive salmon or @falke2012's model on Great Plains stream fishes.
+
+While most articles focus on a single taxa, 40% of the studies in our sample fit models to more than one taxa ('taxa' defined here as one or more species modelled as one).
+Of these, 75% fit independent DOMs to multiple species, and 25% fit explicitly multi-species implementations.
+These multi-species models varied in structure: some use species-specific effects to describe metapopulations by including hundreds of species in a single model [@dorazio2010; @hendershot2020], while others fit models which explicly account for species interactions and estimate conditional occupancy, colonisation, extinction, and detection [@lesmeister2015; @fidino2019].
+While they did not fit multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012].
+Working with large numbers of taxa does raise additional questions, as the level of complexity which can realistically be applied to each individual taxa is likely to be reduced for practicality's sake.
+With multi-species community frameworks, environmental covariates must be relatively general; and when fitting large numbers of independent models covariate selection processes are likely to be more restricted due to computational limitations.
+
+### Data collection and inputs
+
+DOMs show exceptional diversity in the scales at which they can be implemented.
+The smallest study area in our sample was less than one kilometre squared, and the largest over one million.
+Study duration shows similar variation — data collection ran for for less than a month in the shortest study, and for 40 years in the longest, with a median duration of 3.25 years.
+This reiterates the flexibility of dynamic occupancy models, with explorations of hyper-local patterns of occupancy and analyses of continental-scale distribution using the same underlying model structure.
+The detection data collected from these surveys came from a variety of detection methods: 69% of articles used human observations, 20% conducted live trapping, 12% used camera traps, and 1% used bioacoustic monitors (@fig-StudyDetails).
+
 It is important to note that the data input requirements for dynamic occupancy models are not as arduous as may be commonly perceived.
@@
 In one example, @riddell2021 used data from as long ago as 1908 by treating grid cells as sites and counting historical surveys within each cell as repeat observations.
-In a similar vein, data does not necessarily need to come from the same source.
-Detections from multiple methods, with associated variation in detection probabilities, have also been incorporated by authors [@warrier2020; @pitman2017].
+In a similar vein, all detection data does not necessarily need to come from the same source — detections from multiple methods, with associated variation in detection probabilities, may be incorporated within a single model [@warrier2020; @pitman2017].
 
-The proliferation of modern survey techniques like camera traps and bioacoustic monitors has also led to an increase in datasets suitable for the dynamic occupancy model.
-The DOM is a valuable tool for analysing these data [@balantic2019]
+Such broadness and flexibility in data inputs is not amenable to a one-size-fits-all definition of occupancy.
+Authors must carefully consider precisely what they are modelling and address questions on the 'scale' represented by their model [@chave2013].
+Drivers of occupancy may differ depending on whether a site is represented by a single point on the landscape or as a grid cell, with the former likely to depend on more local, small scale factors rather than landscape-level trends [@stevens2019].
+This is also true for the *temporal* scale of occupancy: whether a site is occupied within a week or within a year is a non-arbitrary distinction representing vastly different conceptions of occupancy.
+This discussion is particularly pertinent in cases where the selection of season length *is* to some extent arbitrary, as with camera-trap or bioacoustic data where continuous recordings can be broken down into distinct 'seasons' of any length.
+DOMs are well suited to these data types [@balantic2019], and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data.
+For example, @kleiven2020 and @mölle2022 divide their camera trap records into seasons of just a few days.
+While this provides exciting insights of occupancy at extremely fine temporal scale, further research is needed on how to determine appropriate season and survey durations with respect to research questions.
 
-Modern survey techniques like camera traps and bioacoustic monitors create other opportunities for modelling, and questions remain on the best ways to approach handling such large quantities of data.
-One interesting possibility is the use of large numbers of extremely short seasons, which is rarely feasible when using human survey data — @kleiven2020 and @mölle2022 divided their camera trap records into seasons of just a few days, modelling patterns of occupancy at a fine temporal scale.
-It is important to note that this substantially alters how model outputs should be interpreted, and a larger discussion on this topic is necessary.
+### Objectives and applications
 
-The DOMs ability to make predictions has been often underexploited in the existing literature, given its broad potential for the use-case [@kéry2013; @briscoe2021].
-Articles which do make predictions provide some of the most directly applicable outputs found in the review sample: @mcgowan2020 provides projections for a threatened species under multiple putative management scenarios, and @pollentier2021 generates maps of distributions more frequently associated with species distribution models (SDMs).
+Authors have used the DOM in many ways to achieve their research aims.
+The most frequent objective for studies using DOMs was hypothesis testing, with 50% of authors expressing interest in a specific relationship.
+31% were interested in identifying nondescript drivers of occupancy, 30% in estimating models parameters like occupancy themselves, and 29% worked to develop and expand on methods for DOMs.
+Just 10% of authors expressed interest in making predictions to unsurveyed locations or into the future.
+The DOMs ability to make predictions has been often underexploited in the existing literature, given its noted potential for forecasting occupancy in non-equilibrium scenarios [@kéry2013; @briscoe2021].
+Articles which *do* make predictions provide some of the most directly applicable outputs found in the review sample.
+As examples, @mcgowan2020 provides projections for a threatened species under multiple putative management scenarios, and @pollentier2021 generates maps of distributions resembling those made with species distribution models (SDMs).
 
-# Current modelling practices
+The authors of the studies captured in our sample come from a variety of backgrounds.
+86% of articles had at least one author based at an academic institution, 46% had author(s) at government institutions and 25% had author(s) from non-profit organisations.
+The high proportion of articles with government and NGO participants suggests an applied focus for many users of DOMs.
+This has important implications: when a model is used to assess a critically endangered species [@carvalho2023], guide public health management of a disease vector [@mores2020], or track rapidly-developing biological invasions [@wood2020], it is imperative that models can be trusted to accurately represent a system.
+Regardless of the reasons why they turn to the DOM to analyse their data, authors must navigate the model building process.
 
-Reviewed articles have approached the modelling pro
+# Approaches to implementation
 
-A summary of covariates considered in the review sample is presented in @tbl-covariates.
-For each category of covariate relative frequency and characteristics are given.
-For all categories, the rate of consideration of polynomial terms does not exceed 16%, and the rate of consideration of interactions does not exceed 34%.
+Approaches to building any type of model will necessarily depend on the possibilities of the data at hand and on the priorities of the model-builder.
+This precludes any prescription of the 'best' way to build a model; however, there may still be commonalities in the modelling workflow.
+The modelling process for both DOMs and SDMs can be divided into a few components.
+First, based on prior knowledge and ecological theory a suite of covariates with hypothetical relationships with model parameters is assembled.
+Next, some form of model selection procedure is followed to identify either a single optimal model, or an ensemble of suitable models to be used to make conclusions.
+Following this, a model's performance should be somehow assessed so that the validity of conclusions can be adequately judged.
+In this section we examine how DOM users have approaches each of these stages, and make comparisons to common practice in the more extensive SDM literature.
+
+### Model complexity
+
+'Complexity' is a broad term which encompasses many aspects of a model [@merow2014].
+Opinions on simple versus complex models can be divided - some advocate for the simplest possible models, arguing that they are most generaliseable; others insist that overly-simple models cannot adequately represent the most important drivers in a system [@evans2013; @lonergan2014].
+To account for imperfect detection and describe change in occupancy, DOMs must be somewhat more complex relative to static correlative occupancy models.
+This added complexity helps mitigate the bias introduced when important elements like imperfect detection are ignored.
+Within DOMs, however, further complexity is to some degree up to the modeller: they can choose how many covariates to consider for inclusion on the various parameters, and how to represent the nature of the relationship between those covariates and parameters.
+
+The variation in the quality of covariates considered for each model parameter amongst our reviewed articles is shown in @fig-covariates.
+Notably, many studies expressed some parameters as a constant without including any covariates: of the four standard model parameters initial occupancy has the most articles considering zero covariates (35%) and detection probability the least (19%).
+A lack of any covariates on the initial occupancy parameter raises some questions.
+Presumably any taxa will have some amount of non-random variation in how it is distributed across a landscape; by expressing initial occupancy as a constant this heterogeneity is not captured.
+Omission of factors which drive occupancy can introduce bias into SDMs [@barry2006].
+Initial occupancy is itself effectively a static correlative SDM, and the effects of poorly estimating this parameter on predicted occupancy in ensuing seasons is uncertain but unlikely to be favourable.
+
+![Distribution of covariate quantity considered for each model parameter.](Figures/CovParamPlot.jpeg){#fig-covariates}
+
+The types of covariates which were considered for modelling were also wide-ranging, and a summary of their inclusions in articles is presented in @tbl-covariates.
+The most frequently used covariates described aspects of habitat, with 57% of articles considering these terms for at least one parameter.
+Data for covariates may be either collected directly, requiring measurement at individual sites, or indirectly via remote sensing — where directly measured covariates may offer more direct insights, indirect measurements are far more easily projected to unsurveyed sites when making predictions [@austin2002].
+Some covariates, like biotic interactions, must nearly always be measured at sites, while others such as climate and weather covariates are generally indirectly observed.
+There is variation in which covariates are most likely to be dynamic and change through time.
 
@@
 
-The number of covariates considered varied considerably, both between studies and between parameters as shown in @fig-covariates.
-Of the four standard model parameters, models were least likely to consider covariates for initial occupancy (35% of articles with zero covariates considered), and most likely to consider them for detection (81% of articles with at least one covariate considered).
+For the vast majority of covariates authors considered only linear relationships with model parameters.
+Polynomial responses are rare — even in the categories where they are most frequently used, the rate of consideration for non-linear responses does not exceed 16%.
+This is a notable departure from SDMs; while GLM-based methods with linear responses are used, many common methods used to fit SDMs like MAXENT or boosted regression trees typically permit flexibility in response curves between covariates and parameters [@merow2013; @elith2008].
+Non-linear relationships are the norm, not the exception in nature, and in many cases a linear response will not adequately capture a covariate's influence on model parameters [@austin2002].
+Interactions between covariates were also rarely incorporate in our sample, and their rate of inclusion does not exceed 34% for any category.
+While they should not necessarily be used liberally, where these interactions exist their exclusion can influence model performance [@guisan2006].
 
-![Distribution of covariate quantity considered for model parameters across articles.](Figures/CovParamPlot.jpeg){#fig-covariates}
+Any model must balance simplicity and realism, but it is important that the ecological relationships quantitatively expressed in a model be grounded in ecological theory and accurately represent the modeller's hypothesised response [@austin2002; @austin2007].
+Complexity in the quantity and nature of covariate responses is also controlled during model selection, but these procedures can only work with the information provided to them.
+Adequately considering the plausible drivers of occupancy and detection by identifying quality candidate covariates representing those drivers is an important step in developing robust models that appropriately account for heterogeneity in these parameters.
 
-Model complexity varies with the stated objectives of articles, with those focussed on making predictions and identifying drivers of occupancy being more likely to consider a larger number of covariates (@tbl-objectives).
-These articles were also more likely to represent covariates in more complex ways by considering non-linear responses and interactions between covariates.
+### Covariate selection
+
+Variation in approaches to modelling is particularly apparent in the model selection process, where there is little consensus on how to determine which covariates to include in final models (@fig-selection).
+In the reviewed frequentist implementations, a range of approaches have been used for this purpose including comparison suites of candidate models defined *a priori* and various stepwise selection protocols.
+Bayesian implementations were far less likely to use model selection procedures — 70% of these fit only a single model defined *a priori,* relative to 10% amongst the frequentist models.
+
+![Frequency of use for model selection methods.
+'Single model' indicates no model selection occurred, 'candidate suite' includes studies which compared a predefined set of models, articles with 'procedural selection' fit parameters one-by-one, 'exhaustive selection' fit all possible combinations of models, and 'simple precursor' describes where results of a less-complex model were used to inform covariate choice.](Figures/SelectionFigure.jpeg){#fig-selection}
+
+Model selection is a particularly challenging aspect of fitting DOMs, with substantial uncertainties in the best ways to conduct the process.
+Limited research has been conducted on the topic, but what has been done has indicated that choices in this process can influence the quality of model outputs [@morin2020].
+This is reflected in research from outside the DOM literature, where stepwise protocols have been found to be unreliable [@olden2000].
+However, others have found that different model selection methods can end up arriving at roughly the same answer [@maggini2006].
+Model selection can be particularly difficult for DOMs given the increased complexity of the underlying model structure, which will often preclude exhaustive selection from amongst all possibilities.
+
+Bayesian model selection is an effectively distinct question in DOMs as the framework is not well suited to the more conventional model selection methods.
+@hooten2015 provides a guide to some of the possibilities for optimising Bayesian models, including the use of regularisation priors.
+
+### Model validation
+
+Regardless of whether a model was implemented under the Bayesian or frequentist frameworks, evaluation of model fit or performance was rare.
+Only 8% of articles directly addressed either aspect of modelling, be it via goodness-of-fit tests or validation with either internal or external data.
+This is a striking departure from the SDM literature, where roughly half of studies sufficiently assess their models performance [@araújo2019].
+While the appropriate method of model evaluation may depend on data availability and research objectives, assessing models by some method is always important to ensure reliability of model outputs [@araújo2005; @guisan2005].
+Most models in our sample conduct model selection using AIC as a scoring criteria, and often using model averages.
+While these approaches provide *relative* models of fit, they cannot be considered evaluation themselves - as @mackenzie2004 write, "the selection of a 'best' model(s) does not guarantee the selection of a "good" model.
+
+The first goodness-of-fit (GOF) test for occupancy model was given in @mackenzie2004, implemented via a parametric bootstrap.
+There is no GOF test available for DOM, however, common packages including 'AICcModAvg' and 'unmarked' offer parametric bootstraps derived from @mackenzie2004's test which have been used by some authors in our sample.
+While these tests may not suit Bayesian implementations, @broms2016 discusses possibilities for model evaluation in that context - while their article focuses on single-season multi-species occupancy models, extensions of their approach may suit the dynamic multi-season model as well.
+
+All aspects of modelling varied with the stated objectives of articles (@tbl-objectives).
+Those focussed on making predictions and identifying drivers of occupancy were more likely to consider a larger number of covariates and represented covariates in more complex ways by considering non-linear responses and interactions between covariates.
 Articles with objectives involving methods development considered fewer covariates and were less likely to use these more complex formulations.
@@
 
-Bayesian models were far less likely to be implemented using a model selection protocol compared to frequentist models: 70% of the former fit only a single model defined *a priori,* compared to 10% of the latter (@fig-selection).
-In frequentist models, sequential selection protocols in which parameters are fixed one by one were the most popular choice.
-Nearly all frequentist models used AIC or AICc to identify their best model.
+# Conclusions and key considerations
 
-![Frequency of use for model selection methods.
-'Single model' indicates no model selection occurred, 'candidate suite' includes studies which compared a predefined set of models, articles with 'procedural selection' fit parameters one-by-one, 'exhaustive selection' fit all possible combinations of models, and 'simple precursor' describes where results of a less-complex model were used to inform covariate choice.](Figures/SelectionFigure.jpeg){#fig-selection}
+Our review highlights the dynamic occupancy model's exceptional flexibility and suitability to addressing a range of ecological questions.
+The DOM may be a more achievable option than many think with feasible options to incorporate much existing presence/absence data from a variety of sources.
+Fertile areas for future research remain, with methods and guidance for making predictions and for modelling autonomous detection data from camera-traps and acoustic monitors particularly promising opportunities.
+In light of this potential and the DOMs status as a popular tool for applied ecology, it is increasingly important that best-practice standards be established for building these models.
+Major questions remain on fundamental questions in model building — there is need for further research on the appropriate levels of complexity for DOMs in various scenarios, on the methods for selecting models from amongst candidates, and on the best ways to evaluate the quality of the models chosen.
+In these areas, DOMs lag well behind SDMs where these standards have been more seriously considered [@araújo2006; @araújo2019; @zurell2020].
+Given the parallels between these model classes, we may be able to extrapolate some lessons from SDMs to support building better DOMs.
+In light of these uncertainties, and based upon existing research on SDMs, we discuss some key considerations for authors seeking to use DOMs in their work:
 
-Model evaluation and absolute assessment of model fit was rare throughout the sample regardless of selection protocol.
-Only 8% of articles calculated non-relative metrics, either via in-sample validation, bootstrapped goodness-of-fit tests, or out-of-sample validation.
-
-# Conclusions and key priorities
-
-# OLD —
-
-# Discussion
-
-One of the strengths of the dynamic occupancy model is its considerable flexibility — as seen in our review, applications of DOMs over the past two decades have been fit to data for a diverse cohort of species, using a broad range of detection methods and survey designs.
-This flexibility permits quite a bit of creativity on the part of authors and has allowed for innovative use-cases well beyond simple estimations of occupancy.
-Despite this promise, questions do remain on the best practices for fitting DOMs.
-Studies in our sample are discordant on their approaches the modelling process, and there the literature gives little insight on how these approaches may differ in their ability to identify reliable models.
-
-## Uses and broader potential
-
-## Outstanding questions on approaches to modelling
-
-## Key considerations for users
-
-# Conclusion
-
-### Applications of dynamic occupancy models
-
-DOMs display great variation of scale in all senses.
-Analyses in our samples range from the hyper-local at individual parks [@zúñiga-vega2019] to continental-scale studies [@zuckerberg2011].
-Similar diversity in study duration and site quantity indicates further emphasises the breadth of studies which DOMs may be fit to.
-Method of species detection also need not be a limitation for the implementation of DOMs, as articles in our sample show that many approaches may be coerced into data suitable for this model class.
-While studies pre-designed to match hierarchical data inputs may be simplest to align with model assumptions, post-hoc manipulation of other data such as long-term citizen monitoring programs [@zuckerberg2011; @peach2019], disparate agency modelling programs [@mcgowan2020], or camera trapping data [@davis2018] is feasible with careful consideration.
-Detections from multiple sources, with associated variation in efficiency, have also been incorporated by authors [@warrier2020; @pitman2017].
-
-Reviewed implementations of dynamic occupancy models show widespread representation from mammals, birds, and herptiles.
-This assemblage contrasts with the related stochastic patch occupancy models (SPOMs, @gutiérrez-arellano), where the majority of models are fit to invertebrates, plants, and fungi.
-This difference may reflect SPOMs' assumption of perfect detection and discrete habitat patches, limitations which are unlikely to be fulfilled by most vertebrate taxa.
-Aside from differences in assumptions around detection, SPOMs also differ from DOMs in that they are generally spatially explicit, with probabilities of colonisation and extinction dependent on patch size and the occupancy state of nearby patches.
-Where desired, these features may also added to dynamic occupancy models — @sutherland2014 presents an implementation which may be considered a fusion of the DOM and SPOM methods, and several other authors incorporate patch characteristics aspects into their models as covariates [@duggan2011; @broms2016].
-
-While objectives vary in our sample, DOMs are rarely used to extrapolate beyond study areas — use of prediction remains the least popular use-case in our sample.
-Articles which do make predictions provide some of the most directly applicable outputs found in the review sample: @mcgowan2020 provides projections for a threatened species under multiple putative management scenarios, and @pollentier2021 generates maps of distributions more frequently associated with species distribution models (SDMs).
-DOMs offer advantages over other SDMs for this purpose, particularly where populations are assumed not to be at equilibrium as is the case for many threatened and invasive species.
-By explicitly accounting for changes in occupancy (i.e., colonisation and extinction) DOMs may better capture future range contraction or expansion — an aspect of prediction where other SDMs may struggle.
-
-### Practices in the modelling process
-
-A diverse set of covariates were identified in the model sample reflecting the diversity of systems and ecological questions described by DOMs.
-Notably, many models lack covariates on the initial occupancy parameter, raising questions on how the parameterisation of the model in it's first time-step may affect the quality of estimates at later time-steps.
-The distinction in covariates considered for the ecological parameters vs those considered for detection is also of note: ecological parameters more frequently contained covariates describing environmental features, where detection more frequently contained more structural covariates such as season.
-Covariate complexity is relatively low across the board, with only a fraction of models implementing either non-linear responses to covariates or interactions between covariates.
-While this addition complexity does have implications for model selection, it may be necessary to accurately describe many ecological relationships.
-
-Model selection approaches diverge between Bayesian and frequentist implementations of the model.
-While model selection will naturally be practised differently between these two categories, it is unclear what effect a lack of model selection may have on the outputs produced by Bayesian models, and it is likely that Bayesian model selection approaches should be further explored @hooten2015.
-Model selection approaches in the Frequentist applications show limited consensus, although most do consider multiple models.
-Notably, @morin2020 found that the popular procedural approaches to model selection (where one sub-model is assessed at a time) may not provide optimal results.
-Considerably more research is required on this topic to clarify how authors should best approach covariate selection.
-
-### Key considerations for model users
-
-Implementation of dynamic occupancy models requires considerable thought to ensure that model outputs correspond with author expectations and ecological realities.
-Based upon our review of the existing literature we discuss key questions authors should carefully engage with during the modelling process.
-
-1.
-**Are seasons ecologically meaningful, and how do they affect the definition of occupancy?** As previously discussed, the duration of the season has important implications for both the reliability and proper interpretation of model results.
-Seasons which are not ecologically meaningful may run afoul of the closure assumption resulting in associated biases.
-Also relevant is how the definition of 'occupancy' may shift with season length and life-history of a species — for many species, occupancy within a year and within a week may be considerably different concepts, with shorter intervals more closely relating to the idea of 'use.' With the proliferation of continuous detection methods such as camera traps and acoustic monitors where the definition of 'season' is to some degree arbitrary, it is increasingly important to thoughtfully consider how this element of the DOM is defined.
-
-2.
+I.
 **Are key drivers of occupancy and detection incorporated?** While it is effectively impossible to *fully* describe heterogeneity across a landscape, authors should make their best efforts to capture the drivers of occupancy and detection most likely to influence their study species.
@@
 
-3.
-**Ensure covariates reflect ecological complexity.** Many ecological relationships are non-linear and may require polynomial terms to be accurately represented.
+II.
+**Do covariates relationships reflect ecological complexity?** Many ecological relationships are non-linear and may require polynomial terms to be accurately represented.
 This is particularly true at large spatial scales where a larger portion of a species ecological niche is represented in surveyed data.
-Authors should question whether a linear term is sufficient to capture hypothetical relationships in their study systems, and consider quadratic terms as well as interactions between covariates where appropriate.
+Authors should question whether a linear term is sufficient to capture hypothetical relationships in their study systems, and consider representing species responses as polynomials terms where appropriate.
+Interactions between covariates should also be explored where a plausible ecological relationship between factors exists.
 
-4.
-**Consider model evaluation.** Model evaluation remains rare in ecological research relative to other fields embracing complex modelling, likely due to a reluctance to 'throw away' data for use in testing.
-This concept should be challenged, particularly for larger scale long-term projects where reserving a fraction of data for testing is unlikely to substantially reduce precision of parameter estimates.
-For studies where this is genuinely not possible, reporting results of goodness-of-fit tests may still provide valuable information to readers.
+III.
+**How should model evaluation be conducted?** Regardless of why a model is used, it is important to assess its quality and its ability to support the conclusions drawn from it.
+While more research is needed on the most robust ways to assess models, existing methods should be used and reported in published studies.
+Dependent on data availability and model objectives, this can be achieved through bootstrapped goodness-of-fit tests or metrics of predictive performance on reserved data.
+This is an important step in establishing best-practices for the field, and should realistically be conducted in some form for the vast majority of studies.
 
-### Priorities for future research
-
-Work on dynamic occupancy modelling over the past 20 years has developed this model class into an important tool for describing species distributions and ecological relationships.
-Historically, methods-development research for dynamic occupancy models has focused on questions of model structure — namely ideas around the closure assumption and imperfect detection.
-This emphasis likely reflects the DOM's roots in the mark-recapture literature, where these topics have been discussed for several decades.
-Considerably less emphasis has been placed on the relation of model parameters to environmental factors via covariates and the model selection processes used to do so, despite this being an objective in a considerable fraction of studies fitting DOMs.
-This stands in stark contrast to related correlative species distribution models, where these topics have been the subject of considerable research over the years.
-There are opportunities for the DOM literature to learn from species distribution models, and more targeted research on the influence of the model selection process is necessary.
-
-An important opportunity for future research relates to autonomous detection methods, such as camera traps and bio-acoustic monitors — these technologies have permitted collection of enormous quantities of data particularly well-suited to dynamic occupancy models, but have also raised additional questions around their implementation.
-The definition of seasons is of increased importance for these data types, where division of data into 'seasons' and 'observations' is arbitrary.
-Further research should explore the implications of this phenomenon and the additional modelling opportunities made possible by this novel data type.
-
-The past two decades have shown that dynamic occupancy models are an excellent tool for addressing a wide range of ecological questions relating to species occupancy throughout space and time.
-The flexibility afforded by model extensions and the considerable advantages of the class over other species distribution models is likely to ensure that these models continue to increase in popularity.
-Given their strong use on threatened taxa and excellent potential to support environmental management, it is important that research continues to support model users in creating the most reliable models possible.
-
-## old
-
-First-authors of studies came from a total of 15 countries, with authors in the United States producing 64% of articles.
-86% of articles had at least one author based at an academic institution, 46% had author(s) at government institutions, and 25% had author(s) from non-profit organisations.
-
-## Study locations, focal species, and data collection
-
-Data for DOMs were collected from study areas on all continents and in 21 countries, with studies in the United States again representing a majority of the total articles published.
-These study areas ranged in size from less than 1 to over 1 million square kilometres (@fig-StudyDetails a).
-
-Focal species were largely vertebrate taxa, although 6 articles in the sample fit models to invertebrates.
-A single non-animal taxa was represented in the sample, from an article fitting a DOM to a rare species of lichen [@belinchón2017].
-35% of articles modelled threatened taxa, and 10% modelled species invasive to the study area.
-Most articles (50, 60%) fit single-species DOMs to one taxa.
-Of the remainder, 26 (31%) fit independent single-species models to more than one taxa.
-8 articles fit (10%) multi-species implementations with explicit interactions between species (@fig-StudyDetails b).
-
-Data used to fit DOMs included human survey data such as transects or point counts (69% of articles), live-trapping data (20%), camera traps (12%) or bioacoustic monitors (1%), and citizen science data (12%); some articles incorporate multiple detection methods to build their detection datasets.
-Study duration, defined as the time elapsed between the first and last surveys conducted, ranged from 1 month to 40 years with the median study running for 3.25 years.
-The number of seasons ranged from 2 to 204 with a median of 5; the number of observations per season ranged from 2 to 153 with a median of 5.
-The number of sites surveyed per article ranged from 7 to 9394, with the median study including 98 sites (@fig-StudyDetails c).
+As previously stated, DOMs and their outputs are important — in a biodiversity and conservation context where quantitative representations of complex natural systems are used to guide policy and decision making, it is necessary to ensure models can be trusted.
+We hope that this review can open a broader discussion on how to best approach dynamic occupancy modelling as a first step towards establishing best practices in the discipline.

### 2024-07-28 Randomisation/Archive/Randomise_Papers.qmd

--- 2023-11-27 Randomisation/Randomise_Papers.qmd
@@
     toc: true
+    embed-resources: true
 bibliography: references.bib
@@
 #| message: false
+
+options(tidyverse.quiet = TRUE)
 library("knitr")

### 2024-08-10 Analysis/ReviewFigures/Final_Figures.qmd
updates to manuscript and figures

---
title: "DOM Review Stats and Figures"
author: "Saoirse Kelleher"
date: today
abstract: "Compiles key statistics and figures for use in the DOM Review manuscript"
abstract-title: "Summary"
format: 
  html:
    theme: minty
    toc: true
embed-resources: true
---

# Final DOM review figures

```{r load packages}
#| include: false
library(tidyverse)
library(sf)
library(readxl)
```

# Load in review pages and randomisation sheet

Read in each page of the spreadsheet for analysis

## Article data

```{r Load article data page}
sheet_ArticleData <- read_xlsx("Spreadsheets/DOM_Spreadsheet.xlsx",
                               sheet = "Article Data")
```

## Objectives

```{r Load objective page}
sheet_Objectives <- read_xlsx("Spreadsheets/DOM_Spreadsheet.xlsx",
                              sheet = "Objectives") |>
  filter(`Review ID` != 0)
```

## Focal taxa

```{r Load focal taxa page}
sheet_FocalTaxa <- read_xlsx("Spreadsheets/DOM_Spreadsheet.xlsx",
                             sheet = "Focal Taxa") |>
  filter(`Review ID` != 0)
```

## Study area

```{r Load study area page}
sheet_StudyArea <- read_xlsx("Spreadsheets/DOM_Spreadsheet.xlsx",
                             sheet = "Study Area") |>
  filter(`Review ID` != 0)
```

## Data collection

```{r Load data collection page}
sheet_DataCollection <- read_xlsx("Spreadsheets/DOM_Spreadsheet.xlsx",
                                  sheet = "Data Collection") |>
  filter(`Review ID` != 0)
```

## Covariates

```{r Load covariates page}
sheet_Covariates <- read_xlsx("Spreadsheets/DOM_Spreadsheet.xlsx",
                              sheet = "Covariates") |>
  filter(`Review ID` != 0)
```

## Modelling

```{r Load modelling}
sheet_Modelling <- read_xlsx("Spreadsheets/DOM_Spreadsheet.xlsx",
                             sheet = "Modelling") |>
  filter(`Review ID` != 0)
```

## Randomisation sheets

```{r Load randomisation sheet}
randomisationSheets <- list(read_xlsx("Randomisation/Final_Randomisation.xlsx", sheet = "2004-2007"),
                            read_xlsx("Randomisation/Final_Randomisation.xlsx", sheet = "2008-2011"),
                            read_xlsx("Randomisation/Final_Randomisation.xlsx", sheet = "2012-2015"),
                            read_xlsx("Randomisation/Final_Randomisation.xlsx", sheet = "2016-2019"),
                            read_xlsx("Randomisation/Final_Randomisation.xlsx", sheet = "2020-2023")) |> 
  list_rbind()
```

# Taxa data

```{r Taxa bar plot}
focalTaxaPlot <- sheet_FocalTaxa |>
  mutate(Status = case_when(Status %in% c("CR", "EN", "NT", "Threatened_Other") ~ "Threatened",
                            Status == "NA" ~ "Multispecies",
                            Status == "Invasive" ~ "Invasive",
                            Status == "LC" ~ "Stable")) |>
  mutate(Status = fct(Status, levels = c("Multispecies", "Invasive",
                                         "Stable", "Threatened"))) |>
  summarise(Articles = n(), .by = c(`Taxa keywords`, Status)) |>
  select(Articles, Taxa = `Taxa keywords`, Status) |>
  mutate(Taxa = fct_reorder(Taxa, Articles)) |>

  ggplot() +
  geom_col(aes(x = Articles, y = Taxa, fill = Status)) +
  scale_fill_manual("", values = c("gray40", "plum4", "aquamarine4", "firebrick4")) +
  labs(x = "Articles", y = "") +
  theme(panel.grid = element_blank(),
        #legend.position = "top",
        legend.text = element_text(size = 20), axis.text = element_text(size = 21),
        axis.title = element_text(size = 19), axis.ticks.y = element_blank(),
        panel.background = element_rect(fill='transparent', color = NA),
        plot.background = element_rect(fill='transparent', color = NA),
        legend.background = element_rect(fill = 'transparent', color = NA))
focalTaxaPlot

ggsave("Manuscript/Figures/TaxaPlot.png",
       focalTaxaPlot, width = 10, height = 4, bg='transparent')

```

# Article data

Plot countries of first authors.

```{r Authorship countries}
sheet_ArticleData |>
  summarise(Articles = n(), .by = `First-author country`) |>
  arrange(-Articles) |>
  select(AuthorCountry = `First-author country`, Articles)
```

Determine estimated coverage

```{r Article coverage}
coveragePlot <- randomisationSheets |>
  summarise(N = n(), .by = c(Strata, Included)) |>
  pivot_wider(names_from = Included, values_from = N) |>
  mutate(HitRate = YES/(NO+YES)) |>
  mutate(EstRemaining = `NA`*HitRate) |>
  pivot_longer(cols = c(YES, EstRemaining), 
               names_to = 'Group', values_to = 'Articles') |>

  ggplot() +
  geom_col(aes(x = Strata, y = Articles, fill = Group), colour = "white") +
  scale_fill_manual("",
                    limits = c("YES", "EstRemaining"),
                    labels = c("Included articles", 
                               "Estimated qualifying articles"),
                    values = c("aquamarine4", "gray70")) +
  labs(x = "", y = "Number of articles") +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "top", axis.text = element_text(size = 12),
        axis.title = element_text(size = 12), legend.text = element_text(size = 12))

ggsave(plot = coveragePlot, "Manuscript/Figures/CoveragePlot.jpeg",
       width = 8, height = 4)
```

# Study areas

Plot study locations

```{r Study area maps}
# Load country shapefile
countries_shp <- st_read("Analysis/SpatialData/Countries/World_Countries__Generalized_.shp", quiet = TRUE) |>
  st_transform(crs = "ESRI:54030") |>
  filter(COUNTRY != "Antarctica")

studyArea_Map <- sheet_StudyArea |>
  st_as_sf(coords = c("Centre Point (lon)", "Centre Point (lat)"),
           crs = "EPSG:4326") |>
  st_transform(crs = "ESRI:54030") |>

  ggplot() +
  geom_sf(data = countries_shp, colour = "gray90", fill = "gray80") +
  geom_sf(shape = 21,fill = "aquamarine4", colour = "white", alpha = 0.7, size = 4) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.background = element_rect(fill='transparent', color=NA),
        plot.background = element_rect(fill='transparent', color=NA)) 
studyArea_Map
ggsave(plot = studyArea_Map, "Manuscript/Figures/StudiesMap.png")
```

Study area size

```{r Study area size histogram}
studyAreaPlot <- sheet_StudyArea |>
  summarise(Articles = n(), .by = Size) |>
  filter(Size != "Unspecified") |>

  ggplot() +
  geom_col(aes(x = Size, y = Articles),
           fill = "aquamarine4") +
  labs(x = "", y = "") +
  scale_x_discrete(labels = c("0-1", "1-10", "10-100",
                              "100-\n1000", "1000-\n10000",
                              "10000-\n100000", "100000-\n1000000",
                              "1000000+")) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(size = 20,
                                   angle = 90,
                                   vjust = 1, hjust = 1),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.background = element_rect(fill='transparent', color=NA),
        plot.background = element_rect(fill='transparent', color=NA))
studyAreaPlot

ggsave("Manuscript/Figures/StudyAreaPlot.png",
       studyAreaPlot, width = 8, height = 4, bg='transparent') 
```

# Data collection

Plot study duration

```{r plot study duration}
studyDurationPlot <- sheet_DataCollection |>
  mutate(StartDay = as.numeric(as.Date(`Start month`)),
         EndDay = as.numeric(as.Date(`End month`))) |>
  mutate(StudyDuration = EndDay - StartDay) |>
  mutate(StudyMonths = (StudyDuration/30)) |>
  mutate(StudyMonths = case_when(StudyMonths == 0 ~ 1,
                                 .default = StudyMonths)) |>

  ggplot() +
  geom_histogram(aes(y = StudyMonths), 
                 binwidth = 0.2, fill = "aquamarine4", colour = "white") +
  geom_hline(aes(yintercept = median(StudyMonths)), 
             colour = "goldenrod3", linetype = 2, linewidth = 2) +
  labs(y = "", x = "") +
  scale_y_continuous(transform = "log10",
                     limits = c(0.5,725),
                     breaks = c(1, 6, 12, 60, 
                                120, 240, 480, 720), 
                     labels = c("1 month", "6 months", "1 year", "5 years", 
                                "10 years", "20 years", "40 years", "60 Years")) +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill='transparent', color=NA),
        plot.background = element_rect(fill='transparent', color=NA),
        axis.text.y = element_text(size = 23),
        axis.text.x = element_blank(), axis.ticks.x = element_blank())
studyDurationPlot
ggsave(plot = studyDurationPlot, filename = "Manuscript/Figures/Duration.png",
       height = 8, width = 4)
```

Plot site quantity

```{r plot site quantity}
siteQuantityPlot <- sheet_DataCollection |>
  ggplot() +
  geom_histogram(aes(y = `Site quantity`),
                 binwidth = 0.25, fill = "aquamarine4", colour = "white") +
  geom_hline(aes(yintercept = median(`Site quantity`)), 
             colour = "goldenrod3", linetype = 2, linewidth = 2) +
  scale_y_continuous(transform = "log10",
                     limits = c(5, 10000),
                     breaks = c(10, 25, 50, 100, 250, 500,
                                1000, 2500, 5000, 10000)) +
  scale_x_continuous(expand = c(0,0)) +
  labs(x = "", y = "") +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill='transparent', color=NA),
        plot.background = element_rect(fill='transparent', color=NA),
        axis.text.y = element_text(size = 23),
        axis.text.x = element_blank(), axis.ticks.x = element_blank())
siteQuantityPlot
ggsave(plot = siteQuantityPlot, filename = "Manuscript/Figures/Sites.png",
       height = 8, width = 4)
```

Site quantity boxplots

```{r site quantity boxplots}
sheet_DataCollection |>
  left_join(select(sheet_ArticleData, c(`Review ID`, `Year Strata`)), by = "Review ID") |>
  mutate(outlier = case_when(`Site quantity` >= 500 ~ TRUE,
                             .default = FALSE)) |>
  mutate(`Site quantity` = case_when(`Site quantity` > 500 ~ 500,
                                     .default = `Site quantity`)) |>

  ggplot(aes(x = `Year Strata`, y = `Site quantity`)) +
  geom_jitter(aes(size = outlier, shape = outlier),
              colour = "cadetblue4", width = 0.1) +
  geom_boxplot(alpha = 0.4, colour = "cadetblue4",
               width = 0.4, outliers = FALSE) +
  scale_shape_manual(limits = c(TRUE, FALSE),
                     values = c(17,16)) +
  scale_size_manual(limits = c(TRUE, FALSE),
                    values = c(3,1)) +
  labs(x = "", y = "Sites") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())

```

Plot primary occasion quantity

```{r plot primary occasions}
sheet_DataCollection |>
  summarise(Articles = n(), .by = `Primary occasions`) |>

  ggplot() +
  geom_histogram(aes(x = `Primary occasions`), fill = "cadetblue4") +
  labs(x = "Number of primary occasions", y = "Frequency") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

Plot secondary occasion quantity

```{r plot secondary occasions}

sheet_DataCollection |>
  filter(!`Secondary occasions` %in% c('Unclear', 'Unspecified')) |>
  separate_wider_delim(`Secondary occasions`, delim = "-",
                       names = c("minSecondary", "maxSecondary"),
                       too_few = 'align_end') |>
  mutate(maxSecondary = as.numeric(maxSecondary)) |>
  summarise(Articles = n(), .by = maxSecondary) |>

  ggplot() +
  geom_histogram(aes(x = maxSecondary), fill = "cadetblue4") +
  labs(x = "Maximum secondary occasions", y = "Frequency") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())

```

Study scale over time

```{r study scale over time}
sheet_DataCollection |>
  left_join(select(sheet_ArticleData, c(`Review ID`, `Year Strata`)), by = "Review ID") |>
  filter(!`Secondary occasions` %in% c('Unclear', 'Unspecified')) |>
  separate_wider_delim(`Secondary occasions`, delim = "-",
                       names = c("minSecondary", "maxSecondary"),
                       too_few = 'align_end') |>
  mutate(maxSecondary = as.numeric(maxSecondary)) |>
  mutate(studyScale = `Site quantity`*`Primary occasions`) |>
  mutate(studyScale = case_when(studyScale > 5000 ~ 5000,
                                .default = studyScale)) |>
  mutate(outlier = case_when(studyScale >= 5000 ~ TRUE,
                             .default = FALSE)) |>

  ggplot(aes(x = `Year Strata`, y = studyScale)) +
  geom_jitter(aes(shape = outlier, size = outlier),
              colour = "cadetblue4", width = 0.1) +
  geom_boxplot(alpha = 0.4, colour = "cadetblue4", width = 0.4) +
  scale_shape_manual(limits = c(TRUE, FALSE),
                     values = c(17,16)) +
  scale_size_manual(limits = c(TRUE, FALSE),
                    values = c(3,1)) +
  labs(x = "", y = "Sites x Seasons") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

# Covariates

Covariate counts

```{r Covariate counts}
covParamPlot <- sheet_Covariates |> 
  separate_longer_delim(Covariates, delim = ", ") |>
  separate_wider_delim(Covariates, delim = "_", 
                       names = c("Covariate", "Variation", 
                                 "Observation", "Relation"),
                       too_few = "align_start") |>
  separate_wider_delim(Covariate, delim = "-",
                       names = c("Category", "Covariate"),
                       too_few = "align_start") |>
  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
                                 "Colonisation", "Extinction_Persistence",
                                 "Detection")) |>
  summarise(Covariates = n(),
            .by = c(`Review ID`, `Model ID`, `Parameter type`, Category)) |>
  complete(nesting(`Review ID`, `Model ID`, `Parameter type`), 
           Category, fill = list(Covariates = 0)) |>
  filter(Category != "None") |>
  mutate(`Parameter type` = case_when(`Parameter type` == "Extinction_Persistence" ~ "Extinction",
                                      `Parameter type` == "Occupancy" ~ "Occupancy*",
                                      .default = `Parameter type`)) |>
  mutate(Parameter = fct(`Parameter type`, 
                         levels = c("Initial occupancy", "Occupancy*",
                                    "Colonisation", "Extinction",
                                    "Detection"))) |>
  mutate(Covariates = case_when(Covariates >= 20 ~ 20,
                                .default = Covariates)) |>

  ggplot(aes(x = Category, y = Covariates)) +
  geom_jitter(aes(colour = Category), 
              alpha = 0.4, width = 0.2, height = 0.2) +
  geom_boxplot(aes(fill = Category), 
               outliers = FALSE, alpha = 0.3, colour = "gray60") +
  scale_fill_manual("", labels = c("Environmental covariates", "Structural covariates"),
                    values = c("aquamarine4", "plum4")) +
  scale_colour_manual("", labels = c("Environmental covariates", 
                                     "Structural covariates"),
                      values = c("aquamarine4", "plum4")) +
  scale_y_continuous(breaks = c(0, 5, 10, 15, 20),
                     labels = c("0", "5", "10", "15", "20+")) +
  labs(x = "", y = "Number of covariates considered") +
  facet_grid(cols = vars(Parameter), switch = "x") +
  theme(legend.position = "bottom", 
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(), 
        strip.text = element_text(size = 12, family = "Helvetica"), 
        axis.title.y = element_text(size = 13, family = "Helvetica"), 
        axis.text = element_text(size = 12, family = "Helvetica"),
        legend.text = element_text(size = 12, family = "Helvetica"))

covParamPlot
ggsave(plot = covParamPlot, filename = "Manuscript/Figures/CovParamPlot.jpeg",
       width = 8, height = 4)
```

Covariate table

```{r Covariate table}
# Covariate level summaries ------------------------------------------------
# Extract details for each covariate and get parameter counts
covariate_details <- sheet_Covariates |>
  separate_longer_delim(Covariates, delim = ", ") |>
  separate_wider_delim(Covariates, delim = "_", 
                       names = c("Covariate", "Variation", 
                                 "Observation", "Relation"),
                       too_few = "align_start") |>
  separate_wider_delim(Covariate, delim = "-",
                       names = c("Category", "Covariate"),
                       too_few = "align_start") |>
  mutate(n_overall = length(unique(`Review ID`))) |>
  mutate(n_param = length(unique(`Review ID`)), 
         .by = `Parameter type`) |>
  mutate(Dynamic = case_when(Variation %in% c("SE", "SISE", "SU") ~ TRUE,
                             .default = FALSE),
         DirectObs = case_when(Observation == "D" ~ TRUE, 
                               .default = FALSE),
         Nonlinear = case_when(Relation %in% c("N", "B") ~ TRUE,
                               .default = FALSE),
         Interaction = case_when(Relation %in% c("I", "B") ~ TRUE,
                                 .default = FALSE)) |>
  filter(!is.na(Covariate))

# Proportion of articles with covariate on any parameter 
covariates_PropAny <- covariate_details |>
  select(`Review ID`, Covariate, Category, n_overall) |>
  distinct() |>
  summarise(Prop_Any = n()/mean(n_overall), .by = c(Covariate, Category))

# Proportion of articles with covariate on each parameter
covariates_PropParameter <- covariate_details |>
  select(`Review ID`, Covariate, Category, `Parameter type`, n_param) |>
  distinct() |>
  summarise(Prop_Param = n()/mean(n_param), 
            .by = c(Covariate, Category, `Parameter type`)) |>
  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
                                 "Colonisation", "Extinction_Persistence",
                                 "Detection")) |>
  pivot_wider(names_from = `Parameter type`, values_from = Prop_Param,
              values_fill = 0)

# Proportion of covariates which are dynamic
covariates_dynamic <- covariate_details |>
  select(`Review ID`, Covariate, Category, Parameter, Dynamic) |>
  filter(Parameter != "Initial occupancy") |>
  summarise(Prop_Dynamic = sum(Dynamic)/n(),
            .by = c(Category, Covariate))

# Proportion of covariates which are directly observed
covariates_direct <- covariate_details |>
  select(`Review ID`, Covariate, Category, Parameter, DirectObs) |>
  summarise(Prop_Direct = sum(DirectObs)/n(),
            .by = c(Category, Covariate))

# Proportion of articles which use each cov non-linearly
covariates_nonlinear <- covariate_details |> 
  select(`Review ID`, Covariate, Category, Nonlinear) |>
  summarise(Nonlinear = max(Nonlinear), 
            .by = c(Covariate, Category, `Review ID`)) |>
  summarise(Prop_Nonlinear = sum(Nonlinear)/n(),
            .by = c(Covariate, Category))

# Proportion of articles which use each cov in an interaction  
covariates_interact <- covariate_details |> 
  select(`Review ID`, Covariate, Category, Interaction) |>
  summarise(Interaction = max(Interaction), 
            .by = c(Covariate, Category, `Review ID`)) |>
  summarise(Prop_Interact = sum(Interaction)/n(),
            .by = c(Covariate, Category))

# Combine all for covariates
covariates_table <- covariates_PropAny |>
  full_join(covariates_PropParameter, by = join_by(Covariate, Category)) |>
  full_join(covariates_dynamic, by = join_by(Covariate, Category)) |>
  full_join(covariates_direct, by = join_by(Covariate, Category)) |>
  full_join(covariates_nonlinear, by = join_by(Covariate, Category)) |>
  full_join(covariates_interact, by = join_by(Covariate, Category)) |>
  filter(!is.na(Covariate))

# Category level summary ----------------------------------------------
categories_PropAny <- covariate_details |>
  select(`Review ID`, Category, n_overall) |>
  distinct() |>
  summarise(Prop_Any = n()/mean(n_overall), .by = c(Category))

categories_PropParameter <- covariate_details |>
  select(`Review ID`, Category, `Parameter type`, n_param) |>
  distinct() |>
  summarise(Prop_Param = n()/mean(n_param), 
            .by = c(Category, `Parameter type`)) |>
  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
                                 "Colonisation", "Extinction_Persistence",
                                 "Detection")) |>
  pivot_wider(names_from = `Parameter type`, values_from = Prop_Param,
              values_fill = 0)

# Proportion of covariates which are dynamic
categories_dynamic <- covariate_details |>
  select(`Review ID`, Category, Parameter, Dynamic) |>
  filter(Parameter != "Initial occupancy") |>
  summarise(Prop_Dynamic = sum(Dynamic)/n(),
            .by = c(Category))

# Proportion of covariates which are directly observed
categories_direct <- covariate_details |>
  select(`Review ID`, Category, Parameter, DirectObs) |>
  summarise(Prop_Direct = sum(DirectObs)/n(),
            .by = c(Category))

# Proportion of articles which use each cov non-linearly
categories_nonlinear <- covariate_details |> 
  select(`Review ID`, Category, Nonlinear) |>
  summarise(Nonlinear = max(Nonlinear), 
            .by = c(Category, `Review ID`)) |>
  summarise(Prop_Nonlinear = sum(Nonlinear)/n(),
            .by = c(Category))

# Proportion of articles which use each cov in an interaction  
categories_interact <- covariate_details |> 
  select(`Review ID`, Category, Interaction) |>
  summarise(Interaction = max(Interaction), 
            .by = c(Category, `Review ID`)) |>
  summarise(Prop_Interact = sum(Interaction)/n(),
            .by = Category)

categories_table <- categories_PropAny |>
  full_join(categories_PropParameter, by = join_by(Category)) |>
  full_join(categories_dynamic, by = join_by(Category)) |>
  full_join(categories_direct, by = join_by(Category)) |>
  full_join(categories_nonlinear, by = join_by(Category)) |>
  full_join(categories_interact, by = join_by(Category)) |>
  mutate(Covariate = case_when(Category == "E" ~ "Environmental",
                               Category == "S" ~ "Structural"))

# Overall summary row ----------------------------------------------
overall_PropAny <- covariate_details |>
  select(`Review ID`, n_overall) |>
  distinct() |>
  summarise(Prop_Any = n()/mean(n_overall)) |>
  mutate(Category = "O", Covariate = "Overall")

overall_PropParameter <- covariate_details |>
  select(`Review ID`, `Parameter type`, n_param) |>
  distinct() |>
  summarise(Prop_Param = n()/mean(n_param), 
            .by = c(`Parameter type`)) |>
  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
                                 "Colonisation", "Extinction_Persistence",
                                 "Detection")) |>
  pivot_wider(names_from = `Parameter type`, values_from = Prop_Param,
              values_fill = 0) |>
  mutate(Category = "O", Covariate = "Overall")

# Proportion of covariates which are dynamic
overall_dynamic <- covariate_details |>
  select(`Review ID`, Parameter, Dynamic) |>
  filter(Parameter != "Initial occupancy") |>
  summarise(Prop_Dynamic = sum(Dynamic)/n()) |>
  mutate(Category = "O", Covariate = "Overall")

# Proportion of covariates which are directly observed
overall_direct <- covariate_details |>
  select(`Review ID`, Parameter, DirectObs) |>
  summarise(Prop_Direct = sum(DirectObs)/n()) |>
  mutate(Category = "O", Covariate = "Overall")

# Proportion of articles which use each cov non-linearly
overall_nonlinear <- covariate_details |> 
  select(`Review ID`, Nonlinear) |>
  summarise(Nonlinear = max(Nonlinear), 
            .by = `Review ID`) |>
  summarise(Prop_Nonlinear = sum(Nonlinear)/n()) |>
  mutate(Category = "O", Covariate = "Overall")

# Proportion of articles which use each cov in an interaction  
overall_interact <- covariate_details |> 
  select(`Review ID`, Interaction) |>
  summarise(Interaction = max(Interaction), 
            .by = `Review ID`) |>
  summarise(Prop_Interact = sum(Interaction)/n()) |>
  mutate(Category = "O", Covariate = "Overall")

overall_row <- overall_PropAny |>
  full_join(overall_PropParameter, by = join_by(Category, Covariate)) |>
  full_join(overall_dynamic, by = join_by(Category, Covariate)) |>
  full_join(overall_direct, by = join_by(Category, Covariate)) |>
  full_join(overall_nonlinear, by = join_by(Category, Covariate)) |>
  full_join(overall_interact, by = join_by(Category, Covariate))

# Combine all summaries ----------------------------------------------------
fullCovariateSummary <- covariates_table |>
  rbind(categories_table) |>
  rbind(overall_row) |>
  mutate(across(where(is.numeric), ~ round(.x, 2)))

write_csv(fullCovariateSummary, "Manuscript/Figures/CovariateTable.csv")
```

# Objectives

Objective proportions

```{r Objective proportions}
sheet_Objectives |>
  pivot_longer(c("Demonstrate methods?", "Observe trends?", 
                 "Identify drivers?", "Test relationships?",
                 "Predict spatially?", "Predict temporally?"),
               names_to = "Objective", values_to = "ObjectiveIncluded") |>
  filter(ObjectiveIncluded == "YES") |>
  mutate(Articles = n()) |>
  summarise(ObjArticles = n(), .by = c(Objective, Articles)) |>
  mutate(PropArticles = ObjArticles/Articles) |>
  mutate(Objective = fct_reorder(Objective, PropArticles)) |>

  ggplot() + 
  geom_col(aes(y = Objective, x = PropArticles),
           fill = "cadetblue4") +
  labs(x = "Proportion of articles", y = "") +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) 
```

Objective proportions through time

```{r Objectives through time}
sheet_Objectives |>
  left_join(select(sheet_ArticleData, c(`Review ID`, `Year Strata`)), by = "Review ID") |>
  pivot_longer(c("Demonstrate methods?", "Observe trends?", 
                 "Identify drivers?", "Test relationships?",
                 "Predict spatially?", "Predict temporally?"),
               names_to = "Objective", values_to = "ObjectiveIncluded") |>
  filter(ObjectiveIncluded == "YES") |>
  mutate(Articles = length(unique(`Review ID`)), .by = `Year Strata`) |>
  summarise(ObjArticles = n(), .by = c(Objective, Articles, `Year Strata`)) |>
  mutate(PropArticles = ObjArticles/Articles) |>
  mutate(Objective = fct_reorder(Objective, PropArticles)) |>
  complete(Objective, `Year Strata`, fill = list(PropArticles = 0)) |>
  mutate(Objective = factor(Objective, levels = c("Demonstrate methods?",
                                               "Observe trends?", 
                                               "Identify drivers?", 
                                               "Test relationships?",
                                               "Predict spatially?",
                                               "Predict temporally?"))) |>

  ggplot() +
  geom_col(aes(x = `Year Strata`, y = PropArticles, fill = Objective),
           position = "dodge", colour = "white") + 
  scale_fill_manual(values = c("plum4", "cadetblue3", "aquamarine3",
                               "aquamarine4", "goldenrod2", "goldenrod3")) +
  labs(y = "Proportion of studies") +
  lims(y = c(0, 1)) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = "top")
```

# Modelling

Bayes/Freq/All table

need:

n articles in each category

median covariates per core param

n using each selection method

n which look at fit

n which evaluate

```{r modelling table}
# get number of covariates per parameter for each model
modelling_Covariates <- sheet_Covariates |>
  separate_longer_delim(Covariates, delim = ", ") |>
  mutate(n_models = length(unique(`Model ID`)), .by = `Review ID`) |>
  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
                                 "Colonisation", "Extinction_Persistence",
                                 "Detection")) |>
  mutate(n_params = length(unique(`Parameter type`)), .by = `Review ID`) |>
  filter(Covariates != "None") |>
  summarise(CovsPerParam = n()/(max(n_params)*max(n_models)), .by = `Review ID`)

# compose table
modelling_details <- sheet_Modelling |>
  mutate(n_Overall = length(unique(`Review ID`))) |>
  mutate(n_Class = length(unique(`Review ID`)),
         .by = `Model class`) |>
  separate_longer_delim(`Performance evaluation`, ", ") |>
  separate_longer_delim(`Selection method`, ", ") |>
  mutate(GOF = case_when(`Performance evaluation` %in% c("GOF", "BayesianP") ~ TRUE,
                         .default = FALSE),
         Evaluate = case_when(`Performance evaluation` %in% c("Validation_In", "Validation_Out") ~ TRUE,
                         .default = FALSE)) |>
  mutate(Select = 1) |>
  pivot_wider(names_from = `Selection method`, 
              values_from = Select, values_fill = 0) |>
  summarise(GOF = max(GOF), Evaluate = max(Evaluate),
            `Model averaging` = max(`Model averaging`),
            `A priori` = max(`A priori`), 
            Sequential = max(Sequential), Simple = max(Simple),
            .by = c(`Review ID`, `Model class`, n_Class, n_Overall)) |>
  left_join(modelling_Covariates, by = join_by(`Review ID`))

modelling_byClass <- modelling_details |>
  summarise(Number = mean(n_Class),
            MedianCovs = median(CovsPerParam, na.rm = TRUE),
            Prop_ModAvg = sum(`Model averaging`)/mean(n_Class),
            Prop_AnySelect = 1-(sum(`A priori`)/mean(n_Class)),
            Prop_Sequential = sum(Sequential)/mean(n_Class),
            Prop_Simple = sum(Simple)/mean(n_Class),
            Prop_GOF = sum(GOF)/mean(n_Class),
            Prop_Eval = sum(Evaluate)/mean(n_Class),
            .by = `Model class`) |>
  mutate(across(where(is.numeric), ~ round(.x, 2)))

modelling_overall <- modelling_details |>
  summarise(Number = mean(n_Overall),
            MedianCovs = median(CovsPerParam, na.rm = TRUE),
            Prop_ModAvg = sum(`Model averaging`)/mean(n_Overall),
            Prop_AnySelect = 1-(sum(`A priori`)/mean(n_Overall)),
            Prop_Sequential = sum(Sequential)/mean(n_Overall),
            Prop_Simple = sum(Simple)/mean(n_Overall),
            Prop_GOF = sum(GOF)/mean(n_Overall),
            Prop_Eval = sum(Evaluate)/mean(n_Overall)) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  mutate(`Model class` = "All Models")

modelling_summary <- rbind(modelling_byClass, modelling_overall) |>
  t() |> as.data.frame()
```

### 2024-08-10 Manuscript/domReview.qmd
updates to manuscript and figures

--- 2024-04-12 Manuscript/domReview.qmd
@@
 
-bibliography: references.bib
-embed-resources: true
+abstract: | 
+  Describing patterns of species occupancy across landscapes and throughout time is a fundamental aspect of much ecological research.
+The dynamic multi-season occupancy model is an important tool for analysing wildlife occurence data -- by explicitly estimating colonisation and extinction probabilities while accounting for imperfect detection, it strikes a favourable balance between realism and feasibility.
+These models can be used in similar ways to more conventional species distribution models, which can struggle to deal with imperfect detection and non-equilibrium populations.
+
+  We present a review of applications of the dynamic occupancy model in the twenty years since its initial publication.
+Our findings indicate that this model demonstrates exceptional flexibility with suitability to address a wide range of ecological questions, and that it may have  broader potential than commonly thought.
+We also note that practices in fitting dynamic occupancy models are highly variable, and that the existing literature lacks clear answers for best practices.
+This contrasts with the strong focus on modelling practices in the species distribution model literature, which we suggest may be a valuable resource for improving how authors implement dynamic occupancy models.
 ---
@@
 
-Since its introduction two decades ago by @mackenzie2003, the dynamic multi-season occupancy model has become an important tool for analysis of species occupancy throughout space and time.
-Dynamic occupancy models (DOMs) represent a happy medium in the ecological modelling landscape.
-In explicitly estimating the processes which guide occupancy (i.e., colonisation and extinction) this model class offers more advanced insights than simpler static models of species distribution which have noted deficits in their ability to describe non-equilibrium states **\[cite\]**.
-The capacity to account for imperfect detection provides a strong additional benefit given the near-ubiquity of this source of bias in ecological field data.
-Even with these advantages, data input requirements for DOMs remain achievable for many research programs, particularly compared to other more complex process-explicit models.
+Ecologists have long sought to describe the processes by which species are distributed across landscapes and throughout time [@humboldt1849], and an understanding of where a species exists, which factors lead to its presence, and how its distribution may change is increasingly important in a dynamic and changing world.
+Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs assessing conservation status, and identifying potential drivers of occurrence can help inform potential management actions.
+Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur, both under present conditions and hypothetical future scenarios.
 
-Potential use cases of this model class are wide-ranging.
-Simple estimates of occupancy through time have value in many monitoring programs, particularly for threatened or invasive species where distributions are likely to exist in a non-equilibrium state **\[cite?\]**.
-These models can also be used in hypothesis testing of specific relationships, with examples including the effect of fire on extinction probability and the influence of species interactions on occupancy **\[cite,cite\]**.
-A less frequently pursued yet promising application of DOMs lies in their ability to make predictions, either to unsurveyed areas or into the future [@briscoe2021; @kéry2013a].
+The inherent value in understanding these patterns of occurrence has led to the development of several techniques for modelling occupancy.
+However, several factors can make occupancy challenging to describe quantitatively.
+For instance, when conducting surveys it is often impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer.
+This 'imperfect detection' is a near-ubiquitous characteristic of wildlife presence/absence data and is known to bias estimates of occupancy in conventional SDMs [@lahoz-monfort2014; @gu2004].
+Despite this, many popular models make no adjustments for this source of bias [@kellner2014].
+Another challenge lies in modelling occupancy under non-equilibrium conditions where a species' distribution or relationship to its environment is in flux, as occurs during range shifts or biological invasions.
+Conventional correlative SDMs often perform poorly in these scenarios [@dormann2007; @elith2010], a noteworthy challenge given the increased frequency of these events in the Anthopocene [@bertelsmeier2013; @lenoir2015].
 
-This review aims to provide an overview of the first two decades of the dynamic occupancy models, including its development and how authors have implemented it in their investigations of ecological questions.
-We present key guidelines for authors seeking to build these models in their work, and outline the key priorities and opportunities for future research incorporating this flexible and important model class.
+Occupancy models, first introduced in @mackenzie2002, offer an alternative framework for modelling occurrence while accounting for these challenges.
+By using repeat visits to each site, static occupancy models explicitly describe the observation process and account for imperfect detection to correct for bias in estimates of site occupancy.
+@mackenzie2003 extended this model to account for changes in occupancy across time periods by explicitly describing processes of colonisation and extinction, relaxing the assumption of equilibrium present in other models of species occurrence.
+This dynamic occupancy model (DOM) is well-suited to address many research questions in ecology, offering a balance between achievable input data requirements and the complexity necessary to describe intricate natural systems.
 
-### The dynamic occupancy model
+This review aims to provide an overview of dynamic occupancy models, their uses, and the ways that they are implemented.
+Following an introduction to the model form and assumptions, we present a systematic review exploring how authors have used the DOM in the two decades since their development, with emphasis on how they collected data, selected covariates, and evaluated their models.
+Based on these results we highlight the DOMs flexibility as a tool for understanding occupancy, examine how authors have approached the modelling process, and outline key priorities for future research involving this important model class.
 
-Dynamic occupancy models jointly describe two processes: the ecological process of site occupancy describing the presence or absence of a species at a given site in a given point in time; and the observational process of detection describing whether a species is observed given that a site is occupied.
-The basic model form as described by @mackenzie2003 is depicted in @fig-modelform.
-Under this parameterisation sites exist in either occupied or unoccupied states which they may transition between through time.
-Occupancy state in the first time-step is given by the initial occupancy parameter, and in ensuing time-steps occupancy is described as a Markovian process governed by probabilities of colonisation and extinction.
-Given that a site is occupied, the probability of observing the species during a survey is given by the detection parameter.
-Under this formulation, it is assumed that false-positive detections do not occur.
+## Dynamic occupancy model form and assumptions
 
-![The original parameterisation of the dynamic occupancy model.
-The ecological process of occupancy is guided by the initial occupancy, colonisation, and extinction parameters; the observational process is guided only by the detection parameter.](Figures/ModelForm.png){#fig-modelform width="1000"}
+Dynamic occupancy models unite two processes: the ecological process of site occupancy describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed given that a site is occupied (@fig-modelform).
+Under this parameterisation, sites exist in either occupied or unoccupied states.
+In the first step-step, occupancy state at each site is determined by the probability of initial occupancy.In subsequent time steps occupancy state is described as a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation and extinction.
+The observation component of the DOM accounts for imperfect detection - at occupied sites, the detection probability describes whether or not the species is observed during a survey.
+Under the DOM's standard parameterisation, it is assumed that false-positive detections never occur at unoccupied sites.
 
-To disentangle the ecological and observation processes and estimate the detection parameter, the DOM relies on a hierarchical sampling design shown in @fig-surveys.
-Under this design, observations at a site occur during distinct, time-bound seasons in which sites are considered 'closed' to changes in occupancy.
-Within seasons multiple observations are conducted, which permit estimation of the probability of detection conditional on occupancy.
-These repeat observations may be conducted by a variety of methods.
-Most frequently surveys are simply repeated on separate occasions, although alternative methods exist.
-These include surveys conducted at multiple locations within a site **\[cite\]**, use of multiple observers during a survey **\[cite\]**, or by recording time elapsed before a detection is recorded **\[cite\]**.
+![The dynamic occupancy model as described by @mackenzie2003.
+In the ecological process submodel, the occupancy state of any given site is determined by the *initial occupancy* parameter.
+In subsequent time-steps, unoccupied sites may become occupied according to the *colonisation probability* and occupied sites may become unoccupied according to the *extinction probability*.
+The observational process submodel accounts for imperfect detection: during a survey at occupied site, the species may be detected or not according to the *detection probability*.
+At sites which are truly unoccupied, is it assumed that the species is never detected.](Figures/ModelForm.png){#fig-modelform}
 
-![The sampling design of the dynamic occupancy model.](Figures/SurveyDesign.png){#fig-surveys fig-align="center" width="589"}
+To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys.
+Under this design, observations at a site occur during distinct, time-bound seasons within which sites are considered closed to changes in occupancy.
+In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy.
+Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@mackenzie2005].
 
-Dynamic occupancy models require a few assumptions with important implications for users.
+![The sampling design of the standard dynamic occupancy model.
+During seasons, also called primary occasions, sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons.
+Within each season, multiple observations ('secondary occasions') are conducted to record the observed presence or absence of the species at each site.
+These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded.
+It is not necessary for the same number of observations to occur in each year or each site, allowing for greater flexibility in data inputs.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}
 
-1.
-**False positive detections do not occur.** While this assumption can be safely met in some studies, it is not guaranteed when working with more cryptic species.
-Several authors have commented on the perils of not accounting for false positives where they occur (@miller2015, @mcclintock2010).
-Furthermore, @berigan2019 notes that in some cases genuine identifications of a species may be considered false positives if they do not represent the right kind of occupancy - in this case, transient individuals where the authors were interested in breeding occupany.
-Extensions to the dynamic occupancy model which account for false positive error are available (@royle2006, @miller2011, @miller2013) and should be carefully considered where this assumption cannot be met.
+The descriptions above are based on the original form of the dynamic occupancy model as prescribed in @mackenzie2003.
+Numerous model extensions and alternative formulations are also available, including implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models[@dorazio2010].
+For a comprehensive discussion of the most common extensions and their applications see @bailey2014, as well as @devarajan2020 for a review of multi-species occupancy models.
 
-2.
-**Sites are closed to changes in occupancy between seasons.** This assumption, best known as the 'closure assumption,' has also been subject to considerable discussion on the bias which is introduced when it is violated (@rota2009, @otto2013).
-Closure is dependent not only on the behaviour of the species, but also the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discusses the difficultly in distinguishing temporary emigration and local extinction.
-Model extensions to relax the closure assumption have been developed, including @kendall2013 's approach using staggered arrival and depature periods between sites.
-More pertinent for most studies, however, is a careful consideration of ecologically and observationally relevant seasons which will allow the estimation of an appropriate 'type' of occupancy, whether that be shorter or longer term.
+Dynamic occupancy models make a small number of assumptions with important implications:
 
-3.
-**No modelled heterogeneity in ecological or observational parameters.** Relative to the former two assumptions, considerably less research has been conducted on the impact of unmodelled heterogeneity on parameter estimates.
-Particularly at larger spatial scales, environmental factors are near-guaranteed to influence species occupancy and observation.
+I.
+**False positive detections do not occur**.
+While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods.
+@mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need for authors to consider how certain their detections truly are.
+Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019].
+Where this assumption can not reasonably met, aforementioned extensions designed to account for false positive error should be considered.
 
-# Methods
+II.
+**Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion around the bias which is introduced when it is violated [@rota2009; @otto2013].
+Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discuss the difficultly in distinguishing between temporary emigration and local extinction.
+Model extensions to relax the closure assumption have been developed, including @kendall2013's approach using staggered arrival and departure periods between sites.
+A more pertinent approach for most studies, however, is careful consideration of ecologically relevant seasons corresponding to an appropriate definition of occupancy.
 
-## Article solicitation
+III.
+**Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species distributions, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system.
+Generally, this is achieved by allowing model parameters to vary with respect to covariates representing environmental factors which may be expected to influence these characteristics.
+No model will fully account for the complexity inherent in patterns of species occupancy; however, failure to capture key drivers can introduce bias.
+Compared to the first two assumptions, this aspect of DOMs has been less thoroughly discussed and comparatively little is known how this latent heterogeneity can influence model estimates.
 
-This review is focused on the dynamic, multi-season, implementation of the occupancy model of @mackenzie2003.
-To qualify for inclusion articles must fit a model which fulfils these criteria: i) multiple sites which can exist in at least two states, including an occupied and unoccupied states; ii) multiple survey periods, between which sites may change state conditional on the prior survey period's occupancy state and at least one transition probability; iii) contains at least one detection parameter; iv) uses real (i.e., not simulated) field ecological data.
+## Building dynamic occupancy models
 
-Papers were solicited for inclusion in two stages.
-In a preliminary analysis, four search strings ("dynamic occupancy model," "occupancy dynamics model," "incidence function model," and "multi-season occupancy model" were fed into Google Scholar **\[cite\]**.
-Up to 100 articles were downloaded for each string in order of appearance.
-Articles were divided into four temporal strata: 2000-2005, 2006-2010, 2011-2015, and 2016-2021.
-For each combination of search string and year strata, either a random 25% or 5 articles (whichever was greater) were processed.
-Articles which did not meet inclusion criteria were replaced with another from the same search string and strata where possible.
+A key aim of this review is to investigate how authors have approached the steps of the modelling building process, and how these practices have changed since the introduction of these models 20 years ago.
+The process of fitting DOMs requires decisions by model builders at several key points with meaningful implications for how outputs should be interpreted.
+This process begins with the project's inception and plans for data collection.
+The definition of occupancy is contingent on
 
-After completion of the preliminary sample authors recognised significant uncertainty in how the model of interest was described, particularly in their first decade of use, and noted that the incidence-function model is a distinct model class **\[cite\].** To increase sample size, a complete list of articles which have cited @mackenzie2003 was downloaded from Web of Science.
-Articles in this list were split into the same year strata, plus an additional 2022-2023 category.
-15 articles were drawn from each strata and again processed with replacement from their strata.
+Authors must also consider the temporal aspect of occupancy by defining primary and secondary occasions.
+In many cases, this is a simple decision driven by data collection - take a typical bird survey program, where sites may be visited on a few occasions in the breeding season of each year; years and individual visits within years will generally be used as primary and secondary occasions, respectively.
+However, this choice may not be as clear with other data collection methods.
+Citizen science programs, camera traps, acoustic monitors can all yield somewhat continuous data which must be arbitrarily segregated into primary and secondary occasions.
+In this context, it is important to consider both the temporal scale of interest for occupancy and whether or not closure can reasonably be assumed between primary occasions.
 
-Details on methods for each article were recorded in detail, including the full suite of covariates considered for inclusion in modelling.
-The full spreadsheet of reviewed articles is available in **Appendix I**.
+Once data has been coerced into a suitable structure for fitting DOMs, modellers must also consider which environmental factors to incorporate into the model, as well as how to incorporate them.
+These factors will vary considerably by study, but should generally be plausible drivers of variation in either occupancy dynamics or survey observability.
+Determining which candidate covariates should be included in a final model for drawing inference will generally involve some form of model selection.
+Covariate selection can be particularly challenging for DOMs relative to other models like SDMs due to their increased complexity.
+Consider a model with 6 candidate covariates, each of which can be included in any parameter.
+The number of permutations of this model is given as (2^n^)^p^, where n is the number of candidate covariates and p the number of parameters.
+A simple single-parameter SDM would have 64 possible formulations, whereas a typical 4-parameter DOM would have over 16 million candidate models given the same covariate pool.
+For many studies, this precludes exhaustive comparisons of candidate models and demands some form of simplified model selection protocol [@doherty2012].
+Multi-model inference is also a common strategy with DOMs, allowing for uncertainty in the difference in fit by averaging estimates from across competing models [@burnham2004].
 
-## Analysis
+Several computational tools exist for implementing the DOM and related models.
+Widely-cited frequentist options include the R package 'unmarked' and the GUI-based programs PRESENCE and MARK.
+Bayesian DOMs can be written in programs such as JAGS, BUGS, or STAN and fit via R packages such as rJAGS or brms*.* The recently developed R package 'ubms' offers a straightforward Bayesian implementation through an interface similar to *unmarked,* without requiring writing the DOM from scratch.
 
-For each included article we collected data collected on authorship, stated research objectives, study taxa and field survey locations, data collection methods, covariates, and model selection and evaluation.
-Particular emphasis was placed on the covariates considered by authors, with no distinction between covariates which were or were not included in final models.
-As we were principally interested in the modelling process, our analysis did not directly explore final models or the results generated by those models.
-A spreadsheet of reviewed articles and collected data is provided in Appendix I.
+Upon selecting a final model, authors may evaluate model performance by different methods dependent on the objectives of their studies.
+While no dedicated goodness-of-fit test exists for DOMs, a version is available for single-season models [@mackenzie2004] which has been extended to multiple seasons; functions are available in *unmarked* and *aicCmodavg.* Another possibility for assessing model fit and predictive ability is via validation.
+As true occupancy state is generally not known for DOM data, the typical response variable is the observed occupancy state.
 
-# Results
+# Systematic review methods
 
-A total of **\[number\]** articles were included in the review, **\[number\]** from the preliminary sample and **\[number\]** from the secondary sample.
-Based on the acceptance rate from the secondary sample, an estimated **\[hit number\]** of the **\[total number\]** articles citing @mackenzie2003 would have met inclusion criteria.
-Total published articles have increased considerably over time.
+To assess how DOMs have been applied in the years since their introduction, we gathered a representative sample of articles fitting them to field ecological data.
+A pool of candidate articles was generated using two queries on Web of Science.
+The first included all articles from 2004-2023 citing @mackenzie2003.
+The second query searched articles matching "dynamic occupancy model\*", "multi-season occupancy model\*", and "occupancy dynamic\*" in any field, articles with each of "occupancy", "colonization", "extinction/persistence", and "detection" in any field, and articles with the term "occupancy" near "dynamic".
 
-![Estimated coverage of articles included in this review](Figures/CoveragePlot.jpeg){#fig-coverage}
+As we were interested in how DOMs use has changed through time, we divided articles across five four-year strata: 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024.
+From each of these strata we randomly selected 20 articles for inclusions in the review; articles which did not meet inclusion criteria were replaced from within their own strata.
 
-Articles were published with first-authors from a total of **\[number\]** countries, with authors from the United States producing **\[zz\]**% of articles.
-**\[x\]**% of articles had at least one author based at an academic institution, **\[y\]**% had authors at government institutions, and \[**z\]**% had authors from non-profit organisations.
+Our review is focused on applications of the dynamic, multi-season occupancy model of @mackenzie2003 and its extensions.
+To reflect this, we included articles which fit a model meeting the following criteria:
 
-## Study locations, focal species, and data collection
+I.  Uses non-simulated, field-collected presence-absence data.
 
-Data for DOMs were collected from study areas on all continents and in **\[xx\]** countries, with the United States again representing a majority of the total articles published.
-These study areas ranged in size from less than 1 to over 1 million square kilometres (@fig-StudyDetails a).
+II. Describes multiple sites which can exist in at least two states including *occupied* and *unoccupied.*
 
-Focal species were largely vertebrate taxa, although 6 articles fit models to invertebrates.
-A single non-animal taxa was represented in the sample, from an article fitting a DOM to a rare species of lichen @belinchón2017.
-[aa%] of articles modelled threatened taxa, and \[bb%\] modelled species invasive to the study area.
-The majority of articles (50, X%) fit single-species DOMs to one taxa.
-Of the remainder, 26 (X%) fit independent single-species models to more than one taxa.
-8 articles fit (X%) multi-species implementations with explicit interactions between species (@fig-StudyDetails b).
+III. Has multiple seasons, between which sites may change states conditional on the prior season's occupancy state and transition probabilities such as colonisation and extinction.
 
-Data used to fit dynamic occupancy models included human survey data such as transects or point counts (X% of articles), live-trapping data (X%), camera traps (X%) or bioacoustic monitors (Y%), and citizen science data (Z%); some articles incorporate multiple detection methods to build their detection datasets.
-Study duration, defined as the time elapsed between the first and last surveys conducted, ranged from 1 month to 40 years with the median study running for X years.
-The number of seasons ranged from 2 to Y with a median of Z; the number of observations per season ranged from X to Y with a median of Z.
-The number of sites surveyed per article ranged from X% to Y%, with the median study including Z sites (@fig-StudyDetails c).
+IV. Contains at least one parameter describing the detection process.
 
-![](Figures/ReviewBox.png){#fig-StudyDetails}
+For each article we recorded key details on authorship, research objectives, study taxa and system, survey methods, and modelling approach.
+To assess patterns of authorship, the location of each article's first author's primary affiliation was recorded.
+*All* affiliations for all authors were classified into one of four categories: academic institutions, governmental organisations and institutes, non-governmental organisations, and private sector companies.
 
-## Covariates
+We recorded details on the type of taxa (bird, mammal, etc.) modelled in each article, and how multiple species were modelled where applicable.
+Taxa were denoted as threatened either when they are listed on the IUCN Red List of Species or when authors otherwise indicate that they are threatened.
+This deference to authors' representation of conservation status was made to account for sub-species which lack listings or species which are of more local concern.
+Study locality and size was documented; the size of the study area being defined as the intended area of inference containing all sites, recorded as an order of magnitude to partially account for uncertainty in reporting.
 
-A summary of the covariates classes considered in the review sample is presented in @tbl-covariates. The most frequent
+To examine the reasons why authors used DOMs, we allocated each article to one or more category of objective based on the study's stated aims.
+The possible categories were *Estimating parameters*, where authors express interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; *Testing hypotheses*, where authors explore specific predefined relationships between covariates and model parameters; *Identifying drivers*, where authors attempt to find which covariates influence model parameters; *Generating predictions*, where authors estimate model parameters either for unsurveyed sites or under possible future conditions; and *Methods development*, where authors introduce, test, or demonstrate aspects of dynamic occupancy models.
 
-| Covariate type | Proportion of articles considering | Proportion including on Initial Occupancy | Proportion dynamic | Proportion directly observed | Frequency of interaction terms | Frequency of polynomial terms |
-|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
-|                |                                    |                                           |                    |                              |                                |                               |
-|                |                                    |                                           |                    |                              |                                |                               |
-|                |                                    |                                           |                    |                              |                                |                               |
-|                |                                    |                                           |                    |                              |                                |                               |
-|                |                                    |                                           |                    |                              |                                |                               |
-|                |                                    |                                           |                    |                              |                                |                               |
-|                |                                    |                                           |                    |                              |                                |                               |
-|                |                                    |                                           |                    |                              |                                |                               |
-|                |                                    |                                           |                    |                              |                                |                               |
-|                |                                    |                                           |                    |                              |                                |                               |
+Particular emphasis was placed on describing how authors conducted their modelling from covariate selection through model evaluation.
+We recorded all covariates considered in each study, regardless of whether they were or were not included in final models.
+Key traits of each covariate were recorded including their general category, whether they were directly observed or remotely sensed, whether they were static or varied between seasons, and how they were represented in the model (as a linear term, a polynomial term, or as part of an interaction with another covariate).
+Model selection procedures were sorted into non-exclusive categories including *A priori*, where only one model was considered; *candidate suite*, where a predefined set of models was considered; *procedural*, where covariates were selected parameter-by-parameter; *exhaustive*, where all possible model combinations were fit; *simple precursors*, where selection was preceded by another simpler model implementation; and *other*.
+Model evaluation was similarly categorised - options in this case included *goodness of fit*, where absolute metrics of fit were used; and *out-of-sample validation* and *in-sample-validation*, where predictive performance was evaluated using in-sample or out-of-sample data respectively.
+For the full spreadsheet of data collected and further details on categorisation, see **Appendix I.**
 
-: Covariates considered in dynamic occupancy models {#tbl-covariates}
+A total of 84 articles were included in this review.
+Based on the acceptance rate from the secondary sample, an estimated 448 of the 1066 articles in our sample would have met inclusion criteria (@fig-coverage).
 
-![](Figures/CovParamPlot.jpeg){#fig-covariates}
+![Estimated coverage of articles included in this review, based on the secondary sample's rate of inclusion for each strata.](Figures/CoveragePlot.jpeg){#fig-coverage}
 
-### Complexity
+# Applications of DOMs
 
-![](Figures/ModelComplexity.png){#fig-complexity}
+Dynamic occupancy models are considerably flexible in many ways: they may used to achieve several different objectives, fit to presence-absence data from a variety of sources, and applied to studies at a wide range of spatial and temporal scales.
+Study locations included in our sample are globally distributed [@fig-StudyDetails A] and span a diverse range of ecosystems.
+Notably, these study areas vary considerably in size — the smallest study locality included in our sample studied insect occurrence in a rainforest plot smaller than one square kilometre [@basset2023], while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@clement2019].
 
-### Selection and evaluation
+![A) Study areas: Data for a majority of studies were collected from study locations in the United States.
+The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres.
+\n B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles.
+50 articles fit models to only a single taxa, 26 fit independent models to multiple taxa, and 8 fit multi-species models with explicit interactions between taxa.
+\n C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors.
+10 articles incorporated data from citizen science projects.
+Project scale was also variable, with the median study running for 3.25 years and covering 98 sites.](Figures/ReviewBox.png){#fig-StudyDetails}
 
-Bayesian models were far less likely to implement a model selection protocol compared to frequentist models: 30% of the former fit only a single model defined *a priori,* compared to 10% of the latter (@fig-selection).
-In frequentist models, stepwise selection protocols were the most popular choice.
-The vast majority of frequentist models (XX%) used AIC or AICc to identify the best model.
+### Focal systems
 
-![](Figures/SelectionFigure.jpeg){#fig-selection}
+Although studies were conducted in 21 countries and on all continents except for Antarctica, the majority of reviewed articles used data collected in the United States of America.
+This strong geographic trend starkly contrasts with SDMs far more international reach; @araújo2019's review of SDM applications found that no more than 20% of studies came from any single continent.
+The reasons behind this disparity are unclear, but a possible explanation could relate to the DOM's roots in the mark-recapture literature and a history of publication in American journals.
+These differences in where users conduct their research may also contribute to the seemingly limited overlap between these models' respective bodies of literature.
 
-Objective evaluation of model fit was rare throughout the sample, regardless of selection protocol.
-XX% of articles calculated non-relative metrics, either via in-sample validation, bootstrapped goodness-of-fit tests, or out-of-sample validation.
+DOMs have more rarely been applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community.
+However, there are exceptions — @belinchón2017 fits a DOM to lichen data, and @cook2022 uses them to model the spread of chronic wasting disease.
+The latter's application to disease dynamics is not unique, and DOMs have been a reasonable choice for modelling disease dynamics.
+@mores2020 and @padilla-torres2013 have each used DOMs to model mosquito dynamics due to their importance as a human disease vector, and @bailey2014 discusses the feasibility of applying the DOM to study the spread of chytrid fungus in amphibian habitat.
 
-# Discussion
+Differences in use between SDMs and DOMs are also evident in the species targeted for modelling.
+Where SDMs show broader coverage of plants, invertebrates, and aquatic species [@araújo2019], 93% of studies in our sample of DOMs modelled terrestrial vertebrate fauna with @belinchón2017's study on lichen the sole non-animal taxa representative.
+These differences are most easily explained by the DOM's emphasis on imperfect detection, a concern which is less prioritised (though not non-existent [@chen2013]) in studies on non-mobile species like plants.
+We find DOMs have been frequently used to model threatened taxa (32% of studies) and invasive species (11% of studies), taxa for which they may be expected to excel due to their more explicit handling of changes in occupancy.
+One group for which DOMs have only rarely been used is aquatic species, as in @fisher2014's application to invasive salmon or @falke2012's model on Great Plains stream fishes.
 
-### Usage of dynamic occupancy models
+While most articles focus on a single taxa, 40% of the studies in our sample fit models to more than one taxa ('taxa' defined here as one or more species modelled as one).
+Of these, 75% fit independent DOMs to multiple species, and 25% fit explicitly multi-species implementations.
+These multi-species models varied in structure: some use species-specific effects to describe metapopulations by including hundreds of species in a single model [@dorazio2010; @hendershot2020], while others fit models which explicly account for species interactions and estimate conditional occupancy, colonisation, extinction, and detection [@lesmeister2015; @fidino2019].
+While they did not fit multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012].
+Working with large numbers of taxa does raise additional questions, as the level of complexity which can realistically be applied to each individual taxa is likely to be reduced for practicality's sake.
+With multi-species community frameworks, environmental covariates must be relatively general; and when fitting large numbers of independent models covariate selection processes are likely to be more restricted due to computational limitations.
 
-Our review sample displays considerable diversity in implementations of dynamic occupancy models with widespread representation from mammals, birds, and herptiles.
-This assemblage contrasts with the related stochastic patch occupancy models (SPOMs, @gutiérrez-arellano), where the majority of models are fit to invertebrates, plants, and fungi.
-This may reflect SPOMs lack of consideration for imperfect detection and assumption of discrete habitat patches, limitations which are unlikely to be fulfilled by most vertebrate taxa.
+### Data collection and inputs
 
-DOMs also display great variation in scale, from local studies with a handful of sites to continental-scale studies spanning decades.
+DOMs show exceptional diversity in the scales at which they can be implemented.
+The smallest study area in our sample was less than one kilometre squared, and the largest over one million.
+Study duration shows similar variation — data collection ran for for less than a month in the shortest study, and for 40 years in the longest, with a median duration of 3.25 years.
+This reiterates the flexibility of dynamic occupancy models, with explorations of hyper-local patterns of occupancy and analyses of continental-scale distribution using the same underlying model structure.
+The detection data collected from these surveys came from a variety of detection methods: 69% of articles used human observations, 20% conducted live trapping, 12% used camera traps, and 1% used bioacoustic monitors (@fig-StudyDetails).
 
-Several types of detection data have been successfully incorporated DOMs.
-This flexibility may be one of the more notable aspects of dynamic occupancy models; while the heirarchical data inputs may be more complex than those required by other models of species distributions, authors have been able to successfully manipulate datasets from many sources to meet requirements.
+The flexibility of DOM inputs data allows for creative uses of detection data from a range of sources.
+@marescot2020 models a relationship between wildlife and human poachers, using ranger-reported signs of poacher activity to create detection histories.
+In a rare example of a DOM applied to a marine species, @pendleton2022 uses aerial transects broken up into grid cells to study the occupancy dynamics of whales in the Northeastern United States.
 
-### Practices in the modelling process
+Notably, many reviewed articles use data which was not originally collected in a robust design framework for occupancy modelling.
+In these articles, authors manipulated their data into a hierarchical format post-hoc using a variety of methods.
+Some defined primary seasons as a discrete time interval, treating all surveys occurring within the season as secondary occasions.
+Others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate.
+These manipulations permit authors to use data which far predate the DOM, with one study using surveys from as far back as 1908 to model century-long changes in occupancy [@riddell2021].
 
-Model selection approaches diverge between Bayesian and frequentist implementations of the model.
-While model selection will naturally look different between these two categories, it is unclear what effect a lack of model selection may have on the outputs produced by Bayesian models and it is likely that Bayesian model selection approaches should be further explored @hooten2015.
+Additionally, not all articles rely on a single source of detection data - some integrate multiple sources of data to maximise sample size, combining data from camera traps, sign surveys, and citizen science reports.
+These integrated detection method models do require additional care in estimating detection probability, which is generally expected to vary based on observation method; however, this can be reasonably achieved by including survey method as a covariate on detection probability [@warrier2020; @pitman2017].
+A special case exists when different detection methods are used where one has the potential for false positive detections; e.g., where less-certain citizen science detections are combined with certain detections from field surveys.
+In this context, the certain detections are used to help account for false-positive detection probability, as in @miller2011's study integrating GPS collars and hunter reports to estimate wolf occupancy in Montana.
 
-Notably, the popular sequential approach to model selection (where one sub-model is assessed at a time) may not provide optimal results as shown in @morin2020.
-Considerably more research is required on this topic to clarify how authors should best approach the model building process.
+It is important to note that the data input requirements for dynamic occupancy models are not as arduous as may be commonly perceived.
+While it *is* necessary to record multiple observations at individual sites, this does not necessarily require that data be collected according to a strict, preconceived sampling scheme.
+Post-hoc manipulation of other data such as long-term citizen monitoring programs [@zuckerberg2011; @peach2019], disparate agency modelling programs [@mcgowan2020], or camera trapping data [@davis2018] is feasible with careful consideration.
+In one example, @riddell2021 used data from as long ago as 1908 by treating grid cells as sites and counting historical surveys within each cell as repeat observations.
+In a similar vein, all detection data does not necessarily need to come from the same source — detections from multiple methods, with associated variation in detection probabilities, may be incorporated within a single model
 
-### Key considerations for model users
+Such broadness and flexibility in data inputs is not amenable to a one-size-fits-all definition of occupancy.
+Authors must carefully consider precisely what they are modelling and address questions on the 'scale' represented by their model [@chave2013].
+Drivers of occupancy may differ depending on whether a site is represented by a single point on the landscape or as a grid cell, with the former likely to depend on more local, small scale factors rather than landscape-level trends [@stevens2019].
+This is also true for the *temporal* scale of occupancy: whether a site is occupied within a week or within a year is a non-arbitrary distinction representing vastly different conceptions of occupancy.
+This discussion is particularly pertinent in cases where the selection of season length *is* to some extent arbitrary, as with camera-trap or bioacoustic data where continuous recordings can be broken down into distinct 'seasons' of any length.
+DOMs are well suited to these data types [@balantic2019], and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data.
+For example, @kleiven2020 and @mölle2022 divide their camera trap records into seasons of just a few days.
+While this provides exciting insights of occupancy at extremely fine temporal scale, further research is needed on how to determine appropriate season and survey durations with respect to research questions.
 
-Implementation of dynamic occupancy models requires considerable thought to ensure that model outputs correspond with author expectations and ecological realities.
-Based upon our review of the existing literature we discuss key questions authors should carefully engage with during the modelling process.
+### Objectives and applications
 
-1.
-**Are seasons ecologically meaningful, and how do they affect the definition of occupancy?** As previously discussed, the duration of the season has important implications for both the reliability and proper interpretation of model results.
-Seasons which are not ecologically meaningful may run afoul of the closure assumption resulting in associated biases.
-Also relevant is how the definition of 'occupancy' may shift with season length and life-history of a species — for many species, occupancy within a year and within a week may be considerably different concepts, with shorter intervals more closely relating to the idea of 'use.' With the increasing proliferation of continuous detection methods such as camera traps and acoustic monitors where the definition of 'season' is to some degree arbitrary, it is increasingly important to thoughtfully consider how this element of the DOM is defined.
+Authors have used the DOM in many ways to achieve their research aims.
+The most frequent objective for studies using DOMs was hypothesis testing, with 50% of authors expressing interest in a specific relationship.
+31% were interested in identifying nondescript drivers of occupancy, 30% in estimating models parameters like occupancy themselves, and 29% worked to develop and expand on methods for DOMs.
+Just 10% of authors expressed interest in making predictions to unsurveyed locations or into the future.
+The DOMs ability to make predictions has been often underexploited in the existing literature, given its noted potential for forecasting occupancy in non-equilibrium scenarios [@kéry2013; @briscoe2021].
+Articles which *do* make predictions provide some of the most directly applicable outputs found in the review sample.
+As examples, @mcgowan2020 provides projections for a threatened species under multiple putative management scenarios, and @pollentier2021 generates maps of distributions resembling those made with species distribution models (SDMs).
 
-2.
+The authors of the studies captured in our sample come from a variety of backgrounds.
+86% of articles had at least one author based at an academic institution, 46% had author(s) at government institutions and 25% had author(s) from non-profit organisations.
+The high proportion of articles with government and NGO participants suggests an applied focus for many users of DOMs.
+This has important implications: when a model is used to assess a critically endangered species [@carvalho2023], guide public health management of a disease vector [@mores2020], or track rapidly-developing biological invasions [@wood2020], it is imperative that models can be trusted to accurately represent a system.
+Regardless of the reasons why they turn to the DOM to analyse their data, authors must navigate the model building process.
+
+# Practices in implementing DOMs
+
+Approaches to building any type of model will necessarily depend on the possibilities of the data at hand and on the priorities of the model-builder.
+This precludes any prescription of the 'best' way to build a model; however, there may still be commonalities in the modelling workflow.
+The modelling process for both DOMs and SDMs can be divided into a few components.
+First, based on prior knowledge and ecological theory a suite of covariates with hypothetical relationships with model parameters is assembled.
+Next, some form of model selection procedure is followed to identify either a single optimal model, or an ensemble of suitable models to be used to make conclusions.
+Following this, a model's performance should be somehow assessed so that the validity of conclusions can be adequately judged.
+In this section we examine how DOM users have approaches each of these stages, and make comparisons to common practice in the more extensive SDM literature.
+
+### Model complexity
+
+'Complexity' is a broad term which encompasses many aspects of a model [@merow2014].
+Opinions on simple versus complex models can be divided - some advocate for the simplest possible models, arguing that they are most generaliseable; others insist that overly-simple models cannot adequately represent the most important drivers in a system [@evans2013; @lonergan2014].
+To account for imperfect detection and describe change in occupancy, DOMs must be somewhat more complex relative to static correlative occupancy models.
+This added complexity helps mitigate the bias introduced when important elements like imperfect detection are ignored.
+Within DOMs, however, further complexity is to some degree up to the modeller: they can choose how many covariates to consider for inclusion on the various parameters, and how to represent the nature of the relationship between those covariates and parameters.
+
+The variation in the quality of covariates considered for each model parameter amongst our reviewed articles is shown in @fig-covariates.
+Notably, many studies expressed some parameters as a constant without including any covariates: of the four standard model parameters initial occupancy has the most articles considering zero covariates (35%) and detection probability the least (19%).
+A lack of any covariates on the initial occupancy parameter raises some questions.
+Presumably any taxa will have some amount of non-random variation in how it is distributed across a landscape; by expressing initial occupancy as a constant this heterogeneity is not captured.
+Omission of factors which drive occupancy can introduce bias into SDMs [@barry2006].
+Initial occupancy is itself effectively a static correlative SDM, and the effects of poorly estimating this parameter on predicted occupancy in ensuing seasons is uncertain but unlikely to be favourable.
+
+![Distribution of covariate quantity considered for each model parameter.](Figures/CovParamPlot.jpeg){#fig-covariates}
+
+The types of covariates which were considered for modelling were also wide-ranging, and a summary of their inclusions in articles is presented in @tbl-covariates.
+The most frequently used covariates described aspects of habitat, with 57% of articles considering these terms for at least one parameter.
+Data for covariates may be either collected directly, requiring measurement at individual sites, or indirectly via remote sensing — where directly measured covariates may offer more direct insights, indirect measurements are far more easily projected to unsurveyed sites when making predictions [@austin2002].
+Some covariates, like biotic interactions, must nearly always be measured at sites, while others such as climate and weather covariates are generally indirectly observed.
+There is variation in which covariates are most likely to be dynamic and change through time.
+
+::: column-page-inset
+```{r covariate table}
+#| echo: false
+#| message: false
+#| label: tbl-covariates
+#| tbl-cap: "All covariates considered for inclusion in a study were classified into mutually exclusive categories.
+For each parameter (initial occupancy, colonisation, extinction, and detection), the percentage of studies including at least one covariate in a given category is presented.
+We also present the percentage of covariates in each category which are dynamic (change throughout seasons) or require direct observation (e.g.
+are not remotely sensed).
+The percentage of covariates in each category for which non-linear representations or interactions with other covariates were considered is given to indicate which were most likely to be included as more complex terms."
+
+library(tidyverse)
+library(gt)
+
+covTable <- read_csv("Figures/CovariateTable.csv")
+
+covTable |>
+  arrange(Category, -Prop_Any) |>
+  gt(rowname_col = "Covariate", groupname_col = "Category") |>
+  tab_options(
+    row_group.as_column = TRUE
+  ) |>
+  tab_row_group(
+    label = md("Environmental"),
+    rows = Category == "E",
+    id = "Env"
+  ) |>
+  tab_row_group(
+    label = md("Structural"),
+    rows = Category == "S",
+    id = "Struct"
+  ) |>
+  tab_row_group(
+    label = md("All"),
+    rows = Category == "O",
+    id = "All"
+  ) |>
+  tab_style(
+    style = cell_fill(color = "#82AD8E"),
+    locations = cells_row_groups(groups = "Env")
+  ) |>
+  tab_style(
+    style = list(
+      cell_fill(color = "plum4")
+    ),
+    locations = cells_row_groups(groups = "Struct")
+  ) |>
+  tab_style(
+    style = list(
+      cell_fill(color = "gray60")
+    ),
+    locations = cells_row_groups(groups = "All")
+  ) |>
+  row_group_order(groups = c("Env", "Struct", "All")) |>
+  tab_spanner(label = "Articles with covariate on:",
+              columns = c("Prop_Any", "Initial occupancy", "Occupancy",
+                          "Colonisation", "Extinction_Persistence",
+                          "Detection"),
+              id = "inclusions") |>
+  tab_spanner(label = "Covariates which are:",
+              columns = c(Prop_Dynamic, Prop_Direct),
+              id = "covariates") |>
+  tab_spanner(label = "Articles representing with:",
+              columns = c(Prop_Nonlinear, Prop_Interact),
+              id = "representations") |>
+   cols_label(
+    Prop_Any = "Any parameter",
+    `Initial occupancy` = html("&psi;1"),
+    Occupancy = html("&psi;"),
+    Colonisation = html("&gamma;"),
+    Extinction_Persistence = html("&epsilon;"),
+    Detection = html("&rho;"),
+    Prop_Dynamic = "Dynamic",
+    Prop_Direct = "Directly observed",
+    Prop_Nonlinear = "Non-linear",
+    Prop_Interact = "Interaction"
+  ) |>
+  text_case_match(
+    "Environmental" ~ md("*Any Environmental*"),
+    "HABT" ~ "Habitat type",
+    .locations = cells_stub()
+  ) |>
+  cols_hide(Category) |>
+  fmt_percent(decimals = 0) |>
+  sub_missing(missing_text = "-") |>
+  as_raw_html()
+
+```
+:::
+
+For the vast majority of covariates authors considered only linear relationships with model parameters.
+Polynomial responses are rare — even in the categories where they are most frequently used, the rate of consideration for non-linear responses does not exceed 16%.
+This is a notable departure from SDMs; while GLM-based methods with linear responses are used, many common methods used to fit SDMs like MAXENT or boosted regression trees typically permit flexibility in response curves between covariates and parameters [@merow2013; @elith2008].
+Non-linear relationships are the norm, not the exception in nature, and in many cases a linear response will not adequately capture a covariate's influence on model parameters [@austin2002].
+Interactions between covariates were also rarely incorporate in our sample, and their rate of inclusion does not exceed 34% for any category.
+While they should not necessarily be used liberally, where these interactions exist their exclusion can influence model performance [@guisan2006].
+
+Any model must balance simplicity and realism, but it is important that the ecological relationships quantitatively expressed in a model be grounded in ecological theory and accurately represent the modeller's hypothesised response [@austin2002; @austin2007].
+Complexity in the quantity and nature of covariate responses is also controlled during model selection, but these procedures can only work with the information provided to them.
+Adequately considering the plausible drivers of occupancy and detection by identifying quality candidate covariates representing those drivers is an important step in developing robust models that appropriately account for heterogeneity in these parameters.
+
+### Covariate selection
+
+Variation in approaches to modelling is particularly apparent in the model selection process, where there is little consensus on how to determine which covariates to include in final models (@fig-selection).
+In the reviewed frequentist implementations, a range of approaches have been used for this purpose including comparison suites of candidate models defined *a priori* and various stepwise selection protocols.
+Bayesian implementations were far less likely to use model selection procedures — 70% of these fit only a single model defined *a priori,* relative to 10% amongst the frequentist models.
+
+![Frequency of use for model selection methods.
+'Single model' indicates no model selection occurred, 'candidate suite' includes studies which compared a predefined set of models, articles with 'procedural selection' fit parameters one-by-one, 'exhaustive selection' fit all possible combinations of models, and 'simple precursor' describes where results of a less-complex model were used to inform covariate choice.](Figures/SelectionFigure.jpeg){#fig-selection}
+
+Model selection is a particularly challenging aspect of fitting DOMs, with substantial uncertainties in the best ways to conduct the process.
+Limited research has been conducted on the topic, but what has been done has indicated that choices in this process can influence the quality of model outputs [@morin2020].
+This is reflected in research from outside the DOM literature, where stepwise protocols have been found to be unreliable [@olden2000].
+However, others have found that different model selection methods can end up arriving at roughly the same answer [@maggini2006].
+Model selection can be particularly difficult for DOMs given the increased complexity of the underlying model structure, which will often preclude exhaustive selection from amongst all possibilities.
+
+Bayesian model selection is an effectively distinct question in DOMs as the framework is not well suited to the more conventional model selection methods.
+@hooten2015 provides a guide to some of the possibilities for optimising Bayesian models, including the use of regularisation priors.
+
+Some reviewed studies included substantially more complex model selection protocols.
+
+### Model validation
+
+Regardless of whether a model was implemented under the Bayesian or frequentist frameworks, evaluation of model fit or performance was rare.
+Only 8% of articles directly addressed either aspect of modelling, be it via goodness-of-fit tests or validation with either internal or external data.
+This is a striking departure from the SDM literature, where roughly half of studies sufficiently assess their models performance [@araújo2019].
+While the appropriate method of model evaluation may depend on data availability and research objectives, assessing models by some method is always important to ensure reliability of model outputs [@araújo2005; @guisan2005].
+Most models in our sample conduct model selection using AIC as a scoring criteria, and often using model averages.
+While these approaches provide *relative* models of fit, they cannot be considered evaluation themselves - as @mackenzie2004 write, "the selection of a 'best' model(s) does not guarantee the selection of a "good" model.
+
+The first goodness-of-fit (GOF) test for occupancy model was given in @mackenzie2004, implemented via a parametric bootstrap.
+There is no GOF test available for DOM, however, common packages including 'AICcModAvg' and 'unmarked' offer parametric bootstraps derived from @mackenzie2004's test which have been used by some authors in our sample.
+While these tests may not suit Bayesian implementations, @broms2016 discusses possibilities for model evaluation in that context - while their article focuses on single-season multi-species occupancy models, extensions of their approach may suit the dynamic multi-season model as well.
+
+All aspects of modelling varied with the stated objectives of articles (@tbl-objectives).
+Those focussed on making predictions and identifying drivers of occupancy were more likely to consider a larger number of covariates and represented covariates in more complex ways by considering non-linear responses and interactions between covariates.
+Articles with objectives involving methods development considered fewer covariates and were less likely to use these more complex formulations.
+
+<div>
+
+```{r Objectives table}
+#| echo: false
+#| message: false
+#| label: tbl-objectives
+#| tbl-cap: "Implementation of models by study objective.
+Models were assessed based on the number of covariates considered, the inclusion of non-linear responses and interactions between covaraites, the manner of model selection, and whether or not they objectively evaluated model performance."
+
+objTable <- read_csv("Figures/ObjectiveTable.csv")
+
+objTable |>
+  mutate(Objective = fct(Objective,
+                         levels = c("Hypothesis", "Drivers", "Parameters",
+                                    "Methods", "Predict", "Any"))) |>
+  arrange(Objective) |>
+  mutate(Objective = case_when(Objective == "Hypothesis" ~ "Testing hypotheses",
+                               Objective == "Drivers" ~ "Identifying drivers",
+                               Objective == "Predict" ~ "Making predictions",
+                               Objective == "Methods" ~ "Developing methods",
+                               Objective == "Parameters" ~ "Estimating parameters",
+                               Objective == "Any" ~ "Any objective")) |>
+  gt(rowname_col = "Objective") |>
+  tab_spanner(label = "Articles",
+              columns = c("Articles", "ArticlesProp"),
+              id = "articles") |>
+  tab_spanner(label = "Percent which have:",
+              columns = c("ProbPoly", "ProbInt",
+                          "ProbEval", "ProbSelect"),
+              id = "percents") |>
+  cols_label(Articles = "n", ArticlesProp = "%",
+             MedianCovs = "Median covariates considered",
+             ProbPoly = "Non-linear responses",
+             ProbInt = "Covariate interactions",
+             ProbSelect = "Model selection",
+             ProbEval = "Model evaluation / goodness of fit") |>
+  fmt_percent(columns = c(ArticlesProp, 
+                          ProbPoly, ProbInt, 
+                          ProbEval, ProbSelect),
+              decimals = 0) |>
+  tab_style(style = list(cell_fill(color = "gray80"),
+                         cell_text(style = "italic")),
+            locations = cells_body(rows = Objective == "Any objective")) |>
+  tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = MedianCovs,
+                                   rows = MedianCovs == max(MedianCovs, na.rm = TRUE))) |>
+  tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = ProbPoly,
+                                   rows = ProbPoly == max(ProbPoly, na.rm = TRUE))) |>
+  tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = ProbInt,
+                                   rows = ProbInt == max(ProbInt, na.rm = TRUE))) |>
+  tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = ProbEval,
+                                   rows = ProbEval == max(ProbEval, na.rm = TRUE))) |>
+  tab_style(style = cell_fill(color = "lavender"),
+            locations = cells_body(columns = ProbSelect,
+                                   rows = ProbSelect == max(ProbSelect, na.rm = TRUE))) |>
+  as_raw_html()
+```
+
+</div>
+
+# Conclusions and key considerations
+
+Our review highlights the dynamic occupancy model's exceptional flexibility and suitability to addressing a range of ecological questions.
+The DOM may be a more achievable option than many think with feasible options to incorporate much existing presence/absence data from a variety of sources.
+Fertile areas for future research remain, with methods and guidance for making predictions and for modelling autonomous detection data from camera-traps and acoustic monitors particularly promising opportunities.
+In light of this potential and the DOMs status as a popular tool for applied ecology, it is increasingly important that best-practice standards be established for building these models.
+Major questions remain on fundamental questions in model building — there is need for further research on the appropriate levels of complexity for DOMs in various scenarios, on the methods for selecting models from amongst candidates, and on the best ways to evaluate the quality of the models chosen.
+In these areas, DOMs lag well behind SDMs where these standards have been more seriously considered [@araújo2006; @araújo2019; @zurell2020].
+Given the parallels between these model classes, we may be able to extrapolate some lessons from SDMs to support building better DOMs.
+In light of these uncertainties, and based upon existing research on SDMs, we discuss some key considerations for authors seeking to use DOMs in their work:
+
+I.
 **Are key drivers of occupancy and detection incorporated?** While it is effectively impossible to *fully* describe heterogeneity across a landscape, authors should make their best efforts to capture the drivers of occupancy and detection most likely to influence their study species.
@@
 
-3.
-**Ensure covariates reflect ecological complexity.** Many ecological relationships are non-linear and may require polynomials terms to be accurately represented.
-This is particularly true at large spatial scales where a larger portion of a species ecological niche is shown in surveyed data.
-Authors should question whether a linear term is sufficient to capture hypothetical relationships in their study systems and consider quadratic terms where appropriate.
+II.
+**Do covariates relationships reflect ecological complexity?** Many ecological relationships are non-linear and may require polynomial terms to be accurately represented.
+This is particularly true at large spatial scales where a larger portion of a species ecological niche is represented in surveyed data.
+Authors should question whether a linear term is sufficient to capture hypothetical relationships in their study systems, and consider representing species responses as polynomials terms where appropriate.
+Interactions between covariates should also be explored where a plausible ecological relationship between factors exists.
 
-4.
-**Consider model evaluation.** Model evaluation remains rare in ecological research relative to other fields embracing complex modelling, likely due to a reluctance to 'throw away' data for use in testing.
-This concept should be challenged, particularly for larger scale long-term projects where reserving a fraction of data for testing is unlikely to substantially reduce precision of parameter estimates.
-For studies where this is genuinely not possible, reporting results of goodness-of-fit tests may still provide valuable information to readers.
+III.
+**How should model evaluation be conducted?** Regardless of why a model is used, it is important to assess its quality and its ability to support the conclusions drawn from it.
+While more research is needed on the most robust ways to assess models, existing methods should be used and reported in published studies.
+Dependent on data availability and model objectives, this can be achieved through bootstrapped goodness-of-fit tests or metrics of predictive performance on reserved data.
+This is an important step in establishing best-practices for the field, and should realistically be conducted in some form for the vast majority of studies.
 
-### Priorities for future research
-
-Work on dynamic occupancy modelling over the past 20 years has developed this model class into an important tool for describing species distributions and ecological relationships.
-Historically, methods-development research for dynamic occupancy models has focused on questions of model structure — namely ideas around the closure assumption and imperfect detection.
-This emphasis likely reflects the DOM's roots in the mark-recapture literature, where these topics have been discussed for several decades.
-Considerably less emphasis has been placed on the relation of model parameters to environmental factors via covariates and the model selection processes used to do so.
-This stands in stark contrast to related correlative species distribution models, where these topics have been the subject of considerable research over the years.
-There are opportunities for the DOM literature to learn from species distribution models, and more targeted research on the influence of the model selection process is necessary.
-
-The use of DOMs in prediction has remained fairly limited despite significant potential for this use-case.
-Papers in our sample which have taken advantage of this potential included @mcgowan2020 who used DOMs to make predictions of black rail distributions under various management scenarios, and @saunders2020, who validated predictions from ecological niche models using continent-scale DOMs.
-Further implementations of these models may benefit from the aforementioned necessary research on model selection to ensure robust outputs.
-
-A final opportunity for future research relates to autonomous detection methods, such as camera traps and bio-acoustic monitors — these technologies have permitted collection of enormous quantities of data particularly well-suited to dynamic occupancy models, but have also raised additional questions.
-The definition of seasons is of increased importance for these data types, where division of data into 'seasons' and 'observations' is arbitrary.
-Further research should explore the implications of this phenomenon and additional modelling opportunities made possible by this novel data type.
-
-The past two decades have shown that dynamic occupancy models are an excellent tool for addressing a wide range of ecological questions relating to species occupancy throughout space and time.
-The flexibility afforded by model extensions and the considerable advantages of the class over other species distribution models is likely to ensure that these models continue to increase in popularity.
-Given their strong use on threatened taxa and excellent potential to support environemental management, it is important that research continues to support model users in creating the most reliable models possible.
+As previously stated, DOMs and their outputs are important — in a biodiversity and conservation context where quantitative representations of complex natural systems are used to guide policy and decision making, it is necessary to ensure models can be trusted.
+We hope that this review can open a broader discussion on how to best approach dynamic occupancy modelling as a first step towards establishing best practices in the discipline.

### 2024-08-25 Analysis/ReviewFigures/Final_Figures.qmd
Big changes from ESA travel period
Added approx 30 articles, re-wrote manuscript draft

@@
 library(tidyverse)
+library(patchwork)
 library(sf)
@@
 focalTaxaPlot <- sheet_FocalTaxa |>
-  mutate(Status = case_when(Status %in% c("CR", "EN", "NT", "Threatened_Other") ~ "Threatened",
-                            Status == "NA" ~ "Multispecies",
+  mutate(Status = case_when(Status %in% c("CR", "EN", "NT", "VU", "Threatened_Other") ~ "Threatened",
+                            Status == "NA" ~ "Multiple species",
                             Status == "Invasive" ~ "Invasive",
                             Status == "LC" ~ "Stable")) |>
-  mutate(Status = fct(Status, levels = c("Multispecies", "Invasive",
+  mutate(Status = fct(Status, levels = c("Multiple species", "Invasive",
                                          "Stable", "Threatened"))) |>
@@
   select(Articles, Taxa = `Taxa keywords`, Status) |>
-  mutate(Taxa = fct_reorder(Taxa, Articles)) |>
+  mutate(Taxa = case_when(Taxa == "NA" ~ "Other", .default = Taxa)) |>
+  mutate(Taxa = fct(Taxa, levels = rev(c("Bird", "Mammal", "Herptile",
+                                     "Invertebrate", "Other")))) |>
 
@@
 ```{r Article coverage}
-coveragePlot <- randomisationSheets |>
-  summarise(N = n(), .by = c(Strata, Included)) |>
-  pivot_wider(names_from = Included, values_from = N) |>
-  mutate(HitRate = YES/(NO+YES)) |>
+coveragePlot <- 
+  randomisationSheets |>
+  summarise(N = n(), .by = c(Year, Strata, Included)) |>
+  pivot_wider(names_from = Included, values_from = N, values_fill = 0) |>
+  mutate(HitRate = sum(YES)/(sum(NO)+sum(YES)),
+         .by = Strata) |>
   mutate(EstRemaining = `NA`*HitRate) |>
@@
   ggplot() +
-  geom_col(aes(x = Strata, y = Articles, fill = Group), colour = "white") +
+  geom_col(aes(x = Year, y = Articles, fill = Group), colour = "white") +
   scale_fill_manual("",
@@
         axis.title = element_text(size = 12), legend.text = element_text(size = 12))
-
+coveragePlot
 ggsave(plot = coveragePlot, "Manuscript/Figures/CoveragePlot.jpeg",
@@
   geom_sf(data = countries_shp, colour = "gray90", fill = "gray80") +
-  geom_sf(shape = 21,fill = "aquamarine4", colour = "white", alpha = 0.7, size = 4) +
+  geom_sf(shape = 21,fill = "aquamarine4", colour = "white", alpha = 0.7, size = 2.5) +
   theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
         panel.background = element_rect(fill='transparent', color=NA),
-        plot.background = element_rect(fill='transparent', color=NA)) 
+        plot.background = element_rect(fill='transparent', color=NA),
+        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 
 studyArea_Map
@@
                                    angle = 90,
-                                   vjust = 1, hjust = 1),
-        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
+                                   vjust = 0.5, hjust = 1),
+        axis.text.y = element_blank(), axis.ticks = element_blank(),
         panel.background = element_rect(fill='transparent', color=NA),
@@
 siteQuantityPlot <- sheet_DataCollection |>
+  mutate(`Site quantity` = as.numeric(`Site quantity`)) |>
   ggplot() +
@@
                  binwidth = 0.25, fill = "aquamarine4", colour = "white") +
-  geom_hline(aes(yintercept = median(`Site quantity`)), 
+  geom_hline(aes(yintercept = median(`Site quantity`, na.rm = TRUE)), 
              colour = "goldenrod3", linetype = 2, linewidth = 2) +
@@
 sheet_DataCollection |>
+  mutate(`Site quantity` = as.numeric(`Site quantity`)) |>
   left_join(select(sheet_ArticleData, c(`Review ID`, `Year Strata`)), by = "Review ID") |>
@@
   mutate(studyScale = `Site quantity`*`Primary occasions`) |>
-  mutate(studyScale = case_when(studyScale > 5000 ~ 5000,
+  mutate(studyScale = case_when(studyScale > 10000 ~ 10000,
                                 .default = studyScale)) |>
-  mutate(outlier = case_when(studyScale >= 5000 ~ TRUE,
+  mutate(outlier = case_when(studyScale >= 10000 ~ TRUE,
                              .default = FALSE)) |>
@@
 ```{r Objectives through time}
-sheet_Objectives |>
+objectiveTrendPlot <- sheet_Objectives |>
   left_join(select(sheet_ArticleData, c(`Review ID`, `Year Strata`)), by = "Review ID") |>
@@
                names_to = "Objective", values_to = "ObjectiveIncluded") |>
-  filter(ObjectiveIncluded == "YES") |>
+  mutate(Objective = str_replace(Objective, "\\?", "")) |>
+  filter(ObjectiveIncluded == "YES") %>%
+  rbind(., mutate(., `Year Strata` = "All Years")) |>
   mutate(Articles = length(unique(`Review ID`)), .by = `Year Strata`) |>
   summarise(ObjArticles = n(), .by = c(Objective, Articles, `Year Strata`)) |>
-  mutate(PropArticles = ObjArticles/Articles) |>
+  mutate(PropArticles = (ObjArticles/Articles)*100) |>
   mutate(Objective = fct_reorder(Objective, PropArticles)) |>
   complete(Objective, `Year Strata`, fill = list(PropArticles = 0)) |>
-  mutate(Objective = factor(Objective, levels = c("Demonstrate methods?",
-                                               "Observe trends?", 
-                                               "Identify drivers?", 
-                                               "Test relationships?",
-                                               "Predict spatially?",
-                                               "Predict temporally?"))) |>
+  mutate(Strata = case_when(`Year Strata` == "All Years" ~ "left",
+                            .default = "right")) |>
+  mutate(Strata = fct(Strata, levels = c("left", "right"))) |>
+  mutate(Objective = factor(Objective, levels = c("Demonstrate methods",
+                                                  "Observe trends", 
+                                                  "Identify drivers", 
+                                                  "Test relationships",
+                                                  "Predict spatially",
+                                                  "Predict temporally"))) |>
+  mutate(`Year Strata` = fct(`Year Strata`, levels = c("2004-2007",
+                                                       "2008-2011",
+                                                       "2012-2015", 
+                                                       "2016-2019",
+                                                       "2020-2023",
+                                                       "All Years"))) |>
 
@@
            position = "dodge", colour = "white") + 
-  scale_fill_manual(values = c("plum4", "cadetblue3", "aquamarine3",
+  scale_fill_manual("",
+                    values = c("plum4", "cadetblue3", "aquamarine3",
                                "aquamarine4", "goldenrod2", "goldenrod3")) +
-  labs(y = "Proportion of studies") +
-  lims(y = c(0, 1)) +
+  scale_y_continuous(breaks = c(0, 20, 40, 60, 80),
+                     limits = c(0, 80),
+                     labels = c("0%", "20%", "40%", "60%", "80%")) +
+  labs(y = "Percentage of studies\nwith objective", x = "") +
   theme(panel.grid.major.x = element_blank(),
         panel.grid.minor.x = element_blank(),
-        legend.position = "top")
+        legend.text = element_text(size = 15),
+        axis.text = element_text(size = 15), axis.title = element_text(size = 15),
+        legend.position = "bottom", strip.background = element_blank(),
+        strip.text = element_blank()) +
+  facet_grid(cols = vars(Strata),
+             scales = "free_x", space = "free_x", drop = TRUE)
+objectiveTrendPlot
+ggsave(plot = objectiveTrendPlot, "Manuscript/Figures/ObjectiveTrend.jpeg",
+       width = 8, height = 4)
+```
+
+Avg covariate count by objective
+
+```{r Objective covariate count}
+covsPerParam <- sheet_Covariates |>
+  separate_longer_delim(Covariates, delim = ", ") |>
+  mutate(n_models = length(unique(`Model ID`)), .by = `Review ID`) |>
+  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
+                                 "Colonisation", "Extinction_Persistence",
+                                 "Detection")) |>
+  mutate(n_params = length(unique(`Parameter type`)), .by = `Review ID`) |>
+  filter(Covariates != "None") |>
+  summarise(CovsPerParam = n()/(max(n_params)*max(n_models)), .by = `Review ID`)
+
+objectiveCovariatePlot <- sheet_Objectives |>
+  pivot_longer(c("Demonstrate methods?", "Observe trends?", 
+                 "Identify drivers?", "Test relationships?",
+                 "Predict spatially?", "Predict temporally?"),
+               names_to = "Objective", values_to = "ObjectiveIncluded") |>
+  mutate(Objective = str_replace(Objective, "\\?", "")) |>
+  filter(ObjectiveIncluded == "YES") %>%
+  rbind(., mutate(., Objective = "Any Objective")) |>
+  select(`Review ID`, Objective) |>
+  distinct() |>
+  left_join(covsPerParam, by = join_by(`Review ID`)) |>
+
+  mutate(Objective = factor(Objective, levels = c("Any Objective", "Demonstrate methods",
+                                                  "Observe trends", 
+                                                  "Identify drivers", 
+                                                  "Test relationships",
+                                                  "Predict spatially",
+                                                  "Predict temporally"))) |>
+  mutate(Objective = fct_relabel(Objective, 
+                                 ~ str_replace(.x, " ", "\n"))) |>
+  mutate(facet = case_when(Objective == "Any\nObjective" ~ "left",
+                           .default = "right")) |>
+  mutate(CovsPerParam = case_when(CovsPerParam > 10 ~ 10,
+                                  .default = CovsPerParam)) |>
+
+  ggplot(aes(x = Objective, y = CovsPerParam, 
+             fill = Objective, colour = Objective)) +
+  geom_boxplot(colour = "gray50", alpha = 0.6) +
+  geom_jitter(alpha = 0.6, width = 0.1) +
+  scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10),
+                     labels = c("0", "2", "4", "6", "8", "10+")) +
+  scale_fill_manual(limits = c("Any\nObjective", "Demonstrate\nmethods", 
+                               "Observe\ntrends", "Identify\ndrivers", 
+                               "Test\nrelationships", "Predict\nspatially",
+                               "Predict\ntemporally"),
+                    values = c("gray60", "plum4", "cadetblue3", "aquamarine3",
+                               "aquamarine4", "goldenrod2", "goldenrod3")) +
+  scale_colour_manual(limits = c("Any\nObjective", "Demonstrate\nmethods", 
+                                 "Observe\ntrends", "Identify\ndrivers", 
+                                 "Test\nrelationships", "Predict\nspatially",
+                                 "Predict\ntemporally"),
+                      values = c("gray60", "plum4", "cadetblue3", "aquamarine3",
+                                 "aquamarine4", "goldenrod2", "goldenrod3")) +
+  labs(x = "", y = "Number of covariates\nconsidered per parameter") +
+  facet_grid(cols = vars(facet), scales = "free_x", space = "free_x") +
+  theme(legend.position = "none",
+        panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(),
+        legend.text = element_text(size = 15), axis.text = element_text(size = 13),
+        axis.title = element_text(size = 15), strip.background = element_blank(),
+        strip.text = element_blank()) 
+objectiveCovariatePlot
+ggsave(plot = objectiveCovariatePlot, "Manuscript/Figures/ObjectiveCovariates.jpeg",
+       width = 8, height = 4)
+```
+
+Combined plot
+
+```{r combined objective plot}
+stackedObjectivePlot <- objectiveTrendPlot / objectiveCovariatePlot +
+   plot_annotation(tag_levels = 'A')
+stackedObjectivePlot
+
+ggsave(plot = stackedObjectivePlot, "Manuscript/Figures/ObjectiveFigure.jpeg",
+       width = 12, height = 8)
 ```
@@
             Sequential = max(Sequential), Simple = max(Simple),
+            Candidate = max(`Candidate set`),
             .by = c(`Review ID`, `Model class`, n_Class, n_Overall)) |>
-  left_join(modelling_Covariates, by = join_by(`Review ID`))
+  left_join(modelling_Covariates, by = join_by(`Review ID`)) |>
+  filter(`Model class` != "Other")
 
@@
             Prop_Simple = sum(Simple)/mean(n_Class),
+            Prop_Candidate = sum(Candidate)/mean(n_Class),
             Prop_GOF = sum(GOF)/mean(n_Class),
@@
             Prop_Simple = sum(Simple)/mean(n_Overall),
+            Prop_Candidate = sum(Candidate)/mean(n_Overall),
             Prop_GOF = sum(GOF)/mean(n_Overall),
@@
 modelling_summary <- rbind(modelling_byClass, modelling_overall) |>
-  t() |> as.data.frame()
+  t() |> as.data.frame() %>%
+  mutate(Variable = row.names(.)) |>
+  set_names(c("Frequentist", "Bayesian", "All models", "Variable")) |>
+  filter(Variable != "Model class")
+
+write_csv(modelling_summary, "Manuscript/Figures/ModellingSummary.csv")
 ```

### 2024-08-25 Manuscript/domReview.qmd
Big changes from ESA travel period
Added approx 30 articles, re-wrote manuscript draft

--- 2024-04-17 Manuscript/domReview.qmd
@@
 abstract: | 
-  Dynamic occupancy models are an increasingly important tool for the analysis of wildlife occurence data due to their favourable balance of complexity and ease of implementation.
-Uses of this model class are diverse, including explorations of the drivers of occupancy as well as predictions of species distributions across landscapes and into the future.
-The application of these models does require careful navigation of the modelling process including important decisions on structure, covariate inclusion, and assessment of model performance.
+  Describing patterns of species occupancy across landscapes and throughout time is a key objective of much ecological research.
+Nonetheless, reliably estimating occupancy can be a challenge, particularly when common issues such as imperfect detection and shifting distributions are at play.
+The dynamic, multi-season occupancy model (DOM) offers a framework for occupancy estimation which explicitly accounts for observation error, while also incorporating additional mechanistic process by explicitly estimating colonisation and extinction rates.
+In the two decades since their introduction the DOM has become a popular tool for a popular tool for describing patterns in occupancy, exploring the environmental drivers of species occurence, and, more recently, for predicting species occupancy to new locations and to future conditions.
 
-  In this article we present a systematic review of implementations of the dynamic occupancy model, with particular emphasis on how authors have approached the modelling process.
-Our findings indicate that authors have found this model class suitable for a considerably diverse set of ecological questions and have successfully fit models to data collected under a wide range of conditions, highlighting the high degree of flexibility available.
-However, the manner of application appears to be quite variable: approaches to covariate selection vary considerably, and model complexity generally speaking is low.
-The implications of these decisions as they relate to model outputs are yet to be explored in detail and should be a priority for future research.
+  In this review we examine how ecologists have applied the DOM in its first two decades of use.
+Based on a sample of articles fitting DOMs to field ecological data, we explore the systems in which these models have been used, how data was collected for them, and how authors have built their models to achieve their objectives.
+Our findings indicate that the DOM is an particularly flexible tool readily applied to a range of diverse taxa from around the globe, capable of being used to estimate occupancy at study sites ranging from local to continental scales over time periods from months to decades.
+The DOM framework is also amenable to extension, further broadening their utility to address a wide range of ecological questions.
 
+  A key aim of this review is to explore how users of DOMs have incorporated covariates into their models to describe variation in occupancy through space and time.
+Broadly speaking, model complexity in DOMs tends to be low, with relatively few covariates considered for inclusion and most relationships represented as simple linear terms.
+Considerable variation also exists in the procedures used for covariate selection, and limited research has been conducted on how these choices may influence model performance and inference.
+Furthermore, only a fraction of articles appear to evaluate models (either by goodness-of-fit testing or prediction validation).
+Likewise, little guidance exists on how to approach this task in DOMs.
+These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.
+
+bibliography: references.bib
+csl: apa.csl
 embed-resources: true
-bibliography: references.bib
 ---
@@
 
-Since its introduction two decades ago by @mackenzie2003 the dynamic multi-season occupancy model has become an important tool for the analysis of species occurrence throughout space and time.
-In the ecological modelling landscape dynamic occupancy models (DOMs) occupy a favourable position between model complexity and ease of data collection.
-In explicitly estimating the processes which guide occupancy while accounting for imperfect detection, the model class offers more advanced insights than simpler static species distribution models; despite this, input data requirements remain achievable compared to other more complex process-explicit models.
+The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849].
+An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing can be applied in many ways.
+Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, and identifying potential drivers of occurrence can help inform potential management actions.
+Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios.
 
-The use-cases of this model class are wide-ranging, and overlap considerably with other models such as species distribution models [@elith2009] and stochastic patch occupancy models [@gutiérrez-arellano].
-Estimates of occupancy through time are independently valuable for many monitoring programs, and DOMs are a particularly favourable choice when species are not detected perfectly [@guillera-arroita2017].
-DOMs may also be used in testing specific hypotheses of relationships between occupancy and the environment, or in identifying the drivers of occupancy in a system.
-Although infrequently applied thus far, a promising application of DOMs is their use to make predictions to unsurveyed sites or into the future [@briscoe2021; @kéry2013a].
-In each of these use cases, authors must navigate the modelling process and make important decisions on aspects of model structure, the process of determining which covariates to incorporate, and the critical assessment of model outputs such as parameter estimates or predictions.
+While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate.
+The need to understand and describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly used to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales.
+However, several factors can make occupancy a difficult quantity to estimate.
+For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with most wildlife field data.
+Generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004].
+Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014].
+Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010].
+These conditions are prominent during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].
 
-This review seeks to provide an overview dynamic occupancy model's first two decades of use.
-We discuss the DOMs key features, limitations, and implementation process, before presenting a systematic review of articles implementing the model.
-Based on these results, we present key considerations for authors seeking to build these models in their work and outline priorities and opportunities for future research involving this flexible and important model class.
+The site occupancy models first introduced by @mackenzie2002 offer a powerful framework for modelling presence/absence data while accounting for each of these challenges.
+Originating in the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy.
+Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time.
+This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly incorporating the key processes which drive occupancy while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic data required by more process-explicit models [@briscoe2019].
+These features make the DOM an important tool, with uses for tasks including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].
 
-## Model form
+In this review we examine how dynamic occupancy models have been used by ecologists in the two decades since their inception.
+Following an introduction to the model's form and assumptions, we present the results of a systematic review exploring how researchers have applied the DOM to ecological data, with emphasis on how they collected their data, selected covariates, and evaluated their models.
+Based on these results we highlight the DOM's flexibility as a tool for understanding species occurrence, examine approaches to the model building process, and outline key priorities for future research involving this model class.
 
-Dynamic occupancy models unite two processes: the ecological process of site occupancy describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed given that a site is occupied.
-The basic model form as described by @mackenzie2003 is depicted in @fig-modelform.
-Under this parameterisation, sites exist in either occupied or unoccupied states and may transition between these states between each time-step.
-Occupancy state in the first time-step is given by the initial occupancy parameter, and in ensuing time-steps occupancy is described as a Markovian process governed by probabilities of colonisation and extinction.
-Given that a site is occupied, the probability of observing the species during a survey is given by the detection parameter.
-Under this formulation, it is assumed that false-positive detections do not occur.
+### Dynamic occupancy model form and assumptions
 
-![The original parameterisation of the dynamic occupancy model as described by @mackenzie2003.
-The ecological process of occupancy is guided by the initial occupancy, colonisation, and extinction parameters; the observational process is guided only by the detection parameter.](Figures/ModelForm.png){#fig-modelform width="1000"}
+The DOM unites two processes: the ecological process of site occupancy describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed given that a site is occupied (@fig-modelform).
+Under this parameterisation, sites exist in either an occupied or unoccupied state.
+In the first time-step, occupancy state at each site is determined by the probability of initial occupancy.
+In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation and extinction.
+In the observation component of the DOM, the model explicitly accounts for imperfect detection.
+At occupied sites, the detection probability describes whether or not the species is observed during a survey; under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur.
 
-To disentangle the ecological and observation processes and estimate the probability of detection, the DOM relies on a hierarchical sampling design based on Pollock's robust design [@pollock1982] and depicted in @fig-surveys.
-Under this design, observations at a site occur during distinct, time-bound seasons in which sites are considered 'closed' to changes in occupancy.
-Within seasons multiple observations are conducted, which permit estimation of the probability of detection conditional on occupancy.
-Most frequently these observations are collected by revisiting the site on separate occasions, although alternative methods to generate repeat observations exist [@mackenzie2005].
-These alternatives include surveys conducted at multiple locations within a site, use of multiple observers during a survey, or recording of the time elapsed before a detection is recorded.
+![The form of the dynamic occupancy model as described by @mackenzie2003.
+The ecological process submodel describes changes in occupancy over time, where occupancy in the first season is given by the probability of initial occupancy and occupancy in ensuing seasons is governed by colonisation and extinction probabilities.
+The observational process submodel describes detectability, where the probability of detecting a species given that a site is occupied is given by the detection probability.
+At sites which are truly unoccupied, is it assumed that the species is never detected.](Figures/ModelForm.png){#fig-modelform}
 
-![The sampling design of the dynamic occupancy model.](Figures/SurveyDesign.png){#fig-surveys fig-align="center" width="589"}
+To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys.
+Under this design, observations at a site occur during distinct, time-bound seasons within which it is assumed site occupancy does not change (that is, sites are closed to changes in occupancy).
+In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy.
+Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@mackenzie2005].
 
-## Key assumptions
+![The sampling design of the standard dynamic occupancy model.
+During seasons, also called primary occasions, sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons.
+Within each season, multiple observations ('secondary occasions') are conducted to record the observed presence or absence of the species at each site.
+These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded.
+It is not necessary for the same number of observations to occur in each year or each site, allowing for greater flexibility in data inputs.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}
 
-Dynamic occupancy models require a small number of assumptions with important implications for users.
+This description of the DOM model form and sampling design are based on the original form of the dynamic occupancy model as prescribed in @mackenzie2003.
+Numerous model extensions and alternative formulations are also available, including implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010].
+For a comprehensive discussion of the most common extensions and their applications see @bailey2014, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.
 
-1.
-**False positive detections do not occur.** While this assumption can be safely met in some studies, it is not guaranteed when working with more cryptic species.
-Several authors have commented on the perils of not accounting for false positives where they occur [@miller2015; @mcclintock2010].
-Furthermore, @berigan2019 notes that in some cases genuine identifications of a species may be considered false positives if they do not represent the intended definition of occupancy - in the authors' case, transient individuals where the authors were interested in breeding occupancy.
-Extensions to the dynamic occupancy model which account for false positive error are available [@royle2006; @miller2011; @miller2013] and should be carefully considered where this assumption cannot be met.
+DOMs make a number of assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.
 
-2.
-**Sites are closed to changes in occupancy between seasons.** This assumption, best known as the 'closure assumption,' has also been subject to considerable discussion on the bias which is introduced when it is violated [@rota2009; @otto2013].
-Closure is dependent not only on the behaviour of the species, but also the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discusses the difficultly in distinguishing temporary emigration and local extinction.
-Model extensions to relax the closure assumption have been developed, including @kendall2013 's approach using staggered arrival and departure periods between sites.
-More pertinent for most studies, however, is a careful consideration of ecologically and observationally relevant seasons which will allow the estimation of an appropriate 'type' of occupancy, whether that be shorter or longer term.
+I.
+**False positive detections do not occur**.
+While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods.
+@mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need for authors to consider how certain their detections truly are.
+Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019].
+Where this assumption can not reasonably met, aforementioned extensions designed to account for false positive error should be considered.
 
-3.
-**No modelled heterogeneity in ecological or observational parameters.** Relative to the former two assumptions, considerably less research has been conducted on the impact of modelled heterogeneity on parameter estimates [@mackenzie2017].
-Environmental factors are near-guaranteed to influence species occupancy and observation in any system, particularly at larger spatial scales.
-While studies should attempt to incorporate the key drivers of occupancy via inclusion of covariates, there is limited discussion on the best methods to achieve this and the implications of failure to capture this variation.
+II.
+**Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion around the bias which is introduced when it is violated [@rota2009; @otto2013].
+Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discuss the difficultly in distinguishing between temporary emigration and local extinction.
+Model extensions to relax the closure assumption have been developed, including @kendall2013's approach using staggered arrival and departure periods between sites.
+A more pertinent approach for most studies, however, is careful consideration of ecologically relevant seasons corresponding to an appropriate definition of occupancy.
 
-## Model extensions
+III.
+**Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species occurrence, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system.
+Generally, this is achieved by allowing model parameters to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters.
+An important element of this assumption is that the likelihood of detection of a species depends not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017].
+While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates.
+Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known how this latent heterogeneity can influence model estimates.
 
-The DOM has been frequently extended over the years, with alternate parameterisations available to better address many needs.
-In addition to the false-positive implementations previously discussed, prominent model extensions include multi-state DOMs which allow sites to exist in states other than occupied and unoccupied [@nichols2007], multi-scale models describing occupancy at multiple spatial scales [@nichols2008], and multi-species models permitting joint estimation of species with interactions [@devarajan2020].
-For an in-depth discussion of additional model extensions and their applications see @bailey2014.
+# Systematic review methods
 
-# Methods
+To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data.
+A pool of candidate articles was generated using two queries on Web of Science: the first of these included all articles from 2004-2023 which cite @mackenzie2003; the second query searched articles in the same time-span matching "dynamic occupancy model\*", "multi-season occupancy model\*", and "occupancy dynamic\*", articles including each of "occupancy", "colonization", "extinction/persistence", and "detection", and articles with the term "occupancy" located near "dynamic" in the title, key terms, or abstract.
+As we were interested in how DOMs use has changed through time, we divided these articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024.
+From each of these strata we randomly selected 20 articles for inclusion in the review.
+Articles which did not meet inclusion criteria were replaced from within their own strata.
 
-## Article collection
+As our review focuses only on applications of the dynamic multi-season occupancy model of @mackenzie2003 and its extensions, we included articles which fit a model meeting the following criteria: i) uses non-simulated, field-collected, presence-absence data; ii) relies on data from multiple sites which can exist in at least two states, including an occupied and unoccupied state; iii) has multiple seasons, between which sites may change states conditional on the prior season's occupancy state and transition probabilities such as colonisation and extinction; and iv) contains at least one parameter describing the detection process.
 
-This systematic review is focused on the dynamic, multi-season, implementation of the occupancy model of @mackenzie2003.
-To qualify for inclusion articles must have fit a model which fulfils these criteria: i) describes multiple sites which can exist in at least two states, including an occupied and unoccupied states; ii) has multiple survey periods, between which sites may change state conditional on the prior survey period's occupancy state and at least one transition probability; iii) contains at least one detection parameter; and iv) uses real (i.e., not simulated) field ecological data.
+For each article we recorded key details on authorship, research objectives, study taxa and system, survey methods, and modelling approach.
+To examine the reasons why authors used DOMs, we allocated each article to one or more category of objective based on the study's stated aims.
+The possible categories were *Observing trends*, where authors express interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; *Testing relationships*, where authors explore specific predefined hypotheses of relationships between covariates and model parameters; *Identifying drivers*, where authors attempt to find which covariates influence model parameters; *Predicting temporally*, where authors predict site occupancy under future conditions, *Predicting spatially*, where authors predict site occupancy to unsurveyed locations, and *Developing methods*, where authors introduce, test, or demonstrate aspects of dynamic occupancy models.
 
-Papers were gather for inclusion in two stages.
+We recorded details on the type of organisms (bird, mammal, etc.) modelled in each article, and the means by which multiple species were modelled where applicable.
+Taxa were denoted as threatened either when they are listed on the IUCN Red List of Species or when authors otherwise indicate that they are threatened.
+This deference to authors' representation of conservation status was made to account for sub-species which lack listings or species which are of more local concern.
+Study locality and size was documented; the size of the study area being defined as the intended area of inference which contained all site — this was measured to the order of magnitude to account for uncertainty in reporting.
 
-1.
-In a preliminary analysis, four search strings ("dynamic occupancy model," "occupancy dynamics model," "incidence function model," and "multi-season occupancy model" were entered into Google Scholar.
-Up to 100 articles were downloaded for each string in order of appearance.
-Articles were divided into four temporal strata: 2000-2005, 2006-2010, 2011-2015, and 2016-2021.
-For each combination of search string and year strata, a random 25% or 5 articles (whichever was greater) was processed.
-Articles which did not meet inclusion criteria were replaced with another from the same search string and strata where possible.After completion of the preliminary sample authors recognised significant uncertainty in how the model of interest was described, particularly in their first decade of use, and noted that the incidence-function model is best interpreted as a distinct model class**.**
+We were particularly interested in how authors navigated the modelling process, from covariate selection through model evaluation.
+To this end, we recorded all covariates considered in each study regardless of whether they were or were not included in final models.
+Key traits of each covariate were recorded including their general category, whether they were directly observed or remotely sensed, whether they were static or varied between seasons, and how they were presented in the model (as a linear term, a polynomial term, or as part of an interaction with another covariate [@james2021]).
+Model selection procedures were also sorted into non-exclusive categories including *A priori*, where only one model was considered; *candidate suite*, where a predefined set of models was considered; *sequential,* where covariates were selected parameter-by-parameter; and *simple precursors*, where selection was preceded by another simpler model implementation.
+Finally, for each model we documented whether goodness-of-fit was tested and reported, as well as whether model performance was assessed by validation with either in or out-of-sample data.
+For the full spreadsheet of data collected and further details on categorisation, see **Appendix I.**
 
-2.
-To increase sample size a secondary sample of articles was generated, using the complete list of articles which have cited @mackenzie2003 (downloaded via Web of Science).
-Articles in this list were split into the same year strata, plus an additional 2022-2023 category.
-15 articles were drawn from each strata and again processed with replacement from their strata.
+# Applications of dynamic occupancy models
 
-## Article processing
+A total of 60 articles were included for this review.
+Based on the acceptance rates within each strata, an estimated 516 of the 1300 unreviewed articles in our sample would have met inclusion criteria (@fig-coverage).
 
-For each included article data was collected on authorship, stated research objectives, study taxa and field survey locations, data collection methods, covariates, and model selection and evaluation.
-A full spreadsheet of data collected is given in **Appendix I.**
+![Estimated coverage of the articles included in this review.
+A total of 60 articles from 2000-2023 were processed; bars indicating the estimated number of qualifying articles remaining are based on the rejection rate of each strata and the number of remaining unread articles in the pool.
+The proportion of articles included from each strata were as follows: 2004-2007, 12.5%; 2008-2011, 21.3%; 2012-2015, 44.4%; 2016-2019, 32.4%; 2020-2023, 57.1%.](Figures/CoveragePlot.jpeg){#fig-coverage}
 
-### Authorship
+Articles included in our review demonstrate considerable diversity in scope, scale, and objectives.
+A selection of key attributes of these studies is presented in @fig-StudyDetails, and we provide notable examples of DOMs throughout this section to highlight how some this variation appears in practice.
+Study systems included in our sample are globally distributed; while a majority use data collected in the United States of America, representatives are included from all geographic realms and an exceptional diversity of ecosystems.
+A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale.
+The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre [@basset2023], while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@clement2019].
 
-To assess patterns of authorship, the location of each articles first author's first-listed affiliation was denoted.
-For *all* authors[,]{.underline} each affiliation was denoted as one of four categories: academic institutions, governmental organisations and institutes, non-governmental organisations, and private sector companies.
+The temporal scale of studies shows similar variability.
+The duration between first and last survey ranging from under one month to forty years, with a median of 7 years.
+More meaningful for model interpretation, however, is the duration of the primary occasion, as this is presumed to be the temporal scale at which changes in occupancy are observed.
+Most applications describe the primary occasion as a year, although some studies divide years into multiple primary occasions to describe seasonal variation in occupancy.
+In the most extreme cases the primary occasion may be as brief as a week, capturing much finer scale changes in occupancy.
+These shortened-season occupancy models are most common with camera trapping or acoustic monitoring data, which can be arbitrarily divided into primary and secondary sampling occasions.
+This is seen in @kleiven2020's study of rodent and mustelid interactions using camera trap data, where the primary occasions are less than one week in length.
+On the other end of the spectrum, some studies modelled primary occasions which were decades apart and represented generational changes in occupancy.
+This is illustrated in @couturier2023's study on long-term otter recovery in France, which used data from two primary occasions: one in 2003 and one in 2012.
 
-### Objectives
+DOMs have been used to study a variety of different species, though the majority of studies have been conducted on data for vertebrate taxa.
+They have been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community.
+However, there are exceptions — @belinchón2017 fits a DOM to lichen data, and @cook2022 uses them to model the spread of chronic wasting disease.
+The latter's application to disease dynamics is not unique, and DOMs have been touted as a valuable tool for modelling disease dynamics.
+@mores2020 and @padilla-torres2013 have each used DOMs to model mosquito dynamics as a human disease vector, and @bailey2014 discusses the concept of applying the DOM to the spread of chytrid fungus in amphibian habitats.
+The vast majority of applications have been fit for terrestrial species, though there have been a limited number of studies on aquatic systems, including @fisher2014's application to invasive salmon, @falke2012's on Great Plains stream fishes, and @pendleton2022's on whale occupancy dynamics.
 
-For each article five non-exclusive research objectives were indicated based on the authors' stated aims: 'Estimating parameters,' where authors express interest in estimates of site occupancy, colonisation, or extinction; 'Testing hypotheses,' where authors explore *specific* relationships between covariates and model parameters; 'identifying drivers,' where authors attempt to find *which* covariates influence model parameters; 'generating predictions,' where which authors estimate model parameters either for unsurveyed sites or under possible future conditions; and 'methods development,' where authors introduce or test aspects of dynamic occupancy models.
+While most articles in our sample fit a model to a single species, 26 fit models to multiple species either as independent models or explicitly multi-species implementations.
+The multi-species models used various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects[@dorazio2010; @hendershot2020], as well as explicitly interactive models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019].
+While they did not fit multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012].
+Working with large numbers of taxa does raise additional challenges, as the level of complexity which can realistically be considered for each individual taxa is likely to be reduced for practicality's sake.
 
-### Taxa and study areas
+![A) Study areas: Data for a majority of studies were collected from study locations in the United States.
+The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres.
+\n B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles.
+38 articles fit models to only a single taxa, 20 fit independent models to multiple taxa, and 6 fit multi-species models with explicit interactions between taxa.
+\n C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors.
+10 articles incorporated data from citizen science projects.
+Project scale was also variable, with the median study running for 3.25 years and covering 98 sites.](Figures/ReviewBox.png){#fig-StudyDetails}
 
-Focal taxa were divided into broad categories (bird, mammal, etc.).
-Taxa were denoted as threatened either where the authors refer to them as so or where the species is listed on the IUCN red list (IUCN 2024) - this determination was made to account for unlisted sub-species or species of local concern.
-Study areas are defined as the intended area of inference for a given study in which all sites are contained; size of these areas are reported as orders of magnitude in square kilometres to partly account for uncertainty in reporting.
+The presence/absence data required for DOMs has come from a breadth of different sources and detection methods.
+In our sample 76% of articles used human observational data, 19% used live-trapping methods, and 5% used detections from camera trapping or acoustic monitoring.
+Several articles also use citizen-science data to build detection histories, including @zuckerberg2011's model using bird-feeder data and @warrier2020's using citizen reports of tiger sign.
+Using citizen-science data may require additional consideration of assumptions (particularly false-positives), as discussed in greater detail by @cruickshank2019.
+Within these broad categories of detection method there is additional diversity, with several studies using unique survey methods determining the presence of a species at a location.
+For example, in our only application of a DOM for a marine species @pendleton2022 uses aerial transects broken up into grid cells to observe whale occupancy; and @marescot2020's fits a unique 'multi-species' treating poachers as species which uses ranger reports to create detection histories.
 
-### Covariates
+Notably, many of these studies use data which was not originally collected in a robust design framework for occupancy modelling.
+In these articles, authors manipulated their data into a hierarchical format post-hoc using a variety of methods.
+Some defined primary seasons as a discrete time interval, treating all surveys occurring within the season as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate.
+These manipulations permit authors to use data which far predate the DOM, with one study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].
 
-Particular emphasis was placed on the covariates considered by authors, with no distinction between covariates which were or were not included in final models.
-For each parameter all covariates were recorded in broad mutually-excluded categories.
-For each of these covariates, data was collected on whether the covariate was remotely collected or required direct observation, whether polynomial terms were considered, whether interaction terms included this covariate were considered, and whether the covariate was dynamic – i.e, varied between seasons.
+Additionally, not all articles rely on a single source of detection data - some integrate multiple sources of data to maximise sample size, combining data from camera traps, sign surveys, and citizen science reports.
+These integrated detection method models do require additional care in estimating detection probability, which is generally expected to vary based on observation method; however, this can be reasonably achieved by including survey method as a covariate on detection probability [@warrier2020; @pitman2017].
+A special case exists when different detection methods are used where one has the potential for false positive detections; e.g., where less-certain citizen science detections are combined with certain detections from field surveys.
+In this context, the certain detections are used to help account for false-positive detection probability, as in @miller2011's study integrating GPS collars and hunter reports to estimate wolf occupancy in Montana.
 
-### Model selection and evaluation
+The flexibility in the data used for DOMs, including the model structure and the scale of observations, is not amenable to a one-size-fits-all definition of occupancy.
+Users of DOMs must carefully consider precisely what they are modelling and address questions on the scale represented by their model [@chave2013].
+Drivers of occupancy may differ depending on whether a site is represented by a single point on the landscape or as a grid cell, with the former likely to depend on more local, small scale factors rather than landscape-level trends [@stevens2019].
+This is also true for the temporal scale of occupancy: whether a site is occupied within a week or within a year leads to vastly different conceptions of occupancy.
+This idea is particularly relevant in cases where the selection of season length *is* to some extent arbitrary, as with camera-trap or bioacoustic data where continuous recordings can be broken down into distinct 'seasons' of any length.
+DOMs are well suited to these data types [@balantic2019], and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data.
+For example, @kleiven2020 and @mölle2022 divide their camera trap records into seasons of just a few days.
+While this provides interesting insights of occupancy at extremely fine temporal scale, further research may be needed on how to determine appropriate season and survey durations with respect to research questions and the closure assumption.
 
-To describe the complex model selection process, non-exclusive categories were applied to each article including: 'A priori,' where only one model is considered; 'candidate suite,' where a predefined set of models are considered; 'procedural,' where covariates are selected parameter-by-parameter; 'exhaustive,' where all possible model combinations are fit; 'simple precursors,' where selection is predicated on results of another, simpler model implementation; and 'other.'
+# Practices in implementation and model building
 
-Model evaluation was similarly categorised - options in this case included 'relative fit,' for relative metrics of fit such as AIC; 'model averaging,' where model weights are applied to candidate models; 'goodness of fit,' where absolute metrics of fit are used; 'out-of-sample validation' and 'in-sample-validation,' where predictive performance is evaluated using in-sample or out-of-sample data respectively.
-
-# Results
-
-A total of 84 articles were included in this review, 47 from the preliminary sample and 37from the secondary sample.
-Based on the acceptance rate from the secondary sample, an estimated 448 of the 1066 articles citing @mackenzie2003 would have met inclusion criteria.
-
-![Estimated coverage of articles included in this review, estimated based upon the secondary sample's rate of inclusion for each strata.](Figures/CoveragePlot.jpeg){#fig-coverage}
-
-Articles were published with first-authors from a total of 15countries, with authors from the United States producing 64% of articles.
-86% of articles had at least one author based at an academic institution, 46% had authors at government institutions, and 25% had authors from non-profit organisations.
-
-## Study locations, focal species, and data collection
-
-![An overview of study areas, focal species, and data collection for reviewed articles.](Figures/ReviewBox.png){#fig-StudyDetails}
-
-Data for DOMs were collected from study areas on all continents and in 21 countries, with studies in the United States again representing a majority of the total articles published.
-These study areas ranged in size from less than 1 to over 1 million square kilometres (@fig-StudyDetails a).
-
-Focal species were largely vertebrate taxa, although 6 articles fit models to invertebrates.
-A single non-animal taxa was represented in the sample, from an article fitting a DOM to a rare species of lichen @belinchón2017.
-35% of articles modelled threatened taxa, and 10% modelled species invasive to the study area.
-The majority of articles (50, 60%) fit single-species DOMs to one taxa.
-Of the remainder, 26 (31%) fit independent single-species models to more than one taxa.
-8 articles fit (10%) multi-species implementations with explicit interactions between species (@fig-StudyDetails b).
-
-Data used to fit dynamic occupancy models included human survey data such as transects or point counts (69% of articles), live-trapping data (20%), camera traps (12%) or bioacoustic monitors (1%), and citizen science data (12%); some articles incorporate multiple detection methods to build their detection datasets.
-Study duration, defined as the time elapsed between the first and last surveys conducted, ranged from 1 month to 40 years with the median study running for 39 months.
-The number of seasons ranged from 2 to 204 with a median of 5; the number of observations per season ranged from 2 to 153 with a median of 5.
-The number of sites surveyed per article ranged from 7 to 9394, with the median study including 98 sites (@fig-StudyDetails c).
-
-## Covariates
-
-A summary of covariates considered in the review sample is presented in @tbl-covariates.
-For each category of covariate relative frequency and characteristics are given.
-For all categories, the rate of consideration of polynomial terms does not exceed 16%, and the rate of consideration of interactions does not exceed 34%.
+Building dynamic occupancy models can be a challenging process requiring careful consideration of which environmental factors to incorporate to adequately represent occupancy and detection in complex natural systems.
+In this review, we have recorded all covariates considered for each model in our sample.
+These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of articles considering each type of covariate in their models, the means by which that covariate data is collected, and the ways in which those covariates are statistically represented in DOMs.
+We further categorise these covariates into two groups: environmental covariates, which represent plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environmental itself.
+Our findings indicate that users of DOMs have incorporated an exceptionally wide diversity of different covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample.
+Many models incorporate covariates representing aspects of site geometry (31% of articles) and connectivity (26% of articles), adding spatial components to these DOMs which bear similarities to related stochastic patch occupancy models.
+Notable examples of these spatially-explicit models include @mortelliti2007's model fit to rodent data in patchy landscapes and @risk2011's hybrid DOM/incidence function model used to explore meta-population dynamics of rails.
+Several studies also include biotic interactions with other species as covariates, particularly those which describe taxa threatened by invasive species.
+Examples include models with barred owl presence as a covariate on spotted owl occupancy [@mangan2019] and models with bullfrog covariates on native amphibian occupancy [@rowe2019], each of which effectively incorporate species interactions without fitting explicitly multi-species models.
 
@@
 #| label: tbl-covariates
-#| tbl-cap: "Attributes of covariates found in dynamic occupancy models"
+#| tbl-cap: "All covariates considered for inclusion in a study were classified into mutually exclusive categories.
+We calculate the percentage of articles which include at least one covariate of a given category on any parameter, Initial Occupancy (&psi;1), Occupancy (&psi;), Colonisation (&gamma;), Extinction(&epsilon;), and Detection (&rho;).
+We also present the percentage of all implementations of each covariate category which are dynamic (varying through seasons) and directly observed, as well as the percentage of articles which model each category of covariate as a non-linear relationship or as part of an interaction with another covariate."
 
@@
 covTable |>
-  arrange(Category, -Overall_Proportion) |>
-  gt(rowname_col = "Type") |>
-  tab_header("Covariate attributes") |>
+  filter(!Covariate %in% c("DIST", "PROT")) |>
+  mutate(Summary = case_when(Covariate %in% c("Environmental", "Structural", "Overall") ~ TRUE,
+                             .default = FALSE)) |>
+  arrange(Category, Summary, -Prop_Any) |>
+  gt(rowname_col = "Covariate", groupname_col = "Category") |>
+  tab_options(
+    row_group.as_column = TRUE
+  ) |>
   tab_row_group(
     label = md("Environmental covariates"),
-    rows = Category == "Environmental",
+    rows = Category == "E",
     id = "Env"
@@
   tab_row_group(
-    label = md("Other covariates"),
-    rows = Category == "Other",
-    id = "Other"
+    label = md("Structural covariates"),
+    rows = Category == "S",
+    id = "Struct"
+  ) |>
+  tab_row_group(
+    label = md("All covariates"),
+    rows = Category == "O",
+    id = "All"
   ) |>
   tab_style(
-    style = cell_fill(color = "#82AD8E"),
+    style = cell_fill(color = "aquamarine4"),
     locations = cells_row_groups(groups = "Env")
@@
     style = list(
-      cell_fill(color = "gray70")
+      cell_fill(color = "plum4")
     ),
-    locations = cells_row_groups(groups = "Other")
+    locations = cells_row_groups(groups = "Struct")
   ) |>
-  row_group_order(groups = c("Env", "Other")) |>
-  tab_spanner(label = "Articles with covariate on:",
-              columns = c("Overall_Proportion", "InitOcc_Proportion",
-                          "Colonisation_Proportion", 
-                          "Extinction_Proportion",
-                          "Detection_Proportion"),
-              id = "articles") |>
-  tab_spanner(label = "Covariates which are:",
-              columns = c(Proportion_Dynamic, Proportion_Observed,
-                          Proportion_Polynomial, Proportion_Interact),
+  tab_style(
+    style = list(
+      cell_fill(color = "gray60")
+    ),
+    locations = cells_row_groups(groups = "All")
+  ) |>
+  row_group_order(groups = c("Env", "Struct", "All")) |>
+  tab_spanner(label = "Percentage of articles with covariate on parameters:",
+              columns = c("Prop_Any", "Initial occupancy", "Occupancy",
+                          "Colonisation", "Extinction_Persistence",
+                          "Detection"),
+              id = "inclusions") |>
+  tab_spanner(label = "Percentage which are:",
+              columns = c(Prop_Dynamic, Prop_Direct),
               id = "covariates") |>
+  tab_spanner(label = "Articles representing this covariate with:",
+              columns = c(Prop_Nonlinear, Prop_Interact),
+              id = "representations") |>
    cols_label(
-    Overall_Proportion = "Any parameter",
-    InitOcc_Proportion = "Initial Occ.",
-    Colonisation_Proportion = "Colonisation",
-    Extinction_Proportion = "Extinction",
-    Detection_Proportion = "Detection",
-    Proportion_Dynamic = "Dynamic",
-    Proportion_Observed = "Directly observed",
-    Proportion_Polynomial = "Polynomial",
-    Proportion_Interact = "Interactive"
+    Prop_Any = "Any",
+    `Initial occupancy` = html("&psi;1"),
+    Occupancy = html("&psi;"),
+    Colonisation = html("&gamma;"),
+    Extinction_Persistence = html("&epsilon;"),
+    Detection = html("&rho;"),
+    Prop_Dynamic = "Dynamic",
+    Prop_Direct = "Directly observed",
+    Prop_Nonlinear = "Non-linear relationships",
+    Prop_Interact = "Interactions between covariates"
   ) |>
-  cols_hide(Category) |>
+  text_case_match(
+    "Environmental" ~ html("<b>Any Environmental</b>"),
+    "HABT" ~ html("<b>Habitat type (land)</b>\n<i>Aspects of terrestrial habitat</i>"),
+    "HABA" ~ html("<b>Habitat type (aquatic)</b>\n<i>Aspects of aquatic habitats</i>"),
+    "PHEN" ~ html("<b>Phenology</b>\n<i>Time-varying elements distinct from sampling occasions</i>"),
+    "GEOM" ~ html("<b>Geometry</b>\n<i>Site geometry, such as patch size or shape</i>"),
+    "CONN" ~ html("<b>Connectivity</b>\n<i>Connectivity to other sites<i/>"),
+    "LACO" ~ html("<b>Land cover</b>\n<i>Land cover classes (excluding agriculture)</i>"),
+    "OTHER" ~ html("<b>Other environmental</b>\n<i>Other environmental covariate not otherwise listed</i>"),
+    "TOPO" ~ html("<b>Topography</b>\n<i>Elements of landscape topography</i>"),
+    "CLWE" ~ html("<b>Climate/Weather</b>\n<i>Climate and weather covariates</i>"),
+    "ANTH" ~ html("<b>Anthropogenic</b>\n<i>Relations to human activity</i>"),
+    "HYDR" ~ html("<b>Hydrology</b>\n<i>Hydrological features like rivers or lakes</i>"),
+    "SPAT" ~ html("<b>Spatial</b>\n<i>Spatial variation, such as latitude</i>"),
+    "AGRI" ~ html("<b>Agricultural</b>\n<i>Relating to agricultural features</i>"),
+    "DISA" ~ html("<b>Disturbance</b>\n<i>Natural disasters or disturbances</i>"),
+    "BIOT" ~ html("<b>Biotic interaction</b>\n<i>Interactions with other (non-plant) species</i>"),
+    "Structural" ~ html("<b>Any Structural</b>"),
+    "SEAS" ~ html("<b>Primary occasion</b>\n<i>Effect of the primary occasion</i>"),
+    "OBSE" ~ html("<b>Observation</b>\n<i>Details of the survey</i>"),
+    "SURV" ~ html("<b>Secondary occasion</b>\n<i>Effect of the secondary occasion</i>"),
+    "SPEC" ~ html("<b>Species effect</b>\n<i>Species-level effects</i>"),
+    "SITE" ~ html("<b>Site effect</b>\n<i>Site-level effects</i>"),
+    "Overall" ~ html("<b>All covariates</b>"),
+    .locations = cells_stub()
+  ) |>
+  cols_hide(c(Category, Summary)) |>
   fmt_percent(decimals = 0) |>
-  sub_missing(missing_text = "-")|>
-  tab_style(style = cell_fill(color = "lavender"),
-            locations = cells_body(columns = Overall_Proportion,
-                                   rows = Overall_Proportion == max(Overall_Proportion, na.rm = TRUE))) |>
-    tab_style(style = cell_fill(color = "lavender"),
-            locations = cells_body(columns = InitOcc_Proportion,
-                                   rows = InitOcc_Proportion == max(InitOcc_Proportion, na.rm = TRUE))) |>
-    tab_style(style = cell_fill(color = "lavender"),
-            locations = cells_body(columns = Colonisation_Proportion,
-                                   rows = Colonisation_Proportion == max(Colonisation_Proportion, na.rm = TRUE))) |>
-    tab_style(style = cell_fill(color = "lavender"),
-            locations = cells_body(columns = Extinction_Proportion,
-                                   rows = Extinction_Proportion == max(Extinction_Proportion, na.rm = TRUE))) |>
-    tab_style(style = cell_fill(color = "lavender"),
-            locations = cells_body(columns = Detection_Proportion,
-                                   rows = Detection_Proportion == max(Detection_Proportion, na.rm = TRUE))) |>
-    tab_style(style = cell_fill(color = "lavender"),
-            locations = cells_body(columns = Proportion_Dynamic,
-                                   rows = Proportion_Dynamic == max(Proportion_Dynamic, na.rm = TRUE))) |>
-    tab_style(style = cell_fill(color = "lavender"),
-            locations = cells_body(columns = Proportion_Observed,
-                                   rows = Proportion_Observed == max(Proportion_Observed, na.rm = TRUE))) |>
-    tab_style(style = cell_fill(color = "lavender"),
-            locations = cells_body(columns = Proportion_Polynomial,
-                                   rows = Proportion_Polynomial == max(Proportion_Polynomial, na.rm = TRUE))) |>
-    tab_style(style = cell_fill(color = "lavender"),
-            locations = cells_body(columns = Proportion_Interact,
-                                   rows = Proportion_Interact == max(Proportion_Interact, na.rm = TRUE))) |>
+  sub_missing(missing_text = "-") |>
+  data_color(
+    columns = where(is.numeric),
+    method = "numeric",
+    palette = "viridis",
+    alpha = 0.5
+  ) |>
+  tab_style(locations = cells_stub(rows = "Environmental"),
+            style = cell_fill(color = "aquamarine4", alpha = 0.5)) |>
+  tab_style(locations = cells_body(rows = "Environmental"),
+            style = list(cell_fill(color = "aquamarine4", alpha = 0.5),
+                         cell_text(weight = "bold", style = "italic",
+                                   color = "black"))) |>
+  tab_style(locations = cells_stub(rows = "Structural"),
+            style = cell_fill(color = "plum4", alpha = 0.5)) |>
+  tab_style(locations = cells_body(rows = "Structural"),
+            style = list(cell_fill(color = "plum4", alpha = 0.5),
+                         cell_text(weight = "bold", style = "italic",
+                                   color = "black"))) |>
+  tab_style(locations = cells_stub(rows = "Overall"),
+            style = cell_fill(color = "gray60", alpha = 0.5)) |>
+  tab_style(locations = cells_body(rows = "Overall"),
+            style = list(cell_fill(color = "gray60", alpha = 0.5),
+                         cell_text(weight = "bold",
+                                   color = "black"))) |>
+  tab_style(locations = cells_row_groups(),
+            style = cell_text(color = "white")) |>
   as_raw_html()
-
 ```
@@
 
-Quantity of covariates considered varies considerably both between articles and between parameters as shown in @fig-covariates.
-Of the four standard model parameters, models were least likely to consider covariates for initial occupancy (35% of articles with zero covariates considered), and most likely to consider them for detection (81% of articles with at least one covariate considered).
+Covariate data for studies in our sample was either collected directly by researchers (19% of environmental covariate data), or derived from pre-existing remotely sensed datasets (81%); this of course varies depending on the category of covariate.
+Directly collected data can generally be expected to represent finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely but can often be more proximal drivers of occupancy.
+These high quality covariates do come with trade-offs, as they can be expensive to collect and can often preclude projection to locations where data is unavailable.
+For projects interested in making these projections, remotely-sensed covariates are generally more feasible despite their generally more distal nature [@austin2002].
+29% of environmental factors and 87% of structural factors included in reviewed articles were dynamic covariates which varied through time — this again varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static.
 
-![Distribution of covariate quantity considered for model parameters across articles.](Figures/CovParamPlot.jpeg){#fig-covariates}
+In the standard DOM covariates are incorporated for each parameter via a logit link, meaning that the statistical relationship between covariates and model parameters is represented as a linear term unless otherwise specified.
+Of course, many ecological relationships are non-linear and require more complex forms to be realistically represented in a model.
+@austin2007 discussed the importance of modelling ecologically-realistic responses to covariates, advocating for careful consideration of the most appropriate statistical form for hypothesised relationships.
+However, linear models are quite flexible, and non-linear responses can be easily accommodated using polynomial terms and interactions between covariates.
+Despite this, in our sample just 38% of articles considered using at least one non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear relationships.
+Interactions between covariates were similarly rare, with only 28% of studies considering at least one relationship between terms.
+The relatively low emphasis on more complex non-linear responses contrasts with other popular methods for modelling species occupancy.
+Many common approaches for SDMs, which include MAXENT and Boosted Regression Trees, permit considerable flexibility in the shape of their covariate response curves where supported by the data [@merow2013; @elith2008].
+This emphasis on more complex responses in SDMs may be due to their typically larger range-wide scale which encompasses the full species niche, where environmental relations may be expected to be non-linear.
+However, as previously indicated DOMs have been frequently implemented at similarly large spatial scales where the same assumptions should exist and similar responses might be expected.
+Similar concerns exist for the relatively low level of interactions between covariates: while they should not necessarily be used liberally, where these exist their exclusion can influence model performance [@guisan2006].
 
-Model complexity varies with the stated objectives of articles, with those interested in making predictions and identifying drivers of occupancy being more likely to consider a larger number of covariates (@tbl-objectives).
-These articles were also more likely to incorporate polynomial and interactive terms.
-Articles with objectives involving methods development considered fewer and less complex covariates.
+In addition to the types of covariates considered for modelling we also measured the size of the candidate covariate pool in each reviewed study, tallying the number of environmental and structure covariates which were available for use on each parameter ( @fig-covariates).
+This is an area of considerable variation in modelling practices — the number of covariates considered for each parameter ranges from 0 (effectively modelling the parameter as a constant) to over 40 possible factors.
 
-<div>
+![The number of covariates considered for each parameter across all studies in our sample.
+'Occupancy' given here represents an alternative parameterisation of the DOM which jointly estimates Occupancy, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation.
+Here, a 'covariate' is defined as each distinct covariate considered for inclusion.
+Linear and quadratic representations of the same covariate are counted as 1 covariate.](Figures/CovParamPlot.jpeg){#fig-covariates}
 
-```{r Objectives table}
+The median number of covariates varies strongly by parameter, with transition probabilities (colonisation and extinction) more likely to consider a broader range of environmental covariates relative to initial occupancy and detection.
+The lack of any covariates considered for initial occupancy in many studies (37% of articles) is particularly notable — unless a study is conducted at very small extents or study sites are truly uniform in their suitability, one would expect some amount of non-random variation in occupancy probability across any study system that will not be captured when representing initial occupancy as a constant.
+Omission of the factors which drive occupancy is a known source of bias in static SDMs [@barry2006], which the initial occupancy component of the DOM conceptually resembles.
+Furthermore, any bias in occupancy estimation in the first time-step will perpetuate into future seasons due to the DOM's Markovian nature with implications for the reliability of model outputs.
+Detection probability is also typically represented with low quantities of environmental covariates.
+Recall that detectability is dependent on not only observation factors, but also by drivers of abundance and species use; the low number of candidate covariates here relative to colonisation and extinction raises additional questions on whether this variation is always appropriately captured in models.
+
+To this point in our review, we have discussed only the covariates which were *considered* for inclusion in DOMs without regard for which terms were actually *included* in the final models used by authors to make their inference.
+Identifying the 'best' model from a range of possible candidates can be a challenging process, especially for hierarchical models like DOMs [@doherty2012].
+Consider that the quantity of candidate models involving up to *n* covariates on *p* parameters can be given by the formula $(2^n)^p$; where a simple single-parameter may have have up to 64 possible combinations of 6 covariates, a 4-parameter dynamic occupancy model would have 16 million combinations.
+This explosion of candidate models generally precludes exhaustive comparisons of possible models and requires a reduction of the models tested.
+Models in our sample include both Frequentist (n = 47) and Bayesian (n = 13), which differ considerably in their manner of implementation.
+A summary of modelling practices in DOMs is given in @tbl-modelling below.
+
+```{r Modelling summary table}
 #| echo: false
 #| message: false
-#| label: tbl-objectives
-#| tbl-cap: "Complexity of models by article objective"
+#| label: tbl-modelling
+#| tbl-cap: "Modelling practices in dynamic occupancy models, subset by frequentist or bayesian implementations.
+The median covariate count presented here represents the median quantity of covariates considered for each model parameter across the studies in our review.
+The model selection methods represented in this table are non-exclusive and some articles employ multiple approaches."
 
-objTable <- read_csv("Figures/ObjectiveTable.csv")
+modellingSummary <- read_csv("Figures/ModellingSummary.csv")
 
-objTable |>
-  mutate(Objective = fct(Objective,
-                         levels = c("Hypothesis", "Drivers", "Parameters",
-                                    "Methods", "Predict", "Any"))) |>
-  arrange(Objective) |>
-  mutate(Objective = case_when(Objective == "Hypothesis" ~ "Testing hypotheses",
-                               Objective == "Drivers" ~ "Identifying drivers",
-                               Objective == "Predict" ~ "Making predictions",
-                               Objective == "Methods" ~ "Developing methods",
-                               Objective == "Parameters" ~ "Estimating parameters",
-                               Objective == "Any" ~ "Any objective")) |>
-  gt(rowname_col = "Objective") |>
-  tab_header("Model complexity by article objectives") |>
-  tab_spanner(label = "Articles",
-              columns = c("Articles", "ArticlesProp"),
-              id = "articles") |>
-  tab_spanner(label = "Percent with:",
-              columns = c("ProbPoly", "ProbInt",
-                          "ProbEval", "ProbSelect"),
-              id = "percents") |>
-  cols_label(Articles = "n", ArticlesProp = "%",
-             MedianCovs = "Median covariates considered",
-             ProbPoly = "Polynomial terms",
-             ProbInt = "Interaction terms",
-             ProbEval = "Objective evaluation",
-             ProbSelect = "Model selection") |>
-  fmt_percent(columns = c(ArticlesProp, 
-                          ProbPoly, ProbInt, 
-                          ProbEval, ProbSelect),
-              decimals = 0) |>
-  tab_style(style = list(cell_fill(color = "gray80"),
-                         cell_text(style = "italic")),
-            locations = cells_body(rows = Objective == "Any objective")) |>
-  tab_style(style = cell_fill(color = "lavender"),
-            locations = cells_body(columns = MedianCovs,
-                                   rows = MedianCovs == max(MedianCovs, na.rm = TRUE))) |>
-  tab_style(style = cell_fill(color = "lavender"),
-            locations = cells_body(columns = ProbPoly,
-                                   rows = ProbPoly == max(ProbPoly, na.rm = TRUE))) |>
-  tab_style(style = cell_fill(color = "lavender"),
-            locations = cells_body(columns = ProbInt,
-                                   rows = ProbInt == max(ProbInt, na.rm = TRUE))) |>
-  tab_style(style = cell_fill(color = "lavender"),
-            locations = cells_body(columns = ProbEval,
-                                   rows = ProbEval == max(ProbEval, na.rm = TRUE))) |>
-  tab_style(style = cell_fill(color = "lavender"),
-            locations = cells_body(columns = ProbSelect,
-                                   rows = ProbSelect == max(ProbSelect, na.rm = TRUE))) |>
+modellingSummary |>
+  mutate(Variable = fct(Variable, 
+                        levels = c("Number", "MedianCovs", 
+                                   "Prop_AnySelect", "Prop_Candidate",
+                                   "Prop_Sequential", "Prop_Simple",
+                                   "Prop_ModAvg", "Prop_GOF", 
+                                   "Prop_Eval"))) |>
+  arrange(Variable) |>
+  mutate(Variable = as.character(Variable)) |>
+  gt(rowname_col = "Variable") |> 
+  tab_row_group(
+    label = "Covariate selection methods",
+    rows = c("Prop_AnySelect", "Prop_Candidate",
+             "Prop_Sequential", "Prop_Simple", "Prop_ModAvg"),
+    id = "select"
+  ) |>
+  tab_row_group(
+    label = "Model evaluation conducted",
+    rows = c("Prop_GOF", "Prop_Eval"),
+    id = "eval"
+  ) |>
+  row_group_order(groups = c(NA, "select", "eval")) |>
+  text_case_match("Number" ~ "Number of studies",
+                  "MedianCovs" ~ "Median covariates considered per parameter",
+                  "Prop_ModAvg" ~ "Percentage using model-averaging",
+                  "Prop_AnySelect" ~ "Percentage using any model selection approach",
+                  "Prop_Sequential" ~ "Percentage using procedural model selection",
+                  "Prop_Simple" ~ "Percentage selecting covariates with simpler models",
+                  "Prop_Candidate" ~ "Percentage comparing models in a candidate set",
+                  "Prop_GOF" ~ "Percentage calculating goodness-of-fit",
+                  "Prop_Eval" ~ "Percentage assessing predictive performance",
+                  .locations = cells_stub()) |>
+  fmt_percent(rows = 3:9, decimals = 0) |>
+  fmt_integer(rows = 1) |>
+  tab_stub_indent(
+    rows = 3:9,
+    indent = 3
+  ) |>
+  tab_style(
+    locations = cells_row_groups(),
+    style = cell_text(style = "italic")
+  ) |>
+  tab_style(
+    locations = cells_column_labels(),
+    style = list(cell_fill(color = "aquamarine4"),
+                 cell_text(color = "white"))
+  ) |>
+  tab_style(
+    locations = cells_body(columns = "All models"),
+    style = cell_text(style = "italic")
+  ) |>
   as_raw_html()
@@
 
-</div>
+The largest differences between frequentist and Bayesian models were in their approaches to model selection.
+Where 91% of frequentist models performed some manner of model selection to determine covariate inclusions for their models, only 31% of Bayesian models did so, with the majority instead fitting a single model defined *a priori*.
+Methods used for model selection varied considerably both within and between modelling frameworks.
+For frequentist models, the most popular and conventional approach to model selection as used in 49% of articles involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score.
+Most frequentist studies which did not follow this protocol instead used procedural model selection methods (40% of articles), where the structure for each model parameter is fit one-by-one.
+For example, this protocol might first identify the best structure for detection probability before moving on to initial occupancy and so on.
+A smaller portion of frequentist studies (4%) used more unique approaches, fitting simpler models such as single season occupancy models for candidate covariates to identify the most informative terms to use in a DOM.
+Across all frequentist implementations, 70% of articles conducted multi-model inference and model-averaging using AIC weights [@burnham2004].
 
-Bayesian models were far less likely to implement a model selection protocol compared to frequentist models: 70% of the former fit only a single model defined *a priori,* compared to 10% of the latter (@fig-selection).
-In frequentist models, procedural selection protocols in which parameters are fixed one by one were the most popular choice.
-Nearly all frequentist models used AIC or AICc to identify their best model.
+Those Bayesian models which *did* perform model selection took distinctly different approaches, with largely idiosyncratic methods across these studies.
+While direct comparison of model fit is rare, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data.
+Another approach used by @cook2022 fit a global model including all covariates before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the model.
+@ahumada2013 takes a particularly interesting hybrid approach, in which model selection is conducted by a procedural method in the frequentist framework before refitting the best structure in a Bayesian model.
 
-![](Figures/SelectionFigure.jpeg){#fig-selection}
+Limited research has been conducted on advantages of different methods for covariate selection in DOMs, and there is surely no one-size-fit all approach which will be appropriate for all possible use-cases.
+However, it is important to consider the implications of the different model selection approaches in common usage, and research from the SDM literature highlights how covariate selection can influence our interpretation of model outputs @brodie2020.
+Within the occupancy modelling literature, @stewart2023's article on covariate selection in single-season occupancy model discusses important attributes of the information-criteria approaches widely used in frequentist models in our review, noting that these approaches can make inaccurate estimates and influence the inference which we make from our models.
+@morin2020's work raises different concerns on procedural model selection methods, demonstrating how the fine details of modelling protocols can determine whether the best-fitting possible model is identified and which covariates appear in final models.
+The literature on performance of Bayesian model selection methods is more sparse, although @hooten2015's guide to Bayesian model selection in ecology is a valuable resource for possible methods of fitting those models.
 
-Objective evaluation of model fit was rare throughout the sample regardless of selection protocol.
-Just 8% of articles calculated non-relative metrics, either via in-sample validation, bootstrapped goodness-of-fit tests, or out-of-sample validation.
+With few exceptions, most of the model selection methods identified in our review are designed only to identify the best model of those considered, and do not necessarily guarantee that selected models are well-fit to the data.
+Assessment of the performance of DOMs generally requires additional steps, and the best ways to achieve this are not necessarily clear.
+There is no broadly-accepted goodness-of-fit test for dynamic occupancy models, although @mackenzie2004 describes an approach for single-season occupancy models using a parametric bootstrap which can been extended to DOMs; this test is implemented in the *AICcModAvg* and *unmarked* R packages.
+While these tests may not suit Bayesian implementations, @broms2016 discusses possibilities for model evaluation in that context; their article focuses on single-season multi-species occupancy models but extensions of their approach may also be applicable to the DOM.
+Model evaluation for DOMs in general can be difficult and somewhat uncertain compared to other models, as the primary response variable of interest (species occupancy) is a latent variable where the true state is generally not known.
+Predictive performance evaluation is thus typically based on *observed* occupancy, where a DOM is used to simulate detection histories to be compared with field survey results.
 
-# Discussion
+Perhaps as a result of this uncertainty, assessment of model fit and model performance was rare amongst the articles in our sample.
+Only 19% of studies tested for goodness-of-fit, and just 10% calculated predictive performance.
+These rates are considerably lower than for models like SDMs; for which @araújo2019's review of applications found closer to 50% articles testing both fit and performance.
 
-### Applications of dynamic occupancy models
+# Modelling objectives
 
-DOMs display great variation of scale in all senses.
-Analyses in our samples range from the hyper-local at individual parks [@zúñiga-vega2019] to continental-scale studies \[@zuckerberg2011\].
-Similar diversity in study duration and site quantity indicates further emphasises the breadth of studies which DOMs may be fit to.
-Method of species detection also need not be a limitation for the implementation of DOMs, as articles in our sample show that many approaches may be coerced into data suitable for this model class.
-While studies pre-designed to match hierarchical data inputs may be simplest to align with model assumptions, post-hoc manipulation of other data such as long-term citizen monitoring programs \[@zuckerberg2011; @peach2019\], disparate agency modelling programs [@mcgowan2020], or camera trapping data [@davis2018] is feasible with careful consideration.
-Detections from multiple sources, with associated variation in efficiency, have also been incorporated by authors [@warrier2020; @pitman2017].
+Studies in our sample were classified according to their objectives as stated by authors to explore the use-cases which DOMs have been applied for.
+These objective categories were non-exclusive, and many studies fell into multiple categories; the frequency at which studies pursued each of these objectives is shown in @fig-objectives A.
+Across our study period DOMs have been frequently employed to observe trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions.
+Within each of these broad categories lies even more variation in objectives, emphasising the DOM's flexibility as a tool for making ecological inference in diverse contexts.
 
-Reviewed implementations of dynamic occupancy models show widespread representation from mammals, birds, and herptiles.
-This assemblage contrasts with the related stochastic patch occupancy models (SPOMs, @gutiérrez-arellano), where the majority of models are fit to invertebrates, plants, and fungi.
-This difference may reflect SPOMs' assumption of perfect detection and discrete habitat patches, limitations which are unlikely to be fulfilled by most vertebrate taxa.
-Aside from differences in assumptions around detection, SPOMs also differ from DOMs in that they are generally spatially explicit, with probabilities of colonisation and extinction dependent on patch size and the occupancy state of nearby patches.
-Where desired, these features may also added to dynamic occupancy models — @sutherland2014 presents an implementation which may be considered a fusion of the DOM and SPOM methods, and several other authors incorporate patch characteristics aspects into their models as covariates [@duggan2011; @broms2016].
+Many articles in our sample used the DOM and sought to monitor trends in occupancy state through time.
+These studies include @ahumada2013's paper assessing populations trends of 13 forest mammals and @scott2015's assessment of the conservation status of the endangered flattened musk turtle.
+These studies offer valuable insights on the state of species across landscapes and through time, and DOMs are a strong candidate for many monitoring oriented projects — including those where many species need to be assessed at once.
+Other articles were more focused on examining the factors which influence species occupancy, either testing a pre-specified hypothetical relationships or taking a broader tack to identify drivers of occupancy without preconception.
+The distinction between the two is important, as it guides which covariates may be considered and how model selection may be used to fit models.
+Many of these studies target core conservation priorities for their focal species, like @olson2005's early DOM assessing the influence of barred owls on threatened spotted owl.
+Explorations of these pivotal relationships are important for guiding management action, and may also be used to test the effectiveness of these actions as in @miller2023's study on American Kestrel nest boxes.
+In scenarios where less is known about species habitat preference, DOMs may be used to examine the influence of a wider variety of factors as illustrated in @huber2017's study testing the relative influence of dozens of habitat covariates on Wood warbler occupancy.
+In recent years, increasing numbers of articles have used to DOM to generate predictions of species occupancy.
+As examples, @mcgowan2020 provides projections for the threatened Black rail under multiple management scenarios, and @pollentier2021 generates maps of distributions resembling those made with SDMs, outputs which can immediately be used by managers to inform action.
 
-While objectives vary in our sample, DOMs are rarely used to extrapolate beyond study areas — use of prediction remains the least popular use-case in our sample.
-Articles which do make predictions provide some of the most directly applicable outputs found in the review sample: @mcgowan2020 provides projections for a threatened species under multiple putative management scenarios, and @pollentier2021 generates maps of distributions more frequently associated with species distribution models (SDMs).
-DOMs offer advantages over other SDMs for this purpose, particularly where populations are assumed not to be at equilibrium as is the case for many threatened and invasive species.
-By explicitly accounting for changes in occupancy (i.e., colonisation and extinction) DOMs may better capture future range contraction or expansion — an aspect of prediction where other SDMs may struggle.
+![A) Shows the proportion of articles in each year-strata and across all years which match each of six non-exclusive objective categories.
+B) Shows the quantity of covariates considered per parameter for models which pursued each objectives.
+Studies focused on methods development typically considered the fewest covariates, where those attempting to identify the drivers of occupancy considered the most.](Figures/ObjectiveFigure.jpeg){#fig-objectives}
 
-### Practices in the modelling process
+The objectives of studies fitting DOMs strongly inform approaches to the model building process, particularly with respect to covariate selection.
+@fig-objectives B shows the quantity of covariates considered for models applied to each of our objective categories, and the differences between objectives are apparent.
+Not unexpectedly, articles which focused on describing new methods for DOMs had fewer covariates than those studies focused on more applied objectives.
+However, differences persist between those articles observing trends, identifying trends in occupancy, testing relationships, and making predictions.
 
-A diverse set of covariates were identified in the model sample reflecting the diversity of systems and ecological questions described by DOMs.
-Notably, many models lack covariates on the initial occupancy parameter, raising questions on how the parameterisation of the model in it's first time-step may affect the quality of estimates at later time-steps.
-The distinction in covariates considered for the ecological parameters vs those considered for detection is also of note: ecological parameters more frequently contained covariates describing environmental features, where detection more frequently contained more structural covariates such as season.
-Covariate complexity is relatively low across the board, with only a fraction of models implementing either non-linear responses to covariates or interactions between covariates.
-While this addition complexity does have implications for model selection, it may be necessary to accurately describe many ecological relationships.
+# Synthesis and key priorities
 
-Model selection approaches diverge between Bayesian and frequentist implementations of the model.
-While model selection will naturally be practised differently between these two categories, it is unclear what effect a lack of model selection may have on the outputs produced by Bayesian models, and it is likely that Bayesian model selection approaches should be further explored @hooten2015.
-Model selection approaches in the Frequentist applications show limited consensus, although most do consider multiple models.
-Notably, @morin2020 found that the popular procedural approaches to model selection (where one sub-model is assessed at a time) may not provide optimal results.
-Considerably more research is required on this topic to clarify how authors should best approach covariate selection.
+Approaches to building any type of model will necessarily depend on the possibilities of the data at hand and on the priorities of the model-builder.
+This precludes any prescription of the 'best' way to build a model, however, there are still important discussions to be had on decisions made in the modelling process.
+One aspect of fitting DOMs meriting broader discussion revolves around 'model complexity,' and how much must be incorporated into models to reliable model species occupancy under different contexts and use cases.
+Complexity is a broad term which encompasses many aspects of a model [@merow2014], and opinions on simplicity versus complexity in ecological models can be divisive.
+Where some advocate for the simplest possible models, arguing that they are most generaliseable; others insist that overly-simple models cannot adequately represent the most important drivers in a system [@evans2013; @lonergan2014].
+By their nature DOMs are somewhat more complex than simpler models for studying occupancy thanks to their hierarchical structure, attributes which are necessary to control for detectability and to capture occupancy dynamics.
+Within this structure, however, further complexity is to some degree up to the modeller: one can choose how many covariates to consider for inclusion on the various parameters, and how to represent the nature of the relationship between those covariates and parameters.
+Research from SDMs indicates that allowing for more complex relationships can improve model performance in predicting occupancy [@valavi2023], an increasingly popular use-case for DOMs.
+Within the DOM framework, there are exciting developments on that front — @joseph2020 presents a novel neural-network occupancy model which allows for exponentially higher levels of complexity and may offer improved performance for prediction-oriented studies.
 
-### Key considerations for model users
+Generally speaking, covariate selection seems to be a particularly important area for further investigations into building DOMs.
+In our review, we see little consensus around which approaches are most applicable for any given use case, and existing work on covariate selection in DOMs raises concerns on whether common methods always produce the most suitable models.
+This is true for both frequentist and Bayesian implementations, and comparative research on covariate selection under both frameworks may help to inform model users on which method may be most appropriate for their use-cases.
+In a similar vein, the low number of articles in our review which conducted model evaluation or assessed model fit raises different questions.
+While the appropriate method of model evaluation may depend on data availability and research objectives, assessing models by some method is generally important to understand how reliable model outputs may be [@araújo2005; @guisan2005].
+Existing uncertainties around whether current methods are suitable for the task surely discourage users from calculating these metrics, and further research is needed to establish trusted practices for assessing the quality of DOMs.
 
-Implementation of dynamic occupancy models requires considerable thought to ensure that model outputs correspond with author expectations and ecological realities.
-Based upon our review of the existing literature we discuss key questions authors should carefully engage with during the modelling process.
-
-1.
-**Are seasons ecologically meaningful, and how do they affect the definition of occupancy?** As previously discussed, the duration of the season has important implications for both the reliability and proper interpretation of model results.
-Seasons which are not ecologically meaningful may run afoul of the closure assumption resulting in associated biases.
-Also relevant is how the definition of 'occupancy' may shift with season length and life-history of a species — for many species, occupancy within a year and within a week may be considerably different concepts, with shorter intervals more closely relating to the idea of 'use.' With the proliferation of continuous detection methods such as camera traps and acoustic monitors where the definition of 'season' is to some degree arbitrary, it is increasingly important to thoughtfully consider how this element of the DOM is defined.
-
-2.
-**Are key drivers of occupancy and detection incorporated?** While it is effectively impossible to *fully* describe heterogeneity across a landscape, authors should make their best efforts to capture the drivers of occupancy and detection most likely to influence their study species.
-This is true even when another covariate is of principal interest to investigators — failure to include other major determinants of occupancy is likely to bias parameter estimates.
-One place where this is of particular importance is in determining drivers of detection, which is not only subject to observational conditions but also ecological determinants which may cause the species to spend more time at a location and thus be more frequently available for detection.
-
-3.
-**Ensure covariates reflect ecological complexity.** Many ecological relationships are non-linear and may require polynomial terms to be accurately represented.
-This is particularly true at large spatial scales where a larger portion of a species ecological niche is represented in surveyed data.
-Authors should question whether a linear term is sufficient to capture hypothetical relationships in their study systems, and consider quadratic terms as well as interactions between covariates where appropriate.
-
-4.
-**Consider model evaluation.** Model evaluation remains rare in ecological research relative to other fields embracing complex modelling, likely due to a reluctance to 'throw away' data for use in testing.
-This concept should be challenged, particularly for larger scale long-term projects where reserving a fraction of data for testing is unlikely to substantially reduce precision of parameter estimates.
-For studies where this is genuinely not possible, reporting results of goodness-of-fit tests may still provide valuable information to readers.
-
-### Priorities for future research
-
-Work on dynamic occupancy modelling over the past 20 years has developed this model class into an important tool for describing species distributions and ecological relationships.
-Historically, methods-development research for dynamic occupancy models has focused on questions of model structure — namely ideas around the closure assumption and imperfect detection.
-This emphasis likely reflects the DOM's roots in the mark-recapture literature, where these topics have been discussed for several decades.
-Considerably less emphasis has been placed on the relation of model parameters to environmental factors via covariates and the model selection processes used to do so, despite this being an objective in a considerable fraction of studies fitting DOMs.
-This stands in stark contrast to related correlative species distribution models, where these topics have been the subject of considerable research over the years.
-There are opportunities for the DOM literature to learn from species distribution models, and more targeted research on the influence of the model selection process is necessary.
-
-An important opportunity for future research relates to autonomous detection methods, such as camera traps and bio-acoustic monitors — these technologies have permitted collection of enormous quantities of data particularly well-suited to dynamic occupancy models, but have also raised additional questions around their implementation.
-The definition of seasons is of increased importance for these data types, where division of data into 'seasons' and 'observations' is arbitrary.
-Further research should explore the implications of this phenomenon and the additional modelling opportunities made possible by this novel data type.
-
-The past two decades have shown that dynamic occupancy models are an excellent tool for addressing a wide range of ecological questions relating to species occupancy throughout space and time.
-The flexibility afforded by model extensions and the considerable advantages of the class over other species distribution models is likely to ensure that these models continue to increase in popularity.
-Given their strong use on threatened taxa and excellent potential to support environmental management, it is important that research continues to support model users in creating the most reliable models possible.
+These priorities are particularly important given the frequent applied objectives of DOM users, tackling challenges which include assessing critically endangered species @carvalho2023, guiding public health management of disease vectors @moreira2016, and tracking rapidly developing biological invasions @wood2020.
+DOMs are well suited to these situations, and it is to be expected that as these types of applications are more commonly attempted understanding the sensitivity of model outputs to decisions made in the model fitting process becomes increasingly important.
+In the two decades since the publication of @mackenzie2003 the dynamic occupancy model has become a widely used tool for ecological inference, with numerous extensions to the modelling framework further broadening the scope of questions and use-cases for which it may be applied.
+Given their increasing popularity, further research and guidelines around issues of model building may help to make the DOM format more accessible to newcomers and support confidence in interpretation of models.
+Parallels to existing work in the SDM literature on guidelines for reporting and modelling standards could be valuable contributions in achieving these aims [@zurell2020; @araújo2019] while specifically addressing model attributes unique to DOMs.

### 2024-08-28 Analysis/ReviewFigures/Final_Figures.qmd
Re-rendered figures

@@
   mutate(Taxa = fct(Taxa, levels = rev(c("Bird", "Mammal", "Herptile",
-                                     "Invertebrate", "Other")))) |>
+                                         "Fish", "Invertebrate", "Other")))) |>
 

### 2024-08-29 Manuscript.qmd
shuffled directory for Manuscript

--- 2024-08-25 Manuscript/domReview.qmd
@@
 The dynamic, multi-season occupancy model (DOM) offers a framework for occupancy estimation which explicitly accounts for observation error, while also incorporating additional mechanistic process by explicitly estimating colonisation and extinction rates.
-In the two decades since their introduction the DOM has become a popular tool for a popular tool for describing patterns in occupancy, exploring the environmental drivers of species occurence, and, more recently, for predicting species occupancy to new locations and to future conditions.
+In the two decades since its introduction, the DOM has become a popular tool for describing patterns in occupancy, exploring the environmental drivers of species occurence, and, more recently, for predicting species occupancy to new locations and to future conditions.
 
   In this review we examine how ecologists have applied the DOM in its first two decades of use.
-Based on a sample of articles fitting DOMs to field ecological data, we explore the systems in which these models have been used, how data was collected for them, and how authors have built their models to achieve their objectives.
+Based on a sample of articles fitting DOMs to field ecological data, we explore the systems in which these models have been used, how data were collected for them, and how authors have built their models to achieve their objectives.
 Our findings indicate that the DOM is an particularly flexible tool readily applied to a range of diverse taxa from around the globe, capable of being used to estimate occupancy at study sites ranging from local to continental scales over time periods from months to decades.
@@
 These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.
-
 bibliography: references.bib
-csl: apa.csl
-embed-resources: true
 ---
@@
 
-The site occupancy models first introduced by @mackenzie2002 offer a powerful framework for modelling presence/absence data while accounting for each of these challenges.
-Originating in the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy.
+The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017].
+Drawing on principles from the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy.
 Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time.
-This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly incorporating the key processes which drive occupancy while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic data required by more process-explicit models [@briscoe2019].
+This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes which drive occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic data required by more process-explicit models [@briscoe2019].
 These features make the DOM an important tool, with uses for tasks including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].
@@
 
-The DOM unites two processes: the ecological process of site occupancy describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed given that a site is occupied (@fig-modelform).
-Under this parameterisation, sites exist in either an occupied or unoccupied state.
-In the first time-step, occupancy state at each site is determined by the probability of initial occupancy.
-In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation and extinction.
+The DOM structure encompasses two processes: the ecological process of site occupancy dynamics describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed at a site where it is present (@fig-modelform).
+Sites exist in either an occupied or unoccupied state.
+In the first time step, occupancy state at each site is determined by the probability of initial occupancy.
+In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation and local extinction.
 In the observation component of the DOM, the model explicitly accounts for imperfect detection.
@@
 The observational process submodel describes detectability, where the probability of detecting a species given that a site is occupied is given by the detection probability.
-At sites which are truly unoccupied, is it assumed that the species is never detected.](Figures/ModelForm.png){#fig-modelform}
+At sites which are truly unoccupied, it is assumed that the species is never detected (i.e., no false positives).](Figures/ModelForm.png){#fig-modelform}
 
@@
 Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@mackenzie2005].
+It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this allows for considerably greater flexibility in data inputs.
 
@@
 These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded.
-It is not necessary for the same number of observations to occur in each year or each site, allowing for greater flexibility in data inputs.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}
+Note that it is not necessary for each site or season to have the same number of observations, and that missing data can be accommodated.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}
 
-This description of the DOM model form and sampling design are based on the original form of the dynamic occupancy model as prescribed in @mackenzie2003.
-Numerous model extensions and alternative formulations are also available, including implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010].
-For a comprehensive discussion of the most common extensions and their applications see @bailey2014, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.
-
-DOMs make a number of assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.
+DOMs make a number of key assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.
 
@@
 Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019].
-Where this assumption can not reasonably met, aforementioned extensions designed to account for false positive error should be considered.
+Where this assumption can not reasonably met, model extensions designed to account for false positive error should be considered.
 
@@
 While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates.
-Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known how this latent heterogeneity can influence model estimates.
+Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known about how this latent heterogeneity can influence model performance.
+
+Since its original description in @mackenzie2003, the DOM has been further developed with numerous model extensions and alternative formulations including implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010].
+For a comprehensive discussion of the most common extensions and their applications see @bailey2014 and @guillera-arroita2017, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.
 
@@
 To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data.
-A pool of candidate articles was generated using two queries on Web of Science: the first of these included all articles from 2004-2023 which cite @mackenzie2003; the second query searched articles in the same time-span matching "dynamic occupancy model\*", "multi-season occupancy model\*", and "occupancy dynamic\*", articles including each of "occupancy", "colonization", "extinction/persistence", and "detection", and articles with the term "occupancy" located near "dynamic" in the title, key terms, or abstract.
-As we were interested in how DOMs use has changed through time, we divided these articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024.
+A pool of candidate articles was generated using two queries on Web of Science.
+The first of these included all articles from 2004-2023 which cite @mackenzie2003.
+To ensure that relevant articles which did not directly cite @mackenzie2003 were captured, a second query searched articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract.
+As we were interested in how DOMs use has changed through time, we divided all articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024.
 From each of these strata we randomly selected 20 articles for inclusion in the review.
@@
 We recorded details on the type of organisms (bird, mammal, etc.) modelled in each article, and the means by which multiple species were modelled where applicable.
-Taxa were denoted as threatened either when they are listed on the IUCN Red List of Species or when authors otherwise indicate that they are threatened.
+Taxa were denoted as threatened either when they are listed on the IUCN Red List of Species or when authors indicate that they are threatened.
 This deference to authors' representation of conservation status was made to account for sub-species which lack listings or species which are of more local concern.
-Study locality and size was documented; the size of the study area being defined as the intended area of inference which contained all site — this was measured to the order of magnitude to account for uncertainty in reporting.
+Study locality and size was documented; the size of the study area being defined as the intended area of inference which contained all sites — this was measured to the order of magnitude to account for uncertainty in reporting.
 
@@
 Finally, for each model we documented whether goodness-of-fit was tested and reported, as well as whether model performance was assessed by validation with either in or out-of-sample data.
-For the full spreadsheet of data collected and further details on categorisation, see **Appendix I.**
+For the full spreadsheet of data collected and further details on categorisation, see supplementary material.
 

### 2024-08-30 Manuscript.qmd
Next draft version

@@
 These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.
+
 bibliography: references.bib
+csl: apa.csl 
+embed-resources: true
 ---
@@
 In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy.
-Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@mackenzie2005].
+Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@guillera-arroita2017].
 It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this allows for considerably greater flexibility in data inputs.
@@
 The first of these included all articles from 2004-2023 which cite @mackenzie2003.
-To ensure that relevant articles which did not directly cite @mackenzie2003 were captured, a second query searched articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract.
+To capture any additiona relevant articles which did not directly cite @mackenzie2003, a second query was generated searching articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract.
 As we were interested in how DOMs use has changed through time, we divided all articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024.
@@
 
-A total of 60 articles were included for this review.
-Based on the acceptance rates within each strata, an estimated 516 of the 1300 unreviewed articles in our sample would have met inclusion criteria (@fig-coverage).
+A total of 92 articles were included for this review.
+Based on the acceptance rates within each strata and quantity of unprocessed articles, an estimated 496 of the 1152 unreviewed articles in our sample would have met inclusion criteria (@fig-coverage).
 
-![Estimated coverage of the articles included in this review.
-A total of 60 articles from 2000-2023 were processed; bars indicating the estimated number of qualifying articles remaining are based on the rejection rate of each strata and the number of remaining unread articles in the pool.
-The proportion of articles included from each strata were as follows: 2004-2007, 12.5%; 2008-2011, 21.3%; 2012-2015, 44.4%; 2016-2019, 32.4%; 2020-2023, 57.1%.](Figures/CoveragePlot.jpeg){#fig-coverage}
+![Bars indicate the 92 articles included in our review as a proportion of the estimated number of published articles fitting DOMs, based on the qualification rate for articles in each strata.
+The proportion of articles included from each strata were as follows: 12.5% from 2005-2008; 24.4% from 2008-2011; 42.6% from 2012-2015; 35.1% from 2016-2019; and 57.1% from 2020-2023.](Figures/CoveragePlot.jpeg){#fig-coverage}
 
@@
 A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale.
-The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre [@basset2023], while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@clement2019].
+The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre, while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@basset2023; @clement2019].
 
 The temporal scale of studies shows similar variability.
-The duration between first and last survey ranging from under one month to forty years, with a median of 7 years.
+The duration between first and last survey ranging from under one month to forty years, with a median of 6 years.
 More meaningful for model interpretation, however, is the duration of the primary occasion, as this is presumed to be the temporal scale at which changes in occupancy are observed.
@@
 
-While most articles in our sample fit a model to a single species, 26 fit models to multiple species either as independent models or explicitly multi-species implementations.
-The multi-species models used various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects[@dorazio2010; @hendershot2020], as well as explicitly interactive models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019].
+While most articles in our sample fit a model to a single species, 41 fit models to multiple species either as independent models or explicitly multi-species implementations.
+The multi-species models use various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects[@dorazio2010; @hendershot2020], as well as explicit interaction models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019].
 While they did not fit multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012].
-Working with large numbers of taxa does raise additional challenges, as the level of complexity which can realistically be considered for each individual taxa is likely to be reduced for practicality's sake.
+Working with large numbers of taxa does raise additional challenges, as the level of complexity which can realistically be considered for each individual taxon is likely to be reduced for the sake of practicality.
 
@@
 \n B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles.
-38 articles fit models to only a single taxa, 20 fit independent models to multiple taxa, and 6 fit multi-species models with explicit interactions between taxa.
+55 articles fit models to only a single taxa, 30 fit independent models to multiple taxa, and 11 fit multi-species models with explicit interactions between taxa.
 \n C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors.
-10 articles incorporated data from citizen science projects.
-Project scale was also variable, with the median study running for 3.25 years and covering 98 sites.](Figures/ReviewBox.png){#fig-StudyDetails}
+13 articles incorporated data from citizen science projects.
+Project scale was also variable, with the median study running for 6 years and covering 101 sites.](Figures/ReviewBox.png){#fig-StudyDetails}
 
 The presence/absence data required for DOMs has come from a breadth of different sources and detection methods.
-In our sample 76% of articles used human observational data, 19% used live-trapping methods, and 5% used detections from camera trapping or acoustic monitoring.
-Several articles also use citizen-science data to build detection histories, including @zuckerberg2011's model using bird-feeder data and @warrier2020's using citizen reports of tiger sign.
+In our sample 79% of articles used human observational data, 9% used live-trapping methods, and 14% used detections from camera traps.
+Several articles also use citizen-science data to build detection histories, including @zuckerberg2011's model using bird-feeder data and @warrier2020's using citizen reports of tiger signs.
 Using citizen-science data may require additional consideration of assumptions (particularly false-positives), as discussed in greater detail by @cruickshank2019.
 Within these broad categories of detection method there is additional diversity, with several studies using unique survey methods determining the presence of a species at a location.
-For example, in our only application of a DOM for a marine species @pendleton2022 uses aerial transects broken up into grid cells to observe whale occupancy; and @marescot2020's fits a unique 'multi-species' treating poachers as species which uses ranger reports to create detection histories.
+For example, in the only application of a DOM for a marine species in our sample, @pendleton2022 uses aerial transects broken up into grid cells to observe whale occupancy; and @marescot2020's fits a unique 'multi-species' model treating poachers and using ranger reports to create detection histories.
 
-Notably, many of these studies use data which was not originally collected in a robust design framework for occupancy modelling.
-In these articles, authors manipulated their data into a hierarchical format post-hoc using a variety of methods.
+Notably, many of these studies use data which were not originally collected in a robust design framework for occupancy modelling.
+In these articles, authors formatted their data into a hierarchical format post-hoc using a variety of methods.
 Some defined primary seasons as a discrete time interval, treating all surveys occurring within the season as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate.
@@
 Additionally, not all articles rely on a single source of detection data - some integrate multiple sources of data to maximise sample size, combining data from camera traps, sign surveys, and citizen science reports.
-These integrated detection method models do require additional care in estimating detection probability, which is generally expected to vary based on observation method; however, this can be reasonably achieved by including survey method as a covariate on detection probability [@warrier2020; @pitman2017].
+These integrated detection method models do require additional care and consideration; users must ensure that different detection methods represent comparable spatial and temporal scales, and that any variation in perceptibility between detection methods is accounted for [@warrier2020; @pitman2017].
 A special case exists when different detection methods are used where one has the potential for false positive detections; e.g., where less-certain citizen science detections are combined with certain detections from field surveys.
@@
 Users of DOMs must carefully consider precisely what they are modelling and address questions on the scale represented by their model [@chave2013].
-Drivers of occupancy may differ depending on whether a site is represented by a single point on the landscape or as a grid cell, with the former likely to depend on more local, small scale factors rather than landscape-level trends [@stevens2019].
+Intepretation and drivers of occupancy may differ depending on whether a site is represented by a single point on the landscape or as a grid cell, with the former likely to depend on more local, small scale factors rather than landscape-level trends [@stevens2019].
 This is also true for the temporal scale of occupancy: whether a site is occupied within a week or within a year leads to vastly different conceptions of occupancy.
@@
 In this review, we have recorded all covariates considered for each model in our sample.
-These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of articles considering each type of covariate in their models, the means by which that covariate data is collected, and the ways in which those covariates are statistically represented in DOMs.
-We further categorise these covariates into two groups: environmental covariates, which represent plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environmental itself.
-Our findings indicate that users of DOMs have incorporated an exceptionally wide diversity of different covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample.
-Many models incorporate covariates representing aspects of site geometry (31% of articles) and connectivity (26% of articles), adding spatial components to these DOMs which bear similarities to related stochastic patch occupancy models.
+These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of articles considering each type of covariate in their models, the means by which that covariate data are collected, and how covariates responses are represented in DOMs.
+We further categorise these covariates into two groups: environmental covariates, which represent plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environment.
+Our findings indicate that users of DOMs have incorporated an exceptionally wide diversity of covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample.
+Many models incorporate covariates representing aspects of site geometry and connectivity (37% of articles), adding spatial components to these DOMs which bear similarities to related stochastic patch occupancy models.
 Notable examples of these spatially-explicit models include @mortelliti2007's model fit to rodent data in patchy landscapes and @risk2011's hybrid DOM/incidence function model used to explore meta-population dynamics of rails.
 Several studies also include biotic interactions with other species as covariates, particularly those which describe taxa threatened by invasive species.
-Examples include models with barred owl presence as a covariate on spotted owl occupancy [@mangan2019] and models with bullfrog covariates on native amphibian occupancy [@rowe2019], each of which effectively incorporate species interactions without fitting explicitly multi-species models.
+Examples include models with barred owl presence as a covariate on spotted owl occupancy and models with bullfrog covariates on native amphibian occupancy, each of which effectively incorporate species interactions without fitting explicitly multi-species models [@mangan2019; @rowe2019].
 
@@
 covTable |>
-  filter(!Covariate %in% c("DIST", "PROT")) |>
   mutate(Summary = case_when(Covariate %in% c("Environmental", "Structural", "Overall") ~ TRUE,
@@
     "Environmental" ~ html("<b>Any Environmental</b>"),
-    "HABT" ~ html("<b>Habitat type (land)</b>\n<i>Aspects of terrestrial habitat</i>"),
-    "HABA" ~ html("<b>Habitat type (aquatic)</b>\n<i>Aspects of aquatic habitats</i>"),
+    "HABT" ~ html("<b>Habitat </b>\n<i>Aspects of habitat and land cover</i>"),
     "PHEN" ~ html("<b>Phenology</b>\n<i>Time-varying elements distinct from sampling occasions</i>"),
-    "GEOM" ~ html("<b>Geometry</b>\n<i>Site geometry, such as patch size or shape</i>"),
-    "CONN" ~ html("<b>Connectivity</b>\n<i>Connectivity to other sites<i/>"),
-    "LACO" ~ html("<b>Land cover</b>\n<i>Land cover classes (excluding agriculture)</i>"),
-    "OTHER" ~ html("<b>Other environmental</b>\n<i>Other environmental covariate not otherwise listed</i>"),
-    "TOPO" ~ html("<b>Topography</b>\n<i>Elements of landscape topography</i>"),
-    "CLWE" ~ html("<b>Climate/Weather</b>\n<i>Climate and weather covariates</i>"),
+    "SPAT" ~ html("<b>Spatial</b>\n<i>Site geometry, connectivity, or other spatial elements</i>"),
+    "OTHER-E" ~ html("<b>Other environmental</b>\n<i>Other environmental covariate not otherwise listed</i>"),
+    "TOPO" ~ html("<b>Topography</b>\n<i>Elements of landscape topography, including hydrology</i>"),
+    "CLWE" ~ html("<b>Climate/Weather</b>\n<i>Climate, weather, and natural disasters</i>"),
     "ANTH" ~ html("<b>Anthropogenic</b>\n<i>Relations to human activity</i>"),
-    "HYDR" ~ html("<b>Hydrology</b>\n<i>Hydrological features like rivers or lakes</i>"),
-    "SPAT" ~ html("<b>Spatial</b>\n<i>Spatial variation, such as latitude</i>"),
-    "AGRI" ~ html("<b>Agricultural</b>\n<i>Relating to agricultural features</i>"),
-    "DISA" ~ html("<b>Disturbance</b>\n<i>Natural disasters or disturbances</i>"),
     "BIOT" ~ html("<b>Biotic interaction</b>\n<i>Interactions with other (non-plant) species</i>"),
@@
     "SEAS" ~ html("<b>Primary occasion</b>\n<i>Effect of the primary occasion</i>"),
-    "OBSE" ~ html("<b>Observation</b>\n<i>Details of the survey</i>"),
+    "OBSE" ~ html("<b>Observation</b>\n<i>Details on the observation process</i>"),
     "SURV" ~ html("<b>Secondary occasion</b>\n<i>Effect of the secondary occasion</i>"),
@@
     "SITE" ~ html("<b>Site effect</b>\n<i>Site-level effects</i>"),
+    "OTHER-S" ~ html("<b>Other structural</b>\n<i>Other structural covariate not otherwise listed</i>"),
     "Overall" ~ html("<b>All covariates</b>"),
@@
 
-Covariate data for studies in our sample was either collected directly by researchers (19% of environmental covariate data), or derived from pre-existing remotely sensed datasets (81%); this of course varies depending on the category of covariate.
+Covariate data for studies in our sample was either collected directly by researchers (17% of environmental covariate data), or derived from pre-existing remotely sensed datasets (83%); this of course varies depending on the category of covariate.
 Directly collected data can generally be expected to represent finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely but can often be more proximal drivers of occupancy.
-These high quality covariates do come with trade-offs, as they can be expensive to collect and can often preclude projection to locations where data is unavailable.
+These high quality covariates do come with trade-offs, as they can be expensive to collect and can often preclude projection to locations where data are unavailable.
 For projects interested in making these projections, remotely-sensed covariates are generally more feasible despite their generally more distal nature [@austin2002].
-29% of environmental factors and 87% of structural factors included in reviewed articles were dynamic covariates which varied through time — this again varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static.
+42% of environmental factors and 87% of structural factors included in reviewed articles were dynamic covariates which varied through time — this again varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static.
 
-In the standard DOM covariates are incorporated for each parameter via a logit link, meaning that the statistical relationship between covariates and model parameters is represented as a linear term unless otherwise specified.
+In the standard DOM, covariates for each parameter are most commonly incorporated via a logistic regression (i.e., a linear regression through a logit link function).
+Statistical relationships between covariates and model parameters are represented as linear terms unless otherwise specified.
 Of course, many ecological relationships are non-linear and require more complex forms to be realistically represented in a model.
@@
 However, linear models are quite flexible, and non-linear responses can be easily accommodated using polynomial terms and interactions between covariates.
-Despite this, in our sample just 38% of articles considered using at least one non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear relationships.
-Interactions between covariates were similarly rare, with only 28% of studies considering at least one relationship between terms.
+Despite this, in our sample just 35% of articles considered using at least one non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear relationships.
+Interactions between covariates were similarly rare, with only 26% of studies considering at least one relationship between terms.
 The relatively low emphasis on more complex non-linear responses contrasts with other popular methods for modelling species occupancy.
@@
 ![The number of covariates considered for each parameter across all studies in our sample.
-'Occupancy' given here represents an alternative parameterisation of the DOM which jointly estimates Occupancy, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation.
+'Occupancy' given here represents an alternative parameterisation of the DOM which jointly estimates Occupancy for every season, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation.
 Here, a 'covariate' is defined as each distinct covariate considered for inclusion.
@@
 Identifying the 'best' model from a range of possible candidates can be a challenging process, especially for hierarchical models like DOMs [@doherty2012].
-Consider that the quantity of candidate models involving up to *n* covariates on *p* parameters can be given by the formula $(2^n)^p$; where a simple single-parameter may have have up to 64 possible combinations of 6 covariates, a 4-parameter dynamic occupancy model would have 16 million combinations.
+Consider that the quantity of candidate models involving up to *n* covariates on *p* parameters can be given by $(2^n)^p$; where a simple single-parameter may have up to 64 possible combinations of 6 covariates, a 4-parameter dynamic occupancy model would have 16 million combinations.
 This explosion of candidate models generally precludes exhaustive comparisons of possible models and requires a reduction of the models tested.
-Models in our sample include both Frequentist (n = 47) and Bayesian (n = 13), which differ considerably in their manner of implementation.
+Models in our sample include both Frequentist (n = 70) and Bayesian (n = 22), which differ considerably in their manner of implementation.
 A summary of modelling practices in DOMs is given in @tbl-modelling below.
@@
 
-The largest differences between frequentist and Bayesian models were in their approaches to model selection.
-Where 91% of frequentist models performed some manner of model selection to determine covariate inclusions for their models, only 31% of Bayesian models did so, with the majority instead fitting a single model defined *a priori*.
-Methods used for model selection varied considerably both within and between modelling frameworks.
-For frequentist models, the most popular and conventional approach to model selection as used in 49% of articles involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score.
-Most frequentist studies which did not follow this protocol instead used procedural model selection methods (40% of articles), where the structure for each model parameter is fit one-by-one.
+The largest differences between the frequentist and Bayesian models in our sample lie in their approaches to model selection.
+Where 94% of frequentist models perform some manner of model selection to determine covariate inclusions for their models, only 36% of Bayesian models do so, with the majority instead fitting a single model defined *a priori*.
+Methods used for model selection vary considerably both within and between modelling frameworks.
+For frequentist models, the most popular and conventional approach to model selection as used in 43% of articles involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score.
+Most frequentist studies which do not follow this protocol instead use procedural model selection methods (39% of articles), where the structure for each model parameter is fit one-by-one.
 For example, this protocol might first identify the best structure for detection probability before moving on to initial occupancy and so on.
-A smaller portion of frequentist studies (4%) used more unique approaches, fitting simpler models such as single season occupancy models for candidate covariates to identify the most informative terms to use in a DOM.
-Across all frequentist implementations, 70% of articles conducted multi-model inference and model-averaging using AIC weights [@burnham2004].
+A smaller portion of frequentist studies (9%) use more unique approaches, fitting simpler models such as single season occupancy models for candidate covariates to identify the most informative terms to use in a DOM.
+Across all frequentist implementations, 71% of articles conduct multi-model inference and model-averaging using AIC weights [@burnham2004].
 
-Those Bayesian models which *did* perform model selection took distinctly different approaches, with largely idiosyncratic methods across these studies.
+Those Bayesian models which *do* perform model selection tend to take distinctly different approaches, with largely idiosyncratic methods across these studies.
 While direct comparison of model fit is rare, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data.
 Another approach used by @cook2022 fit a global model including all covariates before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the model.
-@ahumada2013 takes a particularly interesting hybrid approach, in which model selection is conducted by a procedural method in the frequentist framework before refitting the best structure in a Bayesian model.
+@ahumada2013 takes a hybrid approach, in which model selection is conducted by a procedural method in the frequentist framework before refitting the best structure in a Bayesian model.
 
 Limited research has been conducted on advantages of different methods for covariate selection in DOMs, and there is surely no one-size-fit all approach which will be appropriate for all possible use-cases.
-However, it is important to consider the implications of the different model selection approaches in common usage, and research from the SDM literature highlights how covariate selection can influence our interpretation of model outputs @brodie2020.
-Within the occupancy modelling literature, @stewart2023's article on covariate selection in single-season occupancy model discusses important attributes of the information-criteria approaches widely used in frequentist models in our review, noting that these approaches can make inaccurate estimates and influence the inference which we make from our models.
+However, it is important to consider the implications of the different model selection approaches in common usage, and research from the SDM literature highlights how covariate selection can influence our interpretation of model outputs [@brodie2020].
+Within the occupancy modelling literature, @stewart2023's article on covariate selection in single-season occupancy model discusses important attributes of the information-criteria approaches widely used in frequentist models in our review, noting that these approaches can lead to inaccurate estimates and influence model inference.
 @morin2020's work raises different concerns on procedural model selection methods, demonstrating how the fine details of modelling protocols can determine whether the best-fitting possible model is identified and which covariates appear in final models.
@@
 There is no broadly-accepted goodness-of-fit test for dynamic occupancy models, although @mackenzie2004 describes an approach for single-season occupancy models using a parametric bootstrap which can been extended to DOMs; this test is implemented in the *AICcModAvg* and *unmarked* R packages.
-While these tests may not suit Bayesian implementations, @broms2016 discusses possibilities for model evaluation in that context; their article focuses on single-season multi-species occupancy models but extensions of their approach may also be applicable to the DOM.
-Model evaluation for DOMs in general can be difficult and somewhat uncertain compared to other models, as the primary response variable of interest (species occupancy) is a latent variable where the true state is generally not known.
-Predictive performance evaluation is thus typically based on *observed* occupancy, where a DOM is used to simulate detection histories to be compared with field survey results.
+@kéry2021 describe the test and present an alternative based on separately assessing fit to static and dynamic components of the model.
+In Bayesian implementations, posterior predictive checks offer means to assess model fit [@gelman2014].
+@broms2016 discusses possibilities for model evaluation in the Bayesian context focusing on single-season multi-species occupancy models but extensions of their approach may also be applicable to the DOM.
+As in other hierarchical models, model evaluation for DOMs can in general can be difficult and somewhat uncertain compared to other model types, as the primary response variable of interest (species occupancy) is a latent variable where the true state is generally not known.
+Predictive performance evaluation is thus typically based on *observed* occupancy data, where a DOM is used to simulate detection histories to be compared with field survey results.
 
@@
 
-Many articles in our sample used the DOM and sought to monitor trends in occupancy state through time.
+Many articles in our sample used the DOM seeking to monitor trends in occupancy state through time.
 These studies include @ahumada2013's paper assessing populations trends of 13 forest mammals and @scott2015's assessment of the conservation status of the endangered flattened musk turtle.
-These studies offer valuable insights on the state of species across landscapes and through time, and DOMs are a strong candidate for many monitoring oriented projects — including those where many species need to be assessed at once.
+These studies offer valuable insights on the state of species across landscapes and through time, demonstrating DOMs' suitability for many monitoring oriented projects — including those where many species need to be assessed at once.
 Other articles were more focused on examining the factors which influence species occupancy, either testing a pre-specified hypothetical relationships or taking a broader tack to identify drivers of occupancy without preconception.
@@
 
-These priorities are particularly important given the frequent applied objectives of DOM users, tackling challenges which include assessing critically endangered species @carvalho2023, guiding public health management of disease vectors @moreira2016, and tracking rapidly developing biological invasions @wood2020.
+These priorities are particularly important given the frequent applied objectives of DOM users, tackling challenges which include assessing critically endangered species [@carvalho2023], guiding public health management of disease vectors [@moreira2016], and tracking rapidly developing biological invasions [@wood2020].
 DOMs are well suited to these situations, and it is to be expected that as these types of applications are more commonly attempted understanding the sensitivity of model outputs to decisions made in the model fitting process becomes increasingly important.

### 2024-09-10 _manuscript/Manuscript.qmd
Manuscript rendering

@@
 These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.
+
 bibliography: references.bib
+csl: apa.csl
 ---
@@
 In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy.
-Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@mackenzie2005].
+Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@guillera-arroita2017].
+It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this allows for considerably greater flexibility in data inputs.
 
@@
 To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data.
-A pool of candidate articles was generated using two queries on Web of Science: the first of these included all articles from 2004-2023 which cite @mackenzie2003; the second query searched articles in the same time-span matching "dynamic occupancy model\*", "multi-season occupancy model\*", and "occupancy dynamic\*", articles including each of "occupancy", "colonization", "extinction/persistence", and "detection", and articles with the term "occupancy" located near "dynamic" in the title, key terms, or abstract.
-As we were interested in how DOMs use has changed through time, we divided these articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024.
+A pool of candidate articles was generated using two queries on Web of Science.
+The first of these included all articles from 2004-2023 which cite @mackenzie2003.
+To capture any additiona relevant articles which did not directly cite @mackenzie2003, a second query was generated searching articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract.
+As we were interested in how DOMs use has changed through time, we divided all articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024.
 From each of these strata we randomly selected 20 articles for inclusion in the review.

### 2024-09-10 Manuscript.qmd
testing
Minor edits to manuscript, fiddling with pdf options

@@
 csl: apa.csl
-embed-resources: true
 ---
@@
 An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing can be applied in many ways.
-Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, and identifying potential drivers of occurrence can help inform potential management actions.
+Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, while identifying potential drivers of occurrence can help inform potential management actions.
 Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios.
@@
 Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010].
-These conditions are prominent during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].
+These conditions often occur during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].
 
@@
 Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time.
-This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes which drive occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic data required by more process-explicit models [@briscoe2019].
+This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes which drive occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic or abundance data required by more process-explicit models [@briscoe2019].
 These features make the DOM an important tool, with uses for tasks including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].
@@
 The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres.
-\n B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles.
+B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles.
 55 articles fit models to only a single taxa, 30 fit independent models to multiple taxa, and 11 fit multi-species models with explicit interactions between taxa.
-\n C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors.
+C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors.
 13 articles incorporated data from citizen science projects.
@@
     style = cell_text(style = "italic")
-  ) |>
-  as_raw_html()
+  ) 
+#|>
+ # as_raw_html()
 ```

### 2024-09-15 Manuscript.qmd
Addressed introduction, methods, and some result comments

@@
 The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849].
-An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing can be applied in many ways.
+An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing is applicable in many ways.
 Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, while identifying potential drivers of occurrence can help inform potential management actions.
@@
 The need to understand and describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly used to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales.
-However, several factors can make occupancy a difficult quantity to estimate.
-For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with most wildlife field data.
-Generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004].
+However, several factors can make occupancy difficult to estimate.
+For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with much wildlife field data — generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004].
 Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014].
-Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010].
+Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species' occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010].
 These conditions often occur during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].
 
-The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017].
+The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017a] .
 Drawing on principles from the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy.
@@
 This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes which drive occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic or abundance data required by more process-explicit models [@briscoe2019].
-These features make the DOM an important tool, with uses for tasks including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].
+These features make the DOM an important tool, with uses including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].
 
@@
 The DOM structure encompasses two processes: the ecological process of site occupancy dynamics describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed at a site where it is present (@fig-modelform).
-Sites exist in either an occupied or unoccupied state.
-In the first time step, occupancy state at each site is determined by the probability of initial occupancy.
-In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation and local extinction.
+Sites exist in either an occupied (z = 1) or unoccupied state (z = 0).
+In the first time step, occupancy state at each site is determined by the probability of initial occupancy ψ~1~.
+In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation γ and local extinction ε (@eq-occupancy).
+
+$$
+\begin{align*}
+z_1 &\sim Bernoulli(\psi_1)\\\\
+z_t &\sim Bernoulli(\psi)\\
+&\psi = z_{t-1}(1-\epsilon)+(1-z_{t-1})\gamma
+\end{align*}
+$$ {#eq-occupancy}
+
 In the observation component of the DOM, the model explicitly accounts for imperfect detection.
-At occupied sites, the detection probability describes whether or not the species is observed during a survey; under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur.
+At occupied sites, the detection probability ρ describes whether the species is observed during a survey (y = 1) or not (y = 0).
+Under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur (@eq-detection).
+
+$$
+y_{t,j} \sim Bernouilli(z_t\rho)
+$$ {#eq-detection}
 
@@
 In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy.
-Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@guillera-arroita2017].
-It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this allows for considerably greater flexibility in data inputs.
+Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@guillera-arroita2017a; @guillera-arroita2017b].
+It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this enables flexibility in data inputs.
+In this review we use the term "parameter" to refer to the quantities being estimate: in most cases, these are initial occupancy ψ~1~, colonisation γ, extinction ε, and detection ρ (see @fig-modelform, @eq-occupancy, and @eq-detection), though some alternative model forms may directly estimate occupancy ψ as a parameter.
 
@@
 
-DOMs make a number of key assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.
+DOMs make several key assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.
 
@@
 While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods.
-@mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need for authors to consider how certain their detections truly are.
+@mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need to consider how certain detections truly are for any given study system.
 Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019].
-Where this assumption can not reasonably met, model extensions designed to account for false positive error should be considered.
+Where this assumption can not reasonably be met, model extensions designed to account for false positive error should be considered [@royle2006; @miller2011; @miller2015].
 
 II.
-**Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion around the bias which is introduced when it is violated [@rota2009; @otto2013].
+**Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion in terms of the bias which is introduced when it is violated [@rota2009; @otto2013].
 Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discuss the difficultly in distinguishing between temporary emigration and local extinction.
 Model extensions to relax the closure assumption have been developed, including @kendall2013's approach using staggered arrival and departure periods between sites.
-A more pertinent approach for most studies, however, is careful consideration of ecologically relevant seasons corresponding to an appropriate definition of occupancy.
+A more pertinent approach for most studies, however, is careful consideration of the appropriate definition of occupancy under the survey design used [@kéry2021].
 
@@
 Generally, this is achieved by allowing model parameters to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters.
-An important element of this assumption is that the likelihood of detection of a species depends not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017].
+An important element of this assumption is that the likelihood of detection of a species depends not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017a].
 While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates.
@@
 
-Since its original description in @mackenzie2003, the DOM has been further developed with numerous model extensions and alternative formulations including implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010].
-For a comprehensive discussion of the most common extensions and their applications see @bailey2014 and @guillera-arroita2017, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.
+Since its original description in @mackenzie2003, the DOM has been further developed with numerous model extensions and alternative formulations including aforementioned implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010].
+For a comprehensive discussion of the most common extensions and their applications see @bailey2014 and @guillera-arroita2017a, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.
 
@@
 The first of these included all articles from 2004-2023 which cite @mackenzie2003.
-To capture any additiona relevant articles which did not directly cite @mackenzie2003, a second query was generated searching articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract.
+To capture any additional relevant articles that did not directly cite @mackenzie2003, a second query was generated searching articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract.
 As we were interested in how DOMs use has changed through time, we divided all articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024.
 From each of these strata we randomly selected 20 articles for inclusion in the review.
-Articles which did not meet inclusion criteria were replaced from within their own strata.
+Articles that did not meet inclusion criteria were replaced from within their own strata.
 
-As our review focuses only on applications of the dynamic multi-season occupancy model of @mackenzie2003 and its extensions, we included articles which fit a model meeting the following criteria: i) uses non-simulated, field-collected, presence-absence data; ii) relies on data from multiple sites which can exist in at least two states, including an occupied and unoccupied state; iii) has multiple seasons, between which sites may change states conditional on the prior season's occupancy state and transition probabilities such as colonisation and extinction; and iv) contains at least one parameter describing the detection process.
+As our review focuses only on applications of the dynamic multi-season occupancy model of @mackenzie2003 and its extensions, we included articles that fit a model meeting the following criteria: i) uses non-simulated, field-collected, presence-absence data; ii) relies on data from multiple sites which can exist in at least two states, including an occupied and unoccupied state; iii) has multiple seasons, between which sites may change states conditional on the prior season's occupancy state and transition probabilities such as colonisation and extinction; and iv) contains at least one parameter describing the detection process.
 
@@
 To examine the reasons why authors used DOMs, we allocated each article to one or more category of objective based on the study's stated aims.
-The possible categories were *Observing trends*, where authors express interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; *Testing relationships*, where authors explore specific predefined hypotheses of relationships between covariates and model parameters; *Identifying drivers*, where authors attempt to find which covariates influence model parameters; *Predicting temporally*, where authors predict site occupancy under future conditions, *Predicting spatially*, where authors predict site occupancy to unsurveyed locations, and *Developing methods*, where authors introduce, test, or demonstrate aspects of dynamic occupancy models.
+The possible categories were *Estimating trends*, where authors express interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; *Testing relationships*, where authors explore specific predefined hypotheses of relationships between covariates and model parameters; *Identifying drivers*, where authors attempt to find which covariates influence model parameters; *Predicting temporally*, where authors predict site occupancy under future conditions, *Predicting spatially*, where authors predict site occupancy to unsurveyed locations, and *Developing methods*, where authors introduce, test, or demonstrate aspects of dynamic occupancy models.
 
@@
 Key traits of each covariate were recorded including their general category, whether they were directly observed or remotely sensed, whether they were static or varied between seasons, and how they were presented in the model (as a linear term, a polynomial term, or as part of an interaction with another covariate [@james2021]).
-Model selection procedures were also sorted into non-exclusive categories including *A priori*, where only one model was considered; *candidate suite*, where a predefined set of models was considered; *sequential,* where covariates were selected parameter-by-parameter; and *simple precursors*, where selection was preceded by another simpler model implementation.
-Finally, for each model we documented whether goodness-of-fit was tested and reported, as well as whether model performance was assessed by validation with either in or out-of-sample data.
+Model selection procedures were also sorted into non-exclusive categories including *A priori*, where only one model was considered; *candidate suite*, where a predefined set of models was considered; *sequential,* where covariates were selected parameter-by-parameter (e.g., fitting detection first, followed by initial occupancy and so on); and *simple precursors*, where selection was preceded by another simpler model implementation.
+Finally, for each model we documented whether goodness-of-fit was tested and reported, and whether model performance was assessed by validation with either in or out-of-sample data.
 For the full spreadsheet of data collected and further details on categorisation, see supplementary material.
@@
 A total of 92 articles were included for this review.
-Based on the acceptance rates within each strata and quantity of unprocessed articles, an estimated 496 of the 1152 unreviewed articles in our sample would have met inclusion criteria (@fig-coverage).
+Based on the acceptance rates within each stratum and quantity of unprocessed articles, an estimated 496 of the 1152 unreviewed articles in our sample would have met inclusion criteria (@fig-coverage).
 
@@
 Articles included in our review demonstrate considerable diversity in scope, scale, and objectives.
-A selection of key attributes of these studies is presented in @fig-StudyDetails, and we provide notable examples of DOMs throughout this section to highlight how some this variation appears in practice.
-Study systems included in our sample are globally distributed; while a majority use data collected in the United States of America, representatives are included from all geographic realms and an exceptional diversity of ecosystems.
+A selection of key attributes of these studies is presented in @fig-StudyDetails, and we provide notable examples of DOMs throughout this section to highlight how the noted variation appears in practice.
+Study systems included in our sample are globally distributed; while a majority use data collected in the United States of America, representatives are included from all geographic realms and a broad diversity of ecosystems.
 A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale.
@@
 The temporal scale of studies shows similar variability.
-The duration between first and last survey ranging from under one month to forty years, with a median of 6 years.
+The duration between first and last survey ranged from under one month to forty years, with a median of 6 years.
 More meaningful for model interpretation, however, is the duration of the primary occasion, as this is presumed to be the temporal scale at which changes in occupancy are observed.
@@
 
-DOMs have been used to study a variety of different species, though the majority of studies have been conducted on data for vertebrate taxa.
-They have been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community.
-However, there are exceptions — @belinchón2017 fits a DOM to lichen data, and @cook2022 uses them to model the spread of chronic wasting disease.
-The latter's application to disease dynamics is not unique, and DOMs have been touted as a valuable tool for modelling disease dynamics.
-@mores2020 and @padilla-torres2013 have each used DOMs to model mosquito dynamics as a human disease vector, and @bailey2014 discusses the concept of applying the DOM to the spread of chytrid fungus in amphibian habitats.
-The vast majority of applications have been fit for terrestrial species, though there have been a limited number of studies on aquatic systems, including @fisher2014's application to invasive salmon, @falke2012's on Great Plains stream fishes, and @pendleton2022's on whale occupancy dynamics.
+The DOM has been used to study a variety of different species, though the majority of studies have focussed on vertebrate taxa (@fig-StudyDetails B).
+The DOM has been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community.
+However, there are exceptions — @belinchón2017 fits a DOM to lichen data, and @cook2022 model the spread of chronic wasting disease.
+The latter's application to disease dynamics is not unique, and the DOM has been touted as a valuable tool for such applications.
+@mores2020 and @padilla-torres2013 have used the DOM to model mosquito dynamics as a human disease vector, and @bailey2014 discusses the concept of applying the DOM to the spread of chytrid fungus in amphibian habitats.
+The vast majority of studies model terrestrial species, though there have been a limited number of studies on aquatic systems, including invasive salmon [@fisher2014], Great Plains stream fishes [@falke2012], and whales [@pendleton2022].
 
@@
 The multi-species models use various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects[@dorazio2010; @hendershot2020], as well as explicit interaction models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019].
-While they did not fit multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012].
+Rather than using multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012].
 Working with large numbers of taxa does raise additional challenges, as the level of complexity which can realistically be considered for each individual taxon is likely to be reduced for the sake of practicality.

### 2024-09-17 _manuscript/Manuscript.qmd
Nearly finished text updates

@@
 The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849].
-An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing can be applied in many ways.
-Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, and identifying potential drivers of occurrence can help inform potential management actions.
+An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing is applicable in many ways.
+Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, while identifying potential drivers of occurrence can help inform potential management actions.
 Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios.
@@
 The need to understand and describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly used to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales.
-However, several factors can make occupancy a difficult quantity to estimate.
-For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with most wildlife field data.
-Generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004].
+However, several factors can make occupancy difficult to estimate.
+For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with much wildlife field data — generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004].
 Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014].
-Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010].
-These conditions are prominent during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].
+Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species' occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010].
+These conditions often occur during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].
 
-The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017].
+The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017a] .
 Drawing on principles from the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy.
 Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time.
-This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes which drive occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic data required by more process-explicit models [@briscoe2019].
-These features make the DOM an important tool, with uses for tasks including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].
+This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes which drive occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic or abundance data required by more process-explicit models [@briscoe2019].
+These features make the DOM an important tool, with uses including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].
 
@@
 The DOM structure encompasses two processes: the ecological process of site occupancy dynamics describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed at a site where it is present (@fig-modelform).
-Sites exist in either an occupied or unoccupied state.
-In the first time step, occupancy state at each site is determined by the probability of initial occupancy.
-In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation and local extinction.
+Sites exist in either an occupied (z = 1) or unoccupied state (z = 0).
+In the first time step, occupancy state at each site is determined by the probability of initial occupancy ψ~1~.
+In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation γ and local extinction ε (@eq-occupancy).
+
+$$
+\begin{align*}
+z_1 \sim& Bernoulli(\psi_1)\\\\
+z_t \sim& Bernoulli(\psi)\\\\
+&\psi = z_{t-1}(1-\epsilon)+(1-z_{t-1})\gamma
+\end{align*}
+$$ {#eq-occupancy}
+
 In the observation component of the DOM, the model explicitly accounts for imperfect detection.
-At occupied sites, the detection probability describes whether or not the species is observed during a survey; under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur.
+At occupied sites, the detection probability ρ describes whether the species is observed during a survey (y = 1) or not (y = 0).
+Under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur (@eq-detection).
+
+$$
+y_{t,j} \sim Bernouilli(z_t\rho)
+$$ {#eq-detection}
 
@@
 The multi-species models use various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects[@dorazio2010; @hendershot2020], as well as explicit interaction models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019].
-While they did not fit multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012].
-Working with large numbers of taxa does raise additional challenges, as the level of complexity which can realistically be considered for each individual taxon is likely to be reduced for the sake of practicality.
+Rather than using multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012].
+Working with large numbers of taxa does raise additional challenges, as the degree to which models are tailored to the taxa is likely to be limited for the sake of practicality.
 
@@
 The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres.
-\n B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles.
-55 articles fit models to only a single taxa, 30 fit independent models to multiple taxa, and 11 fit multi-species models with explicit interactions between taxa.
-\n C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors.
+B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles.
+55 articles fit models to only a single taxon, 30 fit independent models to multiple taxa, and 11 fit multi-species models with explicit interactions between taxa.
+C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as emerging methods like camera traps and acoustic monitors.
 13 articles incorporated data from citizen science projects.
@@
 
-The presence/absence data required for DOMs has come from a breadth of different sources and detection methods.
+The presence/absence data required for DOMs has come from a breadth of different sources and detection methods (@fig-StudyDetails, panel C).
 In our sample 79% of articles used human observational data, 9% used live-trapping methods, and 14% used detections from camera traps.
@@
 Within these broad categories of detection method there is additional diversity, with several studies using unique survey methods determining the presence of a species at a location.
-For example, in the only application of a DOM for a marine species in our sample, @pendleton2022 uses aerial transects broken up into grid cells to observe whale occupancy; and @marescot2020's fits a unique 'multi-species' model treating poachers and using ranger reports to create detection histories.
+For example, in the only application of a DOM for a marine species in our sample, @pendleton2022 use aerial transects broken up into grid cells to observe whale occupancy; and @marescot2020 fit a unique 'multi-species' model treating poachers as a taxon and using ranger reports to create detection histories.
 
@@
 Some defined primary seasons as a discrete time interval, treating all surveys occurring within the season as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate.
-These manipulations permit authors to use data which far predate the DOM, with one study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].
+These manipulations permit authors to use data which predate the DOM, with one study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].
 
@@
 A special case exists when different detection methods are used where one has the potential for false positive detections; e.g., where less-certain citizen science detections are combined with certain detections from field surveys.
-In this context, the certain detections are used to help account for false-positive detection probability, as in @miller2011's study integrating GPS collars and hunter reports to estimate wolf occupancy in Montana.
+In this context, the "certain" detections are used to help account for false-positive detection probability, as in @miller2011's study integrating GPS collars and hunter reports to estimate wolf occupancy in Montana.
 
@@
 Users of DOMs must carefully consider precisely what they are modelling and address questions on the scale represented by their model [@chave2013].
-Intepretation and drivers of occupancy may differ depending on whether a site is represented by a single point on the landscape or as a grid cell, with the former likely to depend on more local, small scale factors rather than landscape-level trends [@stevens2019].
+The interpretation of what 'occupancy' means and the factors which drive it may differ depending on the scale of what is considered a site, and how that relates to the ecology of a species.
 This is also true for the temporal scale of occupancy: whether a site is occupied within a week or within a year leads to vastly different conceptions of occupancy.
 This idea is particularly relevant in cases where the selection of season length *is* to some extent arbitrary, as with camera-trap or bioacoustic data where continuous recordings can be broken down into distinct 'seasons' of any length.
-DOMs are well suited to these data types [@balantic2019], and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data.
+The DOM may be readily applied to these data types [@balantic2019], and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data.
 For example, @kleiven2020 and @mölle2022 divide their camera trap records into seasons of just a few days.
-While this provides interesting insights of occupancy at extremely fine temporal scale, further research may be needed on how to determine appropriate season and survey durations with respect to research questions and the closure assumption.
+While this provides interesting insights of occupancy at extremely fine temporal scale, further research could help provide general guidance on determining appropriate season and survey durations with respect to research questions and the closure assumption.
 
@@
 In this review, we have recorded all covariates considered for each model in our sample.
-These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of articles considering each type of covariate in their models, the means by which that covariate data are collected, and how covariates responses are represented in DOMs.
-We further categorise these covariates into two groups: environmental covariates, which represent plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environment.
-Our findings indicate that users of DOMs have incorporated an exceptionally wide diversity of covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample.
+These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of articles considering each type of covariate in their models, the means by which covariate data are collected, and how covariates responses are represented in DOMs.
+We further delineate their use in the model by classifying covariates into two groups: environmental covariates, representing plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environment.
+Our findings indicate that users of DOMs have incorporated a wide diversity of covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample.
 Many models incorporate covariates representing aspects of site geometry and connectivity (37% of articles), adding spatial components to these DOMs which bear similarities to related stochastic patch occupancy models.
-Notable examples of these spatially-explicit models include @mortelliti2007's model fit to rodent data in patchy landscapes and @risk2011's hybrid DOM/incidence function model used to explore meta-population dynamics of rails.
-Several studies also include biotic interactions with other species as covariates, particularly those which describe taxa threatened by invasive species.
+Notable examples of these spatially-explicit models include @mortelliti2007's model fit to rodent data in patchy landscapes and @risk2011's hybrid DOM/incidence function model used to explore meta-population dynamics of wetland birds.
+Several studies also include biotic interactions with other species as covariates, particularly those which focussing on taxa threatened by invasive species.
 Examples include models with barred owl presence as a covariate on spotted owl occupancy and models with bullfrog covariates on native amphibian occupancy, each of which effectively incorporate species interactions without fitting explicitly multi-species models [@mangan2019; @rowe2019].
@@
 
-Covariate data for studies in our sample was either collected directly by researchers (17% of environmental covariate data), or derived from pre-existing remotely sensed datasets (83%); this of course varies depending on the category of covariate.
+Covariate data for studies in our sample was either collected directly by researchers (17% of environmental covariate data), or derived from pre-existing remotely sensed datasets (83%); this of course varies depending on the category of covariate (@tbl-covariates).
 Directly collected data can generally be expected to represent finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely but can often be more proximal drivers of occupancy.
-These high quality covariates do come with trade-offs, as they can be expensive to collect and can often preclude projection to locations where data are unavailable.
-For projects interested in making these projections, remotely-sensed covariates are generally more feasible despite their generally more distal nature [@austin2002].
-42% of environmental factors and 87% of structural factors included in reviewed articles were dynamic covariates which varied through time — this again varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static.
+These species-specific covariates do come with trade-offs, as they can be expensive to collect and can often preclude projection to locations where data are unavailable.
+For projects interested in making such projections, remotely-sensed covariates are generally more feasible despite their generally more distal nature [@austin2002].
+Forty two percent of environmental factors and 87% of structural factors included in reviewed articles were dynamic covariates which varied through time — this again varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static (@tbl-covariates).
 
 In the standard DOM, covariates for each parameter are most commonly incorporated via a logistic regression (i.e., a linear regression through a logit link function).
-Statistical relationships between covariates and model parameters are represented as linear terms unless otherwise specified.
+Statistical relationships between model parameters and covariates (e.g., between initial occupancy and its environmental covariates) are represented as linear terms unless more complexity is specified.
 Of course, many ecological relationships are non-linear and require more complex forms to be realistically represented in a model.
-@austin2007 discussed the importance of modelling ecologically-realistic responses to covariates, advocating for careful consideration of the most appropriate statistical form for hypothesised relationships.
-However, linear models are quite flexible, and non-linear responses can be easily accommodated using polynomial terms and interactions between covariates.
-Despite this, in our sample just 35% of articles considered using at least one non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear relationships.
-Interactions between covariates were similarly rare, with only 26% of studies considering at least one relationship between terms.
+@austin2007 discusses the importance of modelling ecologically realistic responses to covariates, advocating for careful consideration of the most appropriate statistical form for hypothesised relationships.
+Non-linear responses can be easily accommodated in the DOM by using polynomial transformations and interactions between covariates.
+Despite this, in our sample only 35% of articles tested one or more non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear terms (@tbl-covariates).
+Interactions between covariates were similarly rare, with only 26% of studies considering at least one interaction between terms (@tbl-covariates).
 The relatively low emphasis on more complex non-linear responses contrasts with other popular methods for modelling species occupancy.
-Many common approaches for SDMs, which include MAXENT and Boosted Regression Trees, permit considerable flexibility in the shape of their covariate response curves where supported by the data [@merow2013; @elith2008].
-This emphasis on more complex responses in SDMs may be due to their typically larger range-wide scale which encompasses the full species niche, where environmental relations may be expected to be non-linear.
-However, as previously indicated DOMs have been frequently implemented at similarly large spatial scales where the same assumptions should exist and similar responses might be expected.
-Similar concerns exist for the relatively low level of interactions between covariates: while they should not necessarily be used liberally, where these exist their exclusion can influence model performance [@guisan2006].
+Many common approaches for SDMs, which include MAXENT and Boosted Regression Trees, permit considerable flexibility in the shape of their covariate response curves and use of interactions where supported by the data [@merow2013; @elith2008].
+This emphasis on more complex responses in SDMs may be due to their frequent application across relatively large geographic extents which might encompass the full species niche, where environmental relationships may be expected to be non-linear.
+However, as previously indicated the DOM has also been implemented across large spatial extents where the same assumptions should exist and similar responses might be expected.
+Similar concerns exist for the relatively low level of interactions between covariates: these relationships are commonly expected based on ecological theory, and their exclusion can negatively impact model performance [@guisan2006].
 
-In addition to the types of covariates considered for modelling we also measured the size of the candidate covariate pool in each reviewed study, tallying the number of environmental and structure covariates which were available for use on each parameter ( @fig-covariates).
+In addition to the types of covariates considered for modelling we also recorded the size of the candidate covariate pool in each reviewed study, tallying the number of environmental and structural covariates which were available for use on each parameter ( @fig-covariates).
 This is an area of considerable variation in modelling practices — the number of covariates considered for each parameter ranges from 0 (effectively modelling the parameter as a constant) to over 40 possible factors.
@@
 Here, a 'covariate' is defined as each distinct covariate considered for inclusion.
-Linear and quadratic representations of the same covariate are counted as 1 covariate.](Figures/CovParamPlot.jpeg){#fig-covariates}
+Linear and quadratic representations of the same covariate are counted as one covariate.](Figures/CovParamPlot.jpeg){#fig-covariates}
 
@@
 The lack of any covariates considered for initial occupancy in many studies (37% of articles) is particularly notable — unless a study is conducted at very small extents or study sites are truly uniform in their suitability, one would expect some amount of non-random variation in occupancy probability across any study system that will not be captured when representing initial occupancy as a constant.
-Omission of the factors which drive occupancy is a known source of bias in static SDMs [@barry2006], which the initial occupancy component of the DOM conceptually resembles.
+Omission of the factors that drive occupancy is a known source of bias in static SDMs [@barry2006], which the initial occupancy component of the DOM conceptually resembles.
 Furthermore, any bias in occupancy estimation in the first time-step will perpetuate into future seasons due to the DOM's Markovian nature with implications for the reliability of model outputs.
-Detection probability is also typically represented with low quantities of environmental covariates.
+Detection probability is also typically represented with low numbers of environmental covariates (@tbl-covariates).
 Recall that detectability is dependent on not only observation factors, but also by drivers of abundance and species use; the low number of candidate covariates here relative to colonisation and extinction raises additional questions on whether this variation is always appropriately captured in models.
@@
 Identifying the 'best' model from a range of possible candidates can be a challenging process, especially for hierarchical models like DOMs [@doherty2012].
-Consider that the quantity of candidate models involving up to *n* covariates on *p* parameters can be given by $(2^n)^p$; where a simple single-parameter may have up to 64 possible combinations of 6 covariates, a 4-parameter dynamic occupancy model would have 16 million combinations.
+Consider that the quantity of candidate models involving up to *n* covariates on *p* parameters can be given by $(2^n)^p$; where a simple linear regression may have up to 64 possible combinations of 6 covariates, a 4-parameter dynamic occupancy model would have 16 million combinations.
 This explosion of candidate models generally precludes exhaustive comparisons of possible models and requires a reduction of the models tested.
-Models in our sample include both Frequentist (n = 70) and Bayesian (n = 22), which differ considerably in their manner of implementation.
-A summary of modelling practices in DOMs is given in @tbl-modelling below.
+Models in our sample include both Frequentist (n = 70) and Bayesian (n = 22) frameworks, which differ considerably in their manner of implementation.
+A summary of modelling practices in DOMs is given in @tbl-modelling.
 
@@
 Where 94% of frequentist models perform some manner of model selection to determine covariate inclusions for their models, only 36% of Bayesian models do so, with the majority instead fitting a single model defined *a priori*.
-Methods used for model selection vary considerably both within and between modelling frameworks.
+Methods used for model selection vary considerably both within and between the two modelling frameworks.
 For frequentist models, the most popular and conventional approach to model selection as used in 43% of articles involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score.
-Most frequentist studies which do not follow this protocol instead use procedural model selection methods (39% of articles), where the structure for each model parameter is fit one-by-one.
+The next most popular method in frequentist studies is to use procedural model selection methods (39% of articles), where the structure for each model parameter is fit one-by-one.
 For example, this protocol might first identify the best structure for detection probability before moving on to initial occupancy and so on.
-A smaller portion of frequentist studies (9%) use more unique approaches, fitting simpler models such as single season occupancy models for candidate covariates to identify the most informative terms to use in a DOM.
-Across all frequentist implementations, 71% of articles conduct multi-model inference and model-averaging using AIC weights [@burnham2004].
+The remainder of frequentist studies (9%) use a variety of approaches, such as fitting simpler models like single season occupancy models to identify the most informative terms to use in a DOM.
+Across all frequentist implementations, 71% of articles summaruise a final subset of models using multi-model inference and model-averaging with AIC weights [@burnham2004].
 
-Those Bayesian models which *do* perform model selection tend to take distinctly different approaches, with largely idiosyncratic methods across these studies.
-While direct comparison of model fit is rare, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data.
+Those Bayesian models which *do* perform model selection tend to take various approaches, with largely idiosyncratic methods across these studies.
+While direct comparison of model fit in Bayesian methods is rare, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data.
 Another approach used by @cook2022 fit a global model including all covariates before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the model.
@@
 
-Limited research has been conducted on advantages of different methods for covariate selection in DOMs, and there is surely no one-size-fit all approach which will be appropriate for all possible use-cases.
+Limited research has been conducted on advantages of different methods for covariate selection in DOMs, and there is unlikely to be a one-size-fits-all approach which will be appropriate for all possible use-cases.
 However, it is important to consider the implications of the different model selection approaches in common usage, and research from the SDM literature highlights how covariate selection can influence our interpretation of model outputs [@brodie2020].
-Within the occupancy modelling literature, @stewart2023's article on covariate selection in single-season occupancy model discusses important attributes of the information-criteria approaches widely used in frequentist models in our review, noting that these approaches can lead to inaccurate estimates and influence model inference.
-@morin2020's work raises different concerns on procedural model selection methods, demonstrating how the fine details of modelling protocols can determine whether the best-fitting possible model is identified and which covariates appear in final models.
+Within the occupancy modelling literature, @stewart2023's article on covariate selection in single-season occupancy models discusses important attributes of the information-criteria approaches widely used in frequentist models in our review, noting that these approaches can lead to inaccurate coefficient estimates that may influence model inference.
+@morin2020's work raises other concerns on procedural model selection methods, demonstrating how the fine details of modelling protocols can determine whether the best-fitting possible model is identified and which covariates appear in final models.
 The literature on performance of Bayesian model selection methods is more sparse, although @hooten2015's guide to Bayesian model selection in ecology is a valuable resource for possible methods of fitting those models.
 
-With few exceptions, most of the model selection methods identified in our review are designed only to identify the best model of those considered, and do not necessarily guarantee that selected models are well-fit to the data.
+The model selection methods identified in our review are designed only to identify the best model of those considered, and do not necessarily guarantee that selected models are well-fit to the data.
 Assessment of the performance of DOMs generally requires additional steps, and the best ways to achieve this are not necessarily clear.
@@
 
-Perhaps as a result of this uncertainty, assessment of model fit and model performance was rare amongst the articles in our sample.
+Perhaps as a result of these difficulties, assessment of model fit and model performance was rare amongst the articles in our sample.
 Only 19% of studies tested for goodness-of-fit, and just 10% calculated predictive performance.
-These rates are considerably lower than for models like SDMs; for which @araújo2019's review of applications found closer to 50% articles testing both fit and performance.
+These rates are considerably lower than those reported for SDMs; for which closer to 50% of articles were found to test both fit and performances [@araújo2019].
 
@@
 
-Studies in our sample were classified according to their objectives as stated by authors to explore the use-cases which DOMs have been applied for.
+Studies in our sample were classified according to their objectives as stated by authors to explore the use-cases which the DOM has been applied to (@fig-objectives).
 These objective categories were non-exclusive, and many studies fell into multiple categories; the frequency at which studies pursued each of these objectives is shown in @fig-objectives A.
-Across our study period DOMs have been frequently employed to observe trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions.
+Across our study period DOMs have been frequently employed to estimate trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions.
 Within each of these broad categories lies even more variation in objectives, emphasising the DOM's flexibility as a tool for making ecological inference in diverse contexts.
 
-Many articles in our sample used the DOM seeking to monitor trends in occupancy state through time.
+Many articles in our sample used the DOM seeking to monitor trends in occupancy state through time ("Estimate trends" in @fig-objectives).
 These studies include @ahumada2013's paper assessing populations trends of 13 forest mammals and @scott2015's assessment of the conservation status of the endangered flattened musk turtle.
-These studies offer valuable insights on the state of species across landscapes and through time, demonstrating DOMs' suitability for many monitoring oriented projects — including those where many species need to be assessed at once.
-Other articles were more focused on examining the factors which influence species occupancy, either testing a pre-specified hypothetical relationships or taking a broader tack to identify drivers of occupancy without preconception.
+These studies offer valuable insights on the state of species across landscapes and through time, demonstrating the DOM's suitability for monitoring oriented projects; including those where multiple species need to be assessed at once.
+Other articles were more focused on examining the factors which influence species occupancy, either testing pre-specified hypothetical relationships or taking a broader tack to identify drivers of occupancy without preconception.
 The distinction between the two is important, as it guides which covariates may be considered and how model selection may be used to fit models.
@@
 In recent years, increasing numbers of articles have used to DOM to generate predictions of species occupancy.
-As examples, @mcgowan2020 provides projections for the threatened Black rail under multiple management scenarios, and @pollentier2021 generates maps of distributions resembling those made with SDMs, outputs which can immediately be used by managers to inform action.
+As examples, @mcgowan2020 provides projections of future occupancy for the threatened Black rail under multiple management scenarios, and @pollentier2021 generates maps of distributions resembling those made with SDMs, outputs which can immediately be used by managers to inform action.
 

### 2024-09-20 Manuscript.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{sidenotes}{}{\usepackage{sidenotes}}
\@ifpackageloaded{marginnote}{}{\usepackage{marginnote}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Twenty years of dynamic occupancy models: a review of applications and a look towards the future},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Twenty years of dynamic occupancy models: a review of
applications and a look towards the future}
\author{Saoirse Kelleher \and Natalie Briscoe \and Gurutzeta
Guillera-Arroita \and Jane Elith}
\date{2024-09-10}

\begin{document}
\maketitle
\begin{abstract}
Describing patterns of species occupancy across landscapes and
throughout time is a key objective of much ecological research.
Nonetheless, reliably estimating occupancy can be a challenge,
particularly when common issues such as imperfect detection and shifting
distributions are at play. The dynamic, multi-season occupancy model
(DOM) offers a framework for occupancy estimation which explicitly
accounts for observation error, while also incorporating additional
mechanistic process by explicitly estimating colonisation and extinction
rates. In the two decades since its introduction, the DOM has become a
popular tool for describing patterns in occupancy, exploring the
environmental drivers of species occurence, and, more recently, for
predicting species occupancy to new locations and to future conditions.

In this review we examine how ecologists have applied the DOM in its
first two decades of use. Based on a sample of articles fitting DOMs to
field ecological data, we explore the systems in which these models have
been used, how data were collected for them, and how authors have built
their models to achieve their objectives. Our findings indicate that the
DOM is an particularly flexible tool readily applied to a range of
diverse taxa from around the globe, capable of being used to estimate
occupancy at study sites ranging from local to continental scales over
time periods from months to decades. The DOM framework is also amenable
to extension, further broadening their utility to address a wide range
of ecological questions.

A key aim of this review is to explore how users of DOMs have
incorporated covariates into their models to describe variation in
occupancy through space and time. Broadly speaking, model complexity in
DOMs tends to be low, with relatively few covariates considered for
inclusion and most relationships represented as simple linear terms.
Considerable variation also exists in the procedures used for covariate
selection, and limited research has been conducted on how these choices
may influence model performance and inference. Furthermore, only a
fraction of articles appear to evaluate models (either by
goodness-of-fit testing or prediction validation). Likewise, little
guidance exists on how to approach this task in DOMs. These
uncertainties in the modelling process should be key priorities for
future research on DOMs given their increasing use in applied ecological
research.
\end{abstract}

\section{Introduction}\label{introduction}

The description of patterns of species occupancy across landscapes has
been a long-standing subject of ecological research (Humboldt, 1849). An
understanding of where a species exists, which factors lead to its
presence, and how its distribution may be changing can be applied in
many ways. Descriptions of how widespread a species is and where it
occurs are the foundation of monitoring programs and important for
assessing conservation status, while identifying potential drivers of
occurrence can help inform potential management actions. Robust
knowledge of the occupancy patterns of a species also helps us to
predict where a species is most likely to occur, both under present
conditions and in hypothetical future scenarios.

While occupancy is a useful concept, it is also a challenging quantity
to describe, measure, and estimate. The need to understand and describe
species occupancy has led to the development of several popular
modelling approaches, including stochastic patch occupancy models
(Gutiérrez-Arellano et al., 2024) commonly used to study meta-population
dynamics, and species distribution models (SDMs, Franklin, 2010) widely
used to explore species occurrence at larger scales. However, several
factors can make occupancy a difficult quantity to estimate. For
instance, simple presence/absence data of species observations can be
biased when species are detected imperfectly, as is the case with most
wildlife field data. Generally speaking, it is impossible to determine
whether a location is truly unoccupied or whether the species occurs but
was not detected by the observer (Gu \& Swihart, 2004; Lahoz-Monfort et
al., 2014). Despite the ubiquity of imperfect detection in data
collection, many models fit to presence/absence data make no adjustment
for this source of bias (Kellner \& Swihart, 2014). Another challenge
for modelling occupancy is the difficulty in describing populations
under non-equilibrium conditions, where a species occurrence pattern and
relationship to its environment is in flux (Dormann, 2007; Elith et al.,
2010). These conditions often occur during biological invasions and
climate change driven range shifts, both conservation priorities and
increasingly common scenarios in the Anthropocene (Bertelsmeier et al.,
2013; Lenoir \& Svenning, 2015).

The site occupancy models first introduced by MacKenzie et al. (2002)
set the foundation for a powerful framework for modelling
presence/absence data while accounting for each of these challenges
(Guillera-Arroita \& Lahoz-Monfort, 2017). Drawing on principles from
the mark-recapture literature, occupancy models use multiple resurveys
of sites to estimate detection probabilities and correct for bias in
estimates of site occupancy. Originally a static model, MacKenzie et al.
(2003) extended this model for use in multiple time-steps by explicitly
describing the process of changing occupancy via colonisation and
extinction, relaxing assumptions of equilibrium and allowing description
of patterns of site occupancy through time. This dynamic occupancy model
(DOM) balances complexity and feasibility, explicitly describing the key
processes which drive occupancy dynamics while requiring reasonably
simple-to-collect presence absence/data instead of the detailed
demographic or abundance data required by more process-explicit models
(Briscoe et al., 2019). These features make the DOM an important tool,
with uses for tasks including hypothesis testing of relationships
between occupancy and the environment, explorations of the key drivers
of occupancy, and even prediction of occupancy under future conditions
(Briscoe et al., 2021; Kéry et al., 2013).

In this review we examine how dynamic occupancy models have been used by
ecologists in the two decades since their inception. Following an
introduction to the model's form and assumptions, we present the results
of a systematic review exploring how researchers have applied the DOM to
ecological data, with emphasis on how they collected their data,
selected covariates, and evaluated their models. Based on these results
we highlight the DOM's flexibility as a tool for understanding species
occurrence, examine approaches to the model building process, and
outline key priorities for future research involving this model class.

\subsubsection{Dynamic occupancy model form and
assumptions}\label{dynamic-occupancy-model-form-and-assumptions}

The DOM structure encompasses two processes: the ecological process of
site occupancy dynamics describing the presence or absence of a species
at a site at any point in time, and the observational process of
detection describing whether a species is observed at a site where it is
present (Figure~\ref{fig-modelform}). Sites exist in either an occupied
or unoccupied state. In the first time step, occupancy state at each
site is determined by the probability of initial occupancy. In following
time steps, occupancy state can change according to a Markovian process
where occupancy is predicated on the site's state in the prior time-step
and the probabilities of colonisation and local extinction. In the
observation component of the DOM, the model explicitly accounts for
imperfect detection. At occupied sites, the detection probability
describes whether or not the species is observed during a survey; under
the DOM's original parameterisation, it is assumed that false-positive
detections at unoccupied sites never occur.

\begin{figure}

\centering{

\includegraphics{Figures/ModelForm.png}

}

\caption{\label{fig-modelform}The form of the dynamic occupancy model as
described by MacKenzie et al. (2003). The ecological process submodel
describes changes in occupancy over time, where occupancy in the first
season is given by the probability of initial occupancy and occupancy in
ensuing seasons is governed by colonisation and extinction
probabilities. The observational process submodel describes
detectability, where the probability of detecting a species given that a
site is occupied is given by the detection probability. At sites which
are truly unoccupied, it is assumed that the species is never detected
(i.e., no false positives).}

\end{figure}%

To disentangle the ecological and observational processes the DOM
requires a hierarchical sampling design as depicted in
Figure~\ref{fig-surveys}. Under this design, observations at a site
occur during distinct, time-bound seasons within which it is assumed
site occupancy does not change (that is, sites are closed to changes in
occupancy). In each season multiple observations are conducted,
permitting estimation of the probability of detection conditional on
occupancy. Most frequently these repeat observations are collected by
revisiting the site on separate occasions, although they can also be
attained by alternative means: examples include conducting surveys at
multiple locations within a site, using multiple observers during a
survey, or recording the time elapsed until a detection is recorded
(Guillera-Arroita \& Lahoz-Monfort, 2017). It is important to note that
it is not necessary for the same number of observations to occur in each
year or for each site, and that the DOM accommodates for missing data;
this allows for considerably greater flexibility in data inputs.

\begin{figure}

\centering{

\includegraphics{Figures/SurveyDesign.png}

}

\caption{\label{fig-surveys}The sampling design of the standard dynamic
occupancy model. During seasons, also called primary occasions, sites
are considered closed to changes in true occupancy state; occupancy
state may only change between seasons. Within each season, multiple
observations (`secondary occasions') are conducted to record the
observed presence or absence of the species at each site. These multiple
observations may be recorded in many ways: sites can be revisited
multiple times within a season, surveys can be conducted at multiple
points within a larger site, multiple observers can conduct surveys
contemporaneously, or the time elapsed prior to a detection occurring
can be recorded. Note that it is not necessary for each site or season
to have the same number of observations, and that missing data can be
accommodated.}

\end{figure}%

DOMs make a number of key assumptions requiring careful consideration by
model users, which we outline here as our review interrogates related
aspects of model building.

\begin{enumerate}
\def\labelenumi{\Roman{enumi}.}
\item
  \textbf{False positive detections do not occur}. While this assumption
  can be safely met in many studies, it is not necessarily guaranteed
  when working with more cryptic species or less reliable survey
  methods. McClintock et al. (2010) and D. A. W. Miller et al. (2015)
  comment on the bias induced when false positives occur and are not
  accounted for, highlighting the need for authors to consider how
  certain their detections truly are. Significantly, even genuine
  detections of a species can be considered `false positives' when they
  do not represent the intended definition of occupancy, such as
  detections of transient individuals when the intent is to estimate
  breeding occupancy (Berigan et al., 2019). Where this assumption can
  not reasonably met, model extensions designed to account for false
  positive error should be considered.
\item
  \textbf{Sites are closed to occupancy between seasons.} This
  requirement, best known as the `closure assumption,' has also been
  subject to considerable discussion around the bias which is introduced
  when it is violated (Otto et al., 2013; Rota et al., 2009). Closure is
  dependent not only on the life history of the species, but also on the
  definition of occupancy used by researchers --- short seasons may
  represent dynamics more representative of species `use,' and Valente
  et al. (2017) discuss the difficultly in distinguishing between
  temporary emigration and local extinction. Model extensions to relax
  the closure assumption have been developed, including Kendall et al.
  (2013)'s approach using staggered arrival and departure periods
  between sites. A more pertinent approach for most studies, however, is
  careful consideration of ecologically relevant seasons corresponding
  to an appropriate definition of occupancy.
\item
  \textbf{Heterogeneity in occupancy and detection is accounted for.} As
  with any approach for modelling species occurrence, it is assumed that
  DOMs appropriately capture variation in occupancy patterns and species
  detectability across the study system. Generally, this is achieved by
  allowing model parameters to vary with respect to covariates
  representing the environmental factors which may be expected to
  influence these parameters. An important element of this assumption is
  that the likelihood of detection of a species depends not only on the
  observability of the species, but also on factors like habitat
  suitability which influence species abundance and activity
  (Guillera-Arroita \& Lahoz-Monfort, 2017). While no model will fully
  account for the complexity inherent in patterns of species occupancy
  and detection, failure to capture key drivers is likely to introduce
  bias and confound inference made from model estimates. Compared to the
  first two assumptions mentioned, this aspect of DOMs has been less
  thoroughly discussed and comparatively little is known about how this
  latent heterogeneity can influence model performance.
\end{enumerate}

Since its original description in MacKenzie et al. (2003), the DOM has
been further developed with numerous model extensions and alternative
formulations including implementations accounting for false positives
(D. Miller et al., 2011; D. A. W. Miller et al., 2015; Royle \& Link,
2006), multiple states beyond occupied and unoccupied (Nichols et al.,
2007), and jointly estimated multi-species models (Dorazio et al.,
2010). For a comprehensive discussion of the most common extensions and
their applications see Bailey et al. (2014) and Guillera-Arroita \&
Lahoz-Monfort (2017), as well as Devarajan et al. (2020) for a more
detailed review of multi-species occupancy models.

\section{Systematic review methods}\label{systematic-review-methods}

To assess how DOMs have been applied in the years since their
introduction we gathered a representative sample of articles fitting
them to field ecological data. A pool of candidate articles was
generated using two queries on Web of Science. The first of these
included all articles from 2004-2023 which cite MacKenzie et al. (2003).
To capture any additiona relevant articles which did not directly cite
MacKenzie et al. (2003), a second query was generated searching articles
in the same time-span matching the terms ``\emph{dynamic occupancy
model*}'', ``\emph{multi-season occupancy model*}'', or
``\emph{occupancy dynamic*}''; articles including each of
``\emph{occupancy}'', ``\emph{colonization}'',
``\emph{extinction/persistence}'', and ``\emph{detection}''; and
articles with the term ``\emph{occupancy}'' located near
``\emph{dynamic}'' in the title, key terms, or abstract. As we were
interested in how DOMs use has changed through time, we divided all
articles across four-year-long strata spanning 2004-2007, 2008-2011,
2012-2015, 2016-2019, and 2020-2024. From each of these strata we
randomly selected 20 articles for inclusion in the review. Articles
which did not meet inclusion criteria were replaced from within their
own strata.

As our review focuses only on applications of the dynamic multi-season
occupancy model of MacKenzie et al. (2003) and its extensions, we
included articles which fit a model meeting the following criteria: i)
uses non-simulated, field-collected, presence-absence data; ii) relies
on data from multiple sites which can exist in at least two states,
including an occupied and unoccupied state; iii) has multiple seasons,
between which sites may change states conditional on the prior season's
occupancy state and transition probabilities such as colonisation and
extinction; and iv) contains at least one parameter describing the
detection process.

For each article we recorded key details on authorship, research
objectives, study taxa and system, survey methods, and modelling
approach. To examine the reasons why authors used DOMs, we allocated
each article to one or more category of objective based on the study's
stated aims. The possible categories were \emph{Observing trends}, where
authors express interest in estimates of site occupancy, colonisation,
extinction, or detection probabilities; \emph{Testing relationships},
where authors explore specific predefined hypotheses of relationships
between covariates and model parameters; \emph{Identifying drivers},
where authors attempt to find which covariates influence model
parameters; \emph{Predicting temporally}, where authors predict site
occupancy under future conditions, \emph{Predicting spatially}, where
authors predict site occupancy to unsurveyed locations, and
\emph{Developing methods}, where authors introduce, test, or demonstrate
aspects of dynamic occupancy models.

We recorded details on the type of organisms (bird, mammal, etc.)
modelled in each article, and the means by which multiple species were
modelled where applicable. Taxa were denoted as threatened either when
they are listed on the IUCN Red List of Species or when authors indicate
that they are threatened. This deference to authors' representation of
conservation status was made to account for sub-species which lack
listings or species which are of more local concern. Study locality and
size was documented; the size of the study area being defined as the
intended area of inference which contained all sites --- this was
measured to the order of magnitude to account for uncertainty in
reporting.

We were particularly interested in how authors navigated the modelling
process, from covariate selection through model evaluation. To this end,
we recorded all covariates considered in each study regardless of
whether they were or were not included in final models. Key traits of
each covariate were recorded including their general category, whether
they were directly observed or remotely sensed, whether they were static
or varied between seasons, and how they were presented in the model (as
a linear term, a polynomial term, or as part of an interaction with
another covariate (James et al., 2021)). Model selection procedures were
also sorted into non-exclusive categories including \emph{A priori},
where only one model was considered; \emph{candidate suite}, where a
predefined set of models was considered; \emph{sequential,} where
covariates were selected parameter-by-parameter; and \emph{simple
precursors}, where selection was preceded by another simpler model
implementation. Finally, for each model we documented whether
goodness-of-fit was tested and reported, as well as whether model
performance was assessed by validation with either in or out-of-sample
data. For the full spreadsheet of data collected and further details on
categorisation, see supplementary material.

\section{Applications of dynamic occupancy
models}\label{applications-of-dynamic-occupancy-models}

A total of 92 articles were included for this review. Based on the
acceptance rates within each strata and quantity of unprocessed
articles, an estimated 496 of the 1152 unreviewed articles in our sample
would have met inclusion criteria (Figure~\ref{fig-coverage}).

\begin{figure}

\centering{

\includegraphics{Figures/CoveragePlot.jpeg}

}

\caption{\label{fig-coverage}Bars indicate the 92 articles included in
our review as a proportion of the estimated number of published articles
fitting DOMs, based on the qualification rate for articles in each
strata. The proportion of articles included from each strata were as
follows: 12.5\% from 2005-2008; 24.4\% from 2008-2011; 42.6\% from
2012-2015; 35.1\% from 2016-2019; and 57.1\% from 2020-2023.}

\end{figure}%

Articles included in our review demonstrate considerable diversity in
scope, scale, and objectives. A selection of key attributes of these
studies is presented in Figure~\ref{fig-StudyDetails}, and we provide
notable examples of DOMs throughout this section to highlight how some
this variation appears in practice. Study systems included in our sample
are globally distributed; while a majority use data collected in the
United States of America, representatives are included from all
geographic realms and an exceptional diversity of ecosystems. A
particularly notable aspect of these study areas is their variation in
size, which ranges from the hyper-local to the continental scale. The
smallest study locality included in our sample studied insect occurrence
in a rainforest plot less than one square kilometre, while the largest
analysis modelled avian range shifts across the entire eastern half of
the United States (Basset et al., 2023; Clement et al., 2019).

The temporal scale of studies shows similar variability. The duration
between first and last survey ranging from under one month to forty
years, with a median of 6 years. More meaningful for model
interpretation, however, is the duration of the primary occasion, as
this is presumed to be the temporal scale at which changes in occupancy
are observed. Most applications describe the primary occasion as a year,
although some studies divide years into multiple primary occasions to
describe seasonal variation in occupancy. In the most extreme cases the
primary occasion may be as brief as a week, capturing much finer scale
changes in occupancy. These shortened-season occupancy models are most
common with camera trapping or acoustic monitoring data, which can be
arbitrarily divided into primary and secondary sampling occasions. This
is seen in Kleiven et al. (2020)'s study of rodent and mustelid
interactions using camera trap data, where the primary occasions are
less than one week in length. On the other end of the spectrum, some
studies modelled primary occasions which were decades apart and
represented generational changes in occupancy. This is illustrated in
Couturier et al. (2023)'s study on long-term otter recovery in France,
which used data from two primary occasions: one in 2003 and one in 2012.

DOMs have been used to study a variety of different species, though the
majority of studies have been conducted on data for vertebrate taxa.
They have been less frequently applied to non-animal organisms, perhaps
due to a reduced emphasis on imperfect detection outside of the wildlife
modelling community. However, there are exceptions --- Belinchón et al.
(2017) fits a DOM to lichen data, and Cook et al. (2022) uses them to
model the spread of chronic wasting disease. The latter's application to
disease dynamics is not unique, and DOMs have been touted as a valuable
tool for modelling disease dynamics. Mores et al. (2020) and
Padilla-Torres et al. (2013) have each used DOMs to model mosquito
dynamics as a human disease vector, and Bailey et al. (2014) discusses
the concept of applying the DOM to the spread of chytrid fungus in
amphibian habitats. The vast majority of applications have been fit for
terrestrial species, though there have been a limited number of studies
on aquatic systems, including Fisher et al. (2014)'s application to
invasive salmon, Falke et al. (2012)'s on Great Plains stream fishes,
and Pendleton et al. (2022)'s on whale occupancy dynamics.

While most articles in our sample fit a model to a single species, 41
fit models to multiple species either as independent models or
explicitly multi-species implementations. The multi-species models use
various extensions of the conventional DOM, including hierarchical
models which fit hundreds of species in a single implementation with
species-level effects(Dorazio et al., 2010; Hendershot et al., 2020), as
well as explicit interaction models which estimate conditional
occupancy, colonisation, extinction, and detection probabilities (Fidino
et al., 2019; Lesmeister et al., 2015). While they did not fit
multi-species models, several other authors fit large numbers of
independent models to different species (Otto \& Roloff, 2012; Peach et
al., 2019). Working with large numbers of taxa does raise additional
challenges, as the level of complexity which can realistically be
considered for each individual taxon is likely to be reduced for the
sake of practicality.

\begin{figure}

\centering{

\includegraphics{Figures/ReviewBox.png}

}

\caption{\label{fig-StudyDetails}A) Study areas: Data for a majority of
studies were collected from study locations in the United States. The
size of study areas was log-normally distributed, with the median study
area falling between 1000 and 10000 square kilometres. B) Study taxa:
Most species modelled in our sample were terrestrial vertebrates
including birds, mammals, or herptiles. 55 articles fit models to only a
single taxa, 30 fit independent models to multiple taxa, and 11 fit
multi-species models with explicit interactions between taxa. C) Survey
methods: Approaches to data collected were varied and included
conventional surveys like point counts and transects as well as more
modern methods like camera traps and acoustic monitors. 13 articles
incorporated data from citizen science projects. Project scale was also
variable, with the median study running for 6 years and covering 101
sites.}

\end{figure}%

The presence/absence data required for DOMs has come from a breadth of
different sources and detection methods. In our sample 79\% of articles
used human observational data, 9\% used live-trapping methods, and 14\%
used detections from camera traps. Several articles also use
citizen-science data to build detection histories, including Zuckerberg
et al. (2011)`s model using bird-feeder data and Warrier et al. (2020)'s
using citizen reports of tiger signs. Using citizen-science data may
require additional consideration of assumptions (particularly
false-positives), as discussed in greater detail by Cruickshank et al.
(2019). Within these broad categories of detection method there is
additional diversity, with several studies using unique survey methods
determining the presence of a species at a location. For example, in the
only application of a DOM for a marine species in our sample, Pendleton
et al. (2022) uses aerial transects broken up into grid cells to observe
whale occupancy; and Marescot et al. (2020)'s fits a unique
'multi-species' model treating poachers and using ranger reports to
create detection histories.

Notably, many of these studies use data which were not originally
collected in a robust design framework for occupancy modelling. In these
articles, authors formatted their data into a hierarchical format
post-hoc using a variety of methods. Some defined primary seasons as a
discrete time interval, treating all surveys occurring within the season
as secondary occasions; others defined sites as larger grid cells,
treating any survey falling within the grid as a spatial replicate.
These manipulations permit authors to use data which far predate the
DOM, with one study using surveys conducted by Joseph Grinnell in 1908
to model century-long changes in occupancy (Riddell et al., 2021).

Additionally, not all articles rely on a single source of detection data
- some integrate multiple sources of data to maximise sample size,
combining data from camera traps, sign surveys, and citizen science
reports. These integrated detection method models do require additional
care and consideration; users must ensure that different detection
methods represent comparable spatial and temporal scales, and that any
variation in perceptibility between detection methods is accounted for
(Pitman et al., 2017; Warrier et al., 2020). A special case exists when
different detection methods are used where one has the potential for
false positive detections; e.g., where less-certain citizen science
detections are combined with certain detections from field surveys. In
this context, the certain detections are used to help account for
false-positive detection probability, as in D. Miller et al. (2011)'s
study integrating GPS collars and hunter reports to estimate wolf
occupancy in Montana.

The flexibility in the data used for DOMs, including the model structure
and the scale of observations, is not amenable to a one-size-fits-all
definition of occupancy. Users of DOMs must carefully consider precisely
what they are modelling and address questions on the scale represented
by their model (Chave, 2013). Intepretation and drivers of occupancy may
differ depending on whether a site is represented by a single point on
the landscape or as a grid cell, with the former likely to depend on
more local, small scale factors rather than landscape-level trends
(Stevens \& Conway, 2019). This is also true for the temporal scale of
occupancy: whether a site is occupied within a week or within a year
leads to vastly different conceptions of occupancy. This idea is
particularly relevant in cases where the selection of season length
\emph{is} to some extent arbitrary, as with camera-trap or bioacoustic
data where continuous recordings can be broken down into distinct
`seasons' of any length. DOMs are well suited to these data types
(Balantic \& Donovan, 2019), and the proliferation of autonomous survey
techniques provides novel opportunities for analysis that is simply not
possible with human-collected data. For example, Kleiven et al. (2020)
and Mölle et al. (2022) divide their camera trap records into seasons of
just a few days. While this provides interesting insights of occupancy
at extremely fine temporal scale, further research may be needed on how
to determine appropriate season and survey durations with respect to
research questions and the closure assumption.

\section{Practices in implementation and model
building}\label{practices-in-implementation-and-model-building}

Building dynamic occupancy models can be a challenging process requiring
careful consideration of which environmental factors to incorporate to
adequately represent occupancy and detection in complex natural systems.
In this review, we have recorded all covariates considered for each
model in our sample. These are summarised in a taxonomy presented in
Table~\ref{tbl-covariates}, which states the proportion of articles
considering each type of covariate in their models, the means by which
that covariate data are collected, and how covariates responses are
represented in DOMs. We further categorise these covariates into two
groups: environmental covariates, which represent plausible ecological
correlates of parameters; and structural covariates, representing
aspects of model form and observation functionally distinct from the
environment. Our findings indicate that users of DOMs have incorporated
an exceptionally wide diversity of covariates in their models --- while
the most common varieties of covariates include aspects of habitat and
land cover, a range of other unique factors were considered by authors
in our sample. Many models incorporate covariates representing aspects
of site geometry and connectivity (37\% of articles), adding spatial
components to these DOMs which bear similarities to related stochastic
patch occupancy models. Notable examples of these spatially-explicit
models include Mortelliti \& Boitani (2007)'s model fit to rodent data
in patchy landscapes and Risk et al. (2011)'s hybrid DOM/incidence
function model used to explore meta-population dynamics of rails.
Several studies also include biotic interactions with other species as
covariates, particularly those which describe taxa threatened by
invasive species. Examples include models with barred owl presence as a
covariate on spotted owl occupancy and models with bullfrog covariates
on native amphibian occupancy, each of which effectively incorporate
species interactions without fitting explicitly multi-species models
(Mangan et al., 2019; Rowe et al., 2019).

\begin{figure*}

\begin{longtable}[]{@{}llrrrrrrrrrr@{}}

\caption{\label{tbl-covariates}All covariates considered for inclusion
in a study were classified into mutually exclusive categories. We
calculate the percentage of articles which include at least one
covariate of a given category on any parameter, Initial Occupancy (ψ1),
Occupancy (ψ), Colonisation (γ), Extinction(ε), and Detection (ρ). We
also present the percentage of all implementations of each covariate
category which are dynamic (varying through seasons) and directly
observed, as well as the percentage of articles which model each
category of covariate as a non-linear relationship or as part of an
interaction with another covariate.}

\tabularnewline

\toprule\noalign{}
\multicolumn{2}{@{}l}{%
\multirow{2}{=}{}} & \multicolumn{6}{c}{%
{Percentage of articles with covariate on parameters:}} &
\multicolumn{2}{c}{%
{Percentage which are:}} & \multicolumn{2}{c@{}}{%
{Articles representing this covariate with:}} \\
& & Any & ψ1 & ψ & γ & ε & ρ & Dynamic & Directly observed & Non-linear
relationships & Interactions between covariates \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\multirow{9}{=}{Environmental covariates} & \textbf{Habitat}
\emph{Aspects of habitat and land cover} & 56\% & 41\% & 21\% & 43\% &
48\% & 27\% & 21\% & 14\% & 12\% & 26\% \\
& \textbf{Phenology} \emph{Time-varying elements distinct from sampling
occasions} & 38\% & 1\% & 0\% & 6\% & 5\% & 37\% & 100\% & 0\% & 41\% &
12\% \\
& \textbf{Spatial} \emph{Site geometry, connectivity, or other spatial
elements} & 37\% & 22\% & 36\% & 34\% & 30\% & 11\% & 19\% & 25\% & 18\%
& 21\% \\
& \textbf{Climate/Weather} \emph{Climate, weather, and natural
disasters} & 33\% & 13\% & 14\% & 17\% & 18\% & 22\% & 65\% & 15\% &
33\% & 30\% \\
& \textbf{Topography} \emph{Elements of landscape topography, including
hydrology} & 30\% & 24\% & 36\% & 21\% & 19\% & 8\% & 44\% & 16\% & 26\%
& 19\% \\
& \textbf{Anthropogenic} \emph{Relations to human activity} & 26\% &
19\% & 7\% & 23\% & 21\% & 4\% & 55\% & 4\% & 9\% & 26\% \\
& \textbf{Other environmental} \emph{Other environmental covariate not
otherwise listed} & 21\% & 4\% & 21\% & 5\% & 11\% & 12\% & 77\% & 82\%
& 0\% & 0\% \\
& \textbf{Biotic interaction} \emph{Interactions with other (non-plant)
species} & 16\% & 6\% & 0\% & 14\% & 14\% & 6\% & 73\% & 76\% & 7\% &
21\% \\
& \textbf{Any Environmental} & 91\% & 61\% & 93\% & 74\% & 75\% & 72\% &
42\% & 17\% & 34\% & 28\% \\
\multirow{7}{=}{Structural covariates} & \textbf{Primary occasion}
\emph{Effect of the primary occasion} & 66\% & 1\% & 43\% & 38\% & 38\%
& 62\% & 99\% & 0\% & 15\% & 10\% \\
& \textbf{Observation} \emph{Details on the observation process} & 21\%
& 0\% & 0\% & 0\% & 0\% & 21\% & 91\% & 9\% & 5\% & 11\% \\
& \textbf{Secondary occasion} \emph{Effect of the secondary occasion} &
16\% & 0\% & 0\% & 0\% & 0\% & 16\% & 100\% & 15\% & 14\% & 0\% \\
& \textbf{Species effect} \emph{Species-level effects} & 6\% & 5\% & 0\%
& 5\% & 5\% & 4\% & 0\% & 0\% & 0\% & 60\% \\
& \textbf{Site effect} \emph{Site-level effects} & 3\% & 0\% & 0\% & 2\%
& 2\% & 2\% & 0\% & 0\% & 0\% & 0\% \\
& \textbf{Other structural} \emph{Other structural covariate not
otherwise listed} & 3\% & 1\% & 0\% & 0\% & 0\% & 2\% & 60\% & 60\% &
0\% & 33\% \\
& \textbf{Any Structural} & 82\% & 6\% & 43\% & 42\% & 42\% & 81\% &
87\% & 4\% & 15\% & 12\% \\
All covariates & \textbf{All covariates} & 99\% & 63\% & 100\% & 85\% &
86\% & 97\% & 46\% & 16\% & 35\% & 26\% \\

\end{longtable}

\end{figure*}%

Covariate data for studies in our sample was either collected directly
by researchers (17\% of environmental covariate data), or derived from
pre-existing remotely sensed datasets (83\%); this of course varies
depending on the category of covariate. Directly collected data can
generally be expected to represent finer-scale factors like prey species
occurrence or details of habitat structure, which can be difficult to
measure remotely but can often be more proximal drivers of occupancy.
These high quality covariates do come with trade-offs, as they can be
expensive to collect and can often preclude projection to locations
where data are unavailable. For projects interested in making these
projections, remotely-sensed covariates are generally more feasible
despite their generally more distal nature (M. P. Austin, 2002). 42\% of
environmental factors and 87\% of structural factors included in
reviewed articles were dynamic covariates which varied through time ---
this again varied with the category of covariate in question, with terms
relating to climate or weather most frequently dynamic and topographic
covariates universally static.

In the standard DOM, covariates for each parameter are most commonly
incorporated via a logistic regression (i.e., a linear regression
through a logit link function). Statistical relationships between
covariates and model parameters are represented as linear terms unless
otherwise specified. Of course, many ecological relationships are
non-linear and require more complex forms to be realistically
represented in a model. M. Austin (2007) discussed the importance of
modelling ecologically-realistic responses to covariates, advocating for
careful consideration of the most appropriate statistical form for
hypothesised relationships. However, linear models are quite flexible,
and non-linear responses can be easily accommodated using polynomial
terms and interactions between covariates. Despite this, in our sample
just 35\% of articles considered using at least one non-linear responses
to an environmental covariate, with the majority of studies representing
all covariates as simple linear relationships. Interactions between
covariates were similarly rare, with only 26\% of studies considering at
least one relationship between terms. The relatively low emphasis on
more complex non-linear responses contrasts with other popular methods
for modelling species occupancy. Many common approaches for SDMs, which
include MAXENT and Boosted Regression Trees, permit considerable
flexibility in the shape of their covariate response curves where
supported by the data (Elith et al., 2008; Merow et al., 2013). This
emphasis on more complex responses in SDMs may be due to their typically
larger range-wide scale which encompasses the full species niche, where
environmental relations may be expected to be non-linear. However, as
previously indicated DOMs have been frequently implemented at similarly
large spatial scales where the same assumptions should exist and similar
responses might be expected. Similar concerns exist for the relatively
low level of interactions between covariates: while they should not
necessarily be used liberally, where these exist their exclusion can
influence model performance (Guisan et al., 2006).

In addition to the types of covariates considered for modelling we also
measured the size of the candidate covariate pool in each reviewed
study, tallying the number of environmental and structure covariates
which were available for use on each parameter (
Figure~\ref{fig-covariates}). This is an area of considerable variation
in modelling practices --- the number of covariates considered for each
parameter ranges from 0 (effectively modelling the parameter as a
constant) to over 40 possible factors.

\begin{figure}

\centering{

\includegraphics{Figures/CovParamPlot.jpeg}

}

\caption{\label{fig-covariates}The number of covariates considered for
each parameter across all studies in our sample. `Occupancy' given here
represents an alternative parameterisation of the DOM which jointly
estimates Occupancy for every season, Colonisation, and Detection, where
Extinction is a derived parameter; this differs from the more popular
Initial occupancy/Colonisation/Extinction/Detection parameterisation.
Here, a `covariate' is defined as each distinct covariate considered for
inclusion. Linear and quadratic representations of the same covariate
are counted as 1 covariate.}

\end{figure}%

The median number of covariates varies strongly by parameter, with
transition probabilities (colonisation and extinction) more likely to
consider a broader range of environmental covariates relative to initial
occupancy and detection. The lack of any covariates considered for
initial occupancy in many studies (37\% of articles) is particularly
notable --- unless a study is conducted at very small extents or study
sites are truly uniform in their suitability, one would expect some
amount of non-random variation in occupancy probability across any study
system that will not be captured when representing initial occupancy as
a constant. Omission of the factors which drive occupancy is a known
source of bias in static SDMs (Barry \& Elith, 2006), which the initial
occupancy component of the DOM conceptually resembles. Furthermore, any
bias in occupancy estimation in the first time-step will perpetuate into
future seasons due to the DOM's Markovian nature with implications for
the reliability of model outputs. Detection probability is also
typically represented with low quantities of environmental covariates.
Recall that detectability is dependent on not only observation factors,
but also by drivers of abundance and species use; the low number of
candidate covariates here relative to colonisation and extinction raises
additional questions on whether this variation is always appropriately
captured in models.

To this point in our review, we have discussed only the covariates which
were \emph{considered} for inclusion in DOMs without regard for which
terms were actually \emph{included} in the final models used by authors
to make their inference. Identifying the `best' model from a range of
possible candidates can be a challenging process, especially for
hierarchical models like DOMs (Doherty et al., 2012). Consider that the
quantity of candidate models involving up to \emph{n} covariates on
\emph{p} parameters can be given by \((2^n)^p\); where a simple
single-parameter may have up to 64 possible combinations of 6
covariates, a 4-parameter dynamic occupancy model would have 16 million
combinations. This explosion of candidate models generally precludes
exhaustive comparisons of possible models and requires a reduction of
the models tested. Models in our sample include both Frequentist (n =
70) and Bayesian (n = 22), which differ considerably in their manner of
implementation. A summary of modelling practices in DOMs is given in
Table~\ref{tbl-modelling} below.

\begin{longtable}{l|rrr}

\caption{\label{tbl-modelling}Modelling practices in dynamic occupancy
models, subset by frequentist or bayesian implementations. The median
covariate count presented here represents the median quantity of
covariates considered for each model parameter across the studies in our
review. The model selection methods represented in this table are
non-exclusive and some articles employ multiple approaches.}

\tabularnewline

\toprule
\multicolumn{1}{l}{} & Frequentist & Bayesian & All models \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{4}{l}{\vspace*{-5mm}} \\ 
\midrule\addlinespace[2.5pt]
Number of studies & $70$ & $22$ & $91$ \\ 
Median covariates considered per parameter & 3.12 & 2.5 & 2.75 \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{4}{l}{Covariate selection methods} \\ 
\midrule\addlinespace[2.5pt]
\hspace*{11.25px} Prop\_AnySelect & $94\%$ & $36\%$ & $80\%$ \\ 
\hspace*{11.25px} Prop\_Candidate & $43\%$ & $14\%$ & $36\%$ \\ 
\hspace*{11.25px} Prop\_Sequential & $39\%$ & $0\%$ & $30\%$ \\ 
\hspace*{11.25px} Prop\_Simple & $9\%$ & $5\%$ & $8\%$ \\ 
\hspace*{11.25px} Prop\_ModAvg & $71\%$ & $5\%$ & $56\%$ \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{4}{l}{Model evaluation conducted} \\ 
\midrule\addlinespace[2.5pt]
\hspace*{11.25px} Prop\_GOF & $20\%$ & $14\%$ & $19\%$ \\ 
\hspace*{11.25px} Prop\_Eval & $4\%$ & $18\%$ & $8\%$ \\ 
\bottomrule

\end{longtable}

The largest differences between the frequentist and Bayesian models in
our sample lie in their approaches to model selection. Where 94\% of
frequentist models perform some manner of model selection to determine
covariate inclusions for their models, only 36\% of Bayesian models do
so, with the majority instead fitting a single model defined \emph{a
priori}. Methods used for model selection vary considerably both within
and between modelling frameworks. For frequentist models, the most
popular and conventional approach to model selection as used in 43\% of
articles involves the creation of a fully pre-defined model set
containing some number of hypothesis candidate models, where the best
model is selected according to the lowest AIC score. Most frequentist
studies which do not follow this protocol instead use procedural model
selection methods (39\% of articles), where the structure for each model
parameter is fit one-by-one. For example, this protocol might first
identify the best structure for detection probability before moving on
to initial occupancy and so on. A smaller portion of frequentist studies
(9\%) use more unique approaches, fitting simpler models such as single
season occupancy models for candidate covariates to identify the most
informative terms to use in a DOM. Across all frequentist
implementations, 71\% of articles conduct multi-model inference and
model-averaging using AIC weights (Burnham \& Anderson, 2004).

Those Bayesian models which \emph{do} perform model selection tend to
take distinctly different approaches, with largely idiosyncratic methods
across these studies. While direct comparison of model fit is rare, it
is feasible --- Urban et al. (2023) identifies the best model from a
Bayesian candidate set using the predictive performance on both in and
out-of-sample validation data. Another approach used by Cook et al.
(2022) fit a global model including all covariates before removing each
covariate where the 95\% credible interval of the posterior distribution
overlapped zero and refitting the model. Ahumada et al. (2013) takes a
hybrid approach, in which model selection is conducted by a procedural
method in the frequentist framework before refitting the best structure
in a Bayesian model.

Limited research has been conducted on advantages of different methods
for covariate selection in DOMs, and there is surely no one-size-fit all
approach which will be appropriate for all possible use-cases. However,
it is important to consider the implications of the different model
selection approaches in common usage, and research from the SDM
literature highlights how covariate selection can influence our
interpretation of model outputs (Brodie et al., 2020). Within the
occupancy modelling literature, Stewart et al. (2023)'s article on
covariate selection in single-season occupancy model discusses important
attributes of the information-criteria approaches widely used in
frequentist models in our review, noting that these approaches can lead
to inaccurate estimates and influence model inference. Morin et al.
(2020)'s work raises different concerns on procedural model selection
methods, demonstrating how the fine details of modelling protocols can
determine whether the best-fitting possible model is identified and
which covariates appear in final models. The literature on performance
of Bayesian model selection methods is more sparse, although Hooten \&
Hobbs (2015)'s guide to Bayesian model selection in ecology is a
valuable resource for possible methods of fitting those models.

With few exceptions, most of the model selection methods identified in
our review are designed only to identify the best model of those
considered, and do not necessarily guarantee that selected models are
well-fit to the data. Assessment of the performance of DOMs generally
requires additional steps, and the best ways to achieve this are not
necessarily clear. There is no broadly-accepted goodness-of-fit test for
dynamic occupancy models, although MacKenzie \& Bailey (2004) describes
an approach for single-season occupancy models using a parametric
bootstrap which can been extended to DOMs; this test is implemented in
the \emph{AICcModAvg} and \emph{unmarked} R packages. Kéry \& Royle
(2021) describe the test and present an alternative based on separately
assessing fit to static and dynamic components of the model. In Bayesian
implementations, posterior predictive checks offer means to assess model
fit (Gelman, 2014). Broms et al. (2016) discusses possibilities for
model evaluation in the Bayesian context focusing on single-season
multi-species occupancy models but extensions of their approach may also
be applicable to the DOM. As in other hierarchical models, model
evaluation for DOMs can in general can be difficult and somewhat
uncertain compared to other model types, as the primary response
variable of interest (species occupancy) is a latent variable where the
true state is generally not known. Predictive performance evaluation is
thus typically based on \emph{observed} occupancy data, where a DOM is
used to simulate detection histories to be compared with field survey
results.

Perhaps as a result of this uncertainty, assessment of model fit and
model performance was rare amongst the articles in our sample. Only 19\%
of studies tested for goodness-of-fit, and just 10\% calculated
predictive performance. These rates are considerably lower than for
models like SDMs; for which M. B. Araújo et al. (2019)'s review of
applications found closer to 50\% articles testing both fit and
performance.

\section{Modelling objectives}\label{modelling-objectives}

Studies in our sample were classified according to their objectives as
stated by authors to explore the use-cases which DOMs have been applied
for. These objective categories were non-exclusive, and many studies
fell into multiple categories; the frequency at which studies pursued
each of these objectives is shown in Figure~\ref{fig-objectives} A.
Across our study period DOMs have been frequently employed to observe
trends in species occupancy, to explore relationships between
environmental factors and occupancy, and, increasingly often, to make
predictions of occupancy spatially to unsurveyed locations or temporally
to hypothetical future conditions. Within each of these broad categories
lies even more variation in objectives, emphasising the DOM's
flexibility as a tool for making ecological inference in diverse
contexts.

Many articles in our sample used the DOM seeking to monitor trends in
occupancy state through time. These studies include Ahumada et al.
(2013)`s paper assessing populations trends of 13 forest mammals and
Scott \& Rissler (2015)'s assessment of the conservation status of the
endangered flattened musk turtle. These studies offer valuable insights
on the state of species across landscapes and through time,
demonstrating DOMs' suitability for many monitoring oriented projects
--- including those where many species need to be assessed at once.
Other articles were more focused on examining the factors which
influence species occupancy, either testing a pre-specified hypothetical
relationships or taking a broader tack to identify drivers of occupancy
without preconception. The distinction between the two is important, as
it guides which covariates may be considered and how model selection may
be used to fit models. Many of these studies target core conservation
priorities for their focal species, like Olson et al. (2005)'s early DOM
assessing the influence of barred owls on threatened spotted owl.
Explorations of these pivotal relationships are important for guiding
management action, and may also be used to test the effectiveness of
these actions as in K. E. Miller \& Brown (2023)'s study on American
Kestrel nest boxes. In scenarios where less is known about species
habitat preference, DOMs may be used to examine the influence of a wider
variety of factors as illustrated in Huber et al. (2017)'s study testing
the relative influence of dozens of habitat covariates on Wood warbler
occupancy. In recent years, increasing numbers of articles have used to
DOM to generate predictions of species occupancy. As examples, McGowan
et al. (2020) provides projections for the threatened Black rail under
multiple management scenarios, and Pollentier et al. (2021) generates
maps of distributions resembling those made with SDMs, outputs which can
immediately be used by managers to inform action.

\begin{figure}

\centering{

\includegraphics{Figures/ObjectiveFigure.jpeg}

}

\caption{\label{fig-objectives}A) Shows the proportion of articles in
each year-strata and across all years which match each of six
non-exclusive objective categories. B) Shows the quantity of covariates
considered per parameter for models which pursued each objectives.
Studies focused on methods development typically considered the fewest
covariates, where those attempting to identify the drivers of occupancy
considered the most.}

\end{figure}%

The objectives of studies fitting DOMs strongly inform approaches to the
model building process, particularly with respect to covariate
selection. Figure~\ref{fig-objectives} B shows the quantity of
covariates considered for models applied to each of our objective
categories, and the differences between objectives are apparent. Not
unexpectedly, articles which focused on describing new methods for DOMs
had fewer covariates than those studies focused on more applied
objectives. However, differences persist between those articles
observing trends, identifying trends in occupancy, testing
relationships, and making predictions.

\section{Synthesis and key
priorities}\label{synthesis-and-key-priorities}

Approaches to building any type of model will necessarily depend on the
possibilities of the data at hand and on the priorities of the
model-builder. This precludes any prescription of the `best' way to
build a model, however, there are still important discussions to be had
on decisions made in the modelling process. One aspect of fitting DOMs
meriting broader discussion revolves around `model complexity,' and how
much must be incorporated into models to reliable model species
occupancy under different contexts and use cases. Complexity is a broad
term which encompasses many aspects of a model (Merow et al., 2014), and
opinions on simplicity versus complexity in ecological models can be
divisive. Where some advocate for the simplest possible models, arguing
that they are most generaliseable; others insist that overly-simple
models cannot adequately represent the most important drivers in a
system (Evans et al., 2013; Lonergan, 2014). By their nature DOMs are
somewhat more complex than simpler models for studying occupancy thanks
to their hierarchical structure, attributes which are necessary to
control for detectability and to capture occupancy dynamics. Within this
structure, however, further complexity is to some degree up to the
modeller: one can choose how many covariates to consider for inclusion
on the various parameters, and how to represent the nature of the
relationship between those covariates and parameters. Research from SDMs
indicates that allowing for more complex relationships can improve model
performance in predicting occupancy (Valavi et al., 2023), an
increasingly popular use-case for DOMs. Within the DOM framework, there
are exciting developments on that front --- Joseph (2020) presents a
novel neural-network occupancy model which allows for exponentially
higher levels of complexity and may offer improved performance for
prediction-oriented studies.

Generally speaking, covariate selection seems to be a particularly
important area for further investigations into building DOMs. In our
review, we see little consensus around which approaches are most
applicable for any given use case, and existing work on covariate
selection in DOMs raises concerns on whether common methods always
produce the most suitable models. This is true for both frequentist and
Bayesian implementations, and comparative research on covariate
selection under both frameworks may help to inform model users on which
method may be most appropriate for their use-cases. In a similar vein,
the low number of articles in our review which conducted model
evaluation or assessed model fit raises different questions. While the
appropriate method of model evaluation may depend on data availability
and research objectives, assessing models by some method is generally
important to understand how reliable model outputs may be (M. Araújo et
al., 2005; Guisan \& Thuiller, 2005). Existing uncertainties around
whether current methods are suitable for the task surely discourage
users from calculating these metrics, and further research is needed to
establish trusted practices for assessing the quality of DOMs.

These priorities are particularly important given the frequent applied
objectives of DOM users, tackling challenges which include assessing
critically endangered species (Carvalho et al., 2023), guiding public
health management of disease vectors (Moreira et al., 2016), and
tracking rapidly developing biological invasions (Wood et al., 2020).
DOMs are well suited to these situations, and it is to be expected that
as these types of applications are more commonly attempted understanding
the sensitivity of model outputs to decisions made in the model fitting
process becomes increasingly important. In the two decades since the
publication of MacKenzie et al. (2003) the dynamic occupancy model has
become a widely used tool for ecological inference, with numerous
extensions to the modelling framework further broadening the scope of
questions and use-cases for which it may be applied. Given their
increasing popularity, further research and guidelines around issues of
model building may help to make the DOM format more accessible to
newcomers and support confidence in interpretation of models. Parallels
to existing work in the SDM literature on guidelines for reporting and
modelling standards could be valuable contributions in achieving these
aims (M. B. Araújo et al., 2019; Zurell et al., 2020) while specifically
addressing model attributes unique to DOMs.

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-ahumada2013}
Ahumada, J. A., Hurtado, J., \& Lizcano, D. (2013). Monitoring the
Status and Trends of Tropical Forest Terrestrial Vertebrate Communities
from Camera Trap Data: A Tool for Conservation. \emph{PLOS ONE},
\emph{8}(9), e73707. \url{https://doi.org/10.1371/journal.pone.0073707}

\bibitem[\citeproctext]{ref-arauxfajo2019}
Araújo, M. B., Anderson, R. P., Márcia Barbosa, A., Beale, C. M.,
Dormann, C. F., Early, R., Garcia, R. A., Guisan, A., Maiorano, L.,
Naimi, B., O'Hara, R. B., Zimmermann, N. E., \& Rahbek, C. (2019).
Standards for distribution models in biodiversity assessments.
\emph{Science Advances}, \emph{5}(1), eaat4858.
\url{https://doi.org/10.1126/sciadv.aat4858}

\bibitem[\citeproctext]{ref-arauxfajo2005}
Araújo, M., Pearson, R., Thuiller, W., \& Erhard, M. (2005). Validation
of species-climate impact models under climate change. \emph{Global
Change Biology}, \emph{11}, 1504--1513.
\url{https://doi.org/10.1111/j.1365-2486.2005.01000.x}

\bibitem[\citeproctext]{ref-austin2007}
Austin, M. (2007). Species distribution models and ecological theory: A
critical assessment and some possible new approaches. \emph{Ecological
Modelling}, \emph{200}(1), 1--19.
\url{https://doi.org/10.1016/j.ecolmodel.2006.07.005}

\bibitem[\citeproctext]{ref-austin2002}
Austin, M. P. (2002). Spatial prediction of species distribution: An
interface between ecological theory and statistical modelling.
\emph{Ecological Modelling}, \emph{157}(2), 101--118.
\url{https://doi.org/10.1016/S0304-3800(02)00205-3}

\bibitem[\citeproctext]{ref-bailey2014}
Bailey, L. L., MacKenzie, D. I., \& Nichols, J. D. (2014). Advances and
applications of occupancy models. \emph{Methods in Ecology and
Evolution}, \emph{5}(12), 1269--1279.
\url{https://doi.org/10.1111/2041-210X.12100}

\bibitem[\citeproctext]{ref-balantic2019}
Balantic, C., \& Donovan, T. (2019). Dynamic wildlife occupancy models
using automated acoustic monitoring data. \emph{Ecological
Applications}, \emph{29}(3). \url{https://doi.org/10.1002/eap.1854}

\bibitem[\citeproctext]{ref-barry2006}
Barry, S., \& Elith, J. (2006). Error and uncertainty in habitat models.
\emph{Journal of Applied Ecology}, \emph{43}(3), 413--423.
\url{https://doi.org/10.1111/j.1365-2664.2006.01136.x}

\bibitem[\citeproctext]{ref-basset2023}
Basset, Y., Butterill, P. T., Donoso, D. A., P. A. Lamarre, G.,
Souto-Vilarós, D., Perez, F., Bobadilla, R., Lopez, Y., Alejandro
Ramírez Silva, J., \& Barrios, H. (2023). Abundance, occurrence and time
series: Long-term monitoring of social insects in a tropical rainforest.
\emph{Ecological Indicators}, \emph{150}, 110243.
\url{https://doi.org/10.1016/j.ecolind.2023.110243}

\bibitem[\citeproctext]{ref-belinchuxf3n2017}
Belinchón, R., Harrison, P. J., Mair, L., Várkonyi, G., \& Snäll, T.
(2017). Local epiphyte establishment and future metapopulation dynamics
in landscapes with different spatiotemporal properties. \emph{Ecology},
\emph{98}(3), 741--750. \url{https://doi.org/10.1002/ecy.1686}

\bibitem[\citeproctext]{ref-berigan2019}
Berigan, W. J., Jones, G. M., Whitmore, S. A., Gutiérrez, R. J., \&
Peery, M. Z. (2019). Cryptic wide-ranging movements lead to upwardly
biased occupancy in a territorial species. \emph{Journal of Applied
Ecology}, \emph{56}(2), 470--480.
\url{https://doi.org/10.1111/1365-2664.13265}

\bibitem[\citeproctext]{ref-bertelsmeier2013}
Bertelsmeier, C., Luque, G. M., \& Courchamp, F. (2013). Increase in
Quantity and Quality of Suitable Areas for Invasive Species as Climate
Changes. \emph{Conservation Biology}, \emph{27}(6), 1458--1467.
\url{https://doi.org/10.1111/cobi.12093}

\bibitem[\citeproctext]{ref-briscoe2019}
Briscoe, N. J., Elith, J., Salguero-Gómez, R., Lahoz-Monfort, J. J.,
Camac, J. S., Giljohann, K. M., Holden, M. H., Hradsky, B. A., Kearney,
M. R., McMahon, S. M., Phillips, B. L., Regan, T. J., Rhodes, J. R.,
Vesk, P. A., Wintle, B. A., Yen, J. D. L., \& Guillera-Arroita, G.
(2019). Forecasting species range dynamics with process-explicit models:
matching methods to applications. \emph{Ecology Letters}, \emph{22}(11),
1940--1956. \url{https://doi.org/10.1111/ele.13348}

\bibitem[\citeproctext]{ref-briscoe2021}
Briscoe, N. J., Zurell, D., Elith, J., König, C., Fandos, G., Malchow,
A.-K., Kéry, M., Schmid, H., \& Guillera-Arroita, G. (2021). Can dynamic
occupancy models improve predictions of species' range dynamics? A test
using Swiss birds. \emph{Global Change Biology}, \emph{27}(18),
4269--4282. \url{https://doi.org/10.1111/gcb.15723}

\bibitem[\citeproctext]{ref-brodie2020}
Brodie, S. J., Thorson, J. T., Carroll, G., Hazen, E. L., Bograd, S.,
Haltuch, M. A., Holsman, K. K., Kotwicki, S., Samhouri, J. F.,
Willis-Norton, E., \& Selden, R. L. (2020). Trade-offs in covariate
selection for species distribution models: a methodological comparison.
\emph{Ecography}, \emph{43}(1), 11--24.
\url{https://doi.org/10.1111/ecog.04707}

\bibitem[\citeproctext]{ref-broms2016}
Broms, K. M., Hooten, M. B., Johnson, D. S., Altwegg, R., \& Conquest,
L. L. (2016). Dynamic occupancy models for explicit colonization
processes. \emph{Ecology}, \emph{97}(1), 194--204.
\url{https://doi.org/10.1890/15-0416.1}

\bibitem[\citeproctext]{ref-burnham2004}
Burnham, K. P., \& Anderson, D. R. (2004). \emph{Model Selection and
Multimodel Inference}. Springer. \url{https://doi.org/10.1007/b97636}

\bibitem[\citeproctext]{ref-carvalho2023}
Carvalho, E. a. R., Mendonça, E. N., Lopes, A. M. C., \& Haugaasen, T.
(2023). Current status of the Critically Endangered Black-winged
Trumpeter Psophia obscura in one of its last strongholds. \emph{Bird
Conservation International}, \emph{33}, e12.
\url{https://doi.org/10.1017/S0959270922000077}

\bibitem[\citeproctext]{ref-chave2013}
Chave, J. (2013). The problem of pattern and scale in ecology: what have
we learned in 20 years? \emph{Ecology Letters}, \emph{16}(s1), 4--16.
\url{https://doi.org/10.1111/ele.12048}

\bibitem[\citeproctext]{ref-clement2019}
Clement, M. J., Nichols, J. D., Collazo, J. A., Terando, A. J., Hines,
J. E., \& Williams, S. G. (2019). Partitioning global change: Assessing
the relative importance of changes in climate and land cover for changes
in avian distribution. \emph{Ecology and Evolution}, \emph{9}(4),
1985--2003. \url{https://doi.org/10.1002/ece3.4890}

\bibitem[\citeproctext]{ref-cook2022}
Cook, J. D., Williams, D. M., Porter, W. F., \& Christensen, S. A.
(2022). Improved predictions and forecasts of chronic wasting disease
occurrence using multiple mechanism dynamic occupancy modeling.
\emph{The Journal of Wildlife Management}, \emph{86}(7), e22296.
\url{https://doi.org/10.1002/jwmg.22296}

\bibitem[\citeproctext]{ref-couturier2023}
Couturier, T., Steinmetz, J., Defos du Rau, P., Marc, D., Trichet, E.,
Gomes, R., \& Besnard, A. (2023). Intensive agriculture as the main
limiting factor of the otter's return in southwest france.
\emph{Biological Conservation}, \emph{279}, 109927.
\url{https://doi.org/10.1016/j.biocon.2023.109927}

\bibitem[\citeproctext]{ref-cruickshank2019}
Cruickshank, S. S., Bühler, C., \& Schmidt, B. R. (2019). Quantifying
data quality in a citizen science monitoring program: False negatives,
false positives and occupancy trends. \emph{Conservation Science and
Practice}, \emph{1}(7). \url{https://doi.org/10.1111/csp2.54}

\bibitem[\citeproctext]{ref-devarajan2020}
Devarajan, K., Morelli, T. L., \& Tenan, S. (2020). Multi{-}species
occupancy models: review, roadmap, and recommendations.
\emph{Ecography}, \emph{43}(11), 1612--1624.
\url{https://doi.org/10.1111/ecog.04957}

\bibitem[\citeproctext]{ref-doherty2012}
Doherty, P. F., White, G. C., \& Burnham, K. P. (2012). Comparison of
model building and selection strategies. \emph{Journal of Ornithology},
\emph{152}(2), 317--323. \url{https://doi.org/10.1007/s10336-010-0598-5}

\bibitem[\citeproctext]{ref-dorazio2010}
Dorazio, R. M., Kéry, M., Royle, J. A., \& Plattner, M. (2010). Models
for inference in dynamic metacommunity systems. \emph{Ecology},
\emph{91}(8), 2466--2475. \url{https://doi.org/10.1890/09-1033.1}

\bibitem[\citeproctext]{ref-dormann2007}
Dormann, C. F. (2007). Promising the future? Global change projections
of species distributions. \emph{Basic and Applied Ecology}, \emph{8}(5),
387--397. \url{https://doi.org/10.1016/j.baae.2006.11.001}

\bibitem[\citeproctext]{ref-elith2010}
Elith, J., Kearney, M., \& Phillips, S. (2010). The art of modelling
range-shifting species. \emph{Methods in Ecology and Evolution},
\emph{1}(4), 330--342.
\url{https://doi.org/10.1111/j.2041-210X.2010.00036.x}

\bibitem[\citeproctext]{ref-elith2008}
Elith, J., Leathwick, J. R., \& Hastie, T. (2008). A working guide to
boosted regression trees. \emph{Journal of Animal Ecology},
\emph{77}(4), 802--813.
\url{https://doi.org/10.1111/j.1365-2656.2008.01390.x}

\bibitem[\citeproctext]{ref-evans2013}
Evans, M. R., Grimm, V., Johst, K., Knuuttila, T., Langhe, R. de,
Lessells, C. M., Merz, M., O'Malley, M. A., Orzack, S. H., Weisberg, M.,
Wilkinson, D. J., Wolkenhauer, O., \& Benton, T. G. (2013). Do simple
models lead to generality in ecology? \emph{Trends in Ecology \&
Evolution}, \emph{28}(10), 578--583.
\url{https://doi.org/10.1016/j.tree.2013.05.022}

\bibitem[\citeproctext]{ref-falke2012}
Falke, J. A., Bailey, L. L., Fausch, K. D., \& Bestgen, K. R. (2012).
Colonization and extinction in dynamic habitats: an occupancy approach
for a Great Plains stream fish assemblage. \emph{Ecology}, \emph{93}(4),
858--867. \url{https://doi.org/10.1890/11-1515.1}

\bibitem[\citeproctext]{ref-fidino2019}
Fidino, M., Simonis, J. L., \& Magle, S. B. (2019). A multistate dynamic
occupancy model to estimate local colonization{\textendash}extinction
rates and patterns of co{-}occurrence between two or more interacting
species. \emph{Methods in Ecology and Evolution}, \emph{10}(2),
233--244. \url{https://doi.org/10.1111/2041-210X.13117}

\bibitem[\citeproctext]{ref-fisher2014}
Fisher, A. C., Volpe, J. P., \& Fisher, J. T. (2014). Occupancy dynamics
of escaped farmed Atlantic salmon in Canadian Pacific coastal salmon
streams: implications for sustained invasions. \emph{Biological
Invasions}, \emph{16}(10), 2137--2146.
\url{https://doi.org/10.1007/s10530-014-0653-x}

\bibitem[\citeproctext]{ref-franklin2010}
Franklin, J. (2010). \emph{Mapping species distributions: Spatial
inference and prediction}. Cambridge University Press.
\url{https://www.cambridge.org/core/books/mapping-species-distributions/58225AE5693AED8BD812F7CEBE35378A}

\bibitem[\citeproctext]{ref-gelman2014}
Gelman, A. (2014). \emph{Bayesian data analysis}.
\url{https://research.ebsco.com/linkprocessor/plink?id=86df87c2-efc9-3fae-8c85-32142daec7af}

\bibitem[\citeproctext]{ref-gu2004}
Gu, W., \& Swihart, R. K. (2004). Absent or undetected? Effects of
non-detection of species occurrence on wildlife{\textendash}habitat
models. \emph{Biological Conservation}, \emph{116}(2), 195--203.
\url{https://doi.org/10.1016/S0006-3207(03)00190-3}

\bibitem[\citeproctext]{ref-guillera-arroita2017}
Guillera-Arroita, G., \& Lahoz-Monfort, J. J. (2017). Species occupancy
estimation and imperfect detection: shall surveys continue after the
first detection? \emph{AStA Advances in Statistical Analysis},
\emph{101}(4), 381--398. \url{https://doi.org/10.1007/s10182-017-0292-5}

\bibitem[\citeproctext]{ref-guisan2006}
Guisan, A., Lehmann, A., Ferrier, S., Austin, M., Overton, J. Mc. C.,
Aspinall, R., \& Hastie, T. (2006). Making better biogeographical
predictions of species{'} distributions. \emph{Journal of Applied
Ecology}, \emph{43}(3), 386--392.
\url{https://doi.org/10.1111/j.1365-2664.2006.01164.x}

\bibitem[\citeproctext]{ref-guisan2005}
Guisan, A., \& Thuiller, W. (2005). Predicting species distribution:
offering more than simple habitat models. \emph{Ecology Letters},
\emph{8}(9), 993--1009.
\url{https://doi.org/10.1111/j.1461-0248.2005.00792.x}

\bibitem[\citeproctext]{ref-gutiuxe9rrez-arellano2024}
Gutiérrez-Arellano, C., Crone, E. E., Pettorelli, N., \& Hodgson, J. A.
(2024). Broadening applications of stochastic patch occupancy models
over three decades. \emph{Diversity and Distributions}, \emph{n/a}(n/a),
e13822. \url{https://doi.org/10.1111/ddi.13822}

\bibitem[\citeproctext]{ref-hendershot2020}
Hendershot, J. N., Smith, J. R., Anderson, C. B., Letten, A. D.,
Frishkoff, L. O., Zook, J. R., Fukami, T., \& Daily, G. C. (2020).
Intensive farming drives long-term shifts in avian community
composition. \emph{Nature}, \emph{579}(7799), 393--396.
\url{https://doi.org/10.1038/s41586-020-2090-6}

\bibitem[\citeproctext]{ref-hooten2015}
Hooten, M. B., \& Hobbs, N. T. (2015). A guide to Bayesian model
selection for ecologists. \emph{Ecological Monographs}, \emph{85}(1),
3--28. \url{https://doi.org/10.1890/14-0661.1}

\bibitem[\citeproctext]{ref-huber2017}
Huber, N., Kéry, M., \& Pasinelli, G. (2017). Occupancy dynamics of the
Wood Warbler Phylloscopus sibilatrix assessed with habitat and remote
sensing data. \emph{Ibis}, \emph{159}(3), 623--637.
\url{https://doi.org/10.1111/ibi.12472}

\bibitem[\citeproctext]{ref-humboldt1849}
Humboldt, A. von. (1849). \emph{Cosmos : a sketch of a physical
description of the universe}.

\bibitem[\citeproctext]{ref-james2021}
James, G., Witten, D., Hastie, T., \& Tibshirani, R. (2021). \emph{An
Introduction to Statistical Learning: with Applications in R}. Springer
US. \url{https://doi.org/10.1007/978-1-0716-1418-1}

\bibitem[\citeproctext]{ref-joseph2020}
Joseph, M. B. (2020). Neural hierarchical models of ecological
populations. \emph{Ecology Letters}, \emph{23}(4), 734--747.
\url{https://doi.org/10.1111/ele.13462}

\bibitem[\citeproctext]{ref-kellner2014}
Kellner, K. F., \& Swihart, R. K. (2014). Accounting for Imperfect
Detection in Ecology: A Quantitative Review. \emph{PLOS ONE},
\emph{9}(10), e111436.
\url{https://doi.org/10.1371/journal.pone.0111436}

\bibitem[\citeproctext]{ref-kendall2013}
Kendall, W. L., Hines, J. E., Nichols, J. D., \& Grant, E. H. C. (2013).
Relaxing the closure assumption in occupancy models: staggered arrival
and departure times. \emph{Ecology}, \emph{94}(3), 610--617.
\url{https://doi.org/10.1890/12-1720.1}

\bibitem[\citeproctext]{ref-kuxe9ry2013}
Kéry, M., Guillera-Arroita, G., \& Lahoz-Monfort, J. J. (2013).
Analysing and mapping species range dynamics using occupancy models.
\emph{Journal of Biogeography}, \emph{40}(8), 1463--1474.
\url{https://doi.org/10.1111/jbi.12087}

\bibitem[\citeproctext]{ref-kuxe9ry2021}
Kéry, M., \& Royle, J. A. (2021). \emph{Applied hierarchical modeling in
ecology : analysis of distribution, abundance and species richness in R
and BUGS. Volume 2, Dynamic and advanced models}.
\url{https://research.ebsco.com/linkprocessor/plink?id=acdec215-f169-3c05-a996-ad3f64d6341e}

\bibitem[\citeproctext]{ref-kleiven2020}
Kleiven, E. F., Barraquand, F., Gimenez, O., Henden, J.-A., Ims, R. A.,
Soininen, E. M., \& Yoccoz, N. G. (2020). \emph{A dynamic occupancy
model for interacting species with two spatial scales}.
\url{https://doi.org/10.1101/2020.12.16.423067}

\bibitem[\citeproctext]{ref-lahoz-monfort2014}
Lahoz-Monfort, J. J., Guillera-Arroita, G., \& Wintle, B. A. (2014).
Imperfect detection impacts the performance of species distribution
models. \emph{Global Ecology and Biogeography}, \emph{23}(4), 504--515.
\url{https://doi.org/10.1111/geb.12138}

\bibitem[\citeproctext]{ref-lenoir2015}
Lenoir, J., \& Svenning, J.-C. (2015). Climate-related range shifts
{\textendash} a global multidimensional synthesis and new research
directions. \emph{Ecography}, \emph{38}(1), 15--28.
\url{https://doi.org/10.1111/ecog.00967}

\bibitem[\citeproctext]{ref-lesmeister2015}
Lesmeister, D. B., Nielsen, C. K., Schauber, E. M., \& Hellgren, E. C.
(2015). Spatial and temporal structure of a mesocarnivore guild in
midwestern north America: Midwestern Carnivore Guild Structure.
\emph{Wildlife Monographs}, \emph{191}(1), 1--61.
\url{https://doi.org/10.1002/wmon.1015}

\bibitem[\citeproctext]{ref-lonergan2014}
Lonergan, M. (2014). Data availability constrains model complexity,
generality, and utility: A response to evans {\emph{et al.}}
\emph{Trends in Ecology \& Evolution}, \emph{29}(6), 301--302.
\url{https://doi.org/10.1016/j.tree.2014.03.005}

\bibitem[\citeproctext]{ref-mackenzie2004}
MacKenzie, D. I., \& Bailey, L. L. (2004). Assessing the fit of
site-occupancy models. \emph{Journal of Agricultural, Biological, and
Environmental Statistics}, \emph{9}(3), 300--318.
\url{https://doi.org/10.1198/108571104X3361}

\bibitem[\citeproctext]{ref-mackenzie2003}
MacKenzie, D. I., Nichols, J. D., Hines, J. E., Knutson, M. G., \&
Franklin, A. B. (2003). Estimating site occupancy, colonization, and
local extinction when a species is detected imperfectly. \emph{Ecology},
\emph{84}(8), 2200--2207. \url{https://doi.org/10.1890/02-3090}

\bibitem[\citeproctext]{ref-mackenzie2002}
MacKenzie, D. I., Nichols, J. D., Lachman, G. B., Droege, S., Andrew
Royle, J., \& Langtimm, C. A. (2002). Estimating Site Occupancy Rates
When Detection Probabilities Are Less Than One. \emph{Ecology},
\emph{83}(8), 2248--2255.
\url{https://doi.org/10.1890/0012-9658(2002)083\%5B2248:ESORWD\%5D2.0.CO;2}

\bibitem[\citeproctext]{ref-mangan2019}
Mangan, A. O., Chestnut, T., Vogeler, J. C., Breckheimer, I. K., King,
W. M., Bagnall, K. E., \& Dugger, K. M. (2019). Barred Owls reduce
occupancy and breeding propensity of Northern Spotted Owl in a
Washington old-growth forest. \emph{The Condor}, \emph{121}(3), duz031.
\url{https://doi.org/10.1093/condor/duz031}

\bibitem[\citeproctext]{ref-marescot2020}
Marescot, L., Lyet, A., Singh, R., Carter, N., \& Gimenez, O. (2020).
Inferring wildlife poaching in southeast Asia with multispecies dynamic
occupancy models. \emph{Ecography}, \emph{43}(2), 239--250.
\url{https://doi.org/10.1111/ecog.04536}

\bibitem[\citeproctext]{ref-mcclintock2010}
McClintock, B. T., Bailey, L. L., Pollock, K. H., \& Simons, T. R.
(2010). Unmodelred observation error induces bias when inferring
patterns and dynamics of species occurrence via aural detections.
\emph{Ecology}, \emph{91}(8), 2446--2454.
\url{https://www.jstor.org/stable/27860809}

\bibitem[\citeproctext]{ref-mcgowan2020}
McGowan, C. P., Angeli, N., Beisler, W., Snyder, C. W., Rankin, N. M.,
Woodrow, J., Wilson, J., Rivenbark, E., Schwarzer, A., Hand, C.,
Anthony, R. M., Griffin, R., Barrett, K., Haverland, A., Roach, N.,
Schneider, T., Smith, A. J., Smith, F., Tolliver, J., \& Watts, B. D.
(2020). Linking monitoring and data analysis to predictions and
decisions for the range-wide eastern black rail status assessment.
\emph{Endangered Species Research}, \emph{43}, 209--222.
\url{https://doi.org/10.3354/esr01063}

\bibitem[\citeproctext]{ref-merow2014}
Merow, C., Smith, M. J., Edwards Jr, T. C., Guisan, A., McMahon, S. M.,
Normand, S., Thuiller, W., Wüest, R. O., Zimmermann, N. E., \& Elith, J.
(2014). What do we gain from simplicity versus complexity in species
distribution models? \emph{Ecography}, \emph{37}(12), 1267--1281.
\url{https://doi.org/10.1111/ecog.00845}

\bibitem[\citeproctext]{ref-merow2013}
Merow, C., Smith, M. J., \& Silander Jr, J. A. (2013). A practical guide
to MaxEnt for modeling species' distributions: what it does, and why
inputs and settings matter. \emph{Ecography}, \emph{36}(10), 1058--1069.
\url{https://doi.org/10.1111/j.1600-0587.2013.07872.x}

\bibitem[\citeproctext]{ref-miller2015}
Miller, D. A. W., Bailey, L. L., Grant, E. H. C., McClintock, B. T.,
Weir, L. A., \& Simons, T. R. (2015). Performance of species occurrence
estimators when basic assumptions are not met: a test using field data
where true occupancy status is known. \emph{Methods in Ecology and
Evolution}, \emph{6}(5), 557--565.
\url{https://doi.org/10.1111/2041-210X.12342}

\bibitem[\citeproctext]{ref-miller2011}
Miller, D., Nichols, J., Mcclintock, B., Grant, E., Bailey, L., \& Weir,
L. (2011). Improving occupancy estimation when two types of
observational error occur: Non-detection and species misidentification.
\emph{Ecology}, \emph{92}, 1422--1428.
\url{https://doi.org/10.2307/23035095}

\bibitem[\citeproctext]{ref-miller2023}
Miller, K. E., \& Brown, J. L. (2023). Maximizing nest box monitoring
effort to detect american kestrel site occupancy. \emph{Journal of
Raptor Research}, \emph{57}(2), 176--184.
\url{https://doi.org/10.3356/JRR-22-46}

\bibitem[\citeproctext]{ref-muxf6lle2022}
Mölle, J. P., Kleiven, E. F., Ims, R. A., \& Soininen, E. M. (2022).
Using subnivean camera traps to study arctic small mammal community
dynamics during winter. \emph{Arctic Science}, \emph{8}(1), 183--199.
\url{https://doi.org/10.1139/as-2021-0006}

\bibitem[\citeproctext]{ref-moreira2016}
Moreira, L. F. B., Moura, R. G., \& Maltchik, L. (2016). Stop and ask
for directions: factors affecting anuran detection and occupancy in
Pampa farmland ponds. \emph{Ecological Research}, \emph{31}(1), 65--74.
\url{https://doi.org/10.1007/s11284-015-1316-9}

\bibitem[\citeproctext]{ref-mores2020}
Mores, G. B., Schuler-Faccini, L., Hasenack, H., Fetzer, L. O., Souza,
G. D., \& Ferraz, G. (2020). Site Occupancy by Aedes aegypti in a
Subtropical City is Most Sensitive to Control during Autumn and Winter
Months. \emph{The American Journal of Tropical Medicine and Hygiene},
\emph{103}(1), 445--454. \url{https://doi.org/10.4269/ajtmh.19-0366}

\bibitem[\citeproctext]{ref-morin2020}
Morin, D. J., Yackulic, C. B., Diffendorfer, J. E., Lesmeister, D. B.,
Nielsen, C. K., Reid, J., \& Schauber, E. M. (2020). Is your ad hoc
model selection strategy affecting your multimodel inference?
\emph{Ecosphere}, \emph{11}(1), e02997.
\url{https://doi.org/10.1002/ecs2.2997}

\bibitem[\citeproctext]{ref-mortelliti2007}
Mortelliti, A., \& Boitani, L. (2007). Estimating species' absence,
colonization and local extinction in patchy landscapes: an application
of occupancy models with rodents. \emph{Journal of Zoology},
\emph{273}(3), 244--248.
\url{https://doi.org/10.1111/j.1469-7998.2007.00320.x}

\bibitem[\citeproctext]{ref-nichols2007}
Nichols, J. D., Hines, J. E., Mackenzie, D. I., Seamans, M. E., \&
Gutiérrez, R. J. (2007). Occupancy estimation and modeling with multiple
states and state uncertainty. \emph{Ecology}, \emph{88}(6), 1395--1400.
\url{https://www.jstor.org/stable/27651247}

\bibitem[\citeproctext]{ref-olson2005}
Olson, G. S., Anthony, R. G., Forsman, E. D., Ackers, S. H., Loschl, P.
J., Reid, J. A., Dugger, K. M., Glenn, E. M., \& Ripple, W. J. (2005).
Modeling of Site Occupancy Dynamics for Northern Spotted Owls, with
Emphasis on the Effects of Barred Owls. \emph{The Journal of Wildlife
Management}, \emph{69}(3), 918--932.
\url{https://doi.org/10.2193/0022-541X(2005)069\%5B0918:MOSODF\%5D2.0.CO;2}

\bibitem[\citeproctext]{ref-otto2013}
Otto, C. R. V., Bailey, L. L., \& Roloff, G. J. (2013). Improving
species occupancy estimation when sampling violates the closure
assumption. \emph{Ecography}, \emph{36}(12), 1299--1309.
\url{https://doi.org/10.1111/j.1600-0587.2013.00137.x}

\bibitem[\citeproctext]{ref-otto2012}
Otto, C. R. V., \& Roloff, G. J. (2012). Songbird response to green-tree
retention prescriptions in clearcut forests. \emph{Forest Ecology and
Management}, \emph{284}, 241--250.
\url{https://doi.org/10.1016/j.foreco.2012.07.016}

\bibitem[\citeproctext]{ref-padilla-torres2013}
Padilla-Torres, S. D., Ferraz, G., Luz, S. L. B., Zamora-Perea, E., \&
Abad-Franch, F. (2013). Modeling Dengue Vector Dynamics under Imperfect
Detection: Three Years of Site-Occupancy by Aedes aegypti and Aedes
albopictus in Urban Amazonia. \emph{PLoS ONE}, \emph{8}(3), e58420.
\url{https://doi.org/10.1371/journal.pone.0058420}

\bibitem[\citeproctext]{ref-peach2019}
Peach, M. A., Cohen, J. B., Frair, J. L., Zuckerberg, B., Sullivan, P.,
Porter, W. F., \& Lang, C. (2019). Value of protected areas to avian
persistence across 20 years of climate and land{-}use change.
\emph{Conservation Biology}, \emph{33}(2), 423--433.
\url{https://doi.org/10.1111/cobi.13205}

\bibitem[\citeproctext]{ref-pendleton2022}
Pendleton, D. E., Tingley, M. W., Ganley, L. C., Friedland, K. D., Mayo,
C., Brown, M. W., McKenna, B. E., Jordaan, A., \& Staudinger, M. D.
(2022). Decadal-scale phenology and seasonal climate drivers of
migratory baleen whales in a rapidly warming marine ecosystem.
\emph{Global Change Biology}, \emph{28}(16), 4989--5005.
\url{https://doi.org/10.1111/gcb.16225}

\bibitem[\citeproctext]{ref-pitman2017}
Pitman, R. T., Fattebert, J., Williams, S. T., Williams, K. S., Hill, R.
A., Hunter, L. T. B., Robinson, H., Power, J., Swanepoel, L., Slotow,
R., \& Balme, G. A. (2017). Cats, connectivity and conservation:
incorporating data sets and integrating scales for wildlife management.
\emph{Journal of Applied Ecology}, \emph{54}(6), 1687--1698.
\url{https://doi.org/10.1111/1365-2664.12851}

\bibitem[\citeproctext]{ref-pollentier2021}
Pollentier, C. D., Hardy, M. A., Lutz, R. S., Hull, S. D., \&
Zuckerberg, B. (2021). Gobbling across landscapes: Eastern wild turkey
distribution and occupancy{\textendash}habitat associations.
\emph{Ecology and Evolution}, \emph{11}(24), 18248--18270.
\url{https://doi.org/10.1002/ece3.8419}

\bibitem[\citeproctext]{ref-riddell2021}
Riddell, E. A., Iknayan, K. J., Hargrove, L., Tremor, S., Patton, J. L.,
Ramirez, R., Wolf, B. O., \& Beissinger, S. R. (2021). Exposure to
climate change drives stability or collapse of desert mammal and bird
communities. \emph{Science}, \emph{371}(6529), 633--636.
\url{https://doi.org/10.1126/science.abd4605}

\bibitem[\citeproctext]{ref-risk2011}
Risk, B. B., Valpine, P. de, \& Beissinger, S. R. (2011). A
robust-design formulation of the incidence function model of
metapopulation dynamics applied to two species of rails. \emph{Ecology},
\emph{92}(2), 462--474. \url{https://doi.org/10.1890/09-2402.1}

\bibitem[\citeproctext]{ref-rota2009}
Rota, C. T., Fletcher Jr, R. J., Dorazio, R. M., \& Betts, M. G. (2009).
Occupancy estimation and the closure assumption. \emph{Journal of
Applied Ecology}, \emph{46}(6), 1173--1181.
\url{https://doi.org/10.1111/j.1365-2664.2009.01734.x}

\bibitem[\citeproctext]{ref-rowe2019}
Rowe, J. C., Duarte, A., Pearl, C. A., McCreary, B., Galvan, S. K.,
Peterson, J. T., \& Adams, M. J. (2019). Disentangling effects of
invasive species and habitat while accounting for observer error in a
long{-}term amphibian study. \emph{Ecosphere}, \emph{10}(4).
\url{https://doi.org/10.1002/ecs2.2674}

\bibitem[\citeproctext]{ref-royle2006}
Royle, J. A., \& Link, W. A. (2006). GENERALIZED SITE OCCUPANCY MODELS
ALLOWING FOR FALSE POSITIVE AND FALSE NEGATIVE ERRORS. \emph{Ecology},
\emph{87}(4), 835--841.
\url{https://doi.org/10.1890/0012-9658(2006)87\%5B835:GSOMAF\%5D2.0.CO;2}

\bibitem[\citeproctext]{ref-scott2015}
Scott, P. A., \& Rissler, L. J. (2015). Integrating dynamic occupancy
modeling and genetics to infer the status of the imperiled flattened
musk turtle. \emph{Biological Conservation}, \emph{192}, 294--303.
\url{https://doi.org/10.1016/j.biocon.2015.10.004}

\bibitem[\citeproctext]{ref-stevens2019}
Stevens, B. S., \& Conway, C. J. (2019). Identifying important military
installations for continental-scale conservation of marsh bird breeding
habitat. \emph{Journal of Environmental Management}, \emph{252}, 109664.
\url{https://doi.org/10.1016/j.jenvman.2019.109664}

\bibitem[\citeproctext]{ref-stewart2023}
Stewart, P. S., Stephens, P. A., Hill, R. A., Whittingham, M. J., \&
Dawson, W. (2023). Model selection in occupancy models: Inference versus
prediction. \emph{Ecology}, \emph{104}(3), e3942.
\url{https://doi.org/10.1002/ecy.3942}

\bibitem[\citeproctext]{ref-urban2023}
Urban, M. C., Nadeau, C. P., \& Giery, S. T. (2023). Using mechanistic
insights to predict the climate-induced expansion of a key aquatic
predator. \emph{Ecological Monographs}, \emph{93}(3), e1575.
\url{https://doi.org/10.1002/ecm.1575}

\bibitem[\citeproctext]{ref-valavi2023}
Valavi, R., Elith, J., Lahoz-Monfort, J. J., \& Guillera-Arroita, G.
(2023). Flexible species distribution modelling methods perform well on
spatially separated testing data. \emph{Global Ecology and
Biogeography}, \emph{32}(3), 369--383.
\url{https://doi.org/10.1111/geb.13639}

\bibitem[\citeproctext]{ref-valente2017}
Valente, J. J., Hutchinson, R. A., \& Betts, M. G. (2017).
Distinguishing distribution dynamics from temporary emigration using
dynamic occupancy models. \emph{Methods in Ecology and Evolution},
\emph{8}(12), 1707--1716. \url{https://doi.org/10.1111/2041-210X.12840}

\bibitem[\citeproctext]{ref-warrier2020}
Warrier, R., Noon, B. R., \& Bailey, L. (2020). Agricultural lands offer
seasonal habitats to tigers in a human{-}dominated and fragmented
landscape in India. \emph{Ecosphere}, \emph{11}(7).
\url{https://doi.org/10.1002/ecs2.3080}

\bibitem[\citeproctext]{ref-wood2020}
Wood, C. M., Gutiérrez, R. J., Keane, J. J., \& Peery, M. Z. (2020).
Early detection of rapid Barred Owl population growth within the range
of the California Spotted Owl advises the Precautionary Principle.
\emph{The Condor}, \emph{122}(1), duz058.
\url{https://doi.org/10.1093/condor/duz058}

\bibitem[\citeproctext]{ref-zuckerberg2011}
Zuckerberg, B., Bonter, D. N., Hochachka, W. M., Koenig, W. D.,
DeGaetano, A. T., \& Dickinson, J. L. (2011). Climatic constraints on
wintering bird distributions are modified by urbanization and weather:
Wintering birds, weather, food, and climate. \emph{Journal of Animal
Ecology}, \emph{80}(2), 403--413.
\url{https://doi.org/10.1111/j.1365-2656.2010.01780.x}

\bibitem[\citeproctext]{ref-zurell2020}
Zurell, D., Franklin, J., König, C., Bouchet, P. J., Dormann, C. F.,
Elith, J., Fandos, G., Feng, X., Guillera-Arroita, G., Guisan, A.,
Lahoz-Monfort, J. J., Leitão, P. J., Park, D. S., Peterson, A. T.,
Rapacciuolo, G., Schmatz, D. R., Schröder, B., Serra-Diaz, J. M.,
Thuiller, W., \ldots{} Merow, C. (2020). A standard protocol for
reporting species distribution models. \emph{Ecography}, \emph{43}(9),
1261--1277. \url{https://doi.org/10.1111/ecog.04960}

\end{CSLReferences}

\end{document}

### 2024-09-20 _manuscript/Manuscript.qmd

@@
 Where 94% of frequentist models perform some manner of model selection to determine covariate inclusions for their models, only 36% of Bayesian models do so, with the majority instead fitting a single model defined *a priori*.
-Methods used for model selection vary considerably both within and between the two modelling frameworks.
+Methods used for model selection vary considerably both within and between the two modelling frameworks (@tbl-modelling).
 For frequentist models, the most popular and conventional approach to model selection as used in 43% of articles involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score.
@@
 The remainder of frequentist studies (9%) use a variety of approaches, such as fitting simpler models like single season occupancy models to identify the most informative terms to use in a DOM.
-Across all frequentist implementations, 71% of articles summaruise a final subset of models using multi-model inference and model-averaging with AIC weights [@burnham2004].
+Across all frequentist implementations, 71% of articles summarise a final subset of models using multi-model inference and model-averaging with AIC weights [@burnham2004].
 
 Those Bayesian models which *do* perform model selection tend to take various approaches, with largely idiosyncratic methods across these studies.
-While direct comparison of model fit in Bayesian methods is rare, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data.
+While direct comparison of model fit is rare in Bayesian methods, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data.
 Another approach used by @cook2022 fit a global model including all covariates before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the model.
@@
 The model selection methods identified in our review are designed only to identify the best model of those considered, and do not necessarily guarantee that selected models are well-fit to the data.
-Assessment of the performance of DOMs generally requires additional steps, and the best ways to achieve this are not necessarily clear.
+Assessment of the performance of DOMs generally requires additional steps, and the best ways to achieve this are not yet clear.
 There is no broadly-accepted goodness-of-fit test for dynamic occupancy models, although @mackenzie2004 describes an approach for single-season occupancy models using a parametric bootstrap which can been extended to DOMs; this test is implemented in the *AICcModAvg* and *unmarked* R packages.
@@
 Studies in our sample were classified according to their objectives as stated by authors to explore the use-cases which the DOM has been applied to (@fig-objectives).
-These objective categories were non-exclusive, and many studies fell into multiple categories; the frequency at which studies pursued each of these objectives is shown in @fig-objectives A.
-Across our study period DOMs have been frequently employed to estimate trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions.
+These objective categories were non-exclusive, and many studies fell into multiple categories.
+Across the study period the DOM has been frequently employed to estimate trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions.
 Within each of these broad categories lies even more variation in objectives, emphasising the DOM's flexibility as a tool for making ecological inference in diverse contexts.
@@
 
-The objectives of studies fitting DOMs strongly inform approaches to the model building process, particularly with respect to covariate selection.
+The objectives of studies fitting a DOM strongly informs approaches to model building, particularly with respect to covariate selection.
 @fig-objectives B shows the quantity of covariates considered for models applied to each of our objective categories, and the differences between objectives are apparent.
@@
 This precludes any prescription of the 'best' way to build a model, however, there are still important discussions to be had on decisions made in the modelling process.
-One aspect of fitting DOMs meriting broader discussion revolves around 'model complexity,' and how much must be incorporated into models to reliable model species occupancy under different contexts and use cases.
-Complexity is a broad term which encompasses many aspects of a model [@merow2014], and opinions on simplicity versus complexity in ecological models can be divisive.
-Where some advocate for the simplest possible models, arguing that they are most generaliseable; others insist that overly-simple models cannot adequately represent the most important drivers in a system [@evans2013; @lonergan2014].
-By their nature DOMs are somewhat more complex than simpler models for studying occupancy thanks to their hierarchical structure, attributes which are necessary to control for detectability and to capture occupancy dynamics.
+One aspect of fitting DOMs meriting broader discussion revolves around 'model complexity,' and how much must be incorporated into models to reliably model species occupancy under different contexts and use cases.
+Complexity is a broad term which encompasses many aspects of a model [@merow2014], and opinions on simplicity versus complexity in ecological models are diverse.
+Where some advocate for the simplest possible models, arguing that they are most generalisable; others insist that overly-simple models cannot adequately represent the most important drivers in a system [@evans2013; @lonergan2014].
+By their nature the DOM is somewhat more complex than simpler models for studying occupancy thanks to their hierarchical structure, which is necessary to control for detectability and to capture occupancy dynamics.
 Within this structure, however, further complexity is to some degree up to the modeller: one can choose how many covariates to consider for inclusion on the various parameters, and how to represent the nature of the relationship between those covariates and parameters.
-Research from SDMs indicates that allowing for more complex relationships can improve model performance in predicting occupancy [@valavi2023], an increasingly popular use-case for DOMs.
-Within the DOM framework, there are exciting developments on that front — @joseph2020 presents a novel neural-network occupancy model which allows for exponentially higher levels of complexity and may offer improved performance for prediction-oriented studies.
+Research from SDMs indicates that allowing for more complex relationships can improve model performance in predicting occupancy [@valavi2023], an increasingly popular use-case for the DOM.
+Within the DOM framework, there are promising developments on that front — for example, @joseph2020 presents a novel neural-network occupancy model which allows for exponentially higher levels of complexity and may offer improved performance for prediction-oriented studies.
 
-Generally speaking, covariate selection seems to be a particularly important area for further investigations into building DOMs.
+Covariate selection seems to be a particularly important area for further investigations into building DOMs.
 In our review, we see little consensus around which approaches are most applicable for any given use case, and existing work on covariate selection in DOMs raises concerns on whether common methods always produce the most suitable models.
-This is true for both frequentist and Bayesian implementations, and comparative research on covariate selection under both frameworks may help to inform model users on which method may be most appropriate for their use-cases.
-In a similar vein, the low number of articles in our review which conducted model evaluation or assessed model fit raises different questions.
+This is true for both frequentist and Bayesian implementations, and comparative research on covariate selection under both frameworks may help to inform model users on which method is most appropriate for their use-cases.
+In a similar vein, the low number of articles in our review that conducted model evaluation or assessed model fit raises different questions.
 While the appropriate method of model evaluation may depend on data availability and research objectives, assessing models by some method is generally important to understand how reliable model outputs may be [@araújo2005; @guisan2005].
-Existing uncertainties around whether current methods are suitable for the task surely discourage users from calculating these metrics, and further research is needed to establish trusted practices for assessing the quality of DOMs.
+Existing uncertainties around whether current methods are suitable for the task likely discourage users from calculating these metrics, and further research is needed to establish trusted practices for assessing the quality of DOMs.
 
 These priorities are particularly important given the frequent applied objectives of DOM users, tackling challenges which include assessing critically endangered species [@carvalho2023], guiding public health management of disease vectors [@moreira2016], and tracking rapidly developing biological invasions [@wood2020].
-DOMs are well suited to these situations, and it is to be expected that as these types of applications are more commonly attempted understanding the sensitivity of model outputs to decisions made in the model fitting process becomes increasingly important.
+The DOM is well suited to these situations, and it is to be expected that as these types of applications are more commonly attempted understanding the sensitivity of model outputs to decisions made in the model fitting process becomes increasingly important.
 In the two decades since the publication of @mackenzie2003 the dynamic occupancy model has become a widely used tool for ecological inference, with numerous extensions to the modelling framework further broadening the scope of questions and use-cases for which it may be applied.
-Given their increasing popularity, further research and guidelines around issues of model building may help to make the DOM format more accessible to newcomers and support confidence in interpretation of models.
-Parallels to existing work in the SDM literature on guidelines for reporting and modelling standards could be valuable contributions in achieving these aims [@zurell2020; @araújo2019] while specifically addressing model attributes unique to DOMs.
+Given their increasing popularity, further research and guidelines around issues of model building may help to make the DOM more accessible to newcomers and support confidence in interpretation of models.
+Parallels to existing work in the SDM literature on guidelines for reporting and modelling standards could be valuable contributions in achieving these aims [@zurell2020; @araújo2019] could provide useful examples in developing methods and guidelines specifically focussed on the DOM.

### 2024-09-23 Analysis/Analysis.qmd
Clear archived files

--- 2024-08-25 Analysis/ReviewFigures/Final_Figures.qmd
@@
 library(readxl)
+library(ggtext)
 ```
@@
   select(Articles, Taxa = `Taxa keywords`, Status) |>
-  mutate(Taxa = case_when(Taxa == "NA" ~ "Other", .default = Taxa)) |>
+  mutate(Taxa = case_when(Taxa %in% c("NA", "Fish") ~ "Other", .default = Taxa)) |>
   mutate(Taxa = fct(Taxa, levels = rev(c("Bird", "Mammal", "Herptile",
@@
   geom_col(aes(x = Articles, y = Taxa, fill = Status)) +
-  scale_fill_manual("", values = c("gray40", "plum4", "aquamarine4", "firebrick4")) +
+  scale_fill_manual("", values = c("gray70", "plum4", "aquamarine4", "goldenrod2")) +
   labs(x = "Articles", y = "") +
@@
 
-ggsave("Manuscript/Figures/TaxaPlot.png",
+ggsave("Figures/TaxaPlot.png",
        focalTaxaPlot, width = 10, height = 4, bg='transparent')
 
+sheet_FocalTaxa |>
+  select(`Review ID`, `Multispecies method`) |>
+  distinct() |>
+  summarise(Articles = n(), .by = `Multispecies method`)
 ```
@@
 Determine estimated coverage
+
+```{r strata hit rates}
+sheet_ArticleData |>
+  pull(`Review ID`) |>
+  length()
+
+randomisationSheets |>
+  summarise(N = n(), .by = c(Year, Strata, Included)) |>
+  pivot_wider(names_from = Included, values_from = N, values_fill = 0) |>
+  summarise(HitRate = sum(YES)/(sum(NO)+sum(YES)),
+            .by = Strata)
+
+  randomisationSheets |>
+  summarise(N = n(), .by = c(Year, Strata, Included)) |>
+  pivot_wider(names_from = Included, values_from = N, values_fill = 0) |>
+  mutate(HitRate = sum(YES)/(sum(NO)+sum(YES)),
+         .by = Strata) |>
+  summarise(HitRate = mean(HitRate), `NA` = sum(`NA`),
+            .by = Year) |>
+  mutate(EstRemaining = `NA`*HitRate) |>
+  summarise(`NA` = sum(`NA`), EstRemaining = sum(EstRemaining))
+```
 
@@
                     values = c("aquamarine4", "gray70")) +
+  scale_y_continuous(expand = c(0,0), limits = c(0, 65)) +
   labs(x = "", y = "Number of articles") +
   theme(panel.grid.major.x = element_blank(),
-        legend.position = "top", axis.text = element_text(size = 12),
-        axis.title = element_text(size = 12), legend.text = element_text(size = 12))
+        panel.grid.minor.x = element_blank(),
+        legend.position = "bottom", axis.text = element_text(size = 12),
+        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
+        axis.title = element_text(size = 12), legend.text = element_text(size = 12),
+        strip.text.x = element_text(size = 12)) +
+  facet_grid(cols = vars(Strata), scales = "free_x", switch = "x")
 coveragePlot
-ggsave(plot = coveragePlot, "Manuscript/Figures/CoveragePlot.jpeg",
+ggsave(plot = coveragePlot, "Figures/CoveragePlot.jpeg",
        width = 8, height = 4)
@@
 studyArea_Map
-ggsave(plot = studyArea_Map, "Manuscript/Figures/StudiesMap.png")
+ggsave(plot = studyArea_Map, "Figures/StudiesMap.png")
+
+sheet_StudyArea |>
+  select(`Review ID`, Country) |>
+  separate_longer_delim(Country, ", ") |>
+  distinct() |>
+  summarise(Articles = n(), .by = Country) |>
+  arrange(-Articles)
+
 ```
@@
   geom_col(aes(x = Size, y = Articles),
-           fill = "aquamarine4") +
+           fill = "aquamarine4", colour = "white",
+           width = 1, linewidth = 2) +
   labs(x = "", y = "") +
-  scale_x_discrete(labels = c("0-1", "1-10", "10-100",
-                              "100-\n1000", "1000-\n10000",
-                              "10000-\n100000", "100000-\n1000000",
-                              "1000000+")) +
+  scale_x_discrete(labels = c("< 1", "1 - 10", "10 - 10<sup>2</sup>",
+                              "10<sup>2</sup> - 10<sup>3</sup>",
+                              "10<sup>3</sup> - 10<sup>4</sup>",
+                              "10<sup>4</sup> - 10<sup>5</sup>",
+                              "10<sup>5</sup> - 10<sup>6</sup>",
+                              "10<sup>6</sup> +")) +
   theme(panel.grid = element_blank(),
-        axis.text.x = element_text(size = 20,
-                                   angle = 90,
+        axis.text.x = element_markdown(size = 20, angle = 90,
                                        vjust = 0.5, hjust = 1),
@@
 
-ggsave("Manuscript/Figures/StudyAreaPlot.png",
-       studyAreaPlot, width = 8, height = 4, bg='transparent') 
+ggsave("Figures/StudyAreaPlot.png",
+       studyAreaPlot, width = 7, height = 4, bg='transparent') 
 ```
@@
 # Data collection
+
+survey methods
+
+```{r survey methods}
+sheet_DataCollection |>
+  select(`Review ID`, `Capture method`) |>
+  separate_longer_delim(`Capture method`, ", ") |>
+  distinct() |>
+  summarise(Articles = n(), .by = `Capture method`)
+
+sheet_DataCollection |>
+  select(`Review ID`, Citizen) |>
+  distinct() |>
+  summarise(Articles = n(), .by = Citizen)
+
+```
 
@@
   ggplot() +
-  geom_histogram(aes(y = StudyMonths), 
-                 binwidth = 0.2, fill = "aquamarine4", colour = "white") +
-  geom_hline(aes(yintercept = median(StudyMonths)), 
-             colour = "goldenrod3", linetype = 2, linewidth = 2) +
-  labs(y = "", x = "") +
-  scale_y_continuous(transform = "log10",
+  geom_histogram(aes(x = StudyMonths), 
+                 binwidth = 0.2, fill = "aquamarine4", colour = "white",
+                 linewidth = 2) +
+  geom_vline(aes(xintercept = median(StudyMonths)), 
+             colour = "goldenrod2", linetype = 1, linewidth = 2) +
+  labs(x = "", y = "Study\nduration") +
+  scale_x_continuous(transform = "log10",
                      limits = c(0.5,725),
                      breaks = c(1, 6, 12, 60, 
-                                120, 240, 480, 720), 
+                                120, 240, 480), 
                      labels = c("1 month", "6 months", "1 year", "5 years", 
-                                "10 years", "20 years", "40 years", "60 Years")) +
+                                "10 years", "20 years", "40 years"),
+                     expand = c(0,0)) +
+  scale_y_continuous(expand = c(0,0)) +
   theme(panel.grid = element_blank(),
@@
         plot.background = element_rect(fill='transparent', color=NA),
-        axis.text.y = element_text(size = 23),
-        axis.text.x = element_blank(), axis.ticks.x = element_blank())
+        axis.text.x = element_text(size = 23, angle = 90,
+                                   vjust = 0.5, hjust = 1),
+        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
+        axis.line.x = element_line(), 
+        axis.title.y = element_text(size = 35, colour = "aquamarine4"))
 studyDurationPlot
-ggsave(plot = studyDurationPlot, filename = "Manuscript/Figures/Duration.png",
-       height = 8, width = 4)
+ggsave(plot = studyDurationPlot, filename = "Figures/Duration.png",
+       height = 3, width = 12)
+
+sheet_DataCollection |>
+  mutate(StartDay = as.numeric(as.Date(`Start month`)),
+         EndDay = as.numeric(as.Date(`End month`))) |>
+  mutate(StudyDuration = EndDay - StartDay) |>
+  mutate(StudyMonths = (StudyDuration/30)) |>
+  mutate(StudyMonths = case_when(StudyMonths == 0 ~ 1,
+                                 .default = StudyMonths)) |>
+  summarise(StudyMonths = mean(StudyMonths, na.rm = TRUE),
+            .by = `Review ID`) |>
+  pull(StudyMonths) |>
+  median(na.rm = TRUE)
 ```
@@
   ggplot() +
-  geom_histogram(aes(y = `Site quantity`),
-                 binwidth = 0.25, fill = "aquamarine4", colour = "white") +
-  geom_hline(aes(yintercept = median(`Site quantity`, na.rm = TRUE)), 
-             colour = "goldenrod3", linetype = 2, linewidth = 2) +
-  scale_y_continuous(transform = "log10",
+  geom_histogram(aes(x = `Site quantity`),
+                 binwidth = 0.25, fill = "aquamarine4", colour = "white",
+                 linewidth = 2) +
+  geom_vline(aes(xintercept = median(`Site quantity`, na.rm = TRUE)), 
+             colour = "goldenrod2", linetype = 1, linewidth = 2) +
+  scale_x_continuous(transform = "log10",
                      limits = c(5, 10000),
                      breaks = c(10, 25, 50, 100, 250, 500,
-                                1000, 2500, 5000, 10000)) +
-  scale_x_continuous(expand = c(0,0)) +
-  labs(x = "", y = "") +
+                                1000, 2500, 5000),
+                     expand = c(0,0)) +
+  scale_y_continuous(expand = c(0,0)) +
+  labs(y = "Site\nquantity", x = "") +
   theme(panel.grid = element_blank(),
@@
         plot.background = element_rect(fill='transparent', color=NA),
-        axis.text.y = element_text(size = 23),
-        axis.text.x = element_blank(), axis.ticks.x = element_blank())
+        axis.text.x = element_text(size = 23, angle = 90,
+                                   vjust = 0.5, hjust = 1),
+        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
+        axis.line.x = element_line(), 
+        axis.title.y = element_text(size = 35, colour = "aquamarine4"))
 siteQuantityPlot
-ggsave(plot = siteQuantityPlot, filename = "Manuscript/Figures/Sites.png",
-       height = 8, width = 4)
+ggsave(plot = siteQuantityPlot, filename = "Figures/Sites.png",
+       height = 3, width = 12)
+
+sheet_DataCollection |>
+  mutate(`Site quantity` = as.numeric(`Site quantity`)) |>
+  summarise(Sites = mean(`Site quantity`, na.rm = TRUE), 
+            .by = `Review ID`) |>
+  pull(Sites) |>
+  median(na.rm = TRUE)
+```
+
+Combine sites and durations
+
+```{r site and duration combination}
+combinedStudyScale <- studyDurationPlot / siteQuantityPlot
+combinedStudyScale <- combinedStudyScale +
+  theme(panel.background = element_rect(fill='transparent', color=NA),
+        plot.background = element_rect(fill='transparent', color=NA))
+ggsave(plot = combinedStudyScale, filename = "Figures/StudyScale.png",
+       height = 6, width = 12)
 ```
@@
 covParamPlot
-ggsave(plot = covParamPlot, filename = "Manuscript/Figures/CovParamPlot.jpeg",
+ggsave(plot = covParamPlot, filename = "Figures/CovParamPlot.jpeg",
        width = 8, height = 4)
@@
                        too_few = "align_start") |>
+  mutate(Covariate = case_when(Covariate == "OTHER" & Category == "E" ~ "OTHER-E",
+                               Covariate == "OTHER" & Category == "S" ~ "OTHER-S",
+                               Covariate %in% c("HABT", "HABA", "LACO") ~ "HABT",
+                               Covariate %in% c("CLWE", "DISA", "DIST") ~ "CLWE",
+                               Covariate %in% c("SPAT", "GEOM", "CONN") ~ "SPAT",
+                               Covariate %in% c("TOPO", "HYDR") ~ "TOPO",
+                               Covariate %in% c("ANTH", "AGRI", "PROT") ~ "ANTH",
+                               .default = Covariate)) |>
   mutate(n_overall = length(unique(`Review ID`))) |>
@@
 
-write_csv(fullCovariateSummary, "Manuscript/Figures/CovariateTable.csv")
+write_csv(fullCovariateSummary, "Figures/CovariateTable.csv")
 ```
@@
 objectiveTrendPlot
-ggsave(plot = objectiveTrendPlot, "Manuscript/Figures/ObjectiveTrend.jpeg",
+ggsave(plot = objectiveTrendPlot, "Figures/ObjectiveTrend.jpeg",
        width = 8, height = 4)
@@
 objectiveCovariatePlot
-ggsave(plot = objectiveCovariatePlot, "Manuscript/Figures/ObjectiveCovariates.jpeg",
+ggsave(plot = objectiveCovariatePlot, "Figures/ObjectiveCovariates.jpeg",
        width = 8, height = 4)
@@
 
-ggsave(plot = stackedObjectivePlot, "Manuscript/Figures/ObjectiveFigure.jpeg",
+ggsave(plot = stackedObjectivePlot, "Figures/ObjectiveFigure.jpeg",
        width = 12, height = 8)
@@
 # Modelling
-
-Bayes/Freq/All table
-
-need:
-
-n articles in each category
-
-median covariates per core param
-
-n using each selection method
-
-n which look at fit
-
-n which evaluate
 
@@
 
-write_csv(modelling_summary, "Manuscript/Figures/ModellingSummary.csv")
+write_csv(modelling_summary, "Figures/ModellingSummary.csv")
 ```

### 2024-09-23 Analysis/Analysis.qmd
Converted most of analysis supplement

--- 2024-08-25 Analysis/ReviewFigures/Final_Figures.qmd
@@
 ---
-title: "DOM Review Stats and Figures"
+title: "Analysis and figures"
 author: "Saoirse Kelleher"
 date: today
-abstract: "Compiles key statistics and figures for use in the DOM Review manuscript"
+abstract: "Calculates key statistics and renders figures for \"*Twenty years of dynamic occupancy models: a review of applications and look to the future*\""
 abstract-title: "Summary"
-format: 
-  html:
-    theme: minty
-    toc: true
-embed-resources: true
 ---
 
-# Final DOM review figures
+## Load packages
 
-```{r load packages}
-#| include: false
+Required packages for analysis and plotting figures.
+
+```{r Load packages}
+#| message: false
+
 library(tidyverse)
@@
 library(readxl)
+library(ggtext)
 ```
 
-# Load in review pages and randomisation sheet
+## Load data
 
-Read in each page of the spreadsheet for analysis
+Two sources of data are loaded in for analysis:
 
-## Article data
+-   Randomised list of all queried articles
 
-```{r Load article data page}
-sheet_ArticleData <- read_xlsx("Spreadsheets/DOM_Spreadsheet.xlsx",
-                               sheet = "Article Data")
-```
+-   Spreadsheet of included articles
 
-## Objectives
+#### Randomised articles
 
-```{r Load objective page}
-sheet_Objectives <- read_xlsx("Spreadsheets/DOM_Spreadsheet.xlsx",
-                              sheet = "Objectives") |>
-  filter(`Review ID` != 0)
-```
+All queried articles were assigned random values in *'Randomisation.qmd'* to determine the order in which they were considered for inclusion in this review.
+These articles are read in from *'Randomisation.xlsx'*.
 
-## Focal taxa
-
-```{r Load focal taxa page}
-sheet_FocalTaxa <- read_xlsx("Spreadsheets/DOM_Spreadsheet.xlsx",
-                             sheet = "Focal Taxa") |>
-  filter(`Review ID` != 0)
-```
-
-## Study area
-
-```{r Load study area page}
-sheet_StudyArea <- read_xlsx("Spreadsheets/DOM_Spreadsheet.xlsx",
-                             sheet = "Study Area") |>
-  filter(`Review ID` != 0)
-```
-
-## Data collection
-
-```{r Load data collection page}
-sheet_DataCollection <- read_xlsx("Spreadsheets/DOM_Spreadsheet.xlsx",
-                                  sheet = "Data Collection") |>
-  filter(`Review ID` != 0)
-```
-
-## Covariates
-
-```{r Load covariates page}
-sheet_Covariates <- read_xlsx("Spreadsheets/DOM_Spreadsheet.xlsx",
-                              sheet = "Covariates") |>
-  filter(`Review ID` != 0)
-```
-
-## Modelling
-
-```{r Load modelling}
-sheet_Modelling <- read_xlsx("Spreadsheets/DOM_Spreadsheet.xlsx",
-                             sheet = "Modelling") |>
-  filter(`Review ID` != 0)
-```
-
-## Randomisation sheets
-
-```{r Load randomisation sheet}
-randomisationSheets <- list(read_xlsx("Randomisation/Final_Randomisation.xlsx", sheet = "2004-2007"),
-                            read_xlsx("Randomisation/Final_Randomisation.xlsx", sheet = "2008-2011"),
-                            read_xlsx("Randomisation/Final_Randomisation.xlsx", sheet = "2012-2015"),
-                            read_xlsx("Randomisation/Final_Randomisation.xlsx", sheet = "2016-2019"),
-                            read_xlsx("Randomisation/Final_Randomisation.xlsx", sheet = "2020-2023")) |> 
+```{r Load query sheets}
+sheet_AllArticles <- list(read_xlsx("Randomisation/Randomisation.xlsx", 
+                                    sheet = "2004-2007"),
+                          read_xlsx("Randomisation/Randomisation.xlsx", 
+                                    sheet = "2008-2011"),
+                          read_xlsx("Randomisation/Randomisation.xlsx", 
+                                    sheet = "2012-2015"),
+                          read_xlsx("Randomisation/Randomisation.xlsx", 
+                                    sheet = "2016-2019"),
+                          read_xlsx("Randomisation/Randomisation.xlsx", 
+                                    sheet = "2020-2023")) |> 
   list_rbind()
@@
 
-# Taxa data
+#### Reviewed data
 
-```{r Taxa bar plot}
-focalTaxaPlot <- sheet_FocalTaxa |>
-  mutate(Status = case_when(Status %in% c("CR", "EN", "NT", "VU", "Threatened_Other") ~ "Threatened",
-                            Status == "NA" ~ "Multiple species",
-                            Status == "Invasive" ~ "Invasive",
-                            Status == "LC" ~ "Stable")) |>
-  mutate(Status = fct(Status, levels = c("Multiple species", "Invasive",
-                                         "Stable", "Threatened"))) |>
-  summarise(Articles = n(), .by = c(`Taxa keywords`, Status)) |>
-  select(Articles, Taxa = `Taxa keywords`, Status) |>
-  mutate(Taxa = case_when(Taxa == "NA" ~ "Other", .default = Taxa)) |>
-  mutate(Taxa = fct(Taxa, levels = rev(c("Bird", "Mammal", "Herptile",
-                                     "Invertebrate", "Other")))) |>
+Each sheet from *'Review_Spreadsheet.xlsx'* is loaded individually.
 
-  ggplot() +
-  geom_col(aes(x = Articles, y = Taxa, fill = Status)) +
-  scale_fill_manual("", values = c("gray40", "plum4", "aquamarine4", "firebrick4")) +
-  labs(x = "Articles", y = "") +
-  theme(panel.grid = element_blank(),
-        #legend.position = "top",
-        legend.text = element_text(size = 20), axis.text = element_text(size = 21),
-        axis.title = element_text(size = 19), axis.ticks.y = element_blank(),
-        panel.background = element_rect(fill='transparent', color = NA),
-        plot.background = element_rect(fill='transparent', color = NA),
-        legend.background = element_rect(fill = 'transparent', color = NA))
-focalTaxaPlot
-
-ggsave("Manuscript/Figures/TaxaPlot.png",
-       focalTaxaPlot, width = 10, height = 4, bg='transparent')
-
+```{r Load review sheets}
+sheet_ArticleData <- read_xlsx("Review_Spreadsheet.xlsx",
+                               sheet = "Article Data")
+sheet_Objectives <- read_xlsx("Review_Spreadsheet.xlsx",
+                              sheet = "Objectives") 
+sheet_FocalTaxa <- read_xlsx("Review_Spreadsheet.xlsx",
+                             sheet = "Focal Taxa") 
+sheet_StudyArea <- read_xlsx("Review_Spreadsheet.xlsx",
+                             sheet = "Study Area")
+sheet_DataCollection <- read_xlsx("Review_Spreadsheet.xlsx",
+                                  sheet = "Data Collection")
+sheet_Covariates <- read_xlsx("Review_Spreadsheet.xlsx",
+                              sheet = "Covariates")
+sheet_Modelling <- read_xlsx("Review_Spreadsheet.xlsx",
+                             sheet = "Modelling")
 ```
 
-# Article data
+## Article inclusion
 
-Plot countries of first authors.
+#### Articles by query
 
-```{r Authorship countries}
-sheet_ArticleData |>
-  summarise(Articles = n(), .by = `First-author country`) |>
-  arrange(-Articles) |>
-  select(AuthorCountry = `First-author country`, Articles)
+Calculate the number of articles sourced from each query, and the proportion of articles which were included in both queries.
+
+```{r Calculate query statistics}
+sheet_AllArticles |>
+  mutate(Query = case_when(KeyTerms == TRUE & CitesMackenzie == TRUE ~ "Both",
+                           KeyTerms == FALSE & CitesMackenzie == TRUE ~ "Mackenzie",
+                           KeyTerms == TRUE & CitesMackenzie == FALSE ~ "KeyTerms")) |>
+  summarise(Articles = n(), .by = Query) |>
+  mutate(Proportion = round(Articles/sum(Articles), 2))
 ```
 
-Determine estimated coverage
+#### Total articles
 
-```{r Article coverage}
-coveragePlot <- 
-  randomisationSheets |>
+Get the number of included articles
+
+```{r Total article count}
+sheet_AllArticles |>
+  filter(Included == "YES") |>
+  nrow()
+```
+
+#### Article hit rates
+
+Calculate the hit rate for articles in each stratum, and determine how many articles are estimated to remain in each stratum based off of those hit rates.
+
+```{r Calculate article hit rates}
+# Hit rates within each stratum
+stratumHitRates <- sheet_AllArticles |>
+  summarise(Articles = n(), .by = c(Year, Strata, Included)) |>
+  pivot_wider(names_from = Included, 
+              values_from = Articles, values_fill = 0) |>
+  summarise(HitRate = round(sum(YES)/(sum(NO)+sum(YES)), 2),
+            .by = Strata)
+stratumHitRates
+
+# Number of remaining articles per year
+remainingArticles <- sheet_AllArticles |>
+  summarise(Articles = n(), .by = c(Year, Strata, Included)) |>
+  pivot_wider(names_from = Included, 
+              values_from = Articles, values_fill = 0) |>
+  mutate(HitRate = sum(YES)/(sum(NO)+sum(YES)),
+         .by = Strata) |>
+  mutate(EstRemaining = round(`NA`*HitRate, 2)) |>
+  select(Year, EstRemaining)
+remainingArticles
+
+# Estimating remaining qualified articles and total remaining articles
+remainingProportion <- data.frame(Group = c("EstRemaining", "TotalUnreviewed"),
+                                  Articles = c(sum(remainingArticles$EstRemaining),
+                                               sheet_AllArticles |>
+                                                 summarise(Articles = n(), .by = c(Year, Strata, Included)) |>
+                                                 pivot_wider(names_from = Included, 
+                                                             values_from = Articles, values_fill = 0) |>
+                                                 pull(`NA`) |> 
+                                                 sum()))
+remainingProportion
+```
+
+#### Coverage figure
+
+```{r Coverage figure}
+#| fig-width: 8
+#| fig-height: 4
+
+plot_coverage <- 
+  sheet_AllArticles |>
   summarise(N = n(), .by = c(Year, Strata, Included)) |>
@@
                     values = c("aquamarine4", "gray70")) +
+  scale_y_continuous(expand = c(0,0), limits = c(0, 65)) +
   labs(x = "", y = "Number of articles") +
   theme(panel.grid.major.x = element_blank(),
-        legend.position = "top", axis.text = element_text(size = 12),
-        axis.title = element_text(size = 12), legend.text = element_text(size = 12))
-coveragePlot
-ggsave(plot = coveragePlot, "Manuscript/Figures/CoveragePlot.jpeg",
+        panel.grid.minor.x = element_blank(),
+        legend.position = "bottom", 
+        axis.text = element_text(size = 12),
+        axis.text.x = element_blank(), 
+        axis.ticks.x = element_blank(),
+        axis.title = element_text(size = 12), 
+        legend.text = element_text(size = 12),
+        strip.text.x = element_text(size = 12)) +
+  facet_grid(cols = vars(Strata), scales = "free_x", switch = "x")
+plot_coverage
+ggsave(plot = plot_coverage, "Figures/CoveragePlot.jpeg",
        width = 8, height = 4)
@@
 
-# Study areas
+## Study areas
 
-Plot study locations
+#### Study countries table
 
-```{r Study area maps}
+```{r Study location table}
+sheet_StudyArea |>
+  select(`Review ID`, Country) |>
+  separate_longer_delim(Country, ", ") |>
+  distinct() |>
+  summarise(Articles = n(), .by = Country) |>
+  arrange(-Articles)
+```
+
+#### Unique countries
+
+```{r Unique countries}
+sheet_StudyArea |>
+  select(`Review ID`, Country) |>
+  separate_longer_delim(Country, ", ") |>
+  pull(Country) |>
+  unique() |>
+  length()
+```
+
+#### Map of study locations
+
+```{r Study location map}
 # Load country shapefile
-countries_shp <- st_read("Analysis/SpatialData/Countries/World_Countries__Generalized_.shp", quiet = TRUE) |>
+countries_shp <- st_read("Analysis/Countries_Spatial/World_Countries__Generalized_.shp", quiet = TRUE) |>
   st_transform(crs = "ESRI:54030") |>
@@
   geom_sf(data = countries_shp, colour = "gray90", fill = "gray80") +
-  geom_sf(shape = 21,fill = "aquamarine4", colour = "white", alpha = 0.7, size = 2.5) +
+  geom_sf(shape = 21, fill = "aquamarine4", colour = "white", alpha = 0.5, size = 3.5) +
   theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
@@
 studyArea_Map
-ggsave(plot = studyArea_Map, "Manuscript/Figures/StudiesMap.png")
+ggsave(plot = studyArea_Map, "Figures/StudiesMap.png")
 ```
 
-Study area size
+#### Study area size
 
-```{r Study area size histogram}
+```{r Study area size}
+#| fig-width: 7
+#| fig-height: 4
+
 studyAreaPlot <- sheet_StudyArea |>
@@
   geom_col(aes(x = Size, y = Articles),
-           fill = "aquamarine4") +
+           fill = "aquamarine4", colour = "white",
+           width = 1, linewidth = 2) +
   labs(x = "", y = "") +
-  scale_x_discrete(labels = c("0-1", "1-10", "10-100",
-                              "100-\n1000", "1000-\n10000",
-                              "10000-\n100000", "100000-\n1000000",
-                              "1000000+")) +
+  scale_x_discrete(labels = c("< 1", "1 - 10", "10 - 10<sup>2</sup>",
+                              "10<sup>2</sup> - 10<sup>3</sup>",
+                              "10<sup>3</sup> - 10<sup>4</sup>",
+                              "10<sup>4</sup> - 10<sup>5</sup>",
+                              "10<sup>5</sup> - 10<sup>6</sup>",
+                              "\\> 10<sup>6</sup>")) +
   theme(panel.grid = element_blank(),
-        axis.text.x = element_text(size = 20,
-                                   angle = 90,
+        axis.text.x = element_markdown(size = 20, angle = 90,
                                        vjust = 0.5, hjust = 1),
@@
 
-ggsave("Manuscript/Figures/StudyAreaPlot.png",
-       studyAreaPlot, width = 8, height = 4, bg='transparent') 
+ggsave("Figures/StudyAreaPlot.png",
+       studyAreaPlot, width = 7, height = 4, bg='transparent')
 ```
 
-# Data collection
+## Study taxa
 
-Plot study duration
+#### Multi-species statistics
 
-```{r plot study duration}
-studyDurationPlot <- sheet_DataCollection |>
-  mutate(StartDay = as.numeric(as.Date(`Start month`)),
-         EndDay = as.numeric(as.Date(`End month`))) |>
+```{r Multi-species stats}
+sheet_FocalTaxa |>
+  separate_longer_delim(`Multispecies method`, ", ") |>
+  select(`Review ID`, `Multispecies method`) |>
+  distinct() |>
+  mutate(TotalArticles = length(unique(`Review ID`))) |>
+  summarise(Articles = n(), 
+            .by = c(`Multispecies method`, TotalArticles)) |>
+  mutate(Proportion = round(Articles/TotalArticles, 2))
+```
+
+#### Study species
+
+```{r Focal species figure}
+#| fig-width: 10
+#| fig-height: 4
+
+focalTaxaPlot <- 
+  sheet_FocalTaxa |>
+  separate_longer_delim(`Taxa keywords`, delim = ", ") |>
+  separate_longer_delim(Status, delim = ", ") |>
+  mutate(`Taxa keywords` = case_when(`Taxa keywords` == "Fish" ~ "Other", 
+                                     .default = `Taxa keywords`)) |>
+  mutate(Status = case_when(Status %in% c("CR", "EN", "NT", "VU", "Threatened_Other") ~ "Threatened",
+                            Status %in% c("NA", "Unclear", "DD") ~ "NA",
+                            Status == "Invasive" ~ "Invasive",
+                            Status == "LC" ~ "Stable")) |>
+  mutate(Threatened = case_when("Threatened" %in% Status ~ TRUE,
+                                .default = FALSE),
+         Invasive = case_when("Invasive" %in% Status ~ TRUE,
+                              .default = FALSE),
+         Overall = TRUE,
+         .by = c(`Review ID`, `Taxa keywords`)) |>
+  select(`Review ID`, `Taxa keywords`, Threatened, Invasive, Overall) |>
+  distinct() |>
+  pivot_longer(cols = c(Threatened, Invasive, Overall), 
+               names_to = "Group", values_to = "Value") |>
+  summarise(Articles = sum(Value),
+            .by = c(`Taxa keywords`, Group)) |>
+  mutate(Group = fct(Group, levels = c("Overall", "Threatened",
+                                       "Invasive"))) |>
+  mutate(`Taxa keywords` = fct(`Taxa keywords`, 
+                               levels = rev(c("Bird", "Mammal", "Herptile",
+                                              "Invertebrate", "Other")))) |>
+  mutate(Articles = case_when(Articles == 0 ~ 0.1, .default = Articles)) |>
+
+  ggplot() +
+  geom_col(aes(x = Articles, y = `Taxa keywords`, fill = Group),
+           position = position_dodge2(reverse = TRUE), width = 0.8) +
+  scale_fill_manual("", values = c("aquamarine4", "goldenrod2", "gray70")) +
+  scale_x_continuous(expand = c(0,0)) +
+  labs(x = "Articles", y = "") +
+  theme(panel.grid = element_blank(),
+        legend.text = element_text(size = 20), axis.text = element_text(size = 21),
+        axis.title = element_text(size = 19), axis.ticks.y = element_blank(),
+        panel.background = element_rect(fill='transparent', color = NA),
+        plot.background = element_rect(fill='transparent', color = NA),
+        legend.background = element_rect(fill = 'transparent', color = NA))
+focalTaxaPlot
+
+ggsave("Figures/TaxaPlot.png",
+       focalTaxaPlot, width = 10, height = 4, bg='transparent')
+```
+
+## Data collection
+
+#### Survey method statistics
+
+```{r survey methods}
+sheet_DataCollection |>
+  separate_longer_delim(`Capture method`, ", ") |>
+  select(`Review ID`, `Capture method`) |>
+  distinct() |>
+  mutate(TotalArticles = length(unique(`Review ID`))) |>
+  summarise(Articles = n(), .by = c(`Capture method`, TotalArticles)) |>
+  mutate(Proportion = round(Articles/TotalArticles, 2))
+
+sheet_DataCollection |>
+  select(`Review ID`, Citizen) |>
+  distinct() |>
+  mutate(Citizen = case_when(Citizen == "YES" ~ TRUE,
+                             Citizen == "NO" ~ FALSE)) |>
+  pull(Citizen) |>
+  sum()
+```
+
+#### Site quantity
+
+```{r site quantity plot}
+#| fig-width: 12
+#| fig-height: 3
+
+siteQuantityPlot <- sheet_DataCollection |>
+  mutate(`Site quantity` = as.numeric(`Site quantity`)) |>
+  ggplot() +
+  geom_histogram(aes(x = `Site quantity`),
+                 binwidth = 0.25, fill = "aquamarine4", colour = "white",
+                 linewidth = 2) +
+  geom_vline(aes(xintercept = median(`Site quantity`, na.rm = TRUE)), 
+             colour = "goldenrod2", linetype = 1, linewidth = 2) +
+  scale_x_continuous(transform = "log10",
+                     limits = c(5, 10000),
+                     breaks = c(10, 25, 50, 100, 250, 500,
+                                1000, 2500, 5000),
+                     expand = c(0,0)) +
+  scale_y_continuous(expand = c(0,0), position = "right") +
+  labs(y = "Site\nquantity", x = "") +
+  theme(panel.grid = element_blank(),
+        panel.background = element_rect(fill='transparent', color=NA),
+        plot.background = element_rect(fill='transparent', color=NA),
+        axis.text.x = element_text(size = 23, angle = 90,
+                                   vjust = 0.5, hjust = 1),
+        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
+        axis.line.x = element_line(), 
+        axis.title.y = element_text(size = 35, colour = "aquamarine4"))
+siteQuantityPlot
+ggsave(plot = siteQuantityPlot, filename = "Figures/Sites.png",
+       height = 3, width = 12)
+```
+
+```{r site quantity statistics}
+sheet_DataCollection |>
+  mutate(`Site quantity` = as.numeric(`Site quantity`)) |>
+  summarise(Sites = mean(`Site quantity`, na.rm = TRUE), 
+            .by = `Review ID`) |>
+  pull(Sites) |>
+  median(na.rm = TRUE)
+```
+
+#### Study duration
+
+```{r site duration figure}
+#| fig-width: 12
+#| fig-height: 3
+
+studyDurationPlot <-
+  sheet_DataCollection |>
+    mutate(`Start month` = case_when(str_detect(`Start month`, "[:digit:]{5}") ~ `Start month`,
+                                     .default = NA),
+           `End month` = case_when(str_detect(`End month`, "[:digit:]{5}") ~ `End month`,
+                                     .default = NA)) |>
+  mutate(StartDay = as.numeric(as.Date(as.numeric(`Start month`))),
+         EndDay = as.numeric(as.Date(as.numeric(`End month`)))) |>
   mutate(StudyDuration = EndDay - StartDay) |>
@@
   ggplot() +
-  geom_histogram(aes(y = StudyMonths), 
-                 binwidth = 0.2, fill = "aquamarine4", colour = "white") +
-  geom_hline(aes(yintercept = median(StudyMonths)), 
-             colour = "goldenrod3", linetype = 2, linewidth = 2) +
-  labs(y = "", x = "") +
-  scale_y_continuous(transform = "log10",
+  geom_histogram(aes(x = StudyMonths), 
+                 binwidth = 0.2, fill = "aquamarine4", colour = "white",
+                 linewidth = 2) +
+  geom_vline(aes(xintercept = median(StudyMonths, na.rm = TRUE)), 
+             colour = "goldenrod2", linetype = 1, linewidth = 2) +
+  labs(x = "", y = "Study\nduration") +
+  scale_x_continuous(transform = "log10",
                      limits = c(0.5,725),
                      breaks = c(1, 6, 12, 60, 
-                                120, 240, 480, 720), 
+                                120, 240, 480), 
                      labels = c("1 month", "6 months", "1 year", "5 years", 
-                                "10 years", "20 years", "40 years", "60 Years")) +
+                                "10 years", "20 years", "40 years"),
+                     expand = c(0,0)) +
+  scale_y_continuous(expand = c(0,0), position = "right") +
   theme(panel.grid = element_blank(),
@@
         plot.background = element_rect(fill='transparent', color=NA),
-        axis.text.y = element_text(size = 23),
-        axis.text.x = element_blank(), axis.ticks.x = element_blank())
+        axis.text.x = element_text(size = 23, angle = 90,
+                                   vjust = 0.5, hjust = 1),
+        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
+        axis.line.x = element_line(), 
+        axis.title.y = element_text(size = 35, colour = "aquamarine4"))
 studyDurationPlot
-ggsave(plot = studyDurationPlot, filename = "Manuscript/Figures/Duration.png",
-       height = 8, width = 4)
+ggsave(plot = studyDurationPlot, filename = "Figures/Duration.png",
+       height = 3, width = 12)
 ```
 
-Plot site quantity
+```{r combine sites and duration}
+#| fig-width: 12
+#| fig-height: 6
 
-```{r plot site quantity}
-siteQuantityPlot <- sheet_DataCollection |>
-  mutate(`Site quantity` = as.numeric(`Site quantity`)) |>
-  ggplot() +
-  geom_histogram(aes(y = `Site quantity`),
-                 binwidth = 0.25, fill = "aquamarine4", colour = "white") +
-  geom_hline(aes(yintercept = median(`Site quantity`, na.rm = TRUE)), 
-             colour = "goldenrod3", linetype = 2, linewidth = 2) +
-  scale_y_continuous(transform = "log10",
-                     limits = c(5, 10000),
-                     breaks = c(10, 25, 50, 100, 250, 500,
-                                1000, 2500, 5000, 10000)) +
-  scale_x_continuous(expand = c(0,0)) +
-  labs(x = "", y = "") +
-  theme(panel.grid = element_blank(),
-        panel.background = element_rect(fill='transparent', color=NA),
-        plot.background = element_rect(fill='transparent', color=NA),
-        axis.text.y = element_text(size = 23),
-        axis.text.x = element_blank(), axis.ticks.x = element_blank())
-siteQuantityPlot
-ggsave(plot = siteQuantityPlot, filename = "Manuscript/Figures/Sites.png",
-       height = 8, width = 4)
+combinedStudyScale <- siteQuantityPlot / studyDurationPlot
+combinedStudyScale <- combinedStudyScale +
+  theme(panel.background = element_rect(fill='transparent', color=NA),
+        plot.background = element_rect(fill='transparent', color=NA))
+combinedStudyScale
+ggsave(plot = combinedStudyScale, filename = "Figures/StudyScale.png",
+       height = 6, width = 12)
 ```
-
-Site quantity boxplots
-
-```{r site quantity boxplots}
-sheet_DataCollection |>
-  mutate(`Site quantity` = as.numeric(`Site quantity`)) |>
-  left_join(select(sheet_ArticleData, c(`Review ID`, `Year Strata`)), by = "Review ID") |>
-  mutate(outlier = case_when(`Site quantity` >= 500 ~ TRUE,
-                             .default = FALSE)) |>
-  mutate(`Site quantity` = case_when(`Site quantity` > 500 ~ 500,
-                                     .default = `Site quantity`)) |>
-
-  ggplot(aes(x = `Year Strata`, y = `Site quantity`)) +
-  geom_jitter(aes(size = outlier, shape = outlier),
-              colour = "cadetblue4", width = 0.1) +
-  geom_boxplot(alpha = 0.4, colour = "cadetblue4",
-               width = 0.4, outliers = FALSE) +
-  scale_shape_manual(limits = c(TRUE, FALSE),
-                     values = c(17,16)) +
-  scale_size_manual(limits = c(TRUE, FALSE),
-                    values = c(3,1)) +
-  labs(x = "", y = "Sites") +
-  theme(legend.position = "none",
-        panel.grid.major.x = element_blank(),
-        panel.grid.minor.x = element_blank())
-
-```
-
-Plot primary occasion quantity
-
-```{r plot primary occasions}
-sheet_DataCollection |>
-  summarise(Articles = n(), .by = `Primary occasions`) |>
-
-  ggplot() +
-  geom_histogram(aes(x = `Primary occasions`), fill = "cadetblue4") +
-  labs(x = "Number of primary occasions", y = "Frequency") +
-  theme(panel.grid.major.x = element_blank(),
-        panel.grid.minor.x = element_blank())
-```
-
-Plot secondary occasion quantity
-
-```{r plot secondary occasions}
-
-sheet_DataCollection |>
-  filter(!`Secondary occasions` %in% c('Unclear', 'Unspecified')) |>
-  separate_wider_delim(`Secondary occasions`, delim = "-",
-                       names = c("minSecondary", "maxSecondary"),
-                       too_few = 'align_end') |>
-  mutate(maxSecondary = as.numeric(maxSecondary)) |>
-  summarise(Articles = n(), .by = maxSecondary) |>
-
-  ggplot() +
-  geom_histogram(aes(x = maxSecondary), fill = "cadetblue4") +
-  labs(x = "Maximum secondary occasions", y = "Frequency") +
-  theme(panel.grid.major.x = element_blank(),
-        panel.grid.minor.x = element_blank())
-
-```
-
-Study scale over time
-
-```{r study scale over time}
-sheet_DataCollection |>
-  mutate(`Site quantity` = as.numeric(`Site quantity`)) |>
-  left_join(select(sheet_ArticleData, c(`Review ID`, `Year Strata`)), by = "Review ID") |>
-  filter(!`Secondary occasions` %in% c('Unclear', 'Unspecified')) |>
-  separate_wider_delim(`Secondary occasions`, delim = "-",
-                       names = c("minSecondary", "maxSecondary"),
-                       too_few = 'align_end') |>
-  mutate(maxSecondary = as.numeric(maxSecondary)) |>
-  mutate(studyScale = `Site quantity`*`Primary occasions`) |>
-  mutate(studyScale = case_when(studyScale > 10000 ~ 10000,
-                                .default = studyScale)) |>
-  mutate(outlier = case_when(studyScale >= 10000 ~ TRUE,
-                             .default = FALSE)) |>
-
-  ggplot(aes(x = `Year Strata`, y = studyScale)) +
-  geom_jitter(aes(shape = outlier, size = outlier),
-              colour = "cadetblue4", width = 0.1) +
-  geom_boxplot(alpha = 0.4, colour = "cadetblue4", width = 0.4) +
-  scale_shape_manual(limits = c(TRUE, FALSE),
-                     values = c(17,16)) +
-  scale_size_manual(limits = c(TRUE, FALSE),
-                    values = c(3,1)) +
-  labs(x = "", y = "Sites x Seasons") +
-  theme(legend.position = "none",
-        panel.grid.major.x = element_blank(),
-        panel.grid.minor.x = element_blank())
-```
-
-# Covariates
-
-Covariate counts
-
-```{r Covariate counts}
-covParamPlot <- sheet_Covariates |> 
-  separate_longer_delim(Covariates, delim = ", ") |>
-  separate_wider_delim(Covariates, delim = "_", 
-                       names = c("Covariate", "Variation", 
-                                 "Observation", "Relation"),
-                       too_few = "align_start") |>
-  separate_wider_delim(Covariate, delim = "-",
-                       names = c("Category", "Covariate"),
-                       too_few = "align_start") |>
-  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
-                                 "Colonisation", "Extinction_Persistence",
-                                 "Detection")) |>
-  summarise(Covariates = n(),
-            .by = c(`Review ID`, `Model ID`, `Parameter type`, Category)) |>
-  complete(nesting(`Review ID`, `Model ID`, `Parameter type`), 
-           Category, fill = list(Covariates = 0)) |>
-  filter(Category != "None") |>
-  mutate(`Parameter type` = case_when(`Parameter type` == "Extinction_Persistence" ~ "Extinction",
-                                      `Parameter type` == "Occupancy" ~ "Occupancy*",
-                                      .default = `Parameter type`)) |>
-  mutate(Parameter = fct(`Parameter type`, 
-                         levels = c("Initial occupancy", "Occupancy*",
-                                    "Colonisation", "Extinction",
-                                    "Detection"))) |>
-  mutate(Covariates = case_when(Covariates >= 20 ~ 20,
-                                .default = Covariates)) |>
-
-  ggplot(aes(x = Category, y = Covariates)) +
-  geom_jitter(aes(colour = Category), 
-              alpha = 0.4, width = 0.2, height = 0.2) +
-  geom_boxplot(aes(fill = Category), 
-               outliers = FALSE, alpha = 0.3, colour = "gray60") +
-  scale_fill_manual("", labels = c("Environmental covariates", "Structural covariates"),
-                    values = c("aquamarine4", "plum4")) +
-  scale_colour_manual("", labels = c("Environmental covariates", 
-                                     "Structural covariates"),
-                      values = c("aquamarine4", "plum4")) +
-  scale_y_continuous(breaks = c(0, 5, 10, 15, 20),
-                     labels = c("0", "5", "10", "15", "20+")) +
-  labs(x = "", y = "Number of covariates considered") +
-  facet_grid(cols = vars(Parameter), switch = "x") +
-  theme(legend.position = "bottom", 
-        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
-        panel.grid.major.x = element_blank(),
-        panel.grid.minor.x = element_blank(), 
-        strip.text = element_text(size = 12, family = "Helvetica"), 
-        axis.title.y = element_text(size = 13, family = "Helvetica"), 
-        axis.text = element_text(size = 12, family = "Helvetica"),
-        legend.text = element_text(size = 12, family = "Helvetica"))
-
-covParamPlot
-ggsave(plot = covParamPlot, filename = "Manuscript/Figures/CovParamPlot.jpeg",
-       width = 8, height = 4)
-```
-
-Covariate table
-
-```{r Covariate table}
-# Covariate level summaries ------------------------------------------------
-# Extract details for each covariate and get parameter counts
-covariate_details <- sheet_Covariates |>
-  separate_longer_delim(Covariates, delim = ", ") |>
-  separate_wider_delim(Covariates, delim = "_", 
-                       names = c("Covariate", "Variation", 
-                                 "Observation", "Relation"),
-                       too_few = "align_start") |>
-  separate_wider_delim(Covariate, delim = "-",
-                       names = c("Category", "Covariate"),
-                       too_few = "align_start") |>
-  mutate(n_overall = length(unique(`Review ID`))) |>
-  mutate(n_param = length(unique(`Review ID`)), 
-         .by = `Parameter type`) |>
-  mutate(Dynamic = case_when(Variation %in% c("SE", "SISE", "SU") ~ TRUE,
-                             .default = FALSE),
-         DirectObs = case_when(Observation == "D" ~ TRUE, 
-                               .default = FALSE),
-         Nonlinear = case_when(Relation %in% c("N", "B") ~ TRUE,
-                               .default = FALSE),
-         Interaction = case_when(Relation %in% c("I", "B") ~ TRUE,
-                                 .default = FALSE)) |>
-  filter(!is.na(Covariate))
-
-# Proportion of articles with covariate on any parameter 
-covariates_PropAny <- covariate_details |>
-  select(`Review ID`, Covariate, Category, n_overall) |>
-  distinct() |>
-  summarise(Prop_Any = n()/mean(n_overall), .by = c(Covariate, Category))
-
-# Proportion of articles with covariate on each parameter
-covariates_PropParameter <- covariate_details |>
-  select(`Review ID`, Covariate, Category, `Parameter type`, n_param) |>
-  distinct() |>
-  summarise(Prop_Param = n()/mean(n_param), 
-            .by = c(Covariate, Category, `Parameter type`)) |>
-  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
-                                 "Colonisation", "Extinction_Persistence",
-                                 "Detection")) |>
-  pivot_wider(names_from = `Parameter type`, values_from = Prop_Param,
-              values_fill = 0)
-
-# Proportion of covariates which are dynamic
-covariates_dynamic <- covariate_details |>
-  select(`Review ID`, Covariate, Category, Parameter, Dynamic) |>
-  filter(Parameter != "Initial occupancy") |>
-  summarise(Prop_Dynamic = sum(Dynamic)/n(),
-            .by = c(Category, Covariate))
-
-# Proportion of covariates which are directly observed
-covariates_direct <- covariate_details |>
-  select(`Review ID`, Covariate, Category, Parameter, DirectObs) |>
-  summarise(Prop_Direct = sum(DirectObs)/n(),
-            .by = c(Category, Covariate))
-
-# Proportion of articles which use each cov non-linearly
-covariates_nonlinear <- covariate_details |> 
-  select(`Review ID`, Covariate, Category, Nonlinear) |>
-  summarise(Nonlinear = max(Nonlinear), 
-            .by = c(Covariate, Category, `Review ID`)) |>
-  summarise(Prop_Nonlinear = sum(Nonlinear)/n(),
-            .by = c(Covariate, Category))
-
-# Proportion of articles which use each cov in an interaction  
-covariates_interact <- covariate_details |> 
-  select(`Review ID`, Covariate, Category, Interaction) |>
-  summarise(Interaction = max(Interaction), 
-            .by = c(Covariate, Category, `Review ID`)) |>
-  summarise(Prop_Interact = sum(Interaction)/n(),
-            .by = c(Covariate, Category))
-
-# Combine all for covariates
-covariates_table <- covariates_PropAny |>
-  full_join(covariates_PropParameter, by = join_by(Covariate, Category)) |>
-  full_join(covariates_dynamic, by = join_by(Covariate, Category)) |>
-  full_join(covariates_direct, by = join_by(Covariate, Category)) |>
-  full_join(covariates_nonlinear, by = join_by(Covariate, Category)) |>
-  full_join(covariates_interact, by = join_by(Covariate, Category)) |>
-  filter(!is.na(Covariate))
-
-# Category level summary ----------------------------------------------
-categories_PropAny <- covariate_details |>
-  select(`Review ID`, Category, n_overall) |>
-  distinct() |>
-  summarise(Prop_Any = n()/mean(n_overall), .by = c(Category))
-
-categories_PropParameter <- covariate_details |>
-  select(`Review ID`, Category, `Parameter type`, n_param) |>
-  distinct() |>
-  summarise(Prop_Param = n()/mean(n_param), 
-            .by = c(Category, `Parameter type`)) |>
-  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
-                                 "Colonisation", "Extinction_Persistence",
-                                 "Detection")) |>
-  pivot_wider(names_from = `Parameter type`, values_from = Prop_Param,
-              values_fill = 0)
-
-# Proportion of covariates which are dynamic
-categories_dynamic <- covariate_details |>
-  select(`Review ID`, Category, Parameter, Dynamic) |>
-  filter(Parameter != "Initial occupancy") |>
-  summarise(Prop_Dynamic = sum(Dynamic)/n(),
-            .by = c(Category))
-
-# Proportion of covariates which are directly observed
-categories_direct <- covariate_details |>
-  select(`Review ID`, Category, Parameter, DirectObs) |>
-  summarise(Prop_Direct = sum(DirectObs)/n(),
-            .by = c(Category))
-
-# Proportion of articles which use each cov non-linearly
-categories_nonlinear <- covariate_details |> 
-  select(`Review ID`, Category, Nonlinear) |>
-  summarise(Nonlinear = max(Nonlinear), 
-            .by = c(Category, `Review ID`)) |>
-  summarise(Prop_Nonlinear = sum(Nonlinear)/n(),
-            .by = c(Category))
-
-# Proportion of articles which use each cov in an interaction  
-categories_interact <- covariate_details |> 
-  select(`Review ID`, Category, Interaction) |>
-  summarise(Interaction = max(Interaction), 
-            .by = c(Category, `Review ID`)) |>
-  summarise(Prop_Interact = sum(Interaction)/n(),
-            .by = Category)
-
-categories_table <- categories_PropAny |>
-  full_join(categories_PropParameter, by = join_by(Category)) |>
-  full_join(categories_dynamic, by = join_by(Category)) |>
-  full_join(categories_direct, by = join_by(Category)) |>
-  full_join(categories_nonlinear, by = join_by(Category)) |>
-  full_join(categories_interact, by = join_by(Category)) |>
-  mutate(Covariate = case_when(Category == "E" ~ "Environmental",
-                               Category == "S" ~ "Structural"))
-
-# Overall summary row ----------------------------------------------
-overall_PropAny <- covariate_details |>
-  select(`Review ID`, n_overall) |>
-  distinct() |>
-  summarise(Prop_Any = n()/mean(n_overall)) |>
-  mutate(Category = "O", Covariate = "Overall")
-
-overall_PropParameter <- covariate_details |>
-  select(`Review ID`, `Parameter type`, n_param) |>
-  distinct() |>
-  summarise(Prop_Param = n()/mean(n_param), 
-            .by = c(`Parameter type`)) |>
-  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
-                                 "Colonisation", "Extinction_Persistence",
-                                 "Detection")) |>
-  pivot_wider(names_from = `Parameter type`, values_from = Prop_Param,
-              values_fill = 0) |>
-  mutate(Category = "O", Covariate = "Overall")
-
-# Proportion of covariates which are dynamic
-overall_dynamic <- covariate_details |>
-  select(`Review ID`, Parameter, Dynamic) |>
-  filter(Parameter != "Initial occupancy") |>
-  summarise(Prop_Dynamic = sum(Dynamic)/n()) |>
-  mutate(Category = "O", Covariate = "Overall")
-
-# Proportion of covariates which are directly observed
-overall_direct <- covariate_details |>
-  select(`Review ID`, Parameter, DirectObs) |>
-  summarise(Prop_Direct = sum(DirectObs)/n()) |>
-  mutate(Category = "O", Covariate = "Overall")
-
-# Proportion of articles which use each cov non-linearly
-overall_nonlinear <- covariate_details |> 
-  select(`Review ID`, Nonlinear) |>
-  summarise(Nonlinear = max(Nonlinear), 
-            .by = `Review ID`) |>
-  summarise(Prop_Nonlinear = sum(Nonlinear)/n()) |>
-  mutate(Category = "O", Covariate = "Overall")
-
-# Proportion of articles which use each cov in an interaction  
-overall_interact <- covariate_details |> 
-  select(`Review ID`, Interaction) |>
-  summarise(Interaction = max(Interaction), 
-            .by = `Review ID`) |>
-  summarise(Prop_Interact = sum(Interaction)/n()) |>
-  mutate(Category = "O", Covariate = "Overall")
-
-overall_row <- overall_PropAny |>
-  full_join(overall_PropParameter, by = join_by(Category, Covariate)) |>
-  full_join(overall_dynamic, by = join_by(Category, Covariate)) |>
-  full_join(overall_direct, by = join_by(Category, Covariate)) |>
-  full_join(overall_nonlinear, by = join_by(Category, Covariate)) |>
-  full_join(overall_interact, by = join_by(Category, Covariate))
-
-# Combine all summaries ----------------------------------------------------
-fullCovariateSummary <- covariates_table |>
-  rbind(categories_table) |>
-  rbind(overall_row) |>
-  mutate(across(where(is.numeric), ~ round(.x, 2)))
-
-write_csv(fullCovariateSummary, "Manuscript/Figures/CovariateTable.csv")
-```
-
-# Objectives
-
-Objective proportions
-
-```{r Objective proportions}
-sheet_Objectives |>
-  pivot_longer(c("Demonstrate methods?", "Observe trends?", 
-                 "Identify drivers?", "Test relationships?",
-                 "Predict spatially?", "Predict temporally?"),
-               names_to = "Objective", values_to = "ObjectiveIncluded") |>
-  filter(ObjectiveIncluded == "YES") |>
-  mutate(Articles = n()) |>
-  summarise(ObjArticles = n(), .by = c(Objective, Articles)) |>
-  mutate(PropArticles = ObjArticles/Articles) |>
-  mutate(Objective = fct_reorder(Objective, PropArticles)) |>
-
-  ggplot() + 
-  geom_col(aes(y = Objective, x = PropArticles),
-           fill = "cadetblue4") +
-  labs(x = "Proportion of articles", y = "") +
-  theme(panel.grid.major.y = element_blank(),
-        panel.grid.minor.y = element_blank()) 
-```
-
-Objective proportions through time
-
-```{r Objectives through time}
-objectiveTrendPlot <- sheet_Objectives |>
-  left_join(select(sheet_ArticleData, c(`Review ID`, `Year Strata`)), by = "Review ID") |>
-  pivot_longer(c("Demonstrate methods?", "Observe trends?", 
-                 "Identify drivers?", "Test relationships?",
-                 "Predict spatially?", "Predict temporally?"),
-               names_to = "Objective", values_to = "ObjectiveIncluded") |>
-  mutate(Objective = str_replace(Objective, "\\?", "")) |>
-  filter(ObjectiveIncluded == "YES") %>%
-  rbind(., mutate(., `Year Strata` = "All Years")) |>
-  mutate(Articles = length(unique(`Review ID`)), .by = `Year Strata`) |>
-  summarise(ObjArticles = n(), .by = c(Objective, Articles, `Year Strata`)) |>
-  mutate(PropArticles = (ObjArticles/Articles)*100) |>
-  mutate(Objective = fct_reorder(Objective, PropArticles)) |>
-  complete(Objective, `Year Strata`, fill = list(PropArticles = 0)) |>
-  mutate(Strata = case_when(`Year Strata` == "All Years" ~ "left",
-                            .default = "right")) |>
-  mutate(Strata = fct(Strata, levels = c("left", "right"))) |>
-  mutate(Objective = factor(Objective, levels = c("Demonstrate methods",
-                                                  "Observe trends", 
-                                                  "Identify drivers", 
-                                                  "Test relationships",
-                                                  "Predict spatially",
-                                                  "Predict temporally"))) |>
-  mutate(`Year Strata` = fct(`Year Strata`, levels = c("2004-2007",
-                                                       "2008-2011",
-                                                       "2012-2015", 
-                                                       "2016-2019",
-                                                       "2020-2023",
-                                                       "All Years"))) |>
-
-  ggplot() +
-  geom_col(aes(x = `Year Strata`, y = PropArticles, fill = Objective),
-           position = "dodge", colour = "white") + 
-  scale_fill_manual("",
-                    values = c("plum4", "cadetblue3", "aquamarine3",
-                               "aquamarine4", "goldenrod2", "goldenrod3")) +
-  scale_y_continuous(breaks = c(0, 20, 40, 60, 80),
-                     limits = c(0, 80),
-                     labels = c("0%", "20%", "40%", "60%", "80%")) +
-  labs(y = "Percentage of studies\nwith objective", x = "") +
-  theme(panel.grid.major.x = element_blank(),
-        panel.grid.minor.x = element_blank(),
-        legend.text = element_text(size = 15),
-        axis.text = element_text(size = 15), axis.title = element_text(size = 15),
-        legend.position = "bottom", strip.background = element_blank(),
-        strip.text = element_blank()) +
-  facet_grid(cols = vars(Strata),
-             scales = "free_x", space = "free_x", drop = TRUE)
-objectiveTrendPlot
-ggsave(plot = objectiveTrendPlot, "Manuscript/Figures/ObjectiveTrend.jpeg",
-       width = 8, height = 4)
-```
-
-Avg covariate count by objective
-
-```{r Objective covariate count}
-covsPerParam <- sheet_Covariates |>
-  separate_longer_delim(Covariates, delim = ", ") |>
-  mutate(n_models = length(unique(`Model ID`)), .by = `Review ID`) |>
-  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
-                                 "Colonisation", "Extinction_Persistence",
-                                 "Detection")) |>
-  mutate(n_params = length(unique(`Parameter type`)), .by = `Review ID`) |>
-  filter(Covariates != "None") |>
-  summarise(CovsPerParam = n()/(max(n_params)*max(n_models)), .by = `Review ID`)
-
-objectiveCovariatePlot <- sheet_Objectives |>
-  pivot_longer(c("Demonstrate methods?", "Observe trends?", 
-                 "Identify drivers?", "Test relationships?",
-                 "Predict spatially?", "Predict temporally?"),
-               names_to = "Objective", values_to = "ObjectiveIncluded") |>
-  mutate(Objective = str_replace(Objective, "\\?", "")) |>
-  filter(ObjectiveIncluded == "YES") %>%
-  rbind(., mutate(., Objective = "Any Objective")) |>
-  select(`Review ID`, Objective) |>
-  distinct() |>
-  left_join(covsPerParam, by = join_by(`Review ID`)) |>
-
-  mutate(Objective = factor(Objective, levels = c("Any Objective", "Demonstrate methods",
-                                                  "Observe trends", 
-                                                  "Identify drivers", 
-                                                  "Test relationships",
-                                                  "Predict spatially",
-                                                  "Predict temporally"))) |>
-  mutate(Objective = fct_relabel(Objective, 
-                                 ~ str_replace(.x, " ", "\n"))) |>
-  mutate(facet = case_when(Objective == "Any\nObjective" ~ "left",
-                           .default = "right")) |>
-  mutate(CovsPerParam = case_when(CovsPerParam > 10 ~ 10,
-                                  .default = CovsPerParam)) |>
-
-  ggplot(aes(x = Objective, y = CovsPerParam, 
-             fill = Objective, colour = Objective)) +
-  geom_boxplot(colour = "gray50", alpha = 0.6) +
-  geom_jitter(alpha = 0.6, width = 0.1) +
-  scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10),
-                     labels = c("0", "2", "4", "6", "8", "10+")) +
-  scale_fill_manual(limits = c("Any\nObjective", "Demonstrate\nmethods", 
-                               "Observe\ntrends", "Identify\ndrivers", 
-                               "Test\nrelationships", "Predict\nspatially",
-                               "Predict\ntemporally"),
-                    values = c("gray60", "plum4", "cadetblue3", "aquamarine3",
-                               "aquamarine4", "goldenrod2", "goldenrod3")) +
-  scale_colour_manual(limits = c("Any\nObjective", "Demonstrate\nmethods", 
-                                 "Observe\ntrends", "Identify\ndrivers", 
-                                 "Test\nrelationships", "Predict\nspatially",
-                                 "Predict\ntemporally"),
-                      values = c("gray60", "plum4", "cadetblue3", "aquamarine3",
-                                 "aquamarine4", "goldenrod2", "goldenrod3")) +
-  labs(x = "", y = "Number of covariates\nconsidered per parameter") +
-  facet_grid(cols = vars(facet), scales = "free_x", space = "free_x") +
-  theme(legend.position = "none",
-        panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(),
-        legend.text = element_text(size = 15), axis.text = element_text(size = 13),
-        axis.title = element_text(size = 15), strip.background = element_blank(),
-        strip.text = element_blank()) 
-objectiveCovariatePlot
-ggsave(plot = objectiveCovariatePlot, "Manuscript/Figures/ObjectiveCovariates.jpeg",
-       width = 8, height = 4)
-```
-
-Combined plot
-
-```{r combined objective plot}
-stackedObjectivePlot <- objectiveTrendPlot / objectiveCovariatePlot +
-   plot_annotation(tag_levels = 'A')
-stackedObjectivePlot
-
-ggsave(plot = stackedObjectivePlot, "Manuscript/Figures/ObjectiveFigure.jpeg",
-       width = 12, height = 8)
-```
-
-# Modelling
-
-Bayes/Freq/All table
-
-need:
-
-n articles in each category
-
-median covariates per core param
-
-n using each selection method
-
-n which look at fit
-
-n which evaluate
-
-```{r modelling table}
-# get number of covariates per parameter for each model
-modelling_Covariates <- sheet_Covariates |>
-  separate_longer_delim(Covariates, delim = ", ") |>
-  mutate(n_models = length(unique(`Model ID`)), .by = `Review ID`) |>
-  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
-                                 "Colonisation", "Extinction_Persistence",
-                                 "Detection")) |>
-  mutate(n_params = length(unique(`Parameter type`)), .by = `Review ID`) |>
-  filter(Covariates != "None") |>
-  summarise(CovsPerParam = n()/(max(n_params)*max(n_models)), .by = `Review ID`)
-
-# compose table
-modelling_details <- sheet_Modelling |>
-  mutate(n_Overall = length(unique(`Review ID`))) |>
-  mutate(n_Class = length(unique(`Review ID`)),
-         .by = `Model class`) |>
-  separate_longer_delim(`Performance evaluation`, ", ") |>
-  separate_longer_delim(`Selection method`, ", ") |>
-  mutate(GOF = case_when(`Performance evaluation` %in% c("GOF", "BayesianP") ~ TRUE,
-                         .default = FALSE),
-         Evaluate = case_when(`Performance evaluation` %in% c("Validation_In", "Validation_Out") ~ TRUE,
-                         .default = FALSE)) |>
-  mutate(Select = 1) |>
-  pivot_wider(names_from = `Selection method`, 
-              values_from = Select, values_fill = 0) |>
-  summarise(GOF = max(GOF), Evaluate = max(Evaluate),
-            `Model averaging` = max(`Model averaging`),
-            `A priori` = max(`A priori`), 
-            Sequential = max(Sequential), Simple = max(Simple),
-            Candidate = max(`Candidate set`),
-            .by = c(`Review ID`, `Model class`, n_Class, n_Overall)) |>
-  left_join(modelling_Covariates, by = join_by(`Review ID`)) |>
-  filter(`Model class` != "Other")
-
-modelling_byClass <- modelling_details |>
-  summarise(Number = mean(n_Class),
-            MedianCovs = median(CovsPerParam, na.rm = TRUE),
-            Prop_ModAvg = sum(`Model averaging`)/mean(n_Class),
-            Prop_AnySelect = 1-(sum(`A priori`)/mean(n_Class)),
-            Prop_Sequential = sum(Sequential)/mean(n_Class),
-            Prop_Simple = sum(Simple)/mean(n_Class),
-            Prop_Candidate = sum(Candidate)/mean(n_Class),
-            Prop_GOF = sum(GOF)/mean(n_Class),
-            Prop_Eval = sum(Evaluate)/mean(n_Class),
-            .by = `Model class`) |>
-  mutate(across(where(is.numeric), ~ round(.x, 2)))
-
-modelling_overall <- modelling_details |>
-  summarise(Number = mean(n_Overall),
-            MedianCovs = median(CovsPerParam, na.rm = TRUE),
-            Prop_ModAvg = sum(`Model averaging`)/mean(n_Overall),
-            Prop_AnySelect = 1-(sum(`A priori`)/mean(n_Overall)),
-            Prop_Sequential = sum(Sequential)/mean(n_Overall),
-            Prop_Simple = sum(Simple)/mean(n_Overall),
-            Prop_Candidate = sum(Candidate)/mean(n_Overall),
-            Prop_GOF = sum(GOF)/mean(n_Overall),
-            Prop_Eval = sum(Evaluate)/mean(n_Overall)) |>
-  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
-  mutate(`Model class` = "All Models")
-
-modelling_summary <- rbind(modelling_byClass, modelling_overall) |>
-  t() |> as.data.frame() %>%
-  mutate(Variable = row.names(.)) |>
-  set_names(c("Frequentist", "Bayesian", "All models", "Variable")) |>
-  filter(Variable != "Model class")
-
-write_csv(modelling_summary, "Manuscript/Figures/ModellingSummary.csv")
-```

### 2024-09-23 Manuscript.qmd
Converted most of analysis supplement

@@
 These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.
-
-bibliography: references.bib
-csl: apa.csl
 ---
@@
 \begin{align*}
-z_1 &\sim Bernoulli(\psi_1)\\\\
-z_t &\sim Bernoulli(\psi)\\
+z_1 \sim& Bernoulli(\psi_1)\\\\
+z_t \sim& Bernoulli(\psi)\\\\
 &\psi = z_{t-1}(1-\epsilon)+(1-z_{t-1})\gamma
@@
 Rather than using multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012].
-Working with large numbers of taxa does raise additional challenges, as the level of complexity which can realistically be considered for each individual taxon is likely to be reduced for the sake of practicality.
+Working with large numbers of taxa does raise additional challenges, as the degree to which models are tailored to the taxa is likely to be limited for the sake of practicality.
 
@@
 B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles.
-55 articles fit models to only a single taxa, 30 fit independent models to multiple taxa, and 11 fit multi-species models with explicit interactions between taxa.
-C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors.
+55 articles fit models to only a single taxon, 30 fit independent models to multiple taxa, and 11 fit multi-species models with explicit interactions between taxa.
+C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as emerging methods like camera traps and acoustic monitors.
 13 articles incorporated data from citizen science projects.
@@
 Where 94% of frequentist models perform some manner of model selection to determine covariate inclusions for their models, only 36% of Bayesian models do so, with the majority instead fitting a single model defined *a priori*.
-Methods used for model selection vary considerably both within and between modelling frameworks.
+Methods used for model selection vary considerably both within and between the two modelling frameworks (@tbl-modelling).
 For frequentist models, the most popular and conventional approach to model selection as used in 43% of articles involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score.
-Most frequentist studies which do not follow this protocol instead use procedural model selection methods (39% of articles), where the structure for each model parameter is fit one-by-one.
+The next most popular method in frequentist studies is to use procedural model selection methods (39% of articles), where the structure for each model parameter is fit one-by-one.
 For example, this protocol might first identify the best structure for detection probability before moving on to initial occupancy and so on.
-A smaller portion of frequentist studies (9%) use more unique approaches, fitting simpler models such as single season occupancy models for candidate covariates to identify the most informative terms to use in a DOM.
-Across all frequentist implementations, 71% of articles conduct multi-model inference and model-averaging using AIC weights [@burnham2004].
+The remainder of frequentist studies (9%) use a variety of approaches, such as fitting simpler models like single season occupancy models to identify the most informative terms to use in a DOM.
+Across all frequentist implementations, 71% of articles summarise a final subset of models using multi-model inference and model-averaging with AIC weights [@burnham2004].
 
-Those Bayesian models which *do* perform model selection tend to take distinctly different approaches, with largely idiosyncratic methods across these studies.
-While direct comparison of model fit is rare, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data.
+Those Bayesian models which *do* perform model selection tend to take various approaches, with largely idiosyncratic methods across these studies.
+While direct comparison of model fit is rare in Bayesian methods, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data.
 Another approach used by @cook2022 fit a global model including all covariates before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the model.
@@
 
-Limited research has been conducted on advantages of different methods for covariate selection in DOMs, and there is surely no one-size-fit all approach which will be appropriate for all possible use-cases.
+Limited research has been conducted on advantages of different methods for covariate selection in DOMs, and there is unlikely to be a one-size-fits-all approach which will be appropriate for all possible use-cases.
 However, it is important to consider the implications of the different model selection approaches in common usage, and research from the SDM literature highlights how covariate selection can influence our interpretation of model outputs [@brodie2020].
-Within the occupancy modelling literature, @stewart2023's article on covariate selection in single-season occupancy model discusses important attributes of the information-criteria approaches widely used in frequentist models in our review, noting that these approaches can lead to inaccurate estimates and influence model inference.
-@morin2020's work raises different concerns on procedural model selection methods, demonstrating how the fine details of modelling protocols can determine whether the best-fitting possible model is identified and which covariates appear in final models.
+Within the occupancy modelling literature, @stewart2023's article on covariate selection in single-season occupancy models discusses important attributes of the information-criteria approaches widely used in frequentist models in our review, noting that these approaches can lead to inaccurate coefficient estimates that may influence model inference.
+@morin2020's work raises other concerns on procedural model selection methods, demonstrating how the fine details of modelling protocols can determine whether the best-fitting possible model is identified and which covariates appear in final models.
 The literature on performance of Bayesian model selection methods is more sparse, although @hooten2015's guide to Bayesian model selection in ecology is a valuable resource for possible methods of fitting those models.
 
-With few exceptions, most of the model selection methods identified in our review are designed only to identify the best model of those considered, and do not necessarily guarantee that selected models are well-fit to the data.
-Assessment of the performance of DOMs generally requires additional steps, and the best ways to achieve this are not necessarily clear.
+The model selection methods identified in our review are designed only to identify the best model of those considered, and do not necessarily guarantee that selected models are well-fit to the data.
+Assessment of the performance of DOMs generally requires additional steps, and the best ways to achieve this are not yet clear.
 There is no broadly-accepted goodness-of-fit test for dynamic occupancy models, although @mackenzie2004 describes an approach for single-season occupancy models using a parametric bootstrap which can been extended to DOMs; this test is implemented in the *AICcModAvg* and *unmarked* R packages.
@@
 
-Studies in our sample were classified according to their objectives as stated by authors to explore the use-cases which DOMs have been applied for.
-These objective categories were non-exclusive, and many studies fell into multiple categories; the frequency at which studies pursued each of these objectives is shown in @fig-objectives A.
-Across our study period DOMs have been frequently employed to observe trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions.
+Studies in our sample were classified according to their objectives as stated by authors to explore the use-cases which the DOM has been applied to (@fig-objectives).
+These objective categories were non-exclusive, and many studies fell into multiple categories.
+Across the study period the DOM has been frequently employed to estimate trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions.
 Within each of these broad categories lies even more variation in objectives, emphasising the DOM's flexibility as a tool for making ecological inference in diverse contexts.
 
-Many articles in our sample used the DOM seeking to monitor trends in occupancy state through time.
+Many articles in our sample used the DOM seeking to monitor trends in occupancy state through time ("Estimate trends" in @fig-objectives).
 These studies include @ahumada2013's paper assessing populations trends of 13 forest mammals and @scott2015's assessment of the conservation status of the endangered flattened musk turtle.
-These studies offer valuable insights on the state of species across landscapes and through time, demonstrating DOMs' suitability for many monitoring oriented projects — including those where many species need to be assessed at once.
-Other articles were more focused on examining the factors which influence species occupancy, either testing a pre-specified hypothetical relationships or taking a broader tack to identify drivers of occupancy without preconception.
+These studies offer valuable insights on the state of species across landscapes and through time, demonstrating the DOM's suitability for monitoring oriented projects; including those where multiple species need to be assessed at once.
+Other articles were more focused on examining the factors which influence species occupancy, either testing pre-specified hypothetical relationships or taking a broader tack to identify drivers of occupancy without preconception.
 The distinction between the two is important, as it guides which covariates may be considered and how model selection may be used to fit models.

### 2024-09-23 Randomisation/Randomisation.qmd
Converted most of analysis supplement

--- 2024-07-28 Randomisation/Final_Randomisation.qmd
@@
 1.
-ALL=(dynamic occupancy model\*) OR ALL=(occupancy dynamic\$ model\*) OR ALL=(multi\$season occupancy model\*) OR (ALL=(occupancy) AND ALL=(Coloni\$ation) AND ((ALL = (Extinction)) OR (ALL = (Persistence))) AND All=(Detection)) and Ecology or Biodiversity Conservation or Environmental Sciences or Zoology or Evolutionary Biology or Biology or Ornithology or Forestry or Plant Sciences or Environmental Studies or Fisheries or Entomology (Web of Science Categories)
+ALL=("dynamic\* occupancy model\*") OR TS=(dynamic\* NEAR "occupancy model\*") OR TI=(dynamic\* NEAR "occupancy model\*") OR AB=(dynamic\* NEAR "occupancy model\*") OR ALL=("occupancy dynamic\*") OR ALL=("multi\*season occupancy model\*") OR ALL=("multi-season occupancy model\*") OR (ALL=(occupancy) AND ALL=(coloni?ation) AND (ALL=(extinction) OR ALL=(persistence)) AND ALL=(detect\*)) and Ecology or Biodiversity Conservation or Environmental Sciences or Zoology or Ornithology or Evolutionary Biology or Biology or Marine Freshwater Biology or Forestry or Fisheries or Entomology or Plant Sciences or Environmental Studies (Web of Science Categories)
 2.  All articles citing MacKenzie et al. 2003
@@
 ```{r load WOS queries}
-keyTerms_datasheet <- read_xlsx("Queries/Final Queries/KeyTerms_A.xlsx",
+keyTerms_datasheet <- read_xlsx("Queries/Final Queries/Keywords.xlsx",
                                 skip = 10) |>
-  rbind(read_xlsx("Queries/Final Queries/KeyTerms_B.xlsx",
-                  skip = 10)) |>
   select(Title, Year = "Publication Year", Journal = "Source Title", 

### 2024-09-23 _manuscript/Analysis/Analysis.qmd
Converted most of analysis supplement

--- 2024-09-23 Analysis/Analysis.qmd
@@
         plot.background = element_rect(fill='transparent', color=NA))
-combinedStudyScale
 ggsave(plot = combinedStudyScale, filename = "Figures/StudyScale.png",

### 2024-09-23 Analysis/Analysis.qmd
updated review box

@@
 #| fig-width: 12
-#| fig-height: 3
+#| fig-height: 3.5
 
@@
 ggsave(plot = studyDurationPlot, filename = "Figures/Duration.png",
-       height = 3, width = 12)
+       height = 3.5, width = 12)
 ```
@@
 combinedStudyScale <- combinedStudyScale +
-  theme(panel.background = element_rect(fill='transparent', color=NA),
-        plot.background = element_rect(fill='transparent', color=NA))
+  theme(panel.background = element_rect(fill='transparent', color = NA),
+        plot.background = element_rect(fill='transparent', color = NA))
 combinedStudyScale
 ggsave(plot = combinedStudyScale, filename = "Figures/StudyScale.png",
-       height = 6, width = 12)
+       height = 6, width = 12, bg='transparent')
 ```

### 2024-09-26 Analysis.qmd
Completed analysis rerun

---
title: "Analysis and figures"
author: "Saoirse Kelleher"
date: today
abstract: "Calculates key statistics and renders figures for \"*Twenty years of dynamic occupancy models: a review of applications and look to the future*\""
abstract-title: "Summary"
format:
  html:
    theme: minty
---

## Load packages

First, we load in a handful of packages required for analysis and visualisation.

```{r Load packages}
#| message: false

library(tidyverse)
library(patchwork)
library(sf)
library(readxl)
library(ggtext)
```

## Read in data

Two data sources are required to render the analyses and visualisation for this article:

i.  Randomised list of all queried articles *(Randomisation.xlsx)*

ii. Spreadsheet of articles included in the review *(Review_Spreadsheet.xlsx)*

#### Randomised article list

All queried articles were collated and assigned random values in *'Randomisation.qmd'* to determine the order in which they would be considered for inclusion. '*Randomisation.xlsx'* contains all of these articles, and also documents which articles were included in the review, were excluded based on our review criteria, or were not processed at all. This spreadsheet is split across sheets by year strata. All sheets are read in using `readxl` and combined.

```{r Load query sheets}
sheet_AllArticles <- list(read_xlsx("Randomisation/Randomisation.xlsx", 
                                    sheet = "2004-2007"),
                          read_xlsx("Randomisation/Randomisation.xlsx", 
                                    sheet = "2008-2011"),
                          read_xlsx("Randomisation/Randomisation.xlsx", 
                                    sheet = "2012-2015"),
                          read_xlsx("Randomisation/Randomisation.xlsx", 
                                    sheet = "2016-2019"),
                          read_xlsx("Randomisation/Randomisation.xlsx", 
                                    sheet = "2020-2023")) |> 
  list_rbind()
```

#### Reviewed data

*'Review_Spreadsheet.xlsx'* contains all recorded details from the 92 reviewed articles. This spreadsheet contains several sheets, each including one category of variable (e.g., details on 'Focal Taxa' or 'Covariates'). Each of these is loaded individually with `readxl`.

```{r Load review sheets}
sheet_ArticleData <- read_xlsx("Review_Spreadsheet.xlsx",
                               sheet = "Article Data")
sheet_Objectives <- read_xlsx("Review_Spreadsheet.xlsx",
                              sheet = "Objectives") 
sheet_FocalTaxa <- read_xlsx("Review_Spreadsheet.xlsx",
                             sheet = "Focal Taxa") 
sheet_StudyArea <- read_xlsx("Review_Spreadsheet.xlsx",
                             sheet = "Study Area")
sheet_DataCollection <- read_xlsx("Review_Spreadsheet.xlsx",
                                  sheet = "Data Collection")
sheet_Covariates <- read_xlsx("Review_Spreadsheet.xlsx",
                              sheet = "Covariates")
sheet_Modelling <- read_xlsx("Review_Spreadsheet.xlsx",
                             sheet = "Modelling")
```

## Article inclusion

The first set of numbers we calculate describe the articles included in the review, and the inclusion rates within each of our year strata. These values are used to render a figure showing the coverage of reviewed articles during our study period.

#### Articles by query

Recall that our review sample was drawn from two Web of Science queries: the first being all articles which have cited @mackenzie2003, and the second using a selection of keywords presumed to be used in articles fitting DOMs. Here, we calculate the number of articles which appeared in one or both of these queries.

```{r Calculate query statistics}
stat_queryCounts <- sheet_AllArticles |>
  mutate(Query = case_when(KeyTerms == TRUE & CitesMackenzie == TRUE ~ "Both",
                           KeyTerms == FALSE & CitesMackenzie == TRUE ~ "Mackenzie only",
                           KeyTerms == TRUE & CitesMackenzie == FALSE ~ "Key terms only")) |>
  summarise(Articles = n(), .by = Query) |>
  mutate(Proportion = round(Articles/sum(Articles), 2))
stat_queryCounts
```

#### Total articles

Next, we get the number of articles included in the review.

```{r Total article count}
stat_articleTotal <- sheet_AllArticles |>
  filter(Included == "YES") |>
  nrow()
stat_articleTotal
```

#### Article hit rates

We then calculate the hit rate for articles in each stratum, by dividing the number of articles included by the number of articles processed. These hit rates are then used to estimate how many articles from the remaining pool would have met inclusion criteria — first summed from each year, then summed across years.

```{r Calculate article hit rates}
# Hit rates within each stratum
stat_hitRates <- sheet_AllArticles |>
  summarise(Articles = n(), 
            .by = c(Year, Strata, Included)) |>
  pivot_wider(names_from = Included, 
              values_from = Articles, 
              values_fill = 0) |>
  summarise(HitRate = round(sum(YES)/(sum(NO)+sum(YES)), 2),
            .by = Strata)
stat_hitRates

# Number of remaining articles per year
stat_remainingArticles <- sheet_AllArticles |>
  summarise(Articles = n(), .by = c(Year, Strata, Included)) |>
  pivot_wider(names_from = Included, 
              values_from = Articles, values_fill = 0) |>
  mutate(HitRate = sum(YES)/(sum(NO)+sum(YES)),
         .by = Strata) |>
  mutate(EstRemaining = round(`NA`*HitRate, 2)) |>
  select(Year, EstRemaining)
stat_remainingArticles

# Estimating remaining qualified articles and total remaining articles
stat_remainingProportion <- data.frame(Group = c("EstRemaining", "TotalUnreviewed"),
                                  Articles = c(sum(stat_remainingArticles$EstRemaining),
                                               sheet_AllArticles |>
                                                 summarise(Articles = n(), 
                                                           .by = c(Year, 
                                                                   Strata, 
                                                                   Included)) |>
                                                 pivot_wider(names_from = Included, 
                                                             values_from = Articles,
                                                             values_fill = 0) |>
                                                 pull(`NA`) |> 
                                                 sum()))
stat_remainingProportion
```

#### Coverage figure

Using these values we render a figure presenting the estimated coverage of our review over time (Figure 3 in the article).

```{r Coverage figure}
#| fig-width: 8
#| fig-height: 4

plot_reviewCoverage <- 
  sheet_AllArticles |>
  summarise(N = n(), .by = c(Year, Strata, Included)) |>
  pivot_wider(names_from = Included, values_from = N, values_fill = 0) |>
  mutate(HitRate = sum(YES)/(sum(NO)+sum(YES)),
         .by = Strata) |>
  mutate(EstRemaining = `NA`*HitRate) |>
  pivot_longer(cols = c(YES, EstRemaining), 
               names_to = 'Group', values_to = 'Articles') |>

  ggplot() +
  geom_col(aes(x = Year, y = Articles, fill = Group), colour = "white") +
  scale_fill_manual("",
                    limits = c("YES", "EstRemaining"),
                    labels = c("Included articles", 
                               "Estimated qualifying articles"),
                    values = c("aquamarine4", "gray65")) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 65)) +
  labs(x = "", y = "Number of articles") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = "bottom", 
        axis.text = element_text(size = 12),
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.title = element_text(size = 12), 
        legend.text = element_text(size = 12),
        strip.text.x = element_text(size = 12)) +
  facet_grid(cols = vars(Strata), scales = "free_x", switch = "x")
plot_reviewCoverage
ggsave(plot = plot_reviewCoverage, "Figures/CoveragePlot.jpeg",
       width = 8, height = 4)
```

## Study areas

The next set of statistics and figures focus on the study areas included in reviewed articles. For each article the countries where data was collected, a rough centroid of the study area, and an estimate of the spatial extent of the study area was recorded.

#### Study countries table

First, a table is generated presenting the number of articles using data from each country with at least one article.

```{r Study location table}
stat_studyCountries <- sheet_StudyArea |>
  select(`Review ID`, Country) |>
  separate_longer_delim(Country, ", ") |>
  distinct() |>
  mutate(TotalArticles = length(unique(`Review ID`))) |>
  summarise(Articles = n(), .by = c(Country, TotalArticles)) |>
  mutate(Proportion = round(Articles/TotalArticles,2)) |>
  select(-TotalArticles) |>
  arrange(-Articles)
stat_studyCountries
```

#### Unique countries

The total number of unique countries is extracted.

```{r Unique countries}
stat_uniqueCountries <- stat_studyCountries$Country |>
  unique() |>
  length()
stat_uniqueCountries
```

#### Map of study locations

A figure is rendered plotting the rough centroids for each study area included in the review, corresponding to Figure 4A in the article. We also note the total number of study areas included in the reviewed articles.

::: callout-important
## Articles may include multiple study areas in cases where distinct dataset were modelled separately
:::

```{r Study location map}
# Load country shapefile
countries_shp <- st_read("Analysis/Countries_Spatial/World_Countries__Generalized_.shp", quiet = TRUE) |>
  st_transform(crs = "ESRI:54030") |>
  filter(COUNTRY != "Antarctica")

# Convert coordinates and plot map
plot_studyAreaMap <- sheet_StudyArea |>
  st_as_sf(coords = c("Centre Point (lon)", "Centre Point (lat)"),
           crs = "EPSG:4326") |>
  st_transform(crs = "ESRI:54030") |>

  ggplot() +
  geom_sf(data = countries_shp, colour = "gray90", fill = "gray80") +
  geom_sf(shape = 21, fill = "aquamarine4", colour = "white", alpha = 0.5, size = 3.5) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.background = element_rect(fill='transparent', color=NA),
        plot.background = element_rect(fill='transparent', color=NA),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 
plot_studyAreaMap
ggsave(plot = plot_studyAreaMap, "Figures/StudiesMap.png")

# Get number of unique study areas
stat_totalStudyAreas <- sheet_StudyArea |>
  nrow()
stat_totalStudyAreas
```

#### Study area size

We noted the rough extent of each study area in our review, recording it as an order of magnitude to account for some amount of measurement uncertainty. Here we render a histogram of study area size to be used as Figure 4B in the article.

```{r Study area size}
#| fig-width: 7
#| fig-height: 4

plot_studyAreaSize <- sheet_StudyArea |>
  summarise(Articles = n(), .by = Size) |>
  filter(Size != "Unspecified") |>

  ggplot() +
  geom_col(aes(x = Size, y = Articles),
           fill = "aquamarine4", colour = "white",
           width = 1, linewidth = 2) +
  labs(x = "", y = "") +
  scale_x_discrete(labels = c("< 1", "1 - 10", "10 - 10<sup>2</sup>",
                              "10<sup>2</sup> - 10<sup>3</sup>",
                              "10<sup>3</sup> - 10<sup>4</sup>",
                              "10<sup>4</sup> - 10<sup>5</sup>",
                              "10<sup>5</sup> - 10<sup>6</sup>",
                              "\\> 10<sup>6</sup>")) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_markdown(size = 20, angle = 90,
                                       vjust = 0.5, hjust = 1),
        axis.text.y = element_blank(), axis.ticks = element_blank(),
        panel.background = element_rect(fill='transparent', color=NA),
        plot.background = element_rect(fill='transparent', color=NA))
plot_studyAreaSize

ggsave("Figures/StudyAreaPlot.png",
       plot = plot_studyAreaSize, width = 7, height = 4, bg='transparent')
```

## Study taxa

Details were recorded on the focal taxa for each article. In this section we calculate some numbers on the approaches taken for modelling multiple species and render a figure showing the types of taxa modelled.

#### Multi-species statistics

First we calculate the proportion of articles which modelled i) only one taxa, ii) multiple species independently, iii) multiple species in a hierarchical multi-species implementation, and iv) multiple species in an interactive multi-species implementation.

```{r Multi-species stats}
stat_multiSpecies <- sheet_FocalTaxa |>
  separate_longer_delim(`Multispecies method`, ", ") |>
  select(`Review ID`, `Multispecies method`) |>
  distinct() |>
  mutate(TotalArticles = length(unique(`Review ID`))) |>
  summarise(Articles = n(), 
            .by = c(`Multispecies method`, TotalArticles)) |>
  mutate(Proportion = round(Articles/TotalArticles, 2))
stat_multiSpecies
```

#### Study species

We render a figure corresponding to Figure 4D in the review, showing the number of articles fitting DOMs for each category of taxa, with further detail on invasive and threatened species. Species were considered 'threatened' either if they are currently listed on the IUCN Red List, or if the article indicates the species is otherwise threatened (e.g., a state 'species of special concern.') Range-expanding species like the Barred Owl in California were included in the 'Invasive' tally.

```{r Focal species figure}
#| fig-width: 10
#| fig-height: 4

plot_focalTaxa <- 
  sheet_FocalTaxa |>
  separate_longer_delim(`Taxa keywords`, delim = ", ") |>
  separate_longer_delim(Status, delim = ", ") |>
  mutate(`Taxa keywords` = case_when(`Taxa keywords` == "Fish" ~ "Other", 
                                     .default = `Taxa keywords`)) |>
  mutate(Status = case_when(Status %in% c("CR", "EN", "NT", "VU", "Threatened_Other") ~ "Threatened",
                            Status %in% c("NA", "Unclear", "DD") ~ "NA",
                            Status == "Invasive" ~ "Invasive",
                            Status == "LC" ~ "Stable")) |>
  mutate(Threatened = case_when("Threatened" %in% Status ~ TRUE,
                                .default = FALSE),
         Invasive = case_when("Invasive" %in% Status ~ TRUE,
                              .default = FALSE),
         Overall = TRUE,
         .by = c(`Review ID`, `Taxa keywords`)) |>
  select(`Review ID`, `Taxa keywords`, Threatened, Invasive, Overall) |>
  distinct() |>
  pivot_longer(cols = c(Threatened, Invasive, Overall), 
               names_to = "Group", values_to = "Value") |>
  summarise(Articles = sum(Value),
            .by = c(`Taxa keywords`, Group)) |>
  mutate(Group = fct(Group, levels = c("Overall", "Threatened",
                                       "Invasive"))) |>
  mutate(`Taxa keywords` = fct(`Taxa keywords`, 
                               levels = rev(c("Bird", "Mammal", "Herptile",
                                              "Invertebrate", "Other")))) |>
  # Bump up 0's to 0.1 so they show on the figure
  mutate(Articles = case_when(Articles == 0 ~ 0.1, .default = Articles)) |>

  ggplot() +
  geom_col(aes(x = Articles, y = `Taxa keywords`, fill = Group),
           position = position_dodge2(reverse = TRUE), width = 0.8) +
  scale_fill_manual("", values = c("aquamarine4", "goldenrod2", "gray70")) +
  scale_x_continuous(expand = c(0,0)) +
  labs(x = "Articles", y = "") +
  theme(panel.grid = element_blank(),
        legend.text = element_text(size = 20), axis.text = element_text(size = 21),
        axis.title = element_text(size = 19), axis.ticks.y = element_blank(),
        panel.background = element_rect(fill='transparent', color = NA),
        plot.background = element_rect(fill='transparent', color = NA),
        legend.background = element_rect(fill = 'transparent', color = NA))
plot_focalTaxa

ggsave("Figures/TaxaPlot.png",
       plot_focalTaxa, width = 10, height = 4, bg='transparent')
```

## Data collection

In this section, we calculate key statistics on the survey protocols used to collect data for DOMs in our review. This includes numbers on the most commonly used survey methods, the duration of surveys, and the number of sites. In cases where multiple datasets were used and modelled independently in the same article, these datasets are considered separately.

#### Survey method statistics

We tally how many articles used data collected with each survey method, as well as how many articles use citizen science data irrespective of detection method.

```{r survey methods}
# Number of articles using each method
stat_surveyMethod <- sheet_DataCollection |>
  separate_longer_delim(`Capture method`, ", ") |>
  select(`Review ID`, `Capture method`) |>
  distinct() |>
  mutate(TotalArticles = length(unique(`Review ID`))) |>
  summarise(Articles = n(), .by = c(`Capture method`, TotalArticles)) |>
  mutate(Proportion = round(Articles/TotalArticles, 2))
stat_surveyMethod

# Number of articles with citizen science
stat_citizen <- sheet_DataCollection |>
  select(`Review ID`, Citizen) |>
  distinct() |>
  mutate(Citizen = case_when(Citizen == "YES" ~ TRUE,
                             Citizen == "NO" ~ FALSE)) |>
  pull(Citizen) |>
  sum()
stat_citizen
```

#### Site quantity

We calculate the median number of sites included in DOMs, and render a histogram of site quantity for use in Figure 4F. We also get the sample size, as the number of datasets is greater than the number of reviewed articles due to aforementioned cases where studies contain multiple datasets.

```{r site quantity plot}
#| fig-width: 12
#| fig-height: 3.5

# Get median site quantity
stat_medianSites <- sheet_DataCollection |>
  filter(`Site quantity` != "Unclear") |>
  mutate(`Site quantity` = as.numeric(`Site quantity`)) |>
  pull(`Site quantity`) |>
  median(na.rm = TRUE)
stat_medianSites

# Plot site quantity distribution
plot_siteQuantity <- sheet_DataCollection |>
  filter(`Site quantity` != "Unclear") |>
  mutate(`Site quantity` = as.numeric(`Site quantity`)) |>
  ggplot() +
  geom_histogram(aes(x = `Site quantity`),
                 binwidth = 0.25, fill = "aquamarine4", colour = "white",
                 linewidth = 2) +
  geom_vline(aes(xintercept = median(`Site quantity`, na.rm = TRUE)), 
             colour = "goldenrod2", linetype = 1, linewidth = 2) +
  scale_x_continuous(transform = "log10",
                     breaks = c(0, 10, 25, 50, 100, 250, 500,
                                1000, 2500, 5000, 7500),
                     expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0), position = "right") +
  labs(y = "Site\nquantity", x = "") +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill='transparent', color=NA),
        plot.background = element_rect(fill='transparent', color=NA),
        axis.text.x = element_text(size = 23, angle = 90,
                                   vjust = 0.5, hjust = 1),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.line.x = element_line(), 
        axis.title.y = element_text(size = 35, colour = "aquamarine4"))
plot_siteQuantity
ggsave(plot = plot_siteQuantity, filename = "Figures/Sites.png",
       height = 3, width = 12)

# Site sample size
stat_siteQuantityN <- sheet_DataCollection |>
  filter(`Site quantity` != "Unclear") |>
  nrow()
stat_siteQuantityN
```

#### Study duration

We calculate similar statistics and render a matching plot for study duration for Figure 4F. Study duration is here defined as the time elapsed between the first and last surveys.

```{r site duration figure}
#| fig-width: 12
#| fig-height: 3.5

# Calculate median duration
stat_medianDuration <- sheet_DataCollection |>
  filter(`Start month` != "Unclear" & `End month` != "Unclear") |>
  mutate(StartDay = as.numeric(as.Date(as.numeric(`Start month`))),
         EndDay = as.numeric(as.Date(as.numeric(`End month`)))) |>
  mutate(StudyDuration = EndDay - StartDay) |>
  # Divide days by average length of month
  mutate(StudyMonths = (StudyDuration/30.44)) |>
  # For cases where start/end date are in same month, set to '1 month'
  mutate(StudyMonths = case_when(StudyMonths == 0 ~ 1,
                                 .default = StudyMonths)) |>
  pull(StudyMonths) |>
  median()
stat_medianDuration

# Render duration figure
plot_studyDuration <-
  sheet_DataCollection |>
  filter(`Start month` != "Unclear" & `End month` != "Unclear") |>
  mutate(StartDay = as.numeric(as.Date(as.numeric(`Start month`))),
         EndDay = as.numeric(as.Date(as.numeric(`End month`)))) |>
  mutate(StudyDuration = EndDay - StartDay) |>
  # Divide days by average length of month
  mutate(StudyMonths = (StudyDuration/30.44)) |>
  # For cases where start/end date are in same month, set to '1 month'
  mutate(StudyMonths = case_when(StudyMonths == 0 ~ 1,
                                 .default = StudyMonths)) |>

  ggplot() +
  geom_histogram(aes(x = StudyMonths), 
                 binwidth = 0.2, fill = "aquamarine4", colour = "white",
                 linewidth = 2) +
  geom_vline(aes(xintercept = median(StudyMonths, na.rm = TRUE)), 
             colour = "goldenrod2", linetype = 1, linewidth = 2) +
  labs(x = "", y = "Study\nduration") +
  scale_x_continuous(transform = "log10",
                     breaks = c(1, 6, 12, 60, 
                                120, 240, 480), 
                     labels = c("1 month", "6 months", "1 year", "5 years", 
                                "10 years", "20 years", "40 years"),
                     expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0), position = "right") +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill='transparent', color=NA),
        plot.background = element_rect(fill='transparent', color=NA),
        axis.text.x = element_text(size = 23, angle = 90,
                                   vjust = 0.5, hjust = 1),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.line.x = element_line(), 
        axis.title.y = element_text(size = 35, colour = "aquamarine4"))
plot_studyDuration
ggsave(plot = plot_studyDuration, filename = "Figures/Duration.png",
       height = 3.5, width = 12)

# Calculate sample size
stat_studyDurationN <- sheet_DataCollection |>
  filter(`Start month` != "Unclear" & `End month` != "Unclear") |>
  nrow()
stat_studyDurationN
```

We combine the study duration figure and the site quantity figure using `patchwork`.

```{r combine sites and duration}
#| fig-width: 12
#| fig-height: 7.5

plot_combinedStudyScale <- plot_siteQuantity / plot_spacer() / plot_studyDuration
plot_combinedStudyScale <- plot_combinedStudyScale +
  theme(panel.background = element_rect(fill='transparent', color = NA),
        plot.background = element_rect(fill='transparent', color = NA))
plot_combinedStudyScale
ggsave(plot = plot_combinedStudyScale, filename = "Figures/StudyScale.png",
       height = 7, width = 12, bg='transparent')
```

## Covariates

For each article we recorded *all* covariates considered across all modelled parameters.

#### Covariates figure

We render a plot to be used as Figure 5 in the article, presenting the number of covariates considered for each main parameter in each study.

```{r Covariate quantity figure}
plot_covariateQuantity <- sheet_Covariates |> 
  separate_longer_delim(Covariates, delim = ", ") |>
  separate_wider_delim(Covariates, delim = "_", 
                       names = c("Covariate", "Variation", 
                                 "Observation", "Relation"),
                       too_few = "align_start") |>
  separate_wider_delim(Covariate, delim = "-",
                       names = c("Category", "Covariate"),
                       too_few = "align_start") |>
  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
                                 "Colonisation", "Extinction_Persistence",
                                 "Detection")) |>
  summarise(Covariates = n(),
            .by = c(`Review ID`, `Model ID`, `Parameter type`, Category)) |>
  complete(nesting(`Review ID`, `Model ID`, `Parameter type`), 
           Category, fill = list(Covariates = 0)) |>
  filter(Category != "None") |>
  mutate(`Parameter type` = case_when(`Parameter type` == "Extinction_Persistence" ~ "Extinction",
                                      `Parameter type` == "Occupancy" ~ "Occupancy*",
                                      .default = `Parameter type`)) |>
  mutate(Parameter = fct(`Parameter type`, 
                         levels = c("Initial occupancy", "Occupancy*",
                                    "Colonisation", "Extinction",
                                    "Detection"))) |>
  mutate(Covariates = case_when(Covariates >= 15 ~ 15,
                                .default = Covariates)) |>
  mutate(Outlier = case_when(Covariates == 15 ~ TRUE,
                             .default = FALSE)) |>

  ggplot(aes(x = Category, y = Covariates)) +
  geom_jitter(aes(colour = Category, shape = Outlier), 
              alpha = 0.3, size = 2.5,
              width = 0.25, height = 0) +
  geom_boxplot(aes(fill = Category), 
               outliers = FALSE, alpha = 0.3, colour = "gray60") +
  scale_fill_manual("", labels = c("Environmental covariates", "Structural covariates"),
                    values = c("aquamarine4", "plum4")) +
  scale_colour_manual("", labels = c("Environmental covariates", 
                                     "Structural covariates"),
                      values = c("aquamarine4", "plum4")) +
  scale_shape_manual("", labels = c("",""), 
                     limits = c(TRUE, FALSE), values = c(17, 16)) +
  scale_y_continuous(breaks = c(0, 5, 10, 15),
                     labels = c("0", "5", "10", "15+")) +
  labs(x = "", y = "Number of covariates considered") +
  facet_grid(cols = vars(Parameter), switch = "x") +
  guides(shape = "none") +
  theme(legend.position = "bottom", 
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(), 
        strip.text = element_text(size = 12, family = "Helvetica"), 
        axis.title.y = element_text(size = 13, family = "Helvetica"), 
        axis.text = element_text(size = 12, family = "Helvetica"),
        legend.text = element_text(size = 12, family = "Helvetica"))

plot_covariateQuantity
ggsave(plot = plot_covariateQuantity, filename = "Figures/CovParamPlot.jpeg",
       width = 8, height = 4)
```

#### Covariates table

Here, we compile information on covariates required for Table 1. This is achieved in several steps: first, details on each narrow category of covariate are calculated; second, the same details are calculated for broader categories (environmental and structural); third, they are calculated once more for an overall summary.

```{r Compile covariate table}
# Covariate level summaries ------------------------------------------------
# Extract details for each covariate and get parameter counts
covariate_details <- sheet_Covariates |>
  separate_longer_delim(Covariates, delim = ", ") |>
  separate_wider_delim(Covariates, delim = "_", 
                       names = c("Covariate", "Variation", 
                                 "Observation", "Relation"),
                       too_few = "align_start") |>
  separate_wider_delim(Covariate, delim = "-",
                       names = c("Category", "Covariate"),
                       too_few = "align_start") |>
  # Merge some covariate categories
  mutate(Covariate = case_when(Covariate == "OTHER" & Category == "E" ~ "OTHER-E",
                               Covariate == "OTHER" & Category == "S" ~ "OTHER-S",
                               Covariate %in% c("HABT", "HABA", "LACO") ~ "HABT",
                               Covariate %in% c("CLWE", "DISA", "DIST") ~ "CLWE",
                               Covariate %in% c("SPAT", "GEOM", "CONN") ~ "SPAT",
                               Covariate %in% c("ANTH", "AGRI", "PROT") ~ "ANTH",
                               .default = Covariate)) |>
  # Combine Review ID & Model ID, and calculate the number of studies and number of studies with each parameter
  mutate(studyID = paste(`Review ID`, `Model ID`), sep = "_") |>
  mutate(n_overall = length(unique(studyID))) |>
  mutate(n_param = length(unique(studyID)), 
         .by = `Parameter type`) |>
  mutate(Dynamic = case_when(Variation %in% c("SE", "SISE", "SU") ~ TRUE,
                             .default = FALSE),
         DirectObs = case_when(Observation == "D" ~ TRUE, 
                               .default = FALSE),
         Nonlinear = case_when(Relation %in% c("N", "B") ~ TRUE,
                               .default = FALSE),
         Interaction = case_when(Relation %in% c("I", "B") ~ TRUE,
                                 .default = FALSE)) |>
  filter(!is.na(Covariate))

# Proportion of studies with covariate on any parameter 
covariates_PropAny <- covariate_details |>
  select(studyID, Covariate, Category, n_overall) |>
  distinct() |>
  summarise(Prop_Any = n()/mean(n_overall), .by = c(Covariate, Category))

# Proportion of articles with covariate on each parameter
covariates_PropParameter <- covariate_details |>
  select(studyID, Covariate, Category, `Parameter type`, n_param) |>
  distinct() |>
  summarise(Prop_Param = n()/mean(n_param), 
            .by = c(Covariate, Category, `Parameter type`)) |>
  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
                                 "Colonisation", "Extinction_Persistence",
                                 "Detection")) |>
  pivot_wider(names_from = `Parameter type`, values_from = Prop_Param,
              values_fill = 0)

# Average proportion of covariates which are dynamic
covariates_dynamic <- covariate_details |>
  select(studyID, Covariate, Category, Parameter, Dynamic) |>
  filter(Parameter != "Initial occupancy") |>
  summarise(Prop_Dynamic = sum(Dynamic)/n(),
            .by = c(Category, Covariate, studyID)) |>
  summarise(Prop_Dynamic = mean(Prop_Dynamic),
            .by = c(Category, Covariate))

# Proportion of covariates which are directly observed
covariates_direct <- covariate_details |>
  select(studyID, Covariate, Category, Parameter, DirectObs) |>
  summarise(Prop_Direct = sum(DirectObs)/n(),
            .by = c(Category, Covariate, studyID)) |>
  summarise(Prop_Direct = mean(Prop_Direct),
            .by = c(Category, Covariate))

# Proportion of articles which use each cov non-linearly
covariates_nonlinear <- covariate_details |> 
  select(studyID, Covariate, Category, Nonlinear) |>
  summarise(Nonlinear = max(Nonlinear), 
            .by = c(Covariate, Category, studyID)) |>
  summarise(Prop_Nonlinear = sum(Nonlinear)/n(),
            .by = c(Covariate, Category))

# Proportion of articles which use each cov in an interaction  
covariates_interact <- covariate_details |> 
  select(studyID, Covariate, Category, Interaction) |>
  summarise(Interaction = max(Interaction), 
            .by = c(Covariate, Category, studyID)) |>
  summarise(Prop_Interact = sum(Interaction)/n(),
            .by = c(Covariate, Category))

# Combine all for covariates
covariates_table <- covariates_PropAny |>
  full_join(covariates_PropParameter, by = join_by(Covariate, Category)) |>
  full_join(covariates_dynamic, by = join_by(Covariate, Category)) |>
  full_join(covariates_direct, by = join_by(Covariate, Category)) |>
  full_join(covariates_nonlinear, by = join_by(Covariate, Category)) |>
  full_join(covariates_interact, by = join_by(Covariate, Category)) |>
  filter(!is.na(Covariate))

# Category level summary ----------------------------------------------
categories_PropAny <- covariate_details |>
  select(studyID, Category, n_overall) |>
  distinct() |>
  summarise(Prop_Any = n()/mean(n_overall), .by = c(Category))

categories_PropParameter <- covariate_details |>
  select(studyID, Category, `Parameter type`, n_param) |>
  distinct() |>
  summarise(Prop_Param = n()/mean(n_param), 
            .by = c(Category, `Parameter type`)) |>
  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
                                 "Colonisation", "Extinction_Persistence",
                                 "Detection")) |>
  pivot_wider(names_from = `Parameter type`, values_from = Prop_Param,
              values_fill = 0)

# Proportion of covariates which are dynamic
categories_dynamic <- covariate_details |>
  select(studyID, Category, Parameter, Dynamic) |>
  filter(Parameter != "Initial occupancy") |>
  summarise(Prop_Dynamic = sum(Dynamic)/n(),
            .by = c(Category, studyID)) |>
  summarise(Prop_Dynamic = mean(Prop_Dynamic),
            .by = Category)

# Proportion of covariates which are directly observed
categories_direct <- covariate_details |>
  select(studyID, Category, Parameter, DirectObs) |>
  summarise(Prop_Direct = sum(DirectObs)/n(),
            .by = c(Category, studyID)) |>
  summarise(Prop_Direct = mean(Prop_Direct),
            .by = Category)

# Proportion of articles which use each cov non-linearly
categories_nonlinear <- covariate_details |> 
  select(studyID, Category, Nonlinear) |>
  summarise(Nonlinear = max(Nonlinear), 
            .by = c(Category, studyID)) |>
  summarise(Prop_Nonlinear = sum(Nonlinear)/n(),
            .by = Category)

# Proportion of articles which use each cov in an interaction  
categories_interact <- covariate_details |> 
  select(studyID, Category, Interaction) |>
  summarise(Interaction = max(Interaction), 
            .by = c(Category, studyID)) |>
  summarise(Prop_Interact = sum(Interaction)/n(),
            .by = Category)

categories_table <- categories_PropAny |>
  full_join(categories_PropParameter, by = join_by(Category)) |>
  full_join(categories_dynamic, by = join_by(Category)) |>
  full_join(categories_direct, by = join_by(Category)) |>
  full_join(categories_nonlinear, by = join_by(Category)) |>
  full_join(categories_interact, by = join_by(Category)) |>
  mutate(Covariate = case_when(Category == "E" ~ "Environmental",
                               Category == "S" ~ "Structural"))

# Overall summary row ----------------------------------------------
overall_PropAny <- covariate_details |>
  select(studyID, n_overall) |>
  distinct() |>
  summarise(Prop_Any = n()/mean(n_overall)) |>
  mutate(Category = "O", Covariate = "Overall")

overall_PropParameter <- covariate_details |>
  select(studyID, `Parameter type`, n_param) |>
  distinct() |>
  summarise(Prop_Param = n()/mean(n_param), 
            .by = c(`Parameter type`)) |>
  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
                                 "Colonisation", "Extinction_Persistence",
                                 "Detection")) |>
  pivot_wider(names_from = `Parameter type`, values_from = Prop_Param,
              values_fill = 0) |>
  mutate(Category = "O", Covariate = "Overall")

# Proportion of covariates which are dynamic
overall_dynamic <- covariate_details |>
  select(studyID, Parameter, Dynamic) |>
  filter(Parameter != "Initial occupancy") |>
  summarise(Prop_Dynamic = sum(Dynamic)/n(), 
            .by = studyID) |>
  summarise(Prop_Dynamic = mean(Prop_Dynamic)) |>
  mutate(Category = "O", Covariate = "Overall")

# Proportion of covariates which are directly observed
overall_direct <- covariate_details |>
  select(studyID, Parameter, DirectObs) |>
  summarise(Prop_Direct = sum(DirectObs)/n(),
            .by = studyID) |>
  summarise(Prop_Direct = mean(Prop_Direct)) |>
  mutate(Category = "O", Covariate = "Overall")

# Proportion of articles which use each cov non-linearly
overall_nonlinear <- covariate_details |> 
  select(studyID, Nonlinear) |>
  summarise(Nonlinear = max(Nonlinear), 
            .by = studyID) |>
  summarise(Prop_Nonlinear = sum(Nonlinear)/n()) |>
  mutate(Category = "O", Covariate = "Overall")

# Proportion of articles which use each cov in an interaction  
overall_interact <- covariate_details |> 
  select(studyID, Interaction) |>
  summarise(Interaction = max(Interaction), 
            .by = studyID) |>
  summarise(Prop_Interact = sum(Interaction)/n()) |>
  mutate(Category = "O", Covariate = "Overall")

overall_row <- overall_PropAny |>
  full_join(overall_PropParameter, by = join_by(Category, Covariate)) |>
  full_join(overall_dynamic, by = join_by(Category, Covariate)) |>
  full_join(overall_direct, by = join_by(Category, Covariate)) |>
  full_join(overall_nonlinear, by = join_by(Category, Covariate)) |>
  full_join(overall_interact, by = join_by(Category, Covariate))

# Combine all summaries ----------------------------------------------------
table_covariateSummary <- covariates_table |>
  rbind(categories_table) |>
  rbind(overall_row) |>
  mutate(across(where(is.numeric), ~ round(.x, 2)))

write_csv(table_covariateSummary, "Figures/CovariateTable.csv")
```

## Modelling

#### Modelling table

```{r Compile modelling table}
# get number of covariates per parameter for each model
modelling_Covariates <- sheet_Covariates |>
  separate_longer_delim(`Model ID`, delim = ",") |>
  separate_longer_delim(Covariates, delim = ", ") |>
  mutate(studyID = paste(`Review ID`, `Model ID`, sep = "_")) |>
  # Get number of core parameters
  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
                                 "Colonisation", "Extinction_Persistence",
                                 "Detection")) |>
  mutate(n_params = length(unique(`Parameter type`)), .by = studyID) |>
  # Remove 'none' covariates
  filter(Covariates != "None") |>
  # Calculate number of covariates per parameter
  summarise(CovsPerParam = n()/(max(n_params)), .by = studyID) |>
  # Models with 0 covariates equate to NA, convert these to 0s
  mutate(CovsPerParam = case_when(is.na(CovsPerParam) ~ 0,
                                  .default = CovsPerParam))

# compose table
modelling_details <- sheet_Modelling |>
  separate_longer_delim(`Model ID`, delim = ",") |>
  mutate(studyID = paste(`Review ID`, `Model ID`, sep = "_")) |>
  mutate(n_Overall = length(unique(studyID))) |>
  mutate(n_Class = length(unique(studyID)),
         .by = `Model class`) |>
  separate_longer_delim(`Performance evaluation`, ", ") |>
  separate_longer_delim(`Selection method`, ", ") |>
  mutate(GOF = case_when(`Performance evaluation` %in% c("GOF", "BayesianP") ~ TRUE,
                         .default = FALSE),
         Evaluate = case_when(`Performance evaluation` %in% c("Validation_In", "Validation_Out") ~ TRUE,
                         .default = FALSE)) |>
  mutate(Select = 1) |>
  pivot_wider(names_from = `Selection method`, 
              values_from = Select, values_fill = 0) |>
  summarise(GOF = max(GOF), Evaluate = max(Evaluate),
            `Model averaging` = max(`Model averaging`),
            `A priori` = max(`A priori`), 
            Sequential = max(Sequential), Simple = max(Simple),
            Candidate = max(`Candidate set`),
            .by = c(studyID, `Model class`, n_Class, n_Overall)) |>
  left_join(modelling_Covariates, by = join_by(studyID)) |>
  filter(`Model class` != "Other")

modelling_byClass <- modelling_details |>
  summarise(Number = mean(n_Class),
            MedianCovs = median(CovsPerParam, na.rm = TRUE),
            Prop_ModAvg = sum(`Model averaging`)/mean(n_Class),
            Prop_AnySelect = 1-(sum(`A priori`)/mean(n_Class)),
            Prop_Sequential = sum(Sequential)/mean(n_Class),
            Prop_Simple = sum(Simple)/mean(n_Class),
            Prop_Candidate = sum(Candidate)/mean(n_Class),
            Prop_GOF = sum(GOF)/mean(n_Class),
            Prop_Eval = sum(Evaluate)/mean(n_Class),
            .by = `Model class`) |>
  mutate(across(where(is.numeric), ~ round(.x, 2)))

modelling_overall <- modelling_details |>
  summarise(Number = mean(n_Overall),
            MedianCovs = median(CovsPerParam, na.rm = TRUE),
            Prop_ModAvg = sum(`Model averaging`)/mean(n_Overall),
            Prop_AnySelect = 1-(sum(`A priori`)/mean(n_Overall)),
            Prop_Sequential = sum(Sequential)/mean(n_Overall),
            Prop_Simple = sum(Simple)/mean(n_Overall),
            Prop_Candidate = sum(Candidate)/mean(n_Overall),
            Prop_GOF = sum(GOF)/mean(n_Overall),
            Prop_Eval = sum(Evaluate)/mean(n_Overall)) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  mutate(`Model class` = "All Models")

modelling_summary <- rbind(modelling_byClass, modelling_overall) |>
  t() |> as.data.frame() %>%
  mutate(Variable = row.names(.)) |>
  set_names(c("Frequentist", "Bayesian", "All models", "Variable")) |>
  filter(Variable != "Model class")

write_csv(modelling_summary, "Figures/ModellingSummary.csv")
```

## Objectives

The objectives of each article were documented, split into 6 non-exclusive categories.

#### Objectives statistics

We calculate the proportion of articles which used DOMs for each objective.

```{r Objective proportions}
stat_objectiveProportions <- sheet_Objectives |>
  mutate(TotalArticles = length(unique(`Review ID`))) |>
  pivot_longer(c("Demonstrate methods?", "Estimate trends?", 
                 "Identify drivers?", "Test relationships?",
                 "Predict spatially?", "Predict temporally?"),
               names_to = "Objective", values_to = "ObjectiveIncluded") |>
  filter(ObjectiveIncluded == "YES") |>
  summarise(ObjArticles = n(), 
            .by = c(Objective, TotalArticles)) |>
  mutate(PropArticles = ObjArticles/TotalArticles)
stat_objectiveProportions
```

#### Objectives figure

```{r Objective proportion figure}
#| fig-width: 8
#| fig-height: 4

plot_objectiveTrend <- sheet_Objectives |>
  left_join(select(sheet_ArticleData, c(`Review ID`, `Year Strata`)), by = "Review ID") |>
  pivot_longer(c("Demonstrate methods?", "Estimate trends?", 
                 "Identify drivers?", "Test relationships?",
                 "Predict spatially?", "Predict temporally?"),
               names_to = "Objective", values_to = "ObjectiveIncluded") |>
  mutate(Objective = str_replace(Objective, "\\?", "")) |>
  filter(ObjectiveIncluded == "YES") %>%
  rbind(., mutate(., `Year Strata` = "All Years")) |>
  mutate(StrataArticles = length(unique(`Review ID`)), .by = `Year Strata`) |>
  summarise(ObjArticles = n(), 
            .by = c(Objective, StrataArticles, `Year Strata`)) |>
  mutate(PctArticles = (ObjArticles/StrataArticles)*100) |>
  mutate(Objective = fct_reorder(Objective, PctArticles)) |>
  complete(Objective, `Year Strata`, fill = list(PctArticles = 0)) |>
  mutate(Strata = case_when(`Year Strata` == "All Years" ~ "left",
                            .default = "right")) |>
  mutate(Strata = fct(Strata, levels = c("left", "right"))) |>
  mutate(Objective = factor(Objective, levels = c("Demonstrate methods",
                                                  "Estimate trends", 
                                                  "Identify drivers", 
                                                  "Test relationships",
                                                  "Predict spatially",
                                                  "Predict temporally"))) |>
  mutate(`Year Strata` = fct(`Year Strata`, levels = c("2004-2007",
                                                       "2008-2011",
                                                       "2012-2015", 
                                                       "2016-2019",
                                                       "2020-2023",
                                                       "All Years"))) |>

  ggplot() +
  geom_col(aes(x = `Year Strata`, y = PctArticles, fill = Objective),
           position = "dodge", colour = "white") + 
  scale_fill_manual("",
                    values = c("plum4", "cadetblue3", "aquamarine3",
                               "aquamarine4", "goldenrod2", "goldenrod3")) +
  scale_y_continuous(breaks = c(0, 20, 40, 60, 80),
                     limits = c(0, 80),
                     labels = c("0%", "20%", "40%", "60%", "80%")) +
  labs(y = "Percentage of studies\nwith objective", x = "") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.text = element_text(size = 15),
        axis.text = element_text(size = 15), axis.title = element_text(size = 15),
        legend.position = "bottom", strip.background = element_blank(),
        strip.text = element_blank()) +
  facet_grid(cols = vars(Strata),
             scales = "free_x", space = "free_x", drop = TRUE)
plot_objectiveTrend
ggsave(plot = plot_objectiveTrend, "Figures/ObjectiveTrend.jpeg",
       width = 8, height = 4)
```

We render a second figure for the average number of covariates used in articles approaching each objective

```{r Covariates per objective}
#| fig-width: 8
#| fig-height: 4

covsPerParam <- sheet_Covariates |>
  mutate(studyID = paste(`Review ID`, `Model ID`, sep = "_")) |>
  separate_longer_delim(Covariates, delim = ", ") |>
  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
                                 "Colonisation", "Extinction_Persistence",
                                 "Detection")) |>
  mutate(n_params = length(unique(`Parameter type`)),
         .by = c(studyID, `Review ID`)) |>
  filter(Covariates != "None") |>
  summarise(CovsPerParam = n()/max(n_params), 
            .by = c(studyID, `Review ID`))

plot_objectiveCovariates <- sheet_Objectives |>
  pivot_longer(c("Demonstrate methods?", "Estimate trends?", 
                 "Identify drivers?", "Test relationships?",
                 "Predict spatially?", "Predict temporally?"),
               names_to = "Objective", values_to = "ObjectiveIncluded") |>
  mutate(Objective = str_replace(Objective, "\\?", "")) |>
  filter(ObjectiveIncluded == "YES") %>%
  rbind(., mutate(., Objective = "Any Objective")) |>
  select(`Review ID`, Objective) |>
  distinct() |>
  left_join(covsPerParam, by = join_by(`Review ID`), relationship = "many-to-many") |>
  mutate(CovsPerParam = case_when(is.na(CovsPerParam) ~ 0,
                                  .default = CovsPerParam)) |>
  mutate(Objective = factor(Objective, levels = c("Any Objective", 
                                                  "Demonstrate methods",
                                                  "Estimate trends", 
                                                  "Identify drivers", 
                                                  "Test relationships",
                                                  "Predict spatially",
                                                  "Predict temporally"))) |>
  mutate(Objective = fct_relabel(Objective, 
                                 ~ str_replace(.x, " ", "\n"))) |>
  mutate(facet = case_when(Objective == "Any\nObjective" ~ "left",
                           .default = "right")) |>
  mutate(CovsPerParam = case_when(CovsPerParam > 15 ~ 15,
                                  .default = CovsPerParam)) |>
  mutate(Outlier = case_when(CovsPerParam == 15 ~ TRUE,
                             .default = FALSE)) |>

  ggplot(aes(x = Objective, y = CovsPerParam, 
             fill = Objective, colour = Objective)) +
  geom_boxplot(colour = "gray50", alpha = 0.6, outliers = FALSE) +
  geom_jitter(aes(shape = Outlier), alpha = 0.4, width = 0.3, size = 2.5) +
  scale_y_continuous(breaks = c(0, 5, 10, 15),
                     labels = c("0", "5", "10", "15+")) +
  scale_fill_manual(limits = c("Any\nObjective", "Demonstrate\nmethods", 
                               "Estimate\ntrends", "Identify\ndrivers", 
                               "Test\nrelationships", "Predict\nspatially",
                               "Predict\ntemporally"),
                    values = c("gray60", "plum4", "cadetblue3", "aquamarine3",
                               "aquamarine4", "goldenrod2", "goldenrod3")) +
  scale_colour_manual(limits = c("Any\nObjective", "Demonstrate\nmethods", 
                                 "Estimate\ntrends", "Identify\ndrivers", 
                                 "Test\nrelationships", "Predict\nspatially",
                                 "Predict\ntemporally"),
                      values = c("gray60", "plum4", "cadetblue3", "aquamarine3",
                                 "aquamarine4", "goldenrod2", "goldenrod3")) +
  scale_shape_manual(limits = c(TRUE, FALSE),
                     values = c(17, 16)) +
  labs(x = "", y = "Number of covariates\nconsidered per parameter") +
  facet_grid(cols = vars(facet), scales = "free_x", space = "free_x") +
  guides(shape = "none") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(),
        legend.text = element_text(size = 15), axis.text = element_text(size = 13),
        axis.title = element_text(size = 15), strip.background = element_blank(),
        strip.text = element_blank()) 
plot_objectiveCovariates
ggsave(plot = plot_objectiveCovariates, "Figures/ObjectiveCovariates.jpeg",
       width = 8, height = 4)
```

Combine the objective plots

```{r combine objective plots}
#| fig-width: 12
#| fig-height: 8

plot_stackedObjectives <- plot_objectiveTrend / plot_objectiveCovariates +
   plot_annotation(tag_levels = 'A')
plot_stackedObjectives

ggsave(plot = plot_stackedObjectives, "Figures/ObjectiveFigure.jpeg",
       width = 12, height = 8)

```

### 2024-09-26 _manuscript/Analysis/Analysis.qmd
Completed analysis rerun

@@
 abstract-title: "Summary"
+format:
+  html:
+    theme: minty
 ---
@@
 
-Required packages for analysis and plotting figures.
+First, we load in a handful of packages required for analysis and visualisation.
 
@@
 
-## Load data
+## Read in data
 
-Two sources of data are loaded in for analysis:
+Two data sources are required to render the analyses and visualisation for this article:
 
--   Randomised list of all queried articles
+i.  Randomised list of all queried articles *(Randomisation.xlsx)*
 
--   Spreadsheet of included articles
+ii. Spreadsheet of articles included in the review *(Review_Spreadsheet.xlsx)*
 
-#### Randomised articles
+#### Randomised article list
 
-All queried articles were assigned random values in *'Randomisation.qmd'* to determine the order in which they were considered for inclusion in this review.
-These articles are read in from *'Randomisation.xlsx'*.
+All queried articles were collated and assigned random values in *'Randomisation.qmd'* to determine the order in which they would be considered for inclusion.
+'*Randomisation.xlsx'* contains all of these articles, and also documents which articles were included in the review, were excluded based on our review criteria, or were not processed at all.
+This spreadsheet is split across sheets by year strata.
+All sheets are read in using `readxl` and combined.
 
@@
 
-Each sheet from *'Review_Spreadsheet.xlsx'* is loaded individually.
+*'Review_Spreadsheet.xlsx'* contains all recorded details from the 92 reviewed articles.
+This spreadsheet contains several sheets, each including one category of variable (e.g., details on 'Focal Taxa' or 'Covariates').
+Each of these is loaded individually with `readxl`.
 
@@
 
+The first set of numbers we calculate describe the articles included in the review, and the inclusion rates within each of our year strata.
+These values are used to render a figure showing the coverage of reviewed articles during our study period.
+
 #### Articles by query
 
-Calculate the number of articles sourced from each query, and the proportion of articles which were included in both queries.
+Recall that our review sample was drawn from two Web of Science queries: the first being all articles which have cited @mackenzie2003, and the second using a selection of keywords presumed to be used in articles fitting DOMs.
+Here, we calculate the number of articles which appeared in one or both of these queries.
 
 ```{r Calculate query statistics}
-sheet_AllArticles |>
+stat_queryCounts <- sheet_AllArticles |>
   mutate(Query = case_when(KeyTerms == TRUE & CitesMackenzie == TRUE ~ "Both",
-                           KeyTerms == FALSE & CitesMackenzie == TRUE ~ "Mackenzie",
-                           KeyTerms == TRUE & CitesMackenzie == FALSE ~ "KeyTerms")) |>
+                           KeyTerms == FALSE & CitesMackenzie == TRUE ~ "Mackenzie only",
+                           KeyTerms == TRUE & CitesMackenzie == FALSE ~ "Key terms only")) |>
   summarise(Articles = n(), .by = Query) |>
   mutate(Proportion = round(Articles/sum(Articles), 2))
+stat_queryCounts
 ```
@@
 
-Get the number of included articles
+Next, we get the number of articles included in the review.
 
 ```{r Total article count}
-sheet_AllArticles |>
+stat_articleTotal <- sheet_AllArticles |>
   filter(Included == "YES") |>
   nrow()
+stat_articleTotal
 ```
@@
 
-Calculate the hit rate for articles in each stratum, and determine how many articles are estimated to remain in each stratum based off of those hit rates.
+We then calculate the hit rate for articles in each stratum, by dividing the number of articles included by the number of articles processed.
+These hit rates are then used to estimate how many articles from the remaining pool would have met inclusion criteria — first summed from each year, then summed across years.
 
@@
 # Hit rates within each stratum
-stratumHitRates <- sheet_AllArticles |>
-  summarise(Articles = n(), .by = c(Year, Strata, Included)) |>
+stat_hitRates <- sheet_AllArticles |>
+  summarise(Articles = n(), 
+            .by = c(Year, Strata, Included)) |>
   pivot_wider(names_from = Included, 
-              values_from = Articles, values_fill = 0) |>
+              values_from = Articles, 
+              values_fill = 0) |>
   summarise(HitRate = round(sum(YES)/(sum(NO)+sum(YES)), 2),
             .by = Strata)
-stratumHitRates
+stat_hitRates
 
 # Number of remaining articles per year
-remainingArticles <- sheet_AllArticles |>
+stat_remainingArticles <- sheet_AllArticles |>
   summarise(Articles = n(), .by = c(Year, Strata, Included)) |>
@@
   select(Year, EstRemaining)
-remainingArticles
+stat_remainingArticles
 
 # Estimating remaining qualified articles and total remaining articles
-remainingProportion <- data.frame(Group = c("EstRemaining", "TotalUnreviewed"),
-                                  Articles = c(sum(remainingArticles$EstRemaining),
+stat_remainingProportion <- data.frame(Group = c("EstRemaining", "TotalUnreviewed"),
+                                  Articles = c(sum(stat_remainingArticles$EstRemaining),
                                                sheet_AllArticles |>
-                                                 summarise(Articles = n(), .by = c(Year, Strata, Included)) |>
+                                                 summarise(Articles = n(), 
+                                                           .by = c(Year, 
+                                                                   Strata, 
+                                                                   Included)) |>
                                                  pivot_wider(names_from = Included, 
-                                                             values_from = Articles, values_fill = 0) |>
+                                                             values_from = Articles,
+                                                             values_fill = 0) |>
                                                  pull(`NA`) |> 
                                                  sum()))
-remainingProportion
+stat_remainingProportion
 ```
@@
 #### Coverage figure
+
+Using these values we render a figure presenting the estimated coverage of our review over time (Figure 3 in the article).
 
@@
 
-plot_coverage <- 
+plot_reviewCoverage <- 
   sheet_AllArticles |>
@@
                                "Estimated qualifying articles"),
-                    values = c("aquamarine4", "gray70")) +
+                    values = c("aquamarine4", "gray65")) +
   scale_y_continuous(expand = c(0,0), limits = c(0, 65)) +
@@
   facet_grid(cols = vars(Strata), scales = "free_x", switch = "x")
-plot_coverage
-ggsave(plot = plot_coverage, "Figures/CoveragePlot.jpeg",
+plot_reviewCoverage
+ggsave(plot = plot_reviewCoverage, "Figures/CoveragePlot.jpeg",
        width = 8, height = 4)
@@
 
+The next set of statistics and figures focus on the study areas included in reviewed articles.
+For each article the countries where data was collected, a rough centroid of the study area, and an estimate of the spatial extent of the study area was recorded.
+
 #### Study countries table
 
+First, a table is generated presenting the number of articles using data from each country with at least one article.
+
 ```{r Study location table}
-sheet_StudyArea |>
+stat_studyCountries <- sheet_StudyArea |>
   select(`Review ID`, Country) |>
@@
   distinct() |>
-  summarise(Articles = n(), .by = Country) |>
+  mutate(TotalArticles = length(unique(`Review ID`))) |>
+  summarise(Articles = n(), .by = c(Country, TotalArticles)) |>
+  mutate(Proportion = round(Articles/TotalArticles,2)) |>
+  select(-TotalArticles) |>
   arrange(-Articles)
+stat_studyCountries
 ```
@@
 
+The total number of unique countries is extracted.
+
 ```{r Unique countries}
-sheet_StudyArea |>
-  select(`Review ID`, Country) |>
-  separate_longer_delim(Country, ", ") |>
-  pull(Country) |>
+stat_uniqueCountries <- stat_studyCountries$Country |>
   unique() |>
   length()
+stat_uniqueCountries
 ```
@@
 #### Map of study locations
+
+A figure is rendered plotting the rough centroids for each study area included in the review.
+
+::: callout-important
+## Some articles include multiple study areas which were treated as distinct when they were modelled independently.
+:::
 
@@
 combinedStudyScale <- combinedStudyScale +
-  theme(panel.background = element_rect(fill='transparent', color=NA),
-        plot.background = element_rect(fill='transparent', color=NA))
+  theme(panel.background = element_rect(fill='transparent', color = NA),
+        plot.background = element_rect(fill='transparent', color = NA))
+combinedStudyScale
 ggsave(plot = combinedStudyScale, filename = "Figures/StudyScale.png",
-       height = 6, width = 12)
+       height = 6, width = 12, bg='transparent')
 ```
+
+## Covariates
+
+#### Covariates table
+
+#### Covariates figure
+
+## Modelling
+
+#### Modelling table
+
+## Objectives
+
+#### Objectives statistics
+
+#### Objectives figure

### 2024-09-27 Manuscript.qmd
Completed final version

@@
 ---
-title: "Twenty years of dynamic occupancy models: a review of applications and a look towards the future"
+title: "Twenty years of dynamic occupancy models: a review of applications and look towards the future"
 date: today
@@
 abstract: | 
-  Describing patterns of species occupancy across landscapes and throughout time is a key objective of much ecological research.
-Nonetheless, reliably estimating occupancy can be a challenge, particularly when common issues such as imperfect detection and shifting distributions are at play.
-The dynamic, multi-season occupancy model (DOM) offers a framework for occupancy estimation which explicitly accounts for observation error, while also incorporating additional mechanistic process by explicitly estimating colonisation and extinction rates.
-In the two decades since its introduction, the DOM has become a popular tool for describing patterns in occupancy, exploring the environmental drivers of species occurence, and, more recently, for predicting species occupancy to new locations and to future conditions.
-
-  In this review we examine how ecologists have applied the DOM in its first two decades of use.
-Based on a sample of articles fitting DOMs to field ecological data, we explore the systems in which these models have been used, how data were collected for them, and how authors have built their models to achieve their objectives.
-Our findings indicate that the DOM is an particularly flexible tool readily applied to a range of diverse taxa from around the globe, capable of being used to estimate occupancy at study sites ranging from local to continental scales over time periods from months to decades.
-The DOM framework is also amenable to extension, further broadening their utility to address a wide range of ecological questions.
-
-  A key aim of this review is to explore how users of DOMs have incorporated covariates into their models to describe variation in occupancy through space and time.
-Broadly speaking, model complexity in DOMs tends to be low, with relatively few covariates considered for inclusion and most relationships represented as simple linear terms.
-Considerable variation also exists in the procedures used for covariate selection, and limited research has been conducted on how these choices may influence model performance and inference.
-Furthermore, only a fraction of articles appear to evaluate models (either by goodness-of-fit testing or prediction validation).
-Likewise, little guidance exists on how to approach this task in DOMs.
+ Understanding the patterns by which species occur across landscapes and throughout time is a long-standing objective of ecological research which has inspired the development of numerous quantitative modelling approaches for the task.
+However, estimating occupancy can be a challenge, particularly where issues like imperfect detection and shifting distributions exist.
+Dynamic occupancy models (DOMs) offer a framework for occupancy estimation which explicitly accounts for observation error while incorporating additional mechanistic process by estimating colonisation and local extinction.
+In light of increasing interest in more process-explicit models for understanding species occurrence, here we examine how DOMs have been applied to field ecological data in the two decades since their introduction.
+Following a general introduction to the model, we present the results of a systematic review exploring the research contexts where DOMs have been applied and interrogating how authors have built their models, with particular emphasis on how covariates are incorporated to describe variation in occupancy dynamics.
+Our findings indicate that DOMs are a flexible tool readily applied to diverse study systems and data types, with their usage expanding in recent years as more studies apply them to make spatial and temporal prediction of species occupancy.
+DOMs are also amenable to extension, further broadening their utility.
+Despite their flexibility, model complexity in DOMs tends to be low; most studies consider relatively few covariates, most of which are represented as simple linear relationships.
+Approaches to covariate selection also vary considerably, and there remains little research on how these choices may influence model performance.
+Furthermore, only a fraction of articles report evaluating models and little guidance exists on how to approach this task in DOMs.
 These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.
@@
 The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849].
-An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing is applicable in many ways.
-Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, while identifying potential drivers of occurrence can help inform potential management actions.
-Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios.
+Estimates of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, and identifying potential drivers of occurrence can help inform potential management actions [@mackenzie2013].
+Robust knowledge of the occupancy patterns of a species can also help us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios [@kéry2013].
 
 While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate.
-The need to understand and describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly used to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales.
+The need to understand and quantitatively describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly applied to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales.
 However, several factors can make occupancy difficult to estimate.
@@
 In this review we examine how dynamic occupancy models have been used by ecologists in the two decades since their inception.
-Following an introduction to the model's form and assumptions, we present the results of a systematic review exploring how researchers have applied the DOM to ecological data, with emphasis on how they collected their data, selected covariates, and evaluated their models.
+Following an introduction to the model's form and assumptions, we present the results of a systematic review exploring how researchers have applied DOMs to ecological data, with emphasis on how they collected their data, selected covariates, and evaluated their models.
 Based on these results we highlight the DOM's flexibility as a tool for understanding species occurrence, examine approaches to the model building process, and outline key priorities for future research involving this model class.
@@
 The DOM structure encompasses two processes: the ecological process of site occupancy dynamics describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed at a site where it is present (@fig-modelform).
-Sites exist in either an occupied (z = 1) or unoccupied state (z = 0).
-In the first time step, occupancy state at each site is determined by the probability of initial occupancy ψ~1~.
-In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation γ and local extinction ε (@eq-occupancy).
-
-$$
-\begin{align*}
-z_1 \sim& Bernoulli(\psi_1)\\\\
-z_t \sim& Bernoulli(\psi)\\\\
-&\psi = z_{t-1}(1-\epsilon)+(1-z_{t-1})\gamma
-\end{align*}
-$$ {#eq-occupancy}
-
-In the observation component of the DOM, the model explicitly accounts for imperfect detection.
-At occupied sites, the detection probability ρ describes whether the species is observed during a survey (y = 1) or not (y = 0).
-Under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur (@eq-detection).
-
-$$
-y_{t,j} \sim Bernouilli(z_t\rho)
-$$ {#eq-detection}
+Sites can exist in either an occupied (z = 1) or unoccupied state (z = 0).
+In the first time step, occupancy state at each site is determined by the probability of initial occupancy ψ~1~, i.e., $z_1 \sim Bernouilli(\psi_1)$.
+In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation γ and local extinction ε, such that $z_t \sim Bernoulli(\psi)$ and $\psi = z_{t-1}(1-\epsilon)+(1-z_{t-1})\gamma$.
+The observation component of the DOM explicitly accounts for imperfect detection.
+At occupied sites, the detection probability ρ describes whether the species is observed during a survey (y = 1) or not (y = 0), according to the formula $y_{t,j} \sim Bernouilli(z_t\rho)$.
+Under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur.
+In this review we use the term "parameter" to refer to the quantities being estimate: in most cases, these are initial occupancy ψ~1~, colonisation γ, extinction ε, and detection ρ (see @fig-modelform).
+However, some alternative model forms may directly estimate occupancy ψ as a parameter.
 
 ![The form of the dynamic occupancy model as described by @mackenzie2003.
-The ecological process submodel describes changes in occupancy over time, where occupancy in the first season is given by the probability of initial occupancy and occupancy in ensuing seasons is governed by colonisation and extinction probabilities.
-The observational process submodel describes detectability, where the probability of detecting a species given that a site is occupied is given by the detection probability.
-At sites which are truly unoccupied, it is assumed that the species is never detected (i.e., no false positives).](Figures/ModelForm.png){#fig-modelform}
+The ecological process sub-model describes changes in occupancy over time, while the observational process sub-model describes detectability.](Figures/ModelForm.png){#fig-modelform}
 
@@
 It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this enables flexibility in data inputs.
-In this review we use the term "parameter" to refer to the quantities being estimate: in most cases, these are initial occupancy ψ~1~, colonisation γ, extinction ε, and detection ρ (see @fig-modelform, @eq-occupancy, and @eq-detection), though some alternative model forms may directly estimate occupancy ψ as a parameter.
 
 ![The sampling design of the standard dynamic occupancy model.
-During seasons, also called primary occasions, sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons.
-Within each season, multiple observations ('secondary occasions') are conducted to record the observed presence or absence of the species at each site.
+During seasons (also called primary occasions) sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons.
+Within each season, multiple observations (also called surveys or secondary occasions) are conducted to record the observed presence or absence of the species at each site.
 These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded.
@@
 II.
-**Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion in terms of the bias which is introduced when it is violated [@rota2009; @otto2013].
-Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discuss the difficultly in distinguishing between temporary emigration and local extinction.
-Model extensions to relax the closure assumption have been developed, including @kendall2013's approach using staggered arrival and departure periods between sites.
-A more pertinent approach for most studies, however, is careful consideration of the appropriate definition of occupancy under the survey design used [@kéry2021].
+**Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion in terms of the bias introduced when it is violated [@rota2009; @otto2013].
+Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and it can be difficult to distinguish local extinction from temporary emigration [@valente2017].
+Model extensions to relax the closure assumption have been developed, including approaches using staggered arrival and departure periods between sites [@kendall2013].
+A more pertinent approach for most studies, however, is careful consideration of the appropriate definition of occupancy under the survey design used [@steenweg2018].
 
@@
 **Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species occurrence, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system.
-Generally, this is achieved by allowing model parameters to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters.
-An important element of this assumption is that the likelihood of detection of a species depends not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017a].
-While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates.
+Generally, this is achieved by allowing model parameters (ψ~1~, γ, ε, and ρ) to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters.
+An important element of this assumption is that the likelihood of detection of a species can depend not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017a].
+While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates [@barry2006].
 Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known about how this latent heterogeneity can influence model performance.
@@
 Since its original description in @mackenzie2003, the DOM has been further developed with numerous model extensions and alternative formulations including aforementioned implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010].
-For a comprehensive discussion of the most common extensions and their applications see @bailey2014 and @guillera-arroita2017a, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.
+For a comprehensive discussion of common extensions and their applications see @bailey2014 and @guillera-arroita2017a, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.
 
@@
 To capture any additional relevant articles that did not directly cite @mackenzie2003, a second query was generated searching articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract.
+These queries resulted in 1469 articles: 897 were retrieved only from the MacKenzie citations, 274 only from the keywords search, and 298 which appeared in both queries.
 As we were interested in how DOMs use has changed through time, we divided all articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024.
@@
 
-As our review focuses only on applications of the dynamic multi-season occupancy model of @mackenzie2003 and its extensions, we included articles that fit a model meeting the following criteria: i) uses non-simulated, field-collected, presence-absence data; ii) relies on data from multiple sites which can exist in at least two states, including an occupied and unoccupied state; iii) has multiple seasons, between which sites may change states conditional on the prior season's occupancy state and transition probabilities such as colonisation and extinction; and iv) contains at least one parameter describing the detection process.
+As our review focuses only on applications of the dynamic multi-season occupancy model of @mackenzie2003 and its extensions, we included articles that fit a model meeting the following criteria: i) used non-simulated, field-collected, presence-absence data; ii) relied on data from multiple sites which could exist in at least two states, including an occupied and unoccupied state; iii) had multiple seasons, between which sites could change states conditional on the prior season's occupancy state and transition probabilities such as colonisation and extinction; and iv) contained at least one parameter describing the detection process.
 
@@
 A total of 92 articles were included for this review.
-Based on the acceptance rates within each stratum and quantity of unprocessed articles, an estimated 496 of the 1152 unreviewed articles in our sample would have met inclusion criteria (@fig-coverage).
+Based on the acceptance rates within each stratum and quantity of unprocessed articles, an estimated 496 of the 1152 unreviewed articles in our sample would have met inclusion criteria, with an apparent increase over time in the number of articles fitting DOMs (@fig-coverage).
 
 ![Bars indicate the 92 articles included in our review as a proportion of the estimated number of published articles fitting DOMs, based on the qualification rate for articles in each strata.
-The proportion of articles included from each strata were as follows: 12.5% from 2005-2008; 24.4% from 2008-2011; 42.6% from 2012-2015; 35.1% from 2016-2019; and 57.1% from 2020-2023.](Figures/CoveragePlot.jpeg){#fig-coverage}
+The proportion of articles included from each strata were as follows: 12% from 2005-2008; 24% from 2008-2011; 42% from 2012-2015; 35% from 2016-2019; and 57% from 2020-2023.](Figures/CoveragePlot.jpeg){#fig-coverage}
 
-Articles included in our review demonstrate considerable diversity in scope, scale, and objectives.
-A selection of key attributes of these studies is presented in @fig-StudyDetails, and we provide notable examples of DOMs throughout this section to highlight how the noted variation appears in practice.
-Study systems included in our sample are globally distributed; while a majority use data collected in the United States of America, representatives are included from all geographic realms and a broad diversity of ecosystems.
-A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale.
+The articles included in our review demonstrate considerable diversity in scope, scale, and objectives.
+A selection of attributes of these studies is presented in @fig-StudyDetails, and throughout this section we provide notable examples of DOMs to highlight how the noted variation appears in practice.
+Study systems included in our sample are globally distributed: while a majority of articles use data collected in the United States of America, representatives are included from all geographic realms and a broad diversity of ecosystems (@fig-StudyDetails A).
+A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale (@fig-StudyDetails B) .
 The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre, while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@basset2023; @clement2019].
@@
 The temporal scale of studies shows similar variability.
-The duration between first and last survey ranged from under one month to forty years, with a median of 6 years.
-More meaningful for model interpretation, however, is the duration of the primary occasion, as this is presumed to be the temporal scale at which changes in occupancy are observed.
-Most applications describe the primary occasion as a year, although some studies divide years into multiple primary occasions to describe seasonal variation in occupancy.
-In the most extreme cases the primary occasion may be as brief as a week, capturing much finer scale changes in occupancy.
+The time elapsed between the first and last survey ranged from under one month to over forty years, with a median duration of 8 years (@fig-StudyDetails F).
+More meaningful, however, is the period of the primary occasion, as this represents the temporal scale at which changes in occupancy are assumed to occur.
+Most applications describe primary occasions as occurring annually, although some studies divide years into multiple primary occasions to describe seasonal variation in occupancy.
+In the most extreme cases the primary occasion may be as brief as a week and capture much finer scale changes in occupancy.
 These shortened-season occupancy models are most common with camera trapping or acoustic monitoring data, which can be arbitrarily divided into primary and secondary sampling occasions.
-This is seen in @kleiven2020's study of rodent and mustelid interactions using camera trap data, where the primary occasions are less than one week in length.
+One example of this is @kleiven2020's study of rodent and mustelid interactions using camera trap data, which used primary occasions less than one week in length.
 On the other end of the spectrum, some studies modelled primary occasions which were decades apart and represented generational changes in occupancy.
-This is illustrated in @couturier2023's study on long-term otter recovery in France, which used data from two primary occasions: one in 2003 and one in 2012.
+This is illustrated in @couturier2023's study on long-term otter recovery in France which used data from two primary occasions: one in 2003 and one in 2012.
 
-The DOM has been used to study a variety of different species, though the majority of studies have focussed on vertebrate taxa (@fig-StudyDetails B).
+DOMs have been used to study a variety of species, though the majority of studies have focused on vertebrate taxa (@fig-StudyDetails C).
 The DOM has been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community.
-However, there are exceptions — @belinchón2017 fits a DOM to lichen data, and @cook2022 model the spread of chronic wasting disease.
-The latter's application to disease dynamics is not unique, and the DOM has been touted as a valuable tool for such applications.
-@mores2020 and @padilla-torres2013 have used the DOM to model mosquito dynamics as a human disease vector, and @bailey2014 discusses the concept of applying the DOM to the spread of chytrid fungus in amphibian habitats.
-The vast majority of studies model terrestrial species, though there have been a limited number of studies on aquatic systems, including invasive salmon [@fisher2014], Great Plains stream fishes [@falke2012], and whales [@pendleton2022].
+However, there are exceptions, including DOMs used to model decadel changes in lichen occupancy and the spread of chronic wasting disease [@belinchón2017; @cook2022].
+The latter's application to disease dynamics is not unique, and the DOM has been touted as a valuable tool for such applications [@bailey2014].
+Multiple authors have also used DOMs to model mosquito dynamics, a strongly applied use-case with important human health implications [@mores2020; @padilla-torres2013].
+The vast majority of studies model terrestrial species, though there have been a limited number of studies on aquatic systems, including invasive salmon, Great Plains stream fishes, and whales [@fisher2014; @falke2012; @pendleton2022].
 
-While most articles in our sample fit a model to a single species, 41 fit models to multiple species either as independent models or explicitly multi-species implementations.
-The multi-species models use various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects[@dorazio2010; @hendershot2020], as well as explicit interaction models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019].
+While most studies in our sample fit a model to a single species, 44% fit models to multiple species either as independent models (34%) or explicitly multi-species implementations (12% — some articles attempted multiple approaches; @fig-StudyDetails D).
+The multi-species models use various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects [@dorazio2010; @hendershot2020], as well as explicit interaction models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019].
 Rather than using multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012].
@@
 
-![A) Study areas: Data for a majority of studies were collected from study locations in the United States.
-The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres.
-B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles.
-55 articles fit models to only a single taxon, 30 fit independent models to multiple taxa, and 11 fit multi-species models with explicit interactions between taxa.
-C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as emerging methods like camera traps and acoustic monitors.
-13 articles incorporated data from citizen science projects.
-Project scale was also variable, with the median study running for 6 years and covering 101 sites.](Figures/ReviewBox.png){#fig-StudyDetails}
+![A) Locations of study areas where data was collected for reviewed DOMs.
+B) Spatial extent of study areas, defined as the area of inference, within which all surveyed points were contained.
+C) Number of articles which fit models to each category of taxa.
+Taxa were considered 'threatened' if they are listed on the IUCN Red List, or if authors indicate that they are otherwise threatened.
+D) Explicitly multi-species models include both hierarchical, jointly estimated models as well as more interactive models.
+Some studies fit both independent and multi-species models, such that these values do not sum to our sample size.
+E) Survey methods used to collected presence/absence data.
+Note that some articles employed multiple detection methods, and that some methods (e.g., citizen bird counts) may fall into multiple categories.
+F) Quantity of sites where surveys were conducted and duration of studies.
+Yellow bars indicate median values for site quantity (100) and study duration (8.2 years).
+Study duration is defined as the time elapsed between the first and last survey.](Figures/ReviewBox.png){#fig-StudyDetails}
 
-The presence/absence data required for DOMs has come from a breadth of different sources and detection methods (@fig-StudyDetails, panel C).
-In our sample 79% of articles used human observational data, 9% used live-trapping methods, and 14% used detections from camera traps.
-Several articles also use citizen-science data to build detection histories, including @zuckerberg2011's model using bird-feeder data and @warrier2020's using citizen reports of tiger signs.
+Studies collected the presence/absence data required for DOMs from a breadth of different sources and detection methods (@fig-StudyDetails E).
+In our sample, 79% of studies used human observational data, 11% used live-trapping methods, and 14% used detections from camera traps.
+10% of articles also use citizen-science data to build detection histories, including @zuckerberg2011's model using bird-feeder data and @warrier2020's using citizen reports of tiger signs.
 Using citizen-science data may require additional consideration of assumptions (particularly false-positives), as discussed in greater detail by @cruickshank2019.
-Within these broad categories of detection method there is additional diversity, with several studies using unique survey methods determining the presence of a species at a location.
-For example, in the only application of a DOM for a marine species in our sample, @pendleton2022 use aerial transects broken up into grid cells to observe whale occupancy; and @marescot2020 fit a unique 'multi-species' model treating poachers as a taxon and using ranger reports to create detection histories.
+Within these broad categories of detection method there is additional diversity, with several studies using unique survey methods to determine the presence of a species at a location.
+For example, in the only application of a DOM for a marine species in our sample, @pendleton2022 used aerial transects broken up into grid cells to observe whale occupancy; in another unique implementation, @marescot2020 fit a unique 'multi-species' model treating poachers as a taxon and using ranger reports to create detection histories.
 
@@
 Some defined primary seasons as a discrete time interval, treating all surveys occurring within the season as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate.
-These manipulations permit authors to use data which predate the DOM, with one study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].
+These manipulations permit authors to use data which predate the DOM, as in a study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].
 
-Additionally, not all articles rely on a single source of detection data - some integrate multiple sources of data to maximise sample size, combining data from camera traps, sign surveys, and citizen science reports.
-These integrated detection method models do require additional care and consideration; users must ensure that different detection methods represent comparable spatial and temporal scales, and that any variation in perceptibility between detection methods is accounted for [@warrier2020; @pitman2017].
+Additionally, not all articles rely on a single source of detection data; some integrate multiple sources of data to maximise sample size.
+In one example, @warrier2020 modelled tiger occupancy in India using three sources of detections, including camera traps, sign surveys, and citizen science reports.
+These integrated detection method models do require additional care and consideration; users must ensure that different detection methods represent comparable spatial and temporal scales, and that any variation in perceptibility between detection methods is accounted for [@pitman2017].
 A special case exists when different detection methods are used where one has the potential for false positive detections; e.g., where less-certain citizen science detections are combined with certain detections from field surveys.
@@
 This idea is particularly relevant in cases where the selection of season length *is* to some extent arbitrary, as with camera-trap or bioacoustic data where continuous recordings can be broken down into distinct 'seasons' of any length.
-The DOM may be readily applied to these data types [@balantic2019], and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data.
-For example, @kleiven2020 and @mölle2022 divide their camera trap records into seasons of just a few days.
-While this provides interesting insights of occupancy at extremely fine temporal scale, further research could help provide general guidance on determining appropriate season and survey durations with respect to research questions and the closure assumption.
+The DOM may be readily applied to these data types, and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data [@balantic2019].
+This is already apparent in recent camera trap studies which collapsed their detections into seasons of just a few days, far shorter than is realistic with conventional survey methods [@kleiven2020; @mölle2022].
+While these studies provide interesting insights of occupancy at extremely fine temporal scale, further research could help provide general guidance on determining appropriate season and survey durations with respect to research questions and the closure assumption.
 
@@
 Building dynamic occupancy models can be a challenging process requiring careful consideration of which environmental factors to incorporate to adequately represent occupancy and detection in complex natural systems.
-In this review, we have recorded all covariates considered for each model in our sample.
-These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of articles considering each type of covariate in their models, the means by which covariate data are collected, and how covariates responses are represented in DOMs.
+In this review, we recorded all covariates considered for each model in our sample, including those not used in final models.
+These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of studies considering each type of covariate in their models, the means by which covariate data are collected, and how covariates responses are represented in DOMs.
 We further delineate their use in the model by classifying covariates into two groups: environmental covariates, representing plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environment.
-Our findings indicate that users of DOMs have incorporated a wide diversity of covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample.
-Many models incorporate covariates representing aspects of site geometry and connectivity (37% of articles), adding spatial components to these DOMs which bear similarities to related stochastic patch occupancy models.
-Notable examples of these spatially-explicit models include @mortelliti2007's model fit to rodent data in patchy landscapes and @risk2011's hybrid DOM/incidence function model used to explore meta-population dynamics of wetland birds.
-Several studies also include biotic interactions with other species as covariates, particularly those which focussing on taxa threatened by invasive species.
-Examples include models with barred owl presence as a covariate on spotted owl occupancy and models with bullfrog covariates on native amphibian occupancy, each of which effectively incorporate species interactions without fitting explicitly multi-species models [@mangan2019; @rowe2019].
+Our findings indicate that DOM users have incorporated a wide diversity of covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample.
+Many models incorporate covariates representing aspects of site geometry and connectivity (35% of studies).
+Often these are included as simple covariates, as in @duggan2011's models using landcsape connectivety metrics as covariates on colonisation and extinction.
+Alternatively, more complex parameterisations explicitly model colonisation or extinction as a spatial process, blending attributes of DOMs and stochastic patch occupancy models [@broms2016; @risk2011].
+Several studies also include biotic interactions with other species as covariates, particularly those focused on taxa threatened by invasive species.
+This is observed frequently in the Spotted owl literature, where Barred Owl presence is often considered as a covariate driving Spotted owl occupancy dynamics [@olson2005; ; @sovern2014; @dugger2016].
+The use of these types of covariates effectively incorporate species interactions in DOMs without requiring the use of more complex explicitly multi-species models.
 
@@
 #| tbl-cap: "All covariates considered for inclusion in a study were classified into mutually exclusive categories.
-We calculate the percentage of articles which include at least one covariate of a given category on any parameter, Initial Occupancy (&psi;1), Occupancy (&psi;), Colonisation (&gamma;), Extinction(&epsilon;), and Detection (&rho;).
-We also present the percentage of all implementations of each covariate category which are dynamic (varying through seasons) and directly observed, as well as the percentage of articles which model each category of covariate as a non-linear relationship or as part of an interaction with another covariate."
+We calculate the percentage of studies which include at least one covariate of a given category on any parameter, Initial Occupancy (&psi;1), Occupancy (&psi;), Colonisation (&gamma;), Extinction(&epsilon;), and Detection (&rho;).
+We also present the average percentage of covariates in a category which are dynamic (varying through seasons) and directly observed, as well as the percentage of articles which model each category of covariate with a non-linear relationship or as part of an interaction with another covariate."
 
@@
     "OTHER-E" ~ html("<b>Other environmental</b>\n<i>Other environmental covariate not otherwise listed</i>"),
-    "TOPO" ~ html("<b>Topography</b>\n<i>Elements of landscape topography, including hydrology</i>"),
+    "TOPO" ~ html("<b>Topography</b>\n<i>Elements of landscape topography</i>"),
+    "HYDR" ~ html("<b>Hydrology</b>\n<i>Aspects of hydrology, such as distance to water</i>"),
     "CLWE" ~ html("<b>Climate/Weather</b>\n<i>Climate, weather, and natural disasters</i>"),
@@
 
-Covariate data for studies in our sample was either collected directly by researchers (17% of environmental covariate data), or derived from pre-existing remotely sensed datasets (83%); this of course varies depending on the category of covariate (@tbl-covariates).
-Directly collected data can generally be expected to represent finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely but can often be more proximal drivers of occupancy.
-These species-specific covariates do come with trade-offs, as they can be expensive to collect and can often preclude projection to locations where data are unavailable.
-For projects interested in making such projections, remotely-sensed covariates are generally more feasible despite their generally more distal nature [@austin2002].
-Forty two percent of environmental factors and 87% of structural factors included in reviewed articles were dynamic covariates which varied through time — this again varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static (@tbl-covariates).
+Covariate data for studies in our sample was either collected directly by researchers (an average of 30% of environmental covariates per study), or derived from pre-existing remotely sensed datasets (70% of environmental covariates); this of course varies depending on the category of covariate (@tbl-covariates).
+Directly collected data can often represent finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely but can often be more proximal drivers of occupancy.
+These species-specific covariates do come with trade-offs, as they can be expensive to collect and often preclude projection to locations where these data are unavailable.
+For studies interested in making such projections, remotely-sensed covariates are generally more feasible despite their generally more distal nature [@austin2002].
+An average of 43% percent of environmental factors and 94% of structural factors included in reviewed studies were dynamic covariates which varied through time — this again varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static (@tbl-covariates).
 
-In the standard DOM, covariates for each parameter are most commonly incorporated via a logistic regression (i.e., a linear regression through a logit link function).
+In the standard DOM, covariates for each parameter are most commonly incorporated via a logistic regression (i.e., a linear regression through a logit link function) [@mackenzie2017].
 Statistical relationships between model parameters and covariates (e.g., between initial occupancy and its environmental covariates) are represented as linear terms unless more complexity is specified.
@@
 @austin2007 discusses the importance of modelling ecologically realistic responses to covariates, advocating for careful consideration of the most appropriate statistical form for hypothesised relationships.
-Non-linear responses can be easily accommodated in the DOM by using polynomial transformations and interactions between covariates.
+Non-linear responses can be easily accommodated in DOMs by using polynomial transformations and interactions between covariates.
 Despite this, in our sample only 35% of articles tested one or more non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear terms (@tbl-covariates).
-Interactions between covariates were similarly rare, with only 26% of studies considering at least one interaction between terms (@tbl-covariates).
+Interactions between covariates were similarly rare, with only 24% of studies considering at least one interaction between terms (@tbl-covariates).
 The relatively low emphasis on more complex non-linear responses contrasts with other popular methods for modelling species occupancy.
-Many common approaches for SDMs, which include MAXENT and Boosted Regression Trees, permit considerable flexibility in the shape of their covariate response curves and use of interactions where supported by the data [@merow2013; @elith2008].
+Many common approaches for SDMs, such as MAXENT and Boosted Regression Trees, permit considerable flexibility in the shape of their covariate response curves and use of interactions where supported by the data [@merow2013; @elith2008].
 This emphasis on more complex responses in SDMs may be due to their frequent application across relatively large geographic extents which might encompass the full species niche, where environmental relationships may be expected to be non-linear.
-However, as previously indicated the DOM has also been implemented across large spatial extents where the same assumptions should exist and similar responses might be expected.
-Similar concerns exist for the relatively low level of interactions between covariates: these relationships are commonly expected based on ecological theory, and their exclusion can negatively impact model performance [@guisan2006].
+However, as previously indicated DOMs have also been implemented across large spatial extents where the same assumptions should exist and similar responses might be expected.
+Similar concerns exist for the relatively low level of covariate interactions, given that these relationships are commonly expected based on ecological theory and that their exclusion can negatively impact model performance [@guisan2006].
 
 In addition to the types of covariates considered for modelling we also recorded the size of the candidate covariate pool in each reviewed study, tallying the number of environmental and structural covariates which were available for use on each parameter ( @fig-covariates).
-This is an area of considerable variation in modelling practices — the number of covariates considered for each parameter ranges from 0 (effectively modelling the parameter as a constant) to over 40 possible factors.
+This is an area of considerable variation in modelling practices — the number of covariates considered for each parameter ranges from 0 (effectively modelling the parameter as a constant) to over 40 candidates on a single parameter.
 
 ![The number of covariates considered for each parameter across all studies in our sample.
-'Occupancy' given here represents an alternative parameterisation of the DOM which jointly estimates Occupancy for every season, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation.
+'Occupancy' given here represents the alternative parameterisation of the DOM which jointly estimates Occupancy for every season, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation.
 Here, a 'covariate' is defined as each distinct covariate considered for inclusion.
@@
 
-The median number of covariates varies strongly by parameter, with transition probabilities (colonisation and extinction) more likely to consider a broader range of environmental covariates relative to initial occupancy and detection.
-The lack of any covariates considered for initial occupancy in many studies (37% of articles) is particularly notable — unless a study is conducted at very small extents or study sites are truly uniform in their suitability, one would expect some amount of non-random variation in occupancy probability across any study system that will not be captured when representing initial occupancy as a constant.
+The median number of covariates varies strongly by parameter, with transition probabilities (colonisation and extinction) more likely to consider a broader range of environmental covariates compared to initial occupancy and detection.
+The lack of any covariates considered for initial occupancy in many studies (37% of studies) is particularly notable — unless a study is conducted at very small extents or study sites are truly uniform in their suitability, one would expect some amount of non-random variation in occupancy probability across any study system that will not be captured when representing initial occupancy as a constant.
 Omission of the factors that drive occupancy is a known source of bias in static SDMs [@barry2006], which the initial occupancy component of the DOM conceptually resembles.
@@
 Identifying the 'best' model from a range of possible candidates can be a challenging process, especially for hierarchical models like DOMs [@doherty2012].
-Consider that the quantity of candidate models involving up to *n* covariates on *p* parameters can be given by $(2^n)^p$; where a simple linear regression may have up to 64 possible combinations of 6 covariates, a 4-parameter dynamic occupancy model would have 16 million combinations.
-This explosion of candidate models generally precludes exhaustive comparisons of possible models and requires a reduction of the models tested.
-Models in our sample include both Frequentist (n = 70) and Bayesian (n = 22) frameworks, which differ considerably in their manner of implementation.
+Consider that the quantity of candidate models involving up to *n* covariates on *p* parameters can be given by $(2^n)^p$: where a simple linear regression may have up to 64 possible combinations of 6 covariates, a 4-parameter dynamic occupancy model would have 16 million combinations of these same covariates.
+This explosion of candidate models generally precludes exhaustive comparisons of possible models and requires some reduction of the models tested.
+Models in our sample include both Frequentist (n = 76) and Bayesian (n = 24) frameworks, which differ considerably in their manner of implementation; some articles in our sample included multiple distinct modelling workflows, resulting in a higher total number of models than articles.
 A summary of modelling practices in DOMs is given in @tbl-modelling.
@@
 The median covariate count presented here represents the median quantity of covariates considered for each model parameter across the studies in our review.
-The model selection methods represented in this table are non-exclusive and some articles employ multiple approaches."
+The model selection methods represented in this table are non-exclusive and some articles employ multiple approaches.
+2 models included in the 'Overall' column are neural network based and fall into neither the Frequentist or Bayesian categories."
 
@@
 The largest differences between the frequentist and Bayesian models in our sample lie in their approaches to model selection.
-Where 94% of frequentist models perform some manner of model selection to determine covariate inclusions for their models, only 36% of Bayesian models do so, with the majority instead fitting a single model defined *a priori*.
+Where 95% of frequentist models perform some manner of model selection to determine covariate inclusions for their models, only 33% of Bayesian models do so, with the majority instead fitting a single model defined *a priori*.
 Methods used for model selection vary considerably both within and between the two modelling frameworks (@tbl-modelling).
-For frequentist models, the most popular and conventional approach to model selection as used in 43% of articles involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score.
-The next most popular method in frequentist studies is to use procedural model selection methods (39% of articles), where the structure for each model parameter is fit one-by-one.
-For example, this protocol might first identify the best structure for detection probability before moving on to initial occupancy and so on.
-The remainder of frequentist studies (9%) use a variety of approaches, such as fitting simpler models like single season occupancy models to identify the most informative terms to use in a DOM.
-Across all frequentist implementations, 71% of articles summarise a final subset of models using multi-model inference and model-averaging with AIC weights [@burnham2004].
+For frequentist models, the most popular and conventional approach to model selection (45% of models) involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score.
+The next most popular method in frequentist studies is to use procedural model selection methods (37% of models), where the structure for each model parameter is fit in sequence.
+For example, this protocol might first identify the best structure for detection probability while holding the other parameters constant, before moving on to initial occupancy and so on until all parameters are fixed.
+The remainder of frequentist studies (8%) use a variety of approaches, such as fitting simpler models like single season occupancy models to identify the most informative terms to use in a DOM.
+Across all frequentist implementations, 47% of articles summarise a final subset of models using multi-model inference and model-averaging with AIC weights [@burnham2004].
 
-Those Bayesian models which *do* perform model selection tend to take various approaches, with largely idiosyncratic methods across these studies.
+Those Bayesian models which *do* perform model selection take various approaches, with largely idiosyncratic methods across these studies.
 While direct comparison of model fit is rare in Bayesian methods, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data.
 Another approach used by @cook2022 fit a global model including all covariates before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the model.
-@ahumada2013 takes a hybrid approach, in which model selection is conducted by a procedural method in the frequentist framework before refitting the best structure in a Bayesian model.
+@ahumada2013 takes a hybrid approach, in which model selection is conducted by a procedural method in the frequentist framework before refitting the best structure as a Bayesian model.
 
-Limited research has been conducted on advantages of different methods for covariate selection in DOMs, and there is unlikely to be a one-size-fits-all approach which will be appropriate for all possible use-cases.
+Limited research has been conducted on the advantages of different methods for covariate selection in DOMs, and there is unlikely to be a one-size-fits-all approach which will be appropriate for all possible use-cases.
 However, it is important to consider the implications of the different model selection approaches in common usage, and research from the SDM literature highlights how covariate selection can influence our interpretation of model outputs [@brodie2020].
@@
 
-The model selection methods identified in our review are designed only to identify the best model of those considered, and do not necessarily guarantee that selected models are well-fit to the data.
+Regardless of covariate selection protocol, 'the selection of a best model does not guarantee the selection of a good model' [@mackenzie2004].
 Assessment of the performance of DOMs generally requires additional steps, and the best ways to achieve this are not yet clear.
-There is no broadly-accepted goodness-of-fit test for dynamic occupancy models, although @mackenzie2004 describes an approach for single-season occupancy models using a parametric bootstrap which can been extended to DOMs; this test is implemented in the *AICcModAvg* and *unmarked* R packages.
+There is no broadly-accepted goodness-of-fit test for dynamic occupancy models, although @mackenzie2004's approach for single-season occupancy models using a parametric bootstrap which can been extended to DOMs; this test is implemented in the *AICcModAvg* and *unmarked* R packages.
 @kéry2021 describe the test and present an alternative based on separately assessing fit to static and dynamic components of the model.
 In Bayesian implementations, posterior predictive checks offer means to assess model fit [@gelman2014].
-@broms2016 discusses possibilities for model evaluation in the Bayesian context focusing on single-season multi-species occupancy models but extensions of their approach may also be applicable to the DOM.
-As in other hierarchical models, model evaluation for DOMs can in general can be difficult and somewhat uncertain compared to other model types, as the primary response variable of interest (species occupancy) is a latent variable where the true state is generally not known.
+@broms2016 discusses further possibilities for model evaluation in the Bayesian context focusing on single-season multi-species occupancy models, and extensions of their approach may also be applicable to the DOM.
+As with other hierarchical models, model evaluation for DOMs can be difficult and somewhat uncertain compared to other model types, as the primary response variable of interest (species occupancy) is a latent variable where the true state is generally not known.
 Predictive performance evaluation is thus typically based on *observed* occupancy data, where a DOM is used to simulate detection histories to be compared with field survey results.
@@
 Perhaps as a result of these difficulties, assessment of model fit and model performance was rare amongst the articles in our sample.
-Only 19% of studies tested for goodness-of-fit, and just 10% calculated predictive performance.
+Only 18% of studies tested for goodness-of-fit, and just 7% calculated predictive performance.
 These rates are considerably lower than those reported for SDMs; for which closer to 50% of articles were found to test both fit and performances [@araújo2019].
@@
 Studies in our sample were classified according to their objectives as stated by authors to explore the use-cases which the DOM has been applied to (@fig-objectives).
-These objective categories were non-exclusive, and many studies fell into multiple categories.
+These objective categories are non-exclusive, with many studies falling into multiple categories.
 Across the study period the DOM has been frequently employed to estimate trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions.
@@
 
-Many articles in our sample used the DOM seeking to monitor trends in occupancy state through time ("Estimate trends" in @fig-objectives).
-These studies include @ahumada2013's paper assessing populations trends of 13 forest mammals and @scott2015's assessment of the conservation status of the endangered flattened musk turtle.
+36% of articles used the DOM to monitor trends in occupancy state through time ("Estimate trends" in @fig-objectives A), both for single species of high conservation interest and for broader community assemblages of species [@ahumada2013; @scott2015].
 These studies offer valuable insights on the state of species across landscapes and through time, demonstrating the DOM's suitability for monitoring oriented projects; including those where multiple species need to be assessed at once.
-Other articles were more focused on examining the factors which influence species occupancy, either testing pre-specified hypothetical relationships or taking a broader tack to identify drivers of occupancy without preconception.
+Other articles were more focused on examining the factors which influence species occupancy, either testing pre-specified hypothetical relationships (44% of articles) or taking a broader tack to identify drivers of occupancy without preconception (35%).
 The distinction between the two is important, as it guides which covariates may be considered and how model selection may be used to fit models.
 Many of these studies target core conservation priorities for their focal species, like @olson2005's early DOM assessing the influence of barred owls on threatened spotted owl.
-Explorations of these pivotal relationships are important for guiding management action, and may also be used to test the effectiveness of these actions as in @miller2023's study on American Kestrel nest boxes.
-In scenarios where less is known about species habitat preference, DOMs may be used to examine the influence of a wider variety of factors as illustrated in @huber2017's study testing the relative influence of dozens of habitat covariates on Wood warbler occupancy.
-In recent years, increasing numbers of articles have used to DOM to generate predictions of species occupancy.
-As examples, @mcgowan2020 provides projections of future occupancy for the threatened Black rail under multiple management scenarios, and @pollentier2021 generates maps of distributions resembling those made with SDMs, outputs which can immediately be used by managers to inform action.
+Explorations of these pivotal relationships are important for guiding management action, and may also be used to test the effectiveness of these actions [@miller2023].
+In scenarios where less is known about species habitat preference, DOMs may be used to examine the influence of a wider variety of factors as in @huber2017's study testing the relative influence of dozens of habitat covariates on Wood warbler occupancy.
+In recent years, increasing numbers of articles have used DOMs to generate predictions of species occupancy; this a use-case for which DOMs show considerable promise [@kéry2013; @briscoe2021].
+These predictions often have strong utility for conservation management.
+For example, @mcgowan2020 projects future occupancy for wetland birds under alternative management scenarios, and @pollentier2021 generates distribution maps resembling those made with SDMs.
 
-![A) Shows the proportion of articles in each year-strata and across all years which match each of six non-exclusive objective categories.
-B) Shows the quantity of covariates considered per parameter for models which pursued each objectives.
-Studies focused on methods development typically considered the fewest covariates, where those attempting to identify the drivers of occupancy considered the most.](Figures/ObjectiveFigure.jpeg){#fig-objectives}
+![A) Proportion of articles in each year-strata and across all years which match each of six non-exclusive objective categories.
+B) Quantity of covariates considered per parameter for models which pursued each objectives.](Figures/ObjectiveFigure.jpeg){#fig-objectives}
 
-The objectives of studies fitting a DOM strongly informs approaches to model building, particularly with respect to covariate selection.
-@fig-objectives B shows the quantity of covariates considered for models applied to each of our objective categories, and the differences between objectives are apparent.
+Study objectives strongly inform approaches to model building, particularly with respect to covariate selection.
+@fig-objectives B presents the quantity of covariates considered for models applied to each of our objective categories, with apparent differences between objectives.
 Not unexpectedly, articles which focused on describing new methods for DOMs had fewer covariates than those studies focused on more applied objectives.
@@
 This precludes any prescription of the 'best' way to build a model, however, there are still important discussions to be had on decisions made in the modelling process.
-One aspect of fitting DOMs meriting broader discussion revolves around 'model complexity,' and how much must be incorporated into models to reliably model species occupancy under different contexts and use cases.
-Complexity is a broad term which encompasses many aspects of a model [@merow2014], and opinions on simplicity versus complexity in ecological models are diverse.
+One aspect of fitting DOMs meriting broader discussion centres on 'model complexity,' and how much must be incorporated into models to reliably model species occupancy under different contexts and use cases.
+Complexity is a broad term encompassing many aspects of a model [@merow2014], and opinions on simplicity versus complexity in ecological models are diverse.
 Where some advocate for the simplest possible models, arguing that they are most generalisable; others insist that overly-simple models cannot adequately represent the most important drivers in a system [@evans2013; @lonergan2014].
-By their nature the DOM is somewhat more complex than simpler models for studying occupancy thanks to their hierarchical structure, which is necessary to control for detectability and to capture occupancy dynamics.
+By their nature the DOM is somewhat more complex than simpler models for studying occupancy due to their hierarchical structure, which is necessary to control for detectability and to capture occupancy dynamics.
 Within this structure, however, further complexity is to some degree up to the modeller: one can choose how many covariates to consider for inclusion on the various parameters, and how to represent the nature of the relationship between those covariates and parameters.
 Research from SDMs indicates that allowing for more complex relationships can improve model performance in predicting occupancy [@valavi2023], an increasingly popular use-case for the DOM.
-Within the DOM framework, there are promising developments on that front — for example, @joseph2020 presents a novel neural-network occupancy model which allows for exponentially higher levels of complexity and may offer improved performance for prediction-oriented studies.
+Within the DOM framework, there are promising developments on that front: @joseph2020 presents a novel neural-network occupancy model which allows for exponentially higher levels of complexity and may offer improved performance for prediction-oriented studies.
 
 Covariate selection seems to be a particularly important area for further investigations into building DOMs.
-In our review, we see little consensus around which approaches are most applicable for any given use case, and existing work on covariate selection in DOMs raises concerns on whether common methods always produce the most suitable models.
+In our review, we see little consensus around which approaches are most applicable for any given use case, and existing work on covariate selection in DOMs raises concerns on whether common methods always produce the most suitable models [@morin2020; @stewart2023].
 This is true for both frequentist and Bayesian implementations, and comparative research on covariate selection under both frameworks may help to inform model users on which method is most appropriate for their use-cases.
@@
 
-These priorities are particularly important given the frequent applied objectives of DOM users, tackling challenges which include assessing critically endangered species [@carvalho2023], guiding public health management of disease vectors [@moreira2016], and tracking rapidly developing biological invasions [@wood2020].
+These priorities are particularly important given the frequent applied objectives of DOM users, tackling challenges which include assessing critically endangered species, guiding public health management of disease vectors, and tracking rapidly developing biological invasions [@carvalho2023; @moreira2016; @wood2020].
 The DOM is well suited to these situations, and it is to be expected that as these types of applications are more commonly attempted understanding the sensitivity of model outputs to decisions made in the model fitting process becomes increasingly important.
 In the two decades since the publication of @mackenzie2003 the dynamic occupancy model has become a widely used tool for ecological inference, with numerous extensions to the modelling framework further broadening the scope of questions and use-cases for which it may be applied.
-Given their increasing popularity, further research and guidelines around issues of model building may help to make the DOM more accessible to newcomers and support confidence in interpretation of models.
-Parallels to existing work in the SDM literature on guidelines for reporting and modelling standards could be valuable contributions in achieving these aims [@zurell2020; @araújo2019] could provide useful examples in developing methods and guidelines specifically focussed on the DOM.
+Given their increasing popularity, further research and guidelines around issues of model building may help to make the DOM more accessible to newcomers and support confidence in model interpretation.
+Parallels to existing work in the SDM literature on guidelines for reporting and modelling standards could be valuable contributions in achieving these aims and provide useful examples in developing methods and guidelines specifically focused on DOMs [@zurell2020; @araújo2019].

### 2024-10-14 Supplement/Analysis.qmd
Changes during article submission

--- 2024-09-26 Analysis.qmd
@@
 title: "Analysis and figures"
-author: "Saoirse Kelleher"
-date: today
+date: September 26 2024
 abstract: "Calculates key statistics and renders figures for \"*Twenty years of dynamic occupancy models: a review of applications and look to the future*\""
@@
     theme: minty
+embed-resources: true
 ---
@@
 ```{r Load query sheets}
-sheet_AllArticles <- list(read_xlsx("Randomisation/Randomisation.xlsx", 
+sheet_AllArticles <- list(read_xlsx("Randomisation.xlsx", 
                                     sheet = "2004-2007"),
-                          read_xlsx("Randomisation/Randomisation.xlsx", 
+                          read_xlsx("Randomisation.xlsx", 
                                     sheet = "2008-2011"),
-                          read_xlsx("Randomisation/Randomisation.xlsx", 
+                          read_xlsx("Randomisation.xlsx", 
                                     sheet = "2012-2015"),
-                          read_xlsx("Randomisation/Randomisation.xlsx", 
+                          read_xlsx("Randomisation.xlsx", 
                                     sheet = "2016-2019"),
-                          read_xlsx("Randomisation/Randomisation.xlsx", 
+                          read_xlsx("Randomisation.xlsx", 
                                     sheet = "2020-2023")) |> 
@@
 The next set of statistics and figures focus on the study areas included in reviewed articles.
-For each article the countries where data was collected, a rough centroid of the study area, and an estimate of the spatial extent of the study area was recorded.
+For each article the countries where data was collected, a rough centroid of the study area and an estimate of the spatial extent of the study area was recorded.
 
@@
 
-First, a table is generated presenting the number of articles using data from each country with at least one article.
+First, a table is generated presenting the number of articles using data from each country.
 
@@
 ::: callout-important
-## Articles may include multiple study areas in cases where distinct dataset were modelled separately
+## Articles may include multiple study areas in cases where distinct datasets were modelled separately
 :::
@@
 # Load country shapefile
-countries_shp <- st_read("Analysis/Countries_Spatial/World_Countries__Generalized_.shp", quiet = TRUE) |>
+countries_shp <- st_read("Spatial/World_Countries__Generalized_.shp", quiet = TRUE) |>
   st_transform(crs = "ESRI:54030") |>
@@
 
-#### Modelling table
+In this section, we gather information on practices in the modelling process to generate a summary table.
+This involves aspects including the median number of covariates per parameter, the proportion of studies using different model selection approaches, and the proportion of studies evaluating their models.
+This data is split between Bayesian and Frequentist models.
 
@@
 
-The objectives of each article were documented, split into 6 non-exclusive categories.
+In this section we summarise information on the objectives of each study. Objectives were described in six non-exclusive categories.
 
@@
 #### Objectives figure
+
+We generate a plot (Figure 6 in the review) showing two elements related to objectives. First, we calculate the proportion of articles in each year-strata pursuing each objective.
 
@@
 
-We render a second figure for the average number of covariates used in articles approaching each objective
+We then render a second figure for the average number of covariates used in articles approaching each objective.
 
@@
 
-Combine the objective plots
+We combine the objective plots using the `patchwork` package.
 

### 2025-01-23 Manuscript.qmd
Tweaked figures
preserve the initial version and begin scaffold for new one

@@
 abstract: | 
- Understanding the patterns by which species occur across landscapes and throughout time is a long-standing objective of ecological research which has inspired the development of numerous quantitative modelling approaches for the task.
-However, estimating occupancy can be a challenge, particularly where issues like imperfect detection and shifting distributions exist.
-Dynamic occupancy models (DOMs) offer a framework for occupancy estimation which explicitly accounts for observation error while incorporating additional mechanistic process by estimating colonisation and local extinction.
-In light of increasing interest in more process-explicit models for understanding species occurrence, here we examine how DOMs have been applied to field ecological data in the two decades since their introduction.
-Following a general introduction to the model, we present the results of a systematic review exploring the research contexts where DOMs have been applied and interrogating how authors have built their models, with particular emphasis on how covariates are incorporated to describe variation in occupancy dynamics.
-Our findings indicate that DOMs are a flexible tool readily applied to diverse study systems and data types, with their usage expanding in recent years as more studies apply them to make spatial and temporal prediction of species occupancy.
-DOMs are also amenable to extension, further broadening their utility.
-Despite their flexibility, model complexity in DOMs tends to be low; most studies consider relatively few covariates, most of which are represented as simple linear relationships.
-Approaches to covariate selection also vary considerably, and there remains little research on how these choices may influence model performance.
-Furthermore, only a fraction of articles report evaluating models and little guidance exists on how to approach this task in DOMs.
-These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.
+ Occupancy models.
+
+format:
+  html:
+    theme: simplex
+
+embed-resources: true
+title-block-banner: true
 ---
@@
 
-The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849].
-Estimates of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, and identifying potential drivers of occurrence can help inform potential management actions [@mackenzie2013].
-Robust knowledge of the occupancy patterns of a species can also help us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios [@kéry2013].
+Background Information section
 
-While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate.
-The need to understand and quantitatively describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly applied to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales.
-However, several factors can make occupancy difficult to estimate.
-For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with much wildlife field data — generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004].
-Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014].
-Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species' occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010].
-These conditions often occur during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].
+#### Model form and assumptions
 
-The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017a] .
-Drawing on principles from the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy.
-Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time.
-This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes which drive occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic or abundance data required by more process-explicit models [@briscoe2019].
-These features make the DOM an important tool, with uses including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].
-
-In this review we examine how dynamic occupancy models have been used by ecologists in the two decades since their inception.
-Following an introduction to the model's form and assumptions, we present the results of a systematic review exploring how researchers have applied DOMs to ecological data, with emphasis on how they collected their data, selected covariates, and evaluated their models.
-Based on these results we highlight the DOM's flexibility as a tool for understanding species occurrence, examine approaches to the model building process, and outline key priorities for future research involving this model class.
-
-### Dynamic occupancy model form and assumptions
-
-The DOM structure encompasses two processes: the ecological process of site occupancy dynamics describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed at a site where it is present (@fig-modelform).
-Sites can exist in either an occupied (z = 1) or unoccupied state (z = 0).
-In the first time step, occupancy state at each site is determined by the probability of initial occupancy ψ~1~, i.e., $z_1 \sim Bernouilli(\psi_1)$.
-In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation γ and local extinction ε, such that $z_t \sim Bernoulli(\psi)$ and $\psi = z_{t-1}(1-\epsilon)+(1-z_{t-1})\gamma$.
-The observation component of the DOM explicitly accounts for imperfect detection.
-At occupied sites, the detection probability ρ describes whether the species is observed during a survey (y = 1) or not (y = 0), according to the formula $y_{t,j} \sim Bernouilli(z_t\rho)$.
-Under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur.
-In this review we use the term "parameter" to refer to the quantities being estimate: in most cases, these are initial occupancy ψ~1~, colonisation γ, extinction ε, and detection ρ (see @fig-modelform).
-However, some alternative model forms may directly estimate occupancy ψ as a parameter.
+Model form.
 
@@
 
-To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys.
-Under this design, observations at a site occur during distinct, time-bound seasons within which it is assumed site occupancy does not change (that is, sites are closed to changes in occupancy).
-In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy.
-Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@guillera-arroita2017a; @guillera-arroita2017b].
-It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this enables flexibility in data inputs.
+Survey design.
 
@@
 
-DOMs make several key assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.
+Assumptions.
 
-I.
-**False positive detections do not occur**.
-While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods.
-@mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need to consider how certain detections truly are for any given study system.
-Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019].
-Where this assumption can not reasonably be met, model extensions designed to account for false positive error should be considered [@royle2006; @miller2011; @miller2015].
+# Review methods
 
-II.
-**Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion in terms of the bias introduced when it is violated [@rota2009; @otto2013].
-Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and it can be difficult to distinguish local extinction from temporary emigration [@valente2017].
-Model extensions to relax the closure assumption have been developed, including approaches using staggered arrival and departure periods between sites [@kendall2013].
-A more pertinent approach for most studies, however, is careful consideration of the appropriate definition of occupancy under the survey design used [@steenweg2018].
+Methods.
 
-III.
-**Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species occurrence, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system.
-Generally, this is achieved by allowing model parameters (ψ~1~, γ, ε, and ρ) to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters.
-An important element of this assumption is that the likelihood of detection of a species can depend not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017a].
-While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates [@barry2006].
-Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known about how this latent heterogeneity can influence model performance.
+# Results
 
-Since its original description in @mackenzie2003, the DOM has been further developed with numerous model extensions and alternative formulations including aforementioned implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010].
-For a comprehensive discussion of common extensions and their applications see @bailey2014 and @guillera-arroita2017a, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.
-
-# Systematic review methods
-
-To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data.
-A pool of candidate articles was generated using two queries on Web of Science.
-The first of these included all articles from 2004-2023 which cite @mackenzie2003.
-To capture any additional relevant articles that did not directly cite @mackenzie2003, a second query was generated searching articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract.
-These queries resulted in 1469 articles: 897 were retrieved only from the MacKenzie citations, 274 only from the keywords search, and 298 which appeared in both queries.
-As we were interested in how DOMs use has changed through time, we divided all articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024.
-From each of these strata we randomly selected 20 articles for inclusion in the review.
-Articles that did not meet inclusion criteria were replaced from within their own strata.
-
-As our review focuses only on applications of the dynamic multi-season occupancy model of @mackenzie2003 and its extensions, we included articles that fit a model meeting the following criteria: i) used non-simulated, field-collected, presence-absence data; ii) relied on data from multiple sites which could exist in at least two states, including an occupied and unoccupied state; iii) had multiple seasons, between which sites could change states conditional on the prior season's occupancy state and transition probabilities such as colonisation and extinction; and iv) contained at least one parameter describing the detection process.
-
-For each article we recorded key details on authorship, research objectives, study taxa and system, survey methods, and modelling approach.
-To examine the reasons why authors used DOMs, we allocated each article to one or more category of objective based on the study's stated aims.
-The possible categories were *Estimating trends*, where authors express interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; *Testing relationships*, where authors explore specific predefined hypotheses of relationships between covariates and model parameters; *Identifying drivers*, where authors attempt to find which covariates influence model parameters; *Predicting temporally*, where authors predict site occupancy under future conditions, *Predicting spatially*, where authors predict site occupancy to unsurveyed locations, and *Developing methods*, where authors introduce, test, or demonstrate aspects of dynamic occupancy models.
-
-We recorded details on the type of organisms (bird, mammal, etc.) modelled in each article, and the means by which multiple species were modelled where applicable.
-Taxa were denoted as threatened either when they are listed on the IUCN Red List of Species or when authors indicate that they are threatened.
-This deference to authors' representation of conservation status was made to account for sub-species which lack listings or species which are of more local concern.
-Study locality and size was documented; the size of the study area being defined as the intended area of inference which contained all sites — this was measured to the order of magnitude to account for uncertainty in reporting.
-
-We were particularly interested in how authors navigated the modelling process, from covariate selection through model evaluation.
-To this end, we recorded all covariates considered in each study regardless of whether they were or were not included in final models.
-Key traits of each covariate were recorded including their general category, whether they were directly observed or remotely sensed, whether they were static or varied between seasons, and how they were presented in the model (as a linear term, a polynomial term, or as part of an interaction with another covariate [@james2021]).
-Model selection procedures were also sorted into non-exclusive categories including *A priori*, where only one model was considered; *candidate suite*, where a predefined set of models was considered; *sequential,* where covariates were selected parameter-by-parameter (e.g., fitting detection first, followed by initial occupancy and so on); and *simple precursors*, where selection was preceded by another simpler model implementation.
-Finally, for each model we documented whether goodness-of-fit was tested and reported, and whether model performance was assessed by validation with either in or out-of-sample data.
-For the full spreadsheet of data collected and further details on categorisation, see supplementary material.
-
-# Applications of dynamic occupancy models
-
-A total of 92 articles were included for this review.
-Based on the acceptance rates within each stratum and quantity of unprocessed articles, an estimated 496 of the 1152 unreviewed articles in our sample would have met inclusion criteria, with an apparent increase over time in the number of articles fitting DOMs (@fig-coverage).
+Coverage.
 
@@
 
-The articles included in our review demonstrate considerable diversity in scope, scale, and objectives.
-A selection of attributes of these studies is presented in @fig-StudyDetails, and throughout this section we provide notable examples of DOMs to highlight how the noted variation appears in practice.
-Study systems included in our sample are globally distributed: while a majority of articles use data collected in the United States of America, representatives are included from all geographic realms and a broad diversity of ecosystems (@fig-StudyDetails A).
-A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale (@fig-StudyDetails B) .
-The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre, while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@basset2023; @clement2019].
-
-The temporal scale of studies shows similar variability.
-The time elapsed between the first and last survey ranged from under one month to over forty years, with a median duration of 8 years (@fig-StudyDetails F).
-More meaningful, however, is the period of the primary occasion, as this represents the temporal scale at which changes in occupancy are assumed to occur.
-Most applications describe primary occasions as occurring annually, although some studies divide years into multiple primary occasions to describe seasonal variation in occupancy.
-In the most extreme cases the primary occasion may be as brief as a week and capture much finer scale changes in occupancy.
-These shortened-season occupancy models are most common with camera trapping or acoustic monitoring data, which can be arbitrarily divided into primary and secondary sampling occasions.
-One example of this is @kleiven2020's study of rodent and mustelid interactions using camera trap data, which used primary occasions less than one week in length.
-On the other end of the spectrum, some studies modelled primary occasions which were decades apart and represented generational changes in occupancy.
-This is illustrated in @couturier2023's study on long-term otter recovery in France which used data from two primary occasions: one in 2003 and one in 2012.
-
-DOMs have been used to study a variety of species, though the majority of studies have focused on vertebrate taxa (@fig-StudyDetails C).
-The DOM has been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community.
-However, there are exceptions, including DOMs used to model decadel changes in lichen occupancy and the spread of chronic wasting disease [@belinchón2017; @cook2022].
-The latter's application to disease dynamics is not unique, and the DOM has been touted as a valuable tool for such applications [@bailey2014].
-Multiple authors have also used DOMs to model mosquito dynamics, a strongly applied use-case with important human health implications [@mores2020; @padilla-torres2013].
-The vast majority of studies model terrestrial species, though there have been a limited number of studies on aquatic systems, including invasive salmon, Great Plains stream fishes, and whales [@fisher2014; @falke2012; @pendleton2022].
-
-While most studies in our sample fit a model to a single species, 44% fit models to multiple species either as independent models (34%) or explicitly multi-species implementations (12% — some articles attempted multiple approaches; @fig-StudyDetails D).
-The multi-species models use various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects [@dorazio2010; @hendershot2020], as well as explicit interaction models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019].
-Rather than using multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012].
-Working with large numbers of taxa does raise additional challenges, as the degree to which models are tailored to the taxa is likely to be limited for the sake of practicality.
+Stamp collection.
 
@@
 
-Studies collected the presence/absence data required for DOMs from a breadth of different sources and detection methods (@fig-StudyDetails E).
-In our sample, 79% of studies used human observational data, 11% used live-trapping methods, and 14% used detections from camera traps.
-10% of articles also use citizen-science data to build detection histories, including @zuckerberg2011's model using bird-feeder data and @warrier2020's using citizen reports of tiger signs.
-Using citizen-science data may require additional consideration of assumptions (particularly false-positives), as discussed in greater detail by @cruickshank2019.
-Within these broad categories of detection method there is additional diversity, with several studies using unique survey methods to determine the presence of a species at a location.
-For example, in the only application of a DOM for a marine species in our sample, @pendleton2022 used aerial transects broken up into grid cells to observe whale occupancy; in another unique implementation, @marescot2020 fit a unique 'multi-species' model treating poachers as a taxon and using ranger reports to create detection histories.
-
-Notably, many of these studies use data which were not originally collected in a robust design framework for occupancy modelling.
-In these articles, authors formatted their data into a hierarchical format post-hoc using a variety of methods.
-Some defined primary seasons as a discrete time interval, treating all surveys occurring within the season as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate.
-These manipulations permit authors to use data which predate the DOM, as in a study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].
-
-Additionally, not all articles rely on a single source of detection data; some integrate multiple sources of data to maximise sample size.
-In one example, @warrier2020 modelled tiger occupancy in India using three sources of detections, including camera traps, sign surveys, and citizen science reports.
-These integrated detection method models do require additional care and consideration; users must ensure that different detection methods represent comparable spatial and temporal scales, and that any variation in perceptibility between detection methods is accounted for [@pitman2017].
-A special case exists when different detection methods are used where one has the potential for false positive detections; e.g., where less-certain citizen science detections are combined with certain detections from field surveys.
-In this context, the "certain" detections are used to help account for false-positive detection probability, as in @miller2011's study integrating GPS collars and hunter reports to estimate wolf occupancy in Montana.
-
-The flexibility in the data used for DOMs, including the model structure and the scale of observations, is not amenable to a one-size-fits-all definition of occupancy.
-Users of DOMs must carefully consider precisely what they are modelling and address questions on the scale represented by their model [@chave2013].
-The interpretation of what 'occupancy' means and the factors which drive it may differ depending on the scale of what is considered a site, and how that relates to the ecology of a species.
-This is also true for the temporal scale of occupancy: whether a site is occupied within a week or within a year leads to vastly different conceptions of occupancy.
-This idea is particularly relevant in cases where the selection of season length *is* to some extent arbitrary, as with camera-trap or bioacoustic data where continuous recordings can be broken down into distinct 'seasons' of any length.
-The DOM may be readily applied to these data types, and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data [@balantic2019].
-This is already apparent in recent camera trap studies which collapsed their detections into seasons of just a few days, far shorter than is realistic with conventional survey methods [@kleiven2020; @mölle2022].
-While these studies provide interesting insights of occupancy at extremely fine temporal scale, further research could help provide general guidance on determining appropriate season and survey durations with respect to research questions and the closure assumption.
-
-# Practices in implementation and model building
-
-Building dynamic occupancy models can be a challenging process requiring careful consideration of which environmental factors to incorporate to adequately represent occupancy and detection in complex natural systems.
-In this review, we recorded all covariates considered for each model in our sample, including those not used in final models.
-These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of studies considering each type of covariate in their models, the means by which covariate data are collected, and how covariates responses are represented in DOMs.
-We further delineate their use in the model by classifying covariates into two groups: environmental covariates, representing plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environment.
-Our findings indicate that DOM users have incorporated a wide diversity of covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample.
-Many models incorporate covariates representing aspects of site geometry and connectivity (35% of studies).
-Often these are included as simple covariates, as in @duggan2011's models using landcsape connectivety metrics as covariates on colonisation and extinction.
-Alternatively, more complex parameterisations explicitly model colonisation or extinction as a spatial process, blending attributes of DOMs and stochastic patch occupancy models [@broms2016; @risk2011].
-Several studies also include biotic interactions with other species as covariates, particularly those focused on taxa threatened by invasive species.
-This is observed frequently in the Spotted owl literature, where Barred Owl presence is often considered as a covariate driving Spotted owl occupancy dynamics [@olson2005; ; @sovern2014; @dugger2016].
-The use of these types of covariates effectively incorporate species interactions in DOMs without requiring the use of more complex explicitly multi-species models.
+Covariates.
 
@@
 
-Covariate data for studies in our sample was either collected directly by researchers (an average of 30% of environmental covariates per study), or derived from pre-existing remotely sensed datasets (70% of environmental covariates); this of course varies depending on the category of covariate (@tbl-covariates).
-Directly collected data can often represent finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely but can often be more proximal drivers of occupancy.
-These species-specific covariates do come with trade-offs, as they can be expensive to collect and often preclude projection to locations where these data are unavailable.
-For studies interested in making such projections, remotely-sensed covariates are generally more feasible despite their generally more distal nature [@austin2002].
-An average of 43% percent of environmental factors and 94% of structural factors included in reviewed studies were dynamic covariates which varied through time — this again varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static (@tbl-covariates).
-
-In the standard DOM, covariates for each parameter are most commonly incorporated via a logistic regression (i.e., a linear regression through a logit link function) [@mackenzie2017].
-Statistical relationships between model parameters and covariates (e.g., between initial occupancy and its environmental covariates) are represented as linear terms unless more complexity is specified.
-Of course, many ecological relationships are non-linear and require more complex forms to be realistically represented in a model.
-@austin2007 discusses the importance of modelling ecologically realistic responses to covariates, advocating for careful consideration of the most appropriate statistical form for hypothesised relationships.
-Non-linear responses can be easily accommodated in DOMs by using polynomial transformations and interactions between covariates.
-Despite this, in our sample only 35% of articles tested one or more non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear terms (@tbl-covariates).
-Interactions between covariates were similarly rare, with only 24% of studies considering at least one interaction between terms (@tbl-covariates).
-The relatively low emphasis on more complex non-linear responses contrasts with other popular methods for modelling species occupancy.
-Many common approaches for SDMs, such as MAXENT and Boosted Regression Trees, permit considerable flexibility in the shape of their covariate response curves and use of interactions where supported by the data [@merow2013; @elith2008].
-This emphasis on more complex responses in SDMs may be due to their frequent application across relatively large geographic extents which might encompass the full species niche, where environmental relationships may be expected to be non-linear.
-However, as previously indicated DOMs have also been implemented across large spatial extents where the same assumptions should exist and similar responses might be expected.
-Similar concerns exist for the relatively low level of covariate interactions, given that these relationships are commonly expected based on ecological theory and that their exclusion can negatively impact model performance [@guisan2006].
-
-In addition to the types of covariates considered for modelling we also recorded the size of the candidate covariate pool in each reviewed study, tallying the number of environmental and structural covariates which were available for use on each parameter ( @fig-covariates).
-This is an area of considerable variation in modelling practices — the number of covariates considered for each parameter ranges from 0 (effectively modelling the parameter as a constant) to over 40 candidates on a single parameter.
+More covariates.
 
@@
 
-The median number of covariates varies strongly by parameter, with transition probabilities (colonisation and extinction) more likely to consider a broader range of environmental covariates compared to initial occupancy and detection.
-The lack of any covariates considered for initial occupancy in many studies (37% of studies) is particularly notable — unless a study is conducted at very small extents or study sites are truly uniform in their suitability, one would expect some amount of non-random variation in occupancy probability across any study system that will not be captured when representing initial occupancy as a constant.
-Omission of the factors that drive occupancy is a known source of bias in static SDMs [@barry2006], which the initial occupancy component of the DOM conceptually resembles.
-Furthermore, any bias in occupancy estimation in the first time-step will perpetuate into future seasons due to the DOM's Markovian nature with implications for the reliability of model outputs.
-Detection probability is also typically represented with low numbers of environmental covariates (@tbl-covariates).
-Recall that detectability is dependent on not only observation factors, but also by drivers of abundance and species use; the low number of candidate covariates here relative to colonisation and extinction raises additional questions on whether this variation is always appropriately captured in models.
-
-To this point in our review, we have discussed only the covariates which were *considered* for inclusion in DOMs without regard for which terms were actually *included* in the final models used by authors to make their inference.
-Identifying the 'best' model from a range of possible candidates can be a challenging process, especially for hierarchical models like DOMs [@doherty2012].
-Consider that the quantity of candidate models involving up to *n* covariates on *p* parameters can be given by $(2^n)^p$: where a simple linear regression may have up to 64 possible combinations of 6 covariates, a 4-parameter dynamic occupancy model would have 16 million combinations of these same covariates.
-This explosion of candidate models generally precludes exhaustive comparisons of possible models and requires some reduction of the models tested.
-Models in our sample include both Frequentist (n = 76) and Bayesian (n = 24) frameworks, which differ considerably in their manner of implementation; some articles in our sample included multiple distinct modelling workflows, resulting in a higher total number of models than articles.
-A summary of modelling practices in DOMs is given in @tbl-modelling.
+Modelling summary.
 
@@
 
-The largest differences between the frequentist and Bayesian models in our sample lie in their approaches to model selection.
-Where 95% of frequentist models perform some manner of model selection to determine covariate inclusions for their models, only 33% of Bayesian models do so, with the majority instead fitting a single model defined *a priori*.
-Methods used for model selection vary considerably both within and between the two modelling frameworks (@tbl-modelling).
-For frequentist models, the most popular and conventional approach to model selection (45% of models) involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score.
-The next most popular method in frequentist studies is to use procedural model selection methods (37% of models), where the structure for each model parameter is fit in sequence.
-For example, this protocol might first identify the best structure for detection probability while holding the other parameters constant, before moving on to initial occupancy and so on until all parameters are fixed.
-The remainder of frequentist studies (8%) use a variety of approaches, such as fitting simpler models like single season occupancy models to identify the most informative terms to use in a DOM.
-Across all frequentist implementations, 47% of articles summarise a final subset of models using multi-model inference and model-averaging with AIC weights [@burnham2004].
-
-Those Bayesian models which *do* perform model selection take various approaches, with largely idiosyncratic methods across these studies.
-While direct comparison of model fit is rare in Bayesian methods, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data.
-Another approach used by @cook2022 fit a global model including all covariates before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the model.
-@ahumada2013 takes a hybrid approach, in which model selection is conducted by a procedural method in the frequentist framework before refitting the best structure as a Bayesian model.
-
-Limited research has been conducted on the advantages of different methods for covariate selection in DOMs, and there is unlikely to be a one-size-fits-all approach which will be appropriate for all possible use-cases.
-However, it is important to consider the implications of the different model selection approaches in common usage, and research from the SDM literature highlights how covariate selection can influence our interpretation of model outputs [@brodie2020].
-Within the occupancy modelling literature, @stewart2023's article on covariate selection in single-season occupancy models discusses important attributes of the information-criteria approaches widely used in frequentist models in our review, noting that these approaches can lead to inaccurate coefficient estimates that may influence model inference.
-@morin2020's work raises other concerns on procedural model selection methods, demonstrating how the fine details of modelling protocols can determine whether the best-fitting possible model is identified and which covariates appear in final models.
-The literature on performance of Bayesian model selection methods is more sparse, although @hooten2015's guide to Bayesian model selection in ecology is a valuable resource for possible methods of fitting those models.
-
-Regardless of covariate selection protocol, 'the selection of a best model does not guarantee the selection of a good model' [@mackenzie2004].
-Assessment of the performance of DOMs generally requires additional steps, and the best ways to achieve this are not yet clear.
-There is no broadly-accepted goodness-of-fit test for dynamic occupancy models, although @mackenzie2004's approach for single-season occupancy models using a parametric bootstrap which can been extended to DOMs; this test is implemented in the *AICcModAvg* and *unmarked* R packages.
-@kéry2021 describe the test and present an alternative based on separately assessing fit to static and dynamic components of the model.
-In Bayesian implementations, posterior predictive checks offer means to assess model fit [@gelman2014].
-@broms2016 discusses further possibilities for model evaluation in the Bayesian context focusing on single-season multi-species occupancy models, and extensions of their approach may also be applicable to the DOM.
-As with other hierarchical models, model evaluation for DOMs can be difficult and somewhat uncertain compared to other model types, as the primary response variable of interest (species occupancy) is a latent variable where the true state is generally not known.
-Predictive performance evaluation is thus typically based on *observed* occupancy data, where a DOM is used to simulate detection histories to be compared with field survey results.
-
-Perhaps as a result of these difficulties, assessment of model fit and model performance was rare amongst the articles in our sample.
-Only 18% of studies tested for goodness-of-fit, and just 7% calculated predictive performance.
-These rates are considerably lower than those reported for SDMs; for which closer to 50% of articles were found to test both fit and performances [@araújo2019].
-
-# Modelling objectives
-
-Studies in our sample were classified according to their objectives as stated by authors to explore the use-cases which the DOM has been applied to (@fig-objectives).
-These objective categories are non-exclusive, with many studies falling into multiple categories.
-Across the study period the DOM has been frequently employed to estimate trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions.
-Within each of these broad categories lies even more variation in objectives, emphasising the DOM's flexibility as a tool for making ecological inference in diverse contexts.
-
-36% of articles used the DOM to monitor trends in occupancy state through time ("Estimate trends" in @fig-objectives A), both for single species of high conservation interest and for broader community assemblages of species [@ahumada2013; @scott2015].
-These studies offer valuable insights on the state of species across landscapes and through time, demonstrating the DOM's suitability for monitoring oriented projects; including those where multiple species need to be assessed at once.
-Other articles were more focused on examining the factors which influence species occupancy, either testing pre-specified hypothetical relationships (44% of articles) or taking a broader tack to identify drivers of occupancy without preconception (35%).
-The distinction between the two is important, as it guides which covariates may be considered and how model selection may be used to fit models.
-Many of these studies target core conservation priorities for their focal species, like @olson2005's early DOM assessing the influence of barred owls on threatened spotted owl.
-Explorations of these pivotal relationships are important for guiding management action, and may also be used to test the effectiveness of these actions [@miller2023].
-In scenarios where less is known about species habitat preference, DOMs may be used to examine the influence of a wider variety of factors as in @huber2017's study testing the relative influence of dozens of habitat covariates on Wood warbler occupancy.
-In recent years, increasing numbers of articles have used DOMs to generate predictions of species occupancy; this a use-case for which DOMs show considerable promise [@kéry2013; @briscoe2021].
-These predictions often have strong utility for conservation management.
-For example, @mcgowan2020 projects future occupancy for wetland birds under alternative management scenarios, and @pollentier2021 generates distribution maps resembling those made with SDMs.
+Objectives.
 
@@
 
-Study objectives strongly inform approaches to model building, particularly with respect to covariate selection.
-@fig-objectives B presents the quantity of covariates considered for models applied to each of our objective categories, with apparent differences between objectives.
-Not unexpectedly, articles which focused on describing new methods for DOMs had fewer covariates than those studies focused on more applied objectives.
-However, differences persist between those articles observing trends, identifying trends in occupancy, testing relationships, and making predictions.
+More objectives.
 
-# Synthesis and key priorities
+# Discussion
 
-Approaches to building any type of model will necessarily depend on the possibilities of the data at hand and on the priorities of the model-builder.
-This precludes any prescription of the 'best' way to build a model, however, there are still important discussions to be had on decisions made in the modelling process.
-One aspect of fitting DOMs meriting broader discussion centres on 'model complexity,' and how much must be incorporated into models to reliably model species occupancy under different contexts and use cases.
-Complexity is a broad term encompassing many aspects of a model [@merow2014], and opinions on simplicity versus complexity in ecological models are diverse.
-Where some advocate for the simplest possible models, arguing that they are most generalisable; others insist that overly-simple models cannot adequately represent the most important drivers in a system [@evans2013; @lonergan2014].
-By their nature the DOM is somewhat more complex than simpler models for studying occupancy due to their hierarchical structure, which is necessary to control for detectability and to capture occupancy dynamics.
-Within this structure, however, further complexity is to some degree up to the modeller: one can choose how many covariates to consider for inclusion on the various parameters, and how to represent the nature of the relationship between those covariates and parameters.
-Research from SDMs indicates that allowing for more complex relationships can improve model performance in predicting occupancy [@valavi2023], an increasingly popular use-case for the DOM.
-Within the DOM framework, there are promising developments on that front: @joseph2020 presents a novel neural-network occupancy model which allows for exponentially higher levels of complexity and may offer improved performance for prediction-oriented studies.
+Discussion.
 
-Covariate selection seems to be a particularly important area for further investigations into building DOMs.
-In our review, we see little consensus around which approaches are most applicable for any given use case, and existing work on covariate selection in DOMs raises concerns on whether common methods always produce the most suitable models [@morin2020; @stewart2023].
-This is true for both frequentist and Bayesian implementations, and comparative research on covariate selection under both frameworks may help to inform model users on which method is most appropriate for their use-cases.
-In a similar vein, the low number of articles in our review that conducted model evaluation or assessed model fit raises different questions.
-While the appropriate method of model evaluation may depend on data availability and research objectives, assessing models by some method is generally important to understand how reliable model outputs may be [@araújo2005; @guisan2005].
-Existing uncertainties around whether current methods are suitable for the task likely discourage users from calculating these metrics, and further research is needed to establish trusted practices for assessing the quality of DOMs.
+# Conclusions
 
-These priorities are particularly important given the frequent applied objectives of DOM users, tackling challenges which include assessing critically endangered species, guiding public health management of disease vectors, and tracking rapidly developing biological invasions [@carvalho2023; @moreira2016; @wood2020].
-The DOM is well suited to these situations, and it is to be expected that as these types of applications are more commonly attempted understanding the sensitivity of model outputs to decisions made in the model fitting process becomes increasingly important.
-In the two decades since the publication of @mackenzie2003 the dynamic occupancy model has become a widely used tool for ecological inference, with numerous extensions to the modelling framework further broadening the scope of questions and use-cases for which it may be applied.
-Given their increasing popularity, further research and guidelines around issues of model building may help to make the DOM more accessible to newcomers and support confidence in model interpretation.
-Parallels to existing work in the SDM literature on guidelines for reporting and modelling standards could be valuable contributions in achieving these aims and provide useful examples in developing methods and guidelines specifically focused on DOMs [@zurell2020; @araújo2019].
+Conclusions.

### 2025-01-23 Manuscript_Initial.qmd
preserve the initial version and begin scaffold for new one

--- 2024-09-20 _manuscript/Manuscript.qmd
@@
 abstract: | 
-  Describing patterns of species occupancy across landscapes and throughout time is a key objective of much ecological research.
-Nonetheless, reliably estimating occupancy can be a challenge, particularly when common issues such as imperfect detection and shifting distributions are at play.
-The dynamic, multi-season occupancy model (DOM) offers a framework for occupancy estimation which explicitly accounts for observation error, while also incorporating additional mechanistic process by explicitly estimating colonisation and extinction rates.
-In the two decades since its introduction, the DOM has become a popular tool for describing patterns in occupancy, exploring the environmental drivers of species occurence, and, more recently, for predicting species occupancy to new locations and to future conditions.
-
-  In this review we examine how ecologists have applied the DOM in its first two decades of use.
-Based on a sample of articles fitting DOMs to field ecological data, we explore the systems in which these models have been used, how data were collected for them, and how authors have built their models to achieve their objectives.
-Our findings indicate that the DOM is an particularly flexible tool readily applied to a range of diverse taxa from around the globe, capable of being used to estimate occupancy at study sites ranging from local to continental scales over time periods from months to decades.
-The DOM framework is also amenable to extension, further broadening their utility to address a wide range of ecological questions.
-
-  A key aim of this review is to explore how users of DOMs have incorporated covariates into their models to describe variation in occupancy through space and time.
-Broadly speaking, model complexity in DOMs tends to be low, with relatively few covariates considered for inclusion and most relationships represented as simple linear terms.
-Considerable variation also exists in the procedures used for covariate selection, and limited research has been conducted on how these choices may influence model performance and inference.
-Furthermore, only a fraction of articles appear to evaluate models (either by goodness-of-fit testing or prediction validation).
-Likewise, little guidance exists on how to approach this task in DOMs.
+ Understanding the patterns by which species occur across landscapes and throughout time is a long-standing objective of ecological research which has inspired the development of numerous quantitative modelling approaches for the task.
+However, estimating occupancy can be a challenge, particularly where issues like imperfect detection and shifting distributions exist.
+Dynamic occupancy models (DOMs) offer a framework for occupancy estimation which explicitly accounts for observation error while incorporating additional mechanistic process by estimating colonisation and local extinction.
+In light of increasing interest in more process-explicit models for understanding species occurrence, here we examine how DOMs have been applied to field ecological data in the two decades since their introduction.
+Following a general introduction to the model, we present the results of a systematic review exploring the research contexts where DOMs have been applied and interrogating how authors have built their models, with particular emphasis on how covariates are incorporated to describe variation in occupancy dynamics.
+Our findings indicate that DOMs are a flexible tool readily applied to diverse study systems and data types, with their usage expanding in recent years as more studies apply them to make spatial and temporal prediction of species occupancy.
+DOMs are also amenable to extension, further broadening their utility.
+Despite their flexibility, model complexity in DOMs tends to be low; most studies consider relatively few covariates, most of which are represented as simple linear relationships.
+Approaches to covariate selection also vary considerably, and there remains little research on how these choices may influence model performance.
+Furthermore, only a fraction of articles report evaluating models and little guidance exists on how to approach this task in DOMs.
 These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.
 
-bibliography: references.bib
-csl: apa.csl
+format:
+  html:
+    theme: simplex
+
+embed-resources: true
+title-block-banner: true
 ---

### 2025-01-23 Manuscript.qmd
Continued progress on the introduction section, added box for model form

@@
     theme: simplex
+  docx: default
 
@@
 
-Background Information section
+Ecologists are often asked to describe, explain, and predict where and when species occur on the landscape.
+Research on patterns of species occupancy dates back to the origins of the field, and continues to a priority in quantitative ecological research [@humboldt1849].
+Estimates of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, while identifying potential drivers of occurrence can help inform potential management actions [@mackenzie2013].
+Robust knowledge of the occupancy patterns of a species can also help us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios [@kéry2013].
 
-#### Model form and assumptions
+While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate.
+The need to understand and quantitatively describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models commonly applied to study meta-population dynamics [@gutiérrez-arellano2024], and species distribution models (SDMs) widely used to explore species occurrence at larger scales [@franklin2010].
+However, several factors can make occupancy difficult to estimate.
+For instance, simple presence/absence observations can be biased when species are detected imperfectly – this is often the case in wildlife field data, where it can be impossible to determine from a single observation whether a location is truly occupied or whether the species occurs but was not detected [@gu2004; @lahoz-monfort2014].
+Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014].
+Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species’ occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010].
+These conditions often occur during biological invasions and climate change driven range shifts, each of which are conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].
 
-Model form.
+The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017a] .
+Drawing on principles from the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy.
+Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time.
+For newcomers to DOMs, in [Box @nte-Primer] we present a simple introduction to the basic model form as described in @mackenzie2003.
 
-![The form of the dynamic occupancy model as described by @mackenzie2003.
-The ecological process sub-model describes changes in occupancy over time, while the observational process sub-model describes detectability.](Figures/ModelForm.png){#fig-modelform}
+This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes driving occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic or abundance data required by more process-explicit models [@briscoe2019].
+These features make the DOM an important tool, with uses including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@briscoe2021; @kéry2013].
 
-Survey design.
+::: {#nte-Primer .callout-note icon="false"}
+## An introduction to dynamic occupancy models
 
-![The sampling design of the standard dynamic occupancy model.
-During seasons (also called primary occasions) sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons.
-Within each season, multiple observations (also called surveys or secondary occasions) are conducted to record the observed presence or absence of the species at each site.
-These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded.
-Note that it is not necessary for each site or season to have the same number of observations, and that missing data can be accommodated.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}
+Dynamic occupancy models.
 
-Assumptions.
+![](images/clipboard-1253177906.png){width="11cm"}
+
+![](images/clipboard-3209412869.png){width="11cm"}
+:::
+
+DOMs make several key assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.
+
+I.
+**False positive detections do not occur**.
+While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods.
+@mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need to consider how certain detections truly are for any given study system.
+Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019].
+Where this assumption can not reasonably be met, model extensions designed to account for false positive error should be considered [@royle2006; @miller2011; @miller2015].
+
+II.
+**Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion in terms of the bias introduced when it is violated [@rota2009; @otto2013].
+Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and it can be difficult to distinguish local extinction from temporary emigration [@valente2017].
+Model extensions to relax the closure assumption have been developed, including approaches using staggered arrival and departure periods between sites [@kendall2013].
+A more pertinent approach for most studies, however, is careful consideration of the appropriate definition of occupancy under the survey design used [@steenweg2018].
+
+III.
+**Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species occurrence, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system.
+Generally, this is achieved by allowing model parameters (ψ~1~, γ, ε, and ρ) to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters.
+An important element of this assumption is that the likelihood of detection of a species can depend not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017a].
+While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates [@barry2006].
+Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known about how this latent heterogeneity can influence model performance.
+
+Since its original description in @mackenzie2003, the DOM has been further developed with numerous model extensions and alternative formulations including aforementioned implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010].
+For a comprehensive discussion of common extensions and their applications see @bailey2014 and @guillera-arroita2017a, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.
 
@@
     style = cell_text(style = "italic")
-  ) 
-#|>
- # as_raw_html()
+  ) |>
+  as_raw_html()
 ```

### 2025-01-28 Manuscript.qmd
Updated introduction box

@@
 
-Dynamic occupancy models.
+We give a minimal introduction to the canonical dynamic occupancy model as defined in @mackenzie2003.
+DOMs are hierarchical models which link two sub-models.
+The first of these describes the ecological process of occupancy, in which sites move between occupied and unoccupied states over time.
+Occupancy is unobservable, as we cannot always know the true occupancy state.
+The second sub-model describes the observational process of detection, allowing us to correct for imperfect detection.
 
-![](images/clipboard-1253177906.png){width="11cm"}
+#### Data collection
 
-![](images/clipboard-3209412869.png){width="11cm"}
+DOMs require hierarchical data to separate the occupancy and detection processes.
+Occupancy is measured in discrete, time-bound intervals termed *primary occasions*.
+Within each primary occasion *t*, multiple observations are recorded at the same sites.
+Importantly, all sites or primary occasions do not require the same number of observations and some number of missing observations are permissible.
+While secondary occasions are typically repeat observations, other options including same-visit replicates and spatial replicates are also possible (see @guillera-arroita2010).
+
+![](images/clipboard-2772050481.png){width="600" fig-align="center"}
+
+#### Site occupancy sub-model
+
+Occupancy is described during discrete, time-bound intervals termed *primary occasions.* In each primary occasion *t,* each site *i* may exist in either an occupied (z~i,t~ = 1) or unoccupied state (z~i,t~ = 0).
+In the first primary occasion, occupancy is determined as a Bernoulli trial with initial occupancy probability 𝛙~1~, such that $z_{i,t=1} \sim Bernoulli(\psi_1)$.
+In subsequent primary occasions, site occupancy is determined by the site's state in the previous primary occasion and probability of colonisation 𝛄 and extinction 𝛆, such that $z_{i,t} \sim Bernoulli(z_{i,t-1}(1-\epsilon)+(1-z_{i,t-1})*\gamma)$.
+While we present each of 𝛙~1~, 𝛄, and 𝛆 as constants for simplicity, they may be estimated with respect to covariates which are typically included via a logit link.
+
+![](images/clipboard-3349624601.png){width="600" fig-align="center"}
+
+#### Detection sub-model
+
+During any given survey *j,* observers will either observe (y~i,t,j~ = 1) or not observe (y~i,t,j~ = 0) the target species.
+In each survey, the probability of observing the target species at a site is given as a Bernoulli trial with detection probability *p,* conditional on the site being occupied, such that $y_{i,t,j} \sim Bernoulli(z_{i,t}*p)$.
+This formulation assumes that no false positive detections exist; that is, observations are never recorded at truly unoccupied sites.
+As with the occupancy probabilities, *p* may also be estimated using covariates via a logit link.
+
+![](images/clipboard-411045638.png){width="324" fig-align="center"}
+
+#### Fitting DOMs
+
+DOMs are most often fit with maximum likelihood estimation, which identifies the values of 𝛙~1~, 𝛄, 𝛆, and *p* under which the observed data **y** are most probable.
+Common tools for fitting DOMs include `PRESENCE`, `MARK`, and the r package `unmarked`.
+DOMs may also be fit in a Bayesian framework, with available packages including `ubms`.
 :::
 
-DOMs make several key assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.
+DOMs make several key assumptions requiring careful consideration by model users, which we outline here as our reviews related aspects of model building.
 

### 2025-02-03 Analysis.qmd
Completed first-pass revisions for Ecography

@@
        width = 8, height = 4)
+```
+
+## Authorship
+
+```{r Authorship_categories}
+sheet_Objectives |>
+  separate_longer_delim(`Authorship categories`, delim = ", ") |>
+  mutate(n_Overall = length(unique(`Review ID`))) |>
+  summarise(PropAuthor = (n()/mean(n_Overall))*100,
+            .by = `Authorship categories`)
+
+sheet_Objectives |>
+  separate_longer_delim(`Authorship categories`, delim = ", ") |>
+  summarise(nPer = n(),
+            .by = `Review ID`) |>
+  summarise(AffilCount = n(),
+            .by = nPer) (92-34)/92
+
+sheet_ArticleData |>
+  separate_longer_delim(`Countries of affiliation`, delim = ", ") |>
+  summarise(nCountry = n(),
+            .by = `Countries of affiliation`) 
 ```
@@
 
+## Covariate summaries
+
+#### Quantity
+
+```{r Covariate_Quantity}
+covariateFreq <- sheet_Covariates |>
+  separate_longer_delim(Covariates, delim = ", ") |>
+  separate_wider_delim(Covariates, delim = "_", 
+                       names = c("Covariate", "Variation", 
+                                 "Observation", "Relation"),
+                       too_few = "align_start") |>
+  separate_wider_delim(Covariate, delim = "-",
+                       names = c("Category", "Covariate"),
+                       too_few = "align_start") |>
+  # Merge some covariate categories
+  mutate(Covariate = case_when(Covariate == "OTHER" & Category == "E" ~ "OTHER-E",
+                               Covariate == "OTHER" & Category == "S" ~ "OTHER-S",
+                               Covariate %in% c("HABT", "HABA", "LACO") ~ "Habitat",
+                               Covariate %in% c("CLWE", "DISA", "DIST") ~ "Climate",
+                               Covariate %in% c("SPAT", "GEOM", "CONN") ~ "Spatial",
+                               Covariate %in% c("ANTH", "AGRI", "PROT") ~ "Human",
+                               .default = Covariate)) |>
+  # Combine Review ID & Model ID, and calculate the number of studies and number of studies with each parameter
+  mutate(studyID = paste(`Review ID`, `Model ID`), sep = "_") |>
+  mutate(n_overall = length(unique(studyID))) |>
+  select(studyID, Covariate, Category, n_overall) |>
+  filter(!is.na(Covariate)) %>%
+  rbind(mutate(., Covariate = Category)) %>%
+  rbind(mutate(., Covariate = "ANY", Category = "ANY")) |>
+  distinct() |>
+  summarise(PctCategory = (n()/mean(n_overall))*100,
+            .by = c(Covariate, Category)) |>
+  mutate(Included = TRUE) %>%
+  rbind(mutate(., PctCategory = 100-PctCategory, Included = FALSE)) |>
+  mutate(Covariate = case_when(Covariate == "Habitat" ~ "Habitat",
+                               Covariate == "Climate" ~ "Climate",
+                               Covariate == "Spatial" ~ "Spatial",
+                               Covariate == "Human" ~ "Human",
+                               Covariate == "HYDR" ~ "Water",
+                               Covariate == "BIOT" ~ "Biotic",
+                               Covariate == "TOPO" ~ "Topography",
+                               Covariate == "PHEN" ~ "Timing",
+                               Covariate == "SEAS" ~ "Primary",
+                               Covariate == "SURV" ~ "Secondary",
+                               Covariate == "OBSE" ~ "Observation",
+                               Covariate == "SITE" ~ "Site",
+                               Covariate == "SPEC" ~ "Species",
+                               Covariate == "ANY" ~ "Any",
+                               Covariate == "E" ~ "Any\nEnv.",
+                               Covariate == "S" ~ "Any\nStruct."),
+         .default = Covariate) |>
+  filter(!is.na(Covariate)) |>
+  mutate(Covariate = fct(Covariate,
+                         levels = c("Any", "Any\nEnv.",
+                                    "Habitat", "Spatial", "Climate",
+                                    "Timing", "Human", "Topography",
+                                    "Biotic", "Water",
+                                    "Any\nStruct.", "Primary", "Observation",
+                                    "Secondary", "Site", "Species"))) |>
+  mutate(Category = case_when(Category == "ANY" ~ "Any",
+                              Category == "E" ~ "Environmental",
+                              Category == "S" ~ "Structural")) |>
+
+  ggplot(aes(x = Covariate, y = PctCategory, fill = Included)) +
+  geom_col(colour = "white", 
+           linewidth = 1, width = 1) +
+  scale_fill_manual(limits = c(TRUE, FALSE), 
+                    values = c("goldenrod", "gray50")) +
+  scale_y_continuous(expand = c(0,0)) +
+  facet_grid(cols = vars(Category), scales = "free_x", space = "free_x") +
+  labs(y = "Percent\nof models", x = "") +
+  theme(legend.position = "none",
+        plot.background = element_blank(),
+        panel.grid = element_blank(),
+        axis.text.x = element_text(family = "Helvetica", 
+                                   angle = 90, vjust = 0.5, hjust = 1),
+        strip.text.x = element_text(family = "Helvetica"))
+covariateFreq
+```
+
+#### Dynamism
+
+```{r Covariate_Dynamism}
+covariateDyn <- sheet_Covariates |>
+  separate_longer_delim(Covariates, delim = ", ") |>
+  separate_wider_delim(Covariates, delim = "_", 
+                       names = c("Covariate", "Variation", 
+                                 "Observation", "Relation"),
+                       too_few = "align_start") |>
+  separate_wider_delim(Covariate, delim = "-",
+                       names = c("Category", "Covariate"),
+                       too_few = "align_start") |>
+  # Merge some covariate categories
+  mutate(Covariate = case_when(Covariate == "OTHER" & Category == "E" ~ "OTHER-E",
+                               Covariate == "OTHER" & Category == "S" ~ "OTHER-S",
+                               Covariate %in% c("HABT", "HABA", "LACO") ~ "Habitat",
+                               Covariate %in% c("CLWE", "DISA", "DIST") ~ "Climate",
+                               Covariate %in% c("SPAT", "GEOM", "CONN") ~ "Spatial",
+                               Covariate %in% c("ANTH", "AGRI", "PROT") ~ "Human",
+                               .default = Covariate)) |>
+  # Combine Review ID & Model ID, and calculate the number of studies and number of studies with each parameter
+  mutate(studyID = paste(`Review ID`, `Model ID`), sep = "_") |>
+  mutate(n_overall = length(unique(studyID))) |>
+  select(studyID, `Parameter type`, Covariate, Category, Variation, n_overall) |>
+  filter(!is.na(Covariate)) |>
+  filter(`Parameter type` != "Initial occupancy") %>%
+  rbind(rbind(mutate(., Covariate = Category),
+              mutate(., Covariate = "ANY", Category = "ANY"))) |>
+  mutate(n_Covariate = n(),
+         .by = Covariate) |>
+  mutate(Dynamism = case_when(Variation %in% c("SISE", "SU", "SE") ~ 1,
+                              .default = 0)) |>
+  summarise(PctDyn = (sum(Dynamism)/mean(n_Covariate))*100,
+            .by = c(Covariate, Category)) |>
+  mutate(Dyn = TRUE) %>%
+  rbind(mutate(., PctDyn = 100-PctDyn, Dyn = FALSE)) |>
+  mutate(Covariate = case_when(Covariate == "Habitat" ~ "Habitat",
+                               Covariate == "Climate" ~ "Climate",
+                               Covariate == "Spatial" ~ "Spatial",
+                               Covariate == "Human" ~ "Human",
+                               Covariate == "HYDR" ~ "Water",
+                               Covariate == "BIOT" ~ "Biotic",
+                               Covariate == "TOPO" ~ "Topography",
+                               Covariate == "PHEN" ~ "Timing",
+                               Covariate == "SEAS" ~ "Primary",
+                               Covariate == "SURV" ~ "Secondary",
+                               Covariate == "OBSE" ~ "Observation",
+                               Covariate == "SITE" ~ "Site",
+                               Covariate == "SPEC" ~ "Species",
+                               Covariate == "ANY" ~ "Any",
+                               Covariate == "E" ~ "Any\nEnv.",
+                               Covariate == "S" ~ "Any\nStruct."),
+         .default = Covariate) |>
+  filter(!is.na(Covariate)) |>
+  mutate(Covariate = fct(Covariate,
+                         levels = c("Any", "Any\nEnv.",
+                                    "Habitat", "Spatial", "Climate",
+                                    "Timing", "Human", "Topography",
+                                    "Biotic", "Water",
+                                    "Any\nStruct.", "Primary", "Observation",
+                                    "Secondary", "Site", "Species"))) |>
+  mutate(Category = case_when(Category == "ANY" ~ "Any",
+                              Category == "E" ~ "Environmental",
+                              Category == "S" ~ "Structural")) |>
+
+  ggplot(aes(x = Covariate, y = PctDyn, fill = Dyn)) +
+  geom_col(colour = "white", 
+           linewidth = 1, width = 1) +
+  scale_fill_manual(limits = c(TRUE, FALSE), 
+                    values = c("aquamarine3", "gray50")) +
+  scale_y_continuous(expand = c(0,0)) +
+  facet_grid(cols = vars(Category), scales = "free_x", space = "free_x") +
+  labs(y = "Percent\ndynamic", x = "") +
+  theme(legend.position = "none",
+        plot.background = element_blank(),
+        panel.grid = element_blank(),
+        axis.text.x = element_text(family = "Helvetica", 
+                                   angle = 90, vjust = 0.5, hjust = 1),
+        strip.text.x = element_blank(), strip.background = element_blank())
+covariateDyn
+```
+
+#### Observation
+
+```{r Covariate_Observed}
+covariateObs <- sheet_Covariates |>
+  separate_longer_delim(Covariates, delim = ", ") |>
+  separate_wider_delim(Covariates, delim = "_", 
+                       names = c("Covariate", "Variation", 
+                                 "Observation", "Relation"),
+                       too_few = "align_start") |>
+  separate_wider_delim(Covariate, delim = "-",
+                       names = c("Category", "Covariate"),
+                       too_few = "align_start") |>
+  # Merge some covariate categories
+  mutate(Covariate = case_when(Covariate == "OTHER" & Category == "E" ~ "OTHER-E",
+                               Covariate == "OTHER" & Category == "S" ~ "OTHER-S",
+                               Covariate %in% c("HABT", "HABA", "LACO") ~ "Habitat",
+                               Covariate %in% c("CLWE", "DISA", "DIST") ~ "Climate",
+                               Covariate %in% c("SPAT", "GEOM", "CONN") ~ "Spatial",
+                               Covariate %in% c("ANTH", "AGRI", "PROT") ~ "Human",
+                               .default = Covariate)) |>
+  # Combine Review ID & Model ID, and calculate the number of studies and number of studies with each parameter
+  mutate(studyID = paste(`Review ID`, `Model ID`), sep = "_") |>
+  mutate(n_overall = length(unique(studyID))) |>
+  select(studyID, Covariate, Category, Observation, n_overall) |>
+  filter(!is.na(Covariate)) %>%
+  rbind(rbind(mutate(., Covariate = Category),
+              mutate(., Covariate = "ANY", Category = "ANY"))) |>
+  mutate(n_Covariate = n(),
+         .by = Covariate) |>
+  mutate(Observed = case_when(Observation == "D" ~ 1,
+                              .default = 0)) |>
+  summarise(PctObs = (sum(Observed)/mean(n_Covariate))*100,
+            .by = c(Covariate, Category)) |>
+  mutate(Obs = TRUE) %>%
+  rbind(mutate(., PctObs = 100-PctObs, Obs = FALSE)) |>
+  mutate(Covariate = case_when(Covariate == "Habitat" ~ "Habitat",
+                               Covariate == "Climate" ~ "Climate",
+                               Covariate == "Spatial" ~ "Spatial",
+                               Covariate == "Human" ~ "Human",
+                               Covariate == "HYDR" ~ "Water",
+                               Covariate == "BIOT" ~ "Biotic",
+                               Covariate == "TOPO" ~ "Topography",
+                               Covariate == "PHEN" ~ "Timing",
+                               Covariate == "SEAS" ~ "Primary",
+                               Covariate == "SURV" ~ "Secondary",
+                               Covariate == "OBSE" ~ "Observation",
+                               Covariate == "SITE" ~ "Site",
+                               Covariate == "SPEC" ~ "Species",
+                               Covariate == "ANY" ~ "Any",
+                               Covariate == "E" ~ "Any\nEnv.",
+                               Covariate == "S" ~ "Any\nStruct."),
+         .default = Covariate) |>
+  filter(!is.na(Covariate)) |>
+  mutate(Covariate = fct(Covariate,
+                         levels = c("Any", "Any\nEnv.",
+                                    "Habitat", "Spatial", "Climate",
+                                    "Timing", "Human", "Topography",
+                                    "Biotic", "Water",
+                                    "Any\nStruct.", "Primary", "Observation",
+                                    "Secondary", "Site", "Species"))) |>
+  mutate(Category = case_when(Category == "ANY" ~ "Any",
+                              Category == "E" ~ "Environmental",
+                              Category == "S" ~ "Structural")) |>
+
+  ggplot(aes(x = Covariate, y = PctObs, fill = Obs)) +
+  geom_col(colour = "white", 
+           linewidth = 1, width = 1) +
+  scale_fill_manual(limits = c(TRUE, FALSE), 
+                    values = c("cadetblue3", "gray50")) +
+  scale_y_continuous(expand = c(0,0)) +
+  facet_grid(cols = vars(Category), scales = "free_x", space = "free_x") +
+  labs(y = "Percent directly\nobserved", x = "") +
+  theme(legend.position = "none",
+        plot.background = element_blank(),
+        panel.grid = element_blank(),
+        axis.text.x = element_text(family = "Helvetica", 
+                                   angle = 90, vjust = 0.5, hjust = 1),
+        strip.text.x = element_blank(), strip.background = element_blank())
+covariateObs
+
+```
+
+#### Complexity
+
+```{r Covariate_Complexity}
+covariateComplex <- sheet_Covariates |>
+  separate_longer_delim(Covariates, delim = ", ") |>
+  separate_wider_delim(Covariates, delim = "_", 
+                       names = c("Covariate", "Variation", 
+                                 "Observation", "Relation"),
+                       too_few = "align_start") |>
+  separate_wider_delim(Covariate, delim = "-",
+                       names = c("Category", "Covariate"),
+                       too_few = "align_start") |>
+  # Merge some covariate categories
+  mutate(Covariate = case_when(Covariate == "OTHER" & Category == "E" ~ "OTHER-E",
+                               Covariate == "OTHER" & Category == "S" ~ "OTHER-S",
+                               Covariate %in% c("HABT", "HABA", "LACO") ~ "Habitat",
+                               Covariate %in% c("CLWE", "DISA", "DIST") ~ "Climate",
+                               Covariate %in% c("SPAT", "GEOM", "CONN") ~ "Spatial",
+                               Covariate %in% c("ANTH", "AGRI", "PROT") ~ "Human",
+                               .default = Covariate)) |>
+  # Combine Review ID & Model ID, and calculate the number of studies and number of studies with each parameter
+  mutate(studyID = paste(`Review ID`, `Model ID`), sep = "_") |>
+  mutate(n_overall = length(unique(studyID))) |>
+  select(studyID, Covariate, Category, Relation, n_overall) |>
+  filter(!is.na(Covariate)) %>%
+  rbind(rbind(mutate(., Covariate = Category),
+              mutate(., Covariate = "ANY", Category = "ANY"))) |>
+  mutate(n_Covariate = n(),
+         .by = Covariate) |>
+  mutate(Complex = case_when(Relation %in% c("N", "I", "B") ~ 1,
+                               .default = 0)) |>
+  summarise(PctComplex = (sum(Complex)/mean(n_Covariate))*100,
+            .by = c(Covariate, Category)) |>
+  mutate(Complex = TRUE) %>%
+  rbind(mutate(., PctComplex = 100-PctComplex, Complex = FALSE)) |>
+  mutate(Covariate = case_when(Covariate == "Habitat" ~ "Habitat",
+                               Covariate == "Climate" ~ "Climate",
+                               Covariate == "Spatial" ~ "Spatial",
+                               Covariate == "Human" ~ "Human",
+                               Covariate == "HYDR" ~ "Water",
+                               Covariate == "BIOT" ~ "Biotic",
+                               Covariate == "TOPO" ~ "Topography",
+                               Covariate == "PHEN" ~ "Timing",
+                               Covariate == "SEAS" ~ "Primary",
+                               Covariate == "SURV" ~ "Secondary",
+                               Covariate == "OBSE" ~ "Observation",
+                               Covariate == "SITE" ~ "Site",
+                               Covariate == "SPEC" ~ "Species",
+                               Covariate == "ANY" ~ "Any",
+                               Covariate == "E" ~ "Any\nEnv.",
+                               Covariate == "S" ~ "Any\nStruct."),
+         .default = Covariate) |>
+  filter(!is.na(Covariate)) |>
+  mutate(Covariate = fct(Covariate,
+                         levels = c("Any", "Any\nEnv.",
+                                    "Habitat", "Spatial", "Climate",
+                                    "Timing", "Human", "Topography",
+                                    "Biotic", "Water",
+                                    "Any\nStruct.", "Primary", "Observation",
+                                    "Secondary", "Site", "Species"))) |>
+  mutate(Category = case_when(Category == "ANY" ~ "Any",
+                              Category == "E" ~ "Environmental",
+                              Category == "S" ~ "Structural")) |>
+
+  ggplot(aes(x = Covariate, y = PctComplex, fill = Complex)) +
+  geom_col(colour = "white", 
+           linewidth = 1, width = 1) +
+  scale_fill_manual(limits = c(TRUE, FALSE), 
+                    values = c("plum3", "gray50")) +
+  scale_y_continuous(expand = c(0,0)) +
+  facet_grid(cols = vars(Category), scales = "free_x", space = "free_x") +
+  labs(y = "Percent nonlinear\nor interactive", x = "") +
+  theme(legend.position = "none",
+        plot.background = element_blank(),
+        panel.grid = element_blank(),
+        axis.text.x = element_text(family = "Helvetica", 
+                                   angle = 90, vjust = 0.5, hjust = 1),
+        strip.text.x = element_blank(), strip.background = element_blank())
+covariateComplex
+```
+
+#### Combine
+
+```{r Covariate_Patchwork}
+covariatePatchwork <- covariateFreq / covariateDyn / covariateObs / covariateComplex +
+  plot_layout(axes = "collect") &
+  theme(axis.text = element_text(family = "Helvetica"),
+        axis.title = element_text(family = "Helvetica"))
+covariatePatchwork
+
+ggsave(plot = covariatePatchwork, "Figures/CovariateSummary.png",
+```
+
 ## Objectives
@@
 stat_objectiveProportions
+
+# Obj per study
+sheet_Objectives |>
+  mutate(TotalArticles = length(unique(`Review ID`))) |>
+  pivot_longer(c("Demonstrate methods?", "Estimate trends?", 
+                 "Identify drivers?", "Test relationships?",
+                 "Predict spatially?", "Predict temporally?"),
+               names_to = "Objective", values_to = "ObjectiveIncluded") |>
+  filter(ObjectiveIncluded == "YES") |>
+  summarise(nObj = n(),
+            .by = c(`Review ID`, TotalArticles)) |>
+  summarise(nFreq = n(),
+            .by = c(nObj))
 ```
@@
 ```
+
+## Time Series
+
+#### Number of sites
+
+```{r TimeSeries_Sites}
+timeSeries_Sites <- sheet_DataCollection |>
+  filter(`Site quantity` != "Unclear") |>
+  mutate(`Site quantity` = as.numeric(`Site quantity`)) |>
+  select(`Review ID`, SiteNumber = `Site quantity`) |>
+  left_join(select(sheet_ArticleData, c(`Review ID`, `Year Strata`, Year)), 
+            by = "Review ID") |>
+
+  ggplot(aes(x = `Year Strata`, y = SiteNumber)) +
+  geom_boxplot(fill = "goldenrod", colour = "gray40", staplewidth = 0.4,
+               linewidth = 0.2, , outlier.size = 0.2) +
+  scale_y_log10() +
+  labs(x = "", y = "Number of sites\n(log scale)",
+       title = "Site quantity") +
+  theme(panel.grid.major.x = element_blank(),
+        panel.grid.minor.x = element_blank(), 
+        axis.title = element_text(size = 8, family = "Helvetica"), 
+        axis.text = element_text(size = 8, family = "Helvetica"))
+timeSeries_Sites
+```
+
+#### Number of covariates considered
+
+```{r TimeSeries_Covariates}
+timeSeries_Covariates <- sheet_Covariates |>
+  separate_longer_delim(`Model ID`, delim = ",") |>
+  separate_longer_delim(Covariates, delim = ", ") |>
+  mutate(studyID = paste(`Review ID`, `Model ID`, sep = "_")) |>
+  # Get number of core parameters
+  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
+                                 "Colonisation", "Extinction_Persistence",
+                                 "Detection")) |>
+  mutate(n_params = length(unique(`Parameter type`)), .by = studyID) |>
+  # Remove 'none' covariates
+  filter(Covariates != "None") |>
+  # Calculate number of covariates per parameter
+  summarise(CovsPerParam = n()/(max(n_params)), .by = c(studyID, `Review ID`)) |>
+  # Models with 0 covariates equate to NA, convert these to 0s
+  mutate(CovsPerParam = case_when(is.na(CovsPerParam) ~ 0,
+                                  .default = CovsPerParam)) |>
+  left_join(select(sheet_ArticleData, c(`Review ID`, `Year Strata`, Year)), 
+            by = "Review ID") |>
+
+  ggplot(aes(x = `Year Strata`, y = CovsPerParam)) +
+  geom_boxplot(fill = "goldenrod", colour = "gray40", staplewidth = 0.4,
+               linewidth = 0.2, outlier.size = 0.2) +
+  scale_y_log10() +
+  labs(x = "", y = "Covariates\nper parameter",
+       title = "Covariates considered") +
+  theme(panel.grid.major.x = element_blank(),
+        panel.grid.minor.x = element_blank(), 
+        axis.title = element_text(size = 8, family = "Helvetica"), 
+        axis.text = element_text(size = 8, family = "Helvetica"))
+timeSeries_Covariates
+```
+
+#### Proportion of selection methods (Frequentist)
+
+```{r TimeSeries_Selection}
+timeSeries_Selection <- sheet_Modelling |>
+  separate_longer_delim(`Model ID`, delim = ",") |>
+  mutate(studyID = paste(`Review ID`, `Model ID`, sep = "_")) |>
+  filter(`Model class` == "Frequentist") |>
+  separate_longer_delim(`Selection method`, ", ") |>
+  mutate(Select = 1) |>
+  select(`Selection method`, `Review ID`, studyID, Select) |>
+  left_join(select(sheet_ArticleData, c(`Review ID`, `Year Strata`, Year)), 
+            by = "Review ID") |>
+  mutate(n_strata = length(unique(studyID)),
+         .by = `Year Strata`) |>
+  summarise(sumMethod = sum(Select),
+            .by = c(`Selection method`, `Year Strata`, n_strata)) |>
+  mutate(PropMethod = (sumMethod/n_strata)*100) |>
+  mutate(`Year Strata` = fct(`Year Strata`)) |>
+  filter(`Selection method` %in% c("A priori", "Sequential", 
+                                   "Candidate set", "Simple")) |>
+  complete(`Year Strata`, `Selection method`, fill = list(PropMethod = 0)) |>
+
+  ggplot(aes(x = `Year Strata`, y = PropMethod,
+             colour = `Selection method`, group = `Selection method`)) +
+  geom_line() +
+  geom_point(size = 1) +
+  labs(x = "", y = "Percentage of models",
+       title = "Selection method (MLE*)") +
+  scale_colour_manual("",
+                      limits = c("A priori", "Candidate set", 
+                                 "Sequential", "Simple"),
+                      labels = c("A priori", "Candidate set",
+                                 "Sequential", "Simple precursors"),
+                      values = c("aquamarine4", "cadetblue3",
+                                 "plum4", "goldenrod")) +
+  guides(colour = guide_legend(nrow = 2,
+                               override.aes = list(size = 0.5,
+                                                   linewidth = 0.5))) +
+  theme(legend.position = "top",
+        panel.grid.major.x = element_blank(),
+        panel.grid.minor.x = element_blank(), 
+        axis.title = element_text(size = 8, family = "Helvetica"), 
+        axis.text = element_text(size = 8, family = "Helvetica"),
+        legend.title = element_text(size = 8, family = "Helvetica"), 
+        legend.text = element_text(size = 8, family = "Helvetica"))
+timeSeries_Selection
+```
+
+#### Goodness-of-fit and validation
+
+```{r TimeSeries_Evaluation}
+timeSeries_Validation <- sheet_Modelling |>
+  separate_longer_delim(`Model ID`, delim = ",") |>
+  mutate(studyID = paste(`Review ID`, `Model ID`, sep = "_")) |>
+  separate_longer_delim(`Performance evaluation`, ", ") |>
+  mutate(`Performance evaluation` = case_when(`Performance evaluation` %in% c("GOF", "BayesianP") ~ "GOF",
+                                              `Performance evaluation` %in% c("Validation_In", "Validation_Out") ~ "Validation",
+                                              .default = "None")) |>
+  select(`Performance evaluation`, `Review ID`, studyID) |>
+  left_join(select(sheet_ArticleData, c(`Review ID`, `Year Strata`, Year)), 
+            by = "Review ID") |>
+  mutate(n_strata = length(unique(studyID)),
+         .by = `Year Strata`) |>
+  summarise(sumEval = n(),
+            .by = c(`Performance evaluation`, `Year Strata`, n_strata)) |>
+  mutate(PropEval = (sumEval/n_strata)*100) |>
+  complete(`Year Strata`, `Performance evaluation`, fill = list(PropEval = 0)) |>
+  mutate(`Year Strata` = fct(`Year Strata`)) |>
+  filter(`Performance evaluation` != "None") |>
+
+  ggplot(aes(x = `Year Strata`, y = PropEval, 
+             colour = `Performance evaluation`, group = `Performance evaluation`)) +
+  geom_line() +
+  geom_point(size = 1) +
+  scale_colour_manual("", 
+                      limits = c("GOF", "Validation"),
+                      labels = c("Goodness-of-fit", "Validation"),
+                      values = c("aquamarine4", "goldenrod")) +
+  labs(x = "", y = "Percentage of models",
+       title = "Model assessment") +
+  guides(colour = guide_legend(override.aes = list(size = 0.5,
+                                                   linewidth = 0.5))) +
+  theme(legend.position = "top", 
+        panel.grid.major.x = element_blank(),
+        panel.grid.minor.x = element_blank(), 
+        axis.title = element_text(size = 8, family = "Helvetica"), 
+        axis.text = element_text(size = 8, family = "Helvetica"),
+        legend.title = element_text(size = 8, family = "Helvetica"),
+        legend.text = element_text(size = 8, family = "Helvetica"))
+timeSeries_Validation
+```
+
+#### Framework
+
+```{r TimeSeries_Framework}
+timeSeries_Framework <- sheet_Modelling |>
+  separate_longer_delim(`Model ID`, delim = ",") |>
+  mutate(studyID = paste(`Review ID`, `Model ID`, sep = "_")) |>
+  select(`Model class`, `Review ID`, studyID) |>
+  left_join(select(sheet_ArticleData, c(`Review ID`, `Year Strata`, Year)), 
+            by = "Review ID") |>
+  mutate(n_strata = length(unique(studyID)),
+         .by = `Year Strata`) |>
+  summarise(sumMethod = n(),
+            .by = c(`Model class`, `Year Strata`, n_strata)) |>
+  mutate(PropMethod = (sumMethod/n_strata)*100) |>
+  mutate(`Year Strata` = fct(`Year Strata`)) |>
+  complete(`Year Strata`, `Model class`, fill = list(PropMethod = 0)) |>
+  filter(`Model class` != "Other") |>
+
+  ggplot(aes(x = `Year Strata`, y = PropMethod,
+             colour = `Model class`, group = `Model class`)) +
+  geom_line() +
+  geom_point(size = 1) + 
+  scale_colour_manual("", 
+                      limits = c("Frequentist", "Bayesian"),
+                      labels = c("Maximum Likelihood", "Bayesian"),
+                      values = c("aquamarine4", "goldenrod")) +
+  guides(colour = guide_legend(nrow = 2, 
+                               override.aes = list(size = 0.2,
+                                                   linewidth = 0.2))) +
+  labs(x = "", y = "Percentage of models", 
+       title = "Model framework") +
+  ylim(0,100) +
+  theme(legend.position = "top",
+        panel.grid.major.x = element_blank(),
+        panel.grid.minor.x = element_blank(), 
+        axis.title = element_text(size = 8, family = "Helvetica"), 
+        axis.text = element_text(size = 8, family = "Helvetica"),
+        legend.title = element_text(size = 8, family = "Helvetica"),
+        legend.text = element_text(size = 8, family = "Helvetica"))
+timeSeries_Framework
+```
+
+#### Combine plots
+
+```{r TimeSeries_Combined}
+timeSeries_Combined <- timeSeries_Sites + timeSeries_Covariates + timeSeries_Framework + timeSeries_Selection + free(timeSeries_Validation, "space", "l") +
+  plot_layout(design = "1122
+                        3344
+                        #55#") +
+  plot_annotation(tag_levels = "A") & 
+  theme(legend.position = "top",
+        axis.text.x = element_text(size = 6, family = "Helvetica", 
+                                 angle = 45, vjust = 0.5, hjust = 1),
+        axis.text.y = element_text(size = 6, family = "Helvetica"),
+        axis.title = element_text(size = 6, family = "Helvetica"),
+        legend.text = element_text(size = 6, family = "Helvetica"),
+        legend.title = element_text(size = 6, family = "Helvetica"),
+        legend.key.size = unit(.2, 'cm'),
+        legend.margin=margin(0,0,0,0),
+        legend.box.margin=margin(0,-10,-10,-10),
+        plot.tag = element_text(size = 8, family = "Helvetica"),
+        plot.title = element_text(size = 6, family = "Helvetica",
+                                  hjust = 0.5))
+timeSeries_Combined
+
+ggsave(plot = timeSeries_Combined, "Figures/TimeSeries.png",
+       width = 11, height = 14, units = "cm")
+```

### 2025-02-03 Manuscript.qmd
Completed first-pass revisions for Ecography

--- 2025-01-23 Manuscript.qmd
@@
 abstract: | 
- Occupancy models.
+ Since their introduction over twenty years ago, dynamic occupancy models (DOMs) have developed into a powerful and flexible framework for estimating species occupancy across space and throughout time while accounting for imperfect detection.
+As their popularity has increased and extensions have broadened their capabilities, DOMs have been applied to increasingly diverse datasets and research objectives in applied ecology.
+In light of this expansion, it is necessary to reassess common practices in building DOMs to identify potential vulnerabilities in model development and identify priorities for future research.
+We review how DOMs have been applied since their introduction, examining the contexts of their application and the approaches taken for model building.
+Our findings indicate that DOMs have been readily applied to datasets of wide-ranging size and spatial and temporal scales, resulting in variation in the definition of site occupancy within studies that leads to challenges in building and interpreting DOMs.
+Current practices for model selection and evaluation also vary considerably, and there remains little research on the impact of these choices on model performance.
+Where possible we provide recommendations for current users to develop models appropriate for their research contexts, and where uncertainties remain we highlight key priorities for further research into DOMs to support users in producing the most reliable models possible.
 
@@
     theme: simplex
+    mainfont: Georgia, serif
+  docx: default
 
@@
 
-Background Information section
+A considerable portion of ecological research depends on an ability to accurately estimate where and when species occur on the landscape.
+Understanding the patterns by which various species are distributed has been a fundamental question since the origins of the field, and the subject continues to be an active area of quantitative research and methods development [@humboldt1849].
+This is for good reason — species occupancy is a valuable concept with broad relevance in wildlife management and conservation biology.
+Estimates of how widespread a species is and where it occurs are the foundation of monitoring programs and are vital for assessing conservation status, while identification of the potential drivers of occurrence is a key first step in informing potential management action [@mackenzie2013].
+Robust knowledge of a species' occupancy patterns can also aid prediction of where it is most likely to occur, both under present conditions and in hypothetical future scenarios [@kéry2013].
 
-#### Model form and assumptions
+Irrespective of its importance, 'occupancy' has proven to be a persistently difficult quantity to estimate in practice.
+This is largely due to the challenges inherent in modelling complex and dynamic natural systems with incomplete data.
+It is often impossible to reliably determine whether a species truly occupies a site, and it is now well-established that imperfect detection of organisms results in biased estimates of occupancy [@gu2004; @lahoz-monfort2014].
+Despite this fact and the ubiquity of imperfect detection in field data, many models in widespread use make no adjustment for it [@kellner2014].
+Research has also indicated that many of these tools, including conventional correlative species distribution models, struggle under non-equilibrium conditions where a species’ occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010].
+These conditions, often associated with biological invasions and climate change driven range shifts, are increasingly commonplace in the Anthropocene [@bertelsmeier2013; @lenoir2015].
+In recent years, awareness of the obstacles to accurate occupancy estimation and the limitations of many popular methods has encouraged interest in somewhat more process-explicit models better suited to these contexts [@briscoe2019].
 
-Model form.
+Within this category are the 'dynamic occupancy models' (DOMs), first introduced by @mackenzie2003 as an extension to earlier static site occupancy models [@mackenzie2002].
+In their simplest form, DOMs estimate the probability of site occupancy over multiple time-steps while correcting for imperfect detection using hierarchical observation data.
+For newcomers to the models we provide a brief introduction to their form and requirements in [Box @nte-Primer].
+By accounting for imperfect detection and explicitly modelling colonisation and extinction to capture changes in site occupancy dynamics, DOMs provide answers to common issues in occupancy estimation and offer important advantages over other common models [@guillera-arroita2017a].
+Despite these advantages, they require only reasonably simple-to-collect presence absence/data rather than the detailed demographic or abundance data required by other more process-explicit models [@briscoe2019].
+This balance between complexity and feasibility makes DOMs an appealing option for many applied ecologists.
 
-![The form of the dynamic occupancy model as described by @mackenzie2003.
-The ecological process sub-model describes changes in occupancy over time, while the observational process sub-model describes detectability.](Figures/ModelForm.png){#fig-modelform}
+::: {#nte-Primer .callout-note icon="false"}
+## An introduction to dynamic occupancy models
 
-Survey design.
+DOMs are a hierarchical model which links observed detection data with the underlying latent process of changing site occupancy.
+To do this, they contain two sub-models: one which describes the ecological process of how sites shift between unoccupied and occupied states over time, and one which describes the observation process which describes whether or not a species is detected during surveys at occupied sites.
 
-![The sampling design of the standard dynamic occupancy model.
-During seasons (also called primary occasions) sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons.
-Within each season, multiple observations (also called surveys or secondary occasions) are conducted to record the observed presence or absence of the species at each site.
-These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded.
-Note that it is not necessary for each site or season to have the same number of observations, and that missing data can be accommodated.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}
+To separate these processes DOMs require hierarchical data inputs.
+Occupancy is estimated at independent sites *(i)* during discrete, time-bound intervals termed primary occasions *(t)*.
+Within primary occasions*,* multiple observations are conducted at the same sites during secondary occasions *(j).* During each observation, a species is either detected (y~i,t,j~ = 1) or not (y~i,t,j~ = 0).
+Importantly, it is not necessary for all primary occasions to contain the same number of secondary occasions, and missing observations at sites can be accomodated.
+While secondary occasions are most often represented as repeat observations, other options including same-visit replicates and spatial replicates may also be used – see @guillera-arroita2017a for further discussion on these methods.
 
-Assumptions.
+![](images/clipboard-2772050481.png){fig-align="center" width="1476"}
 
-# Review methods
+DOMs' occupancy sub-model describes how site occupancy changes across primary occasions.
+In each primary occasion *t,* each site *i* may exist in either an occupied (z~i,t~ = 1) or unoccupied state (z~i,t~ = 0).
+In the first primary occasion, occupancy is determined by a Bernoulli trial with initial occupancy probability 𝛙~1~, such that $z_{i,t=1} \sim Bernoulli(\psi_1)$.
+In subsequent primary occasions, site occupancy is determined by the site's state in the previous primary occasion and probabilities of colonisation 𝛄 and extinction 𝛆, such that $z_{i,t} \sim Bernoulli(z_{i,t-1}(1-\epsilon)+(1-z_{i,t-1})*\gamma)$.
 
-Methods.
+![](images/clipboard-3349624601.png){fig-align="center" width="1476"}
+
+The observation sub-model of DOMs describes how the observed detection data (y) relate to the latent occupancy states (z).
+During any given secondary occasion *j* during primary occasion *t* at site *i,* observers will either observe (y~i,t,j~ = 1) or not observe (y~i,t,j~ = 0) the target species.
+In each secondary occasion, the probability of observing the target species at a site is given as a Bernoulli trial with detection probability *p,* conditional on the site being occupied in primary occasion *t*, such that $y_{i,t,j} \sim Bernoulli(z_{i,t}*p)$.
+This formulation assumes that no false positive detections exist; that is, observations are never recorded at truly unoccupied sites.
+
+![](images/clipboard-411045638.png){fig-align="center" width="945"}
+
+In this formulation, DOMs estimate four parameters of interest: Initial occupancy (𝛙~1~), colonisation (𝛄), extinction (𝛆), and detection (*p)*.
+While we have represented each of these parameters as constants for simplicity, they can also be estimated with respect to covariates to describe variation in the occupancy or detection processes - these are typically incorporated via an logit link function, such that DOMs can be considered four interconnected generalised linear models.
+
+DOMs do make some important assumptions of the data used to fit them: i) sites are closed to changes in true occupancy state within each primary occasion, ii) occupancy at each site is independent of other sites, iii) observations within each primary occasion are independent of each other, iv) no false positive detections occur, and v) no un-modelled heterogeneity exists in occupancy or detection probabilities.
+
+For more detailed instruction on DOMs and related models, consider texts by @mackenzie2006 and @kéry2021.
+:::
+
+Over the past twenty-some years, continued research on DOMs (and occupancy models more broadly) has established a powerful and flexible modelling framework.
+Freely available software tools, including program PRESENCE and the R packages *unmarked* and *ubms* among others, have encouraged uptake among applied ecologists [@hines2006; @kery2016; @kellner2022].
+Various extensions to the models have further broadened their capabilities – versions of DOMs are available which can account for false positives [@royle2006; @miller2011], model multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimate occupancy for multiple species [@dorazio2010; @devarajan2020]; @bailey2014 provides an overview of many of these extensions.
+Even with this expansion, there remains room for further growth in applications of DOMs.
+One area of interest is in DOMs' ability to make more robust predictions of species distributions, a use-case for which they have only seen limited use despite their potential [@kéry2013; @briscoe2019].
+
+Regardless of use-case, DOMs have always been somewhat challenging models to build, evaluate, and interpret.
+Even defining the term 'occupancy' poses difficulties – what it represents depends on several factors, including the definition of a site, the duration of a primary occasion, and how these relate to the behaviour of the species of interest [@steenweg2018; @wood2022; @valente2024].
+This definition of occupancy should inform the nature of the covariates considered for each modelled parameter, although determining which of these factors to include in a model leads to additional questions.
+Covariate selection is particularly difficult for hierarchical models; with four parameters capable of being estimated with respect to covariates (initial occupancy, colonisation, extinction, and detection), the number of candidate models can rapidly balloon and render exhaustive comparison impossible.
+The choice of model selection method, and the approach to assessing the resulting models' performance, are important decisions that any user of DOMs must make with respect to their research context and objectives.
+
+The years since the introduction of DOMs have seen enormous changes in how ecologists conduct their research.
+The amount of data available for modelling has grown considerably, aided by improvements in data sharing and large-scale monitoring efforts.
+Increased interest in citizen science efforts has unlocked another source of data with its own set of considerations [@altwegg2019].
+Technological advances have facilitated the widespread deployment of autonomous detection methods such as camera traps and acoustic monitors, generating massive quantities of fine-scale observation data suitable for analysis with DOMs [@balantic2019].
+At the same time, advances in computing have made methods which may have been prohibitively expensive in the past far more accessible for many.
+While these advances create exciting new opportunities, they also introduce new challenges.
+
+### Objectives
+
+DOMs are an increasingly popular tool for occupancy estimation with broadening applications in applied ecological research.
+As model users encounter research questions, data structures, and modelling techniques beyond the context of the models' original development, it is necessary to revisit common approaches for building DOMs to identify any potential vulnerabilities, areas of confusion, and uncertainty in best practises.
+We conduct an in-depth review of studies using DOMs since their introduction, examining how the context of their application and practices for model building and evaluation have changed over the past two decades.
+We use our findings to identify the largest challenges in building DOMs and to provide recommendations for current users where supported by the literature.
+Where uncertainties remain, we highlight key priorities for research to better support ecologists in building robust models suited to their research contexts.
+
+# Methods
+
+Our review was constrained to applications of the dynamic multi-season occupancy model of @mackenzie2003 and its extensions.
+To capture only these studies, our inclusion criteria required that articles:
+
+i.  Used non-simulated, field-collected, presence-absence data.
+
+ii. Included data from multiple sites which could exist in at least two states, including an occupied and unoccupied state.
+
+iii.
+Included data from multiple primary occasions, between which sites change states conditional on the prior occasions occupancy state *and* transition probabilities, such as (but not limited to) colonisation and extinction.
+
+iv. Contained at least one parameter describing the detection process.
+
+A pool of candidate articles was generated using two queries on Web of Science on July 26, 2024.
+The first query included all articles from 2004-2023 which cited @mackenzie2003.
+To capture any additional relevant articles that did not directly cite @mackenzie2003, a second query was generated searching articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract.
+These queries resulted in 1469 articles: 897 retrieved only from the MacKenzie citations, 274 only from the keywords search, and 298 which appeared in both queries.
+To allow comparison of DOMs through time, we divided all articles across five four-year strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024.
+From each strata we randomly selected 20 articles for inclusion in the review.
+Articles that did not meet inclusion criteria were replaced from within their own stratum unless no articles remained.
+For each article, we documented details of the study's research context and their modelling process.
+
+Aspects of research context including authorship, stated research objectives, study taxa and system, and survey methods used to collect data used for modelling.
+For all authors, the category (academic, government, non-government organisation, or private company) and host country of all affiliated institutions was recorded.
+To examine the reasons why authors use DOMs, we allocated each article to one or more research objectives category based on the study's stated aims.
+The six possible categories included "estimating trends*"*, where authors express interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; "testing relationships", where authors examine predefined hypotheses of relationships between covariates and model parameters; "identifying drivers", where authors more broadly explore factors associated with model parameters; "predicting temporally", where authors predict site occupancy under future conditions, "predicting spatially", where authors predict site occupancy to unsurveyed locations, and "developing methods", where authors introduce, test, or demonstrate aspects of DOMs.
+To develop an understanding of the scale of studies using DOMs, we recorded the quantity of sites, the number of primary and secondary occasions, and the spatial scale of the study location.
+Spatial scale was defined as the intended area of inference containing all sites, measured to an order of magnitude to account for uncertainty in reporting.
+Details on the taxa modelled were also recorded, including the quantity of species modelled, their identities, and their conservation status; taxa were denoted as threatened either where they are listed on the IUCN Red List of Species (as of 2024) or when authors explicitly state that they are threatened.
+This deference to authors' representation of conservation status was made to account for sub-species which lack listings or species which are of more local concern, although we acknowledge that in some cases conservation status may have changed since publication.
+
+A key objective of this review was to understand how users of DOMs have navigated the modelling process, through covariate selection through model evaluation.
+One aspect of this is the type of covariates incorporated in the models – to this end, we recorded all covariates considered in each study regardless of whether they were or were not included in final models.
+Key traits of each covariate were recorded including a general categorisation, whether they were directly observed or remotely sensed, whether they were static or varied between primary occasions, and how they were represented in the model (as a linear term, a polynomial term, or as part of an interaction with another covariate [@james2021]).
+We also closely recorded how authors approached model selection.
+These procedures were sorted into non-exclusive categories which included "*a priori"*, where only one model was considered; candidate suite, where a predefined set of models was considered; "sequential", where covariates were selected parameter-by-parameter (e.g., fitting detection first, followed by initial occupancy and so on); and "simple precursors", where covariate selection was preceded by tests with a simpler model implementation (such as a linear regression or single season occupancy model).
+Finally, for each model we documented whether goodness-of-fit was tested and reported, and whether model performance was assessed by validation with either in or out-of-sample data.
 
@@
 
-Coverage.
+92 articles were included for this review, fewer than the 100 possible articles due to a deficit of qualifying papers from the 2004-2007 stratum.
+Based on the acceptance rates within each stratum and the remaining number of unprocessed articles, an estimated 496 of the 1152 unprocessed articles in our sample would have met inclusion criteria, with an apparent increase over time in the number of articles fitting DOMs (@fig-coverage).
 
-![Bars indicate the 92 articles included in our review as a proportion of the estimated number of published articles fitting DOMs, based on the qualification rate for articles in each strata.
-The proportion of articles included from each strata were as follows: 12% from 2005-2008; 24% from 2008-2011; 42% from 2012-2015; 35% from 2016-2019; and 57% from 2020-2023.](Figures/CoveragePlot.jpeg){#fig-coverage}
+![Bars indicate the number of articles included in our review as a proportion of the estimated number of articles fitting DOMs in each year.
+Estimates were calculated using the number of unprocessed articles remaining in each year multiplied by the acceptance rate of the corresponding stratum: 12% from 2005-2008; 24% from 2008-2011; 42% from 2012-2015; 35% from 2016-2019; and 57% from 2020-2023.](Figures/CoveragePlot.jpeg){#fig-coverage width="945"}
 
-Stamp collection.
+### Research contexts
+
+Authors of the articles in our sample represent institutions spanning all major sectors of ecological research.
+79% of studies included at least one affiliate of an academic institution, 59% a government affiliate, 28% an NGO affiliate, and 12% an affiliate of a private company.
+63% of these studies had cross-sector authorship with affiliations from at least two of these categories.
+While the institutions producing these articles were located in 27 countries across all continents barring Antarctica, 68% of the articles included at least one author affiliated with an institution in the United States of America.
+
+To understand how authors used their DOMs, we classified each study according to their stated research objectives [@fig-Objectives].
+These objective categories are non-exclusive, and 37% of studies pursued multiple objectives simultaneously.
+In the simplest use-case, 36% of articles used DOMs to monitor trends in occupancy state through time, both for single species of high conservation interest and for broader community assemblages of species [@ahumada2013; @scott2015].
+Other studies were more focused on the factors influencing species occupancy, either testing pre-defined hypothetical relationships (44% of articles) or taking a more exploratory approach to identifying possible drivers of occupancy (35%).
+Many articles in the former category target core conservation priorities for their focal species, such as @olson2005's early study assessing the influence of barred owls (*Strix varia)* on the threatened spotted owl (*Strix occidentalis*); in the latter grouping authors examine the influence of a wider variety of factors, as in @huber2017's study testing the relative influence of dozens of habitat covariates on wood warbler (*Phylloscopus sibilatrix*) occupancy.
+While only 17% of articles used their DOMs to make predictions, the proportion of articles pursing this objective has increased in recent years.
+These studies tend to have a strong conservation focus – as an example, @mcgowan2020 projects future occupancy for wetland birds under alternative management scenarios.
+22% of papers reviewed focused on methods development, representing continued work in the literature on extending and testing DOMs.
+
+![The percentage of articles which pursue each of six non-exclusive research objectives.
+Percentages are given for each 4-year stratum, as well as for all articles in the study period.](Figures/ObjectiveTrend.jpeg){#fig-Objectives width="945"}
+
+Reviewed studies used data from a diverse range of study systems and focal species.
+The locations of study sites included in our sample are globally distributed: while a majority of articles use data collected in the United States of America, representatives are included from all continents except Antarctica (@fig-StudyDetails A).
+These study sites vary considerably in size, ranging from the hyper-local to the continental scale (@fig-StudyDetails B).
+The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre, while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@clement2019; @basset2023].
+The temporal scale of studies shows similar variability – time elapsed between the first and last survey ranged from under one month to over forty years, with a median duration of 8 years (@fig-StudyDetails F).
+
+The majority of studies in our sample focus on vertebrate taxa [@fig-StudyDetails], particularly birds and mammals.
+DOMs have been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community.
+However, there are exceptions, including DOMs used to model decadel changes in lichen occupancy and the spread of chronic wasting disease [@belinchón2017; @cook2022].
+Others have used DOMs to model mosquito dynamics, a strongly applied use-case with important human health implications [@mores2020; @padilla-torres2013].
+The vast majority of studies model terrestrial species, though there have been a limited number of studies on aquatic systems, including invasive salmon, Great Plains stream fishes, and whales [@fisher2014; @falke2012; @pendleton2022].
+While most studies fit models for a single taxa, 44% fit models to multiple taxa (@fig-StudyDetails D) .
+34% of studies fit independent models to multiple species [@otto2012; @peach2019], wheras 12% used explicitly multi-species extensions of DOMs.
+These included hierarchical models which fit hundreds of species in a single implementation with species-level effects [@dorazio2010; @hendershot2020] as well as explicit interaction models which estimated conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019].
 
@@
 
-Covariates.
+The presence/absence data required for DOMs was collected using various survey methods (@fig-StudyDetails E).
+in our sample, 79% of studies used human observational data, 11% used live-trapping methods, and 14% used detections from camera traps.
+These human-observed data include the 10% of studies that used data collected by citizen-scientists to build detection histories.
+Citizen science data ranged from coordinated surveys at backyard bird-feeders to interviews with civilians seeking reports of tiger (*Panthera tigris*) signs [@zuckerberg2011; @warrier2020].
+Several studies used particularly unique survey methods to determine the presence of a species at a location.
+For example, in the only application of DOMs to marine species in our sample, @pendleton2022 used aerial transects broken up into grid cells to observe whale occupancy.
+In another unique implementation, @marescot2020 fit a unique multi-species model treating poachers as a taxon and using ranger reports to create detection histories.
+Notably, not all of this data was collected in a hierarchical structure with DOMs in mind.
+In these cases, authors formatted their data into a hierarchical format *post-hoc* using a variety of methods.
+Some defined primary occasions as arbitrary discrete time intervals, treating all surveys occurring within the season as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate.
+These manipulations permit authors to use data which predate the DOM, as in a study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].
 
-::: column-page-inset
-```{r covariate table}
-#| echo: false
-#| message: false
-#| label: tbl-covariates
-#| tbl-cap: "All covariates considered for inclusion in a study were classified into mutually exclusive categories.
-We calculate the percentage of studies which include at least one covariate of a given category on any parameter, Initial Occupancy (&psi;1), Occupancy (&psi;), Colonisation (&gamma;), Extinction(&epsilon;), and Detection (&rho;).
-We also present the average percentage of covariates in a category which are dynamic (varying through seasons) and directly observed, as well as the percentage of articles which model each category of covariate with a non-linear relationship or as part of an interaction with another covariate."
+### Modelling practices
 
-library(tidyverse)
-library(gt)
+To consider how authors incorporate covariates in their DOMs, we recorded all covariates considered for each model in our sample, without regard for whether they were included in final models.
+@fig-CovariateSummary presents a taxonomy of these covariates, stating the proportion of studies considering each type of covariate, whether covariate data was directly observed at sites, and how covariates responses were represented in the models.
+We further delineate these covariates by classifying them two groups: environmental covariates, representing plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environment.
+Our findings indicate that DOM users have incorporated a wide diversity of covariates in their models – while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample.
+Many models incorporate covariates representing aspects of site geometry and connectivity (35% of studies).
+Often these are included as simple covariates on colonisation or extinction, such as the distance to other sites or landscape habitat connectivity metrics [@duggan2011].
+Alternatively, more complex parameterisations explicitly model colonisation or extinction as a spatial process and blend attributes of DOMs with patch occupancy models [@broms2016; @risk2011].
+Several studies also include biotic interactions with other species as covariates, particularly those focused on taxa threatened by invasive species.
+These covariates effectively incorporate species interactions in DOMs without requiring the use of more complex explicitly multi-species models.
 
-covTable <- read_csv("Figures/CovariateTable.csv")
+![A summary of the covariates considered for the models in our sample.
+The first row gives the percentage of studies which considered at least one covariate in a given category.
+The second gives the percentage of covariates considered which are dynamic and vary between occasions; this category does not include covariates used for initial occupancy probabilities which are static by definition.
+The third row presents the percentage of covariates which are directly observed at sites, rather than being remotely sensed.
+The final row gives the percentage of covariates in each category where non-linear responses or interactions with other covariates were considered, rather than only testing linear responses.](Figures/CovariateSummary.png){#fig-CovariateSummary}
 
-covTable |>
-  mutate(Summary = case_when(Covariate %in% c("Environmental", "Structural", "Overall") ~ TRUE,
-                             .default = FALSE)) |>
-  arrange(Category, Summary, -Prop_Any) |>
-  gt(rowname_col = "Covariate", groupname_col = "Category") |>
-  tab_options(
-    row_group.as_column = TRUE
-  ) |>
-  tab_row_group(
-    label = md("Environmental covariates"),
-    rows = Category == "E",
-    id = "Env"
-  ) |>
-  tab_row_group(
-    label = md("Structural covariates"),
-    rows = Category == "S",
-    id = "Struct"
-  ) |>
-  tab_row_group(
-    label = md("All covariates"),
-    rows = Category == "O",
-    id = "All"
-  ) |>
-  tab_style(
-    style = cell_fill(color = "aquamarine4"),
-    locations = cells_row_groups(groups = "Env")
-  ) |>
-  tab_style(
-    style = list(
-      cell_fill(color = "plum4")
-    ),
-    locations = cells_row_groups(groups = "Struct")
-  ) |>
-  tab_style(
-    style = list(
-      cell_fill(color = "gray60")
-    ),
-    locations = cells_row_groups(groups = "All")
-  ) |>
-  row_group_order(groups = c("Env", "Struct", "All")) |>
-  tab_spanner(label = "Percentage of articles with covariate on parameters:",
-              columns = c("Prop_Any", "Initial occupancy", "Occupancy",
-                          "Colonisation", "Extinction_Persistence",
-                          "Detection"),
-              id = "inclusions") |>
-  tab_spanner(label = "Percentage which are:",
-              columns = c(Prop_Dynamic, Prop_Direct),
-              id = "covariates") |>
-  tab_spanner(label = "Articles representing this covariate with:",
-              columns = c(Prop_Nonlinear, Prop_Interact),
-              id = "representations") |>
-   cols_label(
-    Prop_Any = "Any",
-    `Initial occupancy` = html("&psi;1"),
-    Occupancy = html("&psi;"),
-    Colonisation = html("&gamma;"),
-    Extinction_Persistence = html("&epsilon;"),
-    Detection = html("&rho;"),
-    Prop_Dynamic = "Dynamic",
-    Prop_Direct = "Directly observed",
-    Prop_Nonlinear = "Non-linear relationships",
-    Prop_Interact = "Interactions between covariates"
-  ) |>
-  text_case_match(
-    "Environmental" ~ html("<b>Any Environmental</b>"),
-    "HABT" ~ html("<b>Habitat </b>\n<i>Aspects of habitat and land cover</i>"),
-    "PHEN" ~ html("<b>Phenology</b>\n<i>Time-varying elements distinct from sampling occasions</i>"),
-    "SPAT" ~ html("<b>Spatial</b>\n<i>Site geometry, connectivity, or other spatial elements</i>"),
-    "OTHER-E" ~ html("<b>Other environmental</b>\n<i>Other environmental covariate not otherwise listed</i>"),
-    "TOPO" ~ html("<b>Topography</b>\n<i>Elements of landscape topography</i>"),
-    "HYDR" ~ html("<b>Hydrology</b>\n<i>Aspects of hydrology, such as distance to water</i>"),
-    "CLWE" ~ html("<b>Climate/Weather</b>\n<i>Climate, weather, and natural disasters</i>"),
-    "ANTH" ~ html("<b>Anthropogenic</b>\n<i>Relations to human activity</i>"),
-    "BIOT" ~ html("<b>Biotic interaction</b>\n<i>Interactions with other (non-plant) species</i>"),
-    "Structural" ~ html("<b>Any Structural</b>"),
-    "SEAS" ~ html("<b>Primary occasion</b>\n<i>Effect of the primary occasion</i>"),
-    "OBSE" ~ html("<b>Observation</b>\n<i>Details on the observation process</i>"),
-    "SURV" ~ html("<b>Secondary occasion</b>\n<i>Effect of the secondary occasion</i>"),
-    "SPEC" ~ html("<b>Species effect</b>\n<i>Species-level effects</i>"),
-    "SITE" ~ html("<b>Site effect</b>\n<i>Site-level effects</i>"),
-    "OTHER-S" ~ html("<b>Other structural</b>\n<i>Other structural covariate not otherwise listed</i>"),
-    "Overall" ~ html("<b>All covariates</b>"),
-    .locations = cells_stub()
-  ) |>
-  cols_hide(c(Category, Summary)) |>
-  fmt_percent(decimals = 0) |>
-  sub_missing(missing_text = "-") |>
-  data_color(
-    columns = where(is.numeric),
-    method = "numeric",
-    palette = "viridis",
-    alpha = 0.5
-  ) |>
-  tab_style(locations = cells_stub(rows = "Environmental"),
-            style = cell_fill(color = "aquamarine4", alpha = 0.5)) |>
-  tab_style(locations = cells_body(rows = "Environmental"),
-            style = list(cell_fill(color = "aquamarine4", alpha = 0.5),
-                         cell_text(weight = "bold", style = "italic",
-                                   color = "black"))) |>
-  tab_style(locations = cells_stub(rows = "Structural"),
-            style = cell_fill(color = "plum4", alpha = 0.5)) |>
-  tab_style(locations = cells_body(rows = "Structural"),
-            style = list(cell_fill(color = "plum4", alpha = 0.5),
-                         cell_text(weight = "bold", style = "italic",
-                                   color = "black"))) |>
-  tab_style(locations = cells_stub(rows = "Overall"),
-            style = cell_fill(color = "gray60", alpha = 0.5)) |>
-  tab_style(locations = cells_body(rows = "Overall"),
-            style = list(cell_fill(color = "gray60", alpha = 0.5),
-                         cell_text(weight = "bold",
-                                   color = "black"))) |>
-  tab_style(locations = cells_row_groups(),
-            style = cell_text(color = "white")) |>
-  as_raw_html()
-```
-:::
+Covariate data for studies in our sample was either collected directly by researchers (an average of 30% of environmental covariates per study), or derived from pre-existing remotely sensed datasets (70% of environmental covariates); this of course varied depending on the category of covariate (@fig-CovariateSummary).
+Directly collected data often represents finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely.
+An average of 43% percent of environmental factors and 94% of structural factors included in reviewed studies were dynamic covariates which varied through time — this too varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static (@fig-CovariateSummary).
 
-More covariates.
+In the conventional DOMs, covariates for each parameter are most commonly incorporated via a logistic regression (i.e., a linear regression through a logit link function) [@mackenzie2017].
+Statistical relationships between model parameters and covariates (e.g., between initial occupancy and its environmental covariates) are represented as linear terms unless more complexity is specified, although non-linear responses can be easily accommodated in DOMs by using polynomial transformations and interactions between covariates.
+Despite this, in our sample only 35% of articles tested one or more non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear terms.
+Interactions between covariates were similarly rare, with only 24% of studies considering at least one interaction between terms.
+
+In addition to the types of covariates considered for modelling we also recorded the size of the candidate covariate pool in each reviewed study, tallying the number of environmental and structural covariates which were available for use on each parameter (@fig-CovariateCount).
+This is an area of considerable variation in modelling practices — the number of covariates considered for each parameter ranges from 0 (effectively modelling the parameter as a constant) to over 40 candidates on a single parameter.
+Note that this does not represent the number of covariates included in the final model formulation used for inference, but it does represent the upper limit of complexity in candidate models.
+The median number of covariates varies by parameter, with transition probabilities (colonisation and extinction) more likely to consider a broader range of environmental covariates compared to initial occupancy and detection.
+The lack of any covariates considered for initial occupancy in 37% of studies is particularly notable.
+Detection probability is also typically represented with low numbers of environmental covariates @fig-CovariateCount.
 
@@
 Here, a 'covariate' is defined as each distinct covariate considered for inclusion.
-Linear and quadratic representations of the same covariate are counted as one covariate.](Figures/CovParamPlot.jpeg){#fig-covariates}
+Linear and quadratic representations of the same covariate are counted as one covariate.](Figures/CovParamPlot.jpeg){#fig-CovariateCount width="945"}
 
-Modelling summary.
+To explore how users of DOMs have approached the modelling process, we documented the methods used for model selection and evaluation in all articles in our sample [@tbl-modelling].
+Some of these articles included multiple distinct modelling workflows, resulting in a higher total number of models than articles (n = 102 vs.
+n = 92).
+Of these models, 76 were fit by maximum likelihood and 24 were Bayesian implementation.
+2 models fell into neither category and used machine learning based methods.
 
-```{r Modelling summary table}
+In most cases, the DOMs fit in a study did not use all covariates considered for inclusion.
+80% of modelling workflows included some form of model selection approach to identify a final model or set of models to make inference from.
+Approaches varied between the maximum-likelihood and Bayesian (n = 24) implementations in our sample – where 95% of frequentist models perform some manner of model selection to determine covariate inclusions for their models, only 33% of Bayesian models do so, with the majority instead fitting a single model defined *a priori*.
+For frequentist models, the most popular and conventional approach to model selection (45% of models) involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score.
+The next most popular method in frequentist studies is to use sequential model selection methods (37% of models), where the structure for each model parameter is fit in sequence.
+For example, this protocol might first identify the best structure for detection probability while holding the other parameters constant, before moving on to initial occupancy and so on until all parameters are fixed.
+The remainder of frequentist studies (8%) use a variety of other approaches, such as fitting simpler models like single season occupancy models to identify the most informative covariates to use in a dynamic occupancy model.
+Across all frequentist implementations, 47% of articles use multi-model inference by model-averaging with AIC weights [@burnham2004].
+
+```{r Table_Modelling}
 #| echo: false
@@
 #| label: tbl-modelling
-#| tbl-cap: "Modelling practices in dynamic occupancy models, subset by frequentist or bayesian implementations.
+#| tbl-cap: "Modelling practices in dynamic occupancy models, subset by maximum likelihood or bayesian implementations.
 The median covariate count presented here represents the median quantity of covariates considered for each model parameter across the studies in our review.
 The model selection methods represented in this table are non-exclusive and some articles employ multiple approaches.
-2 models included in the 'Overall' column are neural network based and fall into neither the Frequentist or Bayesian categories."
+2 models included in the 'Overall' column are neural network based and fall into neither the MLE or Bayesian categories."
 
-modellingSummary <- read_csv("Figures/ModellingSummary.csv")
+library(tidyverse)
+library(gt)
+modellingSummary <- read_csv("Figures/ModellingSummary.csv") |>
+  rename(`Maximum likelihood` = Frequentist)
 
@@
     style = cell_text(style = "italic")
-  ) 
-#|>
- # as_raw_html()
+  ) |>
+  as_raw_html()
 ```
 
-Objectives.
+Those Bayesian models which *do* perform model selection take various approaches, with largely idiosyncratic methods across these studies.
+While direct comparison of model fit is rare in Bayesian methods, it is feasible – @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data.
+Another approach used by @cook2022 fit a global model including all covariates before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the reduced model.
+@ahumada2013 takes a hybrid approach, in which model selection is conducted by a procedural method in the maximum likelihood framework before refitting the best structure as a Bayesian model.
 
-![A) Proportion of articles in each year-strata and across all years which match each of six non-exclusive objective categories.
-B) Quantity of covariates considered per parameter for models which pursued each objectives.](Figures/ObjectiveFigure.jpeg){#fig-objectives}
+Regardless of the implementation, assessment of model fit and model performance was rare amongst the articles in our sample.
+Only 18% of studies tested for goodness-of-fit, and just 7% calculated predictive performance with either in-sample or out-of-sample validation.
 
-More objectives.
+Our temporally stratified sample allows us to consider how certain aspects of model building have changed over two decades of applying DOMs [@fig-timeSeries].
+One notable trend is in the number of sites modelled in each study.
+In our two most recent strata spanning 2016-2023, the range in site quantity was noticeably wide.
+While models continue to be fit with small numbers of sites, the upper quartile contains more sites than in previous strata (@fig-timeSeries A).
+The median number of covariates considered per parameter has also continually increased since the early years of DOMs, from 1.75 in 2004-2007 to 4.75 in the 2020-2023 stratum.
+(@fig-timeSeries B).
+
+![Key trends in application of DOMs over the study period.
+A) The number of sites included in the dataset for each article, presented on the log scale.
+B) The number of covariates considered per model parameter in each study, presented on the log scale.
+C) The percentage of articles fitting DOMs in either the maximum likelihood or Bayesian frameworks in each stratum.
+D) The percentage of articles using each of four non-exclusive covariate selection strategy in each stratum.
+This figure includes only articles fitting DOMs via maximum likelihood.
+E) The percentage of articles reporting goodness-of-fit testing and model validation (with either in or out of sample data) in each year stratum.](Figures/TimeSeries.png){#fig-timeSeries width="1961"}
+
+In a notable shift, Bayesian implementations eclipsed maximum likelihood for the first time in the 2020-2023 stratum (@fig-timeSeries C).
+Prior to this, Bayesian DOMs had been implemented at a more modest rate since the earliest strata in our sample.
+Within the MLE implementations, the most popular model selection methods have also shifted over time.
+The number of studies using predefined sets of candidate models has gradually declined, contemporaneous with an increase in articles using sequential model selection methods (@fig-timeSeries D).
+While there is evidence of improvement from earlier strata, goodness-of-fit testing and model validation using either in- or out-of-sample data remains uncommon even in the most recent strata (@fig-timeSeries E).
 
@@
 
-Discussion.
+Over the past two decades, dynamic occupancy models have been applied to an increasingly broad range of research contexts.
+As their popularity has grown and new tools have become available, authors have implemented DOMs with wide ranging amounts of data at small and large spatial and temporal scales using diverse data collection techniques.
+While each of these studies share the same underlying methodology, their approaches to implementation and interpretation vary considerably.
+The approach to building any type of model will necessarily depend on the character of the data available and on the priorities of the model-builder.
+This precludes any prescription of the 'best' way to build DOMs, however, there are still important discussions to be had on the most important decisions made throughout the modelling process.
+Based on the findings of our review, in this discussion we highlight areas which may merit closer consideration by both the users applying DOMs in practices as well as the quantitative researchers who research the models themselves.
+Where possible, we provide recommendations to current users of DOMs on ways to avoid common pitfalls based on existing research and lessons learned from other related models.
+Where uncertainties remain, we highlight key priorities for future research to clarify these points and support users in approaching DOMs with greater confidence.
+
+### Defining occupancy and detection
+
+A central challenge in occupancy modelling is identifying the correct interpretation of two interconnected processes: occupancy and detection.
+The flexibility of DOMs allows the definitions of a site, a primary occasion, and a secondary occasion to range widely, with implications for what occupancy and detection represent and how those quantities should be modelled.
+In the optimal case where a site is precisely aligned with a species' home range during the primary occasion, occupancy may be easily defined as the probability of a site containing the target species during the primary occasion.
+However, any deviation from this results in violation of the closure assumption and an altered definition of occupancy, with effects carrying over to the detection component of the model.
+When a site does not correspond to the home range of the target species, the definition of occupancy shifts to the probability that the species is *ever* within the site during a primary occasion, an interpration more akin to resource use than occupancy.
+Where this occurs, the detection probability no longer represents only the probability of observation during a secondary occasion, but also the probability that a species is present at the site during a secondary occasion [@efford2012].
+This latter definition of occupancy is often the case, particularly when using *ad hoc* data not collected under a survey protocol designed to ensure closure or when using autonomous detectors with limited spatial coverage [@wood2022].
+This definition of occupancy is still useful, but should be carefully considered to ensure an appropriate interpretation and avoid misrepresenting outputs.
+
+On a similar note, users of DOMs must carefully consider precisely what they are modelling and address questions on the scale represented by their model [@chave2013].
+Whether a site is occupied within a week or within a year will lead to vastly different conceptions of occupancy, likely with differences in the mechanisms determining whether sites are occupied at these time-scales.
+While data from surveys designed for dynamic occupancy modelling comes with a natural definition of the primary occasion, in other cases authors have some ability to determine the primary occasion themselves.
+This is particularly common with camera-trap or bio-acoustic data where continuous recordings can be broken down into distinct primary occasions of any length.
+While this offers unique opportunities to look at fine-scale changes in occupancy, it should be undertaken with great care to ensure correct interpretation of model outputs.
+Given the proliferation of continuous data types and the interest in DOMs for these applications, further research providing general guidance on determining appropriate primary and secondary occasion lengths for different purposes would be of value.
+
+When building DOMs for any reason, it is an imperative to establish a firm definition for both occupancy and detection that aligns with the context that the data was collected in and the life history traits of the target species.
+Doing so not only supports interpretation of the final model, but also helps clarify which covariates should be considered to explain heterogeneity in these processes.
+While this is a productive exercise for clarifying internal assumptions, we recommend that authors also include these definitions in their methodology to clarify interpretation for readers less familiar with the study system.
+
+### Covariates for model parameters
+
+The definitions of occupancy and detection should strongly inform the covariates considered for inclusion in DOMs.
+This is particularly pertinent for detectability, where the appropriate interpretation may differ from the popular conception of the term.
+Where occupancy does not comply with the closure assumption, detectability is an amalgamation of several distinct processes: resource use and abundance determine whether the species is even available for detection at the site during a secondary occasion, such that observation represents only a portion of detection [@guillera-arroita2017a].
+In some cases (e.g., acoustic monitoring for highly vocal species), use and abundance may be the dominant factors with observation playing a comparatively smaller role.
+In these situations, the relatively low number of environmental covariates considered for the detection parameter in many DOMs in our sample raises potential concerns.
+Failing to consider the environmental component of detection, rather than just the observation ones, is likely to lead to biased estimates of occupancy.
+We encourage authors to consider this broader definition of the detection parameter and, where appropriate, to consider additional environmental factors that may influence it.
+
+Covariates on the occupancy parameters (initial occupancy, colonisation, and extinction) should likewise be informed by the scale at which occupancy is measured and presumed to change.
+The temporal scale of occupancy is particularly important, as the underlying mechanisms which guide occupancy at different time-scales are likely to differ for many species.
+As colonisation and extinction are generally treated as first-order Markov processes, modelling transition probabilities should prioritise the factors most likely to be of importance in the intervals between primary occasions.
+Regarding initial occupancy, our review found that many studies did not include any covariates on this parameter — unless a study is conducted at very small extents or study sites are truly uniform in their suitability, one would expect some amount of non-random variation in occupancy probability across any study system that will not be captured when representing initial occupancy as a constant.
+Omission of the factors that drive occupancy is a known source of bias in static SDMs [@barry2006], which the initial occupancy component of DOMs conceptually resemble.
+Furthermore, any bias in occupancy estimation in the first time-step will perpetuate into future seasons due to DOMs Markovian nature, with implications for the reliability of model outputs.
+
+### Complexity and covariate responses
+
+One aspect of fitting DOMs meriting broader discussion centres on 'model complexity,' and how much must be incorporated into models to reliably model species occupancy under different contexts and use cases.
+Complexity is a broad term encompassing many aspects of a model [@merow2014], and opinions on simplicity versus complexity in ecological models are diverse.
+Where some advocate for the simplest possible models, arguing that they are most generalisable; others insist that overly-simple models cannot adequately represent the most important drivers in a system [@evans2013; @lonergan2014].
+By their nature the DOM is somewhat more complex than simpler models for studying occupancy due to their hierarchical structure, which is necessary to control for detectability and to capture occupancy dynamics.
+Within this structure, however, further complexity is to some degree up to the modeller: one can choose how many covariates to consider for inclusion on the various parameters, and how to represent the nature of the relationship between those covariates and parameters.
+
+Research from correlative species distribution models (SDMs) indicates that allowing for more complex relationships can improve model performance in predicting occupancy [@valavi2023], an increasingly popular use-case for the DOM.
+Many common approaches for SDMs, such as MAXENT and Boosted Regression Trees, permit considerable flexibility in the shape of their covariate response curves and use of interactions where supported by the data [@merow2013; @elith2008].
+Within the DOM framework, there are promising developments in methods for incorporating additional complexity, including @joseph2020's novel neural-network occupancy model which allows for exponentially higher levels of complexity, potentially offering improved performance for prediction-oriented studies.
+Further research on these methods may provide new options where predictive performance is a priority.
+
+The emphasis on more complex responses in SDMs may also be attributed to their frequent application across relatively large geographic extents which might encompass the full species niche, where environmental relationships may be expected to be non-linear.
+However, DOMs have also been implemented across large spatial extents where the same assumptions should exist and similar responses might be expected.
+When this is probable, modellers should ensure that covariate responses are ecologically realistic and appropriately reflect the hypothesised relationships [@austin2007].
+Similar consideration should be given to interactions between covariates, given that these relationships are commonly expected based on ecological theory and that their exclusion can negatively impact model performance [@guisan2006].
+
+### Model selection
+
+Limited research has been conducted on the advantages of different methods for covariate selection in DOMs, and there is unlikely to be a one-size-fits-all approach which will be appropriate for all possible use-cases.
+However, it is important to consider the implications of the different model selection approaches in common usage, and research from the SDM literature highlights how covariate selection can influence our interpretation of model outputs [@brodie2020].
+Within the occupancy modelling literature, @stewart2023's article on covariate selection in single-season occupancy models discusses important attributes of the information-criteria approaches widely used in frequentist models in our review, noting that these approaches can lead to inaccurate coefficient estimates that may influence model inference.
+This corresponds with the underlying theory behind AIC-based methods, which select models based on their predictive ability rather than any causal relationships.
+Authors seeking to estimate relationships with particular environmental factors should consider work from the causal modelling literature, which emphasises careful consideration of hypothesised relationships and the construction of directed acyclic graphs (DAGs; [@siegel2025]).
+In these situations, a comparatively smaller number of candidate models should be considered, rather than extended model selection procedures more appropriate for exploratory analyses or prediction [@nichols2024].
+
+Even where inference on covariate relationships is not a main objective, common sequential model selection approaches seem to be somewhat inconsistent in their outcomes.
+In one of the few studies on covariate selection in DOMs, @morin2020 found that the order parameters were fit in and the number of models carried through to each stage determined not only whether the optimal model was identified, but also which covariates were included.
+Where computing power is available, covariate selection methods which fit larger numbers of models may increase chances of identifying the best available model.
+In any case, additional research is necessary to determine whether there are more reliable approaches to covariate selection in maximum likelihood models.
+The literature on performance of Bayesian model selection methods is more sparse, although @hooten2015's guide to Bayesian model selection in ecology is a valuable resource for possible methods of fitting those models.
+
+### Model assessment
+
+Regardless of covariate selection protocol, it is important to determine whether a chosen model appropriately reflects the data and is suitable for its use case.
+Assessment of the performance of DOMs generally requires additional steps, and the best ways to achieve this are rife with uncertainty.
+Current goodness-of-fit tests for DOMs lack broad acceptance, although @mackenzie2004's approach for single-season occupancy models using a parametric bootstrap has been extended to DOMs.
+This test is implemented in the *AICcModAvg* and *unmarked* R packages.
+@kéry2021 describe the test and present an alternative based on separately assessing fit to static and dynamic components of the model.
+In Bayesian implementations, posterior predictive checks offer means to assess model fit [@gelman2014].
+@broms2016 discusses further possibilities for model evaluation in the Bayesian context focusing on single-season multi-species occupancy models, and extensions of their approach may also be applicable to the DOM.
+While further tests of these methods are necessary to fully assuage concerns around their use, current authors should continue to conduct and report goodness-of-fit tests using existing tools.
+
+When predictive performance is a research objective, it is essential to conduct some form of model evaluation to understand how reliable model outputs may be [@araújo2005; @guisan2005].
+As with other hierarchical models, model evaluation for DOMs can be difficult and somewhat uncertain compared to other model types, as the primary response variable of interest (species occupancy) is a latent variable where the true state is generally not known.
+Predictive performance evaluation is thus typically based on *observed* occupancy data, where a DOM is used to simulate detection histories to be compared with field survey results.
+While predictive performance should ideally be tested with independent datasets, this is generally not practiceable with DOMs.
+Best practice should involve setting aside a portion of data for testing final models, where this is not possible, cross-validation should be considered to attain an imperfect measure of performance.
+
++------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------+
+| Topic                              | Recommendations                                                                                                                                                                                          | Priorities                                                                                                                                |
++====================================+==========================================================================================================================================================================================================+===========================================================================================================================================+
+| Defining occupancy and detection   | -   Authors should explicitly define an interpretation of occupancy and detection probabilities consistent with the research context.
+| -   Identify guidelines for delineating primary and secondary occasions in continuous data.
+|
++------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------+
+| Covariates for model parameters    | -   Authors should consider the environmental aspects of detectability, and incorporate additional covariates where appropriate.
+| -   Assess how missing drivers of heterogeneity on one parameter affects estimates of the others, particularly with respect to detection.
+|
+|                                    |                                                                                                                                                                                                          |                                                                                                                                           |
+|                                    | -   Covariates for all parameters should be linked to the temporal scale of occupancy, particularly for transition probabilities.
+|                                                                                                                                           |
++------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------+
+| Complexity and covariate responses | -   Align covariate responses with hypothesised relationships, using polynomial responses where necessary.
+| -   Facilitate implementation of more complex DOMs for prediction, such as machine-learning based methods.
+|
++------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------+
+| Model selection                    | -   Be aware of pitfalls with AIC-based model selection methods.
+| -   Identify best approaches to model selection for various objectives while limiting computational costs.
+|
+|                                    |                                                                                                                                                                                                          |                                                                                                                                           |
+|                                    | -   Consider more constrained candidate models when inference on covariate relationships is an objective.
+| -   Compare performance of emerging Bayesian methods for model selection.
+|
++------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------+
+| Model assessment                   | -   Available goodness-of-fit tests should be conducted and reported in all cases.
+| -   Existing goodness-of-fit tests should be further validated to ease concerns regarding their use.
+|
+|                                    |                                                                                                                                                                                                          |                                                                                                                                           |
+|                                    | -   When prediction is an objective, data should be reserved for out-of-sample validation wherever possible.
+If not feasible, in-sample validation should be performed and performance metrics reported.
+| -   Additional research on methods for performance evaluation in hierarchical models is necessary.
+|
++------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------+
+
+: A summary of recommendations for current users of DOMs, with priorities for further research.
 
@@
 
-Conclusions.
+In the two decades since the publication of @mackenzie2003 dynamic occupancy models have become increasingly important tools for analysing ecological data.
+These models have seen increasingly applied use, tackling challenges which include assessing critically endangered species, guiding public health management of disease vectors, and tracking rapidly developing biological invasions [@carvalho2023; @moreira2016; @wood2020].
+The DOM is well suited to these situations, and it is to be expected that as these types of applications are more commonly attempted understanding the sensitivity of model outputs to decisions made in the model fitting process becomes increasingly important.
+To this end, guidance and tools are necessary to support model users in aligning their data, model, and objectives to produce the most reliable occupancy estimates possible.
+Parallels to existing work in the SDM literature on guidelines for reporting and modelling standards could be valuable contributions in achieving these aims and provide useful examples in developing methods and guidelines specifically focused on DOMs [@zurell2020; @araújo2019].

### 2025-02-03 Manuscript_Initial.qmd
Completed first-pass revisions for Ecography

@@
 Many models incorporate covariates representing aspects of site geometry and connectivity (35% of studies).
-Often these are included as simple covariates, as in @duggan2011's models using landcsape connectivety metrics as covariates on colonisation and extinction.
+Often these are included as simple covariates, as in @duggan2011's models using landscape connectivity metrics as covariates on colonisation and extinction.
 Alternatively, more complex parameterisations explicitly model colonisation or extinction as a spatial process, blending attributes of DOMs and stochastic patch occupancy models [@broms2016; @risk2011].

### 2025-02-07 Analysis.qmd
Updates to review summary sub-plot code

@@
   ggplot() +
-  geom_sf(data = countries_shp, colour = "gray90", fill = "gray80") +
-  geom_sf(shape = 21, fill = "aquamarine4", colour = "white", alpha = 0.5, size = 3.5) +
+  geom_sf(data = countries_shp, colour = "gray90", fill = "gray60", linewidth = 0.1) +
+  geom_sf(shape = 23, 
+          fill = "goldenrod2", colour = "white",
+          size = 1.5) +
   theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
@@
 plot_studyAreaMap
-ggsave(plot = plot_studyAreaMap, "Figures/StudiesMap.png")
+
+ggsave("Figures/StudiesMap.eps", plot = plot_studyAreaMap,
+       width = 85, height = 50, units = "mm", bg = "transparent")
 
@@
 ```{r Study area size}
-#| fig-width: 7
-#| fig-height: 4
+#| fig-width: "6.52cm"
+#| fig-height: "3.7cm"
 
@@
   geom_col(aes(x = Size, y = Articles),
-           fill = "aquamarine4", colour = "white",
-           width = 1, linewidth = 2) +
+           fill = "#4D7668", colour = "white",
+           width = 1, linewidth = .5) +
   labs(x = "", y = "") +
@@
   theme(panel.grid = element_blank(),
-        axis.text.x = element_markdown(size = 20, angle = 90,
-                                       vjust = 0.5, hjust = 1),
-        axis.text.y = element_blank(), axis.ticks = element_blank(),
+        axis.text.x = element_markdown(size = 8, angle = 90,
+                                       vjust = 0.5, hjust = 1,
+                                       family = "Helvetica"),
+        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
         panel.background = element_rect(fill='transparent', color=NA),
@@
 
-ggsave("Figures/StudyAreaPlot.png",
-       plot = plot_studyAreaSize, width = 7, height = 4, bg='transparent')
+ggsave("Figures/StudyAreaPlot.eps",
+       plot = plot_studyAreaSize, 
+       width = 80, height = 45, units = "mm",
+       bg='transparent')
 ```
@@
            position = position_dodge2(reverse = TRUE), width = 0.8) +
-  scale_fill_manual("", values = c("aquamarine4", "goldenrod2", "gray70")) +
-  scale_x_continuous(expand = c(0,0)) +
+  scale_fill_manual("", 
+                    values = c("#4D7668", "goldenrod3", "gray40"),
+                    limits = c("Overall", "Threatened", "Invasive"),
+                    labels = c("Total", "Threatened", "Invasive")) +
+  scale_x_continuous(expand = c(0,0), breaks = c(0, 15, 30, 45)) +
   labs(x = "Articles", y = "") +
-  theme(panel.grid = element_blank(),
-        legend.text = element_text(size = 20), axis.text = element_text(size = 21),
-        axis.title = element_text(size = 19), axis.ticks.y = element_blank(),
+  theme(#legend.position = "top",
+        panel.grid = element_blank(),
+        legend.text = element_text(size = 8, family = "Helvetica", colour = "#585858"),
+        legend.title = element_text(size = 8, family = "Helvetica", colour = "#585858"),
+        axis.text = element_text(size = 8, family = "Helvetica", colour = "#585858"),
+        axis.title = element_text(size = 8, family = "Helvetica", colour = "#585858"), 
+        axis.ticks.y = element_blank(),
         panel.background = element_rect(fill='transparent', color = NA),
         plot.background = element_rect(fill='transparent', color = NA),
-        legend.background = element_rect(fill = 'transparent', color = NA))
+        legend.background = element_rect(fill = 'transparent'),
+        legend.key.size = unit(4, 'mm'),
+        legend.margin=margin(0,0,0,0),
+        legend.box.margin=margin(0, -5, 0, 0),
+        axis.line.x = element_line(colour = "#585858", linewidth = 0.3))
 plot_focalTaxa
 
-ggsave("Figures/TaxaPlot.png",
-       plot_focalTaxa, width = 10, height = 4, bg='transparent')
+ggsave("Figures/TaxaPlot.eps",
+       plot_focalTaxa, width = 90, height = 40, units = "mm",
+       bg = "transparent")
 ```
@@
   geom_histogram(aes(x = `Site quantity`),
-                 binwidth = 0.25, fill = "aquamarine4", colour = "white",
-                 linewidth = 2) +
+                 bins = 12,
+                 fill = "#4D7668", colour = "white",
+                 linewidth = 0.5) +
   geom_vline(aes(xintercept = median(`Site quantity`, na.rm = TRUE)), 
@@
   scale_y_continuous(expand = c(0,0), position = "right") +
-  labs(y = "Site\nquantity", x = "") +
+  labs(y = "Number of\nsites", x = "") +
   theme(panel.grid = element_blank(),
-        panel.background = element_rect(fill='transparent', color=NA),
-        plot.background = element_rect(fill='transparent', color=NA),
-        axis.text.x = element_text(size = 23, angle = 90,
-                                   vjust = 0.5, hjust = 1),
-        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
-        axis.line.x = element_line(), 
-        axis.title.y = element_text(size = 35, colour = "aquamarine4"))
+        panel.background = element_rect(fill = 'transparent',
+                                        color = NA),
+        plot.background = element_rect(fill = 'transparent', 
+                                       color = NA),
+        axis.text.x = element_text(size = 8, angle = 60,
+                                   #vjust = .5,
+                                   hjust = 1,
+                                   colour = "#585858",
+                                   family = "Helvetica"),
+        axis.text.y = element_blank(), 
+        axis.ticks.y = element_blank(), 
+        plot.margin = margin(20,10,0,10),
+        axis.line.x = element_line(colour = "#585858",
+                                   linewidth = .1), 
+        axis.title.y = element_text(size = 10, 
+                                    colour = "#585858",
+                                   family = "Helvetica"))
 plot_siteQuantity
-ggsave(plot = plot_siteQuantity, filename = "Figures/Sites.png",
-       height = 3, width = 12)
+ggsave(plot = plot_siteQuantity, 
+       filename = "Figures/Sites.eps",
+       height = 35, width = 110, units = "mm")
 
@@
   geom_histogram(aes(x = StudyMonths), 
-                 binwidth = 0.2, fill = "aquamarine4", colour = "white",
-                 linewidth = 2) +
+                 bins = 12,
+                 fill = "#4D7668", colour = "white",
+                 linewidth = 0.5) +
   geom_vline(aes(xintercept = median(StudyMonths, na.rm = TRUE)), 
@@
   theme(panel.grid = element_blank(),
-        panel.background = element_rect(fill='transparent', color=NA),
-        plot.background = element_rect(fill='transparent', color=NA),
-        axis.text.x = element_text(size = 23, angle = 90,
-                                   vjust = 0.5, hjust = 1),
-        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
-        axis.line.x = element_line(), 
-        axis.title.y = element_text(size = 35, colour = "aquamarine4"))
+        panel.background = element_rect(fill = 'transparent', 
+                                        color = NA),
+        plot.background = element_rect(fill = 'transparent',
+                                       color = NA),
+        axis.text.x = element_text(size = 8, angle = 60,
+                                   hjust = 1, 
+                                   family = "Helvetica",
+                                   colour = "#585858"),
+        axis.text.y = element_blank(), 
+        axis.ticks.y = element_blank(),
+        axis.line.x = element_line(colour = "#585858",
+                                   linewidth = .1), 
+        axis.title.y = element_text(size = 10, 
+                                    colour = "#585858",
+                                    family = "Helvetica"),
+        plot.margin = margin(20,10,0,10))
 plot_studyDuration
-ggsave(plot = plot_studyDuration, filename = "Figures/Duration.png",
-       height = 3.5, width = 12)
+ggsave(plot = plot_studyDuration, 
+       filename = "Figures/Duration.eps",
+       height = 35, width = 110, units = "mm",
+       bg = "transparent")
 

### 2025-02-17 Analysis.qmd
Figure detail work

@@
                                "Estimated qualifying articles"),
-                    values = c("aquamarine4", "gray65")) +
+                    values = c("#4D7668", "gray50")) +
   scale_y_continuous(expand = c(0,0), limits = c(0, 65)) +
+  scale_x_continuous(#limits = seq(2004, 2023, 1),
+                     breaks = seq(2004, 2023, 4),
+                     labels = seq(2004, 2023, 4)) +
   labs(x = "", y = "Number of articles") +
@@
         panel.grid.minor.x = element_blank(),
-        legend.position = "bottom", 
-        axis.text = element_text(size = 12),
-        axis.text.x = element_blank(), 
-        axis.ticks.x = element_blank(),
-        axis.title = element_text(size = 12), 
-        legend.text = element_text(size = 12),
-        strip.text.x = element_text(size = 12)) +
-  facet_grid(cols = vars(Strata), scales = "free_x", switch = "x")
+        panel.grid.minor.y = element_blank(),
+        legend.margin=margin(0,0,0,0),
+        legend.box.margin=margin(-10,-10,-5,-10),
+        legend.position = "top", 
+        legend.key.size = unit(10, 'points'),
+        legend.direction = "vertical",
+        axis.text.y = element_text(size = 8, family = "helvetica", colour = "#585858"),
+        axis.text.x = element_text(size = 8, family = "helvetica", 
+                                   colour = "#585858", angle = 60,
+                                   vjust = 1, hjust = 1), 
+        axis.title = element_text(size = 8, family = "helvetica", colour = "#585858"), 
+        legend.text = element_text(size = 8, family = "helvetica", colour = "#585858"),
+        strip.text.x = element_text(size = 8, family = "helvetica", colour = "#585858"))
+#+ facet_grid(cols = vars(Strata), scales = "free_x", switch = "x")
 plot_reviewCoverage
-ggsave(plot = plot_reviewCoverage, "Figures/CoveragePlot.jpeg",
-       width = 8, height = 4)
+ggsave(plot = plot_reviewCoverage, "Figures/CoveragePlot.png",
+       width = 945, height = 700, units = "px")
 ```
@@
   summarise(AffilCount = n(),
-            .by = nPer) (92-34)/92
+            .by = nPer)
 
@@
   ggplot(aes(x = Category, y = Covariates)) +
-  geom_jitter(aes(colour = Category, shape = Outlier), 
-              alpha = 0.3, size = 2.5,
-              width = 0.25, height = 0) +
   geom_boxplot(aes(fill = Category), 
-               outliers = FALSE, alpha = 0.3, colour = "gray60") +
+               outliers = TRUE, linewidth = .3, staplewidth = 0.33,
+               colour = "gray25") +
   scale_fill_manual("", labels = c("Environmental covariates", "Structural covariates"),
-                    values = c("aquamarine4", "plum4")) +
+                    values = c("#4D7668", "plum4")) +
   scale_colour_manual("", labels = c("Environmental covariates", 
                                      "Structural covariates"),
-                      values = c("aquamarine4", "plum4")) +
-  scale_shape_manual("", labels = c("",""), 
-                     limits = c(TRUE, FALSE), values = c(17, 16)) +
+                      values = c("#4D7668", "plum4")) +
   scale_y_continuous(breaks = c(0, 5, 10, 15),
                      labels = c("0", "5", "10", "15+")) +
-  labs(x = "", y = "Number of covariates considered") +
+  labs(x = "", y = "Number of covariates\nconsidered") +
   facet_grid(cols = vars(Parameter), switch = "x") +
   guides(shape = "none") +
-  theme(legend.position = "bottom", 
+  theme(legend.position = "top",
+        legend.box.margin = margin(-5,-10,-10,-10),
+        legend.key.size = unit(12.5, 'points'),
         axis.text.x = element_blank(), axis.ticks.x = element_blank(),
@@
         panel.grid.minor.x = element_blank(), 
-        strip.text = element_text(size = 12, family = "Helvetica"), 
-        axis.title.y = element_text(size = 13, family = "Helvetica"), 
-        axis.text = element_text(size = 12, family = "Helvetica"),
-        legend.text = element_text(size = 12, family = "Helvetica"))
+        panel.grid.minor.y = element_blank(),
+        #panel.background = element_rect(fill = "gray80"),
+        strip.text = element_text(size = 7.5, family = "Helvetica", colour = "#585858"), 
+        axis.title.y = element_text(size = 8, family = "Helvetica", colour = "#585858"), 
+        axis.text = element_text(size = 8, family = "Helvetica", colour = "#585858"),
+        legend.text = element_text(size = 8, family = "Helvetica", colour = "#585858"))
 
 plot_covariateQuantity
-ggsave(plot = plot_covariateQuantity, filename = "Figures/CovParamPlot.jpeg",
-       width = 8, height = 4)
+ggsave(plot = plot_covariateQuantity, filename = "Figures/CovParamPlot.png",
+       width = 1476, height = 700, units = "px")
 ```
@@
 
-ggsave(plot = covariatePatchwork, "Figures/CovariateSummary.png",
+ggsave(plot = covariatePatchwork, "Figures/CovariateSummary.png")
 ```

### 2025-02-17 Analysis.qmd
Update method for covariate summary

@@
             .by = c(Covariate, Category)) |>
-  mutate(Included = TRUE) %>%
-  rbind(mutate(., PctCategory = 100-PctCategory, Included = FALSE)) |>
+  mutate(Included = "TRUE") %>%
+  rbind(mutate(., PctCategory = 100-PctCategory, Included = "FALSE")) |>
   mutate(Covariate = case_when(Covariate == "Habitat" ~ "Habitat",
@@
   mutate(Covariate = fct(Covariate,
-                         levels = c("Any", "Any\nEnv.",
+                         levels = rev(c("Any", "Any\nEnv.",
                                     "Habitat", "Spatial", "Climate",
@@
                                     "Any\nStruct.", "Primary", "Observation",
-                                    "Secondary", "Site", "Species"))) |>
+                                    "Secondary", "Site", "Species")))) |>
   mutate(Category = case_when(Category == "ANY" ~ "Any",
@@
                               Category == "S" ~ "Structural")) |>
-
-  ggplot(aes(x = Covariate, y = PctCategory, fill = Included)) +
+  mutate(Cat_Included = paste0(Category, Included)) |>
+
+  ggplot(aes(y = Covariate, x = PctCategory, fill = Cat_Included)) +
   geom_col(colour = "white", 
            linewidth = 1, width = 1) +
-  scale_fill_manual(limits = c(TRUE, FALSE), 
-                    values = c("goldenrod", "gray50")) +
-  scale_y_continuous(expand = c(0,0)) +
-  facet_grid(cols = vars(Category), scales = "free_x", space = "free_x") +
-  labs(y = "Percent\nof models", x = "") +
+  scale_fill_manual(limits = c("AnyTRUE", "EnvironmentalTRUE", "StructuralTRUE",
+                               "AnyFALSE", "EnvironmentalFALSE", "StructuralFALSE"), 
+                    values = c("goldenrod", "#4D7668", "plum4",
+                               "gray80", "gray80", "gray80")) +
+  scale_x_continuous(expand = c(0,0), breaks = c(25,50,75)) +
+  facet_grid(rows = vars(Category), 
+             scales = "free_y", space = "free_y", switch = "y") +
+  labs(x = "", y = "", title = "Percent of\nmodels considering:") +
   theme(legend.position = "none",
@@
         panel.grid = element_blank(),
-        axis.text.x = element_text(family = "Helvetica", 
-                                   angle = 90, vjust = 0.5, hjust = 1),
-        strip.text.x = element_text(family = "Helvetica"))
+        axis.text.y = element_text(family = "Helvetica"),
+                                   #angle = 90, vjust = 0.5, hjust = 1),
+        strip.text.y = element_text(family = "Helvetica"),
+        strip.placement = "outside")
 covariateFreq
@@
             .by = c(Covariate, Category)) |>
-  mutate(Dyn = TRUE) %>%
-  rbind(mutate(., PctDyn = 100-PctDyn, Dyn = FALSE)) |>
+  mutate(Dyn = "TRUE") %>%
+  rbind(mutate(., PctDyn = 100-PctDyn, Dyn = "FALSE")) |>
   mutate(Covariate = case_when(Covariate == "Habitat" ~ "Habitat",
@@
                               Category == "S" ~ "Structural")) |>
-
-  ggplot(aes(x = Covariate, y = PctDyn, fill = Dyn)) +
+  mutate(Cat_Dyn = paste0(Category, Dyn)) |>
+
+  ggplot(aes(y = Covariate, x = PctDyn, fill = Cat_Dyn)) +
   geom_col(colour = "white", 
            linewidth = 1, width = 1) +
-  scale_fill_manual(limits = c(TRUE, FALSE), 
-                    values = c("aquamarine3", "gray50")) +
-  scale_y_continuous(expand = c(0,0)) +
-  facet_grid(cols = vars(Category), scales = "free_x", space = "free_x") +
-  labs(y = "Percent\ndynamic", x = "") +
+  scale_fill_manual(limits = c("AnyTRUE", "EnvironmentalTRUE", "StructuralTRUE",
+                               "AnyFALSE", "EnvironmentalFALSE", "StructuralFALSE"), 
+                    values = c("goldenrod", "#4D7668", "plum4",
+                               "gray80", "gray80", "gray80")) +
+  scale_x_continuous(expand = c(0,0),  breaks = c(25,50,75)) +
+  facet_grid(rows = vars(Category), scales = "free_y", 
+             space = "free_y", switch = "y") +
+  labs(x = "", y = "", title = "Average rate\ndynamic:") +
   theme(legend.position = "none",
@@
         panel.grid = element_blank(),
-        axis.text.x = element_text(family = "Helvetica", 
-                                   angle = 90, vjust = 0.5, hjust = 1),
-        strip.text.x = element_blank(), strip.background = element_blank())
+        axis.text.y = element_text(family = "Helvetica"), 
+                                   #angle = 90, vjust = 0.5, hjust = 1),
+        strip.text.y = element_blank(), 
+        strip.background = element_blank(),
+        strip.placement = "outside")
 covariateDyn
@@
             .by = c(Covariate, Category)) |>
-  mutate(Obs = TRUE) %>%
-  rbind(mutate(., PctObs = 100-PctObs, Obs = FALSE)) |>
+  mutate(Obs = "TRUE") %>%
+  rbind(mutate(., PctObs = 100-PctObs, Obs = "FALSE")) |>
   mutate(Covariate = case_when(Covariate == "Habitat" ~ "Habitat",
@@
                               Category == "S" ~ "Structural")) |>
-
-  ggplot(aes(x = Covariate, y = PctObs, fill = Obs)) +
+  mutate(Cat_Obs = paste0(Category, Obs)) |>
+
+  ggplot(aes(y = Covariate, x = PctObs, fill = Cat_Obs)) +
   geom_col(colour = "white", 
            linewidth = 1, width = 1) +
-  scale_fill_manual(limits = c(TRUE, FALSE), 
-                    values = c("cadetblue3", "gray50")) +
-  scale_y_continuous(expand = c(0,0)) +
-  facet_grid(cols = vars(Category), scales = "free_x", space = "free_x") +
-  labs(y = "Percent directly\nobserved", x = "") +
+  scale_fill_manual(limits = c("AnyTRUE", "EnvironmentalTRUE", "StructuralTRUE",
+                               "AnyFALSE", "EnvironmentalFALSE", "StructuralFALSE"), 
+                    values = c("goldenrod", "#4D7668", "plum4",
+                               "gray80", "gray80", "gray80")) +
+  scale_x_continuous(expand = c(0,0),  breaks = c(25,50,75)) +
+  facet_grid(rows = vars(Category), scales = "free_y", 
+             space = "free_y", switch = "y") +
+  labs(x = "", y = "", title = "Average rate\ndirectly observed:") + 
   theme(legend.position = "none",
@@
         panel.grid = element_blank(),
-        axis.text.x = element_text(family = "Helvetica", 
-                                   angle = 90, vjust = 0.5, hjust = 1),
-        strip.text.x = element_blank(), strip.background = element_blank())
+        axis.text.y = element_text(family = "Helvetica"), 
+                                   #angle = 90, vjust = 0.5, hjust = 1),
+        strip.text.y = element_blank(), 
+        strip.background = element_blank(),
+        strip.placement = "outside")
 covariateObs
@@
             .by = c(Covariate, Category)) |>
-  mutate(Complex = TRUE) %>%
-  rbind(mutate(., PctComplex = 100-PctComplex, Complex = FALSE)) |>
+  mutate(Complex = "TRUE") %>%
+  rbind(mutate(., PctComplex = 100-PctComplex, Complex = "FALSE")) |>
   mutate(Covariate = case_when(Covariate == "Habitat" ~ "Habitat",
@@
                               Category == "S" ~ "Structural")) |>
-
-  ggplot(aes(x = Covariate, y = PctComplex, fill = Complex)) +
+  mutate(Cat_Complex = paste0(Category, Complex)) |>
+
+  ggplot(aes(y = Covariate, x = PctComplex, fill = Cat_Complex)) +
   geom_col(colour = "white", 
            linewidth = 1, width = 1) +
-  scale_fill_manual(limits = c(TRUE, FALSE), 
-                    values = c("plum3", "gray50")) +
-  scale_y_continuous(expand = c(0,0)) +
-  facet_grid(cols = vars(Category), scales = "free_x", space = "free_x") +
-  labs(y = "Percent nonlinear\nor interactive", x = "") +
+  scale_fill_manual(limits = c("AnyTRUE", "EnvironmentalTRUE", "StructuralTRUE",
+                               "AnyFALSE", "EnvironmentalFALSE", "StructuralFALSE"), 
+                    values = c("goldenrod", "#4D7668", "plum4",
+                               "gray80", "gray80", "gray80")) +
+  scale_x_continuous(expand = c(0,0), breaks = c(25,50,75)) +
+  facet_grid(rows = vars(Category), scales = "free_y",
+             space = "free_y", switch = "y") +
+  labs(x = "", y = "", title = "Average rate non-linear\nor interactive:") +
   theme(legend.position = "none",
@@
         panel.grid = element_blank(),
-        axis.text.x = element_text(family = "Helvetica", 
-                                   angle = 90, vjust = 0.5, hjust = 1),
-        strip.text.x = element_blank(), strip.background = element_blank())
+        axis.text.y = element_text(family = "Helvetica"), 
+                                   #angle = 90, vjust = 0.5, hjust = 1),
+        strip.text.y = element_blank(), 
+        strip.background = element_blank(),
+        strip.placement = "outside")
 covariateComplex
@@
 ```{r Covariate_Patchwork}
-covariatePatchwork <- covariateFreq / covariateDyn / covariateObs / covariateComplex +
-  plot_layout(axes = "collect") &
-  theme(axis.text = element_text(family = "Helvetica"),
-        axis.title = element_text(family = "Helvetica"))
+covariatePatchwork <- covariateFreq + covariateDyn + covariateObs + covariateComplex +
+  plot_layout(ncol = 4, axes = "collect") &
+  theme(axis.text = element_text(size = 8, family = "Helvetica", colour = "#585858",
+                                 face = "italic"),
+        axis.title = element_text(size = 8, family = "Helvetica", colour = "#585858"),
+        plot.title = element_text(size = 8, family = "Helvetica", 
+                                  colour = "#585858", hjust = 0.5)) 
 covariatePatchwork
 
-ggsave(plot = covariatePatchwork, "Figures/CovariateSummary.png")
+ggsave(plot = covariatePatchwork, "Figures/CovariateSummary.png",
+       width = 1961, height = 1600, units = "px")
 ```

### 2025-03-04 Analysis.qmd
Major changes to MS over the past two weeks prior to resubmission

@@
         legend.direction = "vertical",
-        axis.text.y = element_text(size = 8, family = "helvetica", colour = "#585858"),
-        axis.text.x = element_text(size = 8, family = "helvetica", 
+        axis.text.y = element_text(size = 10, family = "helvetica", colour = "#585858"),
+        axis.text.x = element_text(size = 10, family = "helvetica", 
                                    colour = "#585858", angle = 60,
                                    vjust = 1, hjust = 1), 
-        axis.title = element_text(size = 8, family = "helvetica", colour = "#585858"), 
-        legend.text = element_text(size = 8, family = "helvetica", colour = "#585858"),
-        strip.text.x = element_text(size = 8, family = "helvetica", colour = "#585858"))
-#+ facet_grid(cols = vars(Strata), scales = "free_x", switch = "x")
+        axis.title = element_text(size = 10, family = "helvetica", colour = "#585858"), 
+        legend.text = element_text(size = 10, family = "helvetica", colour = "#585858"),
+        strip.text.x = element_text(size = 10, family = "helvetica", colour = "#585858"))
 plot_reviewCoverage
-ggsave(plot = plot_reviewCoverage, "Figures/CoveragePlot.png",
+ggsave(plot = plot_reviewCoverage, "Figures/1_CoveragePlot.png",
        width = 945, height = 700, units = "px")
@@
 
-ggsave("Figures/StudiesMap.eps", plot = plot_studyAreaMap,
+ggsave("Figures/3a_StudyMap.eps", plot = plot_studyAreaMap,
        width = 85, height = 50, units = "mm", bg = "transparent")
@@
                                        vjust = 0.5, hjust = 1,
-                                       family = "Helvetica"),
+                                       family = "Helvetica", colour = "#585858"),
         axis.text.y = element_blank(), axis.ticks.y = element_blank(),
@@
 
-ggsave("Figures/StudyAreaPlot.eps",
+ggsave("Figures/3b_StudyArea.eps",
        plot = plot_studyAreaSize, 
@@
 
-ggsave("Figures/TaxaPlot.eps",
-       plot_focalTaxa, width = 90, height = 40, units = "mm",
+ggsave("Figures/3c_Taxa.eps",
+       plot_focalTaxa, width = 95, height = 40, units = "mm",
        bg = "transparent")
@@
 ggsave(plot = plot_siteQuantity, 
-       filename = "Figures/Sites.eps",
+       filename = "Figures/3e_Sites.eps",
        height = 35, width = 110, units = "mm")
@@
 ggsave(plot = plot_studyDuration, 
-       filename = "Figures/Duration.eps",
+       filename = "Figures/3e_Duration.eps",
        height = 35, width = 110, units = "mm",
@@
 # Calculate sample size
-stat_studyDurationN <- sheet_DataCollection |>
+stat_studyDuration <- sheet_DataCollection |>
   filter(`Start month` != "Unclear" & `End month` != "Unclear") |>
   nrow()
-stat_studyDurationN
-```
-
-We combine the study duration figure and the site quantity figure using `patchwork`.
-
-```{r combine sites and duration}
-#| fig-width: 12
-#| fig-height: 7.5
-
-plot_combinedStudyScale <- plot_siteQuantity / plot_spacer() / plot_studyDuration
-plot_combinedStudyScale <- plot_combinedStudyScale +
-  theme(panel.background = element_rect(fill='transparent', color = NA),
-        plot.background = element_rect(fill='transparent', color = NA))
-plot_combinedStudyScale
-ggsave(plot = plot_combinedStudyScale, filename = "Figures/StudyScale.png",
-       height = 7, width = 12, bg='transparent')
+stat_studyDuration
+```
+
+#### Primary occasions
+
+```{r Primary occasions}
+stat_medianPrimary <- sheet_DataCollection |>
+  pull(`Primary occasions`) |>
+  median()
+stat_medianPrimary
+
+stat_primaryRange <- sheet_DataCollection |>
+  pull(`Primary occasions`) |>
+  range()
+stat_primaryRange
 ```
@@
 ```{r Covariate quantity figure}
+
 plot_covariateQuantity <- sheet_Covariates |> 
@@
                                       `Parameter type` == "Occupancy" ~ "Occupancy*",
+                                      `Parameter type` == "Initial occupancy" ~ "Initial\noccupancy",
                                       .default = `Parameter type`)) |>
   mutate(Parameter = fct(`Parameter type`, 
-                         levels = c("Initial occupancy", "Occupancy*",
+                         levels = c("Initial\noccupancy", "Occupancy*",
                                     "Colonisation", "Extinction",
@@
         #panel.background = element_rect(fill = "gray80"),
-        strip.text = element_text(size = 7.5, family = "Helvetica", colour = "#585858"), 
-        axis.title.y = element_text(size = 8, family = "Helvetica", colour = "#585858"), 
-        axis.text = element_text(size = 8, family = "Helvetica", colour = "#585858"),
-        legend.text = element_text(size = 8, family = "Helvetica", colour = "#585858"))
+        strip.text = element_text(size = 10, family = "Helvetica", colour = "#585858"), 
+        axis.title.y = element_text(size = 10, family = "Helvetica", colour = "#585858"), 
+        axis.text = element_text(size = 10, family = "Helvetica", colour = "#585858"),
+        legend.text = element_text(size = 10, family = "Helvetica", colour = "#585858"))
 
 plot_covariateQuantity
-ggsave(plot = plot_covariateQuantity, filename = "Figures/CovParamPlot.png",
+ggsave(plot = plot_covariateQuantity, filename = "Figures/5_CovariateCount.png",
        width = 1476, height = 700, units = "px")
@@
 
-write_csv(table_covariateSummary, "Figures/CovariateTable.csv")
+write_csv(table_covariateSummary, "Figures/S1_CovariateTable.csv")
 ```
@@
 
-write_csv(modelling_summary, "Figures/ModellingSummary.csv")
+write_csv(modelling_summary, "Figures/T2_ModellingSummary.csv")
 ```
@@
              space = "free_y", switch = "y") +
-  labs(x = "", y = "", title = "Average rate non-linear\nor interactive:") +
+  labs(x = "", y = "", title = "Average rate\nnon-linear or interactive:") +
   theme(legend.position = "none",
@@
   plot_layout(ncol = 4, axes = "collect") &
+  #plot_annotation(tag_levels = 'a', tag_prefix = '(', tag_suffix = ")") &
   theme(axis.text = element_text(size = 8, family = "Helvetica", colour = "#585858",
@@
         plot.title = element_text(size = 8, family = "Helvetica", 
-                                  colour = "#585858", hjust = 0.5)) 
+                                  colour = "#585858", hjust = 0.5),
+        plot.tag = element_text(size = 9, family = "Helvetica", colour = "#585858")) 
 covariatePatchwork
 
-ggsave(plot = covariatePatchwork, "Figures/CovariateSummary.png",
-       width = 1961, height = 1600, units = "px")
+#ggsave(plot = covariatePatchwork, "Figures/4_CovariateSummary.png",
+#       width = 1961, height = 1600, units = "px")
 ```
@@
                     values = c("plum4", "cadetblue3", "aquamarine3",
-                               "aquamarine4", "goldenrod2", "goldenrod3")) +
+                               "#4D7668", "goldenrod", "goldenrod4")) +
   scale_y_continuous(breaks = c(0, 20, 40, 60, 80),
@@
         panel.grid.minor.x = element_blank(),
-        legend.text = element_text(size = 15),
-        axis.text = element_text(size = 15), axis.title = element_text(size = 15),
-        legend.position = "bottom", strip.background = element_blank(),
+        panel.grid.minor.y = element_blank(),
+        legend.text = element_text(size = 9, family = "helvetica", colour = "#585858"),
+        axis.text = element_text(size = 9, family = "helvetica", colour = "#585858"), 
+        axis.title = element_text(size = 9, family = "helvetica", colour = "#585858"),
+        legend.box.margin = margin(-5,-10,-10,-10),
+        legend.key.size = unit(12.5, 'points'),
+        legend.position = "top", 
+        strip.background = element_blank(),
         strip.text = element_blank()) +
@@
 plot_objectiveTrend
-ggsave(plot = plot_objectiveTrend, "Figures/ObjectiveTrend.jpeg",
-       width = 8, height = 4)
-```
-
-We render a second figure for the average number of covariates used in articles approaching each objective
-
-```{r Covariates per objective}
-#| fig-width: 8
-#| fig-height: 4
-
-covsPerParam <- sheet_Covariates |>
-  mutate(studyID = paste(`Review ID`, `Model ID`, sep = "_")) |>
-  separate_longer_delim(Covariates, delim = ", ") |>
-  filter(`Parameter type` %in% c("Initial occupancy", "Occupancy",
-                                 "Colonisation", "Extinction_Persistence",
-                                 "Detection")) |>
-  mutate(n_params = length(unique(`Parameter type`)),
-         .by = c(studyID, `Review ID`)) |>
-  filter(Covariates != "None") |>
-  summarise(CovsPerParam = n()/max(n_params), 
-            .by = c(studyID, `Review ID`))
-
-plot_objectiveCovariates <- sheet_Objectives |>
-  pivot_longer(c("Demonstrate methods?", "Estimate trends?", 
-                 "Identify drivers?", "Test relationships?",
-                 "Predict spatially?", "Predict temporally?"),
-               names_to = "Objective", values_to = "ObjectiveIncluded") |>
-  mutate(Objective = str_replace(Objective, "\\?", "")) |>
-  filter(ObjectiveIncluded == "YES") %>%
-  rbind(., mutate(., Objective = "Any Objective")) |>
-  select(`Review ID`, Objective) |>
-  distinct() |>
-  left_join(covsPerParam, by = join_by(`Review ID`), relationship = "many-to-many") |>
-  mutate(CovsPerParam = case_when(is.na(CovsPerParam) ~ 0,
-                                  .default = CovsPerParam)) |>
-  mutate(Objective = factor(Objective, levels = c("Any Objective", 
-                                                  "Demonstrate methods",
-                                                  "Estimate trends", 
-                                                  "Identify drivers", 
-                                                  "Test relationships",
-                                                  "Predict spatially",
-                                                  "Predict temporally"))) |>
-  mutate(Objective = fct_relabel(Objective, 
-                                 ~ str_replace(.x, " ", "\n"))) |>
-  mutate(facet = case_when(Objective == "Any\nObjective" ~ "left",
-                           .default = "right")) |>
-  mutate(CovsPerParam = case_when(CovsPerParam > 15 ~ 15,
-                                  .default = CovsPerParam)) |>
-  mutate(Outlier = case_when(CovsPerParam == 15 ~ TRUE,
-                             .default = FALSE)) |>
-
-  ggplot(aes(x = Objective, y = CovsPerParam, 
-             fill = Objective, colour = Objective)) +
-  geom_boxplot(colour = "gray50", alpha = 0.6, outliers = FALSE) +
-  geom_jitter(aes(shape = Outlier), alpha = 0.4, width = 0.3, size = 2.5) +
-  scale_y_continuous(breaks = c(0, 5, 10, 15),
-                     labels = c("0", "5", "10", "15+")) +
-  scale_fill_manual(limits = c("Any\nObjective", "Demonstrate\nmethods", 
-                               "Estimate\ntrends", "Identify\ndrivers", 
-                               "Test\nrelationships", "Predict\nspatially",
-                               "Predict\ntemporally"),
-                    values = c("gray60", "plum4", "cadetblue3", "aquamarine3",
-                               "aquamarine4", "goldenrod2", "goldenrod3")) +
-  scale_colour_manual(limits = c("Any\nObjective", "Demonstrate\nmethods", 
-                                 "Estimate\ntrends", "Identify\ndrivers", 
-                                 "Test\nrelationships", "Predict\nspatially",
-                                 "Predict\ntemporally"),
-                      values = c("gray60", "plum4", "cadetblue3", "aquamarine3",
-                                 "aquamarine4", "goldenrod2", "goldenrod3")) +
-  scale_shape_manual(limits = c(TRUE, FALSE),
-                     values = c(17, 16)) +
-  labs(x = "", y = "Number of covariates\nconsidered per parameter") +
-  facet_grid(cols = vars(facet), scales = "free_x", space = "free_x") +
-  guides(shape = "none") +
-  theme(legend.position = "none",
-        panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(),
-        legend.text = element_text(size = 15), axis.text = element_text(size = 13),
-        axis.title = element_text(size = 15), strip.background = element_blank(),
-        strip.text = element_blank()) 
-plot_objectiveCovariates
-ggsave(plot = plot_objectiveCovariates, "Figures/ObjectiveCovariates.jpeg",
-       width = 8, height = 4)
-```
-
-Combine the objective plots
-
-```{r combine objective plots}
-#| fig-width: 12
-#| fig-height: 8
-
-plot_stackedObjectives <- plot_objectiveTrend / plot_objectiveCovariates +
-   plot_annotation(tag_levels = 'A')
-plot_stackedObjectives
-
-ggsave(plot = plot_stackedObjectives, "Figures/ObjectiveFigure.jpeg",
-       width = 12, height = 8)
-
+ggsave(plot = plot_objectiveTrend, "Figures/2_Objectives.png",
+       width = 1476, height = 700, units = "px")
 ```
@@
   ggplot(aes(x = `Year Strata`, y = SiteNumber)) +
-  geom_boxplot(fill = "goldenrod", colour = "gray40", staplewidth = 0.4,
-               linewidth = 0.2, , outlier.size = 0.2) +
-  scale_y_log10() +
-  labs(x = "", y = "Number of sites\n(log scale)",
-       title = "Site quantity") +
+  geom_boxplot(fill = "#4D7668", 
+               linewidth = .3, staplewidth = 0.33,
+               colour = "gray25") +
+  scale_y_log10(breaks = c(10,100,1000,10000), limits = c(10,10000)) +
+  labs(x = "", y = "Number of sites\n(log)",
+       title = "") +
   theme(panel.grid.major.x = element_blank(),
@@
   ggplot(aes(x = `Year Strata`, y = CovsPerParam)) +
-  geom_boxplot(fill = "goldenrod", colour = "gray40", staplewidth = 0.4,
-               linewidth = 0.2, outlier.size = 0.2) +
+  geom_boxplot(fill = "#4D7668", linewidth = .3, staplewidth = 0.33,
+               colour = "gray25") +
   scale_y_log10() +
-  labs(x = "", y = "Covariates\nper parameter",
-       title = "Covariates considered") +
+  labs(x = "", y = "Covariates per\nparameter (log)",
+       title = "") +
   theme(panel.grid.major.x = element_blank(),
@@
   geom_point(size = 1) +
-  labs(x = "", y = "Percentage of models",
-       title = "Selection method (MLE*)") +
+  labs(x = "", y = "Percentage of\nmodels",
+       title = "") +
   scale_colour_manual("",
@@
                                  "Sequential", "Simple precursors"),
-                      values = c("aquamarine4", "cadetblue3",
+                      values = c("#4D7668", "cadetblue3",
                                  "plum4", "goldenrod")) +
   guides(colour = guide_legend(nrow = 2,
-                               override.aes = list(size = 0.5,
-                                                   linewidth = 0.5))) +
+                               override.aes = list(size = 1,
+                                                   linewidth = 1))) +
   theme(legend.position = "top",
@@
                       labels = c("Goodness-of-fit", "Validation"),
-                      values = c("aquamarine4", "goldenrod")) +
-  labs(x = "", y = "Percentage of models",
-       title = "Model assessment") +
-  guides(colour = guide_legend(override.aes = list(size = 0.5,
-                                                   linewidth = 0.5))) +
+                      values = c("#4D7668", "goldenrod")) +
+  labs(x = "", y = "Percentage of\nmodels",
+       title = "") +
+  guides(colour = guide_legend(override.aes = list(size = 1,
+                                                   linewidth = 1))) +
   theme(legend.position = "top", 
@@
                       labels = c("Maximum Likelihood", "Bayesian"),
-                      values = c("aquamarine4", "goldenrod")) +
+                      values = c("#4D7668", "goldenrod")) +
   guides(colour = guide_legend(nrow = 2, 
-                               override.aes = list(size = 0.2,
-                                                   linewidth = 0.2))) +
-  labs(x = "", y = "Percentage of models", 
-       title = "Model framework") +
+                               override.aes = list(size = 1,
+                                                   linewidth = 1))) +
+  labs(x = "", y = "Percentage of\nmodels", 
+       title = "") +
   ylim(0,100) +
@@
                         #55#") +
-  plot_annotation(tag_levels = "A") & 
+  plot_annotation(tag_levels = list(c("(a) Number of sites",
+                                      "(b) Number of covariates considered",
+                                      "(c) Modelling framework",
+                                      "(d) Covariate selection method (MLE*)",
+                                      "(e) Model assessment"))) &
   theme(legend.position = "top",
-        axis.text.x = element_text(size = 6, family = "Helvetica", 
-                                 angle = 45, vjust = 0.5, hjust = 1),
-        axis.text.y = element_text(size = 6, family = "Helvetica"),
-        axis.title = element_text(size = 6, family = "Helvetica"),
-        legend.text = element_text(size = 6, family = "Helvetica"),
-        legend.title = element_text(size = 6, family = "Helvetica"),
+        panel.grid.minor.y = element_blank(),
+        axis.text.x = element_text(size = 10, family = "Helvetica", colour = "#585858",
+                                 angle = 45, vjust = 1, hjust = 1),
+        axis.text.y = element_text(size = 10, family = "Helvetica", colour = "#585858"),
+        axis.title = element_text(size = 10, family = "Helvetica", colour = "#585858"),
+        legend.text = element_text(size = 10, family = "Helvetica", colour = "#585858"),
+        legend.title = element_text(size = 10, family = "Helvetica", colour = "#585858"),
         legend.key.size = unit(.2, 'cm'),
         legend.margin=margin(0,0,0,0),
-        legend.box.margin=margin(0,-10,-10,-10),
-        plot.tag = element_text(size = 8, family = "Helvetica"),
-        plot.title = element_text(size = 6, family = "Helvetica",
+        legend.box.margin=margin(0,-10,-5,-10),
+        plot.tag = element_text(size = 10, family = "Helvetica", colour = "#585858"),
+        plot.tag.location = "plot",
+        plot.title = element_text(size = 10, family = "Helvetica", colour = "#585858",
                                   hjust = 0.5))
@@
 
-ggsave(plot = timeSeries_Combined, "Figures/TimeSeries.png",
-       width = 11, height = 14, units = "cm")
-```
+ggsave(plot = timeSeries_Combined, "Figures/6_TimeSeries.png",
+       width = 1961, height = 2200, units = "px")
+```

### 2025-03-04 Manuscript.qmd
Major changes to MS over the past two weeks prior to resubmission

--- 2025-01-23 Manuscript.qmd
@@
 abstract: | 
- Occupancy models.
+ Since their introduction over twenty years ago, dynamic occupancy models (DOMs) have become a powerful and flexible framework for estimating species occupancy across space and time while accounting for imperfect detection.
+As their popularity has increased and extensions have further expanded their capabilities, DOMs have been applied to increasingly diverse datasets and research objectives in applied ecology.
+At the same time, technological advancements have resulted in massive increases in available data, offering both new opportunities and new challenges for users of DOMs.
+Given these developments, it is timely to examine common practices in building these models to understand the breadth of modelling approaches, determine potential vulnerabilities, and identify priorities for future research.
+We reviewed a sample of articles that have fit DOMs in the past 20 years, examining the contexts of their application and the approaches taken to the model building process.
+We find that these models have been used to pursue diverse objectives, based on datasets with wide-ranging spatial and temporal scales collected using a variety of survey methods.
+Our comparisons of modelling approaches indicate that many applications of DOMs considered relatively few covariates on key model parameters, as well as a tendency towards linear responses over more complex non-linear or interactive forms.
+Model selection techniques were largely idiosyncratic with little consensus on the best approaches, and model evaluation was rare across reviewed applications – additionally, existing research on either of these subjects is limited.
+Based on these findings, we identify key areas of the modelling process which merit discussion and further investigation.
+Where possible we provide recommendations for current users of DOMs, and where uncertainties remain, we highlight key priorities for future research to support users in fitting the most reliable and useful models possible.
 
@@
     theme: simplex
+    mainfont: Georgia, serif
   docx: default
@@
 
-Ecologists are often asked to describe, explain, and predict where and when species occur on the landscape.
-Research on patterns of species occupancy dates back to the origins of the field, and continues to a priority in quantitative ecological research [@humboldt1849].
+The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849].
 Estimates of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, while identifying potential drivers of occurrence can help inform potential management actions [@mackenzie2013].
@@
 
-While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate.
-The need to understand and quantitatively describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models commonly applied to study meta-population dynamics [@gutiérrez-arellano2024], and species distribution models (SDMs) widely used to explore species occurrence at larger scales [@franklin2010].
-However, several factors can make occupancy difficult to estimate.
-For instance, simple presence/absence observations can be biased when species are detected imperfectly – this is often the case in wildlife field data, where it can be impossible to determine from a single observation whether a location is truly occupied or whether the species occurs but was not detected [@gu2004; @lahoz-monfort2014].
-Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014].
-Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species’ occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010].
-These conditions often occur during biological invasions and climate change driven range shifts, each of which are conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].
+Irrespective of its importance, ‘occupancy’ – broadly referring here to the presence of a species in a given area and time period – has proven to be a persistently difficult quantity to estimate in practice.
+This difficulty arises from the challenges inherent in modelling complex and dynamic natural systems with incomplete data, exacerbated by the fact that it is often impossible to determine whether a species is truly absent from a site or whether it was simply not detected.
+It is well-established that imperfect detection of organisms results in biased estimates of occupancy [@gu2004; @lahoz-monfort2014].
+Despite this fact and the ubiquity of imperfect detection in field data, many models in widespread use make no adjustments for detection [@kellner2014].
+In addition to this limitation, many models currently in use (such as conventional correlative species distribution models) are ill-suited for predicting to new locations and time periods [@dormann2007; @elith2010].
+These shortcomings are particularly pronounced when modelling biological invasions and climate change driven range shifts, both of which are management priorities that are increasingly commonplace in the Anthropocene [@bertelsmeier2013; @lenoir2015].
 
-The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017a] .
-Drawing on principles from the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy.
-Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time.
-For newcomers to DOMs, in [Box @nte-Primer] we present a simple introduction to the basic model form as described in @mackenzie2003.
-
-This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes driving occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic or abundance data required by more process-explicit models [@briscoe2019].
-These features make the DOM an important tool, with uses including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@briscoe2021; @kéry2013].
+In recent years, awareness of these obstacles to accurate occupancy estimation and the limitations of many popular methods has encouraged interest in a variety of process-explicit models better suited to these contexts [@briscoe2019].
+These include ‘dynamic occupancy models’ (DOMs), first introduced by @mackenzie2003 as an extension to earlier static site occupancy models [@mackenzie2002].
+In their simplest form DOMs use hierarchical observation data to estimate changing site occupancy over time while correcting for imperfect detection; an overview of the basic form and requirements of these models can be found in Box 1.
+By accounting for imperfect detection and explicitly modelling colonisation and extinction to describe changes in site occupancy DOMs provide answers to common issues in occupancy estimation and offer important advantages over other models [@guillera-arroita2017a].
+Alongside these advantages, they require only relatively simple-to-collect detection/non-detection data rather than the detailed demographic or abundance data required by other more process-explicit models [@briscoe2019].
+This balance between complexity and feasibility makes DOMs an appealing option for many applied ecologists.
 
@@
 
-Dynamic occupancy models.
+DOMsare hierarchical models that link observed detection/non-detection data with the underlying latent process of changing site occupancy.
+To do this, they contain two [sub-models]{.underline}: one describing the ecological process of sites shifting between unoccupied and occupied states over time, and one describing the observation process that records whether a species is detected during surveys given its presence at a site.
 
-![](images/clipboard-1253177906.png){width="11cm"}
+**Input data:** To separate these two processes DOMs require hierarchical data inputs.
+Occupancy is estimated at independent sites *(i)* during discrete, time-bound intervals termed ‘primary occasions’ *(t)*.
+Within primary occasions*,* multiple observations are made at the same sites during independent ‘secondary occasions’ *(j).* For each observation, a species is either detected (y~i,t,j~ = 1) or not (y~i,t,j~ = 0), resulting in a three dimensional detection matrix as shown in the example below.
+It is not necessary for all primary occasions to contain the same number of secondary occasions, and missing observations at sites can be accommodated.
+While secondary occasions are most often represented as repeat observations, other options including same-visit replicates and spatial replicates may also be used [@guillera-arroita2017a].
 
-![](images/clipboard-3209412869.png){width="11cm"}
+![](Figures/B1_RobustDesign.png){fig-align="center"}
+
+**Occupancy sub-model:** Changing site occupancy over time is described as a Markovian process as represented in the graphic below.
+In each primary occasion *t,* each site *i* may exist in either an occupied (z~i,t~ = 1) or unoccupied state (z~i,t~ = 0).
+In the first primary occasion, occupancy is determined by a Bernoulli trial with initial occupancy probability 𝛙~1~, such that $z_{i,t=1} \sim Bernoulli(\psi_1)$.
+In subsequent primary occasions, site occupancy is determined by the site's state in the previous primary occasion and probabilities of colonisation 𝛄 and extinction 𝛆, such that $z_{i,t} \sim Bernoulli(z_{i,t-1}(1-\epsilon)+(1-z_{i,t-1})*\gamma)$.
+
+![](Figures/B2_Occupancy.png){fig-align="center"}
+
+**Observation sub-model:** The detection process relates the observed data **y** to the latent occupancy states **z** as shown in the below graphic.
+During any given secondary occasion *j* within primary occasion *t* at site *i,* observers will either observe (y~i,t,j~ = 1) or not observe (y~i,t,j~ = 0) the target species.
+The probability of observation is given as a Bernoulli trial with detection probability *p,* conditioned on the site being occupied in primary occasion *t* such that $y_{i,t,j} \sim Bernoulli(z_{i,t}*p)$.
+
+![](Figures/B3_Detection.png){fig-align="center"}
+
+In their original formulation, DOMs estimate four parameters of interest: initial occupancy (𝛙~1~), colonisation (𝛄), extinction (𝛆), and detection (*p)*.
+While we have represented each of these parameters as constants for simplicity, they are more often modelled as time-dependent or estimated with respect to covariates.
+These covariates are typically incorporated via a logit link function, such that DOMs can be considered four interconnected generalised linear models.
+
+Conventional DOMs make several important assumptions: i) sites are closed to changes in true occupancy state within each primary occasion, ii) occupancy at each site is independent of other sites, iii) observations within each primary occasion are independent of each other, iv) no false positive detections occur, and v) key sources of heterogeneity (including in detectability) are modelled.
+
+For more detailed information on DOMs and related models, see [@mackenzie2017] and [@kéry2021].
 :::
 
-DOMs make several key assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.
+Over the past two decades, continued research on DOMs and occupancy models more broadly has established a powerful and flexible modelling framework suitable for many common tasks in ecological research.
+Uptake of these models has been encouraged by the availability of freely available software tools for fitting DOMs, which include the program PRESENCE and the R package *unmarked* for fitting the models by maximum likelihood estimation [@hines2006; @kellner2023]*.* In the Bayesian context, resources such as @kéry2021 ’s text on hierarchical modelling in JAGS and the *ubms* R package have helped to increase the accessibility of these implementations [@kellner2022].
+Various model extensions have further broadened DOMs’ capabilities – these include models that can account for false positives [@royle2006; @miller2011], model multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimate occupancy for multiple species [@dorazio2010; @devarajan2020].
+For more details on using these extensions, see reviews by @bailey2014 and @guillera-arroita2017a.
 
-I.
-**False positive detections do not occur**.
-While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods.
-@mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need to consider how certain detections truly are for any given study system.
-Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019].
-Where this assumption can not reasonably be met, model extensions designed to account for false positive error should be considered [@royle2006; @miller2011; @miller2015].
+Coinciding with these developments, recent years have seen substantial changes in how ecologists conduct their research.
+The amount of data available for modelling (including species detections as well as environmental data) has grown considerably over time due to a range of factors, including improvements to data sharing, new large-scale monitoring programs, and increased interest in citizen science efforts [@farley2018; @altwegg2019].
+Technological advances have facilitated the widespread deployment of autonomous detection methods including camera traps and acoustic monitors, generating large quantities of observation data suitable for analysis with DOMs [@balantic2019; @lahoz-monfort2021].
+At the same time, advancements in computing have made methods which may have been too computationally expensive in the past far more accessible for many.
+While these advances create exciting new opportunities, they also introduce new challenges for users who must navigate a complex model building process to produce useful and reliable models.
+Where researchers on related species distribution models have built a large body of literature on assessing various approaches to model building including covariate inclusion and model selection, comparatively little research on these topics has been published for DOMs despite additional complications in performing these tasks for hierarchical models [@guisan2017].
 
-II.
-**Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion in terms of the bias introduced when it is violated [@rota2009; @otto2013].
-Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and it can be difficult to distinguish local extinction from temporary emigration [@valente2017].
-Model extensions to relax the closure assumption have been developed, including approaches using staggered arrival and departure periods between sites [@kendall2013].
-A more pertinent approach for most studies, however, is careful consideration of the appropriate definition of occupancy under the survey design used [@steenweg2018].
+### Objectives
 
-III.
-**Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species occurrence, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system.
-Generally, this is achieved by allowing model parameters (ψ~1~, γ, ε, and ρ) to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters.
-An important element of this assumption is that the likelihood of detection of a species can depend not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017a].
-While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates [@barry2006].
-Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known about how this latent heterogeneity can influence model performance.
+To understand how users have applied DOMs over the past two decades, we conducted an in-depth review of studies implementing these models since their introduction.
+We first identify the research contexts where DOMs have been used, characterising the research objectives they have been applied to, the scale and characteristics of the study systems where data was collected, and the methods used to collect and format the data used to fit models.
+We then review approaches taken to the modelling process, including the nature of the covariates considered for inclusion and the form of their relationships in the model, the methods used for model selection, and the reporting of model assessment and evaluation.
+By jointly considering the research contexts to which DOMs are now applied and the approaches taken for model building and evaluation, we aim to highlight challenges in building DOMs, providing recommendations where possible and identifying priorities for future research where uncertainties remain.
 
-Since its original description in @mackenzie2003, the DOM has been further developed with numerous model extensions and alternative formulations including aforementioned implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010].
-For a comprehensive discussion of common extensions and their applications see @bailey2014 and @guillera-arroita2017a, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.
+# Methods
 
-# Review methods
+Our review was constrained to applications of the dynamic occupancy model of @mackenzie2003 and its extensions. To be included, articles must have fit models that met the following criteria:
 
-Methods.
+i.   Used non-simulated, field-collected, detection/non-detection data.
+
+ii. Included data from multiple sites which could exist in at least two states, including an occupied and unoccupied state.
+
+iii.
+Included data from multiple primary occasions, between which sites change states conditional on the prior occasion’s occupancy state and transition probabilities such as colonisation and extinction.
+
+iv. Contained at least one parameter describing the detection process.
+
+A pool of candidate articles was generated using two queries on Web of Science on July 26 and July 29, 2024.
+The first query included all articles from 2004-2023 which cited @mackenzie2003.
+To capture additional relevant articles that did not directly cite @mackenzie2003, a second query was generated to search articles in the same time-span matching the terms “*dynamic occupancy model\**”, “*multi-season occupancy model\**”, or “*occupancy dynamic\**”; articles including each of “*occupancy*”, “*colonization*”, “*extinction/persistence*”, and “*detection*”; and articles with the term “*occupancy*” located near “*dynamic*” in the title, key terms, or abstract.
+These queries resulted in 1469 articles: 897 retrieved only from the MacKenzie citations, 274 only from the keywords search, and 298 which appeared in both queries.
+To allow comparison of DOMs through time, we divided all articles across five four-year strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2023.
+From each stratum we randomly selected 20 articles for inclusion in the review.
+Articles that did not meet inclusion criteria were replaced from within their own stratum unless no articles remained.
+For each article we documented details on the research contexts, datasets, and modelling processes as outlined below.
+
+### Research objectives
+
+We allocated each article to research objective categories based on the aims stated in the text.
+The six non-exclusive categories included i) ‘estimating trends’, where authors expressed interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; ii) ‘testing relationships’, where authors examined predefined hypothesised relationships between covariates and model parameters; iii) ‘identifying drivers’, where authors more broadly explored covariates associated with model parameters; iv) ‘predicting temporally’, where authors predicted site occupancy under future conditions; v) ‘predicting spatially’, where authors predicted site occupancy to unsurveyed locations; and vi) ‘developing methods’, where authors introduced, tested, or demonstrated aspects of DOMs.
+
+### Study systems
+
+We recorded the approximate geographical location, spatial scale, and target taxa for each study system in the reviewed articles.
+In cases where a single article included DOMs fit to data from multiple study systems, we recorded details for each dataset that was analysed independently.
+A study system’s spatial scale was defined as the intended area of inference containing all sites, measured to an order of magnitude to account for uncertainty in reporting.
+Details collected on target taxa included the total number of taxa modelled, their general categorisation (birds, mammals, herptiles, invertebrates, or ‘other’), and their conservation status.
+Taxa were denoted as threatened if they were listed on the IUCN Red List of Species as of 2024 or if authors explicitly stated that they were threatened [@iucn].
+This deference to authors’ representation of conservation status was made to account for sub-species which lack listings or species which are of more local concern, although we acknowledge that in some cases conservation status may have changed since publication.
+
+### Observations and data structure
+
+For each article, we recorded how observation data was collected for use in modelling.
+Categories of survey methods included human observations, physical trapping, camera traps, and acoustic monitors.
+Within these categories we also indicated whether any observations were collected by citizen scientists, either as part of structured survey programs or as more *ad hoc* observations.
+Details collected on each dataset’s structure included the number of primary and secondary occasions, the time elapsed between the first and last survey, and the number of sites used for modelling.
+
+### Covariates and complexity
+
+We were interested in the types and quantities of the covariates considered by authors.
+To this end, we recorded all covariates considered in each study regardless of whether they were included in final models, acknowledging that not all covariates considered are always reported.
+Key traits of each covariate were recorded including a general categorisation (see supporting information), whether they were directly observed or remotely sensed, whether they were static or varied between primary occasions, and how they were represented in the model: as a linear response, a non-linear response, or part of an interaction with another covariate [@james2021].
+In some cases, a single article included distinct modelling workflows with different candidate covariates.
+Covariates were recorded for each of these workflows independently.
+
+### Model selection and assessment
+
+Model selection procedures were sorted into non-exclusive categories including ‘*a priori’*, where only one model was considered; ‘candidate set’, where a predefined set of models was considered; ‘sequential’, where covariates were selected parameter-by-parameter (e.g., fitting detection first, followed by initial occupancy and so on); and ‘simple precursors’, where covariate selection was preceded by tests with a simpler model implementation such as a linear regression or single season occupancy model.
+We also indicated whether model averaging was used for multi-model inference @burnham2004.
+For each modelling workflow, we documented whether goodness-of-fit was tested and reported, and whether model performance was assessed by validation with either in-sample or out-of-sample data.
 
@@
 
-Coverage.
+92 articles were included for this review, fewer than the 100 possible articles due to a deficit of qualifying papers in the first stratum (2004-2007).
+All review data is available in the supporting information.
+Based on the acceptance rates within each stratum and the number of unprocessed articles remaining for each year, an estimated 496 of 1152 unprocessed articles in our sample would have met inclusion criteria, with an apparent increase over time in the number of articles fitting DOMs (@fig-coverage).
 
-![Bars indicate the 92 articles included in our review as a proportion of the estimated number of published articles fitting DOMs, based on the qualification rate for articles in each strata.
-The proportion of articles included from each strata were as follows: 12% from 2005-2008; 24% from 2008-2011; 42% from 2012-2015; 35% from 2016-2019; and 57% from 2020-2023.](Figures/CoveragePlot.jpeg){#fig-coverage}
+![Bars indicate the estimated number of articles fitting DOMs in each year, with green representing articles included in our review and grey projected remaining qualifying articles.
+Projections were calculated using the number of unprocessed articles remaining in each year multiplied by the acceptance rate of the corresponding stratum: 12% of queried articles from 2004-2007; 24% from 2008-2011; 42% from 2012-2015; 35% from 2016-2019; and 57% from 2020-2023.](Figures/1_CoveragePlot.png){#fig-coverage}
 
-Stamp collection.
+### Research objectives
 
-![A) Locations of study areas where data was collected for reviewed DOMs.
-B) Spatial extent of study areas, defined as the area of inference, within which all surveyed points were contained.
-C) Number of articles which fit models to each category of taxa.
-Taxa were considered 'threatened' if they are listed on the IUCN Red List, or if authors indicate that they are otherwise threatened.
-D) Explicitly multi-species models include both hierarchical, jointly estimated models as well as more interactive models.
-Some studies fit both independent and multi-species models, such that these values do not sum to our sample size.
-E) Survey methods used to collected presence/absence data.
-Note that some articles employed multiple detection methods, and that some methods (e.g., citizen bird counts) may fall into multiple categories.
-F) Quantity of sites where surveys were conducted and duration of studies.
-Yellow bars indicate median values for site quantity (100) and study duration (8.2 years).
-Study duration is defined as the time elapsed between the first and last survey.](Figures/ReviewBox.png){#fig-StudyDetails}
+DOMs have been used to achieve varied research objectives, with no one category of objective representing over half of usage and 37% of studies having pursued multiple objectives simultaneously (Figure 2).
+The most frequent use of DOMs (44% of studies) was to test hypothesised relationships between environmental factors and species occupancy, often targeting core conservation priorities for their focal species as demonstrated by @olson2005 ’s early study assessing the influence of barred owls (*Strix varia)* on the threatened spotted owl (*Strix occidentalis*).
+35% of studies took a more exploratory approach to identifying possible drivers of occupancy, such as in @huber2017 ’s study that examined dozens of habitat covariates for wood warbler (*Phylloscopus sibilatrix*) occupancy.
+An additional 36% of articles used DOMs to monitor trends in occupancy state through time, both for single species of high conservation interest and for broader community assemblages of species [@ahumada2013; @scott2015].
+While only 17% of articles used their DOMs to make predictions, the proportion of articles pursing this objective has increased in recent years.
+These studies tended to have a strong conservation focus, as with @mcgowan2020 ’s projections of future occupancy for wetland birds under alternative management scenarios.
+Finally, 22% of the papers reviewed focused on methods development, representing continued focus on extending and testing DOMs.
 
-Covariates.
+The applied nature of studies fitting DOMs is further reflected in their authorship: while 79% of studies included at least one affiliate of an academic institution, 59% included a government affiliate, 28% an NGO affiliate, and 12% an affiliate of a private company.
+63% of studies had cross-sector authorship with affiliations from at least two categories.
 
-::: column-page-inset
-```{r covariate table}
+![Bars represent the percentage of articles which pursued each of six non-exclusive research objectives.
+These objectives include articles which demonstrated methods for DOMs, estimated trends in model parameters, identified drivers of occupancy dynamics, tested hypothesised relationships with the environment, made spatial predictions, and made temporal predictions.
+Percentages are given for each four-year stratum (n = 12 articles for 2004-2007; n = 20 for all other strata), as well as for all articles in the review sample (n = 92).](Figures/2_Objectives.png){#fig-Objectives}
+
+### Study systems
+
+The study systems included in our sample were globally distributed: while a majority of articles used data collected in the United States of America, study systems spanned all continents except Antarctica (@fig-StudyDetails A).
+These study areas varied considerably in size, ranging from hyper-local to continental scales (@fig-StudyDetails B).
+The smallest study locality included in our sample explored insect occurrence in a rainforest plot less than one square kilometre, while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@clement2019; @basset2023].
+
+Most studies focused on vertebrate taxa (@fig-StudyDetails C), particularly birds and mammals.
+DOMs have been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community.
+However, there were exceptions, including studies that used DOMs to model decadal changes in lichen occupancy, the spread of chronic wasting disease, and mosquito dynamics [@belinchón2017; @mores2020; @cook2022].
+The vast majority of studies model terrestrial species, though there have been a limited number of studies on aquatic taxa including invasive salmon, Great Plains stream fishes, and whales [@falke2012; @fisher2014; @pendleton2022].
+
+While most studies fit models for a single taxon (either a single species or multiple species lumped together), 44% fit models to multiple distinct taxa (@fig-StudyDetails D).
+34% of studies fit independent models to multiple species [@otto2012; @peach2019], whereas 12% used explicitly multi-species extensions of DOMs.
+The latter included hierarchical models which fit hundreds of species in a single implementation with species-level effects [@dorazio2010; @hendershot2020] as well as explicit interaction models which estimated conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019].undefined
+
+![A summary of the research contexts in reviewed applications of DOMs.
+a) The approximate locations of the study areas where data was collected.
+b) Spatial extent of the study areas, defined as the area of inference within which all surveyed points were contained.
+c) Number of articles that fit models to each taxa category.
+Taxa are considered ‘threatened’ if indicated by authors or they are listed on the IUCN Red List.
+d) The number of taxa modelled as distinct groups in each application.
+Explicitly multi-species models include both hierarchical jointly estimated models and more interactive models.
+e) Number of articles using each method for capturing detection/non-detection data.
+These are non-exclusive, and citizen science data is a subset of ‘direct observations.’ (f) The number of sites and study duration for each dataset used to fit DOMs – gold lines indicate medians.](Figures/3_StudyContexts.png){#fig-StudyDetails}
+
+### Observations and data collection
+
+The detection/non-detection data required for DOMs was collected using various survey methods across reviewed studies (@fig-StudyDetails E): 79% of studies used direct observation data, 11% used live-trapping methods, and 14% used detections from camera traps.
+Within these categories, 10% of studies used citizen science data.
+Citizen science data included coordinated surveys at backyard bird-feeders as well as interviews with civilians on sightings of tiger (*Panthera tigris*) signs [@zuckerberg2011; @warrier2020].
+Datasets ranged considerably in size, with the smallest including observations at just 10 sites and the largest including over 6000 (@fig-StudyDetails F); the median dataset included 100 sites.
+The temporal scale of studies shows similar variability – time elapsed between the first and last survey ranged from under one month to over forty years (median 8 years), and the number of primary occasions ranged from 2 to 189 (median 6 occasions).
+
+Notably, not all of these datasets were originally collected in a hierarchical structure with DOMs in mind.
+In these cases, authors formatted their data into a hierarchical format *post-hoc* using a variety of methods.
+Some defined primary occasions as arbitrary discrete time intervals, treating all surveys occurring within a window as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate.
+For example, in the only application of DOMs to marine species in our sample, @pendleton2022 used aerial transects broken up into grid cells to observe whale occupancy.
+In another example with grid cells, @marescot2020 fit a multi-species model treating poachers as a taxon and using ranger reports from each cell to create detection histories.
+These manipulations permit use of data predating DOMs, with one study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].
+
+### Covariates and complexity
+
+The most common covariates considered for use in DOMs addressed aspects of habitat and land cover [@fig-CovariateSummary].
+35% of studies incorporated covariates for site geometry and connectivity, such as habitat patch size or distance to other sites.
+Often these were included as simple covariates on colonisation or extinction, such as the distance to other sites or landscape habitat connectivity metrics [@duggan2011].
+Alternatively, more complex parameterisations explicitly modelled colonisation or extinction as a spatial process dependent on patch size or the distance to occupied sites [@risk2011; @broms2016].
+Several studies also included biotic interactions with other species as covariates, often where the target taxon was threatened by invasive species.
+These covariates effectively incorporate species interactions in DOMs without requiring the use of more complex explicitly multi-species models.
+
+![Attributes of the covariates considered for inclusion in DOMs.
+Columns represent the percentage of studies that tested at least one covariate in a category on any model parameter (a), and the average percentage of covariates in each category that were (b) time-varying, excluding covariates on initial occupancy, (c) directly observed at each site, and (d) considered with non-linear terms or interactions with other covariates.
+To avoid the dominance of studies considering large amounts of covariates in b-d, we first calculated the relevant percentage of covariates per study then took the mean.
+Covariates were classified as either environmental factors representing plausible ecological correlates of model parameters, or structural factors related to the model form or observational details that distinct from the environment.
+Environmental categories included ‘Habitat’ – land cover and habitat features; ‘Spatial’ – site dimensions and physical location; ‘Climate’ – weather and long-term climate; ‘Timing’ – chronology with ecological relevance; ‘Human’ – interaction with anthropogenic activity; ‘Topography’ – geologic features; ‘Biotic’ -- any potential predator/prey/competitor interactions; and ‘Water’ – hydrologic features.
+Structural categories included ‘Primary’ – effect of the primary occasion; ‘Observation’ –observation method and characteristics; ‘Secondary’ – effect of the primary occasion; ‘Site’ – site-level effects; and ‘Species’ – species-specific effects.
+We also provide summaries across all covariate categories (‘Any’), all environmental covariates (‘Any Env’), and all structural covariates (‘Any Struct.’).
+For more verbose definitions and analysis code, see supplementally materials.](Figures/4_CovariateSummary.png){#fig-CovariateSummary}
+
+Covariate data for studies in our sample were either collected directly during surveys by researchers (an average of 30% of environmental covariates per study), or derived from pre-existing remotely sensed datasets (70% of environmental covariates); this varied depending on the category of covariate [@fig-CovariateSummary].
+Directly collected data often represented finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely.
+An average of 43% percent of the environmental factors and 94% of the structural factors considered for time-varying parameters were dynamic covariates, with terms relating to climate or weather most frequently dynamic [@fig-CovariateSummary].
+
+In conventional DOMs, covariates for each parameter are incorporated via a logistic regression (i.e., a linear regression through a logit link function; @mackenzie2017).
+Statistical relationships between model parameters and covariates (e.g., between initial occupancy and its environmental covariates) are represented as linear terms unless more complexity is specified, although non-linear responses can be easily accommodated in DOMs by using polynomial transformations and interactions between covariates.
+Despite this, in our sample only 35% of articles tested one or more non-linear responses to an environmental covariate, with most studies representing all covariates as simple linear terms.
+Interactions between covariates were similarly rare, with only 24% of studies considering at least one interaction between terms.
+
+The size of the covariate pool for each parameter varied substantially, with the number of covariates considered ranging from 0 (effectively modelling the parameter as a constant) to over 40 candidates on a single parameter (@fig-CovariateCount).
+Note that this does not represent the number of covariates included in the final model formulation used for inference.
+The median number of covariates considered varied by parameter, with transition probabilities (colonisation and extinction) more likely to have a broader range of environmental covariates considered compared to initial occupancy and detection.
+The lack of any covariates considered for initial occupancy in 37% of studies is particularly notable, as is the observation that 30% of modelling workflows considered no environmental covariates for detection probability.
+
+![The number of covariates considered for each core parameter across all distinct modelling workflows in our sample.
+A covariate is defined here as a distinct variable considered for inclusion, with linear, non-linear, and interactive forms of the same factor counted as a single covariate.
+Covariates are either ‘environmental’, representing plausible ecological correlates of model parameters, or ‘structural’, relating to factors without direct ecological relationships.
+The ‘Occupancy\*’ category here corresponds with alternative parameterisations of DOMs that jointly estimate occupancy, colonisation, and detection with extinction being only a derived parameter; this contrasts with more popular initial occupancy/colonisation/extinction/detection parameterisations.](Figures/5_CovariateCount.png){#fig-CovariateCount}
+
+### Model selection and assessment
+
+The 92 articles in our sample included 102 distinct modelling workflows.
+Of these models, 76 were fit by maximum likelihood estimation (MLE) and 24 were Bayesian implementations.
+2 models fell into neither category and instead used machine learning based methods [@joseph2020].
+
+In most cases, the DOMs fit in a study did not use all covariates initially considered for inclusion: 80% of modelling workflows included some form of model selection approach to identify a final model or set of models to make inferences from [@tbl-modelling].
+Approaches varied between the MLE and Bayesian implementations in our sample; where 95% of MLE models performed some manner of model selection, only 33% of Bayesian models did so, with the majority instead fitting a single model defined *a priori*.
+For MLE models the most popular and conventional approach to model selection (45% of models) involved the creation of a candidate set of models, where the best model(s) was selected according to the lowest AIC score.
+The next most popular method in MLE studies was to use sequential model selection methods (37% of models), where the structure for each model parameter was fit in sequence.
+For example, a protocol might have first identified the best structure for detection probability while holding the other parameters constant, before moving on to initial occupancy and so on until all parameters were fixed.
+The remainder of MLE studies (8%) used a variety of other approaches, such as fitting simpler models like single season occupancy models to identify the most informative covariates to use in a dynamic occupancy model.
+Across all MLE implementations, 47% of articles used multi-model inference by model-averaging with AIC weights [@burnham2004].
+
+```{r Table_Modelling}
 #| echo: false
 #| message: false
-#| label: tbl-covariates
-#| tbl-cap: "All covariates considered for inclusion in a study were classified into mutually exclusive categories.
-We calculate the percentage of studies which include at least one covariate of a given category on any parameter, Initial Occupancy (&psi;1), Occupancy (&psi;), Colonisation (&gamma;), Extinction(&epsilon;), and Detection (&rho;).
-We also present the average percentage of covariates in a category which are dynamic (varying through seasons) and directly observed, as well as the percentage of articles which model each category of covariate with a non-linear relationship or as part of an interaction with another covariate."
+#| label: tbl-modelling
+#| tbl-cap: "A summary of modelling practices in DOMs subset by framework (maximum likelihood or Bayesian).
+Some studies include multiple distinct modelling workflows (n = 102 workflows across 92 articles).
+2 models included in the ‘All models’ column were neural network based and fell into neither the MLE or Bayesian categories.
+The model selection methods represented in this table are non-exclusive and some articles employed multiple approaches."
 
@@
 library(gt)
-
-covTable <- read_csv("Figures/CovariateTable.csv")
-
-covTable |>
-  mutate(Summary = case_when(Covariate %in% c("Environmental", "Structural", "Overall") ~ TRUE,
-                             .default = FALSE)) |>
-  arrange(Category, Summary, -Prop_Any) |>
-  gt(rowname_col = "Covariate", groupname_col = "Category") |>
-  tab_options(
-    row_group.as_column = TRUE
-  ) |>
-  tab_row_group(
-    label = md("Environmental covariates"),
-    rows = Category == "E",
-    id = "Env"
-  ) |>
-  tab_row_group(
-    label = md("Structural covariates"),
-    rows = Category == "S",
-    id = "Struct"
-  ) |>
-  tab_row_group(
-    label = md("All covariates"),
-    rows = Category == "O",
-    id = "All"
-  ) |>
-  tab_style(
-    style = cell_fill(color = "aquamarine4"),
-    locations = cells_row_groups(groups = "Env")
-  ) |>
-  tab_style(
-    style = list(
-      cell_fill(color = "plum4")
-    ),
-    locations = cells_row_groups(groups = "Struct")
-  ) |>
-  tab_style(
-    style = list(
-      cell_fill(color = "gray60")
-    ),
-    locations = cells_row_groups(groups = "All")
-  ) |>
-  row_group_order(groups = c("Env", "Struct", "All")) |>
-  tab_spanner(label = "Percentage of articles with covariate on parameters:",
-              columns = c("Prop_Any", "Initial occupancy", "Occupancy",
-                          "Colonisation", "Extinction_Persistence",
-                          "Detection"),
-              id = "inclusions") |>
-  tab_spanner(label = "Percentage which are:",
-              columns = c(Prop_Dynamic, Prop_Direct),
-              id = "covariates") |>
-  tab_spanner(label = "Articles representing this covariate with:",
-              columns = c(Prop_Nonlinear, Prop_Interact),
-              id = "representations") |>
-   cols_label(
-    Prop_Any = "Any",
-    `Initial occupancy` = html("&psi;1"),
-    Occupancy = html("&psi;"),
-    Colonisation = html("&gamma;"),
-    Extinction_Persistence = html("&epsilon;"),
-    Detection = html("&rho;"),
-    Prop_Dynamic = "Dynamic",
-    Prop_Direct = "Directly observed",
-    Prop_Nonlinear = "Non-linear relationships",
-    Prop_Interact = "Interactions between covariates"
-  ) |>
-  text_case_match(
-    "Environmental" ~ html("<b>Any Environmental</b>"),
-    "HABT" ~ html("<b>Habitat </b>\n<i>Aspects of habitat and land cover</i>"),
-    "PHEN" ~ html("<b>Phenology</b>\n<i>Time-varying elements distinct from sampling occasions</i>"),
-    "SPAT" ~ html("<b>Spatial</b>\n<i>Site geometry, connectivity, or other spatial elements</i>"),
-    "OTHER-E" ~ html("<b>Other environmental</b>\n<i>Other environmental covariate not otherwise listed</i>"),
-    "TOPO" ~ html("<b>Topography</b>\n<i>Elements of landscape topography</i>"),
-    "HYDR" ~ html("<b>Hydrology</b>\n<i>Aspects of hydrology, such as distance to water</i>"),
-    "CLWE" ~ html("<b>Climate/Weather</b>\n<i>Climate, weather, and natural disasters</i>"),
-    "ANTH" ~ html("<b>Anthropogenic</b>\n<i>Relations to human activity</i>"),
-    "BIOT" ~ html("<b>Biotic interaction</b>\n<i>Interactions with other (non-plant) species</i>"),
-    "Structural" ~ html("<b>Any Structural</b>"),
-    "SEAS" ~ html("<b>Primary occasion</b>\n<i>Effect of the primary occasion</i>"),
-    "OBSE" ~ html("<b>Observation</b>\n<i>Details on the observation process</i>"),
-    "SURV" ~ html("<b>Secondary occasion</b>\n<i>Effect of the secondary occasion</i>"),
-    "SPEC" ~ html("<b>Species effect</b>\n<i>Species-level effects</i>"),
-    "SITE" ~ html("<b>Site effect</b>\n<i>Site-level effects</i>"),
-    "OTHER-S" ~ html("<b>Other structural</b>\n<i>Other structural covariate not otherwise listed</i>"),
-    "Overall" ~ html("<b>All covariates</b>"),
-    .locations = cells_stub()
-  ) |>
-  cols_hide(c(Category, Summary)) |>
-  fmt_percent(decimals = 0) |>
-  sub_missing(missing_text = "-") |>
-  data_color(
-    columns = where(is.numeric),
-    method = "numeric",
-    palette = "viridis",
-    alpha = 0.5
-  ) |>
-  tab_style(locations = cells_stub(rows = "Environmental"),
-            style = cell_fill(color = "aquamarine4", alpha = 0.5)) |>
-  tab_style(locations = cells_body(rows = "Environmental"),
-            style = list(cell_fill(color = "aquamarine4", alpha = 0.5),
-                         cell_text(weight = "bold", style = "italic",
-                                   color = "black"))) |>
-  tab_style(locations = cells_stub(rows = "Structural"),
-            style = cell_fill(color = "plum4", alpha = 0.5)) |>
-  tab_style(locations = cells_body(rows = "Structural"),
-            style = list(cell_fill(color = "plum4", alpha = 0.5),
-                         cell_text(weight = "bold", style = "italic",
-                                   color = "black"))) |>
-  tab_style(locations = cells_stub(rows = "Overall"),
-            style = cell_fill(color = "gray60", alpha = 0.5)) |>
-  tab_style(locations = cells_body(rows = "Overall"),
-            style = list(cell_fill(color = "gray60", alpha = 0.5),
-                         cell_text(weight = "bold",
-                                   color = "black"))) |>
-  tab_style(locations = cells_row_groups(),
-            style = cell_text(color = "white")) |>
-  as_raw_html()
-```
-:::
-
-More covariates.
-
-![The number of covariates considered for each parameter across all studies in our sample.
-'Occupancy' given here represents the alternative parameterisation of the DOM which jointly estimates Occupancy for every season, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation.
-Here, a 'covariate' is defined as each distinct covariate considered for inclusion.
-Linear and quadratic representations of the same covariate are counted as one covariate.](Figures/CovParamPlot.jpeg){#fig-covariates}
-
-Modelling summary.
-
-```{r Modelling summary table}
-#| echo: false
-#| message: false
-#| label: tbl-modelling
-#| tbl-cap: "Modelling practices in dynamic occupancy models, subset by frequentist or bayesian implementations.
-The median covariate count presented here represents the median quantity of covariates considered for each model parameter across the studies in our review.
-The model selection methods represented in this table are non-exclusive and some articles employ multiple approaches.
-2 models included in the 'Overall' column are neural network based and fall into neither the Frequentist or Bayesian categories."
-
-modellingSummary <- read_csv("Figures/ModellingSummary.csv")
+modellingSummary <- read_csv("Figures/T2_ModellingSummary.csv") |>
+  rename(`MLE` = Frequentist)
 
@@
   row_group_order(groups = c(NA, "select", "eval")) |>
-  text_case_match("Number" ~ "Number of studies",
+  text_case_match("Number" ~ "Number of workflows",
                   "MedianCovs" ~ "Median covariates considered per parameter",
@@
                   "Prop_AnySelect" ~ "Percentage using any model selection approach",
-                  "Prop_Sequential" ~ "Percentage using procedural model selection",
+                  "Prop_Sequential" ~ "Percentage using sequential model selection",
                   "Prop_Simple" ~ "Percentage selecting covariates with simpler models",
@@
 
-Objectives.
+Those Bayesian models which did perform model selection took various approaches, with largely idiosyncratic methods across these studies.
+While direct comparison of model fit was rare amongst Bayesian methods, it is feasible – @urban2023 identified the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data.
+Another approach used by @cook2022 fit a global model including all covariates, before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the reduced model.
+@ahumada2013 took a hybrid approach, where model selection was conducted by a sequential method fitting models by MLE before refitting the best structure as a Bayesian model.
+Finally, @brown2014 capitalised on the advantages of the Bayesian framework and used reversible jump MCMC routines to perform model selection during model fitting.
 
-![A) Proportion of articles in each year-strata and across all years which match each of six non-exclusive objective categories.
-B) Quantity of covariates considered per parameter for models which pursued each objectives.](Figures/ObjectiveFigure.jpeg){#fig-objectives}
+Regardless of the implementation, assessment of model fit and model performance was rare amongst the articles in our sample.
+Only 18% of studies tested for goodness-of-fit, and just 7% calculated predictive performance with either in-sample or out-of-sample validation.
 
-More objectives.
+### Temporal trends in applications of DOMs
+
+While our temporally stratified sample allows us to consider changes in how DOMs have been applied over the past two decades, modest sample sizes within strata necessitate caution in interpretating these results.
+Nonetheless, there are several clear trends we believe are worth noting (Figure 6).
+Firstly, recent strata saw changes in the number of sites modelled in each study.
+In our two most recent strata spanning 2016-2023, the range in the number of sites was noticeably wide.
+While models continue to be fit with small numbers of sites, the upper quartile contained more sites than in previous strata (@fig-timeSeries A).
+The median number of covariates considered per parameter has also continually increased since the early years of DOMs, from 1.75 covariates in 2004-2007 to 4.75 in the 2020-2023 stratum (@fig-timeSeries B).
+
+![Key trends in applications of dynamic occupancy models over the study period.
+a) The number of sites modelled in each dataset, presented on the log scale.
+b) The number of covariates considered per model parameter in each modelling workflow, presented on the log scale.
+c) The percentage of models in each stratum fit in the maximum likelihood or Bayesian frameworks.
+d) The percentage of articles in each stratum that used each of four non-exclusive covariate selection strategies.
+This panel includes only articles which fit models via maximum likelihood.
+e) The percentage of articles in each stratum that reported goodness-of-fit testing and model validation (with either in or out of sample data).](Figures/6_TimeSeries.png){#fig-timeSeries}
+
+While Bayesian models were included in all strata of our sample, there was a marked increase in their frequency in the most recent stratum (Figure 6 C).
+This may in part be explained by the publication of accessible resources such as @kéry2021, which includes chapters on fitting DOMs in JAGS with associated code.
+Within the MLE implementations, the most popular model selection methods have shifted over time.
+The number of studies using predefined sets of candidate models has gradually declined, contemporaneous with an increase in articles using sequential model selection methods (@fig-timeSeries D).
+While there is evidence of improvement from earlier strata, goodness-of-fit testing and model validation using either in- or out-of-sample data remained uncommon even in more recent strata where tests were readily available in common R packages (@fig-timeSeries E).
 
@@
 
-Discussion.
+Over the past two decades dynamic occupancy models have been applied to an increasingly broad range of objectives and research contexts.
+As their popularity has grown and new tools have become available, authors have implemented DOMs with wide ranging amounts of data, at small and large spatial and temporal scales, using diverse data collection techniques.
+While each of these studies share the same underlying methodology, their approaches to implementation and interpretation vary considerably.
+The approach to building any type of model will necessarily depend on the character of the data available and on the priorities of the model-builder.
+This precludes any prescription of the ‘best’ way to build DOMs; however, we believe that there are several areas of the modelling process that merit closer consideration by both those using and further developing these methods.
+
+### Defining occupancy and detection
+
+Building and interpreting DOMs requires a firm understanding of what probabilities of occupancy and detection conceptually represent for a given dataset.
+However, arriving at appropriate definitions for these terms can be a challenge.
+What site occupancy describes is dependent on a variety of factors, including the spatial extent of sites, the duration of primary occasions, the survey methods used, and the ways that each of these relate to the life history traits of the target species [@valente2024].
+Applications of DOMs in our review used data that varied across each of these axes, resulting in vastly different conceptions of occupancy and detection.
+
+The appropriate definition of occupancy is closely related to the spatial extent of each site, relative to a species’ home range within a primary occasion [@fig-occupancyVariants].
+This relationship determines whether the closure assumption – which states that the occupancy status of sites does not change within primary occasions such that the species is always available for detection at occupied sites – is fulfilled.
+If this is the case, and assuming that the survey method can detect the species even if it occurs in only a portion of the site, occupancy simply represents the probability that the species is present within a site during a primary occasion.
+Conversely, if the home range of species is not wholly contained within a site, occupancy must be interpreted as the probability that a site is *used* by the species for at least some portion of the primary occasion (@fig-occupancyVariants bii-iii).
+
+![An overview of how the size of a site relative to the focal species’ home range during a primary occasion can result in differing definitions of site occupancy.
+In (a), where a site represents a discrete habitat patch, occupancy typically represents whether a site is occupied during a primary occasion.
+If each site corresponds with an individual’s home range (a-i), as can occur with nest box surveys, occupancy is equivalent to abundance.
+If each site can contain multiple home ranges (a-ii) as with surveys of water bodies, this occupancy-abundance relationship is lost.
+In (b), where sites are defined in continuous habitat, occupancy definitions are more variable.
+If sites are larger than the focal species home range (b-i), occupancy will still represent whether a site is occupied – however, occupancy will not relate to abundance.
+Closure violations occur when individuals do not remain within a site for the full primary occasion (b-ii, b-iii).
+Here, occupancy represents whether a site is used during a primary occasion.
+If sites have a similar size to the species home range and home ranges do not overlap (b-ii), occupancy may retain some relationship with abundance.
+If sites are smaller than the home range (b-iii) as typically occurs with camera traps, the occupancy-abundance relationship is lost.](Figures/7_OccupancyVariation.png){#fig-occupancyVariants}
+
+The relationship between sites and home ranges also determines whether occupancy corresponds with abundance [@steenweg2018].
+In cases where each site corresponds with the home range of a single individual or pair, site occupancy may be fully equivalent to abundance.
+For example, this may be the case with many studies of spotted owls *(Strix occidentalis),* where sites typically include all possible owl territories in a region [@olson2005].
+Alternatively, where sites are larger than home ranges, occupancy will only approach abundance at low densities.
+Users of DOMs should be conservative when making inference on abundance, noting that it typically requires making strong assumptions beyond those required for estimating occupancy or site use.
+
+Earlier occupancy model research was often focused on conventional definitions of occupancy which require compliance with the closure assumption.
+This work emphasised how violations of this assumption can bias estimates of model parameters, leading to overestimates of occupancy when sites are not consistently occupied within primary occasions due to non-random mortality or movement of individuals [@otto2013; @valente2017].
+These authors provide guidance for study designs to ensure closure, as well as tests and analysis-based solutions for cases where closure is violated [@rota2009; @kendall2013; @mackenzie2017].
+
+More recently, authors have begun to consider closure violations less as a source of bias and more as a modifier to the definition of occupancy [@efford2012; @valente2024; @goldstein2024].
+When movement of individuals during primary occasions results in closure violations such that a species is only available for detection during a subset of sampling occasions, occupancy represents something more akin to site use: the probability that a species is present at a site *at some point* during a primary occasion.
+Many DOMs in our review would best fit this definition of occupancy: closure is often an unreliable assumption when modelling the mobile birds and mammals which dominate our sample, particularly when sites do not align with species home ranges.
+When data is collected with autonomous units like camera traps or acoustic monitors, small detection ranges make violation of closure very likely, necessitating a definition of occupancy which represents the probability that a species uses habitat that overlaps with the unit’s detection range during the primary occasion [@wood2022].
+
+Alternative definitions of occupancy have knock-on effects for detection probabilities with important implications for model building and interpretation.
+‘Detection’ is often described as the probability that a species is observed during a survey given that a site is occupied, depending largely on the ability of the observer to identify the species.
+However, in practice detection probabilities can include several other components – individual movement patterns can determine whether a species is even available for detection at occupied sites, behaviour alters the perceptibility of a species, and variation in abundance changes the number of individuals available for detection [@guillera-arroita2017a].
+Consider a study using acoustic monitors, where a species is considered detected if a call is identified within a one-day secondary occasion: the probability of detection would be contingent on the number of individuals in the area, whether any of them entered the recording range, whether they made any vocalisations, and whether human listeners or classifier software successfully identified the call.
+In this circumstance, detection would contain both ecological and observational components.
+
+An unusual challenge arises when users seek to fit to DOMs to detection/non-detection data that is continuously recorded – this may occur with camera traps or acoustic monitors which can run for long time periods, or with data from citizen science portals that lacks clearly delineated survey periods.
+In these situations, modellers must arbitrarily determine how to stratify detections into primary and secondary occasions (but note the availability of continuous detection models for secondary “occasions”; [@guillera-arroita2011; @emmet2021; @pautrel2024].
+This stratification affects the temporal scale at which occupancy is measured, with corresponding changes to the definitions of occupancy and detection.
+The appropriate duration of these occasions will of course depend on research objectives and the ecology of the target species [@chave2013], but guidelines for approaching this task in DOMs specifically would hold value for the increasing numbers of authors fitting models to this type of data.
+
+In all cases we echo recommendations made to users of occupancy-detection models more broadly, that they establish clear definitions for both occupancy and detection that correspond to the target species’ biology and the structure of the data used for modelling [@goldstein2024].
+While this is a productive exercise for clarifying internal assumptions and guiding choice of covariates, including these definitions in published articles also provides transparency and supports interpretation for readers less familiar with the study system.
+
+### Identifying covariates for each modelled parameter
+
+The definitions of occupancy and detection should strongly inform decisions on the covariates that are considered for inclusion in DOMs.
+Hypothesised relationships between occupancy dynamics and the environment should correspond with the spatial and temporal scale at which the data was collected, noting that the underlying mechanisms that drive occupancy at different timescales are likely to differ for many species.
+
+Our review found that many studies did not include any covariates on initial occupancy.
+However, unless a study is conducted at very small extents or study sites are truly uniform in their initial suitability, one would expect some amount of non-random variation in occupancy probability across any study system.
+As seen in our review, many DOMs are applied at large spatial scales across significant habitat gradients.
+Omission of the factors that drive occupancy biases spatial predictions in static SDMs [@barry2006], which the initial occupancy component of DOMs conceptually resembles.
+In tests of predictions using DOMs, @briscoe2021 posit that poor spatial predictive performance may have been the result of a constrained covariate pool for parameters including initial occupancy.
+Where spatial predictions are of interest, it is prudent for users of DOMs to consider sufficient covariates to capture the key drivers of occupancy in the first primary occasion – this may be less of a concern in cases where estimating trends in occupancy probability is a greater priority.
+
+When choosing covariates for colonisation and extinction users should prioritise the factors most likely to be of importance in driving change *between* primary occasions.
+DOMs are generally first-order Markov models where occupancy probabilities are linked only to the preceding primary occasions; therefore, covariates should have some link to possible mechanisms of transition at this timescale.
+Interestingly, fewer than 50% of environmental covariates considered for transition probabilities in our review were time-varying.
+While some static covariates provide useful information on habitat suitability in general, dynamic covariates are increasingly available and can provide more nuance on how species respond to fluctuations in environmental conditions.
+
+As previously mentioned, different definitions of detection probability can include aspects of species abundance, behaviour, and resource use which should be reflected in the choice of covariates considered for inclusion on this parameter.
+Most DOMs included in our review considered comparatively few (median = 3) covariates on detection, either environmental or structural.
+We encourage authors to consider fully how environmental factors might influence detection in their study system, and to avoid restricting these covariates to the occupancy dynamics parameters.
+Existing work on unmodelled heterogeneity in detection demonstrates a deleterious effect on estimates of occupancy probability in both SDMs and DOMs, highlighting the need for greater emphasis on this component of the model [@mcclintock2010; @lahoz-monfort2014].
+
+The consequences of unmodelled heterogeneity on model outputs generally remains an understudied aspect of DOMs, despite ongoing acknowledgements that this is an area meriting further research [@mackenzie2006; @mackenzie2017].
+Simulation studies exploring these scenarios would help to inform where this may be of concern and guide future research on model building practices.
+This includes explorations on how missing covariates on detection may influence estimates of other parameters in DOMs, particularly in cases where closure is violated and detection contains a large environmental component.
+
+### Complexity in DOMs
+
+One aspect of fitting DOMs meriting broader discussion centres on ‘model complexity,’ and how much must be incorporated into models to reliably estimate species occupancy under different contexts and use cases.
+Complexity is a broad term encompassing many aspects of a model, including the number of covariates included, the form of their relationship to response variables, and the overall structure of the model [@merow2014].
+Opinions on simplicity versus complexity in ecological models are diverse - where some advocate for the simplest possible models, arguing that they are most generalisable; others insist that overly simple models cannot adequately represent the most important drivers in a system [@evans2013; @lonergan2014].
+By their nature, DOMs are somewhat more complex than simpler models for studying occupancy due to their hierarchical structure, which is necessary to control for detectability and to capture occupancy dynamics.
+Within this structure, however, further complexity is up to the modeller: one can choose how many covariates to consider for inclusion on parameters, and how to represent the nature of the relationship between those covariates and parameters.
+In the applications of DOMs in our review, average complexity was somewhat low along both of these axes – relatively few covariates per parameter were considered for inclusion, and the majority of studies did not consider non-linear responses nor interactions between covariates.
+
+Research from correlative species distribution models (SDMs) indicates that allowing for more complex relationships can improve model performance in spatial predictions of occupancy [@valavi2023], an increasingly popular use-case for DOMs.
+Many common approaches for SDMs, such as MAXENT and Boosted Regression Trees, permit considerable flexibility in the shape of their covariate response curves and use of interactions, where supported by the data [@elith2008; @merow2013].
+Within DOMs, there are promising developments in methods for incorporating additional complexity, including @joseph2020 ’s novel neural-network occupancy model which allows for exponentially higher levels of complexity, potentially offering improved performance for prediction-oriented studies.
+While no studies in our samples used these approaches, other means for flexible response shapes such as splines, BRTs, or machine learning based methods may offer other avenues for increased flexibility [@hutchinson2011; @rushing2019; @joseph2020].
+Further research on these methods may provide new options where spatial predictive performance is a priority.
+
+The emphasis on more complex responses in SDMs may also be attributed to their frequent application across relatively large geographic extents which might encompass the full species niche, where environmental relationships may be expected to be non-linear.
+However, we found that many DOMs were also implemented across large spatial extents where non-linear responses might be expected.
+When this is probable, modellers should ensure that covariate responses are ecologically realistic and appropriately reflect the hypothesised relationships [@austin2007].
+Similar consideration should be given to interactions between covariates, given that these relationships are commonly expected based on ecological theory and that their exclusion can negatively impact model performance [@guisan2006].
+This is not to suggest that modellers should consider arbitrary non-linear responses or covariate interactions that are not grounded in ecological theory, but that they should remain open to more complexity where merited by the data and study system.
+While overfitting models can be a concern, many covariate selection methods (including AIC-based approaches and Bayesian regularisation) will penalise excessive complexity and appropriate model evaluation (currently rare – see below) can detect these issues.
+
+### Covariate Selection
+
+Methods used for covariate selection were particularly varied in reviewed DOMs, with a wide range of approaches used in both MLE and Bayesian frameworks and little consensus on the best techniques.
+Covariate selection is a particularly challenging task for DOMs relative to other common models for occupancy estimation, as testing alternative covariates across multiple parameters can result in a rapidly expanding pool of candidate models.
+Consider that while testing all combinations of 6 covariates for a simple correlative SDM would require fitting just 64 models, possible combinations of 6 covariates on each of the DOMs four parameters would require 24 million models, making exhaustive comparison computationally infeasible.
+Limited research has been conducted on the advantages of different methods for covariate selection in DOMs, and there is unlikely to be a one-size-fits-all approach which will be appropriate for all possible objectives.
+However, it is important to consider the implications of the different model selection approaches in common usage and their suitability for various use-cases.
+
+The few papers that have addressed covariate selection in occupancy models have largely focused on AIC-based methods, which aim to identify models that are useful for prediction by selecting for fit while penalising complexity [@chakrabarti2011].
+In a comparison of static occupancy models fit with three AIC-based approaches using simulated data, @doherty2012 found that while each method achieved similar predictive performance, estimates of covariate weights varied.
+They advise model averaging to mitigate this effect, but acknowledge that this effect is likely to be even larger in more complex models such as DOMs.
+More recent work by @morin2020 using occupancy models fit to field data had similar findings, with sequential model selection approaches often failing to identify the lowest AIC model amongst exhaustive combinations.
+They recommend that modellers carry more candidate models through each stage in sequential selection processes to increase the probability that the top models are identified, given that exhaustive model selection may not be feasible for DOMs with moderate-large numbers of potential covariates.
+
+The most common application of DOMs was to test hypothesised relationships.
+In recent years, work from the causal modelling community has critiqued aspects of model selection in cases where the principal research objective is to test pre-defined environmental relationships [@tredennick2021].
+@stewart2023 discuss this in the context of occupancy modelling, demonstrating the risks posed by certain ‘collider’ variables which have causal relationships with both occupancy and other covariates in the candidate set.
+Where these colliders are present, the top models selected by AIC may produce inaccurate estimates of focal covariates, even where they produce more accurate estimates of occupancy probabilities.
+As a result of these concerns, recent work has reinforced suggestions to limit or avoid covariate selection when inference on hypothesised relationships is the primary objective [@tredennick2021; @arif2022; @popovic2024; @bolker2024].
+These authors instead suggest a focus on more constrained model sets defined *a priori*, with careful consideration of the structural relationships between candidate covariates using tools such as directed acyclic graphs to clarify assumptions [@arif2023].
+These firmer views on causal inference are not universal, and authors including @nichols2024 have defended the use of predictive models (including DOMs) coupled with thoughtful and constrained model selection as useful tools for inference, while noting the importance of carefully considering relationships among covariates.
+
+Where inference on pre-specified hypotheses is not required, approaches to model selection may be more flexible.
+AIC-based model selection starting from a larger pool of candidate covariates is suitable for exploratory research, where the risk of bias in coefficient estimates and spurious correlations may be less of a concern [@tredennick2021].
+Prediction remains somewhat underexplored for DOMs, and further research is needed on the best techniques for fitting models for this purpose [@briscoe2021].
+While we discuss model evaluation in greater detail later in the discussion, model selection by cross-validation is a promising avenue where computationally feasible and is available via the R packages *ubms* and *unmarked* [@yates2023; @kellner2022; @kellner2023]*.*
+
+Regardless of objective, Bayesian model selection for DOMs remains somewhat underexplored.
+@hooten2015 ’s guide to Bayesian model selection in ecology remains a valuable resource for possible methods for fitting and evaluating these models, and a range of promising methods including regularisation priors and reversible-jump MCMC merit future research on their suitability for use with DOMs [@park2008].
+In one of the few comparative studies on model selection for Bayesian occupancy models, @stevens2019 found that models selected using the logarithmic scoring rule rather than WAIC or DIC produced better performing models for prediction.
+As the accessibility and popularity of Bayesian DOMs appears likely to continue to increase, further research and tool development in this space is necessary.
+
+### Model assessment
+
+Regardless of covariate selection protocol, it is important to determine whether a chosen model appropriately reflects the data and is suitable for its use case.
+However, current applications of DOMs report model assessment at rates lower than either single-season occupancy models or SDMs [@goldstein2024; @araújo2019].
+This is perhaps because model assessment methods and guidance remain underdeveloped for DOMs: few goodness-of-fit tests have been proposed, and those which do exist require more empirical assessment for greater acceptance.
+Currently, @mackenzie2004 ’s approach for single-season occupancy models using a parametric bootstrap has been extended to DOMs and is implemented in the *AICcModAvg* and *unmarked* R packages – @kéry2021 describe the test and present an alternative based on separately assessing fit to static and dynamic components of the model.
+In Bayesian implementations, posterior predictive checks offer means to assess model fit [@gelman2014].
+@broms2016 discusses various options for model assessment in Bayesian single-season multi-species occupancy models, with insights that may also be applicable to DOMs.
+While further tests and developments of these methods are necessary to solidify their use, current authors should to the best extent possible continue to conduct and report goodness-of-fit tests using existing tools regardless of their research objective.
+
+Studies interested in making spatial or temporal predictions require more substantial forms of model evaluation to quantify predictive performance [@araújo2005; @guisan2005].
+As with other hierarchical models, model evaluation for DOMs can be difficult and somewhat uncertain compared to other model types, as the primary response variable of interest (species occupancy) is a latent variable where the true state is generally not known.
+Predictive performance evaluation is thus typically based on *observed* occupancy data, where a DOM is used to compute the probability of observing the species and this is compared with the field data.
+This method confounds the occupancy and detection processes, restricting our ability to determine whether any variation is correctly apportioned to each of these processes.
+
+While predictive performance should ideally be tested with independent datasets, this is generally not practicable with DOMs.
+Best practice should involve setting aside a portion of data for testing final models, where this is not possible, cross-validation should be considered to attain an estimate of performance.
+Validation of DOM predictions can be done by splitting datasets temporally and using latter years for evaluation or by setting aside sites [@briscoe2021].
+Site-based cross validation is included in several packages including *unmarked* and *ubms* [@kellner2022; @kellner2023].
+
++------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------+
+| Topic                              | Recommendations                                                                                                                                                                                                             | Priorities                                                                                                                          |
++====================================+=============================================================================================================================================================================================================================+=====================================================================================================================================+
+| Defining occupancy and detection   | -   Explicitly state context-dependent definitions of occupancy and detection probabilities.
+| -   Develop guidelines for delineating primary occasions in continuous data.
+|
++------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------+
+| Covariates for model parameters    | -   Choose relevant covariates for colonization and extinction, aligned with the temporal scale of changes in occupancy.
+| -   Assess how missing drivers of heterogeneity on one parameter affect estimates of the other parameters and model predictions.
+|
+|                                    |                                                                                                                                                                                                                             |                                                                                                                                     |
+|                                    | -   Ensure that key drivers are also considered for initial occupancy, particularly when making spatial predictions.
+|                                                                                                                                     |
+|                                    |                                                                                                                                                                                                                             |                                                                                                                                     |
+|                                    | -   Ensure that environmental aspects of detection are reflected in candidate covariates.
+|                                                                                                                                     |
++------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------+
+| Complexity and covariate responses | -   Align covariate responses with hypothesised relationships, testing non-linear responses or interactions where appropriate.
+| -   Explore the best ways to include greater complexity in covariate responses, such as splines and machine learning based methods.
+|
+|                                    |                                                                                                                                                                                                                             | -   Assess influence of covariate quantity and response complexity on predictive performance and uncertainty.
+|
++------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------+
+| Model selection                    | -   For all objectives, ensure reported methods are comprehensive.
+| -   Identify best approaches to model selection for various objectives while limiting computational costs.
+|
+|                                    |                                                                                                                                                                                                                             |                                                                                                                                     |
+|                                    | -   When testing hypotheses, include clear rationale for covariate inclusions, such as by creating directed acyclic graphs.
+Some authors recommend constraining candidate sets or avoiding covariate selection altogether.
+| -   Compare performance of emerging Bayesian methods for model selection.
+|
+|                                    |                                                                                                                                                                                                                             |                                                                                                                                     |
+|                                    | -   Consider all models with similar support.
+|                                                                                                                                     |
+|                                    |                                                                                                                                                                                                                             |                                                                                                                                     |
+|                                    | -   Be aware that results of model selection are sensitive to implementation.
+When using sequential selection, test more models at each stage to avoid missing the best-performing model.
+|                                                                                                                                     |
++------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------+
+| Model assessment                   | -   Available goodness-of-fit tests should always be conducted and reported.
+| -   Existing goodness-of-fit tests should be further validated to ease concerns regarding their use.
+|
+|                                    |                                                                                                                                                                                                                             |                                                                                                                                     |
+|                                    | -   Data should be reserved for out-of-sample validation wherever possible, and especially when prediction is an objective.
+If not feasible, in-sample validation should be performed and performance metrics reported.
+| -   Additional research on methods for performance evaluation in hierarchical models is necessary                                   |
++------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------+
+
+: A summary of recommendations for current users of DOMs and priorities for further research.
 
@@
 
-Conclusions.
+In the two decades since the publication of @mackenzie2003, use of dynamic occupancy models has increased and they have been applied to a broad range of data types and ecological and applied problems.
+This expansion can pose challenges in navigating the model building processes and aligning data, model and interpretations to produce useful outputs.
+Many of our recommendations thus simply implore users to take stock at key points to clarify their assumptions and expectations.
+With frequent use of DOMs for important problems such as managing critically endangered species, guiding public health decisions and tracking harmful invasive species, understanding the sensitivity of model outputs to decisions made in the model fitting process becomes increasingly important.
+Major uncertainties remain in key areas of the model development process including model selection and evaluation, with much of the existing guidance based on the static occupancy modelling literature [@mackenzie2017].
+Several of the research priorities we identified relate to increasingly common use-cases – such as generating spatial and temporal predictions, and modelling data derived from novel detection techniques.
+Targeted research and guidance specific to DOMs will be needed to ensure that the potential of these tools can be realised, and they can reliably inform conservation and management decisions.

### 2025-03-04 Supplement/Code/Randomisation.qmd
Major changes to MS over the past two weeks prior to resubmission

--- 2024-09-23 Randomisation/Randomisation.qmd
@@
 subtitle: Describes random selection of papers for inclusion in the review
-author: Saoirse Kelleher
-date: today
+date: July 29 2024
 format: 
   html:
-    code-fold: true
     theme: minty
@@
 ```{r load WOS queries}
-keyTerms_datasheet <- read_xlsx("Queries/Final Queries/Keywords.xlsx",
+keyTerms_datasheet <- read_xlsx("Queries/Keywords.xlsx",
                                 skip = 10) |>
@@
 
-citesMackenzie_datasheet <- read_xlsx("Queries/Final Queries/MacKenzie_A.xlsx",
+citesMackenzie_datasheet <- read_xlsx("Queries/MacKenzie_A.xlsx",
                                       skip = 10) |>
-  rbind(read_xlsx("Queries/Final Queries/MacKenzie_B.xlsx",
+  rbind(read_xlsx("Queries/MacKenzie_B.xlsx",
                   skip = 10)) |>
@@
 ```{r Save sheet}
-write_csv(randomised_datasheet, "Randomisation/Final_Randomisation.csv")
+write_csv(randomised_datasheet, "Randomisation.csv")
 ```

### 2025-03-04 Supplement/Code/Supplement.qmd
Major changes to MS over the past two weeks prior to resubmission

---
title: "Supporting information"
date: February 28 2025
abstract: "Supporting information for the article \"Twenty years of dynamic occupancy models: a review of applications and look to the future.\" This is divided into two sections: \"Review methods\", with additional details on how articles were processed, and \"Additional information\", with further figures and analyses which were not included in the published article. Session information from the R session used to run this document is provided at the end."
abstract-title: "Summary"
format:
  html:
    number-sections: true
    number-depth: 3
    theme: minty
embed-resources: true
---

```{r Packages_and_data}
#| include: false

# Packages
library(tidyverse)
library(readxl)

# Data from review spreadsheet pages
sheet_ArticleData <- read_xlsx("Data/Review_Spreadsheet.xlsx",
                               sheet = "Article Data")
sheet_Objectives <- read_xlsx("Data/Review_Spreadsheet.xlsx",
                              sheet = "Objectives") 
sheet_FocalTaxa <- read_xlsx("Data/Review_Spreadsheet.xlsx",
                             sheet = "Focal Taxa") 
sheet_StudyArea <- read_xlsx("Data/Review_Spreadsheet.xlsx",
                             sheet = "Study Area")
sheet_DataCollection <- read_xlsx("Data/Review_Spreadsheet.xlsx",
                                  sheet = "Data Collection")
sheet_Covariates <- read_xlsx("Data/Review_Spreadsheet.xlsx",
                              sheet = "Covariates")
sheet_Modelling <- read_xlsx("Data/Review_Spreadsheet.xlsx",
                             sheet = "Modelling")
```

# Review methods

## Search queries

## Studies, datasets, and workflows

## Defining 'taxa'

## Covariates and parameters

# Additional information

## Authorship

### By country

### By objective

## Study locations and taxa

### By country

### Full taxa list

## Data collection

### Sampling method

### Site count by publication year

### Site count by study area scale

### Primary occasion count by secondary occasion count

### Primary occasion count by site count

### Dataset size (primary occasion count \* site count)

### Dataset size (primary occasion count \* secondary occasion count \* site count)

## Covariates

### Covariates table

### Covariate count by year

### Covariate count by site count

### Covariate count by primary occasion count

### Covariate count by study area scale

### Covariate count by dataset size (primary occasion count \* site count)

### Covariate count by objective

# Session information

```{r Session_Info}
#| echo: false
sessionInfo()
```
